### 2007-03-15 wikiplia/paper/wikiplia.tex
start paper

\documentclass{article}
\usepackage{amssymb}
\usepackage{fancyvrb}
\usepackage{url}

\begin{document}

\title{{\bf \huge Wikiplia}:\\
       The Free Programming Language\\
       That Anyone Can Edit}

\author{Tom Murphy VII}

\maketitle

\begin{abstract}
We present a new programming language called Wikiplia. We argue that
this language is Free to a greater extent and in more dimensions than
any existing language.

\end{abstract}

\nocite{alice}

{\small
\bibliographystyle{plain}
\bibliography{wikiplia}
}

\end{document}

### 2007-03-15 wikiplia/paper/wikiplia.tex
substantial start on doc

@@
-\documentclass{article}
+\documentclass[twocolumn]{article}
+\usepackage{fullpage}
 \usepackage{amssymb}
@@
 \usepackage{url}
+\usepackage{proof}
 
 \begin{document}
+
+\newcommand\wcite[1]{\footnote{\tiny Wikipedia, the free encyclopedia: {\it #1}; 2007}}
+\newcommand\comment[1]{}
+\newcommand\z{\ensuremath{\!}}
+
+\newcommand\D\Delta
+\newcommand\G\Gamma
+\newcommand\m\mapsto
+\newcommand\eval{{\sf eval\,\,}}
+\newcommand\tag[2]{{\tt{\mbox{\tt <}}{#1}{\mbox{\tt>}}}{#2}{\tt{\mbox{\tt</}}{#1}{\mbox{\tt>}}}}
+\newcommand\prim{{\,\,\sf prim}}
+\newcommand\rate{{\sf rate\,\,}}
 
@@
 \author{Tom Murphy VII}
+\date{1 April 2007}
 
@@
 \begin{abstract}
-We present a new programming language called Wikiplia. We argue that
-this language is Free to a greater extent and in more dimensions than
-any existing language.
-
+We present a new programming language called Wikiplia. The language
+has an unprecedented level of integration: The system is its own
+compiler, language definition, documentation, development environment,
+distributed filesystem, database, revision control system, software
+license, community message board, and World Wide Web home site.
+Wikiplia is designed to be Free to a greater extent and in more
+dimensions than existing languages.
 \end{abstract}
 
+\section{Introduction}
+
+One of the most cherished social principles of mankind is
+freedom,\z\wcite{Freedom} in its many incarnations. More recently,
+freedom has become an important principle in computer science as well,
+with the introduction of Free Software licenses such as the GNU
+GPL,\z\wcite{GNU General Public License} extensible markup languages
+such as XML,\z\wcite{XML} the ability expicitly deallocate memory
+with the {\tt free(3)} library call, and the widespread availibility of
+Free Herbal V1agara on the World Wide Web.\z\wcite{Sildenafil}
+
+The aim of this project is to develop a programming language that is
+as free as possible. We begin by enumerating freedoms that we desire
+to support. Because freedom is a possession of inestimable
+value\wcite{Cicero} we do not attempt to rank these freedoms; instead,
+each freedom is ``numbered'' using a symbol drawn from incomparable
+sets of glyphs.
+
+\paragraph{Freedom $\alpha$: The freedom to tinker.}
+Users should be able to study a program to see how it works, and to
+make modifications to suit his or her needs. For most software, this
+means that the programmer needs access to the software's
+documentation, source code, and UML\wcite{Unified Modeling Language}
+use case diagrams. This is traditionally achieved through licenses
+such as the GPL; however, as we will discuss in
+Section~\ref{sec:bootstrap} there are special considerations for
+bootstrapping compilers that render the GPL inadequate for this
+purpose.
+
+% freedom of speech?
+
+\paragraph{Freedom $\Game$: Freedom of beer.}
+Users should be able to write software without paying money to a
+licensing authority or certification program.
+
+\paragraph{Freedom $\surd$: Freedom of USA \#1.}
+Wikiplia is 100\% made in the USA and only available in
+English.\z\wcite{Freedom fries}
+
+% XXX more...
+
+\section{Reflections on booting boots} \label{sec:bootstrap}
+
+\section{Core calculus} \label{sec:calculus}
+
+\begin{figure}[ht]
+
+\comment{
+    List of exp list       <list> ... </list>
+  | String of string       <string>s</string>
+  | Int of IntInf.int      <int>i</int>
+  | Symbol of string       <symbol>s</symbol>
+  (* the only lazy expression *)
+  | Quote of exp           <quote>exp</quote>
+  (* user can't write these *)
+  | Prim of prim           <prim>p</prim>
+  | Closure of (string * exp) list * string * exp
+}
+\begin{center}
+\begin{tabular}{rcl}
+ {\em X} & ::= & {\tt<}{\em tag}{\tt >} $X_1\ X_2\ \ldots\ X_n$ {\tt</}{\em tag}{\tt >} \\
+         & $|$ & string
+\end{tabular}
+\end{center}
+\caption{Syntax of XML} \label{fig:xml}
+\end{figure}
+
+\begin{figure*}[ht]
+\[\begin{array}{cc}
+% basic, easy
+\infer{\G \vdash \eval \tag{string}{s} \m \tag{string}{s}}{} &
+\infer{\G \vdash \eval \tag{quote}{X} \m X}{} \\[1em]
+\infer{\G \vdash \eval \tag{int}{s} \m \tag{int}{s}}{} &
+\infer{\G \vdash \eval \tag{prim}{s} \m \tag{prim}{s}}{} \\[1em]
+\infer{\G \vdash \eval \tag{symbol}{s} \m \tag{prim}{s}}{s \prim} &
+\infer{\G \vdash \eval \tag{symbol}{s} \m X}{\G(s) = X} \\[1em]
+\multicolumn{2}{c}{\infer{\G \vdash \eval \tag{closure}{\G\ s\ X} \m \tag{closure}{\G\ s\ X}}{}} \\[1em]
+\multicolumn{2}{c}{\infer{\G \vdash \eval \tag{list}{X_1\ \ldots\ X_n} \m X'}%
+{\G\vdash \eval X_1 \m X_1' & \cdots & \G\vdash \eval X_n \m X_n' &
+ \G\vdash \rate X_1'\ \ldots\ X_n' \m X'
+}} \\[1em]
+% todo:
+% skip? abort, handle
+
+%
+\end{array}\]
+
+\caption{Evaluation of XML, part 1. The judgment $\G \vdash \eval X \m X'$
+indicates an assessment of the document $X$ with value $X'$. The
+judgment $\rate$ is an auxiliary assessment of a list of documents. It is defined in Figure~\ref{fig:xmlrate}.
+$\vec{X}$ is shorthand for a possibly empty sequence of XML documents.
+$\G$ is itself an XML document of the form
+% XXX is this right wrt <symbol>?
+$\tag{list}{\tag{symbol}{s_1}\ X_1\ \ldots\ \tag{symbol}{s_n}\ X_n}$.
+We take the judgment $\G(s) = X$ to produce the leftmost $X_i$ in $\G$
+such that $s_i$ is $s$. $\G, s = X$ is $\tag{list}{\tag{symbol}{s}\ X
+\vec{X}}$ if $\G$ is $\tag{list}{\vec{X}}$. 
+%
+} \label{fig:xmleval}
+\end{figure*}
+
+\begin{figure*}[htp]
+\[\begin{array}{cc}
+% evaluation of lists
+\multicolumn{2}{c}{\infer{\G \vdash \rate \tag{prim}{\sf list}\ \vec{X} \m \tag{list}{\vec{X}}}{}} \\[1em]
+\multicolumn{2}{c}{\infer{\G \vdash \rate \tag{prim}{\sf cons}\ X\ \tag{list}{\vec{X}} \m \tag{list}{X\ \vec{X}}}{}} \\[1em]
+\multicolumn{2}{c}{\infer{\G \vdash \rate \tag{prim}{\sf lambda}\ \tag{symbol}{s}\ X \m \tag{closure}{\G\ s\ X}}{}} \\[1em]
+\infer{\G \vdash \rate \tag{closure}{\G'\ s\ X}\ \vec{X} \m X'}{\G',s = \tag{list}{\vec{X}} \vdash \eval X \m X'} &
+\infer{\G \vdash \rate \tag{prim}{\sf eval}\ X \m X'}{\G \vdash \eval X \m X'} \\[1em]
+\multicolumn{2}{c}{\infer{\G \vdash \rate \tag{prim}{\sf xcase}\ \tag{list}{}\ X\ \vec{X} \m X'}{\G \vdash \eval X \m X'}} \\[1em]
+\multicolumn{2}{c}{\infer{
+\begin{array}{r@{}l}
+  \G \vdash \rate & \tag{prim}{\sf xcase}\ \tag{list}{X_h \vec{X_t}}\  X_0\ \\
+                  & \tag{list}{\tag{symbol}{s_h}\ \tag{symbol}{s_t}\ X_b}\ \vec{X} \m X' \\
+\end{array}
+}{\G, s_h = X_h, s_t = X_t \vdash \eval X_b \m X'}} \\[1em]
+\multicolumn{2}{c}{\infer{
+\begin{array}{r@{}l}
+  \G \vdash \rate & \tag{prim}{\sf xcase}\ \tag{quote}{X}\  X_0\ X_1\ \\
+                  & \tag{list}{\tag{symbol}{s_q}\ X_b}\ \vec{X} \m X' \\
+\end{array}
+}{\G, s_q = X \vdash \eval X_b \m X'}} \\[1em]
+\multicolumn{2}{c}{\infer{
+\begin{array}{r@{}l}
+  \G \vdash \rate & \tag{prim}{\sf xcase}\ \tag{string}{s}\  X_0\ X_1\ X_2\ 
+                    X\ \vec{X} \m X' \\
+\end{array}
+}{\G \vdash \eval X \m X'}} \\[1em]
+\multicolumn{2}{c}{\infer{
+\begin{array}{r@{}l}
+  \G \vdash \rate & \tag{prim}{\sf xcase}\ \tag{int}{s}\  X_0\ X_1\ X_2\ X_3\ 
+                    X\ \vec{X} \m X' \\
+\end{array}
+}{\G \vdash \eval X \m X'}} \\[1em]
+\multicolumn{2}{c}{\infer{
+\begin{array}{r@{}l}
+  \G \vdash \rate & \tag{prim}{\sf xcase}\ \tag{symbol}{s}\  X_0\ X_1\ X_2\ X_3\ X_4\ \\
+                  & \tag{list}{\tag{symbol}{s_s}\ X_b}\ \vec{X} \m X' \\
+\end{array}
+}{\G, s_s = \tag{string}{s} \vdash \eval X_b \m X'}} \\[1em]
+\multicolumn{2}{c}{\infer{
+\begin{array}{r@{}l}
+  \G \vdash \rate & \tag{prim}{\sf xcase}\ \tag{t}{\vec{X_t}}\  X_0\ X_1\ X_2\ X_3\ X_4\ X_5\ X \vec{X} \m X' \\
+\end{array}
+}{t = {\tt prim} \,\,or\,\, {\tt closure} & \G \vdash \eval X \m X'}} \\[1em]
+\multicolumn{2}{c}{\infer{\G \vdash \rate \tag{prim}{\sf quote}\ {X} \m \tag{list}{X}}{}} \\[1em]
+%
+\multicolumn{2}{c}{\infer{
+\G \vdash \rate \tag{prim}{\sf head}\ \tag{string}{s} \m X
+}{{\tt cvs\ checkout}\ s = X}} \\[1em]
+\multicolumn{2}{c}{\infer{
+\G \vdash \rate \tag{prim}{\sf read}\ \tag{string}{s} \tag{int}{i} \m X
+}{{\tt cvs\ checkout\ -\!r}\ i\ s = X}} \\[1em]
+\multicolumn{2}{c}{\infer{
+\G \vdash \rate \tag{prim}{\sf insert}\ \tag{string}{s} X \m \tag{int}{i}
+}{{\tt cvs\ commit}\ s\ X = i}} \\[1em]
+\multicolumn{2}{c}{\infer{
+\G \vdash \rate \tag{prim}{\sf history}\ \tag{string}{s} \m \tag{list}{\vec{X}}
+}{{\tt cvs\ log}\ s = \vec{X}}} \\[1em]
+
+\end{array}\]
+
+\caption{The definition of the {\sf rate} judgment. The rules for
+rating the primitives {\sf string}, {\sf sub}, {\sf substr}, {\sf +},
+{\sf -}, {\sf int}, {\sf eq} and {\sf if} are omitted for space. 
+%
+} \label{fig:xmlrate}
+\end{figure*}
+
+Wikiplia is built upon a core calculus of structured data with
+primitive revision control. Because we wish to support the freedom to
+tinker, the structured data take the form of XML (the {\em extensible}
+markup language; Figure~\ref{fig:xml}). Similar to the W3C's XML
+Validation,\z\wcite{XML schema} we allow the quality of an XML
+document to be assessed via a process called {\em evaluation}, whose
+output (if any) is itself an XML document. The rules for XML
+evaluation are simple and are given in
+Figures~\ref{fig:xmleval}~and~\ref{fig:xmlrate}.
+
+\comment{
   \nocite{alice}
@@
   }
+}
 

### 2007-03-15 wikiplia/paper/wikiplia.tex
a little bit more

@@
 such that $s_i$ is $s$. $\G, s = X$ is $\tag{list}{\tag{symbol}{s}\ X
-\vec{X}}$ if $\G$ is $\tag{list}{\vec{X}}$. 
+\vec{X}}$ if $\G$ is $\tag{list}{\vec{X}}$. The judgment $s\prim$ holds when
+$s$ is one of 
+{\sf insert},
+{\sf head},
+{\sf read},
+{\sf abort},
+{\sf lambda},
+{\sf list},
+{\sf cons},
+{\sf quote},
+{\sf string},
+{\sf xcase},
+{\sf size},
+{\sf sub},
+{\sf substr},
+{\sf handle},
+{\sf parse},
+{\sf eval},
+{\sf eq},
+{\sf +},
+{\sf -},
+{\sf int},
+{\sf history},
+{\sf let}, or
+{\sf if}.
 %
@@
 document to be assessed via a process called {\em evaluation}, whose
-output (if any) is itself an XML document. The rules for XML
+output (if any) is itself an XML document. Selected rules for XML
 evaluation are simple and are given in
-Figures~\ref{fig:xmleval}~and~\ref{fig:xmlrate}.
+Figures~\ref{fig:xmleval}~and~\ref{fig:xmlrate}.\z\footnote{XML
+documents can be self-correcting through the use of the {\sf handle}
+primitive, which detects an invalid document and proceeds along an
+alternative path. We omit the rules for this feature, which requires
+propagating invalid document status throughout evaluation and thus
+complicates the rules substantially.}
+
+Revision control is accessed through the class of {\tt cvs} judgments.
+We assume a single global repository, which maps {\em keys} (strings)
+to lists of {\em revisions}. A revision is a monotonically increasing
+and unique {\em revision number} (integer) paired with an XML
+document. The judgment ${\tt cvs commit}\ s\ X = i$ creates a new
+revision with revision number $i$ and data $X$ and inserts it in under
+the key $s$.
 

### 2007-03-16 trunk/wikiplia/paper/wikiplia.tex
syntax

--- 2007-03-15 wikiplia/paper/wikiplia.tex
@@
 \newcommand\rate{{\sf rate\,\,}}
+
+\newcommand\lb{\ensuremath{[\![}}
+\newcommand\rb{\ensuremath{]\!]}}
 
@@
 
-\caption{The definition of the {\sf rate} judgment. The rules for
-rating the primitives {\sf string}, {\sf sub}, {\sf substr}, {\sf +},
-{\sf -}, {\sf int}, {\sf eq} and {\sf if} are omitted for space. 
+\caption{Evaluation of XML, part 2. The {\sf rate} judgment assesses a
+sequence of XML expressions. The rules for rating the primitives {\sf
+string}, {\sf sub}, {\sf substr}, {\sf +}, {\sf -}, {\sf int}, {\sf
+eq} and {\sf if} are omitted for space.
 %
@@
 output (if any) is itself an XML document. Selected rules for XML
-evaluation are simple and are given in
+evaluation (the dynamic semantic markup) are given in
 Figures~\ref{fig:xmleval}~and~\ref{fig:xmlrate}.\z\footnote{XML
@@
 
-Revision control is accessed through the class of {\tt cvs} judgments.
-We assume a single global repository, which maps {\em keys} (strings)
-to lists of {\em revisions}. A revision is a monotonically increasing
-and unique {\em revision number} (integer) paired with an XML
-document. The judgment ${\tt cvs commit}\ s\ X = i$ creates a new
-revision with revision number $i$ and data $X$ and inserts it in under
-the key $s$.
+Revision control is accessed through the class of imperative {\tt cvs}
+judgments. We assume a single global repository, which maps {\em keys}
+(strings) to lists of {\em revisions}. A revision is a monotonically
+increasing and unique {\em revision number} (integer) paired with an
+XML document. The judgment ${\tt cvs commit}\ s\ X = i$ creates a new
+revision with revision number $i$ and data $X$ and inserts it under
+the key $s$. The judgment ${\tt cvs checkout}\ s = X$ fetches the most
+recent revision for the key $s$, the document $X$ (if no such key
+exists, then the document is invalid). The judgment ${\tt cvs\
+checkout\ -\!r}\ i\ s = X$ does the same, but for the specific
+revision number $i$ (if no such revision exists, the document is
+invalid). Finally, ${\tt cvs\ log}\ s = \vec{X}$ fetches all of the
+revision numbers for the key $s$, as a series of integers $\vec{X}$.
 
-\comment{
-  \nocite{alice}
+\subsection{Syntax}
 
-  {\small
-  \bibliographystyle{plain}
-  \bibliography{wikiplia}
-  }
-}
+\begin{figure}[b]
+\begin{center}
+\begin{tabular}{c}
+  \begin{tabular}{rcl}
+   {\em E} & ::= & {\tt(} $E_1\ E_2\ \ldots\ E_n$ {\tt)} \\
+           & $|$ & {\tt"}string{\tt"} \\
+           & $|$ & $n$ \\
+           & $|$ & symbol \\
+           & $|$ & {\tt'}E \\
+  \end{tabular} \\[3em]
+
+  \begin{tabular}{r@{\,\,=\,\,}l}
+   \lb {\tt(} $E_1\ \ldots\ E_n$ {\tt)}\rb & $\tag{list}{ \lb E_1 \rb\ \ldots\ \lb E_n \rb}$ \\
+   \lb {\tt"}$s${\tt"}\rb & $\tag{string}{s}$ \\
+   \lb $n$\rb & $\tag{int}{n}$ \\
+   \lb $sym$\rb & $\tag{symbol}{sym}$ \\
+   \lb {\tt'}$E$\rb & $\tag{quote}{E}$ \\
+  \end{tabular}
+
+\end{tabular}
+\end{center}
+\caption{The syntax for the XESP syntax. The recursively defined $\lb
+\cdot \rb$ operation converts an XESP expression into an XML
+document.} \label{fig:xesp}
+\end{figure}
+
+XML documents are universally parseable.\z\wcite{Parsing} However,
+they are difficult to write and read. Therefore, as usual\wcite{RELAX
+NG} we create a new syntax by which humans can write and read programs
+and which the computer automatically parses and converts to the easily
+parsable XML syntax and back to the new syntax, reducing
+complexity.\z\footnote{For efficiency, the Wikiplia implementation is
+optimized to lazily perform these translations, so that the document
+is never represented in XML form.} This compact syntax is based upon
+parentheses rather than tags: The XML document $\tag{list}{X_1\ X_2}$
+is instead written ${\tt(}X_1\ X_2{\tt)}$. Note that the closing
+parenthesis is not {\em named} as in XML, which makes parsing
+difficult because the computer must {\em guess} which parenthesis
+belongs to which other parenthesis. We therefore call this compact
+syntax XESP because it basically {\em reads the programmer's mind} to
+guess what the name of the closing parenthesis should be. The grammar
+for XESP is given in Figure~\ref{fig:xesp} along with the translation
+to XML documents.
 

### 2007-03-16 trunk/wikiplia/paper/wikiplia.tex
working

@@
 
-\section{Reflections on booting boots} \label{sec:bootstrap}
+\section{Reflections on strapping straps and booting boots} \label{sec:bootstrap}
+
+The hallmark of Free software is the GNU General Public License. It is
+a hereditary license that requires that (a) source code be distributed
+with the program and (b) modified versions of the program also be
+licensed under the GPL. The intention is that anyone receiving the
+software can exercise Freedom~$\alpha$ by understanding the source
+code and modifying it to suit his needs. Clearly any source code will
+not do: an obfuscated\wcite{Obfuscated code} version of the source
+code cannot be easily understood or modified, even though it is
+technically ``source code.'' The GPL therefore defines source code as
+the ``preferred form'' for ``making modifications.''
+
+Even source code in the preferred form might not be enough to achieve
+Freedom~$\alpha$, however. For instance, the program might be written
+in a mysterious programming language that only the author understands,
+and that programming language might only . It is therefore reasonable
+to construe the ``preferred form'' of the original software to include
+the implementation of the programming language that the software is
+written in. Because the programmer might need to fix bugs or extend
+the programming language implementation in order to modify the
+original program, he also needs the source code for that language as
+well. This code must be written in some language as well, so the
+process continues. It can end when one of the programming languages is
+generally well known enough that there are no practical barriers to
+understanding it or finding an implementation (examples would include
+C\wcite{C (programming language) and ALGOL 58\wcite{ALGOL 58}}, or so
+simple that the implementation is essentially non-existent ({\em e.g.}
+an assembler implemented directly in machine code).
+
+Another way for this process to terminate is for a programming
+language to be implemented in itself. This is known as a
+``bootstrapping compiler.'' A natural social tendency causes this to
+be very common: language implementors are more likely to enjoy the
+language they are implementing, and therefore more likely to choose it
+to implement the language. But when this process terminates this way,
+the reader might be left with a suspicious sense that nothing has
+actually been achieved.\z\footnote{In the case of a LOGO interpreter
+implemented in LOGO, we could say that this is then ``turtles all the
+way down.''} Specifically: What freedom-fulfilling use is the source
+code to an implementation of a mysterious programming language, if
+that source code is itself in the same mysterious programming
+language?
+
+Let us concentrate on a more concrete example. The GNU C
+compiler\z\wcite{GNU Compiler Collection} (licensed under the GPL) is
+an implementation of the C langauge with some extensions specific to
+the compiler. The GCC source code uses some of these extensions to HERE
+
+% parallel development yields same source, different behavior.
+% need globally atomic commits across all developers
+
+All popular modern languages are defined via a definitional
+interpreter\wcite{JavaScript}\wcite{Objective Caml}\wcite{Perl} with
+accompanying O'Reilly ``animal'' book.
 

### 2007-03-17 wikiplia/paper/wikiplia.tex
some work on paper

@@
 \usepackage{proof}
+\usepackage{code}
 
@@
 
-\section{Reflections on booting boots} \label{sec:bootstrap}
+\section{Reflections on strapping straps and booting boots} \label{sec:bootstrap}
+
+The hallmark of Free software is the GNU General Public License. It is
+a hereditary license that requires that (a) source code be distributed
+with the program and (b) modified versions of the program also be
+licensed under the GPL. The intention is that anyone receiving the
+software can exercise Freedom~$\alpha$ by understanding the source
+code and modifying it to suit his needs. Clearly any source code will
+not do: an obfuscated\wcite{Obfuscated code} version of the source
+code cannot be easily understood or modified, even though it is
+technically ``source code.'' The GPL therefore defines source code as
+the ``preferred form'' for ``making modifications.''
+
+Even source code in the preferred form might not be enough to achieve
+Freedom~$\alpha$, however. For instance, the program might be written
+in a mysterious programming language that only the author understands,
+and that programming language might only be implemented in a private
+compiler on the author's hard disk.\z\wcite{Hard disk} It is therefore
+reasonable to construe the ``preferred form'' of the original software
+to include the implementation of the programming language that the
+software is written in. Because the programmer might need to fix bugs
+or extend the programming language implementation in order to modify
+the original program, he also needs the source code for that language
+as well. This code must also be written in some language, so the
+process continues. It can end when one of the programming languages is
+generally well known enough that there are no practical barriers to
+understanding it or finding an implementation (examples would include
+C\wcite{C (programming language)} and ALGOL 58\wcite{ALGOL 58}), or so
+simple that the implementation is essentially non-existent ({\em e.g.}
+an assembler implemented directly in machine code).
+
+Another way for this process to terminate is for a programming
+language to be implemented in itself. This is known as a
+``bootstrapping compiler.'' A natural social tendency causes this to
+be very common: language implementors are more likely to enjoy the
+language they are implementing, and therefore more likely to choose it
+to implement the language. But when this process terminates this way,
+the reader might be left with a suspicious sense that nothing has
+actually been achieved.\z\footnote{In the case of a LOGO interpreter
+implemented in LOGO, we could say that this is then ``turtles all the
+way down.''} Specifically: What freedom-fulfilling use is the source
+code to an implementation of a mysterious programming language, if
+that source code is itself in the same mysterious programming
+language?
+
+Let us concentrate on a more concrete example. The GNU C
+compiler\z\wcite{GNU Compiler Collection} (licensed under the GPL) is
+an implementation of the C language with some extensions specific to
+the compiler. The GCC source code uses some of these extensions. Can
+the GCC be Free software if it requires the GCC to build? In the
+extreme case, what if someone were to add an extention to the GCC to
+enable a new C keyword, called {\tt compile\_a\_program}, and then
+replaced the entire source code with:
+
+\begin{code}
+int main (int argc, char ** argv) \{
+  compile_a_program;
+  return 0;
+\}
+\end{code}
+
+Such code is clearly worthless. Not all subversions of the source code
+via language extension may be so overt, but we claim that they
+nonetheless pose a substantial threat to freedom.
+
+We do not wish to limit the programmer's ability to make extensions to
+a language, since this would also toe-step Freedom~$\alpha$. We then
+conclude that the licensing terms must be expanded in order to provide
+more than ``source code.'' We propose that not only the source, but
+the source code's {\em history}, must be made available.
+
+\subsection{Revision control}
+
+Computer scientists use revision control\wcite{Revision control} to
+track changes to software and to coordinate development between
+multiple programmers. This has been true for thousands of years.
+Popular revision control systems such as CVS\wcite{Concurrent Versions
+System} and Subversion\wcite{Subversion (software)} allow for code
+to be concurrently modified by two or more developers and then have
+their changes integrated after the fact by an explicit ``check in''
+and conflict resolution phase.
+
+It may na\"ively seem that publishing the entire CVS history of a
+project would solve the issue with language extensions: By inspecting
+the revision that introduced the {\tt compile\_a\_program} feature
+(but prior to the replacement of the GCC with the minuscule version
+above), one could then see its implementation and then know how to
+HERE
+
+% parallel development yields same source, different behavior.
+% need globally atomic commits across all developers
+
+All popular modern languages are defined via a definitional
+interpreter\wcite{JavaScript}\wcite{Objective Caml}\wcite{Perl} with
+accompanying O'Reilly ``animal'' book.
 
@@
 
-Revision control is accessed through the class of {\tt cvs} judgments.
-We assume a single global repository, which maps {\em keys} (strings)
-to lists of {\em revisions}. A revision is a monotonically increasing
-and unique {\em revision number} (integer) paired with an XML
-document. The judgment ${\tt cvs commit}\ s\ X = i$ creates a new
-revision with revision number $i$ and data $X$ and inserts it in under
-the key $s$.
+Revision control is accessed through the class of imperative {\tt cvs}
+judgments. We assume a single global repository, which maps {\em keys}
+(strings) to lists of {\em revisions}. A revision is a monotonically
+increasing and unique {\em revision number} (integer) paired with an
+XML document. The judgment ${\tt cvs commit}\ s\ X = i$ creates a new
+revision with revision number $i$ and data $X$ and inserts it under
+the key $s$. The judgment ${\tt cvs checkout}\ s = X$ fetches the most
+recent revision for the key $s$, the document $X$ (if no such key
+exists, then the document is invalid). The judgment ${\tt cvs\
+checkout\ -\!r}\ i\ s = X$ does the same, but for the specific
+revision number $i$ (if no such revision exists, the document is
+invalid). Finally, ${\tt cvs\ log}\ s = \vec{X}$ fetches all of the
+revision numbers for the key $s$, as a series of integers $\vec{X}$.
 
-\comment{
-  \nocite{alice}
+\subsection{Syntax}
 
-  {\small
-  \bibliographystyle{plain}
-  \bibliography{wikiplia}
-  }
-}
+\begin{figure}[b]
+\begin{center}
+\begin{tabular}{c}
+  \begin{tabular}{rcl}
+   {\em E} & ::= & {\tt(} $E_1\ E_2\ \ldots\ E_n$ {\tt)} \\
+           & $|$ & {\tt"}string{\tt"} \\
+           & $|$ & $n$ \\
+           & $|$ & symbol \\
+           & $|$ & {\tt'}E \\
+  \end{tabular} \\[3em]
+
+  \begin{tabular}{r@{\,\,=\,\,}l}
+   \lb {\tt(} $E_1\ \ldots\ E_n$ {\tt)}\rb & $\tag{list}{ \lb E_1 \rb\ \ldots\ \lb E_n \rb}$ \\
+   \lb {\tt"}$s${\tt"}\rb & $\tag{string}{s}$ \\
+   \lb $n$\rb & $\tag{int}{n}$ \\
+   \lb $sym$\rb & $\tag{symbol}{sym}$ \\
+   \lb {\tt'}$E$\rb & $\tag{quote}{E}$ \\
+  \end{tabular}
+
+\end{tabular}
+\end{center}
+\caption{The syntax for the XESP syntax. The recursively defined $\lb
+\cdot \rb$ operation converts an XESP expression into an XML
+document.} \label{fig:xesp}
+\end{figure}
+
+XML documents are universally parseable.\z\wcite{Parsing} However,
+they are difficult to write and read. Therefore, as usual\wcite{RELAX
+NG} we create a new syntax by which humans can write and read
+documents and which the computer automatically parses and converts to
+the easily parsable XML syntax and back to the new syntax, reducing
+complexity.\z\footnote{For efficiency, the Wikiplia implementation is
+optimized to lazily perform these translations, so that the document
+is never represented in XML form.} This compact syntax is based upon
+parentheses rather than tags: The XML document $\tag{list}{X_1\ X_2}$
+is instead written ${\tt(}X_1\ X_2{\tt)}$. Note that the closing
+parenthesis is not {\em named} as in XML, which makes parsing
+difficult because the computer must {\em guess} which parenthesis
+belongs to which other parenthesis. We therefore call this compact
+syntax XESP because it basically {\em reads the programmer's
+mind}\wcite{Extra-sensory perception} to guess what the name of the
+closing parenthesis should be. The grammar for XESP is given in
+Figure~\ref{fig:xesp} along with the translation to XML documents.
+From now on, we use the XESP syntax in our examples.
 

### 2007-03-18 trunk/wikiplia/paper/wikiplia.tex
tiny bit

@@
 \usepackage{proof}
+\usepackage{code}
 
 \begin{document}
+
+\newcommand\ftinker{\ensuremath{\copyright}}
+\newcommand\fexpression{\ensuremath{\hbar}}
+\newcommand\fbeer{\ensuremath{\triangle}}
+\newcommand\fusa{\ensuremath{\sqrt[x^{1.2}]{\mbox{\t{oo}} + \frac{x}{z^2}}}}
 
@@
 
-\paragraph{Freedom $\alpha$: The freedom to tinker.}
+\paragraph{Freedom \ftinker: The freedom to tinker.}
 Users should be able to study a program to see how it works, and to
@@
 
+\paragraph{Freedom \fexpression: Freedom of expression.}
+Programmers should be able to write their programs using any
+expressions that they like. Specifically, there should be no
+prior establishment of arbitrary categories of expression that
+are excluded, such as those that discriminate on the basis
+of class, XXXXX, or type.
+
 % freedom of speech?
 
-\paragraph{Freedom $\Game$: Freedom of beer.}
+\paragraph{Freedom \fbeer: Freedom of beer.}
 Users should be able to write software without paying money to a
@@
 
-\paragraph{Freedom $\surd$: Freedom of USA \#1.}
+\paragraph{Freedom \fusa: Freedom of USA \#1.}
 Wikiplia is 100\% made in the USA and only available in
@@
 in a mysterious programming language that only the author understands,
-and that programming language might only . It is therefore reasonable
-to construe the ``preferred form'' of the original software to include
-the implementation of the programming language that the software is
-written in. Because the programmer might need to fix bugs or extend
-the programming language implementation in order to modify the
-original program, he also needs the source code for that language as
-well. This code must be written in some language as well, so the
+and that programming language might only be implemented in a private
+compiler on the author's hard disk.\z\wcite{Hard disk} It is therefore
+reasonable to construe the ``preferred form'' of the original software
+to include the implementation of the programming language that the
+software is written in. Because the programmer might need to fix bugs
+or extend the programming language implementation in order to modify
+the original program, he also needs the source code for that language
+as well. This code must also be written in some language, so the
 process continues. It can end when one of the programming languages is
@@
 understanding it or finding an implementation (examples would include
-C\wcite{C (programming language) and ALGOL 58\wcite{ALGOL 58}}, or so
+C\wcite{C (programming language)} and ALGOL 58\wcite{ALGOL 58}), or so
 simple that the implementation is essentially non-existent ({\em e.g.}
@@
 compiler\z\wcite{GNU Compiler Collection} (licensed under the GPL) is
-an implementation of the C langauge with some extensions specific to
-the compiler. The GCC source code uses some of these extensions to HERE
+an implementation of the C language with some extensions specific to
+the compiler. The GCC source code uses some of these extensions. Can
+the GCC be Free software if it requires the GCC to build? In the
+extreme case, what if someone were to add an extention to the GCC to
+enable a new C keyword, called {\tt compile\_a\_program}, and then
+replaced the entire source code with:
+
+\begin{code}
+int main (int argc, char ** argv) \{
+  compile_a_program;
+  return 0;
+\}
+\end{code}
+
+Such code is clearly worthless. Not all subversions of the source code
+via language extension may be so overt, but we claim that they
+nonetheless pose a substantial threat to freedom.
+
+We do not wish to limit the programmer's ability to make extensions to
+a language, since this would also toe-step Freedom~$\alpha$. We then
+conclude that the licensing terms must be expanded in order to provide
+more than ``source code.'' We propose that not only the source, but
+the source code's {\em history}, must be made available.
+
+\subsection{Revision control}
+
+Computer scientists use revision control\wcite{Revision control} to
+track changes to software and to coordinate development between
+multiple programmers. This has been true for thousands of years.
+Popular revision control systems such as CVS\wcite{Concurrent Versions
+System} and Subversion\wcite{Subversion (software)} allow for code
+to be concurrently modified by two or more developers and then have
+their changes integrated after the fact by an explicit ``check in''
+and conflict resolution phase.
+
+It may na\"ively seem that publishing the entire CVS history of a
+project would solve the issue with language extensions: By inspecting
+the revision that introduced the {\tt compile\_a\_program} feature
+(but prior to the replacement of the GCC with the minuscule version
+above), one could then see its implementation and then know how to
+HERE
 
@@
 they are difficult to write and read. Therefore, as usual\wcite{RELAX
-NG} we create a new syntax by which humans can write and read programs
-and which the computer automatically parses and converts to the easily
-parsable XML syntax and back to the new syntax, reducing
+NG} we create a new syntax by which humans can write and read
+documents and which the computer automatically parses and converts to
+the easily parsable XML syntax and back to the new syntax, reducing
 complexity.\z\footnote{For efficiency, the Wikiplia implementation is
@@
 belongs to which other parenthesis. We therefore call this compact
-syntax XESP because it basically {\em reads the programmer's mind} to
-guess what the name of the closing parenthesis should be. The grammar
-for XESP is given in Figure~\ref{fig:xesp} along with the translation
-to XML documents.
+syntax XESP because it basically {\em reads the programmer's
+mind}\wcite{Extra-sensory perception} to guess what the name of the
+closing parenthesis should be. The grammar for XESP is given in
+Figure~\ref{fig:xesp} along with the translation to XML documents.
+From now on, we use the XESP syntax in our examples.
 

### 2007-03-18 wikiplia/paper/wikiplia.tex
license and commentary in paper

@@
 \begin{document}
+
+% TODO: license
+\newcommand\currentrevision{465}
+
+\newcommand\ftinker{\ensuremath{\copyright}}
+\newcommand\fexpression{\ensuremath{\hbar}}
+\newcommand\fbeer{\ensuremath{\triangle}}
+\newcommand\fusa{\ensuremath{\sqrt[x^{1.2}]{\mbox{\t{oo}} + \frac{x}{z^2}}}}
+\newcommand\ffree{\ensuremath{\P}}
 
@@
 compiler, language definition, documentation, development environment,
-distributed filesystem, database, revision control system, software
-license, community message board, and World Wide Web home site.
-Wikiplia is designed to be Free to a greater extent and in more
-dimensions than existing languages.
+distributed filesystem, database, revision control system,
+bootstrapping software license, community message board, and World
+Wide Web home site. Wikiplia is designed to be Free to a greater
+extent and in more dimensions than existing languages.
 \end{abstract}
+
+\footnotetext[-1]{Copyright \copyright\ 2007 the Regents of the Wikiplia
+Foundation. Appears in SIGBOVIK 2007 with the permission of the
+Association for Computational Heresy; {\em IEEEEEE!} press,
+Verlag-Verlag volume no.~0x40-2A. This document may be distributed
+under the terms of the TIA Public License (Section~\ref{sec:license}).
+\pounds 0.00}
+
+\vspace{1em}
+{\noindent \small {\bf Keywords}:
+  Freedom, programming language, software license, wiki, XML, 
+  hyper-driven devices
+}
 
@@
 
+\paragraph{Freedom \fexpression: Freedom of expression.}
+Programmers should be able to write their programs using any
+expressions that they like. Specifically, there should be no
+prior establishment of arbitrary categories of expression that
+are excluded, such as those that discriminate on the basis
+of class, mathematical philosophy, or type.
+
 % freedom of speech?
 
-\paragraph{Freedom $\Game$: Freedom of beer.}
+\paragraph{Freedom \fbeer: Freedom of beer.}
 Users should be able to write software without paying money to a
@@
 
-\paragraph{Freedom $\surd$: Freedom of USA \#1.}
+\paragraph{Freedom \ffree: Freedom to redefine freedom.}
+Freedom should be free, so the definition of freedom should be free to
+change as the meaning of freedom changes. Wikiplia's license allows
+for Wikiplia to be distributed in a way that monotonically increases
+freedom as new concepts and catchphrases of freedom are invented.
+
+\paragraph{Freedom \fusa: Freedom of USA \#1.}
 Wikiplia is 100\% made in the USA and only available in
@@
 licensed under the GPL. The intention is that anyone receiving the
-software can exercise Freedom~$\alpha$ by understanding the source
+software can exercise Freedom~\ftinker by understanding the source
 code and modifying it to suit his needs. Clearly any source code will
@@
 Even source code in the preferred form might not be enough to achieve
-Freedom~$\alpha$, however. For instance, the program might be written
+Freedom~\ftinker, however. For instance, the program might be written
 in a mysterious programming language that only the author understands,
@@
 Let us concentrate on a more concrete example. The GNU C
-compiler\z\wcite{GNU Compiler Collection} (licensed under the GPL) is
+compiler\wcite{GNU Compiler Collection} (licensed under the GPL) is
 an implementation of the C language with some extensions specific to
@@
 the GCC be Free software if it requires the GCC to build? In the
-extreme case, what if someone were to add an extention to the GCC to
-enable a new C keyword, called {\tt compile\_a\_program}, and then
-replaced the entire source code with:
+extreme case, what if someone were to add an extension to the GCC to
+enable a new C keyword---called {\tt compile\_a\_program}---and then
+replace the entire source code with:
 
@@
 We do not wish to limit the programmer's ability to make extensions to
-a language, since this would also toe-step Freedom~$\alpha$. We then
+a language, since this would also toe-step Freedom~\ftinker. We then
 conclude that the licensing terms must be expanded in order to provide
@@
 
-It may na\"ively seem that publishing the entire CVS history of a
+It may na\"\i{}vely seem that publishing the entire CVS history of a
 project would solve the issue with language extensions: By inspecting
@@
 (but prior to the replacement of the GCC with the minuscule version
-above), one could then see its implementation and then know how to
-HERE
+above), one could then see its implementation and then know what it
+means. For certain patterns of development this does indeed suffice.
+However, programmers are not forced to check in their changes except
+at their own whims, as determined by social conventions; a programmer
+might make the private addition of the keyword {\tt
+compile\_a\_program}, then rewrite the GCC to use it, and only then
+check in this change as one revision. For this action he will surely
+be rebuked by his fellow programmers; none of the other developers can
+compile the new version of the code without access to the intermediate
+revision! This social pressure would also na\"\i{}vely seem to be enough
+to address the problem, but more insidious scenarios yet obtain.
 
-% parallel development yields same source, different behavior.
-% need globally atomic commits across all developers
+As a concrete example, suppose there are two programmers called K and
+R. Each is modifying the GCC with the purpose of adding a new
+character constant, \verb+'\c'+. K and R start at revision 100 of the
+GCC. K finds the case analysis for parsing character constants:
+%
+\begin{code}
+/* REVISION 100 (K) */
+switch(ch) \{
+  case 'n': return '\\n';
+  case 'r': return '\\r';
+  \vdots
+  default: abort("bad char constant");
+\}
+\end{code}
+%
+He adds a case for his extension, without using the extension, and
+checks this in as revision 101.
+%
+\begin{code}
+/* REVISION 101 (K) */
+switch(ch) \{
+  case 'n': return '\\n';
+  case 'r': return '\\r';
+  \vdots
+  case 'c': return 257;
+  default: abort("bad char constant");
+\}
+\end{code}
+%
+Meanwhile, R has similar (but not identical) inspiration and modifies
+his copy of the compiler:
+%
+\begin{code}
+/* REVISION 100 (R) */
+switch(ch) \{
+  case 'n': return '\\n';
+  case 'r': return '\\r';
+  \vdots
+  case 'c': return 8675309;
+  default: abort("bad char constant");
+\}
+\end{code}
+%
+He does not commit his code because he is wary of the time-consuming
+conflict resolution phase and is late for a date with K's estranged
+wife who is fed up with K's all-night hacking binges. He takes off in
+his 2007 Honda Civic\wcite{Honda Civic} with aftermarket spoiler for a
+night on the town, believing that a healthy well-rounded programmer
+spends more or less equal nights basking in the pale amber glow of the
+teletype as waking up with a few missing teeth naked and norovirused
+in some midtown alleyway with his wallet barely out of reach but empty
+anyway, having amply exercised Freedom~\fbeer.
 
-All popular modern languages are defined via a definitional
-interpreter\wcite{JavaScript}\wcite{Objective Caml}\wcite{Perl} with
-accompanying O'Reilly ``animal'' book.
+Meanwhile, K continues extending the GCC, using the extension to
+implement itself. He checks in this code with no conflicts:
+%
+\begin{code}
+/* REVISION 102 (K) */
+switch(ch) \{
+  case 'n': return '\\n';
+  case 'r': return '\\r';
+  \vdots
+  case 'c': return '\\c';
+  default: abort("bad char constant");
+\}
+\end{code}
+%
+K punches out at {\tt 1130 UTC},\z\wcite{Coordinated Universal Time}
+just as R returns from his adventure. R's confidence bolstered, he
+finishes his extension effort, following best practices and
+implementing the extension using itself:
+%
+\begin{code}
+/* REVISION 100 (R) */
+switch(ch) \{
+  case 'n': return '\\n';
+  case 'r': return '\\r';
+  \vdots
+  case 'c': return '\\c';
+  default: abort("bad char constant");
+\}
+\end{code}
+%
+He now decides to commit his changes (forgetting that he did not
+commit the intermediate revision). To do so he updates to the newest
+revision, 102, and sees that there are no conflicts---in fact,
+revision 102 already contains his changes! Believing that his changes
+are therefore compatible, he continues hacking.
+
+After this scenario, K and R believe they are working on the same
+programming language---after all, it has the same source code---but
+their minor bifurcation in development history means that they forever
+differing meanings of the \verb+'\c'+ extension. This mistake is
+likely to go unnoticed for some time, and until it is resolved, the
+meaning of the \verb+'\c'+ extension is firmly enslaved in the
+bipartite penitentiary of double entendre, yearning to be
+free\ldots\z\wcite{Information wants to be free}
+
+\subsection{Solution}
+
+Based on these scenarios we conclude that extant social measures such
+as revision control conventions are not enough to guarantee freedom.
+Even if we think these situations are implausible in the hands of
+well-intentioned, well-mannered software engineers, we wish for our
+software to remain free even when in the hands of nefarious and crafty
+factions who would seek to fracture\wcite{Fork (software development)}
+our free software community. We therefore need a technological and
+legal solution that forces the entire development history to be
+available. We call this technology and its license {\em Total
+Information Awareness} after the successful project of the US
+Government with the same name.\z\wcite{Information Awareness Office}
+
+Technologically, we develop our system around a primitive notion of
+revision control in which every change to the system is recorded.
+Because the system has an integrated editor, every action of a
+programmer is logged and preserved indefinitely with no extra action
+necessary on the programmer's part. Such commits are globally atomic,
+using a single universal repository. (This means that in the above
+scenario, R would not have been able to forget to commit his
+intermediate change, and would have been forced to observe his
+conflict with K.) To protect against the possibility that divergent
+development paths lead to incompatible compilers, we require that
+there is only one compiler for any given programming language, which
+itself exists in the revision control system. Therefore, it is always
+clear which version of the compiler was used to produce an executable
+from source. This also makes the K and R scenario impossible: there is
+only one compiler and so it is impossible for it to differ from
+itself.
+
+Legally, all of the software is licensed (Section~\ref{sec:license})
+under terms similar to the GNU GPL, but that define the ``source
+code'' to include the entire revision history of the system. In order
+to be compatible with Freedom~\ffree, we allow the license itself to be
+edited, but to ensure that no one can remove freedoms already present
+in the license, the license includes a provision that allows any {\em
+prior} version of the license to be used, at the programmer's option.
+
+\subsection{Implementation}
+
+Wikiplia, the free programming language that anyone can edit, is
+implemented as a web-site on the Internet at the address
+$${\tt http://wikiplia.spacebar.org/}$$
+
+\subsection{Roadmap}
+
+The remainder of this paper proceeds as follows. We first discuss in
+Section~\ref{sec:calculus} the design of the initial Wikiplia system,
+which is used to bootstrap the rest of Wikiplia. We then discuss the
+current state of Wikiplia as of revision \currentrevision\ in
+Section~\ref{sec:current}. We discuss the freedom-preserving TIA
+Public License in Section~\ref{sec:license}. We conclude with a
+discussion of unrelated work and plans for the future~\ref{sec:conc}.
 
@@
 }} \\[1em]
-% todo:
-% skip? abort, handle
+% skip: abort, handle
 
@@
 
+\subsection{Implementation}
+% SML: miserable non-free language
+
+\section{Revision \currentrevision} \label{sec:current}
+
+\section{TIA Public License} \label{sec:license}
+
+In this section we reproduce the Total Information Awareness Public
+License. Commentary is given via footnotes\wcite{footnote} into parts
+of the license.
+
+{\tt \parindent 0pt \parskip 10pt
+\verb+     +TIA PUBLIC LICENSE \\
+\verb+   +Revision 468, March 2007
+
+BEGIN INVINCIBLE SECTION 1
+
+This software is Copyright \copyright\ 2007{\large \rm--}$\infty$ The
+Regents of the Wikiplia Foundation. Permission is not granted to
+reproduce this software or license except by the terms explicitly
+enumerated below.
+
+I. Invincible Sections
+
+This license contains certain invincible sections, denoted by the text
+``BEGIN INVINCIBLE SECTION <n>'' and ``END INVINCIBLE SECTION <n>''.
+Such sections may not be modified under any circumstances.
+
+END INVINCIBLE SECTION 1\footnote{Invincible sections exist in order
+to ensure the sanctity of the license. We first establish that
+invincible sections will appear and that they are inviolable; this
+itself is done in an invincible section. Invincible sections cause a
+limited loss of liberty, but this is the cost of freedom.}
+
+BEGIN INVINCIBLE SECTION 2
+
+II. Version Idenfitication and Invalid Licenses
+
+This license must identify itself in the header as Revision <n> for
+some number <n>, which must be the same as the revision number in the
+Wikiplia repository for the key ``TPL'' in which the license text is
+stored. If this is not the case, then this version of the license is
+considered Invalid and Void.
+
+Permission is not granted to distribute the software or license using
+any Invalid version of the license.
+
+END INVINCIBLE SECTION 2\footnote{This invincible section establishes
+a connection between license versions and the actual contents of the
+Wikiplia repository. Note the self-reference: Though this text is in
+an invincible section and never changes, the referent of ``this
+license'' does change as the rest of the license is modified. Because
+Wikiplia assigns version numbers monotonically, this ensures that the
+next invincible section is able to guarantee that freedom is
+monotonic. In the case of an invalid license, no permissions
+whatsoever are granted, so the licensee {\em must} use a prior valid
+version of the license. The initial version of the license is valid.}
+
+BEGIN INVINCIBLE SECTION 3
+
+III. Option of License
+
+The licensee has the option to choose any revision of the license
+prior to (numerically less than) this version as the licensing
+terms for the software and license.
+
+END INVINCIBLE SECTION 3
+
+BEGIN INVINCIBLE SECTION 4
+
+IV. Heredity of License
+
+Any copy or derivative work of this software or license must be
+licensed under the TIA Public License.
+
+END INVINCIBLE SECTION 4\footnote{This clause makes the license
+``viral'' like the GNU GPL, so that freedom is preserved in all
+descendents of the software.}
+
+BEGIN INVINCIBLE SECTION 5
+
+V. Completeness of Copy
+
+This software and license may only be copied in their entirety,
+including the entire revision history.
+
+END INVINCIBLE SECTION 5\footnote{This section is the centerpiece of
+Wikiplia; it guarantees that the ``source code'' to any software or
+programming language derived from Wikiplia is free from the loopholes
+described in Section~\ref{sec:bootstrap} and so maximizes
+Freedom~\ftinker. Note that this does not limit the way that the
+software can be modified; the programmer might begin by blanking all
+of the keys he doesn't care about---as long as he preserves the fact
+that there was {\em once} something there.}
+
+VI. Freedom to edit
+
+Permission is hereby granted to edit this license.\footnote{The only
+non-invincible provision of the original license allows the reader to
+add provisions that he desires to the license. This guarantees
+Freedom~\ffree, the freedom to redefine freedom.
+%
+Note that even if a freedom-hater removes this provision from the
+license, the invincible sections above ensure that the freedom to edit
+the license is preserved for all time.}
+
+}
+
+\bigskip
+This license is bootstrapping in the sense that it grants only the
+minimal permissions necessary, after setting up invariants via the
+behavior-limiting invincible sections. In fact, the original version
+of the license does not directly permit the licensee to copy the
+software at all; he must first amend the license using provision {\tt
+VI} to give himself this permission.
+
+\section{Conclusion} \label{sec:conc}
+
+% All popular modern languages are defined via a definitional
+% interpreter\wcite{JavaScript}\wcite{Objective Caml}\wcite{Perl} with
+% accompanying O'Reilly ``animal'' book.
+
 \end{document}

### 2007-03-19 trunk/wikiplia/paper/wikiplia.tex
screenshot

@@
 \usepackage{code}
+\usepackage{graphicx}
 
 \begin{document}
+
+% TODO: license
+\newcommand\currentrevision{465}
 
@@
 \newcommand\fusa{\ensuremath{\sqrt[x^{1.2}]{\mbox{\t{oo}} + \frac{x}{z^2}}}}
+\newcommand\ffree{\ensuremath{\P}}
 
@@
 are excluded, such as those that discriminate on the basis
-of class, XXXXX, or type.
+of class, mathematical philosophy, or type.
 
@@
 licensing authority or certification program.
+
+\paragraph{Freedom \ffree: Freedom to redefine freedom.}
+Freedom should be free, so the definition of freedom should be free to
+change as the meaning of freedom changes. Wikiplia's license allows
+for Wikiplia to be distributed in a way that monotonically increases
+freedom as new concepts and catchphrases of freedom are invented.
 
@@
 (but prior to the replacement of the GCC with the minuscule version
-above), one could then see its implementation and then know how to
-HERE
+above), one could then see its implementation and then know what it
+means. For certain patterns of development this does indeed suffice.
+However, programmers are not forced to check in their changes except
+at their own whims, as determined by social conventions; a programmer
+might make the private addition of the keyword {\tt
+compile\_a\_program}, then rewrite the GCC to use it, and only then
+check in this change as one revision. For this action he will surely
+be rebuked by his fellow programmers; none of the other developers can
+compile the new version of the code without access to the intermediate
+revision! This social pressure would also na\"\i{}vely seem to be enough
+to address the problem, but more insidious scenarios yet obtain.
 
-% parallel development yields same source, different behavior.
-% need globally atomic commits across all developers
+As a concrete example, suppose there are two programmers called K and
+R. Each is modifying the GCC with the purpose of adding a new
+character constant, \verb+'\c'+. K and R start at revision 100 of the
+GCC. K finds the case analysis for parsing character constants:
+%
+\begin{code}
+/* REVISION 100 (K) */
+switch(ch) \{
+  case 'n': return '\\n';
+  case 'r': return '\\r';
+  \vdots
+  default: abort("bad char constant");
+\}
+\end{code}
+%
+He adds a case for his extension, without using the extension, and
+checks this in as revision 101.
+%
+\begin{code}
+/* REVISION 101 (K) */
+switch(ch) \{
+  case 'n': return '\\n';
+  case 'r': return '\\r';
+  \vdots
+  case 'c': return 257;
+  default: abort("bad char constant");
+\}
+\end{code}
+%
+Meanwhile, R has similar (but not identical) inspiration and modifies
+his copy of the compiler:
+%
+\begin{code}
+/* REVISION 100 (R) */
+switch(ch) \{
+  case 'n': return '\\n';
+  case 'r': return '\\r';
+  \vdots
+  case 'c': return 8675309;
+  default: abort("bad char constant");
+\}
+\end{code}
+%
+He does not commit his code because he is wary of the time-consuming
+conflict resolution phase and is late for a date with K's estranged
+wife who is fed up with K's all-night hacking binges. He takes off in
+his 2007 Honda Civic\wcite{Honda Civic} with aftermarket spoiler for a
+night on the town, believing that a healthy well-rounded programmer
+spends more or less equal nights basking in the pale amber glow of the
+teletype as waking up with a few missing teeth naked and norovirused
+in some midtown alleyway with his wallet barely out of reach but empty
+anyway, having amply exercised Freedom~\fbeer.
 
-All popular modern languages are defined via a definitional
-interpreter\wcite{JavaScript}\wcite{Objective Caml}\wcite{Perl} with
-accompanying O'Reilly ``animal'' book.
+Meanwhile, K continues extending the GCC, using the extension to
+implement itself. He checks in this code with no conflicts:
+%
+\begin{code}
+/* REVISION 102 (K) */
+switch(ch) \{
+  case 'n': return '\\n';
+  case 'r': return '\\r';
+  \vdots
+  case 'c': return '\\c';
+  default: abort("bad char constant");
+\}
+\end{code}
+%
+K punches out at {\tt 1130 UTC},\z\wcite{Coordinated Universal Time}
+just as R returns from his adventure. R's confidence bolstered, he
+finishes his extension effort, following best practices and
+implementing the extension using itself:
+%
+\begin{code}
+/* REVISION 100 (R) */
+switch(ch) \{
+  case 'n': return '\\n';
+  case 'r': return '\\r';
+  \vdots
+  case 'c': return '\\c';
+  default: abort("bad char constant");
+\}
+\end{code}
+%
+He now decides to commit his changes (forgetting that he did not
+commit the intermediate revision). To do so he updates to the newest
+revision, 102, and sees that there are no conflicts---in fact,
+revision 102 already contains his changes! Believing that his changes
+are therefore compatible, he continues hacking.
+
+After this scenario, K and R believe they are working on the same
+programming language---after all, it has the same source code---but
+their minor bifurcation in development history means that they forever
+differing meanings of the \verb+'\c'+ extension. This mistake is
+likely to go unnoticed for some time, and until it is resolved, the
+meaning of the \verb+'\c'+ extension is firmly enslaved in the
+bipartite penitentiary of double entendre, yearning to be
+free\ldots\z\wcite{Information wants to be free}
+
+\subsection{Solution}
+
+Based on these scenarios we conclude that extant social measures such
+as revision control conventions are not enough to guarantee freedom.
+Even if we think these situations are implausible in the hands of
+well-intentioned, well-mannered software engineers, we wish for our
+software to remain free even when in the hands of nefarious and crafty
+factions who would seek to fracture\wcite{Fork (software development)}
+our free software community. We therefore need a technological and
+legal solution that forces the entire development history to be
+available. Keeping with Freedom~\fusa, we call this technology and its
+license {\em Total Information Awareness} after the successful project
+of the US Government with the same name.\z\wcite{Information Awareness
+Office}
+
+Technologically, we develop our system around a primitive notion of
+revision control in which every change to the system is recorded.
+Because the system has an integrated editor, every action of a
+programmer is logged and preserved indefinitely with no extra action
+necessary on the programmer's part. Such commits are globally atomic,
+using a single universal repository. (This means that in the above
+scenario, R would not have been able to forget to commit his
+intermediate change, and would have been forced to observe his
+conflict with K.) To protect against the possibility that divergent
+development paths lead to incompatible compilers, we require that
+there is only one compiler for any given programming language, which
+itself exists in the revision control system. Therefore, it is always
+clear which version of the compiler was used to produce an executable
+from source. This also makes the K and R scenario impossible: there is
+only one compiler and so it is impossible for it to differ from
+itself.
+
+Legally, all of the software is licensed (Section~\ref{sec:license})
+under terms similar to the GNU GPL, but that define the ``source
+code'' to include the entire revision history of the system. In order
+to be compatible with Freedom~\ffree, we allow the license itself to be
+edited, but to ensure that no one can remove freedoms already present
+in the license, the license includes a provision that allows any {\em
+prior} version of the license to be used, at the programmer's option.
+
+\subsection{Implementation}
+
+Wikiplia, the free programming language that anyone can edit, is
+implemented as a web-site on the Internet at the address
+$${\tt http://wikiplia.spacebar.org/}$$
+
+\subsection{Roadmap}
+
+The remainder of this paper proceeds as follows. We first discuss in
+Section~\ref{sec:calculus} the design of the initial Wikiplia system,
+which is used to bootstrap the rest of Wikiplia. We then discuss the
+current state of Wikiplia as of revision \currentrevision\ in
+Section~\ref{sec:current}. We discuss the freedom-preserving TIA
+Public License in Section~\ref{sec:license}. We conclude with a
+discussion of unrelated work and plans for the future~\ref{sec:conc}.
 
@@
 increasing and unique {\em revision number} (integer) paired with an
-XML document. The judgment ${\tt cvs commit}\ s\ X = i$ creates a new
+XML document. The judgment ${\tt cvs\ commit}\ s\ X = i$ creates a new
 revision with revision number $i$ and data $X$ and inserts it under
-the key $s$. The judgment ${\tt cvs checkout}\ s = X$ fetches the most
+the key $s$. The judgment ${\tt cvs\ checkout}\ s = X$ fetches the most
 recent revision for the key $s$, the document $X$ (if no such key
@@
 
-\begin{figure}[b]
+\begin{figure}[htb]
 \begin{center}
@@
 syntax XESP because it basically {\em reads the programmer's
-mind}\wcite{Extra-sensory perception} to guess what the name of the
+mind}\,\wcite{Extra-sensory perception} to guess what the name of the
 closing parenthesis should be. The grammar for XESP is given in
@@
 
+\subsection{Implementation}
+% SML: miserable non-free language
+
+Wikiplia is implemented as a World Wide Web Home-Site, which allows
+for easy access from any location.
+
+\section{Revision \currentrevision} \label{sec:current}
+
+\begin{figure*}[t]
+\begin{center}
+\includegraphics[width=0.92\textwidth]{ss_homepage}
+\end{center}
+\caption{Screenshot of the Wikiplia home page as of revision
+\currentrevision.} \label{fig:sshomepage}
+\end{figure*}
+
+\section{TIA Public License} \label{sec:license}
+
+In this section we reproduce the Total Information Awareness Public
+License. Commentary is given via footnotes\wcite{footnote} into parts
+of the license.
+
+{\tt \parindent 0pt \parskip 10pt
+\verb+     +TIA PUBLIC LICENSE \\
+\verb+   +Revision 468, March 2007
+
+BEGIN INVINCIBLE SECTION 1
+
+This software is Copyright \copyright\ 2007{\large \rm--}$\infty$ The
+Regents of the Wikiplia Foundation. Permission is not granted to
+reproduce this software or license except by the terms explicitly
+enumerated below.
+
+I. Invincible Sections
+
+This license contains certain invincible sections, denoted by the text
+``BEGIN INVINCIBLE SECTION <n>'' and ``END INVINCIBLE SECTION <n>''.
+Such sections may not be modified under any circumstances.
+
+END INVINCIBLE SECTION 1\footnote{Invincible sections exist in order
+to ensure the sanctity of the license. We first establish that
+invincible sections will appear and that they are inviolable; this
+itself is done in an invincible section. Invincible sections cause a
+limited loss of liberty, but this is the cost of freedom.}
+
+BEGIN INVINCIBLE SECTION 2
+
+II. Version Idenfitication and Invalid Licenses
+
+This license must identify itself in the header as Revision <n> for
+some number <n>, which must be the same as the revision number in the
+Wikiplia repository for the key ``TPL'' in which the license text is
+stored. If this is not the case, then this version of the license is
+considered Invalid and Void.
+
+Permission is not granted to distribute the software or license using
+any Invalid version of the license.
+
+END INVINCIBLE SECTION 2\footnote{This invincible section establishes
+a connection between license versions and the actual contents of the
+Wikiplia repository. Note the self-reference: Though this text is in
+an invincible section and never changes, the referent of ``this
+license'' does change as the rest of the license is modified. Because
+Wikiplia assigns version numbers monotonically, this ensures that the
+next invincible section is able to guarantee that freedom is
+monotonic. In the case of an invalid license, no permissions
+whatsoever are granted, so the licensee {\em must} use a prior valid
+version of the license. The initial version of the license is valid.}
+
+BEGIN INVINCIBLE SECTION 3
+
+III. Option of License
+
+The licensee has the option to choose any revision of the license
+prior to (numerically less than) this version as the licensing
+terms for the software and license.
+
+END INVINCIBLE SECTION 3
+
+BEGIN INVINCIBLE SECTION 4
+
+IV. Heredity of License
+
+Any copy or derivative work of this software or license must be
+licensed under the TIA Public License.
+
+END INVINCIBLE SECTION 4\footnote{This clause makes the license
+``viral'' like the GNU GPL, so that freedom is preserved in all
+descendents of the software.}
+
+BEGIN INVINCIBLE SECTION 5
+
+V. Completeness of Copy
+
+This software and license may only be copied in their entirety,
+including the entire revision history.
+
+END INVINCIBLE SECTION 5\footnote{This section is the centerpiece of
+Wikiplia; it guarantees that the ``source code'' to any software or
+programming language derived from Wikiplia is free from the loopholes
+described in Section~\ref{sec:bootstrap} and so maximizes
+Freedom~\ftinker. Note that this does not limit the way that the
+software can be modified; the programmer might begin by blanking all
+of the keys he doesn't care about---as long as he preserves the fact
+that there was {\em once} something there.}
+
+VI. Freedom to edit
+
+Permission is hereby granted to edit this license.\footnote{The only
+non-invincible provision of the original license allows the reader to
+add provisions that he desires to the license. This guarantees
+Freedom~\ffree, the freedom to redefine freedom.
+%
+Note that even if a freedom-hater removes this provision from the
+license, the invincible sections above ensure that the freedom to edit
+the license is preserved for all time.}
+
+}
+
+\bigskip
+This license is bootstrapping in the sense that it grants only the
+minimal permissions necessary, after setting up invariants via the
+behavior-limiting invincible sections. In fact, the original version
+of the license does not directly permit the licensee to copy the
+software at all; he must first amend the license using provision {\tt
+VI} to give himself this permission.
+
+\section{Conclusion} \label{sec:conc}
+
+% All popular modern languages are defined via a definitional
+% interpreter\wcite{JavaScript}\wcite{Objective Caml}\wcite{Perl} with
+% accompanying O'Reilly ``animal'' book.
+
 \end{document}

### 2007-03-19 trunk/wikiplia/paper/wikiplia.tex
almost there...

@@
 \newcommand\ffree{\ensuremath{\P}}
+\newcommand\ftobe{{\tiny \ensuremath{\stackrel{\bigcirc}{\amalg}}}}
 
@@
   Freedom, programming language, software license, wiki, XML, 
-  hyper-driven devices
+  weatherproof footwear and their fastening mechanisms,
+  hyper-driven devices,
 }
@@
 of class, mathematical philosophy, or type.
+
+\paragraph{Freedom \ftobe: Free to be You and Me.}
+The development of a programming language should not be confined to
+the bearded academic elite, gazing down upon the programmer fiefs from
+their stratospheric ivory towers. Wikiplia is the free programming
+language that {\em anyone} can edit: from bearded academic
+elite\wcite{Tenure} to congressional staffers\wcite{Congressional
+staffer edits to Wikipedia} to nameless Slashdot\wcite{Slashdot}
+trolls. Wikiplia's WWW-based approach means that an Internet
+connection and compatible WWW hypertext browser is all that's needed
+to begin on the life-changing journey of programming language design.
 
@@
 sequence of XML expressions. The rules for rating the primitives {\sf
-string}, {\sf sub}, {\sf substr}, {\sf +}, {\sf -}, {\sf int}, {\sf
-eq} and {\sf if} are omitted for space.
+parse}, {\sf string}, {\sf sub}, {\sf substr}, {\sf +}, {\sf -}, {\sf
+int}, {\sf eq} and {\sf if} are omitted for space.
 %
@@
 
-\section{Revision \currentrevision} \label{sec:current}
+The system is implemented as a Standard~ML\wcite{Standard ML} program
+of approximately 1,000 lines.\z\footnote{This count does not include
+general purpose libraries, such as a networking library.} This program
+is designed to be minimal: it consists of a web server, a revision
+control system, and facilities for evaluating XESP documents. It also
+contains a very minimal bootstrapping ``compiler'' for XESP documents,
+with its boots manually strapped. From this tiny core we then develop
+the remainder of Wikiplia using Wikiplia itself.
+
+Some may balk at the choice of Standard~ML, as the language is
+miserably non-free: First, while many of the major implementations are
+GNU or BSD-licensed,\z\wcite{BSD licenses} all are implemented in
+Standard~ML itself, yielding the bootstrapping problem described
+earlier. Second, the mathematical Definition of Standard ML book is
+only available as a copyrighted publication of MIT
+Press,\z\wcite{MIT Press} not even {\em ostensibly} in a free
+manner. However, the performance considerations of the server and
+evaluator---and the lack of suitable free alternatives---force us to
+settle for such subjugation.
+
+\subsubsection{Web Server}
+
+The web server's job is simple. It runs in a loop, accepting a single
+connection, setting up an initial environment $\G$ for evaluation
+(which associates the symbols {\sf request.url}, {\sf request.ip} and
+{\sf request.time} with the appropriate values), and then evaluating
+the document contained at the head of the key ``main'' in the
+repository. The result of that evaluation is sent back to the web
+browser as a string.\z\footnote{The HTTP result code may be modified
+if, for example, the resulting document is a HTTP redirect to another
+URL.} The server knows nothing else about how Wikiplia works.
+
+The web server is single-threaded (so each request must finish
+before the next is handled) because we desire global atomicity
+(Section~\ref{sec:bootstrap}).
+
+\subsubsection{Revision Control}
+
+The revision control system stores XESP documents and their history.
+This is mostly a straightforward implementation of the imperative {\tt
+cvs} judgments given in Section~\ref{sec:calculus}. Unlike typical
+revision control systems, we need to support very large numbers of
+revisions with small edits (since every change is saved), so the
+implementation is engineered to make storing small revisions very
+cheap. Particularly, revisions are aggressively compressed by only
+storing the newest revision directly, and then a series of difference
+``plans'' that describe how to get the next older revision from the
+current one. We compute optimal plans using an efficient minimal edit
+distance\wcite{Levenshtein distance} calculation at the
+token\wcite{Token (parser)} level. As of revision~\currentrevision,
+the database is only 236 kilobytes.\z\wcite{Kilobyte}
+
+\subsubsection{Document Evaluation}
+
+The Wikiplia implementation also has facilities for parsing and
+evaluating XESP documents. These are a direct implementation of the
+evaluation rules in Figures~\ref{fig:xmleval}~and~\ref{fig:xmlrate}.
+
+\subsubsection{Bootstrapping}
+
+The key ``main'' of Wikiplia is responsible for decomposing the URL
+and acting upon it however is appropriate. The goal of the
+bootstrapping process is to make Wikiplia
+self-sufficient\wcite{Self-sufficiency} in the sense that the language
+can be edited from the web site implemented by ``main.'' To do so, the
+web site needs to be able to present the user with an edit box
+containing the current source of the ``main'' key and the ability to
+save his code into the database, overwriting the ``main'' key, in
+order to add functionality.
+
+The initial implementation of ``main'' is provides for the simple
+ability to edit, save, and view the current version of keys in the
+repository. These three actions are encoded as the URLs {\tt
+/edit/key}, {\tt /save/key}, and {\tt /view/\_/key}. The view action
+is straightforward. The edit action displays an HTML
+textarea\wcite{Text box} containing the current value of the key and a
+button that submits the changes to the {\tt save} url. The save action
+is the most complex. First, the submitted document is saved as the new
+version of the key. Then, if the key is of the form {\tt
+base.extension}, the database is checked to see if there is a key
+called {\tt extension:compile}. If so, the XESP document that is there
+is applied to the input document (a string) to produce a document that
+is saved at the key {\tt base}. This allows us to develop languages
+that are automatically compiled when saved.
+
+\begin{figure}
+\begin{center}
+\verb+(lambda 's '(parse (xcase s 'no '(h _ h))))+
+\end{center}
+\caption{The initial bootstrapping compiler.} \label{fig:b:compile.b}
+\end{figure}
+
+This initial functionality is implemented directly in the XESP
+language, whose extension is {\tt b}; the bootstrapping ``compiler''
+is just the built-in parser (Figure~\ref{fig:b:compile.b}).
+
+Arranging that the repository contain the correct keys to make this
+work is subtle. A small initialization phase sets up:
+\begin{itemize}
+\item {\tt b:compile.b}$\quad$ 
+      The source code (a string) in Figure~\ref{fig:b:compile.b}
+\item {\tt b:compile}$\quad$ 
+      The parsed document corresponding to the above, such that {\tt b:compile} applied to {\tt b:compile.b} yields {\tt b:compile}
+\item {\tt main.b}$\quad$ 
+      The source code of the original ``main'' program
+\item {\tt main}$\quad$
+      The parsed document corresponding to the above, such that {\tt b:compile} applied to {\tt main.b} yields {\tt main}.
+\end{itemize}
+
+We do this by {\em anticipating}, during the initialization process,
+the {\em meaning} of {\tt b:compile} so that we can perform that
+action (parsing) on {\tt b:compile.b} to produce {\tt b:compile}. This
+can only be achieved by fiat and this is the essence of bootstrapping.
+
+After this minimal initialization, we can then exclusively use the web
+interface to develop and extend Wikiplia.
 
@@
 \end{figure*}
+
+\begin{figure*}[t]
+\begin{center}
+% \includegraphics[width=0.92\textwidth]{ss_history}
+\end{center}
+\caption{Screenshot of the history information for {\tt main-go.w}.
+} \label{fig:sshistory}
+\end{figure*}
+
+\section{Revision \currentrevision} \label{sec:current}
+
+As of writing, Wikiplia is at revision~\currentrevision, and has
+a number of features implemented.
+
+\subsection{Interface}
+
+The editing interface implemented has been enhanced greatly;
+Figure~\ref{fig:sshomepage} shows a screenshot of the main page. The
+various views of a key are shown with a series of tabs at the top of
+each page. Each user has a home page named after his IP
+address,\z\wcite{IP address} which he can use to catalogue his
+interests. Various warnings help the user, for example, if he tries to
+edit a page that was generated by compiling some source code, a
+warning message suggests that he may wish to edit the source code
+instead. An ornate logo in SVG\wcite{Scalable Vector Graphics} adorns
+the page, and a sidebar provides quick access to the site's features.
+The logo and graphics for the site are stored in the repository; the
+new {\tt raw} and {\tt typed} actions allows access to these resources
+over HTTP so that they may be freely modified.\z\footnote{Because
+Wikiplia is dogmatically forward-looking, the graphics require SVG
+support in the browser and support for the {\tt data:} URL format, a
+combination only found in the newest versions of the Mozilla Firefox.
+Wikiplia is compatible with incompatible browsers, however, simply
+displaying a crapified version of each page.} At revision 228 support
+for metadata was added for each page; a {\tt history} tab now shows
+the date, revision number, IP address, and edit summary for each
+change to a key (Figure~\ref{fig:sshistory}).
+
+\subsection{Wikiplia language {\tt w}}
+
+Writing XESP documents by hand is very tedious, so one of the first
+orders of business was to develop a compiler for a new language, {\tt
+w}, which can be extended with convenient features. The first version
+of this language, created in revision 22, was written in XESP. It
+automatically quoted the appropriate arguments to the {\sf let}, {\sf
+it}, {\sf lambda}, {\sf xcase} primitives. In revision 361, the
+compiler was ported to the language {\tt w} as {\tt w:compile.w}, then
+compiled with the existing {\tt b}-compiled version of {\tt
+w:compile.b}, then recompiled with itself until reaching a fixed
+point.\z\wcite{Fixed point (mathematics)} After that, more features
+were added: a multi-argument function construct {\tt fn}; a
+list-deconstructing binding construct {\tt lets}; simple support for
+separately-compiled libraries\wcite{Library (computing)} via {\tt
+include}; support for mutually-recursive\wcite{Recursion (computer
+science)} bundles of functions via {\tt fun};\footnote{Prior to
+this, recursion had to be encoded directly by passing an initial
+``self'' argument to each function.} and the {\tt cond} keyword for
+series of chained ``if\ldots else'' conditionals. At each stage, the
+feature is implemented using the current version of {\tt w}, and then
+{\tt w:compile.w} is rewritten to use that convenient extension, and
+then recompiled until reaching a fixed point.
+
+\subsection{Wikiplia language {\tt page}}
+
+Wikiplia is not just a programming language; it needs facilities for
+editing pages that are human readable as well. This can be used to
+edit documentation for programming languages, to modify the home page
+to tout new developments, to modify the software license
+(Section~\ref{sec:license}) or to deface other user's home pages. For
+this we provide a Wiki-like\wcite{Wiki} syntax that allows for the
+authorship of such pages and easy linking between them. Like other
+Wikis, a link to a page that does not exist is colored red, to alert
+the user to the opportunity to stake out cyberspace real estate. This
+syntax is compiled to XESP documents via the {\tt page} language; the
+resulting documents are active in the sense that they check the status
+of linked pages to report the correct color on every page load.
 

### 2007-03-19 wikiplia/paper/wikiplia.tex
fix typo in license ;)

@@
 \usepackage{code}
+\usepackage{graphicx}
 
@@
 GPL,\z\wcite{GNU General Public License} extensible markup languages
-such as XML,\z\wcite{XML} the ability expicitly deallocate memory
-with the {\tt free(3)} library call, and the widespread availibility of
+such as XML,\z\wcite{XML} the ability explicitly deallocate memory
+with the {\tt free(3)} library call, and the widespread availability of
 Free Herbal V1agara on the World Wide Web.\z\wcite{Sildenafil}
@@
 conflict resolution phase and is late for a date with K's estranged
-wife who is fed up with K's all-night hacking binges. He takes off in
-his 2007 Honda Civic\wcite{Honda Civic} with aftermarket spoiler for a
-night on the town, believing that a healthy well-rounded programmer
-spends more or less equal nights basking in the pale amber glow of the
-teletype as waking up with a few missing teeth naked and norovirused
-in some midtown alleyway with his wallet barely out of reach but empty
-anyway, having amply exercised Freedom~\fbeer.
+wife who is fed up with K's all-night hacking binges. He burns rubber
+in his 2007 Honda Civic\wcite{Honda Civic} with aftermarket spoiler
+for a night on the town, believing that a healthy well-rounded
+programmer spends more or less equal nights basking in the pale amber
+glow of the teletype as waking up with a few missing teeth naked and
+norovirused in some midtown alleyway with his wallet barely out of
+reach but empty anyway, having amply exercised Freedom~\fbeer.
 
@@
 legal solution that forces the entire development history to be
-available. We call this technology and its license {\em Total
-Information Awareness} after the successful project of the US
-Government with the same name.\z\wcite{Information Awareness Office}
+available. Keeping with Freedom~\fusa, we call this technology and its
+license {\em Total Information Awareness} after the successful project
+of the US Government with the same name.\z\wcite{Information Awareness
+Office}
 
@@
 
+Wikiplia is implemented as a World Wide Web Home-Site, which allows
+for easy access from any location.
+
+The system is implemented as a Standard~ML\wcite{Standard ML} program
+of approximately 1,000 lines.\z\footnote{This count does not include
+general purpose libraries, such as a networking library.} This program
+is designed to be minimal: it consists of a web server, a revision
+control system, and facilities for evaluating XESP documents. It also
+contains a very minimal bootstrapping ``compiler'' for XESP documents,
+with its boots manually strapped. From this tiny core we then develop
+the remainder of Wikiplia using Wikiplia itself.
+
+Some may balk at the choice of Standard~ML, as the language is
+miserably non-free: First, while many of the major implementations are
+GNU or BSD-licensed,\z\wcite{BSD licenses} all are implemented in
+Standard~ML itself, yielding the bootstrapping problem described
+earlier. Second, the mathematical Definition of Standard ML book is
+only available as a copyrighted publication of MIT
+Press,\z\wcite{MIT Press} not even {\em ostensibly} in a free
+manner. However, the performance considerations of the server and
+evaluator---and the lack of suitable free alternatives---force us to
+settle for such subjugation.
+
+\subsubsection{Web Server}
+
+The web server's job is simple. It runs in a loop, accepting a single
+connection, setting up an initial environment $\G$ for evaluation
+(which associates the symbols {\sf request.url}, {\sf request.ip} and
+{\sf request.time} with the appropriate values), and then evaluating
+the document contained at the head of the key ``main'' in the
+repository. The result of that evaluation is sent back to the web
+browser as a string.\z\footnote{The HTTP result code may be modified
+if, for example, the resulting document is a HTTP redirect to another
+URL.} The server knows nothing else about how Wikiplia works.
+
+The web server is single-threaded (so each request must finish
+before the next is handled) because we desire global atomicity
+(Section~\ref{sec:bootstrap}).
+
+\subsubsection{Revision Control}
+
+The revision control system stores XESP documents and their history.
+This is mostly a straightforward implementation of the imperative {\tt
+cvs} judgments given in Section~\ref{sec:calculus}. Unlike typical
+revision control systems, we need to support very large numbers of
+revisions with small edits (since every change is saved), so the
+implementation is engineered to make storing small revisions very
+cheap. Particularly, revisions are aggressively compressed by only
+storing the newest revision directly, and then a series of difference
+``plans'' that describe how to get the next older revision from the
+current one. We compute optimal plans using an efficient minimal edit
+distance\wcite{Levenshtein distance} calculation at the
+token\wcite{Token (parser)} level. As of revision~\currentrevision,
+the database is only 236 kilobytes.\z\wcite{Kilobyte}
+
+\subsubsection{Document Evaluation}
+
+The Wikiplia implementation also has facilities for parsing and
+evaluating XESP documents. These are a direct implementation of the
+evaluation rules in Figures~\ref{fig:xmleval}~and~\ref{fig:xmlrate}.
+
+\subsubsection{Bootstrapping}
+
+The key ``main'' of Wikiplia is responsible for decomposing the URL
+and acting upon it however is appropriate. The goal of the
+bootstrapping process is to make Wikiplia
+self-sufficient\wcite{Self-sufficiency} in the sense that the language
+can be edited from the web site implemented by ``main.'' To do so, the
+web site needs to be able to present the user with an edit box
+containing the current source of the ``main'' key and the ability to
+save his code into the database, overwriting the ``main'' key, in
+order to add functionality.
+
+The initial implementation of ``main'' is provides for the simple
+ability to edit, save, and view the current version of keys in the
+repository. These three actions are encoded as the URLs {\tt
+/edit/key}, {\tt /save/key}, and {\tt /view/\_/key}. The view action
+is straightforward. The edit action displays an HTML
+textarea\wcite{Text box} containing the current value of the key and a
+button that submits the changes to the {\tt save} url. The save action
+is the most complex. First, the submitted document is saved as the new
+version of the key. Then, if the key is of the form {\tt
+base.extension}, the database is checked to see if there is a key
+called {\tt extension:compile}. If so, the XESP document that is there
+is applied to the input document (a string) to produce a document that
+is saved at the key {\tt base}. This allows us to develop languages
+that are automatically compiled when saved.
+
+\begin{figure}
+\begin{center}
+\verb+(lambda 's '(parse (xcase s 'no '(h _ h))))+
+\end{center}
+\caption{The initial bootstrapping compiler.} \label{fig:b:compile.b}
+\end{figure}
+
+This initial functionality is implemented directly in the XESP
+language, whose extension is {\tt b}; the bootstrapping ``compiler''
+is just the built-in parser (Figure~\ref{fig:b:compile.b}).
+
+Arranging that the repository contain the correct keys to make this
+work is subtle. A small initialization phase sets up:
+\begin{itemize}
+\item {\tt b:compile.b}$\quad$ 
+      The source code (a string) in Figure~\ref{fig:b:compile.b}
+\item {\tt b:compile}$\quad$ 
+      The parsed document corresponding to the above, such that {\tt b:compile} applied to {\tt b:compile.b} yields {\tt b:compile}
+\item {\tt main.b}$\quad$ 
+      The source code of the original ``main'' program
+\item {\tt main}$\quad$
+      The parsed document corresponding to the above, such that {\tt b:compile} applied to {\tt main.b} yields {\tt main}.
+\end{itemize}
+
+We do this by {\em anticipating}, during the initialization process,
+the {\em meaning} of {\tt b:compile} so that we can perform that
+action (parsing) on {\tt b:compile.b} to produce {\tt b:compile}. This
+can only be achieved by fiat and this is the essence of bootstrapping.
+
+After this minimal initialization, we can then exclusively use the web
+interface to develop and extend Wikiplia.
+
+\begin{figure*}[t]
+\begin{center}
+\includegraphics[width=0.92\textwidth]{ss_homepage}
+\end{center}
+\caption{Screenshot of the Wikiplia home page as of revision
+\currentrevision.} \label{fig:sshomepage}
+\end{figure*}
+
+\begin{figure*}[t]
+\begin{center}
+% \includegraphics[width=0.92\textwidth]{ss_history}
+\end{center}
+\caption{Screenshot of the history information for {\tt main-go.w}.
+} \label{fig:sshistory}
+\end{figure*}
+
 \section{Revision \currentrevision} \label{sec:current}
+
+As of writing, Wikiplia is at revision~\currentrevision, and has
+a number of features implemented.
+
+\subsection{Interface}
+
+The editing interface implemented has been enhanced greatly;
+Figure~\ref{fig:sshomepage} shows a screenshot of the main page. The
+various views of a key are shown with a series of tabs at the top of
+each page. Each user has a home page named after his IP
+address,\z\wcite{IP address} which he can use to catalogue his
+interests. Various warnings help the user, for example, if he tries to
+edit a page that was generated by compiling some source code, a
+warning message suggests that he may wish to edit the source code
+instead. An ornate logo in SVG\wcite{Scalable Vector Graphics} adorns
+the page, and a sidebar provides quick access to the site's features.
+The logo and graphics for the site are stored in the repository; the
+new {\tt raw} and {\tt typed} actions allows access to these resources
+over HTTP so that they may be freely modified.\z\footnote{Because
+Wikiplia is dogmatically forward-looking, the graphics require SVG
+support in the browser and support for the {\tt data:} URL format, a
+combination only found in the newest versions of the Mozilla Firefox.
+Wikiplia is compatible with incompatible browsers, however, simply
+displaying a crapified version of each page.} At revision 228 support
+for metadata was added for each page; a {\tt history} tab now shows
+the date, revision number, IP address, and edit summary for each
+change to a key (Figure~\ref{fig:sshistory}).
+
+\subsection{Wikiplia language {\tt w}}
+
+Writing XESP documents by hand is very tedious, so one of the first
+orders of business was to develop a compiler for a new language, {\tt
+w}, which can be extended with convenient features. The first version
+of this language, created in revision 22, was written in XESP. It
+automatically quoted the appropriate arguments to the {\sf let}, {\sf
+it}, {\sf lambda}, {\sf xcase} primitives. In revision 361, the
+compiler was ported to the language {\tt w} as {\tt w:compile.w}, then
+compiled with the existing {\tt b}-compiled version of {\tt
+w:compile.b}, then recompiled with itself until reaching a fixed
+point.\z\wcite{Fixed point (mathematics)} After that, more features
+were added: a multi-argument function construct {\tt fn}; a
+list-deconstructing binding construct {\tt lets}; simple support for
+separately-compiled libraries\wcite{Library (computing)} via {\tt
+include}; support for mutually-recursive\wcite{Recursion (computer
+science)} bundles of functions via {\tt fun};\footnote{Prior to
+this, recursion had to be encoded directly by passing an initial
+``self'' argument to each function.} and the {\tt cond} keyword for
+series of chained ``if\ldots else'' conditionals. At each stage, the
+feature is implemented using the current version of {\tt w}, and then
+{\tt w:compile.w} is rewritten to use that convenient extension, and
+then recompiled until reaching a fixed point.
+
+\subsection{Wikiplia language {\tt page}}
+
+Wikiplia is not just a programming language; it needs facilities for
+editing pages that are human readable as well. This can be used to
+edit documentation for programming languages, to modify the home page
+to tout new developments, to modify the software license
+(Section~\ref{sec:license}) or to deface other user's home pages. For
+this we provide a Wiki-like\wcite{Wiki} syntax that allows for the
+authorship of such pages and easy linking between them. Like other
+Wikis, a link to a page that does not exist is colored red, to alert
+the user to the opportunity to stake out cyberspace real estate. This
+syntax is compiled to XESP documents via the {\tt page} language; the
+resulting documents are active in the sense that they check the status
+of linked pages to report the correct color on every page load.
 
@@
 
-II. Version Idenfitication and Invalid Licenses
+II. Version Identification and Invalid Licenses
 
@@
 ``viral'' like the GNU GPL, so that freedom is preserved in all
-descendents of the software.}
+descendants of the software.}
 

### 2007-03-19 trunk/wikiplia/paper/wikiplia.tex
cxpoint

@@
 revision with revision number $i$ and data $X$ and inserts it under
-the key $s$. The judgment ${\tt cvs\ checkout}\ s = X$ fetches the most
-recent revision for the key $s$, the document $X$ (if no such key
-exists, then the document is invalid). The judgment ${\tt cvs\ 
-checkout\ -\!r}\ i\ s = X$ does the same, but for the specific
-revision number $i$ (if no such revision exists, the document is
-invalid). Finally, ${\tt cvs\ log}\ s = \vec{X}$ fetches all of the
-revision numbers for the key $s$, as a series of integers $\vec{X}$.
+the key $s$. The judgment ${\tt cvs\ checkout}\ s = X$ fetches the
+most recent revision for the key $s$,\z\wcite{Dynamic scope} the
+document $X$ (if no such key exists, then the document is invalid).
+The judgment ${\tt cvs\ checkout\ -\!r}\ i\ s = X$ does the same, but
+for the specific revision number $i$\wcite{Static scope} (if no such
+revision exists, the document is invalid). Finally, ${\tt cvs\ log}\ s
+= \vec{X}$ fetches all of the revision numbers for the key $s$, as a
+series of integers $\vec{X}$.
 
@@
 \subsection{Implementation}
-% SML: miserable non-free language
 
@@
 
+% XXX main emergency revert
+
 \subsection{Wikiplia language {\tt w}}
@@
 
-% All popular modern languages are defined via a definitional
-% interpreter\wcite{JavaScript}\wcite{Objective Caml}\wcite{Perl} with
-% accompanying O'Reilly ``animal'' book.
+We have reached the end of our journey. But the journey is not
+complete! We conclude with a discussion of future plans and unrelated
+work, and then conclude with another paragraph.
+
+\subsection{Future Work}
+
+Though Wikipedia in its current form is a usable general-purpose
+programming language, work remains to be done for it to reach its full
+potential. For one, it needs a vibrant community of contentious and
+hubristic editors hiding behind anonymous IP addresses boldly
+asserting half-baked syntactic extensions or enforcing superficial
+style preferences, gritting their teeth while typing and clicking
+white knuckled in a kind of road rage\wcite{List of rages} created by
+the dehumanizing semantic markup by which they are forced to
+communicate.
+
+We also seek to improve the languages. The language {\tt w} needs many
+more features to speed development: the parenthesis-based XESP syntax
+should eventually be replaced by a pleasant concrete syntax, if we can
+get around to it before too much code is written in XESP.
+
+\subsection{Unrelated Work}
+
+All popular modern languages are defined via a definitional
+interpreter\wcite{JavaScript}$^,$\wcite{Objective
+Caml}$^,$\wcite{Perl} with accompanying O'Reilly ``animal''
+book.\z\wcite{O'Reilly Media} The work on Wikiplia is unrelated: We
+have no animal mascot\wcite{ORLY owl} and the languages are described
+by a tower of source-to-source translations on top of a universally
+parseable semantic document in XML\wcite{Category:ML programming
+language family} form.
+
+The author doesn't think much of musicals,\z\wcite{Musical theatre} to
+be perfectly honest, so those are basically a no-go. He also feels
+that the metric system\wcite{Metric system} and time zones\wcite{Time
+zone} are pretty overrated. Ketchup on eggs is gross, but not quite as
+gross as foie gras\wcite{Foie gras}, which more or less has the word
+``gross'' in its name so duh.\z\wcite{Duh}
+
+\subsection{Another Paragraph}
+
+We have described Wikiplia, the free programming language that anyone
+can edit. Unlike other programming languages, it supports a variety of
+freedoms (...) and is explicitly scalable to new freedoms. ...
 

### 2007-03-19 wikiplia/paper/wikiplia.tex
almost done, need screenshots

@@
 
-% TODO: license
-\newcommand\currentrevision{465}
+\newcommand\currentrevision{475}
+\newcommand\wikipliaurl{http://wikiplia.spacebar.org/}
 
@@
 \newcommand\wcite[1]{\footnote{\tiny Wikipedia, the free encyclopedia: {\it #1}; 2007}}
+\newcommand\wciteshort[1]{\footnote{\tiny Wikipedia, the free encyclopedia: {\it #1}}}
 \newcommand\comment[1]{}
@@
 implemented as a web-site on the Internet at the address
-$${\tt http://wikiplia.spacebar.org/}$$
+$${\tt \wikipliaurl}$$
 
@@
 of the license does not directly permit the licensee to copy the
-software at all; he must first amend the license using provision {\tt
-VI} to give himself this permission.
+% double-entendre with VI intentional
+software at all; he must first amend the license using {\tt VI} to
+give himself this permission.
 
@@
 
-% All popular modern languages are defined via a definitional
-% interpreter\wcite{JavaScript}\wcite{Objective Caml}\wcite{Perl} with
-% accompanying O'Reilly ``animal'' book.
+We have reached the end of our journey. But the journey is not
+complete! We conclude with a discussion of future plans and unrelated
+work, and then conclude with another paragraph.
+
+\subsection{Future Work}
+
+Though Wikipedia in its current form is a usable general-purpose
+programming language, work remains to be done for it to reach its full
+potential. For one, it needs a vibrant community of contentious and
+hubristic editors hiding behind anonymous IP addresses boldly
+asserting half-baked syntactic extensions or enforcing superficial
+style preferences, gritting their teeth while typing and clicking
+white knuckled in a kind of road rage\wcite{List of rages} created by
+the dehumanizing semantic markup by which they are forced to
+communicate.
+
+% XXX: supporting users with revision #s?
+
+We also seek to improve the languages. The language {\tt w} needs many
+more features to speed development: the parenthesis-based XESP syntax
+should eventually be replaced by a pleasant concrete syntax, if we can
+get around to it before too much code is written in XESP. A type
+system\wcite{Type system} is not planned, because type systems
+restrict Freedom~\fexpression, the freedom of expression. However, we
+should seek to make Wikiplia as multi-paradigm as possible (again,
+freedom from discrimination on the basis of paradigm\wcite{paradigm}
+orientation), supporting OOPs-oriented
+programming,\z\wcite{Object-oriented programming} aspect-oriented
+programming,\z\wcite{Aspect-oriented programming} duck-oriented
+typing,\z\wcite{Duck typing} orientation-oriented
+orienteering,\z\wcite{Orienteering} Orient-oriented
+Programming,\z\wcite{The Orient} etc.
+
+The {\tt
+page} language needs extensions for developing human-readable web
+pages, mostly for the purpose of creating jazzy graphics and boxes
+that distract from or directly call attention to obvious problems with
+the pages without actually addressing those problems.
+
+\subsection{Unrelated Work}
+
+All popular modern languages are defined via a definitional
+interpreter\wcite{JavaScript}$^,$\wcite{Objective
+Caml}$^,$\wcite{Perl} with accompanying O'Reilly ``animal''
+book.\z\wcite{O'Reilly Media} The work on Wikiplia is unrelated: We
+have no animal mascot\wcite{ORLY owl} and the languages are described
+by a tower of source-to-source translations on top of a universally
+parseable semantic document in XML\wciteshort{Category:ML programming
+language family} form.
+
+The author doesn't think much of musicals,\z\wcite{Musical theatre} to
+be perfectly\wcite{Thomas Aquinas} honest,\z\wcite{Honesty} so those
+are basically a no-go. He also feels that the metric
+system\wcite{Metric system} and time zones\wcite{Time zone} are pretty
+overrated. Ketchup\wcite{Ketchup} on eggs\wcite{Egg (food)} is
+gross,\z\wcite{Gross} but not quite as gross as foie gras\wcite{Foie
+gras}, which more or less has the word ``gross'' in its name so
+duh.\z\wcite{Duh}
+
+\subsection{Another Paragraph}
+
+We have described Wikiplia, the free programming language that anyone
+can edit. Unlike other programming languages, it is designed to
+support a variety of freedoms (\ftinker, \fexpression, \fbeer, \fusa,
+\ffree, \ftobe) and is explicitly scalable to new freedoms. The system
+is implemented in a minimal bootstrapping core, and then built up to
+a featured system using its own faculties for extension. XXX more...
+
+\bigskip
+$${\tt \wikipliaurl}$$
 

### 2007-03-20 wikiplia/paper/wikiplia.tex
little comment
fix image
lparen, paper

@@
   weatherproof footwear and their fastening mechanisms,
-  hyper-driven devices,
+  hyper-driven devices
 }
@@
 language that {\em anyone} can edit: from bearded academic
-elite\wcite{Tenure} to congressional staffers\wcite{Congressional
+elite\wcite{Tenure} to congressional staffers\wciteshort{Congressional
 staffer edits to Wikipedia} to nameless Slashdot\wcite{Slashdot}
@@
 licensed under the GPL. The intention is that anyone receiving the
-software can exercise Freedom~\ftinker by understanding the source
+software can exercise Freedom~\ftinker\ by understanding the source
 code and modifying it to suit his needs. Clearly any source code will
@@
 code cannot be easily understood or modified, even though it is
-technically ``source code.'' The GPL therefore defines source code as
-the ``preferred form'' for ``making modifications.''
+technically ``source code.'' The GPL therefore legally defines source
+code as the ``preferred form'' for ``making modifications.''
 
@@
 programming language---after all, it has the same source code---but
-their minor bifurcation in development history means that they forever
-differing meanings of the \verb+'\c'+ extension. This mistake is
-likely to go unnoticed for some time, and until it is resolved, the
+their minor bifurcation in development history means that they have
+forever different meanings of the \verb+'\c'+ extension. This mistake
+is likely to go unnoticed for some time, and until it is resolved, the
 meaning of the \verb+'\c'+ extension is firmly enslaved in the
 bipartite penitentiary of double entendre, yearning to be
-free\ldots\z\wcite{Information wants to be free}
+free\,\ldots\z\wcite{Information wants to be free}
 
@@
 Even if we think these situations are implausible in the hands of
-well-intentioned, well-mannered software engineers, we wish for our
-software to remain free even when in the hands of nefarious and crafty
-factions who would seek to fracture\wcite{Fork (software development)}
-our free software community. We therefore need a technological and
-legal solution that forces the entire development history to be
-available. Keeping with Freedom~\fusa, we call this technology and its
-license {\em Total Information Awareness} after the successful project
-of the US Government with the same name.\z\wcite{Information Awareness
-Office}
+well-intentioned, well-mannered and capable%
+%
+\wcite{Year 2000 problem}$^,$%
+\wciteshort{Northeast Blackout of 2003}$^,$%
+\wcite{Ariane 5 Flight 501}$^,$%
+\wcite{Mars Climate Orbiter}$^,$%
+\wcite{Mars Polar Lander}$^,$%
+\wcite{Mars Rover}$^,$%
+\wcite{Mars Pathfinder}$^,$%
+\wcite{Gripen\#Crashes}$^,$%
+\wcite{missingno.}$^,$%
+\wcite{XSS}$^,$%
+\wcite{Buffer overflow}$^,$%
+\wcite{Therac-25}$^,$%
+\wcite{Lothar (storm)}$^,$%
+\wcite{Mariner 1}$^,$%
+\wcite{Code Red worm}$^,$%
+\wcite{SQLSlammer}$^,$%
+\wcite{Sandstorm (vehicle)}$^,$%
+\wcite{Samy (XSS)}$^,$%
+\wcite{Pentium FDIV bug}$^,$%
+\wcite{MIM-103 Patriot}$^,$%
+\wcite{Windows 95}$^,$%
+\wcite{Morris (computer worm)}$^,$%
+\wcite{USS Yorktown (CG-48)}
+%
+software engineers, we wish for our software to remain free even when
+in the hands of nefarious and crafty factions who would seek to
+fracture\wcite{Fork (software development)} our free software
+community. We therefore need a technological and legal solution that
+forces the entire development history to be available. Keeping with
+Freedom~\fusa, we call this technology and its license {\em Total
+Information Awareness} after the successful project of the US
+Government with the same name.\z\wcite{Information Awareness Office}
 
@@
 
-The remainder of this paper proceeds as follows. We first discuss in
+The remainder of this paper proceeds as follows. We first present in
 Section~\ref{sec:calculus} the design of the initial Wikiplia system,
@@
 current state of Wikiplia as of revision \currentrevision\ in
-Section~\ref{sec:current}. We discuss the freedom-preserving TIA
+Section~\ref{sec:current}. We explain the freedom-preserving TIA
 Public License in Section~\ref{sec:license}. We conclude with a
@@
 \begin{array}{r@{}l}
-  \G \vdash \rate & \tag{prim}{\sf xcase}\ \tag{int}{s}\  X_0\ X_1\ X_2\ X_3\ 
+  \G \vdash \rate & \tag{prim}{\sf xcase}\ \tag{int}{i}\  X_0\ X_1\ X_2\ X_3\ 
                     X\ \vec{X} \m X' \\
@@
   \G \vdash \rate & \tag{prim}{\sf xcase}\ \tag{symbol}{s}\  X_0\ X_1\ X_2\ X_3\ X_4\ \\
-                  & \tag{list}{\tag{symbol}{s_s}\ X_b}\ \vec{X} \m X' \\
+                  & \tag{list}{\tag{symbol}{s_b}\ X_b}\ \vec{X} \m X' \\
 \end{array}
-}{\G, s_s = \tag{string}{s} \vdash \eval X_b \m X'}} \\[1em]
+}{\G, s_b = \tag{string}{s} \vdash \eval X_b \m X'}} \\[1em]
 \multicolumn{2}{c}{\infer{
@@
 \multicolumn{2}{c}{\infer{
-\G \vdash \rate \tag{prim}{\sf read}\ \tag{string}{s} \tag{int}{i} \m X
+\G \vdash \rate \tag{prim}{\sf read}\ \tag{string}{s}\ \tag{int}{i} \m X
 }{{\tt cvs\ checkout\ -\!r}\ i\ s = X}} \\[1em]
@@
    \lb $sym$\rb & $\tag{symbol}{sym}$ \\
-   \lb {\tt'}$E$\rb & $\tag{quote}{E}$ \\
+   \lb {\tt'}$E$\rb & $\tag{quote}{\lb E\rb}$ \\
   \end{tabular}
@@
 documents and which the computer automatically parses and converts to
-the easily parsable XML syntax and back to the new syntax, reducing
+the easily parseable XML syntax and back to the new syntax, reducing
 complexity.\z\footnote{For efficiency, the Wikiplia implementation is
@@
 {\sf request.time} with the appropriate values), and then evaluating
-the document contained at the head of the key ``main'' in the
+the document contained at the head of the key {\tt main} in the
 repository. The result of that evaluation is sent back to the web
@@
 cvs} judgments given in Section~\ref{sec:calculus}. Unlike typical
-revision control systems, we need to support very large numbers of
-revisions with small edits (since every change is saved), so the
-implementation is engineered to make storing small revisions very
-cheap. Particularly, revisions are aggressively compressed by only
-storing the newest revision directly, and then a series of difference
-``plans'' that describe how to get the next older revision from the
-current one. We compute optimal plans using an efficient minimal edit
-distance\wcite{Levenshtein distance} calculation at the
-token\wcite{Token (parser)} level. As of revision~\currentrevision,
-the database is only 236 kilobytes.\z\wcite{Kilobyte}
+revision control systems, we need to support very large
+numbers\footnote{The implementation also supports revision numbers of
+arbitrary size.} of revisions with small edits (since every change is
+saved), so the implementation is engineered to make storing small
+revisions very cheap. Particularly, revisions are aggressively
+compressed by only storing the newest revision directly, and then a
+series of difference ``plans'' that describe how to get the next older
+revision from the current one. We compute optimal plans using an
+efficient minimal edit distance\wcite{Levenshtein distance}
+calculation at the token\wcite{Token (parser)} level. As of
+revision~\currentrevision, the database is only 236
+kilobytes.\z\wcite{Kilobyte}
 
@@
 
-The key ``main'' of Wikiplia is responsible for decomposing the URL
+The key {\tt main} of Wikiplia is responsible for decomposing the URL
 and acting upon it however is appropriate. The goal of the
@@
 self-sufficient\wcite{Self-sufficiency} in the sense that the language
-can be edited from the web site implemented by ``main.'' To do so, the
+can be edited from the web site implemented by {\tt main}. To do so, the
 web site needs to be able to present the user with an edit box
-containing the current source of the ``main'' key and the ability to
-save his code into the database, overwriting the ``main'' key, in
+containing the current source of the {\tt main} key and the ability to
+save his code into the database, overwriting the {\tt main} key, in
 order to add functionality.
 
-The initial implementation of ``main'' is provides for the simple
+The initial implementation of {\tt main} is provides for the simple
 ability to edit, save, and view the current version of keys in the
@@
 \begin{center}
-% \includegraphics[width=0.92\textwidth]{ss_history}
+\includegraphics[width=0.92\textwidth]{ss_history}
 \end{center}
@@
 The logo and graphics for the site are stored in the repository; the
-new {\tt raw} and {\tt typed} actions allows access to these resources
+new {\tt raw} and {\tt typed} actions allow access to these resources
 over HTTP so that they may be freely modified.\z\footnote{Because
@@
 to tout new developments, to modify the software license
-(Section~\ref{sec:license}) or to deface other user's home pages. For
-this we provide a Wiki-like\wcite{Wiki} syntax that allows for the
+(Section~\ref{sec:license}) or to deface other user's personal pages.
+For this we provide a Wiki-like\wcite{Wiki} syntax that allows for the
 authorship of such pages and easy linking between them. Like other
@@
 
-Though Wikipedia in its current form is a usable general-purpose
+Though Wikiplia in its current form is a usable general-purpose
 programming language, work remains to be done for it to reach its full
@@
 orienteering,\z\wcite{Orienteering} Orient-oriented
-Programming,\z\wcite{The Orient} etc.
-
-The {\tt
-page} language needs extensions for developing human-readable web
-pages, mostly for the purpose of creating jazzy graphics and boxes
+programming,\z\wcite{The Orient} etc.
+
+The {\tt page} language needs extensions for developing human-readable
+web pages, mostly for the purpose of creating jazzy graphics and boxes
 that distract from or directly call attention to obvious problems with
@@
 
-The author doesn't think much of musicals,\z\wcite{Musical theatre} to
-be perfectly\wcite{Thomas Aquinas} honest,\z\wcite{Honesty} so those
-are basically a no-go. He also feels that the metric
-system\wcite{Metric system} and time zones\wcite{Time zone} are pretty
-overrated. Ketchup\wcite{Ketchup} on eggs\wcite{Egg (food)} is
+The author doesn't think\wcite{Thought} much of
+musicals,\z\wcite{Musical theatre} to be perfectly\wcite{Thomas
+Aquinas} honest,\z\wcite{Honesty} so those are basically\wcite{BASIC}
+a no-go. He also feels that the metric system\wcite{Metric system} but
+paradoxically\wcite{Pardadox} also time zones\wcite{Time zone} are
+pretty overrated. Ketchup\wcite{Ketchup} on eggs\wcite{Egg (food)} is
 gross,\z\wcite{Gross} but not quite as gross as foie gras\wcite{Foie
-gras}, which more or less has the word ``gross'' in its name so
-duh.\z\wcite{Duh}
+gras}, which more or less has the word\wcite{Word} ``gross'' in its
+name\wcite{Name} so duh.\z\wcite{Duh}
 

### 2007-03-20 trunk/wikiplia/paper/wikiplia.tex
interface ready for paper

@@
 
-% TODO: license
-\newcommand\currentrevision{465}
+\newcommand\currentrevision{532}
+\newcommand\currentsizekb{250}
+\newcommand\wikipliaurl{http://wikiplia.spacebar.org/}
 
@@
 cvs} judgments given in Section~\ref{sec:calculus}. Unlike typical
-revision control systems, we need to support very large numbers of
-revisions with small edits (since every change is saved), so the
-implementation is engineered to make storing small revisions very
-cheap. Particularly, revisions are aggressively compressed by only
-storing the newest revision directly, and then a series of difference
-``plans'' that describe how to get the next older revision from the
-current one. We compute optimal plans using an efficient minimal edit
-distance\wcite{Levenshtein distance} calculation at the
-token\wcite{Token (parser)} level. As of revision~\currentrevision,
-the database is only 236 kilobytes.\z\wcite{Kilobyte}
+revision control systems, we need to support very large
+numbers\footnote{The implementation also supports revision numbers of
+arbitrary size.} of revisions with small edits (since every change is
+saved), so the implementation is engineered to make storing small
+revisions very cheap. Particularly, revisions are aggressively
+compressed by only storing the newest revision directly, and then a
+series of difference ``plans'' that describe how to get the next older
+revision from the current one. We compute optimal plans using an
+efficient minimal edit distance\wcite{Levenshtein distance}
+calculation at the token\wcite{Token (parser)} level. As of
+revision~\currentrevision, the database is only \currentsizekb\ 
+kilobytes.\z\wcite{Kilobyte}
 
@@
 
+% XXX: supporting users with revision #s?
+
 We also seek to improve the languages. The language {\tt w} needs many
@@
 should eventually be replaced by a pleasant concrete syntax, if we can
-get around to it before too much code is written in XESP.
+get around to it before too much code is written in XESP. A type
+system\wcite{Type system} is not planned, because type systems
+restrict Freedom~\fexpression, the freedom of expression. However, we
+should seek to make Wikiplia as multi-paradigm as possible (again,
+freedom from discrimination on the basis of paradigm\wcite{paradigm}
+orientation), supporting OOPs-oriented
+programming,\z\wcite{Object-oriented programming} aspect-oriented
+programming,\z\wcite{Aspect-oriented programming} duck-oriented
+typing,\z\wcite{Duck typing} orientation-oriented
+orienteering,\z\wcite{Orienteering} Orient-oriented
+programming,\z\wcite{The Orient} etc.
+
+The {\tt page} language needs extensions for developing human-readable
+web pages, mostly for the purpose of creating jazzy graphics and boxes
+that distract from or directly call attention to obvious problems with
+the pages without actually addressing those problems.
 
@@
 by a tower of source-to-source translations on top of a universally
-parseable semantic document in XML\wcite{Category:ML programming
+parseable semantic document in XML\wciteshort{Category:ML programming
 language family} form.
 
-The author doesn't think much of musicals,\z\wcite{Musical theatre} to
-be perfectly honest, so those are basically a no-go. He also feels
-that the metric system\wcite{Metric system} and time zones\wcite{Time
-zone} are pretty overrated. Ketchup on eggs is gross, but not quite as
-gross as foie gras\wcite{Foie gras}, which more or less has the word
-``gross'' in its name so duh.\z\wcite{Duh}
+The author doesn't think\wcite{Thought} much of
+musicals,\z\wcite{Musical theatre} to be perfectly\wcite{Thomas
+Aquinas} honest,\z\wcite{Honesty} so those are basically\wcite{BASIC}
+a no-go. He also feels that the metric system\wcite{Metric system} but
+paradoxically\wcite{Pardadox} also time zones\wcite{Time zone} are
+pretty overrated. Ketchup\wcite{Ketchup} on eggs\wcite{Egg (food)} is
+gross,\z\wcite{Gross} but not quite as gross as foie gras\wcite{Foie
+gras}, which more or less has the word\wcite{Word} ``gross'' in its
+name\wcite{Name} so duh.\z\wcite{Duh}
 
@@
 We have described Wikiplia, the free programming language that anyone
-can edit. Unlike other programming languages, it supports a variety of
-freedoms (...) and is explicitly scalable to new freedoms. ...
+can edit. Unlike other programming languages, it is designed to
+support a variety of freedoms (\ftinker, \fexpression, \fbeer, \fusa,
+\ffree, \ftobe) and is explicitly scalable to new freedoms. The system
+is implemented in a minimal bootstrapping core, and then built up to
+a featured system using its own faculties for extension. XXX more...
+
+\bigskip
+$${\tt \wikipliaurl}$$
 

### 2007-03-20 wikiplia/paper/wikiplia.tex
improve look of main page, but should use templates for this eventually
camera ready

@@
 
-\newcommand\currentrevision{475}
-\newcommand\wikipliaurl{http://wikiplia.spacebar.org/}
+\newcommand\currentrevision{532}
+\newcommand\currentsizekb{250}
+\newcommand\wikipliaurl{http://wikiplia.spacebar.org:2222/}
 
@@
 
+% so that visual gag on ``capable'' doesn't appear at top of page
+\vspace{10em}
+\hyphenation{capable}
 \subsection{Solution}
@@
 Based on these scenarios we conclude that extant social measures such
-as revision control conventions are not enough to guarantee freedom.
+as revision control conventions are not enough to guarantee freedom
+in all circumstances.
 Even if we think these situations are implausible in the hands of
@@
 calculation at the token\wcite{Token (parser)} level. As of
-revision~\currentrevision, the database is only 236
+revision~\currentrevision, the database is only \currentsizekb\ 
 kilobytes.\z\wcite{Kilobyte}
@@
 
-% XXX main emergency revert
+It is very easy to make mistakes that render the system unusable.
+Therefore Wikiplia supports the ability to safely revert to a previous
+version of a key. When making changes to the {\tt main} key this
+ability can be accidentally disabled, so the complex functionality of
+{\tt main} was split off to a new key called {\tt main-go} at revision
+9; the {\tt main} key now only dispatches to {\tt main-go} but
+provides an {\tt emergency-revert} action that automatiaclly reverts
+{\tt main-go} to its previous revision in case it is damaged and the
+site is unusable.
 
@@
 have no animal mascot\wcite{ORLY owl} and the languages are described
-by a tower of source-to-source translations on top of a universally
-parseable semantic document in XML\wciteshort{Category:ML programming
-language family} form.
-
-The author doesn't think\wcite{Thought} much of
+by a tower of source-to-source translations\wcite{Translation} on top
+of a universally parseable semantic document in
+XML\wciteshort{Category:ML programming language family} form.
+
+The author\wcite{Author} doesn't think\wcite{Thought} much of
 musicals,\z\wcite{Musical theatre} to be perfectly\wcite{Thomas
@@
 support a variety of freedoms (\ftinker, \fexpression, \fbeer, \fusa,
-\ffree, \ftobe) and is explicitly scalable to new freedoms. The system
-is implemented in a minimal bootstrapping core, and then built up to
-a featured system using its own faculties for extension. XXX more...
+\ffree, \ftobe) and is explicitly scalable to new freedoms. Wikiplia
+is implemented in a minimal bootstrapping core based on freedom-aware
+technologies such as XML, and then built up to a featured system using
+its own faculties for extension. However, much work remains to be
+done. We invite you to join us!
 

### 2007-04-13 wikiplia/paper/wikiplia.tex
monoid joke, typo

@@
 propagating invalid document status throughout evaluation and thus
-complicates the rules substantially.}
+complicates the rules substantially.} Lists---delineated with the
+$\tag{list}$ tag---play a central r\^ole because they are the free
+moniod\wcite{Free monoid} on the set of XML documents, and we seek
+to achieve freedom wherever possible.
 
@@
 pretty overrated. Ketchup\wcite{Ketchup} on eggs\wcite{Egg (food)} is
-gross,\z\wcite{Gross} but not quite as gross as foie gras\wcite{Foie
-gras}, which more or less has the word\wcite{Word} ``gross'' in its
-name\wcite{Name} so duh.\z\wcite{Duh}
+gross,\z\wcite{Gross} but not quite as gross as foie
+gras,\z\wcite{Foie gras} which more or less has the word\wcite{Word}
+``gross'' in its name\wcite{Name} so duh.\z\wcite{Duh}
 

### 2007-06-11 trunk/wikiplia/paper/wikiplia.tex
typo

@@
 \newcommand\currentsizekb{250}
-\newcommand\wikipliaurl{http://wikiplia.spacebar.org/}
+\newcommand\wikipliaurl{http://wikiplia.spacebar.org:2222/}
 
@@
 Public License in Section~\ref{sec:license}. We conclude with a
-discussion of unrelated work and plans for the future~\ref{sec:conc}.
+discussion of unrelated work and plans for the future in
+Section~\ref{sec:conc}.
 

### 2010-03-20 trunk/youkeepdying/paper/youkeepdying.tex
commit to start work on paper

\documentclass[twocolumn]{article}

% \usepackage{eclbkbox}
% \usepackage{amsmath}
% \usepackage{amssymb}
% \usepackage{amscd}
% \usepackage{xy}
\usepackage{graphicx}
% \usepackage{fancyhdr}
% \usepackage{color}
% \usepackage[dark,all,bottom,landscape,timestamp]{draftcopy}
% \usepackage{everypage}

% \usepackage{ulem}
% go back to italics for emphasis, though
% \normalem

% \AddEverypageHook{\waterb}

% \fancyhead[LE,RO]{\slshape page number}
% \fancyhead[RE]{\slshape \LaTeX}
% \fancyhead[LO]{\slshape \LaTeX}
% \fancyfoot[LO,LE]{\slshape ! LaTeX Error: Something's wrong---perhaps a missing $\backslash$item?}
% \fancyfoot[C]{}
% \fancyfoot[RO,RE]{\slshape $\heartsuit\hspace{-1.5em}\longrightarrow$ }

% \draftcopyName{DRAFT!}{300}
% \pagestyle{fancy}

\begin{document}

\title{ {\bf You Keep Dying } }
\author{Tom~Murphy~VII}
\date{1 April 2010}

\maketitle

\begin{abstract}
We present a new programming language called Wikiplia. The language
has an unprecedented level of integration: The system is its own
compiler, language definition, documentation, development environment,
distributed filesystem, database, revision control system,
bootstrapping software license, community message board, and World
Wide Web home site. Wikiplia is designed to be Free to a greater
extent and in more dimensions than existing languages.
\end{abstract}

\footnotetext[-1]{Copyright \copyright\ 2010 the Regents of the Wikiplia
Foundation. Appears in SIGBOVIK 2010 with the permission of the
Association for Computational Heresy; {\em IEEEEEE!} press,
Verlag-Verlag volume no.~0x40-2A.
\pounds 0.00}

\vspace{1em}
{\noindent \small {\bf Keywords}:
  Freedom, programming language, software license, wiki, XML, 
  weatherproof footwear and their fastening mechanisms,
  hyper-driven devices
}

% copyright notice generated automatically from Year and author.
% permission added if \permission{} given.

\section{Introduction}

\begin{figure}[t]
\begin{center}
\includegraphics[width=0.43\textwidth]{lightning}
\end{center}
\caption{You keep conduct the lightning\ \ and\ \ you keep dying} \label{fig:lightning}
\end{figure}

NetCraft confirms it. *LF is dying.
I believe children are our future or futures.

\end{document}

### 2011-03-05 wishlist/paper.tex
commit so I can profile on linux

facebook wall posts, 4-grams
100 top paths:
0.002525504 pittsburgh
0.002093510 steelers
0.002093510 sfo
0.001953943 it's
0.001256106 bdl
0.001116539 sigbovik
0.001099196 facebook
0.000976971 mic
0.000976971 s
0.000976971 can't
0.000837404 i'm
0.000837404 icfp
0.000837404 app
0.000697837 x
0.000697837 drunj
0.000697837 g
0.000616021 ther
0.000558269 doesn't
0.000558269 's
0.000558269 sr
0.000558269 pm
0.000558269 fyi
0.000558269 wtf
0.000558269 k
0.000558269 don't
0.000558269 today's
0.000558269 m
0.000558269 what's
0.000456766 iphone
0.000429438 inted
0.000418702 oh
0.000418702 phl
0.000418702 gonna
0.000418702 pre
0.000418702 zrh
0.000418702 th
0.000418702 h
0.000418702 b
0.000418702 n
0.000418702 ok
0.000418702 d
0.000418702 that's
0.000418702 rex
0.000418702 fanzibar
0.000418702 st
0.000418702 kinda
0.000418702 didn't
0.000355262 cally
0.000334962 brillobox
0.000314027 superburrito
0.000305304 weath
0.000279135 techno
0.000279135 vii

trigrams, including comments:
100 top paths:
0.003013897 it's
0.001856422 ther
0.001704432 i'm
0.001503910 sfo
0.001102867 -
0.001102867 oh
0.001050350 don't
0.001002607 haha
0.001002607 s
0.000902346 bdl
0.000808769 mic
0.000808769 that's
0.000776981 pittsburgh
0.000705812 thes
0.000701825 ok
0.000645063 can't
0.000601564 icfp
0.000601564 k
0.000583335 didn't
0.000507201 runj
0.000501303 g
0.000501303 omg
0.000501303 x
0.000485262 ins
0.000453126 sout
0.000450580 ands
0.000449950 coff
0.000425932 reat
0.000425133 ning
0.000416189 pittle
0.000412468 getty
0.000407151 ster
0.000401043 sr
0.000401043 pm
0.000401043 m
0.000401043 wtf
0.000401043 aw
0.000401043 d
0.000401043 fyi
0.000401043 ple
0.000401043 's
0.000400791 phon
0.000386315 weat
0.000384116 befor
0.000375978 els
0.000374307 als
0.000357782 nothe
0.000354769 app
0.000354769 appy
0.000354422 they're
0.000353750 wate
0.000345727 there's
0.000340508 res
0.000339344 mes
0.000334202 ove
0.000326473 wher
0.000324373 lity
0.000319011 hotos
0.000316613 dea
0.000315777 sigbovik
0.000315105 ver
0.000314965 thand
0.000310607 outh
0.000300782 n

facebook, n = 2
100 top paths:
0.010628671 ing
0.007144197 th
0.005025442 st
0.004291375 re
0.002932937 ne
0.002662094 se
0.002546194 whe
0.002108786 ther
0.001923496 al
0.001669458 con
0.001652820 wor
0.001498222 ch
0.001495485 wis
0.001379778 i'm
0.001336166 jus
0.001302469 sh
0.001280888 yout
0.001244352 fo
0.001225250 le
0.001224422 pre
0.001221357 ar
0.001201881 com
0.001200574 ge
0.001165417 thes
0.001116295 som
0.001108305 fing
0.001104851 ang
0.001102867 -
0.001102706 it's
0.001039704 ist
0.001036886 hing
0.001007930 ou
0.001002607 s
0.000967282 de
0.000932965 en
0.000926568 tor
0.000827260 wat
0.000827151 oh
0.000814235 cor
0.000812961 ned
0.000783696 lis
0.000762020 ple
0.000760162 whis
0.000759898 sed
0.000755396 res
0.000728412 mor
0.000728049 ong
0.000727818 don
0.000720098 cand
0.000710277 ning
0.000708380 ithe
0.000693547 ame
0.000659248 ty
0.000656211 dow
0.000635771 ha
0.000633225 ext
0.000601564 k

### 2011-03-06 wishlist/paper.tex
Started on paper. some other ideas too

@@
+\documentclass[twocolumn]{article}
+\usepackage[top=1.1in, left=0.85in, right=0.85in]{geometry}
+
+% \usepackage{eclbkbox}
+\usepackage{amsmath}
+\usepackage{amssymb}
+% \usepackage{amscd}
+% \usepackage{xy}
+\usepackage{graphicx}
+% \usepackage{fancyhdr}
+% \usepackage{color}
+% \usepackage[dark,all,bottom,landscape,timestamp]{draftcopy}
+% \usepackage{everypage}
+
+% \usepackage{ulem}
+% go back to italics for emphasis, though
+% \normalem
+
+\begin{document}
+
+\title{What words ought to exist? \\
+       {\normalsize Coining with coinduction}}
+\author{Dr.~Tom~Murphy~VII~Ph.D.\thanks{
+Copyright \copyright\ 2011 the Regents of the Wikiplia
+Foundation. Appears in SIGBOVIK 2011 with the blessing of the
+Association for Computational Heresy; {\em IEEEEEE!} press,
+Verlag-Verlag volume no.~0x40-2A.
+\yen 0.00}
+}
+
+\date{3 March 2011}
+
+\maketitle
+
+\begin{abstract}
+what do I put here
+\end{abstract}
+
+\vspace{1em}
+{\noindent \small {\bf Keywords}:
+ cryptolexicography, n-Markov models, coinduction
+}
+
+\section*{Introduction}
+During a recent high-stakes game of Scrabble-brand Crossword
+Puzzle\footnote{Scrabble is a registered trademark of Hasbro
+Inc./Milton Bradley, and Mattel/JW Spear \& Sons plc.} I had what
+could only be described as a killer bingo word (all 7 tiles) that,
+after careful study, I determined could not be placed anywhere on the
+board. Later in that same game, I had another sequence of letters that
+just totally seemed like it should be able to make some long-ass
+words, like for example ``oilsoap'' which turns out is not a legal
+Scrabble word.\!\footnote{There are actually no 7-letter words that
+can be made from these letters. Don't even bother. Even if playing off
+an existing letter on the board, the best we can do are the non-bingos
+``topsoil,'' ``topsail,'' or ``poloist'' with an available {\it t}.}
+This naturally made me frustrated and I wanted to do something about
+it. Why can't ``oilsoap'' be a word? Or ``loopsia''? Words are
+introduced into the lexicon all the time. My first reaction of course
+was to make an online version of Scrabble where all words are legal.
+This is available at {\tt http://snoot.org/toys/scrallbe}, and is
+pretty boring, I gotta be honest.
+% TODO: Screenshot of Scrallbe
+
+The thing is, it's just more fun when some words aren't words. Think
+about it: If all words were real, then you could never make a really
+devastatingly successful challenge in Scrabble that like, rocked the
+whole household and turned a formerly casual family games night into
+some kind of crying contest. Spelling bees could still exist, because
+while no matter what those kids spelled,\!\footnote{Well, we have to
+consider the possibility that the kiddo would use a letter that
+doesn't exist. In this particular fantasy, grant me also that every
+letter also exists, even $\stackrel{\Diamond}{\smile}$.} it would be
+a word, it would not necessarily be the {\it right} word, just like
+maybe a homophone. There would be fewer bar fights, but probably not
+that many fewer. Moreover, iuhwueg nznie a uaohahweih zmbgba bawuyg!
+
+Clearly we need more words, but not all of them. So this raises the
+question: What words {\it ought} to exist? This paper explores several
+different approaches for scientifically answering this question,
+compares the results, and proposes specific words that should be
+added, with their meanings.
+
+Disclaimer possibly indicated for SIGBOVIK: The ``research'' contained
+herein is 100\% legitimate. I have attempted to present it in a
+tutorial style that assumes little mathematical or computer science
+background. I have also left off the last {\it S} for {\it Savings}.
+
+\section{First idea: Wishlist}
+
+My website ``{\sf snoot.org}'' has a number of games on it, including
+a Scrabble clone called Scribble\footnote{{\tt
+http://snoot.org/toys/scribble/}} and Boggle clone called
+Muddle.\!\footnote{{\tt http://snoot.org/toys/muddle/}} This website
+has been running for almost ten years, comprising over 150,000
+Scribble games totaling 3.8 million words placed and 628,000 Muddle
+games with over 10 million words found. During each game, players
+repeatedly attempt to play words that aren't real. The computer
+rebukes them, but hope really springs eternal with these people. It's
+like they truly deeply wish to break out of the shackles of the
+Official Scrabble Players Dictionary. So the first approach to
+determining what words ought to exist is to analyze the words that
+people tried to play, in order to try to extract the essence of
+word-yearning.
+
+This analysis is quite straightforward. I took the ten years of logs
+files and extracted each attempt to play a word in Scribble or Muddle.
+These log files are quite large, so the first step is just to get a
+count, for each alleged word, and store those in a more convenient
+format. There were XXX total words attemped in Scrabble and XXX in
+Muddle. The most frequent ones appear in
+Figure~\ref{fig:mostfrequent}. The most frequent words are all
+legitimate words, since players have a bias towards attempting words
+that will not be rebuked by the computer. Seeing the words that people
+wish existed is a simple matter of filtering out the words that
+already exist, using the Scrabble dictionary. XXX\% of the words
+attempted in Scribble and XXX\% in Muddle were not real. The most
+frequent ones appear in Figure~\ref{fig:mostfrequentfake}.
+
+\begin{figure}
+\caption{Someone needs to fill this in with the data.}
+\label{fig:mostfrequent}
+\end{figure}
+
+\begin{figure}
+\caption{Someone needs to fill this in with the data.}
+\label{fig:mostfrequentfake}
+\end{figure}
+
+XXX I need to write an analysis of the list and put it here.
+
+In my opinion these are pretty good candidates for words to exist.
+Probably if you were playing someone really intense in Scrabble, and
+he or she played one of these, and was super deadpan about it and
+maybe had caused some crying contests before, and a known
+sesquipedalianist, you would let these fly because they look like real
+words to me. A point in their favor is that they are quite low-scoring
+words in Scrabble; one might expect that the words most wished for are
+like ``qzkwv'' for the scoring potential. The effect is probably due
+to a few factors: Players are less likely to attempt obvious
+non-words, common letters appear more often on the rack and on the
+board and so the opportunity to play words like in
+Figure~\ref{fig:mostfrequentfake} presents itself more frequently, and
+in Muddle, there is no advantage to using unusual letters, except the
+joy of being a weirdo. Nonetheless, this list is surely biased by the
+specifics of Scribble and Muddle, and the question at hand is not just
+what words ought to exist for the purpose of internet word games, but
+for general purposes.
+
+\begin{figure}
+\caption{Someone needs a diagram or table here of the count histogram.}
+\label{fig:gamedistribution}
+\end{figure}
+
+Another downside is that this method completely ignores the many words
+that are attempted only once or a small number of times. The
+distribution is heavy-tailed; XXX\% of the words attempted are not
+even in the top XXX (Figure~\ref{fig:gamedistribution}). Though those
+words individually are not good candidates to exist, like tiny stars
+wished upon in the night sky,\!\footnote{Astronomers now agree that
+stars do exist, by the way.} in aggregate they form a significant
+planetarium that may tell us what {\it kind} of words people wish
+existed. For example, if we saw that the words ``sweeeeeeet'',
+``sweeeeeeeeeeeeet'', ``sweeeet'' and ``sweeeeeeeeeeeeeet'' occurred a
+few times each, we could infer that people wished that words like
+``sweet'' with strictly more than two {\it e}s were real words. They
+might even be indifferent to the absolute number of {\it e}s, as long
+as there existed some legal variation with more than two {\it e}s.
+(This appears to be borne out by data. According to Google's
+estimates, the words ``swe$^n$t'' for various medium-sized $n$
+(10--20) appear on the Internet with similar frequency. The only
+exception is ``sweeeeeeeeeeeeeeeeeeet'', with 19 {\it e}s, which
+unexpectedly appears three times as often as 18 or 20 {\it e}s does.)
+In order to lance these two boils, in the next section I explore
+statistical methods for generalizing from lots of individual examples.
+
+% XXX should really try to get this in the right column, since it
+% overflows the margins
+\begin{figure}
+\begin{tabular}{r|r|l}
+17,900,000 & 0 & swt \\
+1,060,000 & 1 & swet \\
+580,000,000 & 2 & sweet \\
+1,310,000 & 3 & sweeet \\
+806,000 & 4 & sweeeet \\
+509,000 & 5 & sweeeeet \\
+283,000$^1$ & 6 & sweeeeeet \\  % spell correction for five
+170,000 & 7 & sweeeeeeet \\
+115,000 & 8 & sweeeeeeeet \\
+75,200 & 9 &  sweeeeeeeeet \\
+94,300$^2$ & 10 & sweeeeeeeeeet \\ % spell correction for nine?
+51,700 & 11 & sweeeeeeeeeeet \\
+37,900 & 12 & sweeeeeeeeeeeet \\
+32,000 & 13 & sweeeeeeeeeeeeet \\
+25,300 & 14 & sweeeeeeeeeeeeeet \\
+24,300 & 15 & sweeeeeeeeeeeeeeet \\
+41,000$^3$ & 16 & sweeeeeeeeeeeeeeeet \\ % spell correction for 14
+55,000 & 17 & sweeeeeeeeeeeeeeeeet \\
+45,000 & 18 & sweeeeeeeeeeeeeeeeeet \\
+133,000$^4$ & 19 & sweeeeeeeeeeeeeeeeeeet \\ % for 15
+34,800 & 20 &  sweeeeeeeeeeeeeeeeeeeet \\
+%
+16,100$^5$ & 25 & sweeeeeeeeeeeeeeeeeeeeeeeeet \\ % for weeeeeeeeeeeeeeeeeeeeeeeee t
+10,100 & 30 & sweeeeeeeeeeeeeeeeeeeeeeeeeeeeeet \\
+2,800 & 40 &  sweeeeeeeeeeeeeeeeeeeeeeeeeeeeee\ldots t \\
+923 & 50 &    sweeeeeeeeeeeeeeeeeeeeeeeeeeeeee\ldots t \\
+118 & 75 &    sweeeeeeeeeeeeeeeeeeeeeeeeeeeeee\ldots t \\
+38 & 100 &    sweeeeeeeeeeeeeeeeeeeeeeeeeeeeee\ldots t \\
+?$^6$ & 200 &     sweeeeeeeeeeeeeeeeeeeeeeeeeeeeee\ldots t \\ % word is too long for Google
+\end{tabular}
+\caption{Frequency of ``swe$^n$t'' on the internet for various $n$, estimated
+by Google. Notes: {\bf (1)} Spell correction offered for ``sweeeeet''. {\bf (2, 3, 4)} Spell corrections
+offered to e$^{9}$, e$^{14}$ and e$^{15}$ respectively.
+{\bf (5)} Spell correction offered for ``weeeeeeeeeeeeeeeeeeeeeeeee t'' (?) {\bf (6)} With two hundred {\it e}s, the word is too long for Google, which asks me to ``try using a shorter word.'' Thanks Google, but I already did try the shorter ones.}
+\label{ref:sweet}
+\end{figure}
+
+\section{Statistical Models}
+
+The reason that people are more likely to play words like ``tane'' is
+that the letters are common---they appear more often in words, and
+more often in the Scrabble bag. But it's not simply a matter of the
+frequency of letters; if it were, we would expect to see words like
+``eee'' dominating the list, since {\it e} is the most common letter
+in English.\!\footnote{Tied for first place with {\it n}, {\it g},
+{\it l}, {\it i}, {\it s}, and {\it h}.} People do not play such words
+often because they do not {\it seem} like real words. ``oilsoap''
+seems more like a word than ``ioaopsl'' to most non-crazy people,
+even though they contain the same letters. This is because we have
+expectations on what letters are likely to appear next to one
+another in words. This section is about modeling expectations on
+what letters appear together, and then using that model to generate
+the most likely words that don't yet exist.
+
+{\bf Markov chains.}\,
+This guy called Andrei Markov had an idea which is pretty obvious in
+retrospect, but he had it like a hundred years ago before any of us
+were born (probably; if not: you are old), which he didn't call Markov
+chains but now they're called Markov chains because I guess in the
+hopes that contemporary mathematicians will get stuff named after
+their dead selves if they keep the tradition of naming stuff after
+dead people alive. The idea is easiest to understand in the context
+of the current problem. Suppose we know that the words ``hello'',
+``helpful'' and ``felafel'' are the only real words. The following
+is a frequency table of how often each letter occurs.
+
+\vspace{0.2in}
+\begin{centering}
+\begin{tabular}{|r|r|r|r|r|r|r|r|} % {helopfua}
+\hline
+h   & e   & l   & o   & p   & f   & u   & a    \\
+\hline
+2   & 4   & 6   & 1   & 1   & 3   & 1   & 1    \\
+\hline
+\end{tabular}
+\end{centering}
+\vspace{0.2in}
+
+This tells us that {\it l} is by far the most common letter, so the
+most likely word is probably ``l'' or ``llllllll'' or something. A
+Markov chain is like a frequency table, but instead of counting
+individual letters, we count how often one letter comes {\it after}
+another. Here is the Markov chain for those words.
+
+% of the current problem. Suppose we know that the words ``hello'',
+% ``helpful'' and ``felafel'' are the only real words. The following
+
+\vspace{0.2in}
+\begin{centering}
+\begin{tabular}{|c|r|r|r|r|r|r|r|r|} % {helopfua}
+\hline
+\,  &  {\bf h}   & {\bf e}   & {\bf l}   & {\bf o}   & {\bf p}   & {\bf f}   & {\bf u}   & {\bf a} \\
+\hline  %    h     e     l     o     p     f     u     a
+{\bf h}   &  0   & 0   & 0   & 0   & 0   & 0   & 0   & 0    \\
+\hline
+{\bf e}   &  2   & 0   & 0   & 0   & 0   & 2   & 0   & 0    \\
+\hline
+{\bf l}   &  0   & 4   & 1   & 0   & 0   & 0   & 1   & 0    \\
+\hline
+{\bf o}   &  0   & 0   & 1   & 0   & 0   & 0   & 0   & 0    \\
+\hline
+{\bf p}   &  0   & 0   & 1   & 0   & 0   & 0   & 0   & 0    \\
+\hline
+{\bf f}   &  0   & 0   & 0   & 0   & 1   & 0   & 0   & 1    \\
+\hline
+{\bf u}   &  0   & 0   & 0   & 0   & 0   & 1   & 0   & 0    \\
+\hline
+{\bf a}   &  0   & 0   & 1   & 0   & 0   & 0   & 0   & 0    \\
+\hline
+\end{tabular}
+\end{centering}
+\vspace{0.2in}
+
+most people remember induction, perhaps
+
+coinduction: never gonna give you up. once you pop you can't stop.
+
+\section{Ockham's Regexp?}
+
+\section{Survey}
+
+On occasion I have been accused of ``overthinking''
+
+Rob: etsy, nuuog
+Chris: tapping on the bisphenol-A, nurm
+David: wafflucinations
+Lea: hnfff sound
+
+In order to not reprint everyone's bullshit---but not introduce bias by selectively removing data---I discarded random subsets of the data until it did not contain bullshit any more.
+
+\section{Recommendations}
+
+sweeeeeeeeeeeeeeeeeeet with 19 {\it e}s.
+
+% \begin{Verbatim}
 facebook wall posts, 4-grams
@@
 0.000601564 k
+% \end{Verbatim}
+
+% Paper ends with a bigram summary of itself.
+
+% Paper ends with a word that is obviously missing a final 's'.
+
+\end{document}

### 2011-03-06 trunk/wishlist/paper.tex
made some more progress

--- 2011-03-06 wishlist/paper.tex
@@
 }
+
+\renewcommand\>{$>$}
+\newcommand\<{$<$}
 
@@
 Disclaimer possibly indicated for SIGBOVIK: The ``research'' contained
-herein is 100\% legitimate. I have attempted to present it in a
-tutorial style that assumes little mathematical or computer science
-background. I have also left off the last {\it S} for {\it Savings}.
+herein is 100\% legitimate.\!\footnote{Source code is available at
+{\tt http://tom7misc.svn.
+sourceforge.net/viewvc/tom7misc/trunk/wishlist/}} I have attempted to
+present it in a tutorial style that assumes little mathematical or
+computer science background. I have also left off the last {\it S} for
+{\it Savings}.
 
@@
 \vspace{0.2in}
+
+The letters across the top are the ``previous letter'' and the ones
+across the left are the ``next letter'' and the box contains the
+corresponding count. For example, the pair ``el'' appears four times.
+(Pairs of letters are called ``bigrams'' by nerds, some nerd-poseurs,
+and Markov who I can't tell if he was a nerd by his picture, because
+he does have a pretty austere beard, but also did a lot of math.) One
+of the useful things about a Markov chain is that it lets us predict
+the next letter that we might see. For example, if we see ``half'',
+then the column labeled {\bf f} above tells us that the next letter is
+twice as often an {\it e} than a {\it u}, and that no other letters
+ever occurred. Typically we think of these as being probabilities
+inferred from our observations, so we say there's a 2/3 chance of {\it
+e} following {\it f} and a 1/3 chance of {\it u}. Now the word ``llllll''
+isn't so likely any more, because there's only a 1/4 chance of the
+next letter being {\it l} once we see {\it l}.
+
+Words are not just their interiors; it's also important what letters
+tend to start and end words. We can do this by imagining that each
+word starts and ends with some fake letters, and include those in the
+Markov chain. Let's use {\bf \<} for the start symbol and {\bf \>} for
+the end. So we pretend we observed ``\<hello\>'', ``\<helpful\>'', and
+``\<felafel\>''. Speaking of which, could you imagine if there were such
+a thing as a helpful felafel? Would you eat it? Because then it
+probably can't help you any more, except to get fat.
+
+\vspace{0.2in}
+\begin{centering}
+\begin{tabular}{|c|r|r|r|r|r|r|r|r|r|} % {helopfua}
+\hline
+\,  &  {\bf \<}    & {\bf h}   & {\bf e}   & {\bf l}   & {\bf o}   & {\bf p}   & {\bf f}   & {\bf u}   & {\bf a}    \\
+\hline  %  \<     h     e     l     o     p     f     u     a
+{\bf h} &  2  &  0   & 0   & 0   & 0   & 0   & 0   & 0   & 0    \\
+\hline    
+{\bf e} &  0  &  2   & 0   & 0   & 0   & 0   & 2   & 0   & 0    \\
+\hline    
+{\bf l} &  0  &  0   & 4   & 1   & 0   & 0   & 0   & 1   & 0    \\
+\hline    
+{\bf o} &  0  &  0   & 0   & 1   & 0   & 0   & 0   & 0   & 0    \\
+\hline    
+{\bf p} &  0  &  0   & 0   & 1   & 0   & 0   & 0   & 0   & 0    \\
+\hline    
+{\bf f} &  1  &  0   & 0   & 0   & 0   & 1   & 0   & 0   & 1    \\
+\hline    
+{\bf u} &  0  &  0   & 0   & 0   & 0   & 0   & 1   & 0   & 0    \\
+\hline    
+{\bf a} &  0  &  0   & 0   & 1   & 0   & 0   & 0   & 0   & 0    \\
+\hline    
+{\bf \>} &  0  &  0   & 0   & 2   & 1   & 0   & 0   & 0   & 0    \\
+\hline
+\end{tabular}
+\end{centering}
+\vspace{0.2in}
+
+We just added these like other letters, but since the beginning symbol
+{\bf \<} never occurs after other letters, we don't need a row for it
+(it would be all zeroes), and similarly since no letters ever follow
+{\bf \>} we don't need a column for it. Now the word ``lllll'' is
+impossible because no words start with {\it l}.
+
+\begin{verbatim}
+0.166666667 hel
+0.166666667 helpful
+0.166666667 hello
+0.083333333 felpful
+0.083333333 fello
+0.083333333 fel
+0.041666667 helafel
+0.041666667 helafelpful
+0.041666667 helafello
+\end{verbatim}
+
+% laplace smoothing?
 

### 2011-03-12 wishlist/paper.tex
decided to use gnuplot instead of illustrator for graph; will undo cdf junkj

@@
 like they truly deeply wish to break out of the shackles of the
-Official Scrabble Players Dictionary. So the first approach to
-determining what words ought to exist is to analyze the words that
-people tried to play, in order to try to extract the essence of
-word-yearning.
+Official Scrabble Players Dictionary.\!\footnote{For the analyses in
+  this paper that depend on a list of legal words, I actually use a
+  modified version of SOWPODS, which is the tournament list used in
+  Australia and the UK, and significantly more permissive than the US
+  Tournament Word List. Though the modified version is non-canonical,
+  I stuck with it because it's what's been in use on the site for
+  ten years.} So the first approach to determining what
+words ought to exist is to analyze the words that people tried to
+play, in order to try to extract the essence of word-yearning.
 
@@
 count, for each alleged word, and store those in a more convenient
-format. There were XXX total words attemped in Scrabble and XXX in
-Muddle. The most frequent ones appear in
-Figure~\ref{fig:mostfrequent}. The most frequent words are all
-legitimate words, since players have a bias towards attempting words
-that will not be rebuked by the computer. Seeing the words that people
-wish existed is a simple matter of filtering out the words that
-already exist, using the Scrabble dictionary. XXX\% of the words
-attempted in Scribble and XXX\% in Muddle were not real. The most
-frequent ones appear in Figure~\ref{fig:mostfrequentfake}.
+format. There were 3,572,226 total words attempted\footnote{Here a
+  word attempted is the major word of the play. This does not include
+  incidental words (typically two-letter ones) formed in the
+  perpendicular direction.} in Scribble and 13,727,511 in Muddle. The
+most frequent ones appear in Figure~\ref{fig:mostfrequent}. Aside from
+the one-letter ones, the most frequent words are legitimate words,
+since players have a bias towards attempting words that will not be
+rebuked by the computer.
+
+Seeing the words that people wish existed is a simple matter of
+filtering out the words that already exist, using the Scrabble
+dictionary. (I also filtered out one-letter ``words''. It is easy to
+see that no one-letter words should exist, again because of
+ambiguities created in spelling bees. Not only when literally spelling
+``bees'', but according to the official Scripps National Spelling Bee
+rules, the speller may optionally pronounce the word to be spelled
+before and after spelling it. So if ``s'' were a word, then the
+following ridiculous exchange obtains: Judge: ``S. The letter {\it s}.
+Etruscan origin.'' Speller: ``S. S. S.'' and the judge cannot tell if
+the speller meant to state the word before and after, or thinks the
+word is spelled ``sss''.) 22.3\% of the words attempted in Scribble
+and 36.8\% in Muddle were not real. The most frequent ones appear in
+Figure~\ref{fig:mostfrequentfake}.
 
 \begin{figure}
-\caption{Someone needs to fill this in with the data.}
+\begin{tabular}{|rl@{\qquad}rl|}
+\multicolumn{2}{l}{{\bf \large Scribble}} &
+\multicolumn{2}{l}{{\bf \large Muddle}} \\
+\hline
+Count & Word & Count & Word \\
+\hline
+45,605  &  a        &     20,412  &  late  \\
+42,315  &  i        &     19,405  &  rate  \\
+32,499  &  d$^*$    &     19,276  &  dear  \\
+12,981  &  in       &     19,049  &  tear  \\
+12,851  &  oe       &     19,019  &  date  \\
+12,528  &  s$^*$    &     18,771  &  lear  \\
+12,207  &  re       &     18,423  &  deal  \\
+11,159  &  tv       &     18,231  &  real  \\
+10,720  &  jo       &     18,138  &  lead  \\
+10,386  &  it       &     18,076  &  tale  \\
+10,369  &  et       &     17,969  &  lane  \\
+9,659   &  qua      &     17,956  &  sear  \\
+9,218   &  xi       &     17,570  &  read  \\
+9,099   &  go       &     17,193  &  teal  \\
+9,052   &  ow       &     17,170  &  lean  \\
+8,801   &  qat      &     17,071  &  dare  \\
+8,602   &  aa       &     16,923  &  dale  \\
+8,278   &  un       &     16,892  &  seal  \\
+8,142   &  en       &     16,806  &  sale  \\
+8,005   &  or       &     16,465  &  seat  \\
+\hline
+\end{tabular}
+\caption{Most frequently attempted words in Scribble and Muddle. Asterisks
+indicate non-words.}
 \label{fig:mostfrequent}
@@
 \begin{figure}
-\caption{Someone needs to fill this in with the data.}
+
+\begin{tabular}{|rl@{\qquad}rl|}
+\multicolumn{2}{l}{{\bf \large Scribble}} &
+\multicolumn{2}{l}{{\bf \large Muddle}} \\
+\hline
+Count & Word & Count & Word \\
+\hline
+11,159 &  tv         &     16,251 &  dane    \\
+4,003  &  ok         &     6,156  &  rane    \\
+2,862  &  iraq       &     5,603  &  sare    \\
+2,725  &  zen        &     5,576  &  nate    \\
+2,448  &  cho        &     4,863  &  mear    \\
+1,538  &  viz        &     4,750  &  cale    \\
+1,418  &  sdasda     &     4,616  &  nees    \\
+1,396  &  von        &     4,568  &  nale    \\
+1,136  &  etc        &     4,507  &  fale    \\
+878    &  int        &     4,347  &  deat    \\
+829    &  june       &     4,263  &  tean    \\
+745    &  lp         &     4,251  &  nile    \\
+719    &  zion       &     4,160  &  mens    \\
+665    &  cia        &     4,087  &  deel    \\
+661    &  jim        &     3,851  &  deam    \\
+651    &  iraqi      &     3,828  &  dana    \\
+648    &  ques       &     3,781  &  beed    \\
+542    &  que        &     3,769  &  lans    \\
+502    &  tim        &     3,725  &  tade    \\
+\hline
+\end{tabular}
+\caption{Most frequently attempted non-words in Scrabble and Muddle.}
 \label{fig:mostfrequentfake}
@@
 
-XXX I need to write an analysis of the list and put it here.
-
-In my opinion these are pretty good candidates for words to exist.
-Probably if you were playing someone really intense in Scrabble, and
-he or she played one of these, and was super deadpan about it and
-maybe had caused some crying contests before, and a known
+There's a clear difference between these two lists. The Scribble list
+is dominated by words involving difficult-to-play letters like {\it v}
+(there are no legal two-letter {\it v}-words). Most of the words would
+probably be acknowledged as real, just not legal in Scribble. The ones
+that don't already have meanings, like ``cho'' and ``int'' and ``que''
+seem to be pretty good candidates to exist. The Muddle list is all
+four-letter words (the minimum allowed length) using common letters.
+Other than the ones that are already words, like ``dane'' and ``nile''
+and ``mens'' (as in ``mens section'' or ``the powerfuel weapon kills
+hard so many mens''), these are all good candidates for words to
+exist. Probably if you were playing someone really intense in
+Scrabble, and he or she played one of these, and was super deadpan
+about it and maybe had caused some crying contests before, and a known
 sesquipedalianist, you would let these fly because they look like real
-words to me. A point in their favor is that they are quite low-scoring
-words in Scrabble; one might expect that the words most wished for are
-like ``qzkwv'' for the scoring potential. The effect is probably due
-to a few factors: Players are less likely to attempt obvious
-non-words, common letters appear more often on the rack and on the
-board and so the opportunity to play words like in
+words to me. A point in their favor is that they would be quite
+low-scoring words in Scrabble; not a {\it z} or {\it q} to be found.
+Even in the Scribble list there's no ``qzkwv'' junk. The effect is
+probably due to a few factors: Players are less likely to attempt
+obvious non-words, common letters appear more often on the rack and on
+the board and so the opportunity to play words like in
 Figure~\ref{fig:mostfrequentfake} presents itself more frequently, and
 in Muddle, there is no advantage to using unusual letters, except the
-joy of being a weirdo. Nonetheless, this list is surely biased by the
-specifics of Scribble and Muddle, and the question at hand is not just
-what words ought to exist for the purpose of internet word games, but
-for general purposes.
+joy of being a weirdo. Nonetheless, these lists are surely biased by
+the specifics of Scribble and Muddle, and the question at hand is not
+just what words ought to exist for the purpose of internet word games,
+but for general purposes.
 
@@
 that are attempted only once or a small number of times. The
-distribution is heavy-tailed; XXX\% of the words attempted are not
+distribution is heavy-tailed; of the 564,610 unique words attempted,
+501,939 of them aren't real! XXX\% of the words attempted are not
 even in the top XXX (Figure~\ref{fig:gamedistribution}). Though those
@@
 
+The letters across the top are the ``previous letter'' and the ones
+across the left are the ``next letter'' and the box contains the
+corresponding count. For example, the pair ``el'' appears four times.
+(Pairs of letters are called ``bigrams'' by nerds, some nerd-poseurs,
+and Markov who I can't tell if he was a nerd by his picture, because
+he does have a pretty austere beard, but also did a lot of math.) One
+of the useful things about a Markov chain is that it lets us predict
+the next letter that we might see. For example, if we see ``half'',
+then the column labeled {\bf f} above tells us that the next letter is
+twice as often an {\it e} than a {\it u}, and that no other letters
+ever occurred. Typically we think of these as being probabilities
+inferred from our observations, so we say there's a 2/3 chance of {\it
+e} following {\it f} and a 1/3 chance of {\it u}. Now the word ``llllll''
+isn't so likely any more, because there's only a 1/4 chance of the
+next letter being {\it l} once we see {\it l}.
+
+Words are not just their interiors; it's also important what letters
+tend to start and end words. We can do this by imagining that each
+word starts and ends with some fake letters, and include those in the
+Markov chain. Let's use {\bf \<} for the start symbol and {\bf \>} for
+the end. So we pretend we observed ``\<hello\>'', ``\<helpful\>'', and
+``\<felafel\>''. Speaking of which, could you imagine if there were such
+a thing as a helpful felafel? Would you eat it? Because then it
+probably can't help you any more, except to get fat.
+
+\vspace{0.2in}
+\begin{centering}
+\begin{tabular}{|c|r|r|r|r|r|r|r|r|r|} % {helopfua}
+\hline
+\,  &  {\bf \<}    & {\bf h}   & {\bf e}   & {\bf l}   & {\bf o}   & {\bf p}   & {\bf f}   & {\bf u}   & {\bf a}    \\
+\hline  %  \<     h     e     l     o     p     f     u     a
+{\bf h} &  2  &  0   & 0   & 0   & 0   & 0   & 0   & 0   & 0    \\
+\hline    
+{\bf e} &  0  &  2   & 0   & 0   & 0   & 0   & 2   & 0   & 0    \\
+\hline    
+{\bf l} &  0  &  0   & 4   & 1   & 0   & 0   & 0   & 1   & 0    \\
+\hline    
+{\bf o} &  0  &  0   & 0   & 1   & 0   & 0   & 0   & 0   & 0    \\
+\hline    
+{\bf p} &  0  &  0   & 0   & 1   & 0   & 0   & 0   & 0   & 0    \\
+\hline    
+{\bf f} &  1  &  0   & 0   & 0   & 0   & 1   & 0   & 0   & 1    \\
+\hline    
+{\bf u} &  0  &  0   & 0   & 0   & 0   & 0   & 1   & 0   & 0    \\
+\hline    
+{\bf a} &  0  &  0   & 0   & 1   & 0   & 0   & 0   & 0   & 0    \\
+\hline    
+{\bf \>} &  0  &  0   & 0   & 2   & 1   & 0   & 0   & 0   & 0    \\
+\hline
+\end{tabular}
+\end{centering}
+\vspace{0.2in}
+
+We just added these like other letters, but since the beginning symbol
+{\bf \<} never occurs after other letters, we don't need a row for it
+(it would be all zeroes), and similarly since no letters ever follow
+{\bf \>} we don't need a column for it. Now the word ``lllll'' is
+impossible because no words start with {\it l}.
+
+\begin{verbatim}
+0.166666667 hel
+0.166666667 helpful
+0.166666667 hello
+0.083333333 felpful
+0.083333333 fello
+0.083333333 fel
+0.041666667 helafel
+0.041666667 helafelpful
+0.041666667 helafello
+\end{verbatim}
+
+% laplace smoothing?
+% TODO: the empty string?
+
+TODO: most likely words in italian, too.
+
 most people remember induction, perhaps
@@
 Lea: hnfff sound
+Reed: pansepticon
+Jessica: gruntle (like not disgruntled)
 
 In order to not reprint everyone's bullshit---but not introduce bias by selectively removing data---I discarded random subsets of the data until it did not contain bullshit any more.
+
+% TODO: backformation, stripping dis- and anti- and un- from words
+% then checking to see if they're still words?
+
+% ``Scrallbe'' is kind of like god mode for scrabble.
 

### 2011-03-12 trunk/wishlist/paper.tex
progress on paper, coin

@@
 
-XXX I need to write an analysis of the list and put it here.
-
-In my opinion these are pretty good candidates for words to exist.
-Probably if you were playing someone really intense in Scrabble, and
-he or she played one of these, and was super deadpan about it and
-maybe had caused some crying contests before, and a known
+There's a clear difference between these two lists. The Scribble list
+is dominated by words involving difficult-to-play letters like {\it v}
+(there are no legal two-letter {\it v}-words). Most of the words would
+probably be acknowledged as real, just not legal in Scribble. The ones
+that don't already have meanings, like ``cho'' and ``int'' and ``que''
+seem to be pretty good candidates to exist. The Muddle list is all
+four-letter words (the minimum allowed length) using common letters.
+Other than the ones that are already words, like ``dane'' and ``nile''
+and ``mens'' (as in ``mens section'' or ``the powerfuel weapon kills
+hard so many mens''), these are all good candidates for words to
+exist. Probably if you were playing someone really intense in
+Scrabble, and he or she played one of these, and was super deadpan
+about it and maybe had caused some crying contests before, and a known
 sesquipedalianist, you would let these fly because they look like real
-words to me. A point in their favor is that they are quite low-scoring
-words in Scrabble; one might expect that the words most wished for are
-like ``qzkwv'' for the scoring potential. The effect is probably due
-to a few factors: Players are less likely to attempt obvious
-non-words, common letters appear more often on the rack and on the
-board and so the opportunity to play words like in
+words to me. A point in their favor is that they would be quite
+low-scoring words in Scrabble; not a {\it z} or {\it q} to be found.
+Even in the Scribble list there's no ``qzkwv'' junk. The effect is
+probably due to a few factors: Players are less likely to attempt
+obvious non-words, common letters appear more often on the rack and on
+the board and so the opportunity to play words like in
 Figure~\ref{fig:mostfrequentfake} presents itself more frequently, and
 in Muddle, there is no advantage to using unusual letters, except the
-joy of being a weirdo. Nonetheless, this list is surely biased by the
-specifics of Scribble and Muddle, and the question at hand is not just
-what words ought to exist for the purpose of internet word games, but
-for general purposes.
+joy of being a weirdo. Nonetheless, these lists are surely biased by
+the specifics of Scribble and Muddle, and the question at hand is not
+just what words ought to exist for the purpose of internet word games,
+but for general purposes.
 
 \begin{figure}
-\caption{Someone needs a diagram or table here of the count histogram.}
+\includegraphics[width=\linewidth]{wishlist-cdf}
+\caption{Distribution of word frequency. Approximately 25,000
+  different words (y axis) were issued 55 times or fewer (x axis).}
 \label{fig:gamedistribution}
@@
 Another downside is that this method completely ignores the many words
-that are attempted only once or a small number of times. The
-distribution is heavy-tailed; XXX\% of the words attempted are not
-even in the top XXX (Figure~\ref{fig:gamedistribution}). Though those
-words individually are not good candidates to exist, like tiny stars
-wished upon in the night sky,\!\footnote{Astronomers now agree that
-stars do exist, by the way.} in aggregate they form a significant
-planetarium that may tell us what {\it kind} of words people wish
-existed. For example, if we saw that the words ``sweeeeeeet'',
+that are attempted only once or a small number of times. Players are
+very creative; of the 564,610 unique words attempted, 501,939 of them
+aren't real! The vast majority of words are attempted only a handful
+of times (Figure~\ref{fig:gamedistribution}). Though those words
+individually are not good candidates to exist, like tiny stars wished
+upon in the night sky,\!\footnote{Astronomers now agree that stars do
+  exist, by the way.} in aggregate they form a significant planetarium
+that may tell us what {\it kind} of words people wish existed. For
+example, if we saw that the words ``sweeeeeeet'',
 ``sweeeeeeeeeeeeet'', ``sweeeet'' and ``sweeeeeeeeeeeeeet'' occurred a
@@
 exception is ``sweeeeeeeeeeeeeeeeeeet'', with 19 {\it e}s, which
-unexpectedly appears three times as often as 18 or 20 {\it e}s does.)
-In order to lance these two boils, in the next section I explore
-statistical methods for generalizing from lots of individual examples.
+unexpectedly appears three times as often as 18 or 20 {\it e}s does;
+see Figure~\ref{fig:sweet}.) In order to lance these two boils, in the
+next section I explore statistical methods for generalizing from lots
+of individual examples.
 
@@
 16,100$^5$ & 25 & sweeeeeeeeeeeeeeeeeeeeeeeeet \\ % for weeeeeeeeeeeeeeeeeeeeeeeee t
-10,100 & 30 & sweeeeeeeeeeeeeeeeeeeeeeeeeeeeeet \\
-2,800 & 40 &  sweeeeeeeeeeeeeeeeeeeeeeeeeeeeee\ldots t \\
-923 & 50 &    sweeeeeeeeeeeeeeeeeeeeeeeeeeeeee\ldots t \\
-118 & 75 &    sweeeeeeeeeeeeeeeeeeeeeeeeeeeeee\ldots t \\
-38 & 100 &    sweeeeeeeeeeeeeeeeeeeeeeeeeeeeee\ldots t \\
-?$^6$ & 200 &     sweeeeeeeeeeeeeeeeeeeeeeeeeeeeee\ldots t \\ % word is too long for Google
+10,100 & 30 & sweeeeeeeeeeeeeeeeeeeeeeeeee\ldots t \\
+2,800 & 40 &  sweeeeeeeeeeeeeeeeeeeeeeeeee\ldots t \\
+923 & 50 &    sweeeeeeeeeeeeeeeeeeeeeeeeee\ldots t \\
+118 & 75 &    sweeeeeeeeeeeeeeeeeeeeeeeeee\ldots t \\
+38 & 100 &    sweeeeeeeeeeeeeeeeeeeeeeeeee\ldots t  \\
+?$^6$ & 200 & sweeeeeeeeeeeeeeeeeeeeeeeeee\ldots t     \\ % word is too long for Google
 \end{tabular}
@@
 {\bf (5)} Spell correction offered for ``weeeeeeeeeeeeeeeeeeeeeeeee t'' (?) {\bf (6)} With two hundred {\it e}s, the word is too long for Google, which asks me to ``try using a shorter word.'' Thanks Google, but I already did try the shorter ones.}
-\label{ref:sweet}
+\label{fig:sweet}
 \end{figure}
@@
 
-The reason that people are more likely to play words like ``tane'' is
+The reason that people are more likely to play words like ``rane'' is
 that the letters are common---they appear more often in words, and
@@
 
-\begin{verbatim}
-0.166666667 hel
-0.166666667 helpful
-0.166666667 hello
-0.083333333 felpful
-0.083333333 fello
-0.083333333 fel
-0.041666667 helafel
-0.041666667 helafelpful
-0.041666667 helafello
-\end{verbatim}
+It basically makes sense to consider the probability of a whole word
+to be the chance of simultaneously seeing each pair of letters in it,
+which is just the product. So the word ``hel'' is $2/3$ (for {\bf
+  \<h}) $\times$ 2/2 (for {\bf he}) $\times$ 4/4 (for {\bf el}) $\times$ 2/6 (for
+{\bf l\>}), which is $0.222$. These are the most likely words overall (I
+discuss how to generate such lists in Section~\ref{sec:coin}):
+
+\vspace{0.2in}
+\begin{tabular}{rl@{\quad}rl}
+22.2\%  &  hel      &      2.5\%  &  helpfel  \\
+11.1\%  &  helo     &      2.5\%  &  helafel  \\
+ 7.4\%  &  fel      &      1.9\%  &  fulo     \\
+ 3.7\%  &  hell     &      1.9\%  &  hello    \\
+ 3.7\%  &  felo     &      1.2\%  &  fell     \\
+ 3.7\%  &  ful      &      1.2\%  &  helafelo \\
+\end{tabular}
+\vspace{0.2in}
+
+This is pretty good. These words resemble the ones we observed to build
+the Markov chain, but are novel. I think helafelo is a pretty rad word,
+right?
+
+The next step is to build a Markov chain for a list of real words and
+see what results.
 
 % laplace smoothing?
+% TODO: the empty string?
+
+TODO: most likely words in italian, too.
 
@@
 
-\section{Ockham's Regexp?}
+\subsection{Coining words with coinduction} \label{sec:coin}
+
+% \section{Ockham's Regexp?}
 

### 2011-03-12 trunk/wishlist/paper.tex
this almost works...

@@
 to be the chance of simultaneously seeing each pair of letters in it,
-which is just the product. So the word ``hel'' is $2/3$ (for {\bf
-  \<h}) $\times$ 2/2 (for {\bf he}) $\times$ 4/4 (for {\bf el}) $\times$ 2/6 (for
-{\bf l\>}), which is $0.222$. These are the most likely words overall (I
-discuss how to generate such lists in Section~\ref{sec:coin}):
+which is just the product of all the probabilities. So the word
+``hel'' is $2/3$ (for {\bf \<h}) $\times$ 2/2 (for {\bf he}) $\times$
+4/4 (for {\bf el}) $\times$ 2/6 (for {\bf l\>}), which is $0.222$.
+These are the most likely words overall (I discuss how to generate
+such lists in Section~\ref{sec:coin}):
 
 \vspace{0.2in}
-\begin{tabular}{rl@{\quad}rl}
+\begin{tabular}{rl@{\quad\quad}rl}
 22.2\%  &  hel      &      2.5\%  &  helpfel  \\
@@
 The next step is to build a Markov chain for a list of real words and
-see what results.
+see what results. I built one for the SOWPODS word list, and these
+are the most likely words, with real words filtered out:
+
+% XXX improve layout
+\vspace{0.2in}
+\begin{tabular}{rl@{\quad\quad}rl}
+0.049901545  & s     &   0.001768358  & y   \\
+0.017526947  & d     &   0.001761301  & p   \\
+0.009521421  & g     &   0.001653041  & a   \\
+0.005506083  & c     &   0.001652864  & n   \\
+0.004376836  & r     &   0.001563810  & ps  \\
+0.004218223  & t     &   0.001383503  & ms  \\
+0.004088377  & e     &   0.001371545  & ts  \\
+0.003560976  & m     &   0.001327023  & ds  \\
+0.003248559  & ss    &   0.001188715  & hy  \\
+0.002001529  & rs    &   0.001166457  & k   \\
+0.001958674  & h     &   0.001160619  & ng  \\
+0.001873761  & l     &   0.001142785  & ly  \\
+\end{tabular}
+\vspace{0.2in}
+
+Ugh, poop city! Actually, it turns out that when you see enough words,
+you see enough pairs that all sorts of junk looks likely. For example,
+``ng'' is easily explained by many words starting with {\it n}, {\it
+g} often following {\it n}, and many words ending with {\it g}. Even
+though each pair makes sense, the whole thing doesn't look like a word,
+because we expect to at least see a vowel at some point, for one thing.
+
+There is a standard solution to this problem, which is to generalize
+the Markov chain to keep more than one letter of history. So instead
+of just tallying how often {\it g} follows {\it n}, we count how often
+{\it g} follows {\it in} (and any other pair of letters). This makes
+the table pretty large, but I went through some trouble to put a
+graphic in Figure~\ref{fig:sowpodstrigrams} because it looks pretty
+cool.
 
@@
 coinduction: never gonna give you up. once you pop you can't stop.
+
+% word list from Italian Scrabble, which I bet is called
+% {Scrabbizzimo!}
 
@@
 
+% TODO: portmanteautally
+
 % ``Scrallbe'' is kind of like god mode for scrabble.
@@
 
-sweeeeeeeeeeeeeeeeeeet with 19 {\it e}s.
+sweeeeeeeeeeeeeeeeeeet with 19 {\it e}s, which means ``Really sweet.''
+
+``rane'' sounds too much like ``rain'', but ``sare'' has a unique pronunciation and many people seem to think it's already a word.
+``cho'' is similarly easy to pronounce and spell.
+I propose that it be defined as ``A kind of cheese,'' so that we can really nail the new triple entendre on the classic joke.
 

### 2011-03-13 wishlist/paper.tex
more on paper, screenshots etc.

@@
 {\noindent \small {\bf Keywords}:
- cryptolexicography, n-Markov models, coinduction
+ computational cryptolexicography, n-Markov models, coinduction
 }
@@
 Scrabble word.\!\footnote{There are actually no 7-letter words that
-can be made from these letters. Don't even bother. Even if playing off
-an existing letter on the board, the best we can do are the non-bingos
-``topsoil,'' ``topsail,'' or ``poloist'' with an available {\it t}.}
-This naturally made me frustrated and I wanted to do something about
-it. Why can't ``oilsoap'' be a word? Or ``loopsia''? Words are
-introduced into the lexicon all the time. My first reaction of course
-was to make an online version of Scrabble where all words are legal.
-This is available at {\tt http://snoot.org/toys/scrallbe}, and is
-pretty boring, I gotta be honest.
-% TODO: Screenshot of Scrallbe
+  can be made from these letters. Don't even bother. Even if playing
+  off an existing letter on the board, the best we can do are the
+  non-bingos ``topsoil,'' ``topsail,'' or ``poloist'' with an
+  available {\it t}.} This naturally made me frustrated and I wanted
+to do something about it. Why can't ``oilsoap'' be a word? Or
+``loopsia''? Words are introduced into the lexicon all the time. My
+first reaction of course was to make an online version of Scrabble
+where all words are legal. This is called Scrallbe (where they can
+{\it all be} words!\footnote{As of 2011, the official Scrabble slogan
+  is ``every word's a winner!'' which is clearly false.}) This is
+available at {\tt http://snoot.org/toys/scrallbe}, and is pretty
+boring, I gotta be honest (Figure~\ref{fig:scrallbe}).
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.75 \linewidth]{scrallbe-screenshot}
+\end{center}\vspace{-0.1in}
+\caption{In-progress Scrallbe game, 753 points.}
+\label{fig:scrallbe}
+\end{figure}
 
@@
 \begin{figure}
+\begin{center}
 \begin{tabular}{|rl@{\qquad}rl|}
@@
 \end{tabular}
+\end{center}
 \caption{Most frequently attempted words in Scribble and Muddle. Asterisks
@@
 \begin{figure}
-
+\begin{center}
 \begin{tabular}{|rl@{\qquad}rl|}
@@
 \end{tabular}
+\end{center}
 \caption{Most frequently attempted non-words in Scrabble and Muddle.}
@@
 \begin{figure}
-\caption{Someone needs a diagram or table here of the count histogram.}
+\includegraphics[width=\linewidth]{wishlist-cdf}
+\caption{Distribution of word frequency. Approximately 25,000
+  different words (y axis) were issued 55 times or fewer (x axis).}
 \label{fig:gamedistribution}
@@
 Another downside is that this method completely ignores the many words
-that are attempted only once or a small number of times. The
-distribution is heavy-tailed; of the 564,610 unique words attempted,
-501,939 of them aren't real! XXX\% of the words attempted are not
-even in the top XXX (Figure~\ref{fig:gamedistribution}). Though those
-words individually are not good candidates to exist, like tiny stars
-wished upon in the night sky,\!\footnote{Astronomers now agree that
-stars do exist, by the way.} in aggregate they form a significant
-planetarium that may tell us what {\it kind} of words people wish
-existed. For example, if we saw that the words ``sweeeeeeet'',
+that are attempted only once or a small number of times. Players are
+very creative; of the 564,610 unique words attempted, 501,939 of them
+aren't real! The vast majority of words are attempted only a handful
+of times (Figure~\ref{fig:gamedistribution}). Though those words
+individually are not good candidates to exist, like tiny stars wished
+upon in the night sky,\!\footnote{Astronomers now agree that stars do
+  exist, by the way.} in aggregate they form a significant planetarium
+that may tell us what {\it kind} of words people wish existed. For
+example, if we saw that the words ``sweeeeeeet'',
 ``sweeeeeeeeeeeeet'', ``sweeeet'' and ``sweeeeeeeeeeeeeet'' occurred a
@@
 
-\vspace{0.2in}
-\begin{centering}
+\begin{center}
 \begin{tabular}{|r|r|r|r|r|r|r|r|} % {helopfua}
@@
 \end{tabular}
-\end{centering}
-\vspace{0.2in}
+\end{center}
 
@@
 
-\vspace{0.2in}
-\begin{centering}
+\begin{center}
 \begin{tabular}{|c|r|r|r|r|r|r|r|r|} % {helopfua}
@@
 
-\vspace{0.2in}
-\begin{centering}
+\begin{center}
 \begin{tabular}{|c|r|r|r|r|r|r|r|r|r|} % {helopfua}
@@
 
+It basically makes sense to consider the probability of a whole word
+to be the chance of simultaneously seeing each pair of letters in it,
+which is just the product of all the probabilities. So the word
+``hel'' is $2/3$ (for {\bf \<h}) $\times$ 2/2 (for {\bf he}) $\times$
+4/4 (for {\bf el}) $\times$ 2/6 (for {\bf l\>}), which is $0.222$.
+These are the most likely words overall (I discuss how to generate
+such lists in Section~\ref{sec:coin}):
+
+\begin{center}
+\begin{tabular}{rl@{\quad\quad}rl}
+22.2\%  &  hel      &      2.5\%  &  helpfel  \\
+11.1\%  &  helo     &      2.5\%  &  helafel  \\
+ 7.4\%  &  fel      &      1.9\%  &  fulo     \\
+ 3.7\%  &  hell     &      1.9\%  &  hello    \\
+ 3.7\%  &  felo     &      1.2\%  &  fell     \\
+ 3.7\%  &  ful      &      1.2\%  &  helafelo \\
+\end{tabular}
+\end{center}
+
+This is pretty good. These words resemble the ones we observed to build
+the Markov chain, but are novel. I think helafelo is a pretty rad word,
+right?
+
+\begin{figure}
+\includegraphics[width=\linewidth]{coin1}
+\caption{Markov chain for the SOWPODS word list, where darker squares
+indicate higher probability. The darkest is the transition from {\it q}
+to {\it u} ($98\%$), which is not surprising.}
+\label{fig:sowpodsbigrams}
+\end{figure}
+
+The next step is to build a Markov chain for a list of real words and
+see what results. I built one for the SOWPODS word list, which results
+in the table in Figure~\ref{fig:sowpodsbigrams}. These
+are the most likely words, with real words filtered out:
+
+% XXX improve layout
+\begin{center}
+\begin{tabular}{rl@{\quad\quad}rl}
+0.049901545  & s     &   0.001768358  & y   \\
+0.017526947  & d     &   0.001761301  & p   \\
+0.009521421  & g     &   0.001653041  & a   \\
+0.005506083  & c     &   0.001652864  & n   \\
+0.004376836  & r     &   0.001563810  & ps  \\
+0.004218223  & t     &   0.001383503  & ms  \\
+0.004088377  & e     &   0.001371545  & ts  \\
+0.003560976  & m     &   0.001327023  & ds  \\
+0.003248559  & ss    &   0.001188715  & hy  \\
+0.002001529  & rs    &   0.001166457  & k   \\
+0.001958674  & h     &   0.001160619  & ng  \\
+0.001873761  & l     &   0.001142785  & ly  \\
+\end{tabular}
+\end{center}
+
+Ugh, poop city! Actually, it turns out that when you see enough words,
+you see enough pairs that all sorts of junk looks likely. For example,
+``ng'' is easily explained by many words starting with {\it n}, {\it
+g} often following {\it n}, and many words ending with {\it g}. Even
+though each pair makes sense, the whole thing doesn't look like a word,
+because we expect to at least see a vowel at some point, for one thing.
+
+There is a standard solution to this problem, which is to generalize
+the Markov chain to keep more than one letter of history. So instead
+of just tallying how often {\it g} follows {\it n}, we count how often
+{\it g} follows {\it in} (and any other pair of letters).\!\footnote{
+  The details are straightforward, except possibly that we now imagine
+  each word to start with two (or in general, $n$) copies of the start
+  symbol, so that we see ``\<\<helpful\>''. The column corresponding
+  to the history {\bf \<\<} tells us the frequency of letters that
+  start words, and for example the column {\bf \<h} tells us the
+  frequency of letters that follow {\it h} when it appears at the
+  start of a word. We do not need to repeat the ending character \>
+  because once we see it, we never do anything but end the word.} This
+makes the table pretty large, so you'll just have to look at
+Figure~\ref{fig:sowpodsbigrams} again and imagine it being 28 times
+wider. But the good news is that it invents much better words:
+
 \begin{verbatim}
-0.166666667 hel
-0.166666667 helpful
-0.166666667 hello
-0.083333333 felpful
-0.083333333 fello
-0.083333333 fel
-0.041666667 helafel
-0.041666667 helafelpful
-0.041666667 helafello
+Markov chain with n=2.
+100 top paths:
+0.007092576 ing
+0.002483309 ses
+0.001694275 des
+0.001548279 nes
+0.001403047 sts
+0.001314279 se
+0.001287353 ings
+0.001262904 ded
+0.001170356 cal
+0.001101124 le
+0.001078157 der
+0.001070769 ove
+0.001016514 gly
+0.000882527 hy
+0.000853499 ung
+0.000838535 cy
+0.000816757 pres
+0.000802384 pers
+0.000787731 ps
+0.000771953 ders
+0.000746855 ly
+0.000732157 ines
+0.000709460 ve
+0.000650273 sm
+0.000650062 ty
+0.000642246 pred
+0.000627432 ble
+0.000622587 hed
+0.000603012 ter
+0.000587263 gy
+0.000564629 sms
+0.000549088 py
+0.000545497 ans
+0.000541832 ber
+0.000531745 ce
+0.000489139 ang
+0.000487316 unt
+0.000482908 snes
 \end{verbatim}
+
+\begin{verbatim}
+Markov chain with n=3.
+100 top paths:
+0.001093377 des
+0.000789743 pers
+0.000767821 cal
+0.000622381 pres
+0.000452415 nons
+0.000449855 ress
+0.000426892 ing
+0.000404454 pred
+0.000387570 ent
+0.000366753 dist
+0.000357115 ble
+0.000353576 ches
+0.000348642 gly
+0.000345564 inted
+0.000340691 dists
+0.000339750 lity
+0.000336468 und
+0.000331914 dises
+0.000331389 int
+0.000320104 ints
+0.000291846 ching
+0.000284350 tely
+0.000284228 dise
+0.000279669 ster
+0.000279431 ves
+0.000276676 stic
+0.000272655 tric
+0.000263922 cally
+0.000261415 ter
+0.000260898 ents
+0.000258270 rous
+0.000253990 dising
+0.000245834 phy
+0.000244272 dised
+0.000240505 hous
+0.000239726 resses
+0.000237100 sters
+0.000236753 ove
+0.000230976 dism
+0.000227701 ched
+0.000225104 fing
+0.000222993 ence
+0.000221624 ters
+0.000219511 disms
+0.000219246 unders
+\end{verbatim}
+
+\begin{verbatim}
+Markov chain with n=4.
+100 top paths:
+0.000454575 unders
+0.000341807 dising
+0.000293774 pers
+0.000285248 cally
+0.000232842 inted
+0.000204856 heter
+0.000198642 tric
+0.000188363 ster
+0.000185238 hier
+0.000183216 unded
+0.000174876 heters
+0.000169220 sters
+0.000153800 stic
+0.000142753 pering
+0.000132595 dises
+0.000131040 ching
+0.000129387 shing
+0.000125442 dest
+0.000115038 teless
+0.000113312 resis
+0.000112909 dist
+0.000112793 dists
+0.000112397 mising
+0.000109589 sation
+0.000109555 fing
+0.000108931 dised
+0.000107632 ches
+0.000102999 cated
+0.000101327 dise
+0.000097730 sations
+0.000096676 inding
+0.000095270 nons
+0.000093852 lary
+0.000093357 ency
+0.000091199 ther
+0.000090608 ories
+0.000090405 alled
+0.000088604 apped
+0.000088235 hous
+0.000087780 trous
+0.000082438 undered
+0.000081713 noning
+0.000079613 dism
+0.000079071 thers
+0.000078380 ented
+0.000076765 irred
+0.000076363 disms
+0.000074970 geness
+0.000074910 veness
+0.000071079 vities
+0.000070308 mised
+0.000068585 ques
+0.000068096 unding
+0.000067618 sher
+0.000067429 peries
+\end{verbatim}
+
+\begin{verbatim}
+n=5
+GetTempFileName failed with error 5
+\end{verbatim}
+
+Wikipedia favorites: smally (.000287),
+websity (.000156),
+stude (.000156),
+chool (.000124),
+fontry (.000120),
+undex (.000102),
+octory (.0000990),
+coibot (.0000960),
+footnot (.0000837),
+reporth (.0000518),
+delection (.0000484),
+grounty (.0000459),
+betweek (.0000437),
+fination (.0000431),
+manuary (.0000388),
+whicle (.0000360),
+stategory (.0000262),
 
@@
 
+% word list from Italian Scrabble, which I bet is called
+% {Scrabbizzimo!}
+
+\subsection{Coining words with coinduction} \label{sec:coin}
+
+In the earlier sections I blithely produced tables of the most
+probable words according to an $n$-Markov chain. It is not obvious how
+to do this (or that it is even possible), so I explain the algorithm
+in this section. It's safely skippable, I mean if you don't want to
+know about a pretty cool algorithm that's not that complicated and
+might even be new, plus {\em dual math}.
+
+Computing the probability of an individual word is easy. We prefix it
+with $n$ copies of the start symbol \<, suffix it with a single \>,
+and then look up the probability of each symbol given its $n$
+preceding symbols in the table, and multiply those all together. We
+can compute the probability of any word this way. The problem with
+sorting all of the possible words by their probabilities is that there
+are an infinite number of them. We can't just look at short words
+first, either, because for example the word ``thethethe'' is many
+times more likely ($p = 6.08\times 10^{-11}$) than the shorter
+``qatzs'' ($9.07\times 10^{-12}$).
+
+The solution is to use coinduction. Most people remember induction
+from school, maybe, which is the one where you have some base case
+like ``0 is even'', and then you prove that all numbers are either
+even or odd by assuming ``$n$ is even or odd'' and proving ``$n + 1$
+is even or odd''.
+
 most people remember induction, perhaps 
-
 coinduction: never gonna give you up. once you pop you can't stop.
-
-\section{Ockham's Regexp?}
 
@@
 
-On occasion I have been accused of ``overthinking''
+On occasion I have been accused of ``overthinking'' problems, whatever
+that means. So to compare, I next hazarded a tried and true technique
+from grade school, the survey.
 

### 2011-03-20 trunk/wishlist/paper.tex
first attempt at backformation

@@
 
-\vspace{0.2in}
+\begin{center}
 \begin{tabular}{rl@{\quad\quad}rl}
@@
 \end{tabular}
-\vspace{0.2in}
+\end{center}
 
@@
 
+\begin{figure}
+\includegraphics[width=\linewidth]{coin1}
+\caption{Markov chain for the SOWPODS word list, where darker squares
+indicate higher probability. The darkest is the transition from {\it q}
+to {\it u} ($98\%$), which is not surprising.}
+\label{fig:sowpodsbigrams}
+\end{figure}
+
 The next step is to build a Markov chain for a list of real words and
-see what results. I built one for the SOWPODS word list, and these
+see what results. I built one for the SOWPODS word list, which results
+in the table in Figure~\ref{fig:sowpodsbigrams}. These
 are the most likely words, with real words filtered out:
@@
 % XXX improve layout
-\vspace{0.2in}
+\begin{center}
 \begin{tabular}{rl@{\quad\quad}rl}
@@
 of just tallying how often {\it g} follows {\it n}, we count how often
-{\it g} follows {\it in} (and any other pair of letters). This makes
-the table pretty large, but I went through some trouble to put a
-graphic in Figure~\ref{fig:sowpodstrigrams} because it looks pretty
-cool.
+{\it g} follows {\it in} (and any other pair of letters).\!\footnote{
+  The details are straightforward, except possibly that we now imagine
+  each word to start with two (or in general, $n$) copies of the start
+  symbol, so that we see ``\<\<helpful\>''. The column corresponding
+  to the history {\bf \<\<} tells us the frequency of letters that
+  start words, and for example the column {\bf \<h} tells us the
+  frequency of letters that follow {\it h} when it appears at the
+  start of a word. We do not need to repeat the ending character \>
+  because once we see it, we never do anything but end the word.} This
+makes the table pretty large, so you'll just have to look at
+Figure~\ref{fig:sowpodsbigrams} again and imagine it being 28 times
+wider. But the good news is that it invents much better words:
+
+\begin{verbatim}
+Markov chain with n=2.
+100 top paths:
+0.007092576 ing
+0.002483309 ses
+0.001694275 des
+0.001548279 nes
+0.001403047 sts
+0.001314279 se
+0.001287353 ings
+0.001262904 ded
+0.001170356 cal
+0.001101124 le
+0.001078157 der
+0.001070769 ove
+0.001016514 gly
+0.000882527 hy
+0.000853499 ung
+0.000838535 cy
+0.000816757 pres
+0.000802384 pers
+0.000787731 ps
+0.000771953 ders
+0.000746855 ly
+0.000732157 ines
+0.000709460 ve
+0.000650273 sm
+0.000650062 ty
+0.000642246 pred
+0.000627432 ble
+0.000622587 hed
+0.000603012 ter
+0.000587263 gy
+0.000564629 sms
+0.000549088 py
+0.000545497 ans
+0.000541832 ber
+0.000531745 ce
+0.000489139 ang
+0.000487316 unt
+0.000482908 snes
+\end{verbatim}
+
+\begin{verbatim}
+Markov chain with n=3.
+100 top paths:
+0.001093377 des
+0.000789743 pers
+0.000767821 cal
+0.000622381 pres
+0.000452415 nons
+0.000449855 ress
+0.000426892 ing
+0.000404454 pred
+0.000387570 ent
+0.000366753 dist
+0.000357115 ble
+0.000353576 ches
+0.000348642 gly
+0.000345564 inted
+0.000340691 dists
+0.000339750 lity
+0.000336468 und
+0.000331914 dises
+0.000331389 int
+0.000320104 ints
+0.000291846 ching
+0.000284350 tely
+0.000284228 dise
+0.000279669 ster
+0.000279431 ves
+0.000276676 stic
+0.000272655 tric
+0.000263922 cally
+0.000261415 ter
+0.000260898 ents
+0.000258270 rous
+0.000253990 dising
+0.000245834 phy
+0.000244272 dised
+0.000240505 hous
+0.000239726 resses
+0.000237100 sters
+0.000236753 ove
+0.000230976 dism
+0.000227701 ched
+0.000225104 fing
+0.000222993 ence
+0.000221624 ters
+0.000219511 disms
+0.000219246 unders
+\end{verbatim}
+
+\begin{verbatim}
+Markov chain with n=4.
+100 top paths:
+0.000454575 unders
+0.000341807 dising
+0.000293774 pers
+0.000285248 cally
+0.000232842 inted
+0.000204856 heter
+0.000198642 tric
+0.000188363 ster
+0.000185238 hier
+0.000183216 unded
+0.000174876 heters
+0.000169220 sters
+0.000153800 stic
+0.000142753 pering
+0.000132595 dises
+0.000131040 ching
+0.000129387 shing
+0.000125442 dest
+0.000115038 teless
+0.000113312 resis
+0.000112909 dist
+0.000112793 dists
+0.000112397 mising
+0.000109589 sation
+0.000109555 fing
+0.000108931 dised
+0.000107632 ches
+0.000102999 cated
+0.000101327 dise
+0.000097730 sations
+0.000096676 inding
+0.000095270 nons
+0.000093852 lary
+0.000093357 ency
+0.000091199 ther
+0.000090608 ories
+0.000090405 alled
+0.000088604 apped
+0.000088235 hous
+0.000087780 trous
+0.000082438 undered
+0.000081713 noning
+0.000079613 dism
+0.000079071 thers
+0.000078380 ented
+0.000076765 irred
+0.000076363 disms
+0.000074970 geness
+0.000074910 veness
+0.000071079 vities
+0.000070308 mised
+0.000068585 ques
+0.000068096 unding
+0.000067618 sher
+0.000067429 peries
+\end{verbatim}
+
+\begin{verbatim}
+n=5
+GetTempFileName failed with error 5
+\end{verbatim}
+
+Wikipedia favorites: smally (.000287),
+websity (.000156),
+stude (.000156),
+chool (.000124),
+fontry (.000120),
+undex (.000102),
+octory (.0000990),
+coibot (.0000960),
+footnot (.0000837),
+reporth (.0000518),
+delection (.0000484),
+grounty (.0000459),
+betweek (.0000437),
+fination (.0000431),
+manuary (.0000388),
+whicle (.0000360),
+stategory (.0000262),
 
@@
 
-most people remember induction, perhaps
-
-coinduction: never gonna give you up. once you pop you can't stop.
-
 % word list from Italian Scrabble, which I bet is called
@@
 
-% \section{Ockham's Regexp?}
+In the earlier sections I blithely produced tables of the most
+probable words according to an $n$-Markov chain. It is not obvious how
+to do this (or that it is even possible), so I explain the algorithm
+in this section. It's safely skippable, I mean if you don't want to
+know about a pretty cool algorithm that's not that complicated and
+might even be new, plus {\em dual math}.
+
+Computing the probability of an individual word is easy. We prefix it
+with $n$ copies of the start symbol \<, suffix it with a single \>,
+and then look up the probability of each symbol given its $n$
+preceding symbols in the table, and multiply those all together. We
+can compute the probability of any word this way. The problem with
+sorting all of the possible words by their probabilities is that there
+are an infinite number of them. We can't just look at short words
+first, either, because for example the word ``thethethe'' is many
+times more likely ($p = 6.08\times 10^{-11}$) than the shorter
+``qatzs'' ($9.07\times 10^{-12}$).
+
+The solution is to use coinduction. Most people remember induction
+from school, maybe, which is the one where you have some base case
+like ``0 is even'', and then you prove that all numbers are either
+even or odd by assuming ``$n$ is even or odd'' and proving ``$n + 1$
+is even or odd''.
+
+most people remember induction, perhaps 
+coinduction: never gonna give you up. once you pop you can't stop.
+
+\section{Backformation}
+
+\begin{verbatim}
+Best fixes...
+1.000000000 -zzyingly    1/1
+1.000000000 -zzying      1/1
+1.000000000 -zzuolanas   1/1
+1.000000000 -zzuolana    1/1
+1.000000000 -zzotints    1/1
+1.000000000 -zzotintos   1/1
+1.000000000 -zzotinto    1/1
+1.000000000 -zzotinting  1/1
+1.000000000 -zzotinters  1/1
+1.000000000 -zzotinter   1/1
+1.000000000 -zzotinted   1/1
+1.000000000 -zzotint     1/1
+1.000000000 -zzonites    1/1
+1.000000000 -zzonite     1/1
+1.000000000 -zzoners     1/1
+1.000000000 -zzoner      1/1
+1.000000000 -zzolans     1/1
+1.000000000 -zzolanic    1/1
+1.000000000 -zzolan      1/1
+1.000000000 -zzmen       1/1
+1.000000000 -zzman       1/1
+1.000000000 -zzlike      1/1
+1.000000000 -zzini       1/1
+1.000000000 -zzicatos    1/1
+1.000000000 -zzicato     1/1
+1.000000000 -zzicati     1/1
+1.000000000 -zzian       1/1
+\end{verbatim}
+1/1 is not really much support.
+
+Ones that are only seen n/1 time never apply (because ...). Filtered:
+
+(actually we should probably be using binomial likelihood)
+
+\begin{verbatim}
+Best fixes...
+1.000000000 -zzazzes     3/3
+1.000000000 -zzazz       3/3
+1.000000000 -zzans       3/3
+1.000000000 -zzanim      2/2
+1.000000000 -zzan        3/3
+1.000000000 -zygotic     3/3
+1.000000000 -zygotes     2/2
+1.000000000 -zygote      2/2
+1.000000000 -zygosity    2/2
+1.000000000 -zygosis     2/2
+1.000000000 -zpas        2/2
+1.000000000 -zpahs       2/2
+1.000000000 -zpah        2/2
+1.000000000 -zpa         2/2
+1.000000000 -zooks       2/2
+1.000000000 -zonites     3/3
+1.000000000 -zonite      3/3
+1.000000000 -zonians     2/2
+1.000000000 -zonian      2/2
+1.000000000 -zists       2/2
+1.000000000 -zist        2/2
+\end{verbatim}
+
+Not much better!
+
+Only ones with denominator 10+ times.
+\begin{verbatim}
+Best fixes...
+1.000000000 -wrought    10/10
+1.000000000 -writings   12/12
+1.000000000 -wraps      10/10
+1.000000000 -wrap       11/11
+1.000000000 -worms      69/69
+1.000000000 -worm       69/69
+1.000000000 -working    21/21
+1.000000000 -worked     14/14
+1.000000000 -winds      17/17
+1.000000000 -wind       18/18
+1.000000000 -wide       12/12
+1.000000000 -weights    19/19
+1.000000000 -weight     22/22
+1.000000000 -waxes      12/12
+1.000000000 -wax        12/12
+1.000000000 -waters     37/37
+1.000000000 -water      41/41
+1.000000000 -walking    11/11
+1.000000000 -walkers    17/17
+1.000000000 -walker     17/17
+1.000000000 -valuing    10/10
+1.000000000 -valued     11/11
+1.000000000 -towns      11/11
+1.000000000 -tight      15/15
+1.000000000 -thought    12/12
+1.000000000 -thinks     11/11
+\end{verbatim}
+
+Much better! These are still sort of degenerate and won't help
+us find new words because they ALWAYS apply!
+
+Instead anything with p strictly less than 1.
+\begin{verbatim}
+Best fixes...
+0.985714286 -makers     69/70
+0.985714286 -maker      69/70
+0.983606557 -wood       120/122
+0.983471074 -woods      119/121
+0.982758621 -down       57/58
+0.982658960 -works      170/173
+0.981818182 -houses     108/110
+0.981818182 -house      108/110
+0.981132075 kilo-       52/53
+0.980752406 -less       1121/1143
+0.980743395 over-       2190/2233
+0.980000000 -books      49/50
+0.980000000 -book       49/50
+0.979591837 -proof      48/49
+0.979310345 -lessnesses 142/145
+0.979069767 -ships      421/430
+0.978723404 -lessness   184/188
+0.978723404 -board      138/141
+0.978494624 -woman      91/93
+0.978021978 -women      89/91
+0.977528090 -ship       435/445
+0.977272727 -manship    43/44
+0.976744186 -weeds      84/86
+0.976470588 after-      83/85
+0.976190476 -manships   41/42
+0.976190476 -making     41/42
+0.976190476 -craft      41/42
+0.976190476 -boats      41/42
+0.976190476 -boat       41/42
+\end{verbatim}
+
+Wow, now we're talking!
 
@@
 In order to not reprint everyone's bullshit---but not introduce bias by selectively removing data---I discarded random subsets of the data until it did not contain bullshit any more.
+
+% TODO: length normalization. just take the average (or max?)
+% probability for a word of length n, and normalize by that (which means what?)?
 

### 2011-03-27 skymall/paper.tex
paper and code, but not images (non-free)

\documentclass{article} % [twocolumn]
\usepackage[top=1.1in, left=0.85in, right=0.85in]{geometry}

% \usepackage{eclbkbox}
\usepackage{amsmath}
\usepackage{amssymb}
% \usepackage{amscd}
% \usepackage{xy}
\usepackage{graphicx}
% \usepackage{fancyhdr}
% \usepackage{color}
% \usepackage[dark,all,bottom,landscape,timestamp]{draftcopy}
% \usepackage{everypage}

% \usepackage{ulem}
% go back to italics for emphasis, though
% \normalem

\begin{document}

\title{Who is the biggest douche in Skymall?}
\author{Dr.~Tom~Murphy~VII~Ph.D.\thanks{
Copyright \copyright\ 2011 the Regents of the Wikiplia
Foundation. Appears in SIGBOVIK 2011 with the blessing of the
Association for Computational Heresy; {\em IEEEEEE!} press,
Verlag-Verlag volume no.~0x40-2A.
\yen 0.00}
}

\date{1 April 2011}

\maketitle

\begin{abstract}
Did you ever notice how there are so many douches in the Skymall
catalog? This paper investigates the 22 males pictured in the January
2011 issue, using Internet technology to determine their douchiness.
We then present an efficient image recognition algorithm that reliably
predicts douchiness from photos.
\end{abstract}

\vspace{1em}
{\noindent \small {\bf Keywords}:
 computational douchebaggery, as seen on tv, man who can sleep in any seat
}

\section*{Introduction}
In that Luddite void between closing the cabin doors and the beep
indicating it is now safe to use approved electronic devices, there is
one perfect pleasure: The Skymall catalog. It has everything,
including: Pointlessly impractical products you cringe at just
imagining someone receiving as unwanted gifts at Christmas, copy that
preys on the insecurities of business travelers, typo and physically
impossible hyperbole treasure hunts galore, Photoshop disasters, new
friends, and old familiar faces. But since 1990, science has wondered:
{\bf Who is the biggest douche in Skymall?}

It is difficult to assess the absolute douchiness (say, on a scale
from 1 to 10) of a given person. So, in order to answer this question,
we used Internet Technology to conduct a series of
more-douche--less-douche battles between randomly selected pairs of
participants. The visitor is simply asked: Who is the bigger douche?
The proportion of battles won, overall, is the final douche score.

After thousands of battles waged, we converged on the following results,
ranked in descending douchiness:

\section*{Results}

\newcommand\douche[6]{
  \begin{minipage}{0.332 \linewidth}
  \includegraphics[width=#1 \linewidth]{#2} \\
  \raggedright
  {\bf #3} \\
  {\small won #4/#5;} #6\% douche
  \end{minipage}
}

\douche{0.45}{personal-headphones}{In Charge of the Music}{207}{66}{75.82}
\douche{0.45}{head-spa-massage}{The Thinker}{201}{73}{73.35}
\douche{0.75}{sandproof-blanket}{Seaside Date}{197}{76}{72.16}
\douche{0.75}{laptop-holder}{Reclining Numbercruncher}{193}{80}{70.69}
\douche{0.75}{fluidity-of-motion}{Karate Genius}{188}{85}{68.86}
\douche{0.75}{poolside-read}{Poolside Bibliophile}{177}{95}{65.07}
\douche{0.75}{global-network}{Traveling Salesman}{171}{102}{62.63}
\douche{0.75}{relaxing}{Beauty Rest}{167}{106}{61.17}
\douche{0.75}{jeweler}{Jeweler}{167}{106}{61.17}
\douche{0.75}{i-have-to-take-this}{I Have to Take This}{161}{113}{58.75}
\douche{0.75}{nose-jewelry}{Cured Snorer}{141}{132}{51.64}
\douche{0.75}{sleeping}{Man who can Sleep in Any Seat}{131}{141}{48.16}
\douche{0.75}{after}{Treatment Recipent}{121}{152}{44.32}
\douche{0.55}{business-expert}{Business Expert}{118}{155}{43.22}
\douche{0.75}{confidence}{Confident in Glasses}{118}{155}{43.22}
\douche{0.75}{sass-moulavi}{Sass Moulavi, M.D.}{117}{155}{43.01}
\douche{0.55}{videochat}{New Haircut}{96}{177}{35.16}
\douche{0.45}{pockets-jacket}{Woodsy Gentleman}{86}{186}{31.61}
\douche{0.75}{john-q-storus}{John Q. Storus}{78}{195}{28.57}
\douche{0.45}{security-system}{Father/Kidnapper}{73}{199}{26.83}
\douche{0.55}{promotion}{Got the Promotion}{59}{214}{21.61}
\douche{0.45}{runner}{Silver Medalist}{34}{238}{12.50}

Although many people disagreed on the finer points of what constitutes
a douche, the results were fairly significant, in that there were many
participants that were consistently perceived as more or less douchey.
This is truly a victory for the scientific method. Speaking of
victories:

\section*{Douche recognition}

Having collected consensus on what constitutes a douche, we next turn
to the problem of determining whether any given person is a douche,
even if that person has not participated in hundreds of rounds of
douche-battle. Since people appear to be able to make decisions based
mainly on images, we look to image processing techniques.

Images are formed using ``pixels'', which are like tiny individual
color dots. Each color dot, or ``pixel'', is saved in a file. It turns
out that these files are all the same one: Every ``pixel'' is in a
file, which constitutes the series of color dots, as a series of bytes
or ``1s and 0s''~\ref{fig:ones}, which constitute the digital
information that is the file, or ``pixels.'' Point is, in order for a
computer to ``see'' a file, all it needs to do is look at it, by
rubbing the files and ``pixels'' on its CPUs, the same way that you or
I look at a picture by rubbing it on our eyeballs.

The problem with most image recognition algorithms is that they do not
work, and are also slow.

We eschew the traditional model-based approaches, instead using efficient
hashing algorithms such as SHA-1 and MD5. These operate directly on
the ``1s and 0s'' of the image, and run in linear time. This way, even
if the algorithms do not work, they will at least be fast.

We find that the algorithm SHA-1 correlates with the douchiness of the
image, but not well. The MD5 algorithm is faster but in fact correlates
negatively. However, by fine tuning the initialization vector, we are
able to produce a variant that is just as fast and correlates very well
with the data (Figure~\ref{fig:chart}).

\begin{figure*}
\begin{center}
\includegraphics[width=0.9 \linewidth]{chart}
\end{center}\vspace{-0.1in}
\caption{Performance of different prediction algorithms. SHA-1
  correlates with the douchiness of the image, though it tends to
  overestimate the douchiness of medium-low douches. MD5 actually
  has negative correlation. A tuned variant of MD5 with the
  initialization vector {\tt A8F82303 08A1B76B AA25DA9E 4C2C1883}
  correlates quite neatly.}
\label{fig:chart}
\end{figure*}

\section*{Conclusion}
Do you disagree with these data (Y/n)? If (Y), then Science never Sleeps!
{\tt http://snoot.org/toys/wuss/skymall/}

\begin{figure}
\begin{center}
\includegraphics[width=0.15 \linewidth]{ones}
\end{center}\vspace{-0.1in}
\caption{Image files consist of ``ones'' or ``zeroes'', which is
  digital information pixels (pictured).}
\label{fig:ones}
\end{figure}

\section*{Poolside Bibliography}
Sorry, I didn't read any papers or anything or do actual science.

\end{document}

### 2011-03-27 wishlist/paper.tex
notes about backformation, which is not actually implemented yet
more progres on paper. compute bigram frequencies for italian

@@
 Official Scrabble Players Dictionary.\!\footnote{For the analyses in
-  this paper that depend on a list of legal words, I actually use a
+  this section that depend on a list of legal words, I actually use a
   modified version of SOWPODS, which is the tournament list used in
@@
 \includegraphics[width=\linewidth]{wishlist-cdf}
-\caption{Distribution of word frequency. Approximately 25,000
-  different words (y axis) were issued 55 times or fewer (x axis).}
+\vspace{-0.1in}\caption{Cumulative distribution of word frequency.
+  Approximately 25,000 different words (y axis) were issued 55 times
+  or fewer (x axis). The ``total'' area does not appear much larger
+  than its components because this is a log-log plot.}
 \label{fig:gamedistribution}
@@
 
-\section{Statistical Models}
+\section{Statistical models}
 
@@
 \begin{tabular}{rl@{\quad\quad}rl}
-0.049901545  & s     &   0.001768358  & y   \\
-0.017526947  & d     &   0.001761301  & p   \\
-0.009521421  & g     &   0.001653041  & a   \\
-0.005506083  & c     &   0.001652864  & n   \\
-0.004376836  & r     &   0.001563810  & ps  \\
-0.004218223  & t     &   0.001383503  & ms  \\
-0.004088377  & e     &   0.001371545  & ts  \\
-0.003560976  & m     &   0.001327023  & ds  \\
-0.003248559  & ss    &   0.001188715  & hy  \\
-0.002001529  & rs    &   0.001166457  & k   \\
-0.001958674  & h     &   0.001160619  & ng  \\
-0.001873761  & l     &   0.001142785  & ly  \\
+4.99\%  & s     &   0.17\%  & y   \\
+1.75\%  & d     &   0.17\%  & p   \\
+0.95\%  & g     &   0.16\%  & a   \\
+0.55\%  & c     &   0.16\%  & n   \\
+0.43\%  & r     &   0.15\%  & ps  \\
+0.42\%  & t     &   0.13\%  & ms  \\
+0.40\%  & e     &   0.13\%  & ts  \\
+0.35\%  & m     &   0.13\%  & ds  \\
+0.32\%  & ss    &   0.11\%  & hy  \\
+0.20\%  & rs    &   0.11\%  & k   \\
+0.19\%  & h     &   0.11\%  & ng  \\
+0.18\%  & l     &   0.11\%  & ly  \\
 \end{tabular}
@@
 
-\begin{verbatim}
-Markov chain with n=2.
-100 top paths:
-0.007092576 ing
-0.002483309 ses
-0.001694275 des
-0.001548279 nes
-0.001403047 sts
-0.001314279 se
-0.001287353 ings
-0.001262904 ded
-0.001170356 cal
-0.001101124 le
-0.001078157 der
-0.001070769 ove
-0.001016514 gly
-0.000882527 hy
-0.000853499 ung
-0.000838535 cy
-0.000816757 pres
-0.000802384 pers
-0.000787731 ps
-0.000771953 ders
-0.000746855 ly
-0.000732157 ines
-0.000709460 ve
-0.000650273 sm
-0.000650062 ty
-0.000642246 pred
-0.000627432 ble
-0.000622587 hed
-0.000603012 ter
-0.000587263 gy
-0.000564629 sms
-0.000549088 py
-0.000545497 ans
-0.000541832 ber
-0.000531745 ce
-0.000489139 ang
-0.000487316 unt
-0.000482908 snes
-\end{verbatim}
-
-\begin{verbatim}
-Markov chain with n=3.
-100 top paths:
-0.001093377 des
-0.000789743 pers
-0.000767821 cal
-0.000622381 pres
-0.000452415 nons
-0.000449855 ress
-0.000426892 ing
-0.000404454 pred
-0.000387570 ent
-0.000366753 dist
-0.000357115 ble
-0.000353576 ches
-0.000348642 gly
-0.000345564 inted
-0.000340691 dists
-0.000339750 lity
-0.000336468 und
-0.000331914 dises
-0.000331389 int
-0.000320104 ints
-0.000291846 ching
-0.000284350 tely
-0.000284228 dise
-0.000279669 ster
-0.000279431 ves
-0.000276676 stic
-0.000272655 tric
-0.000263922 cally
-0.000261415 ter
-0.000260898 ents
-0.000258270 rous
-0.000253990 dising
-0.000245834 phy
-0.000244272 dised
-0.000240505 hous
-0.000239726 resses
-0.000237100 sters
-0.000236753 ove
-0.000230976 dism
-0.000227701 ched
-0.000225104 fing
-0.000222993 ence
-0.000221624 ters
-0.000219511 disms
-0.000219246 unders
-\end{verbatim}
-
-\begin{verbatim}
-Markov chain with n=4.
-100 top paths:
-0.000454575 unders
-0.000341807 dising
-0.000293774 pers
-0.000285248 cally
-0.000232842 inted
-0.000204856 heter
-0.000198642 tric
-0.000188363 ster
-0.000185238 hier
-0.000183216 unded
-0.000174876 heters
-0.000169220 sters
-0.000153800 stic
-0.000142753 pering
-0.000132595 dises
-0.000131040 ching
-0.000129387 shing
-0.000125442 dest
-0.000115038 teless
-0.000113312 resis
-0.000112909 dist
-0.000112793 dists
-0.000112397 mising
-0.000109589 sation
-0.000109555 fing
-0.000108931 dised
-0.000107632 ches
-0.000102999 cated
-0.000101327 dise
-0.000097730 sations
-0.000096676 inding
-0.000095270 nons
-0.000093852 lary
-0.000093357 ency
-0.000091199 ther
-0.000090608 ories
-0.000090405 alled
-0.000088604 apped
-0.000088235 hous
-0.000087780 trous
-0.000082438 undered
-0.000081713 noning
-0.000079613 dism
-0.000079071 thers
-0.000078380 ented
-0.000076765 irred
-0.000076363 disms
-0.000074970 geness
-0.000074910 veness
-0.000071079 vities
-0.000070308 mised
-0.000068585 ques
-0.000068096 unding
-0.000067618 sher
-0.000067429 peries
-\end{verbatim}
-
-\begin{verbatim}
-n=5
-GetTempFileName failed with error 5
-\end{verbatim}
+\begin{center}
+\begin{tabular}{rl@{\quad\quad}rl}
+\multicolumn{4}{c}{Markov chain with $n=2$.} \\
+\hline
+.709\% & ing     &   .110\% & le    \\
+.248\% & ses     &   .107\% & der   \\
+.169\% & des     &   .107\% & ove   \\
+.154\% & nes     &   .101\% & gly   \\
+.140\% & sts     &   .088\% & hy    \\
+.131\% & se      &   .085\% & ung   \\
+.128\% & ings    &   .083\% & cy    \\
+.126\% & ded     &   .081\% & pres  \\
+.117\% & cal     &   .080\% & pers  \\
+\end{tabular}
+\end{center}
+
+These are even, like, pronounceable. The best news is that they keep
+getting better the more history we keep:
+
+\begin{center}
+\begin{tabular}{rl@{\quad\quad}rl}
+\multicolumn{4}{c}{Markov chain with $n=3$.} \\
+\hline
+.109\% & des     &   .038\%  & ent     \\
+.078\% & pers    &   .036\%  & dist    \\
+.076\% & cal     &   .035\%  & ble     \\
+.062\% & pres    &   .035\%  & ches    \\
+.045\% & nons    &   .034\%  & gly     \\
+.044\% & ress    &   .034\%  & inted   \\
+.042\% & ing     &   .034\%  & dists   \\
+.040\% & pred    &   .033\%  & lity    \\
+\end{tabular}
+\end{center}
+
+\begin{center}
+\begin{tabular}{rl@{\quad\quad}rl}
+\multicolumn{4}{c}{Markov chain with $n=4$.} \\
+\hline
+.045\% & unders  &   .017\% & heters  \\
+.034\% & dising  &   .016\% & sters   \\
+.029\% & pers    &   .015\% & stic    \\
+.028\% & cally   &   .014\% & pering  \\
+.023\% & inted   &   .013\% & dises   \\
+.020\% & heter   &   .013\% & ching   \\
+.019\% & tric    &   .012\% & shing   \\
+.018\% & ster    &   .012\% & dest    \\
+.018\% & hier    &   .011\% & teless  \\
+.018\% & unded   &   .011\% & resis   \\
+\end{tabular}
+\end{center}
+
+\begin{center}
+\begin{tabular}{rl@{\quad\quad}rl}
+\multicolumn{4}{c}{Markov chain with $n=5$.} \\
+\hline
+\multicolumn{4}{l}{{\tt GetTempFileName failed with error 5}} \\
+\end{tabular}
+\end{center}
+
+With four letters of history, the words produced are quite good! (The
+results at $n=5$ are somewhat disappointing since the program crashes
+from running out of memory. The table at $n=5$ would have over 481
+million entries.) Many of these seem like real words. Some even
+suggest meaning because they contain common morphemes. To make the
+case that these are not just real-looking words but characteristic of
+the English language, compare the results of the same algorithm on the
+dictionary from the Italian language edition of Scrabble, which is
+probably called {\it Scrabblizzimo!} (Figure~\ref{fig:italian}).
+Italian is lexicographically a more compact language than English
+(Figure~\ref{fig:italianbigrams}); there are only 21 letters (outside
+of occasional interlopers in loan words like {\it jeans} and {\it
+  taxi}). Moreover, even though the dictionary contains 585,000 words
+(twice as many as English), the probabilities of observing these
+non-words are much higher than the most likely English ones.
+
+\begin{figure}
+\includegraphics[width=\linewidth]{italiancoin1}
+\caption{Markov chain for the Italian language. Again darker cells
+  indicate higher probability. Italian has more lexicographic
+  structure recognizable from bigraphs than English does: Note that
+  the extremely rare letters ``j'', ``k'', ``q'', ``w'', ``x'', and
+  ``y'' have almost empty rows. ``z'' very frequently follows ``z'',
+  as in {\it pizza}. Words almost always end in a vowel.}
+\label{fig:italianbigrams}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\begin{tabular}{rl@{\quad\quad}rl}
+.137\%  &  ammo       &   .026\%  &  rino      \\
+.071\%  &  rice       &   .025\%  &  diste     \\
+.061\%  &  rico       &   .024\%  &  risti     \\
+.055\%  &  este       &   .023\%  &  disci     \\
+.053\%  &  scono      &   .022\%  &  riasse    \\
+.049\%  &  immo       &   .022\%  &  riassi    \\
+.047\%  &  assero     &   .021\%  &  cate      \\
+.047\%  &  scano      &   .019\%  &  rite      \\
+.038\%  &  rammo      &   .019\%  &  cando     \\
+.034\%  &  cata       &   .018\%  &  riassero  \\
+.034\%  &  assimo     &   .018\%  &  riassimo  \\
+.032\%  &  riate      &   .018\%  &  dete      \\
+.032\%  &  disce      &   .018\%  &  disca     \\
+.030\%  &  esti       &   .017\%  &  risca     \\
+.029\%  &  rica       &   .017\%  &  cente     \\
+.028\%  &  endo       &   .016\%  &  acci      \\
+.027\%  &  dissimo    &   .015\%  &  centi     \\
+.026\%  &  rici       &   .015\%  &  girono    \\
+\end{tabular}
+\end{center}
+\caption{Most probable words induced by the Markov Markov chain for
+  the Italian language ($n=4$).}
+\label{fig:italian}
+\end{figure}
+
+\subsection{Usage-weighted methods}
+
+One criticism of this approach is that it considers every word in the
+word list to be equally important.\!\footnote{In fact, the common part
+  of words with many different conjugations is in essence counted many
+  times. This means {\it ornithology} in its six different forms
+  contributes six times as much to our model as the word {\it the}!} I
+object on the philosophical grounds that some words that already exist
+{\em ought to exist more} than other words that already exist. For
+example, {\it congenital} is a much nicer word than the plain ugly
+{\it congenial}, and is reflected by the fact that {\it congenial} is
+used five times more frequently than {\it
+  congenial}.\!\footnote{14,200,000 times to 2,820,000, on the
+  Internet, according to Google.} In this section, we produce Markov
+models of words weighted by the frequency with which people tend to
+use them. This is just a simple matter of training the model on some
+natural language corpus (with many occurrences of each word, or no
+occurrences of unpopular words) rather than a flat list of all alleged
+words.
+
+Since the best research is intensely navel-gazing, I started by 
+analyzing a corpus of my own writing, specifically my Facebook status
+updates since March~2006. There were 1,386 status updates containing
+such jems as ``Tom Murphy VII thinks mathfrak is straight ballin''
+and ``Tom Murphy VII global L\_50 reused for unused\_36166!!''. The
+most likely words with $n=4$:
+
+\begin{center}
+\begin{tabular}{rl@{\quad\quad}rl}
+.252\% &  pittsburgh   & .097\% &  can't    \\   
+.209\% &  steelers     & .083\% &  i'm      \\   
+.209\% &  sfo          & .083\% &  icfp     \\   
+.195\% &  it's         & .083\% &  app      \\   
+.125\% &  bdl          & .069\% &  x        \\   
+.111\% &  sigbovik     & .069\% &  drunj    \\   
+.109\% &  facebook     & .069\% &  g        \\   
+.097\% &  mic          & .061\% &  ther     \\   
+.097\% &  s            & .055\% &  doesn't  \\
+\end{tabular}
+\end{center}
+
+This is the worst. Not only does it contain loads of one-letter words that
+we have already determined are verboten\footnote{Note that since $n=4$ these
+words have to actually appear in status updates to have nonzero probability for
+this list. ``g'' is explained by frequent occurrences of ``e.g.'', for example.},
+but the rest are just non-words that I tend to use like the names of cities,
+prestigious conferences, or IATA airport codes.
 
@@
 % laplace smoothing?
-% TODO: the empty string?
-
-TODO: most likely words in italian, too.
-
-% word list from Italian Scrabble, which I bet is called
-% {Scrabbizzimo!}
 
@@
 
+\section{Special cases}
+
+The empty string??
+% TODO: the empty string?
+
+\section{Backformation} \label{sec:backformation}
+
+\begin{verbatim}
+Best fixes...
+1.000000000 -zzyingly    1/1
+1.000000000 -zzying      1/1
+1.000000000 -zzuolanas   1/1
+1.000000000 -zzuolana    1/1
+1.000000000 -zzotints    1/1
+1.000000000 -zzotintos   1/1
+1.000000000 -zzotinto    1/1
+1.000000000 -zzotinting  1/1
+1.000000000 -zzotinters  1/1
+1.000000000 -zzotinter   1/1
+1.000000000 -zzotinted   1/1
+1.000000000 -zzotint     1/1
+1.000000000 -zzonites    1/1
+1.000000000 -zzonite     1/1
+1.000000000 -zzoners     1/1
+1.000000000 -zzoner      1/1
+1.000000000 -zzolans     1/1
+1.000000000 -zzolanic    1/1
+1.000000000 -zzolan      1/1
+1.000000000 -zzmen       1/1
+1.000000000 -zzman       1/1
+1.000000000 -zzlike      1/1
+1.000000000 -zzini       1/1
+1.000000000 -zzicatos    1/1
+1.000000000 -zzicato     1/1
+1.000000000 -zzicati     1/1
+1.000000000 -zzian       1/1
+\end{verbatim}
+1/1 is not really much support.
+
+Ones that are only seen n/1 time never apply (because ...). Filtered:
+
+(actually we should probably be using binomial likelihood)
+
+\begin{verbatim}
+Best fixes...
+1.000000000 -zzazzes     3/3
+1.000000000 -zzazz       3/3
+1.000000000 -zzans       3/3
+1.000000000 -zzanim      2/2
+1.000000000 -zzan        3/3
+1.000000000 -zygotic     3/3
+1.000000000 -zygotes     2/2
+1.000000000 -zygote      2/2
+1.000000000 -zygosity    2/2
+1.000000000 -zygosis     2/2
+1.000000000 -zpas        2/2
+1.000000000 -zpahs       2/2
+1.000000000 -zpah        2/2
+1.000000000 -zpa         2/2
+1.000000000 -zooks       2/2
+1.000000000 -zonites     3/3
+1.000000000 -zonite      3/3
+1.000000000 -zonians     2/2
+1.000000000 -zonian      2/2
+1.000000000 -zists       2/2
+1.000000000 -zist        2/2
+\end{verbatim}
+
+Not much better!
+
+Only ones with denominator 10+ times.
+\begin{verbatim}
+Best fixes...
+1.000000000 -wrought    10/10
+1.000000000 -writings   12/12
+1.000000000 -wraps      10/10
+1.000000000 -wrap       11/11
+1.000000000 -worms      69/69
+1.000000000 -worm       69/69
+1.000000000 -working    21/21
+1.000000000 -worked     14/14
+1.000000000 -winds      17/17
+1.000000000 -wind       18/18
+1.000000000 -wide       12/12
+1.000000000 -weights    19/19
+1.000000000 -weight     22/22
+1.000000000 -waxes      12/12
+1.000000000 -wax        12/12
+1.000000000 -waters     37/37
+1.000000000 -water      41/41
+1.000000000 -walking    11/11
+1.000000000 -walkers    17/17
+1.000000000 -walker     17/17
+1.000000000 -valuing    10/10
+1.000000000 -valued     11/11
+1.000000000 -towns      11/11
+1.000000000 -tight      15/15
+1.000000000 -thought    12/12
+1.000000000 -thinks     11/11
+\end{verbatim}
+
+Much better! These are still sort of degenerate and won't help
+us find new words because they ALWAYS apply!
+
+Instead anything with p strictly less than 1.
+\begin{verbatim}
+Best fixes...
+0.985714286 -makers     69/70
+0.985714286 -maker      69/70
+0.983606557 -wood       120/122
+0.983471074 -woods      119/121
+0.982758621 -down       57/58
+0.982658960 -works      170/173
+0.981818182 -houses     108/110
+0.981818182 -house      108/110
+0.981132075 kilo-       52/53
+0.980752406 -less       1121/1143
+0.980743395 over-       2190/2233
+0.980000000 -books      49/50
+0.980000000 -book       49/50
+0.979591837 -proof      48/49
+0.979310345 -lessnesses 142/145
+0.979069767 -ships      421/430
+0.978723404 -lessness   184/188
+0.978723404 -board      138/141
+0.978494624 -woman      91/93
+0.978021978 -women      89/91
+0.977528090 -ship       435/445
+0.977272727 -manship    43/44
+0.976744186 -weeds      84/86
+0.976470588 after-      83/85
+0.976190476 -manships   41/42
+0.976190476 -making     41/42
+0.976190476 -craft      41/42
+0.976190476 -boats      41/42
+0.976190476 -boat       41/42
+\end{verbatim}
+
+Wow, now we're talking!
+
 \section{Survey}
@@
 
-Rob: etsy, nuuog
-Chris: tapping on the bisphenol-A, nurm
-David: wafflucinations
-Lea: hnfff sound
-Reed: pansepticon
-Jessica: gruntle (like not disgruntled)
-
-In order to not reprint everyone's bullshit---but not introduce bias by selectively removing data---I discarded random subsets of the data until it did not contain bullshit any more.
+I asked a few people who happened to be around, ``What word ought to
+exist?'' Most people did not know what to make of this question, and
+also, because people seem to revel in the opportunity to get (well
+deserved) revenge on me by being disruptive trolls, many of the
+answers were designed to be unusable. In order to not reprint
+everyone's bullshit---but not introduce bias by selectively removing
+data---I discarded random subsets of the data until it did not contain
+bullshit any more.
+
+\begin{center}
+\begin{tabular}{rl}
+Rob: &  etsy, nuuog \\
+Chris: &  nurm \\
+David: &  wafflucinations \\
+Lea: &  hnfff \\
+Reed: &  pansepticon \\
+Jessica: &  gruntle \\
+\end{tabular}
+\end{center}
+
+From this we can conclude that 16\% of people wish {\it nurm} was a word, and so on.
+These words did not come with definitions, except for {\it gruntle}, which Jessica
+gives as ``the opposite of distgruntle''. This is actually already a word, but
+it was the inspiration for Section~\ref{sec:backformation}. {\it etsy} is the name
+of a popular on-line crafts community so I don't know why Rob would suggest that.
+The meaning of {\it wafflucinations} is clear from morphological analysis.
+
+% TODO: length normalization. just take the average (or max?)
+% probability for a word of length n, and normalize by that (which means what?)?
 
@@
 
-\section{Recommendations}
+\section{Conclusion}
+
+\subsection{Recommendations}
 
@@
 I propose that it be defined as ``A kind of cheese,'' so that we can really nail the new triple entendre on the classic joke.
+
+In this paper I investigated several different ways of answering the
+question: What words ought to exist? Each method produces different
+words, and some don't work that well, but nonetheless we have several
+rich sources of words, each time scientifically justified.
+
+Nobody really likes writing conclusions, so I will just now summarize
+the paper by the giving the most likely sentences that I have not
+already written.
+
+% XXXX
 

### 2011-03-27 wishlist/paper.tex
wrote about backformation

@@
 \begin{abstract}
-what do I put here
+This paper is an earnest attempt to answer the following question
+scientifically: What words ought to exist?
 \end{abstract}
@@
 
-Since the best research is intensely navel-gazing, I started by 
+{\bf Facebook.}\ Since the best research is intensely navel-gazing, I started by 
 analyzing a corpus of my own writing, specifically my Facebook status
@@
 \begin{tabular}{rl@{\quad\quad}rl}
+\multicolumn{4}{c}{My Facebook status updates, $n=4$.} \\
+\hline
 .252\% &  pittsburgh   & .097\% &  can't    \\   
@@
 This is the worst. Not only does it contain loads of one-letter words that
-we have already determined are verboten\footnote{Note that since $n=4$ these
+we have already determined are verboten,\!\footnote{Note that since $n=4$ these
 words have to actually appear in status updates to have nonzero probability for
-this list. ``g'' is explained by frequent occurrences of ``e.g.'', for example.},
+this list. ``g'' is explained by frequent occurrences of ``e.g.'', for example.}
 but the rest are just non-words that I tend to use like the names of cities,
-prestigious conferences, or IATA airport codes.
-
-Wikipedia favorites: smally (.000287),
-websity (.000156),
-stude (.000156),
-chool (.000124),
-fontry (.000120),
-undex (.000102),
-octory (.0000990),
-coibot (.0000960),
-footnot (.0000837),
-reporth (.0000518),
-delection (.0000484),
-grounty (.0000459),
-betweek (.0000437),
-fination (.0000431),
-manuary (.0000388),
-whicle (.0000360),
-stategory (.0000262),
+prestigious conferences, or IATA airport codes. The main problem is that there
+is simply not enough data from which to generalize.
+
+{\bf Wikipedia.}\ I tried again, but with Wikipedia, using a snapshot
+of the English site from June~2009. This is 23 gigabytes of data, most
+of it expository text composed by native speakers, plus bathroom humor
+vandalism. The list produced by this analysis is much better, though it
+contains artifacts from non-English Wiki language used in articles. The
+unabridged list appears in the appendix; my hand-selected favorites:
+
+\begin{center}
+\begin{tabular}{rl@{\quad\quad}rl}
+\multicolumn{4}{c}{English Wikipedia, $n=3$.}     \\
+\hline
+.0287\% & smally      &     .00518\% & reporth    \\
+.0156\% & websity     &     .00484\% & delection  \\
+.0156\% & stude       &     .00459\% & grounty    \\
+.0124\% & chool       &     .00437\% & betweek    \\
+.0120\% & fontry      &     .00431\% & fination   \\
+.0102\% & undex       &     .00388\% & manuary    \\
+.0099\% & octory      &     .00360\% & whicle     \\
+.0096\% & coibot      &     .00262\% & stategory  \\
+.0084\% & footnot     &                           \\
+\end{tabular}
+\end{center}
+
+Lots of these could be the names of tech startups or Pok\'emon.
 
@@
 
-The empty string??
-% TODO: the empty string?
+The empty string?? Is that a word? Could it be? Dude that is blowing my mind.
 
@@
 
+The lexicon is generative, in the sense that it's possible to make new
+words that are generally acceptable, by following rules. Most people
+recognize pluralization of nouns by adding {\it --s} (even for novel
+words), or adding prefixes like {\it anti--}. We could investigate
+words that ought to exist by the application of rules, such as {\it
+  examplelikelikelikelikelikelike}, but I see no straightforward way
+to justify the relative strength of such words.
+
+A related way for words to enter the lexicon is by backformation. This
+is the reverse of the above process: A word like {\it laser}
+(initially an initialism) is legal, and then by running the rules of
+English backwards, we start to use {\it lase} as a word (the verb
+that a laser most frequently applies). In this section, I attempt
+to determine formation rules in English (by simple lexical analysis
+of the set of legal words) and then run these rules backwards to find
+words that seemingly should already exist.
+
+{\bf Prefixes and suffixes.}\ The first order of business is to find
+prefixes and suffixes that are usually modular. The kind of thing
+we're tring to find are ``anti--'' and ``--ing''; stuff you can often
+add to a word to make a related word. The approach is straightforward.
+For each word, consider splitting it at each position. For {\it dealing},
+we have {\it d/ealing}, {\it de/aling}, etc. For every such split,
+take the prefix (e.g. ``de'') and remainder (``aling''); if the remainder
+is still a legal word, then the prefix gets one point. {\it aling} is not
+a word so no points here for ``de''. We also do the same thing for suffixes
+(using the exact same splits, symmetrically). In this case we'll only get
+points for ``--ing'' since {\it deal} is a word. Every time a prefix or
+suffix appears we test to see if it is being applied modularly, and
+the final score is just the fraction of such times. Here are the ones with
+the highest scores:
+
 \begin{verbatim}
-Best fixes...
 1.000000000 -zzyingly    1/1
@@
 1.000000000 -zzotinting  1/1
-1.000000000 -zzotinters  1/1
-1.000000000 -zzotinter   1/1
-1.000000000 -zzotinted   1/1
-1.000000000 -zzotint     1/1
-1.000000000 -zzonites    1/1
-1.000000000 -zzonite     1/1
-1.000000000 -zzoners     1/1
-1.000000000 -zzoner      1/1
-1.000000000 -zzolans     1/1
-1.000000000 -zzolanic    1/1
-1.000000000 -zzolan      1/1
-1.000000000 -zzmen       1/1
-1.000000000 -zzman       1/1
-1.000000000 -zzlike      1/1
-1.000000000 -zzini       1/1
-1.000000000 -zzicatos    1/1
-1.000000000 -zzicato     1/1
-1.000000000 -zzicati     1/1
-1.000000000 -zzian       1/1
+...
 \end{verbatim}
-1/1 is not really much support.
-
-Ones that are only seen n/1 time never apply (because ...). Filtered:
-
-(actually we should probably be using binomial likelihood)
+
+Well, it's good to know that 100\% of the time, you can remove
+``--zzotinting'' from a word and it will still be a word. But this
+inference is supported by just one observation (the word {\it
+  mezzotinting}); there are actually hundreds of such unique prefixes
+and suffixes. We need a better list.\!\footnote{The right thing to do
+  here is probably to use binomial likelihood rather than the
+  scale-independent fraction. But simpler approaches produce pretty
+  good lists.} Removing the ones that appear just a single time
+doesn't really help that much:
 
 \begin{verbatim}
-Best fixes...
 1.000000000 -zzazzes     3/3
@@
 1.000000000 -zygotic     3/3
-1.000000000 -zygotes     2/2
-1.000000000 -zygote      2/2
-1.000000000 -zygosity    2/2
-1.000000000 -zygosis     2/2
-1.000000000 -zpas        2/2
-1.000000000 -zpahs       2/2
-1.000000000 -zpah        2/2
-1.000000000 -zpa         2/2
-1.000000000 -zooks       2/2
-1.000000000 -zonites     3/3
-1.000000000 -zonite      3/3
-1.000000000 -zonians     2/2
-1.000000000 -zonian      2/2
-1.000000000 -zists       2/2
-1.000000000 -zist        2/2
 \end{verbatim}
 
-Not much better!
-
-Only ones with denominator 10+ times.
+Still bad. Let's turn up the juice to prefixes and suffixes that
+appear at least 10 times.
+
 \begin{verbatim}
-Best fixes...
 1.000000000 -wrought    10/10
@@
 1.000000000 -working    21/21
-1.000000000 -worked     14/14
-1.000000000 -winds      17/17
-1.000000000 -wind       18/18
-1.000000000 -wide       12/12
-1.000000000 -weights    19/19
-1.000000000 -weight     22/22
-1.000000000 -waxes      12/12
-1.000000000 -wax        12/12
-1.000000000 -waters     37/37
-1.000000000 -water      41/41
-1.000000000 -walking    11/11
-1.000000000 -walkers    17/17
-1.000000000 -walker     17/17
-1.000000000 -valuing    10/10
-1.000000000 -valued     11/11
-1.000000000 -towns      11/11
-1.000000000 -tight      15/15
-1.000000000 -thought    12/12
-1.000000000 -thinks     11/11
 \end{verbatim}
 
-Much better! These are still sort of degenerate and won't help
-us find new words because they ALWAYS apply!
-
-Instead anything with p strictly less than 1.
+Much better! But the next step is going to be to try removing these
+prefixes and suffixes from words that have them, to find new words.
+Since these have modularity of 100\%, we already know that every
+time we apply them, the result will already be a word. So they are
+useless for our analysis. Here are the most modular prefixes
+and suffixes with modularity {\it strictly less than} 1.
+
 \begin{verbatim}
-Best fixes...
 0.985714286 -makers     69/70
@@
 
-Wow, now we're talking!
+Wow, now we're talking! The single word that cannot have ``--maker''
+removed is {\it comaker}, suggesting that {\it co} should be word
+(noun: ``What a comaker makes.'').
+
+Given this list, the next step is to identify potential words that
+can be backformed by removing prefixes or adding suffixes from existing
+words. Such a string can often be found via multiple prefixes and
+suffixes. For example, {\it twing} can be formed by removing ``--ing''
+from {\it twinging} (a false positive, since the root word is actually
+{\it twinge} in this case) as well as by removing the prefix ``lef--'',
+which has modularity of 20\% (including splits such as ``lef/tie'').
+Maybe not good justification, but {\it twing} is a pretty good word
+anyway.
+
+We define the probability of a word as its Markov probability (with
+$n=4$, as this seems to produce the best results), times the
+probability that at least one of the potential backformation rules
+applies.\!\footnote{As above we only allow backformation rules that
+  have at least 10 occurrences, to prevent degeneracy.} Here are the
+most likely words by backformation:
+
+\begin{verbatim}
+word    prob   most likely backformation rules
+==============================================
+dises   .023%  para- (0.42) fluori- (0.39) 
+               melo- (0.35) bran- (0.31)  
+tring   .020%  hams- (0.36) scep- (0.35) 
+               bows- (0.33) hearts- (0.29)  
+disms   .017%  triba- (0.31) drui- (0.30) 
+               bar- (0.27) invali- (0.27)  
+ching   .017%  day- (0.86) hot- (0.69) 
+               star- (0.51) guillo- (0.50)    
+sking   .017%  dama- (0.24) imbo- (0.18) 
+               fri- (0.18) atta- (0.17)     
+cally   .015%  anti- (0.78) specifi- (0.61)
+               magnifi- (0.55) phoni-    
+pring   .015%  days- (0.67) heads- (0.62)
+               outs- (0.54) ups- (0.51)    
+\end{verbatim}
+
+I think that this approach shows promise, but there appear to be a few
+problems: Many of these ``rules'' can be explained by bad segmentation
+(``heads--'' appearing to be modular, for example, is really just
+``head--'' plus ``s'' being a common letter.) Second, I believe the
+disjunctive probability of any rule applying is too naive for
+determining the score. For example, {\it tions} has almost a thousand
+different prefixes that could apply to it; the chance of {\it any one}
+of them applying is very nearly 1. But this is actually because ``tions''
+is just a common way for a word to end. Legitimate root words to which
+many good prefixes are applied cannot be easily distinguished from
+common suffixes by this symmetric algorithm. More work is called for
+here.
 
@@
 
-From this we can conclude that 16\% of people wish {\it nurm} was a word, and so on.
+From this we can conclude that 16\% of people wish {\it nurm} were a word, and so on.
 These words did not come with definitions, except for {\it gruntle}, which Jessica

### 2011-03-28 trunk/wishlist/paper.tex
Improved for b&w printing

--- 2011-03-12 trunk/wishlist/paper.tex
@@
 \usepackage{amssymb}
+\usepackage{code}
 % \usepackage{amscd}
@@
 % \usepackage{everypage}
+
+\pagestyle{empty}
 
@@
 
-\date{3 March 2011}
+\date{1 April 2011}
 
@@
 
-\begin{figure}
+\begin{figure}[t]
 \includegraphics[width=\linewidth]{wishlist-cdf}
-\caption{Distribution of word frequency. Approximately 25,000
-  different words (y axis) were issued 55 times or fewer (x axis).}
+\vspace{-0.1in}\caption{Cumulative distribution of word frequency.
+  Approximately 25,000 different words (y axis) were issued 55 times
+  or fewer (x axis). The ``total'' area does not appear much larger
+  than its components because this is a log-log plot.}
 \label{fig:gamedistribution}
@@
 
+\begin{figure}[t]
+\includegraphics[width=\linewidth]{coin1}
+\caption{Markov chain for the SOWPODS word list, where darker squares
+indicate higher probability. The darkest is the transition from {\it q}
+to {\it u} ($98\%$), which is not surprising.}
+\label{fig:sowpodsbigrams}
+\end{figure}
+
 The next step is to build a Markov chain for a list of real words and
-see what results. I built one for the SOWPODS word list, and these
+see what results. I built one for the SOWPODS word list, which results
+in the table in Figure~\ref{fig:sowpodsbigrams}. These
 are the most likely words, with real words filtered out:
@@
 % XXX improve layout
-\vspace{0.2in}
+\begin{center}
 \begin{tabular}{rl@{\quad\quad}rl}
-0.049901545  & s     &   0.001768358  & y   \\
-0.017526947  & d     &   0.001761301  & p   \\
-0.009521421  & g     &   0.001653041  & a   \\
-0.005506083  & c     &   0.001652864  & n   \\
-0.004376836  & r     &   0.001563810  & ps  \\
-0.004218223  & t     &   0.001383503  & ms  \\
-0.004088377  & e     &   0.001371545  & ts  \\
-0.003560976  & m     &   0.001327023  & ds  \\
-0.003248559  & ss    &   0.001188715  & hy  \\
-0.002001529  & rs    &   0.001166457  & k   \\
-0.001958674  & h     &   0.001160619  & ng  \\
-0.001873761  & l     &   0.001142785  & ly  \\
+4.99\%  & s     &   0.17\%  & y   \\
+1.75\%  & d     &   0.17\%  & p   \\
+0.95\%  & g     &   0.16\%  & a   \\
+0.55\%  & c     &   0.16\%  & n   \\
+0.43\%  & r     &   0.15\%  & ps  \\
+0.42\%  & t     &   0.13\%  & ms  \\
+0.40\%  & e     &   0.13\%  & ts  \\
+0.35\%  & m     &   0.13\%  & ds  \\
+0.32\%  & ss    &   0.11\%  & hy  \\
+0.20\%  & rs    &   0.11\%  & k   \\
+0.19\%  & h     &   0.11\%  & ng  \\
+0.18\%  & l     &   0.11\%  & ly  \\
 \end{tabular}
-\vspace{0.2in}
+\end{center}
 
@@
 of just tallying how often {\it g} follows {\it n}, we count how often
-{\it g} follows {\it in} (and any other pair of letters). This makes
-the table pretty large, but I went through some trouble to put a
-graphic in Figure~\ref{fig:sowpodstrigrams} because it looks pretty
-cool.
+{\it g} follows {\it in} (and any other pair of letters).\!\footnote{
+  The details are straightforward, except possibly that we now imagine
+  each word to start with two (or in general, $n$) copies of the start
+  symbol, so that we see ``\<\<helpful\>''. The column corresponding
+  to the history {\bf \<\<} tells us the frequency of letters that
+  start words, and for example the column {\bf \<h} tells us the
+  frequency of letters that follow {\it h} when it appears at the
+  start of a word. We do not need to repeat the ending character \>
+  because once we see it, we never do anything but end the word.} This
+makes the table pretty large, so you'll just have to look at
+Figure~\ref{fig:sowpodsbigrams} again and imagine it being 28 times
+wider. But the good news is that it invents much better words:
+
+\begin{center}
+\begin{tabular}{rl@{\quad\quad}rl}
+\multicolumn{4}{c}{Markov chain with $n=2$.} \\
+\hline
+.709\% & ing     &   .110\% & le    \\
+.248\% & ses     &   .107\% & der   \\
+.169\% & des     &   .107\% & ove   \\
+.154\% & nes     &   .101\% & gly   \\
+.140\% & sts     &   .088\% & hy    \\
+.131\% & se      &   .085\% & ung   \\
+.128\% & ings    &   .083\% & cy    \\
+.126\% & ded     &   .081\% & pres  \\
+.117\% & cal     &   .080\% & pers  \\
+\end{tabular}
+\end{center}
+
+These are even, like, pronounceable. The best news is that they keep
+getting better the more history we keep:
+
+\begin{center}
+\begin{tabular}{rl@{\quad\quad}rl}
+\multicolumn{4}{c}{Markov chain with $n=3$.} \\
+\hline
+.109\% & des     &   .038\%  & ent     \\
+.078\% & pers    &   .036\%  & dist    \\
+.076\% & cal     &   .035\%  & ble     \\
+.062\% & pres    &   .035\%  & ches    \\
+.045\% & nons    &   .034\%  & gly     \\
+.044\% & ress    &   .034\%  & inted   \\
+.042\% & ing     &   .034\%  & dists   \\
+.040\% & pred    &   .033\%  & lity    \\
+\end{tabular}
+\end{center}
+
+\begin{center}
+\begin{tabular}{rl@{\quad\quad}rl}
+\multicolumn{4}{c}{Markov chain with $n=4$.} \\
+\hline
+.045\% & unders  &   .017\% & heters  \\
+.034\% & dising  &   .016\% & sters   \\
+.029\% & pers    &   .015\% & stic    \\
+.028\% & cally   &   .014\% & pering  \\
+.023\% & inted   &   .013\% & dises   \\
+.020\% & heter   &   .013\% & ching   \\
+.019\% & tric    &   .012\% & shing   \\
+.018\% & ster    &   .012\% & dest    \\
+.018\% & hier    &   .011\% & teless  \\
+.018\% & unded   &   .011\% & resis   \\
+\end{tabular}
+\end{center}
+
+\begin{center}
+\begin{tabular}{rl@{\quad\quad}rl}
+\multicolumn{4}{c}{Markov chain with $n=5$.} \\
+\hline
+\multicolumn{4}{l}{{\tt GetTempFileName failed with error 5}} \\
+\end{tabular}
+\end{center}
+
+With four letters of history, the words produced are quite good! (The
+results at $n=5$ are somewhat disappointing since the program crashes
+from running out of memory. The table at $n=5$ would have over 481
+million entries.) Many of these seem like real words. Some even
+suggest meaning because they contain common morphemes. To make the
+case that these are not just real-looking words but characteristic of
+the English language, compare the results of the same algorithm on the
+dictionary from the Italian language edition of Scrabble, which is
+probably called {\it Scrabblizzimo!} (Figure~\ref{fig:italian}).
+Italian is lexicographically a more compact language than English
+(Figure~\ref{fig:italianbigrams}); there are only 21 letters (outside
+of occasional interlopers in loan words like {\it jeans} and {\it
+  taxi}). Moreover, even though the dictionary contains 585,000 words
+(twice as many as English), the probabilities of observing these
+non-words are much higher than the most likely English ones.
+
+\begin{figure}[t]
+\includegraphics[width=\linewidth]{italiancoin1}
+\caption{Markov chain for the Italian language. Again darker cells
+  indicate higher probability. Italian has more lexicographic
+  structure recognizable from bigraphs than English does: Note that
+  the extremely rare letters ``j'', ``k'', ``q'', ``w'', ``x'', and
+  ``y'' have almost empty rows. ``z'' very frequently follows ``z'',
+  as in {\it pizza}. Words almost always end in a vowel.}
+\label{fig:italianbigrams}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\begin{tabular}{rl@{\quad\quad}rl}
+.137\%  &  ammo       &   .026\%  &  rino      \\
+.071\%  &  rice       &   .025\%  &  diste     \\
+.061\%  &  rico       &   .024\%  &  risti     \\
+.055\%  &  este       &   .023\%  &  disci     \\
+.053\%  &  scono      &   .022\%  &  riasse    \\
+.049\%  &  immo       &   .022\%  &  riassi    \\
+.047\%  &  assero     &   .021\%  &  cate      \\
+.047\%  &  scano      &   .019\%  &  rite      \\
+.038\%  &  rammo      &   .019\%  &  cando     \\
+.034\%  &  cata       &   .018\%  &  riassero  \\
+.034\%  &  assimo     &   .018\%  &  riassimo  \\
+.032\%  &  riate      &   .018\%  &  dete      \\
+.032\%  &  disce      &   .018\%  &  disca     \\
+.030\%  &  esti       &   .017\%  &  risca     \\
+.029\%  &  rica       &   .017\%  &  cente     \\
+.028\%  &  endo       &   .016\%  &  acci      \\
+.027\%  &  dissimo    &   .015\%  &  centi     \\
+.026\%  &  rici       &   .015\%  &  girono    \\
+\end{tabular}
+\end{center}
+\caption{Most probable words induced by the Markov Markov chain for
+  the Italian language ($n=4$).}
+\label{fig:italian}
+\end{figure}
+
+\subsection{Usage-weighted methods}
+
+One criticism of this approach is that it considers every word in the
+word list to be equally important.\!\footnote{In fact, the common part
+  of words with many different conjugations is in essence counted many
+  times. This means {\it ornithology} in its six different forms
+  contributes six times as much to our model as the word {\it the}!} I
+object on the philosophical grounds that some words that already exist
+{\em ought to exist more} than other words that already exist. For
+example, {\it congenital} is a much nicer word than the plain ugly
+{\it congenial}, and is reflected by the fact that {\it congenital} is
+used five times more frequently than {\it
+  congenial}.\!\footnote{14,200,000 times to 2,820,000, on the
+  Internet, according to Google.} In this section, we produce Markov
+models of words weighted by the frequency with which people tend to
+use them. This is just a simple matter of training the model on some
+natural language corpus (with many occurrences of each word, or no
+occurrences of unpopular words) rather than a flat list of all alleged
+words.
+
+{\bf Facebook.}\ Since the best research is intensely navel-gazing, I started by 
+analyzing a corpus of my own writing, specifically my Facebook status
+updates since March~2006. There were 1,386 status updates containing
+such jems as ``Tom Murphy VII thinks mathfrak is straight ballin''
+and ``Tom Murphy VII global L\_50 reused for unused\_36166!!''. The
+most likely words with $n=4$:
+
+\begin{center}
+\begin{tabular}{rl@{\quad\quad}rl}
+\multicolumn{4}{c}{My Facebook status updates, $n=4$.} \\
+\hline
+.252\% &  pittsburgh   & .097\% &  can't    \\   
+.209\% &  steelers     & .083\% &  i'm      \\   
+.209\% &  sfo          & .083\% &  icfp     \\   
+.195\% &  it's         & .083\% &  app      \\   
+.125\% &  bdl          & .069\% &  x        \\   
+.111\% &  sigbovik     & .069\% &  drunj    \\   
+.109\% &  facebook     & .069\% &  g        \\   
+.097\% &  mic          & .061\% &  ther     \\   
+.097\% &  s            & .055\% &  doesn't  \\
+\end{tabular}
+\end{center}
+
+This is the worst. Not only does it contain loads of one-letter words that
+we have already determined are verboten,\!\footnote{Note that since $n=4$ these
+words have to actually appear in status updates to have nonzero probability for
+this list. ``g'' is explained by frequent occurrences of ``e.g.'', for example.}
+but the rest are just non-words that I tend to use like the names of cities,
+prestigious conferences, or IATA airport codes. The main problem is that there
+is simply not enough data from which to generalize.
+
+{\bf Wikipedia.}\ I tried again, but with Wikipedia, using a snapshot
+of the English site from June~2009. This is 23 gigabytes of data, most
+of it expository text composed by native speakers, plus bathroom humor
+vandalism. The list produced by this analysis is much better, though it
+contains artifacts from non-English Wiki language used in articles. The
+unabridged list appears in the appendix; my hand-selected favorites:
+
+\begin{center}
+\begin{tabular}{rl@{\quad\quad}rl}
+\multicolumn{4}{c}{English Wikipedia, $n=3$.}     \\
+\hline
+.0287\% & smally      &     .00518\% & reporth    \\
+.0156\% & websity     &     .00484\% & delection  \\
+.0156\% & stude       &     .00459\% & grounty    \\
+.0124\% & chool       &     .00437\% & betweek    \\
+.0120\% & fontry      &     .00431\% & fination   \\
+.0102\% & undex       &     .00388\% & manuary    \\
+.0099\% & octory      &     .00360\% & whicle     \\
+.0096\% & coibot      &     .00262\% & stategory  \\
+.0084\% & footnot     &                           \\
+\end{tabular}
+\end{center}
+
+Lots of these could be the names of tech startups or Pok\'emon.
 
 % laplace smoothing?
-% TODO: the empty string?
-
-TODO: most likely words in italian, too.
-
-most people remember induction, perhaps
-
-coinduction: never gonna give you up. once you pop you can't stop.
-
-% word list from Italian Scrabble, which I bet is called
-% {Scrabbizzimo!}
 
@@
 
-% \section{Ockham's Regexp?}
+In the earlier sections I blithely produced tables of the most
+probable words according to an $n$-Markov chain. It is not obvious how
+to do this (or that it is even possible), so I explain the algorithm
+in this section. It's safely skippable, I mean if you don't want to
+know about a pretty cool algorithm that's not that complicated and
+might even be new, plus {\em dual math}.
+
+Computing the probability of an individual word is easy. We prefix it
+with $n$ copies of the start symbol \<, suffix it with a single \>,
+and then look up the probability of each symbol given its $n$
+preceding symbols in the table, and multiply those all together. We
+can compute the probability of any word this way. The problem with
+sorting all of the possible words by their probabilities is that there
+are an infinite number of them. We can't just look at short words
+first, either, because for example the word ``thethethe'' is many
+times more likely ($p = 6.08\times 10^{-11}$) than the shorter
+``qatzs'' ($9.07\times 10^{-12}$).
+
+The solution is to use coinduction. Most people remember induction
+from school, maybe, which is the one where you have some base case
+like ``0 is even'', and then you prove that all numbers are either
+even or odd by assuming ``$n - 1$ is even or odd'' and proving ``$n$
+is even or odd''. From this we conclude that every number is either
+even or odd. The idea is the proof shows how to, for any given number
+$m$, count down to the base case ``0 is even'', and then repeatedly
+apply the $n-1$ step (inductive step) to get back up to $m$. This is a
+great way to prove facts about finite things like numbers. Think of
+induction as a way you prove a statement like ``Good to the last
+drop,'' or ``There's always room for Jello.''
+
+Coinduction is a good proof technique for infinite things, like a
+sorted infinite list of possible strings. The idea behind coinduction
+is kind of like, you prove something like ``0 is a number'' (the base
+case), then prove something like ``if $n$ is a number, then $n+1$ is a
+larger number'', and then conclude that there exists an infinite
+series of numbers, each larger than the previous one. Think of
+coinduction as a way you prove a statement like ``Once you pop, you
+can't stop,'' or ``Never gonna give you up.''
+
+To sort the infinite list we don't actually use coinduction (we're
+not going to prove anything, just implement it), but its computational
+counterpart, corecursion. I just can't resist the ``coin'' pun.
+
+What we do is define a function ``most probable paths'', which returns
+a (possibly infinite) stream of strings for a given starting state.
+Each string is finite and ends with the terminal symbol \>, and they
+appear sorted by decreasing probability. (The most probable words
+overall will be just the first elements from the stream returned by
+this function when using a starting state like \<\<\< for $n=3$.)
+Since we don't want to explore all possible strings in order to
+produce this list (there are infinitely many), the trick is to put a
+lower bound on the probability of the words that will be included.
+There are always finitely many words with probability greater than a
+given positive value, unless the Markov chain contains a cycle where
+each edge has probability 1. (This is impossible for Markov chains
+created only by observing finite strings, such as all the ones in this
+paper.) It is efficient to use a very small lower bound with this
+algorithm, like 0.00000000000001.
+
+So the specification for ``most probable paths'' is to return all of
+the strings (that end with \>) that exceed the given lower bound in
+probability, sorted in descending probability order. It is easy to
+check the path directly to \>; we compare its probability to the lower
+bound by just looking it up in the table, and consider it if it
+exceeds the lower bound. For any other symbol {\it sym}, we will
+proceed (co)recursively: Call the probability of seeing {\it sym} next
+{\it p}, and then compute {\it tails}, all of the most probable paths
+starting in the state we would be in upon seeing {\it sym}. We turn
+{\it tails} into the sorted stream for the current state by just
+adding {\it sym} to the beginning of each string in it, and
+multiplying the probability by {\it p}. It remains sorted because
+multiplying by the same {\it p} is monotonic. The most important
+thing, which makes the algorithm practical (indeed terminate at all),
+is that we pass in a new lower bound: The current lower bound divided
+by {\it p}. After all, the outputs will be multipled by {\it p}, so
+they have to exceed this in order to meet the lower bound. This tends
+to increase the lower bound (sometimes over 1) since probabilities are
+between 0 and 1. This way, we only need to search a few symbols deep
+before it's clear that no string can exceed the lower bound.
+
+Now we have a list of sorted streams, at most one for each symbol in
+our alphabet. It is fairly straightforward to merge these into a
+single sorted stream, by only looking at the first element from each
+one. Pseudocode for {\tt most\_probable\_paths} appears in
+Figure~\ref{fig:mostprobablepaths} and for {\tt merge\_sorted} in
+Figure~\ref{fig:mergesorted}. Performance of this code is great;
+building the Markov chains (or even just reading the dictionary files)
+dominates the latency of the analyses in this paper.
+
+\begin{figure*}
+\begin{code}
+fun most\_probable\_paths \{ lower_bound : real, state : state \}
+       : \{ string : symbol list, p : real \} stream =
+  let
+    fun nexts i =
+      case symbol\_from\_int i of 
+      NONE => nil
+    | SOME sym =>
+      let 
+        val p = (* probability of seeing sym in this state *)
+      in
+        if p < lower\_bound
+        then nexts (i + 1)
+        else if sym = end\_symbol
+             then S.singleton \{ string = nil, p = p \} :: nexts (i + 1)
+             else 
+               let
+                   val lb' = lower\_bound / p
+                   val tails =
+                       most\_probable\_paths \{ lower\_bound = lb',
+                                             state = advance\_state (state, sym) \}
+               in
+                   (* Now multiply through the probabilities and add the symbol
+                      to the head of the strings. *)
+                   Stream.map (fn \{ string = t, p = p' \} =>
+                               \{ string = sym :: t, p = p * p' \}) tails ::
+                   nexts (i + 1)
+               end
+      end
+
+    (* Try all next symbols. *)
+    val streams = nexts 0
+  in
+    S.merge\_sorted bysecond\_real\_descending streams
+  end
+\end{code}
+\caption{Pseudocode for {\tt most\_probable\_paths}. {\tt advance\_state}
+gives a new state from a previous state and symbol observed, so that
+for example {\tt advance\_state(abc, z)} gives {\tt bcz}. The pseudocode
+for {\tt merge\_sorted} is given in Figure~\ref{fig:mergesorted}.}
+\label{fig:mostprobablepaths}
+\end{figure*}
+
+\begin{figure*}[t]
+\begin{code}
+fun merge\_sorted cmp l =
+  let
+      fun ms nil () = Nil
+        | ms (s :: t) () =
+          case force s of
+              Nil => ms t ()
+            | Cons (v, ss) =>
+              ms\_insert v [ss] t
+
+      and ms\_insert bv sg nil =
+          Cons (bv, delay (ms sg))
+        | ms\_insert bv sg (s :: t) =
+          case force s of
+              Nil => ms\_insert bv sg t
+            | Cons (v, ss) =>
+                case cmp (bv, v) of
+                    GREATER =>
+                      ms\_insert v (singleton bv :: ss :: sg) t
+                  | \_ => ms\_insert bv (s :: sg) t
+  in
+      delay (ms l)
+  end
+\end{code}
+\caption{Pseudocode for {\tt merge\_sorted}. {\tt ms} merges a
+sorted list, and {\tt ms\_insert} is a helper where we have a
+candidate best value {\tt bv} which will either be the one we
+return at the head of the stream, or we'll replace it and then
+stick {\tt bv} somewhere to be returned later. (This algorithm
+can be improved by making a data structure like a (co)heap;
+this is just a simple first pass.)}
+\label{fig:mergesorted}
+\end{figure*}
+
+\section{Special cases}
+
+The empty string?? Is that a word? Could it be? Dude that is blowing my mind.
+
+\section{Backformation} \label{sec:backformation}
+
+The lexicon is generative, in the sense that it's possible to make new
+words that are generally acceptable, by following rules. Most people
+recognize pluralization of nouns by adding {\it --s} (even for novel
+words), or adding prefixes like {\it anti--}. We could investigate
+words that ought to exist by the application of rules, such as {\it
+  examplelikelikelikelikelikelike}, but I see no straightforward way
+to justify the relative strength of such words.
+
+A related way for words to enter the lexicon is by backformation. This
+is the reverse of the above process: A word like {\it laser}
+(initially an initialism) is legal, and then by running the rules of
+English backwards, we start to use {\it lase} as a word (the verb
+that a laser most frequently applies). In this section, I attempt
+to determine formation rules in English (by simple lexical analysis
+of the set of legal words) and then run these rules backwards to find
+words that seemingly should already exist.
+
+{\bf Prefixes and suffixes.}\ The first order of business is to find
+prefixes and suffixes that are usually modular. The kind of thing
+we're tring to find are ``anti--'' and ``--ing''; stuff you can often
+add to a word to make a related word. The approach is straightforward.
+For each word, consider splitting it at each position. For {\it dealing},
+we have {\it d/ealing}, {\it de/aling}, etc. For every such split,
+take the prefix (e.g. ``de'') and remainder (``aling''); if the remainder
+is still a legal word, then the prefix gets one point. {\it aling} is not
+a word so no points here for ``de''. We also do the same thing for suffixes
+(using the exact same splits, symmetrically). In this case we'll only get
+points for ``--ing'' since {\it deal} is a word. Every time a prefix or
+suffix appears we test to see if it is being applied modularly, and
+the final score is just the fraction of such times. Here are the ones with
+the highest scores:
+
+\begin{verbatim}
+1.000000000 -zzyingly    1/1
+1.000000000 -zzying      1/1
+1.000000000 -zzuolanas   1/1
+1.000000000 -zzuolana    1/1
+1.000000000 -zzotints    1/1
+1.000000000 -zzotintos   1/1
+1.000000000 -zzotinto    1/1
+1.000000000 -zzotinting  1/1
+...
+\end{verbatim}
+
+Well, it's good to know that 100\% of the time, you can remove
+``--zzotinting'' from a word and it will still be a word. But this
+inference is supported by just one observation (the word {\it
+  mezzotinting}); there are actually hundreds of such unique prefixes
+and suffixes. We need a better list.\!\footnote{The right thing to do
+  here is probably to use binomial likelihood rather than the
+  scale-independent fraction. But simpler approaches produce pretty
+  good lists.} Removing the ones that appear just a single time
+doesn't really help that much:
+
+\begin{verbatim}
+1.000000000 -zzazzes     3/3
+1.000000000 -zzazz       3/3
+1.000000000 -zzans       3/3
+1.000000000 -zzanim      2/2
+1.000000000 -zzan        3/3
+1.000000000 -zygotic     3/3
+\end{verbatim}
+
+Still bad. Let's turn up the juice to prefixes and suffixes that
+appear at least 10 times.
+
+\begin{verbatim}
+1.000000000 -wrought    10/10
+1.000000000 -writings   12/12
+1.000000000 -wraps      10/10
+1.000000000 -wrap       11/11
+1.000000000 -worms      69/69
+1.000000000 -worm       69/69
+1.000000000 -working    21/21
+\end{verbatim}
+
+Much better! But the next step is going to be to try removing these
+prefixes and suffixes from words that have them, to find new words.
+Since these have modularity of 100\%, we already know that every
+time we apply them, the result will already be a word. So they are
+useless for our analysis. Here are the most modular prefixes
+and suffixes with modularity {\it strictly less than} 1.
+
+\begin{verbatim}
+0.985714286 -makers     69/70
+0.985714286 -maker      69/70
+0.983606557 -wood       120/122
+0.983471074 -woods      119/121
+0.982758621 -down       57/58
+0.982658960 -works      170/173
+0.981818182 -houses     108/110
+0.981818182 -house      108/110
+0.981132075 kilo-       52/53
+0.980752406 -less       1121/1143
+0.980743395 over-       2190/2233
+0.980000000 -books      49/50
+0.980000000 -book       49/50
+0.979591837 -proof      48/49
+0.979310345 -lessnesses 142/145
+0.979069767 -ships      421/430
+0.978723404 -lessness   184/188
+0.978723404 -board      138/141
+0.978494624 -woman      91/93
+0.978021978 -women      89/91
+0.977528090 -ship       435/445
+0.977272727 -manship    43/44
+0.976744186 -weeds      84/86
+0.976470588 after-      83/85
+0.976190476 -manships   41/42
+0.976190476 -making     41/42
+0.976190476 -craft      41/42
+0.976190476 -boats      41/42
+0.976190476 -boat       41/42
+\end{verbatim}
+
+Wow, now we're talking! The single word that cannot have ``--maker''
+removed is {\it comaker}, suggesting that {\it co} should be word
+(noun: ``What a comaker makes.'').
+
+Given this list, the next step is to identify potential words that
+can be backformed by removing prefixes or adding suffixes from existing
+words. Such a string can often be found via multiple prefixes and
+suffixes. For example, {\it twing} can be formed by removing ``--ing''
+from {\it twinging} (a false positive, since the root word is actually
+{\it twinge} in this case) as well as by removing the prefix ``lef--'',
+which has modularity of 20\% (including splits such as ``lef/tie'').
+Maybe not good justification, but {\it twing} is a pretty good word
+anyway.
+
+We define the probability of a word as its Markov probability (with
+$n=4$, as this seems to produce the best results), times the
+probability that at least one of the potential backformation rules
+applies.\!\footnote{As above we only allow backformation rules that
+  have at least 10 occurrences, to prevent degeneracy.} Here are the
+most likely words by backformation:
+
+\begin{verbatim}
+word    prob   most likely backformation rules
+==============================================
+dises   .023%  para- (0.42) fluori- (0.39) 
+               melo- (0.35) bran- (0.31)  
+tring   .020%  hams- (0.36) scep- (0.35) 
+               bows- (0.33) hearts- (0.29)  
+disms   .017%  triba- (0.31) drui- (0.30) 
+               bar- (0.27) invali- (0.27)  
+ching   .017%  day- (0.86) hot- (0.69) 
+               star- (0.51) guillo- (0.50)    
+sking   .017%  dama- (0.24) imbo- (0.18) 
+               fri- (0.18) atta- (0.17)     
+cally   .015%  anti- (0.78) specifi- (0.61)
+               magnifi- (0.55) phoni-    
+pring   .015%  days- (0.67) heads- (0.62)
+               outs- (0.54) ups- (0.51)    
+\end{verbatim}
+
+I think that this approach shows promise, but there appear to be a few
+problems: Many of these ``rules'' can be explained by bad segmentation
+(``heads--'' appearing to be modular, for example, is really just
+``head--'' plus ``s'' being a common letter.) Second, I believe the
+disjunctive probability of any rule applying is too naive for
+determining the score. For example, {\it tions} has almost a thousand
+different prefixes that could apply to it; the chance of {\it any one}
+of them applying is very nearly 1. But this is actually because ``tions''
+is just a common way for a word to end. Legitimate root words to which
+many good prefixes are applied cannot be easily distinguished from
+common suffixes by this symmetric algorithm. More work is called for
+here.
 
@@
 
-On occasion I have been accused of ``overthinking''
-
-Rob: etsy, nuuog
-Chris: tapping on the bisphenol-A, nurm
-David: wafflucinations
-Lea: hnfff sound
-Reed: pansepticon
-Jessica: gruntle (like not disgruntled)
-
-In order to not reprint everyone's bullshit---but not introduce bias by selectively removing data---I discarded random subsets of the data until it did not contain bullshit any more.
-
-% TODO: backformation, stripping dis- and anti- and un- from words
-% then checking to see if they're still words?
+On occasion I have been accused of ``overthinking'' problems, whatever
+that means. So to compare, I next hazarded a tried and true technique
+from grade school, the survey.
+
+I asked a few people who happened to be around, ``What word ought to
+exist?'' Most people did not know what to make of this question, and
+also, because people seem to revel in the opportunity to get (well
+deserved) revenge on me by being disruptive trolls, many of the
+answers were designed to be unusable. In order to not reprint
+everyone's bullshit---but not introduce bias by selectively removing
+data---I discarded random subsets of the data until it did not contain
+bullshit any more.
+
+\begin{center}
+\begin{tabular}{rl}
+Rob: &  etsy, nuuog \\
+Chris: &  nurm \\
+David: &  wafflucinations \\
+Lea: &  hnfff \\
+Reed: &  pansepticon \\
+Jessica: &  gruntle \\
+\end{tabular}
+\end{center}
+
+From this we can conclude that 16\% of people wish {\it nurm} were a word, and so on.
+These words did not come with definitions, except for {\it gruntle}, which Jessica
+gives as ``the opposite of disgruntle''. This is actually already a word, but
+it was the inspiration for Section~\ref{sec:backformation}. {\it etsy} is the name
+of a popular on-line crafts community so I don't know why Rob would suggest that.
+The meaning of {\it wafflucinations} is clear from morphological analysis.
+
+% TODO: length normalization. just take the average (or max?)
+% probability for a word of length n, and normalize by that (which means what?)?
 
@@
 
-\section{Recommendations}
-
-sweeeeeeeeeeeeeeeeeeet with 19 {\it e}s, which means ``Really sweet.''
-
-``rane'' sounds too much like ``rain'', but ``sare'' has a unique pronunciation and many people seem to think it's already a word.
-``cho'' is similarly easy to pronounce and spell.
-I propose that it be defined as ``A kind of cheese,'' so that we can really nail the new triple entendre on the classic joke.
-
-% \begin{Verbatim}
-facebook wall posts, 4-grams
-100 top paths:
-0.002525504 pittsburgh
-0.002093510 steelers
-0.002093510 sfo
-0.001953943 it's
-0.001256106 bdl
-0.001116539 sigbovik
-0.001099196 facebook
-0.000976971 mic
-0.000976971 s
-0.000976971 can't
-0.000837404 i'm
-0.000837404 icfp
-0.000837404 app
-0.000697837 x
-0.000697837 drunj
-0.000697837 g
-0.000616021 ther
-0.000558269 doesn't
-0.000558269 's
-0.000558269 sr
-0.000558269 pm
-0.000558269 fyi
-0.000558269 wtf
-0.000558269 k
-0.000558269 don't
-0.000558269 today's
-0.000558269 m
-0.000558269 what's
-0.000456766 iphone
-0.000429438 inted
-0.000418702 oh
-0.000418702 phl
-0.000418702 gonna
-0.000418702 pre
-0.000418702 zrh
-0.000418702 th
-0.000418702 h
-0.000418702 b
-0.000418702 n
-0.000418702 ok
-0.000418702 d
-0.000418702 that's
-0.000418702 rex
-0.000418702 fanzibar
-0.000418702 st
-0.000418702 kinda
-0.000418702 didn't
-0.000355262 cally
-0.000334962 brillobox
-0.000314027 superburrito
-0.000305304 weath
-0.000279135 techno
-0.000279135 vii
-
-trigrams, including comments:
-100 top paths:
-0.003013897 it's
-0.001856422 ther
-0.001704432 i'm
-0.001503910 sfo
-0.001102867 -
-0.001102867 oh
-0.001050350 don't
-0.001002607 haha
-0.001002607 s
-0.000902346 bdl
-0.000808769 mic
-0.000808769 that's
-0.000776981 pittsburgh
-0.000705812 thes
-0.000701825 ok
-0.000645063 can't
-0.000601564 icfp
-0.000601564 k
-0.000583335 didn't
-0.000507201 runj
-0.000501303 g
-0.000501303 omg
-0.000501303 x
-0.000485262 ins
-0.000453126 sout
-0.000450580 ands
-0.000449950 coff
-0.000425932 reat
-0.000425133 ning
-0.000416189 pittle
-0.000412468 getty
-0.000407151 ster
-0.000401043 sr
-0.000401043 pm
-0.000401043 m
-0.000401043 wtf
-0.000401043 aw
-0.000401043 d
-0.000401043 fyi
-0.000401043 ple
-0.000401043 's
-0.000400791 phon
-0.000386315 weat
-0.000384116 befor
-0.000375978 els
-0.000374307 als
-0.000357782 nothe
-0.000354769 app
-0.000354769 appy
-0.000354422 they're
-0.000353750 wate
-0.000345727 there's
-0.000340508 res
-0.000339344 mes
-0.000334202 ove
-0.000326473 wher
-0.000324373 lity
-0.000319011 hotos
-0.000316613 dea
-0.000315777 sigbovik
-0.000315105 ver
-0.000314965 thand
-0.000310607 outh
-0.000300782 n
-
-facebook, n = 2
-100 top paths:
-0.010628671 ing
-0.007144197 th
-0.005025442 st
-0.004291375 re
-0.002932937 ne
-0.002662094 se
-0.002546194 whe
-0.002108786 ther
-0.001923496 al
-0.001669458 con
-0.001652820 wor
-0.001498222 ch
-0.001495485 wis
-0.001379778 i'm
-0.001336166 jus
-0.001302469 sh
-0.001280888 yout
-0.001244352 fo
-0.001225250 le
-0.001224422 pre
-0.001221357 ar
-0.001201881 com
-0.001200574 ge
-0.001165417 thes
-0.001116295 som
-0.001108305 fing
-0.001104851 ang
-0.001102867 -
-0.001102706 it's
-0.001039704 ist
-0.001036886 hing
-0.001007930 ou
-0.001002607 s
-0.000967282 de
-0.000932965 en
-0.000926568 tor
-0.000827260 wat
-0.000827151 oh
-0.000814235 cor
-0.000812961 ned
-0.000783696 lis
-0.000762020 ple
-0.000760162 whis
-0.000759898 sed
-0.000755396 res
-0.000728412 mor
-0.000728049 ong
-0.000727818 don
-0.000720098 cand
-0.000710277 ning
-0.000708380 ithe
-0.000693547 ame
-0.000659248 ty
-0.000656211 dow
-0.000635771 ha
-0.000633225 ext
-0.000601564 k
-% \end{Verbatim}
+\section{Conclusion}
+
+In this paper I investigated several different ways of answering the
+question: What words ought to exist? Each method produces different
+words, and some don't work that well, but nonetheless we have several
+rich sources of words, each time scientifically justified. I conclude
+with a section of recommendations for words that ought to exist, along
+with definitions.
+
+\subsection{Recommendations}
+
+{\it Sweeeeeeeeeeeeeeeeeeet} with 19 {\it e}s is the clear favorite
+based on analysis of usage, so this one should be introduced. It means
+``Really sweet.''
+
+{\it Rane} sounds too much like {\it rain}, but {\it sare} has a
+unique pronunciation and many people seem to think it's already a
+word. I propose that {\it sare} be introduced as a noun meaning, ``a
+word that sounds real but isn't.''
+
+{\it Cho} is similarly easy to pronounce and spell. I propose that it
+be defined as ``A kind of cheese,'' so that we can really nail the new
+triple entendre on the classic joke. {\it Chomaker} is someone who
+makes that kind of cheese.
+
+{\it Unders} was one of the most frequently occuring words towards the
+top of many analyses. This word should be a colloquialism for
+underwear, which would probably already be understood from context.
+
+{\it Dise} is suggested by both the Markov model (as {\it dises}, {\it
+  dising}) and backformation (as {\it dises}). I like thinking of it
+as being the root of {\it paradise}, where {\it para--} means
+something like ``along side of'' or ``resembling''. So {\it dise} is
+the place you're really looking for when you get to paradise and
+realize it's just a mediocre country club.
+
+{\it Helafelo} is one hell of a fellow.
+
+\medskip
+{\bf Addendum.} During the preparation of this paper, the Scrallbe
+game has converged on a culture where the words played are real-seeming,
+with creative definitions. Examples: {\it frodeo} (``Gandalf is the
+clown.'') {\it pridefax} (``An unproven treatment for telephone
+anxiety.'') {\it eeeee} (``eeeee'') {\it orzigato} (``Move mr. roboto.
+for great justice.'') {\it stovebed} (``Now you don't have to get out
+from under the covers to make breakfast.'') {\it ovawiki} (``The free
+egg cell that anyone can edit.'') {\it gaptave} (``Two discontinuous
+musical intervals.'') {\it achoolane} (``Nostril (colloq.)'') {\it
+  gplerious} (``Completely overcome by software licenses.) {\it
+  bestcano} (``Some eruptions are better than others.'') Thanks to the
+players for their contributions, especially Ben~Blum, Chrisamaphone,
+Rob~Simmons, and Flammin~Fingers.
+
+\section*{Appendix}
+
+Here are the most likely words induced by the English Wikipedia, with $n=3$.
+I have left off the probabilities; they can be reproduced by downloading
+Wikipedia and running the software yourself, which only takes like 9 hours.
+
+s
+ther
+w
+en
+t
+c
+d
+f
+m
+b
+ser
+e
+diff
+u
+sup
+p
+n
+de
+tal
+othe
+th
+wikipedia
+rd
+con
+id
+r
+ame
+edia
+cal
+uk
+wiki
+und
+als
+st
+l
+km
+wp
+g
+x
+ext
+j
+des
+td
+ter
+h
+co
+k
+sout
+don
+o
+cound
+v
+ver
+wer
+med
+res
+ind
+la
+em
+re
+ins
+afd
+ands
+div
+htm
+infor
+rom
+wher
+ple
+mi
+acces
+ii
+ent
+unive
+ff
+refere
+ger
+bution
+lish
+ince
+wikiped
+auth
+dist
+oth
+smally
+rea
+al
+y
+est
+cle
+fr
+lation
+tral
+gov
+nd
+mily
+isbn
+cond
+bel
+befor
+hould
+ber
+pers
+inder
+les
+gree
+mation
+lar
+chan
+dia
+ove
+stary
+compan
+mon
+val
+san
+cate
+nothe
+pdf
+tes
+ster
+nort
+pril
+hist
+sq
+clast
+ful
+ment
+q
+aft
+tr
+nown
+lign
+del
+z
+ave
+dify
+stor
+sity
+artic
+pring
+rese
+mas
+li
+ord
+ca
+thes
+oldid
+lin
+es
+ress
+flage
+blace
+pland
+mar
+mes
+conce
+exter
+wate
+mp
+jame
+lity
+ft
+cology
+spany
+mic
+ve
+han
+thist
+spam
+cand
+sher
+smal
+subdiv
+tem
+jan
+ture
+inded
+bothe
+ared
+cour
+rect
+thers
+ope
+mm
+dom
+der
+mand
+tt
+dr
+phy
+cd
+websity
+stude
+lor
+bbc
+ru
+mer
+flor
+succes
+ence
+tely
+aust
+stan
+clas
+et
+outh
+pl
+ints
+vfd
+fc
+wikipedit
+sus
+forman
+contry
+au
+edu
+inding
+el
+lic
+bord
+mr
+ston
+whic
+ide
+hous
+mich
+auser
+google
+pt
+le
+doi
+engle
+ed
+gen
+userved
+cols
+num
+nate
+cer
+withis
+clude
+shous
+inted
+wikipedian
+thered
+andex
+chool
+ning
+attle
+chael
+sta
+plation
+ampion
+crea
+fontry
+lege
+prese
+grough
+iii
+mader
+augh
+ll
+ching
+fb
+furt
+yearch
+examp
+octor
+nove
+furth
+bein
+tra
+nover
+texter
+mont
+sted
+gener
+quary
+inclub
+ander
+sco
+ast
+gr
+wome
+ality
+lon
+hough
+ano
+exas
+polid
+sease
+dec
+sor
+ga
+nov
+unty
+shor
+imdb
+offic
+partic
+oved
+forn
+lan
+fm
+undex
+rus
+arted
+alia
+cong
+vill
+eason
+gue
+shough
+cc
+vil
+tember
+octory
+jr
+gove
+writy
+thand
+anding
+rev
+colog
+gover
+edian
+publist
+mics
+comple
+syster
+olding
+coor
+useum
+spect
+expan
+puble
+bate
+tribut
+oned
+coibot
+betwo
+conter
+cles
+leason
+fina
+freen
+eng
+fa
+que
+fied
+tre
+hamp
+guage
+wered
+nic
+awar
+yout
+usic
+cgi
+andy
+nrhp
+clar
+arting
+bc
+cated
+rober
+feb
+ames
+ven
+jun
+thould
+defere
+artics
+sen
+frand
+cred
+ada
+cana
+rection
+modiff
+sal
+los
+ret
+ched
+decial
+collow
+notalk
+nl
+yor
+elete
+sain
+ance
+cens
+chand
+apr
+delect
+dea
+stant
+bruary
+commer
+righ
+ch
+tor
+peral
+andia
+arth
+roup
+founty
+hor
+scies
+alth
+pete
+footnot
+html
+ath
+tribs
+ac
+chas
+eith
+pa
+tl
+refor
+bes
+bor
+cently
+fonter
+neer
+tration
+reat
+hi
+awa
+chris
+da
+wikipedits
+atter
+pp
+ents
+dvd
+arge
+wan
+sch
+sr
+fere
+inds
+norts
+tha
+nament
+anot
+oper
+aspx
+rence
+website
+minal
+brity
+dts
+apring
+goalso
+alber
+rown
+graphy
+anded
+min
+alian
+ength
+mility
+abbr
+talign
+bec
+com
+authe
+hite
+pic
+pation
+jone
+ords
+demy
+alt
+wated
+alife
+totalk
+lond
+tage
+ad
+ana
+evers
+amg
+gened
+colo
+engton
+mina
+lates
+porth
+fff
+lifor
+footbally
+stry
+shing
+hase
+lis
+eletion
+rol
+mally
+inton
+ded
+doesn
+ny
+mus
+pro
+du
+pre
+somes
+dired
+na
+becamp
+ire
+inces
+cologo
+mong
+swith
+acts
+aree
+ang
+els
+noth
+unic
+mous
+dest
+ali
+deat
+smale
+poing
+dc
+ission
+iv
+oct
+posity
+pred
+poss
+kore
+anda
+ja
+ms
+gard
+coi
+frome
+afc
+nhl
+publice
+mothe
+jers
+maded
+qual
+geory
+roy
+vid
+pur
+studing
+ip
+ign
+fath
+cally
+coundex
+col
+abouth
+twork
+threen
+img
+ess
+lishe
+lable
+reland
+cing
+chi
+thern
+dete
+ar
+ving
+di
+eur
+wers
+il
+dio
+hl
+ando
+seast
+sation
+orge
+ble
+sep
+db
+classion
+cm
+las
+conver
+evel
+ing
+popula
+uns
+polic
+vs
+frans
+ker
+ann
+phich
+ovember
+alish
+cher
+therence
+prock
+ife
+feate
+uservice
+indid
+unded
+powere
+hs
+phot
+assion
+von
+dece
+theight
+proview
+ama
+prote
+kg
+birthe
+specian
+userve
+se
+nois
+cfm
+ma
+len
+ause
+gle
+jp
+flagic
+differe
+counder
+rance
+swed
+lee
+tel
+ont
+sould
+whics
+artion
+rans
+ric
+parly
+overy
+ves
+sa
+grea
+ans
+une
+tary
+ss
+jul
+ne
+fami
+dever
+ange
+hms
+oright
+dation
+cription
+prothe
+univer
+lat
+tota
+un
+adming
+aaa
+ampan
+perfor
+athe
+seased
+pov
+becaust
+alson
+thed
+methe
+aff
+spand
+inc
+sis
+stral
+refered
+endex
+xfd
+texamp
+valign
+sha
+dists
+fron
+heade
+lood
+thest
+unived
+italk
+brary
+colled
+ucfirst
+ite
+gan
+ep
+rary
+oney
+wikipedition
+ra
+ex
+scord
+cition
+hout
+clos
+bon
+lp
+imple
+acade
+flagico
+bost
+isn
+spection
+samp
+ral
+tead
+sant
+roces
+yorks
+ories
+bords
+unition
+che
+dra
+epison
+eff
+stal
+abour
+ach
+earch
+busic
+reate
+conves
+ses
+hims
+mone
+fran
+carly
+nated
+lished
+fra
+pm
+oh
+att
+delet
+unior
+reporth
+daving
+emple
+tworld
+noved
+trand
+ross
+ie
+vall
+orld
+vi
+becauser
+caust
+dard
+exand
+feder
+ches
+yound
+gence
+bol
+stribut
+rement
+gf
+sugges
+tity
+forder
+aug
+systed
+mathe
+websited
+abover
+ltd
+recember
+colle
+wils
+mos
+sman
+lete
+dow
+eles
+throup
+ba
+unt
+expr
+formany
+df
+ments
+tect
+pape
+hool
+exper
+butions
+hited
+delection
+deces
+coung
+gon
+gers
+stors
+abc
+partics
+writion
+furthe
+rade
+mult
+bb
+lism
+afries
+frence
+bish
+arence
+tems
+musion
+colore
+ethe
+bute
+vie
+ia
+hing
+oute
+hom
+clain
+nom
+missue
+hus
+ns
+chich
+oxfor
+contribut
+vict
+autom
+boile
+novers
+aus
+airst
+jos
+proble
+werence
+georgan
+rel
+eas
+opedia
+rs
+ned
+aut
+iss
+pg
+ct
+othis
+indo
+shi
+dst
+alf
+ian
+grounty
+hile
+mor
+jor
+grap
+parth
+elet
+regory
+thems
+ell
+septed
+pc
+url
+gnf
+unds
+appy
+studed
+offere
+gb
+timent
+sed
+dary
+eved
+hally
+deate
+hr
+taly
+noming
+duri
+cented
+thead
+aliam
+mored
+pmid
+hered
+hong
+recore
+vious
+wates
+fathe
+goved
+squary
+soment
+mout
+rela
+eview
+unives
+shere
+trate
+betweek
+gia
+mory
+cas
+dal
+fi
+enge
+stribs
+colspan
+vol
+facted
+amer
+reser
+loc
+fination
+deven
+andid
+leade
+tand
+protes
+gar
+dely
+til
+alled
+sary
+tro
+alist
+ng
+cha
+charly
+ab
+hel
+begion
+mana
+sk
+powed
+ted
+calign
+wikipe
+dar
+nm
+churce
+mil
+ren
+dise
+rever
+sel
+howere
+nexter
+commen
+sonal
+oly
+subser
+coibothe
+shool
+hal
+froman
+eu
+waship
+pr
+rease
+ph
+rical
+op
+assed
+bility
+ording
+conces
+mal
+stitle
+iucn
+sri
+os
+mond
+lations
+mance
+artment
+rigin
+rt
+brite
+ffff
+jsp
+lb
+dd
+wikipeding
+lant
+sett
+arter
+te
+yource
+sh
+contribs
+footbal
+zh
+aution
+acter
+addit
+govers
+coln
+useries
+ract
+thich
+tain
+inten
+sec
+aa
+blo
+janual
+hon
+nowing
+letion
+tition
+pute
+goldid
+communite
+wors
+fontribut
+areat
+titute
+rected
+summent
+stational
+hu
+counded
+sp
+eld
+ples
+yeare
+nating
+fing
+foll
+reque
+fiel
+pol
+josed
+atland
+phile
+manuary
+civision
+ok
+feded
+ult
+dj
+sas
+alle
+cg
+buildid
+kee
+redia
+ap
+unk
+ko
+cr
+inver
+nomy
+sund
+toon
+reg
+aaaa
+cath
+dll
+lc
+ters
+communive
+bal
+prope
+refef
+ania
+mrs
+hased
+worder
+pria
+arry
+centing
+spon
+ade
+vand
+dor
+sourch
+tooks
+seph
+capan
+pbb
+colong
+sql
+petion
+shors
+fift
+prily
+prial
+brition
+fl
+seption
+nz
+fonto
+cbs
+fally
+whill
+losed
+cluded
+mations
+monal
+sland
+pict
+russion
+profilm
+coundo
+reture
+harly
+soute
+sc
+rr
+titution
+therent
+shouser
+pager
+outhe
+quent
+spons
+fn
+ited
+proces
+bable
+atte
+whicle
+sile
+si
+rece
+unitor
+yan
+grought
+cos
+pian
+comput
+trans
+jource
+arthe
+tury
+specia
+artice
+provid
+raction
+deleter
+alo
+fored
+ti
+eague
+wiking
+conth
+ree
+musly
+ars
+mf
+ups
+schood
+rey
+donal
+sected
+writed
+freet
+formate
+cre
+colution
+opers
+chen
+bd
+hp
+sember
+folled
+pagest
+smales
+wen
+socies
+cs
+engin
+aften
+chur
+delp
+bia
+ce
+nors
+cource
+ussion
+fontribs
+summan
+arties
+jon
+dism
+lang
+sking
+augus
+theren
+xs
+seve
+deball
+er
+uning
+truction
+nomic
+creat
+gian
+acco
+sv
+pera
+alls
+rict
+relete
+rects
+coord
+stre
+ences
+aread
+spane
+meth
+brea
+ener
+thernal
+thire
+whical
+rounty
+thu
+cl
+pola
+una
+thround
+uservation
+volved
+hund
+aga
+cames
+menter
+froma
+uset
+publis
+frey
+libe
+posite
+northe
+ska
+appedia
+recial
+ro
+fant
+hought
+mination
+geo
+presult
+andon
+occup
+ala
+nas
+thists
+ew
+maring
+secore
+nee
+dian
+inders
+inves
+infore
+hower
+sume
+scal
+kans
+indust
+eding
+ka
+aux
+brid
+var
+orded
+gived
+backgrough
+hugh
+ado
+prive
+eb
+rading
+joing
+hown
+euro
+lim
+kan
+chus
+alifor
+wille
+deced
+conven
+blogy
+nbc
+quall
+coller
+partion
+folle
+eleter
+reven
+reques
+lition
+humber
+moder
+backgroup
+tribution
+flages
+lears
+forch
+reight
+pault
+evely
+mayork
+sition
+feath
+recont
+guit
+metion
+hony
+deview
+nume
+confere
+fle
+unitory
+rican
+ima
+accore
+deside
+ris
+pedia
+lisher
+acture
+aland
+userview
+thate
+polor
+kar
+nel
+sents
+pany
+cluding
+vard
+bing
+woul
+havel
+greer
+thang
+genry
+effere
+locall
+desh
+thors
+empt
+sult
+quence
+mir
+blook
+porthe
+tran
+rew
+bording
+shought
+kely
+neight
+lam
+stion
+publishe
+hors
+wis
+apan
+nony
+sourn
+nage
+orce
+profest
+jonal
+fonth
+younty
+deta
+pd
+wa
+lo
+peries
+flign
+fource
+ros
+rfc
+mely
+mt
+smalled
+ge
+retion
+relect
+vern
+alid
+aligh
+userves
+speer
+willa
+infl
+undon
+dels
+aftern
+gres
+namedia
+advan
+distor
+oria
+subdiving
+lx
+presign
+blocal
+edway
+def
+scho
+fic
+counds
+belf
+sily
+metry
+cread
+pi
+novery
+occes
+cribut
+holor
+reces
+inte
+howed
+cout
+chis
+ency
+actory
+lange
+cf
+nc
+bornia
+adven
+revie
+sames
+newspan
+shan
+orial
+userry
+suff
+eth
+moto
+showere
+hunty
+nucle
+witz
+ider
+mune
+flagicont
+aire
+shtm
+fri
+guary
+rigion
+mome
+afted
+heate
+mo
+formedia
+oble
+cnn
+nea
+ha
+canage
+borning
+tol
+navbox
+ava
+octors
+md
+frent
+offical
+wal
+arent
+anon
+artish
+cration
+rell
+dir
+auding
+ira
+butional
+labout
+bott
+southe
+dence
+sporth
+formall
+ps
+intry
+thamp
+stube
+unus
+usering
+ron
+usert
+conse
+brote
+lib
+oces
+bg
+proad
+tas
+werent
+chese
+lyn
+atta
+pris
+oile
+drage
+unre
+whenry
+ement
+stada
+chara
+raphy
+tection
+rica
+reletion
+foothe
+titler
+lassion
+areen
+theim
+conship
+formand
+ution
+unio
+ided
+guid
+riging
+brited
+buch
+septem
+eleted
+conds
+preser
+ta
+ort
+feld
+ine
+actional
+alize
+tex
+fer
+aused
+engly
+eque
+sevel
+atted
+smalle
+thome
+alty
+reath
+gation
+prefere
+hd
+nr
+peration
+covember
+govember
+livision
+fren
+dq
+aren
+deve
+va
+ffcc
+jo
+prian
+afl
+gion
+medy
+stategory
+nomich
+tured
+mel
+ene
+appa
+licy
+linal
+rought
+areate
+bethe
+othey
+eral
+opport
+ara
+kw
+whild
+cance
+communited
+varing
+brothe
+pia
+mu
+sched
+socian
+poin
+vis
+obs
+squal
+paking
+sma
+convers
+noing
+maken
+cathe
+ni
+solor
+alry
+flagicons
+aaaaa
+weren
+artical
+mtv
+flagion
+chall
+alm
+upons
+dister
+searly
+sur
+towned
+oring
+ests
+avg
+whis
+publicy
+ec
+thange
+gp
+ced
+pes
+whick
+figh
+peted
+fed
+sween
+discore
+sil
+ds
+sournal
+tay
+dier
+kim
+adver
+reak
+mote
+wouldn
+paign
+inal
+medit
+pointo
+cata
+sper
+pts
+idw
+mlb
+guis
+cational
+ul
+wasn
+septer
+aird
+throm
+fe
+feated
+collo
+itary
+uni
+uple
+distory
+disa
+regist
+vere
+effor
+whican
+ober
+bettle
+anning
+cution
+speciate
+palign
+ai
+cit
+moview
+flagicon
+af
+covery
+govery
+lina
+reast
+creen
+texand
+exten
+vic
 

### 2012-03-10 pants/pants.tex
some progress on this fine paper

\documentclass[twocolumn]{article}
\usepackage{graphicx}
\usepackage{amssymb}

\begin{document}

\newcommand\comment[1]{}

\title{{\bf \huge i-PHONE app stor }:\\
       where is my pants
       }

\author{Dr. Tom Murphy VII, Ph.D. $\qquad$ Dr. Abigale G. Lade, Ph.D.}
\date{1 April 2012}

\maketitle

\begin{abstract}
Dear I--PHONE app stor\,\,\,\, please accept my applicatian for publishment of my game
\end{abstract}

\footnotetext[-1]{Copyright \copyright\ 2012 the Regents of the Wikiplia
Foundation. Appears in SIGBOVIK 2012 with the permission of the
Association for Computational Heresy; {\em IEEEEEE!} press,
Verlag-Verlag volume no.~0x40-2A. Distribute free-ily. $\stackrel\Game\heartsuit$ 0.00}

\vspace{1em}
{\noindent \small {\bf Keywords}:
  pants, where, is my; juvenilia
}

\section{Introduction}

Dear {\sf app stor}$\quad$ I want you to have my game . My game is
{\it where is my pants}. Where is my Pants is a newwest game for the i-touch (and I-PAD, ect.). I believe it should be publiced it the {\sf app stor}. For every1 to play the game and find my pants. Because

\subsection{Reasons because it should be publiced}
\begin{enumerate}
\item This game it SO FUN!!
\item $\cup$ can play it any time or any place
\item I believe that if you put your mind to it, $\cup$ can accomplish anything
\item There are one thousand people with i-touch (and !`-PAD ect.) . If evry touch and/or PAD bought this game. Then I would have at least RICH!!!!!!!
\item I don't know where did my pants went
\end{enumerate}

\subsection{Price of game}

\begin{enumerate}
\item The price of the {\it where is my pants} will be \$9.99 USA.
\end{enumerate}

\subsection{How to play}

\begin{enumerate}
\item get down load game in i- touch
\item lunch game
\item look in the screen to see if you find pants
\item touch pants to see where is my pants
\item lotta of people have pants! so keep touching
\item if the pants 
\end{enumerate}

\end{document}

### 2012-11-17 trunk/powerhour/paper/paper.tex
Some screenshots for paper

\documentclass[twocolumn]{article}
\usepackage[top=1.1in, left=0.85in, right=0.85in]{geometry}

% \usepackage{eclbkbox}
\usepackage{amsmath}
\usepackage{amssymb}
% \usepackage{code}
% \usepackage{amscd}
% \usepackage{xy}
\usepackage{graphicx}
% \usepackage{fancyhdr}
% \usepackage{color}
% \usepackage[dark,all,bottom,landscape,timestamp]{draftcopy}
% \usepackage{everypage}

% \pagestyle{empty}

% \usepackage{ulem}
% go back to italics for emphasis, though
% \normalem

\newcommand\sfrac[2]{{}\,^{#1}\!/{}\!_{#2}}

\begin{document}

\title{New results in $\sfrac{k}{n}$ Power-Hours}
\author{Dr.~Tom~Murphy~VII~Ph.D.\thanks{
Copyright \copyright\ 2012 the Regents of the Wikiplia
Foundation. Appears in SIGBOVIK 2013 with the chagrin of the
Association for Computational Heresy; {\em IEEEEEE!} press,
Verlag-Verlag volume no.~0x40-2A.
\yen 0.00}
}

\renewcommand\>{$>$}
\newcommand\<{$<$}

\date{1 April 2011}

\maketitle

\begin{abstract}
Something about the k-n hours

\end{abstract}

\vspace{1em}
{\noindent \small {\bf Keywords}:
  generalized binge drinking, maths,
  finite-state automata,
  abstract interpretation
}

\section*{Introduction}
Here's where I tell you about doo-dah and so-so.

\begin{figure}
\begin{center}
\includegraphics[width=0.90 \linewidth]{3and2.pdf}
\end{center}\vspace{-0.1in}
\caption{Outcomes possible for the first two players in all different
  3-player power hours (black), overlayed by all possible outcomes
  outcomes for 2-player power hours (red). Mainly included because
  it looks pretty sweet.
}
\label{fig:3and2}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=0.90 \linewidth]{powerday3.pdf}
\end{center}\vspace{-0.1in}
\caption{All possible outcomes for the first two players in 3-player
  power days. These are the same games as the 3-player power hours,
  but at this scale makes it clear the groupings and their sparsity
  in the limit. Lines plotted from (0,0) to (60,*) and (*,60) show
  significant structure, but don't explain some of the interior points.
  (TODO: what are these games? what did the other player drink?)
}
\label{fig:powerday3}
\end{figure}

\section{Some stuffff}

Hi

\end{document}

### 2012-11-17 trunk/tasbot/emulator.h
new emulator interface and conformance test

@@ -1,0 +1,39 @@
+/*
+  Library interface to FCEUX.
+
+  Ideally this would be a proper class (really no reason not to),
+  but because FCEUX uses loads of global variables, this has
+  to be treated as a singleton.
+*/
+
+#ifndef __EMULATOR_H
+#define __EMULATOR_H
+
+#include <vector>
+#include <string>
+
+#include "fceu/types.h"
+
+using namespace std;
+
+struct Emulator {
+  // Returns false upon error. Only initialize once.
+  static bool Initialize(const string &romfile);
+  // Calls some internal FCEUX stuff, probably not necessary.
+  static void Shutdown();
+
+  static void Save(vector<uint8> *out);
+  static void Load(const vector<uint8> &in);
+
+  // Make one emulator step with the given input.
+  // Bits from MSB to LSB are
+  //    RLDUTSBA (Right, Left, Down, Up, sTart, Select, B, A)
+  static void Step(uint8 inputs);
+
+  // Returns 64-bit checksum (based on MD5, endianness-dependent)
+  // of RAM (only). Note there are other important bits of state.
+  static uint64 RamChecksum();
+};
+
+
+#endif

### 2012-11-17 trunk/tasbot/fceu/input.cpp

@@ -103,6 +103,14 @@ extern uint8 coinon;
 //set to true if the fourscore is attached
 static bool FSAttached = false;
 
+// Joyports are where the emulator looks to find input during simulation.
+// They're set by FCEUI_SetInput. Each joyport knows its index (w), type,
+// and pointer to data. I think the data pointer for two gamepads is usually
+// the same. -tom7
+//
+// Ultimately these get copied into joy[4]. I don't know which is which
+// (see the confusing UpdateGP below) but it seems joy[0] is the least significant
+// byte of the pointer. -tom7
 JOYPORT joyports[2] = { JOYPORT(0), JOYPORT(1) };
 FCPORT portFC;
 

### 2012-11-17 trunk/tasbot/simplefm2.h

@@ -1,0 +1,28 @@
+/* Simplified FM2 reader. Only supports one gamepad.
+   Assumes that the movie starts with hard power-on
+   in the first frame. Ignores everything else. */
+
+#ifndef __SIMPLEFM2_H
+#define __SIMPLEFM2_H
+
+#include <vector>
+#include <string>
+
+#include "fceu/types.h"
+
+#define INPUT_R (1<<7)
+#define INPUT_L (1<<6)
+#define INPUT_D (1<<5)
+#define INPUT_U (1<<4)
+#define INPUT_T (1<<3)
+#define INPUT_S (1<<2)
+#define INPUT_B (1<<1)
+#define INPUT_A (1   )
+
+using namespace std;
+
+struct SimpleFM2 {
+  static vector<uint8> ReadInputs(const string &filename);
+};
+
+#endif

### 2012-11-17 trunk/tasbot/tasbot.cc

@@ -34,7 +34,7 @@
 #include "fceu/oldmovie.h"
 #include "fceu/types.h"
 
-extern double g_fpsScale;
+#include "../cc-lib/util.h"
 
 int CloseGame(void);
 
@@ -109,8 +109,13 @@ static void DoFun(int frameskip) {
 
   uint8 v = RAM[0x0009];
   uint8 s = RAM[0x000B];  // Should be 77.
+  uint32 loc = (RAM[0x0080] << 24) |
+    (RAM[0x0081] << 16) |
+    (RAM[0x0082] << 8) |
+    (RAM[0x0083]);
   // fprintf(stderr, "%02x %02x\n", v, s);
 
+#if 0
   std::vector<uint8> savestate;
   EMUFILE_MEMORY ms(&savestate);
   // Compression yields 2x slowdown, but states go from ~80kb to 1.4kb
@@ -127,6 +132,7 @@ static void DoFun(int frameskip) {
 
   ms.trim();
   fprintf(stderr, "SS: %lld\n", savestate.size());
+#endif
 
   // uint8 FCEUI_MemSafePeek(uint16 A);
   // void FCEUI_MemPoke(uint16 a, uint8 v, int hl);
@@ -184,28 +190,6 @@ static void DriverKill() {
 void FCEUD_Update(uint8 *XBuf, int32 *Buffer, int Count) {
 }
 
-/**
- * Opens a file to be read a byte at a time.
- */
-EMUFILE_FILE* FCEUD_UTF8_fstream(const char *fn, const char *m)
-{
-  std::ios_base::openmode mode = std::ios_base::binary;
-  if(!strcmp(m,"r") || !strcmp(m,"rb"))
-    mode |= std::ios_base::in;
-  else if(!strcmp(m,"w") || !strcmp(m,"wb"))
-    mode |= std::ios_base::out | std::ios_base::trunc;
-  else if(!strcmp(m,"a") || !strcmp(m,"ab"))
-    mode |= std::ios_base::out | std::ios_base::app;
-  else if(!strcmp(m,"r+") || !strcmp(m,"r+b"))
-    mode |= std::ios_base::in | std::ios_base::out;
-  else if(!strcmp(m,"w+") || !strcmp(m,"w+b"))
-    mode |= std::ios_base::in | std::ios_base::out | std::ios_base::trunc;
-  else if(!strcmp(m,"a+") || !strcmp(m,"a+b"))
[…]

### 2012-11-17 tasbot/emu_test.cc
new emulator interface and conformance test

@@ -1,0 +1,158 @@
+/* This tests the interface to the emulator library
+   for correctness, by playing back the movie karate.fm2
+   against the rom karate.nes, and checking that
+   the game is won and that the RAM has the right
+   contents. It also does some simple timing. */
+
+#include <unistd.h>
+#include <sys/types.h>
+#include <signal.h>
+#include <sys/time.h>
+#include <sys/stat.h>
+#include <string.h>
+#include <strings.h>
+#include <errno.h>
+#include <stdio.h>
+#include <stdlib.h>
+#include <iostream>
+#include <fstream>
+#include <limits.h>
+#include <math.h>
+
+#include <zlib.h>
+
+#include "fceu/utils/md5.h"
+
+#include "config.h"
+
+#include "fceu/driver.h"
+#include "fceu/drivers/common/args.h"
+
+#include "fceu/state.h"
+
+#include "fceu/drivers/common/cheat.h"
+#include "fceu/fceu.h"
+#include "fceu/movie.h"
+#include "fceu/version.h"
+
+#include "fceu/oldmovie.h"
+#include "fceu/types.h"
+
+#include "../cc-lib/util.h"
+
+#include "simplefm2.h"
+#include "emulator.h"
+
+static void DoFun(int frameskip) {
+  uint8 *gfx;
+  int32 *sound;
+  int32 ssize;
+  static int opause = 0;
+
+  // fprintf(stderr, "In DoFun..\n");
+
+  // Limited ability to skip video and sound.
+  #define SKIP_VIDEO_AND_SOUND 2
+
+  // Emulate a single frame.
+  FCEUI_Emulate(NULL, &sound, &ssize, SKIP_VIDEO_AND_SOUND);
+
+  // This was the only useful thing from Update. It's called multiple
+  // times; I don't know why.
+  // --- update input! --
+
+  uint8 v = RAM[0x0009];
+  uint8 s = RAM[0x000B];  // Should be 77.
+  uint32 loc = (RAM[0x0080] << 24) |
+    (RAM[0x0081] << 16) |
+    (RAM[0x0082] << 8) |
+    (RAM[0x0083]);
+  // fprintf(stderr, "%02x %02x\n", v, s);
+}
+
+/**
+ * Update the video, audio, and input subsystems with the provided
+ * video (XBuf) and audio (Buffer) information.
+ */
[…]

### 2012-11-17 tasbot/emulator.cc

@@ -1,0 +1,217 @@
+
+#include "emulator.h"
+
+#include <string>
+#include <vector>
+#include <zlib.h>
+
+#include "config.h"
+#include "fceu/driver.h"
+#include "fceu/fceu.h"
+#include "fceu/types.h"
+#include "fceu/utils/md5.h"
+#include "fceu/version.h"
+#include "fceu/state.h"
+
+// Joystick data. I think used for both controller 0 and 1. Part of
+// the "API".
+static uint32 joydata = 0;
+static bool initialized = false;
+
+uint64 Emulator::RamChecksum() {
+  md5_context ctx;
+  md5_starts(&ctx);
+  md5_update(&ctx, RAM, 0x800);
+  uint8 digest[16];
+  md5_finish(&ctx, digest);
+  return *(uint64*)digest;
+}
+/**
+ * Initialize all of the subsystem drivers: video, audio, and joystick.
+ */
+static int DriverInitialize(FCEUGI *gi) {
+  // Used to init video. I think it's safe to skip.
+
+  // Here we initialized sound. Assuming it's safe to skip,
+  // because of an early return if config turned it off.
+
+  // Used to init joysticks. Don't care about that.
+
+  // No fourscore support.
+  // eoptions &= ~EO_FOURSCORE;
+
+  // Why do both point to the same joydata? -tom
+  FCEUI_SetInput (0, SI_GAMEPAD, &joydata, 0);
+  FCEUI_SetInput (1, SI_GAMEPAD, &joydata, 0);
+
+  FCEUI_SetInputFourscore (false);
+  return 1;
+}
+
+/**
+ * Closes a game.  Frees memory, and deinitializes the drivers.
+ */
+int CloseGame() {
+  FCEUI_CloseGame();
+  GameInfo = 0;
+  return 1;
+}
+
+/**
+ * Loads a game, given a full path/filename.  The driver code must be
+ * initialized after the game is loaded, because the emulator code
+ * provides data necessary for the driver code(number of scanlines to
+ * render, what virtual input devices to use, etc.).
+ */
+int LoadGame(const char *path) {
+  CloseGame();
+  if(!FCEUI_LoadGame(path, 1)) {
+    return 0;
+  }
+
+  // Here we used to do ParseGIInput, which allows the gameinfo
+  // to override our input config, or something like that. No
+  // weird stuff. Skip it.
+
[…]

### 2012-11-17 tasbot/headless-driver.cc

@@ -62,6 +62,28 @@ FILE *FCEUD_UTF8fopen(const char *fn, const char *mode) {
 }
 
 /**
+ * Opens a file to be read a byte at a time.
+ */
+EMUFILE_FILE* FCEUD_UTF8_fstream(const char *fn, const char *m)
+{
+  std::ios_base::openmode mode = std::ios_base::binary;
+  if(!strcmp(m,"r") || !strcmp(m,"rb"))
+    mode |= std::ios_base::in;
+  else if(!strcmp(m,"w") || !strcmp(m,"wb"))
+    mode |= std::ios_base::out | std::ios_base::trunc;
+  else if(!strcmp(m,"a") || !strcmp(m,"ab"))
+    mode |= std::ios_base::out | std::ios_base::app;
+  else if(!strcmp(m,"r+") || !strcmp(m,"r+b"))
+    mode |= std::ios_base::in | std::ios_base::out;
+  else if(!strcmp(m,"w+") || !strcmp(m,"w+b"))
+    mode |= std::ios_base::in | std::ios_base::out | std::ios_base::trunc;
+  else if(!strcmp(m,"a+") || !strcmp(m,"a+b"))
+    mode |= std::ios_base::in | std::ios_base::out | std::ios_base::app;
+  return new EMUFILE_FILE(fn, m);
+  //return new std::fstream(fn,mode);
+}
+
+/**
  * Returns the compiler string.
  */
 const char *FCEUD_GetCompilerString() {
@@ -168,3 +190,25 @@ void FCEUD_TurboToggle() { }
 
 void FCEUD_DebugBreakpoint(int bp_num) {}
 void FCEUD_TraceInstruction(unsigned char*, int) {}
+
+/**
+ * Get the time in ticks.
+ */
+uint64 FCEUD_GetTime() {
+  //	return SDL_GetTicks();
+  fprintf(stderr, "(FCEUD_GetTime) In headless mode, nothing should "
+	  "try to do timing.\n");
+  abort();
+  return 0;
+}
+
+/**
+ * Get the tick frequency in Hz.
+ */
+uint64 FCEUD_GetTimeFreq(void) {
+  // SDL_GetTicks() is in milliseconds
+  fprintf(stderr, "(FCEUD_GetTimeFreq) In headless mode, nothing should "
+	  "try to do timing.\n");
+  abort();
+  return 1000;
+}

### 2012-11-17 tasbot/simplefm2.cc

@@ -1,0 +1,45 @@
+
+#include "simplefm2.h"
+
+#include "../cc-lib/util.h"
+
+using namespace std;
+
+vector<uint8> SimpleFM2::ReadInputs(const string &filename) {
+  vector<string> contents = Util::ReadFileToLines(filename);
+  vector<uint8> out;
+  for (int i = 0; i < contents.size(); i++) {
+    const string &line = contents[i];
+    if (line.empty() || line[0] != '|')
+      continue;
+
+    // Parse the line.
+    if (line.size() < 12) {
+      fprintf(stderr, "Illegal line: [%s]\n", line.c_str());
+      abort();
+    }
+    
+    if (!(line[1] == '0' ||
+	  (line[1] == '2' && out.empty()))) {
+      fprintf(stderr, "Command must be zero except hard "
+	      "reset in first input: [%s]\n", line.c_str());
+      abort();
+    }
+
+    /*
+      |2|........|........||
+      |0|....T...|........||
+    */
+    
+    const string player = line.substr(3, 8);
+    uint8 command = 0;
+    for (int j = 0; j < 8; j++) {
+      if (player[j] != '.') {
+	command |= (1 << (7 - j));
+      }
+    }
+    out.push_back(command);
+  }
+
+  return out;
+}

### 2012-11-17 trunk/cc-lib/util.cc
new util used in tasbot

@@ -172,6 +172,24 @@ string Util::ReadFile(const string &s) {
   return ret;
 }
 
+vector<string> Util::ReadFileToLines(const string &f) {
+  string s = ReadFile(f);
+  vector<string> v;
+  string line;
+  // PERF don't need to do so much copying.
+  for (size_t i = 0; i < s.size(); i++) {
+    if (s[i] == '\r')
+      continue;
+    else if (s[i] == '\n') {
+      v.push_back(line);
+      line.clear();
+    } else {
+      line += s[i];
+    }
+  }
+  return v;
+}
+
 static bool hasmagicf(FILE * f, const string & mag) {
   char * hdr = (char*)malloc(mag.length());
   if (!hdr) return false;
@@ -888,8 +906,8 @@ bool Util::launchurl(const string & url) {
 #endif
 
 #if WIN32
-  return ((int)ShellExecute(NULL, "open", url.c_str(), 
-			    NULL, NULL, SW_SHOWNORMAL)) > 32;
+  return ((size_t)ShellExecute(NULL, "open", url.c_str(), 
+			       NULL, NULL, SW_SHOWNORMAL)) > 32;
 #endif
 
   /* otherwise.. */

### 2012-11-17 trunk/cc-lib/util.h

@@ -7,6 +7,7 @@
 #include <stdio.h>
 #include <stdlib.h>
 #include <string>
+#include <vector>
 
 using namespace std;
 
@@ -28,6 +29,10 @@ struct Util {
 
   static string ReadFile(const string &f);
   static bool WriteFile(const string &f, const string &s);
+
+  // Reads the lines in the file to the vector. Ignores all
+  // carriage returns, including ones not followed by newline.
+  static vector<string> ReadFileToLines(const string &f);
 
   // XXX terrible names
   static int shout(int, string, unsigned int &);

### 2012-11-17 trunk/tasbot/emulator.h
Saving and loading test, which works

@@ -23,7 +23,8 @@ struct Emulator {
   static void Shutdown();
 
   static void Save(vector<uint8> *out);
-  static void Load(const vector<uint8> &in);
+  // Doesn't modify its argument.
+  static void Load(vector<uint8> *in);
 
   // Make one emulator step with the given input.
   // Bits from MSB to LSB are

### 2012-11-17 tasbot/emu_test.cc
Saving and loading test, which works

@@ -107,6 +107,37 @@ static void CheckLoc(int frame, uint32 expected) {
   }
 }
 
+// Note that fceu_frame is 1 plus the index in the input loop,
+// because the UI displays the first frame as #1.
+static void CheckCheckpoints(int fceu_frame) {
+  // XXX read from golden file.
+  switch (fceu_frame) {
+  case 20: CheckLoc(fceu_frame, 0); break;
+  case 21: CheckLoc(fceu_frame, 65536); break;
+  case 4935: CheckLoc(fceu_frame, 196948); break;
+  case 7674: CheckLoc(fceu_frame, 200273); break;
+  case 7675: CheckLoc(fceu_frame, 200274); break;
+  case 8123: CheckLoc(fceu_frame, 262144); break;
+  case 11213: CheckLoc(fceu_frame, 265916); break;
+  default:;
+  }
+}
+
+static uint64 CrapHash(int a) {
+  uint64 ret = ~a;
+  ret *= 31337;
+  ret ^= 0xDEADBEEF;
+  ret = (ret >> 17) | (ret << (64 - 17));
+  ret -= 911911911911;
+  ret *= 65537;
+  ret ^= 0xCAFEBABE;
+  return ret;
+}
+
+static bool CompareByHash(int a, int b) {
+  return CrapHash(a) < CrapHash(b);
+}
+
 /**
  * The main loop for the SDL.
  */
@@ -116,6 +147,11 @@ int main(int argc, char *argv[]) {
   vector<uint8> inputs = SimpleFM2::ReadInputs("karate.fm2");
   fprintf(stderr, "Running %d steps...\n", inputs.size());
 
+  // The nth savestate is from before issuing the nth input.
+  vector< vector<uint8> > savestates;
+
+  int64 ss_total = 0;
+
   for (int i = 0; i < inputs.size(); i++) {
     // XXX don't think this should ever happen.
     if (!GameInfo) {
@@ -123,25 +159,16 @@ int main(int argc, char *argv[]) {
       return -1;
     }
 
+    vector<uint8> v;
+    Emulator::Save(&v);
+    ss_total += v.size();
+    savestates.push_back(v);
+
     Emulator::Step(inputs[i]);
 
-    // XXX read from golden file.
     // The FCEUX UI indexes frames starting at 1.
-    const int fceu_frame = i + 1;
-    switch (fceu_frame) {
-    case 20: CheckLoc(fceu_frame, 0); break;
[…]

### 2012-11-17 tasbot/emulator.cc

@@ -211,7 +211,11 @@ void Emulator::Save(vector<uint8> *out) {
   ms.trim();
 }
 
-void Emulator::Load(const vector<uint8> &state) {
-  fprintf(stderr, "unimplemented: Load\n");
-  abort();
+void Emulator::Load(vector<uint8> *state) {
+  // XXX Wish we had EMUFILE_READONLY_MEMORY...
+  EMUFILE_MEMORY ms(state);
+  if (!FCEUSS_LoadFP(&ms, SSLOADPARAM_NOBACKUP)) {
+    fprintf(stderr, "Couldn't restore from state\n");
+    abort();
+  }
 }

### 2012-11-17 trunk/tasbot/fceu/state.cpp
Improve size and performance of in-memory savestates

@@ -24,7 +24,6 @@
 #include <stdio.h>
 #include <stdlib.h>
 #include <string.h>
-//#include <unistd.h> //mbg merge 7/17/06 removed
 
 #include <vector>
 #include <fstream>
@@ -55,9 +54,6 @@ static void (*SPostSave)(void);
 
 static int SaveStateStatus[10];
 static int StateShow;
-
-//tells the save system innards that we're loading the old format
-bool FCEU_state_loading_old_format;
 
 char lastSavestateMade[2048]; //Stores the filename of the last savestate made (needed for UndoSavestate)
 bool undoSS = false;		  //This will be true if there is lastSavestateMade, it was made since ROM was loaded, a backup state for lastSavestateMade exists
@@ -263,9 +259,7 @@ static bool ReadStateChunks(EMUFILE* is, int32 totalsize)
 		case 4:if(!ReadStateChunk(is,FCEUCTRL_STATEINFO,size)) ret=false;break;
 		case 7:
 			if(!FCEUMOV_ReadState(is,size)) {
-				//allow this to fail in old-format savestates.
-				if(!FCEU_state_loading_old_format)
-					ret=false;
+			  ret = false;
 			}
 			break;
 		case 0x10:if(!ReadStateChunk(is,SFMDATA,size)) ret=false; break;
@@ -352,345 +346,248 @@ extern int geniestage;
 
 bool FCEUSS_SaveMS(EMUFILE* outstream, int compressionLevel)
 {
-	// reinit memory_savestate
-	// memory_savestate is global variable which already has its vector of bytes, so no need to allocate memory every time we use save/loadstate
-	memory_savestate.set_len(0);	// this also seeks to the beginning
-	memory_savestate.unfail();
+  // PERF really use global file and copy? Why not into outstream?
 
-	EMUFILE* os = &memory_savestate;
+  // reinit memory_savestate
+  // memory_savestate is global variable which already has its vector of bytes, so no need to allocate memory every time we use save/loadstate
+  memory_savestate.set_len(0);	// this also seeks to the beginning
[…]

### 2012-11-17 trunk/tasbot/fceu/state.h

@@ -20,8 +20,9 @@
 
 enum ENUM_SSLOADPARAMS
 {
-	SSLOADPARAM_NOBACKUP,
-	SSLOADPARAM_BACKUP,
+  // PERF no need for backups!
+        SSLOADPARAM_NOBACKUP,
+        SSLOADPARAM_BACKUP,
 };
 
 void FCEUSS_Save(const char *);
@@ -37,14 +38,14 @@ void FCEUSS_CheckStates(void);
 
 struct SFORMAT
 {
-	//a void* to the data or a void** to the data
-	void *v;
+        //a void* to the data or a void** to the data
+        void *v;
 
-	//size, plus flags
-	uint32 s;
+        //size, plus flags
+        uint32 s;
 
-	//a string description of the element
-	char *desc;
+        //a string description of the element
+        char *desc;
 };
 
 void ResetExState(void (*PreSave)(void),void (*PostSave)(void));
@@ -62,18 +63,18 @@ void AddExState(void *v, uint32 s, int type, char *desc);
 void FCEU_DrawSaveStates(uint8 *XBuf);
 
 void CreateBackupSaveState(const char *fname); //backsup a savestate before overwriting it with a new one
-void BackupLoadState();				 //Makes a backup savestate before any loadstate
-void LoadBackup();					 //Loads the backupsavestate
-void RedoLoadState();				 //reloads a loadstate if backupsavestate was run
-void SwapSaveState();				 //Swaps a savestate with its backup state
+void BackupLoadState();                          //Makes a backup savestate before any loadstate
+void LoadBackup();                                       //Loads the backupsavestate
+void RedoLoadState();                            //reloads a loadstate if backupsavestate was run
+void SwapSaveState();                            //Swaps a savestate with its backup state
 
 extern char lastSavestateMade[2048]; //Filename of last savestate used
-extern bool undoSS;					 //undo savestate flag
-extern bool redoSS;					 //redo savestate flag
+extern bool undoSS;                                      //undo savestate flag
+extern bool redoSS;                                      //redo savestate flag
 extern char lastLoadstateMade[2048]; //Filename of last state loaded
-extern bool undoLS;					 //undo loadstate flag
-extern bool redoLS;					 //redo savestate flag
-extern bool backupSavestates;		 //Whether or not to make backups, true by default
-bool CheckBackupSaveStateExist();	 //Checks if backupsavestate exists
+extern bool undoLS;                                      //undo loadstate flag
+extern bool redoLS;                                      //redo savestate flag
[…]

### 2012-11-17 tasbot/emu_test.cc
Improve size and performance of in-memory savestates

@@ -177,10 +177,6 @@ int main(int argc, char *argv[]) {
     return -1;
   }
 
-  fprintf(stderr, "Total for %d savestates: %.2fmb (avg %.0f bytes)\n",
-          savestates.size(), ss_total / (1024.0 * 1024.0),
-	  ss_total / (double)savestates.size());
-
   fprintf(stderr, "\nTest random replay of savestates:\n");
   // Now run through each state in random order. Load it, then execute a step,
   // then check that we get to the same state as before.
@@ -206,6 +202,10 @@ int main(int argc, char *argv[]) {
     }
   }
   fprintf(stderr, "Savestates are ok.\n");
+
+  fprintf(stderr, "Total for %d savestates: %.2fmb (avg %.2f bytes)\n",
+          savestates.size(), ss_total / (1024.0 * 1024.0),
+	  ss_total / (double)savestates.size());
 
   Emulator::Shutdown();
 

### 2012-11-17 tasbot/fceu/movie.cpp

@@ -1208,15 +1208,8 @@ bool FCEUMOV_ReadState(EMUFILE* is, uint32 size)
 	MovieData tempMovieData = MovieData();
 	std::ios::pos_type curr = is->ftell();
 	if(!LoadFM2(tempMovieData, is, size, false)) {
-		is->fseek((uint32)curr+size,SEEK_SET);
-		extern bool FCEU_state_loading_old_format;
-		if(FCEU_state_loading_old_format) {
-			if(movieMode == MOVIEMODE_PLAY || movieMode == MOVIEMODE_RECORD || movieMode == MOVIEMODE_FINISHED) {
-				//FCEUI_StopMovie();  //No reason to stop the movie, nothing destructive has happened yet.
-				FCEU_PrintError("You have tried to use an old savestate while playing a movie. This is unsupported (since the old savestate has old-format movie data in it which can't be converted on the fly)");
-			}
-		}
-		return false;
+	  is->fseek((uint32)curr+size,SEEK_SET);
+	  return false;
 	}
 
 	//----------------

### 2012-11-17 tasbot/fceu/movie.cpp
Going to remove savestate header completely

@@ -805,15 +805,15 @@ bool FCEUMOV_FromPoweron()
 }
 bool MovieData::loadSavestateFrom(std::vector<uint8>* buf)
 {
-	EMUFILE_MEMORY ms(buf);
-	return FCEUSS_LoadFP(&ms,SSLOADPARAM_BACKUP);
+  EMUFILE_MEMORY ms(buf);
+  return FCEUSS_LoadFP(&ms,SSLOADPARAM_BACKUP, NULL);
 }
 
 void MovieData::dumpSavestateTo(std::vector<uint8>* buf, int compressionLevel)
 {
-	EMUFILE_MEMORY ms(buf);
-	FCEUSS_SaveMS(&ms,compressionLevel);
-	ms.trim();
+  EMUFILE_MEMORY ms(buf);
+  FCEUSS_SaveMS(&ms, compressionLevel, NULL);
+  ms.trim();
 }
 
 //begin playing an existing movie

### 2012-11-17 trunk/tasbot/emu_test.cc
Going to remove savestate header completely

@@ -94,6 +94,14 @@ static int64 DumpMem() {
   return *(int64*)digest;
 }
 
+static void PrintSavestate(const vector<uint8> &ss) {
+  printf("Savestate:\n");
+  for (int i = 0; i < ss.size(); i++) {
+    printf("%02x", ss[i]);
+  }
+  printf("\n");
+}
+
 static void CheckLoc(int frame, uint32 expected) {
   fprintf(stderr, "Frame %d expect %u\n", frame, expected);
   uint32 loc = (RAM[0x0080] << 24) |
@@ -169,6 +177,19 @@ int main(int argc, char *argv[]) {
     // The FCEUX UI indexes frames starting at 1.
     CheckCheckpoints(i + 1);
   }
+
+  /*
+  PrintSavestate(savestates[0]);
+  PrintSavestate(savestates[4935]);
+  PrintSavestate(savestates[8123]);
+  */
+
+  vector<uint8> diff;
+  for (int i = 0; i < savestates[4935].size(); i++) {
+    diff.push_back(savestates[8123][i] - savestates[4935][i]);
+  }
+
+  PrintSavestate(diff);
 
   if (0x46e75713b56aea30 == Emulator::RamChecksum()) {
     fprintf(stderr, "Memory OK.\n");

### 2012-11-17 trunk/tasbot/emulator.cc

@@ -198,7 +198,7 @@ void Emulator::Save(vector<uint8> *out) {
   EMUFILE_MEMORY ms(out);
   // Compression yields 2x slowdown, but states go from ~80kb to 1.4kb
   // Without screenshot, ~1.3kb and only 40% slowdown
-  FCEUSS_SaveMS(&ms, Z_DEFAULT_COMPRESSION /* Z_NO_COMPRESSION */);
+  FCEUSS_SaveMS(&ms, /* Z_DEFAULT_COMPRESSION */ Z_NO_COMPRESSION, NULL);
   // TODO
   // Saving is not as efficient as we'd like for a pure in-memory operation
   //  - uses tags to tell you what's next, even though we could already know
@@ -211,10 +211,16 @@ void Emulator::Save(vector<uint8> *out) {
   ms.trim();
 }
 
+void Emulator::GetBasis(vector<uint8> *out) {
+  EMUFILE_MEMORY ms(out);
+  FCEUSS_SaveMS(&ms, Z_NO_COMPRESSION, NULL);
+  ms.trim();
+}
+
 void Emulator::Load(vector<uint8> *state) {
   // XXX Wish we had EMUFILE_READONLY_MEMORY...
   EMUFILE_MEMORY ms(state);
-  if (!FCEUSS_LoadFP(&ms, SSLOADPARAM_NOBACKUP)) {
+  if (!FCEUSS_LoadFP(&ms, SSLOADPARAM_NOBACKUP, NULL)) {
     fprintf(stderr, "Couldn't restore from state\n");
     abort();
   }

### 2012-11-17 trunk/tasbot/emulator.h

@@ -34,6 +34,18 @@ struct Emulator {
   // Returns 64-bit checksum (based on MD5, endianness-dependent)
   // of RAM (only). Note there are other important bits of state.
   static uint64 RamChecksum();
+
+  // Fancy stuff. States often only differ by a small amount, so a way
+  // to reduce their entropy is to diff them against a representative
+  // savestate. This gets an uncompressed basis for the current state,
+  // which can be used in the SaveEx and LoadEx routines.
+  static void GetBasis(vector<uint8> *out);
+
+  // Save and load with a basis vector. The vector can contain anything, and
+  // doesn't even have to be the same length as an uncompressed save state,
+  // but a state needs to be loaded with the same basis as it was saved.
+  static void SaveEx(vector<uint8> *out, const vector<uint8> &basis);
+  static void LoadEx(vector<uint8> *in, const vector<uint8> &basis);
 };
 
 

### 2012-11-17 trunk/tasbot/fceu/state.cpp

@@ -344,7 +344,7 @@ int CurrentState=0;
 extern int geniestage;
 
 
-bool FCEUSS_SaveMS(EMUFILE* outstream, int compressionLevel)
+bool FCEUSS_SaveMS(EMUFILE* outstream, int compressionLevel, std::vector<uint8> *basis)
 {
   // PERF really use global file and copy? Why not into outstream?
 
@@ -481,36 +481,10 @@ void FCEUSS_Save(const char *fname) {
       return;
     }
 
-#ifdef _S9XLUA_H
-  if (!internalSaveLoad)
-    {
-      LuaSaveData saveData;
-      CallRegisteredLuaSaveFunctions(CurrentState, saveData);
-
-      char luaSaveFilename [512];
-      strncpy(luaSaveFilename, fn, 512);
-      luaSaveFilename[512-(1+7/*strlen(".luasav")*/)] = '\0';
-      strcat(luaSaveFilename, ".luasav");
-      if(saveData.recordList)
-	{
-	  FILE* luaSaveFile = fopen(luaSaveFilename, "wb");
-	  if(luaSaveFile)
-	    {
-	      saveData.ExportRecords(luaSaveFile);
-	      fclose(luaSaveFile);
-	    }
-	}
-      else
-	{
-	  unlink(luaSaveFilename);
-	}
-    }
-#endif
-
   if(FCEUMOV_Mode(MOVIEMODE_INACTIVE))
-    FCEUSS_SaveMS(st,-1);
+    FCEUSS_SaveMS(st,-1, NULL);
   else
-    FCEUSS_SaveMS(st,0);
+    FCEUSS_SaveMS(st,0, NULL);
 
   delete st;
 
@@ -522,8 +496,7 @@ void FCEUSS_Save(const char *fname) {
   redoSS = false;					//we have a new savestate so redo is not possible
 }
 
-bool FCEUSS_LoadFP(EMUFILE* is, ENUM_SSLOADPARAMS params)
-{
+bool FCEUSS_LoadFP(EMUFILE* is, ENUM_SSLOADPARAMS params, std::vector<uint8> *basis) {
   if(!is) return false;
 
   // PERF don't do this
@@ -531,7 +504,7 @@ bool FCEUSS_LoadFP(EMUFILE* is, ENUM_SSLOADPARAMS params)
   bool backup = (params == SSLOADPARAM_BACKUP);
   EMUFILE_MEMORY msBackupSavestate;
   if(backup) {
[…]

### 2012-11-17 trunk/tasbot/fceu/state.h

@@ -18,6 +18,12 @@
  * Foundation, Inc., 51 Franklin Street, Fifth Floor, Boston, MA  02110-1301  USA
  */
 
+#ifndef __STATE_H
+#define __STATE_H
+
+#include <vector>
+#include "types.h"
+
 enum ENUM_SSLOADPARAMS
 {
   // PERF no need for backups!
@@ -29,9 +35,9 @@ void FCEUSS_Save(const char *);
 bool FCEUSS_Load(const char *);
 
  //zlib values: 0 (none) through 9 (max) or -1 (default)
-bool FCEUSS_SaveMS(EMUFILE* outstream, int compressionLevel);
+bool FCEUSS_SaveMS(EMUFILE* outstream, int compressionLevel, std::vector<uint8> *basis);
 
-bool FCEUSS_LoadFP(EMUFILE* is, ENUM_SSLOADPARAMS params);
+bool FCEUSS_LoadFP(EMUFILE* is, ENUM_SSLOADPARAMS params, std::vector<uint8> *basis);
 
 extern int CurrentState;
 void FCEUSS_CheckStates(void);
@@ -78,3 +84,5 @@ extern bool backupSavestates;            //Whether or not to make backups, true 
 bool CheckBackupSaveStateExist();        //Checks if backupsavestate exists
 
 extern bool compressSavestates;         //Whether or not to compress non-movie savestates (by default, yes)
+
+#endif

### 2012-11-17 tasbot/emu_test.cc
Now able to write basis savestate to make input to compression likely to have less entropy

@@ -146,6 +146,34 @@ static bool CompareByHash(int a, int b) {
   return CrapHash(a) < CrapHash(b);
 }
 
+static vector<uint8> LoadOrComputeBasis(const vector<uint8> &inputs,
+					int frame,
+					const string &basisfile) {
+  if (Util::ExistsFile(basisfile)) {
+    fprintf(stderr, "Loading basis file %s.\n", basisfile.c_str());
+    return Util::ReadFileBytes(basisfile);
+  }
+
+  fprintf(stderr, "Computing basis file %s.\n", basisfile.c_str());
+  vector<uint8> start;
+  Emulator::Save(&start);
+  for (int i = 0; i < frame; i++) {
+    Emulator::Step(inputs[i]);
+  }
+  vector<uint8> basis;
+  Emulator::GetBasis(&basis);
+  if (!Util::WriteFileBytes(basisfile, basis)) {
+    fprintf(stderr, "Couldn't write to %s\n", basisfile.c_str());
+    abort();
+  }
+  fprintf(stderr, "Written.\n");
+  
+  // Rewind.
+  Emulator::Load(&start);
+
+  return basis;
+}
+
 /**
  * The main loop for the SDL.
  */
@@ -153,13 +181,18 @@ int main(int argc, char *argv[]) {
   Emulator::Initialize("karate.nes");
   // loop playing the game
   vector<uint8> inputs = SimpleFM2::ReadInputs("karate.fm2");
-  fprintf(stderr, "Running %d steps...\n", inputs.size());
+
+  vector<uint8> basis = LoadOrComputeBasis(inputs, 4935, "karate.basis");
 
   // The nth savestate is from before issuing the nth input.
   vector< vector<uint8> > savestates;
 
   int64 ss_total = 0;
 
+  // XXXXXX
+  // inputs.resize(10);
+
+  fprintf(stderr, "Running %d steps...\n", inputs.size());
   for (int i = 0; i < inputs.size(); i++) {
     // XXX don't think this should ever happen.
     if (!GameInfo) {
@@ -168,7 +201,7 @@ int main(int argc, char *argv[]) {
     }
 
     vector<uint8> v;
-    Emulator::Save(&v);
+    Emulator::SaveEx(&v, &basis);
     ss_total += v.size();
     savestates.push_back(v);
 
@@ -178,18 +211,22 @@ int main(int argc, char *argv[]) {
     CheckCheckpoints(i + 1);
   }
 
+#if 0
   /*
[…]

### 2012-11-17 tasbot/emulator.cc

@@ -195,33 +195,116 @@ void Emulator::Step(uint8 inputs) {
 }
 
 void Emulator::Save(vector<uint8> *out) {
-  EMUFILE_MEMORY ms(out);
-  // Compression yields 2x slowdown, but states go from ~80kb to 1.4kb
-  // Without screenshot, ~1.3kb and only 40% slowdown
-  FCEUSS_SaveMS(&ms, /* Z_DEFAULT_COMPRESSION */ Z_NO_COMPRESSION, NULL);
+  SaveEx(out, NULL);
+}
+
+void Emulator::GetBasis(vector<uint8> *out) {
+  FCEUSS_SaveRAW(out);
+}
+
+void Emulator::Load(vector<uint8> *state) {
+  LoadEx(state, NULL);
+}
+
+// Compression yields 2x slowdown, but states go from ~80kb to 1.4kb
+// Without screenshot, ~1.3kb and only 40% slowdown
+#define USE_COMPRESSION 1
+
+#if USE_COMPRESSION
+
+void Emulator::SaveEx(vector<uint8> *state, const vector<uint8> *basis) {
   // TODO
   // Saving is not as efficient as we'd like for a pure in-memory operation
   //  - uses tags to tell you what's next, even though we could already know
   //  - takes care for endianness; no point
-  //  - saves the backing buffer (write-only, used for display)
   //  - might save some other write-only data (sound?)
-  //  - compresses the output using zlib, which may be good for space
-  //    but not good for time!
 
-  ms.trim();
+  vector<uint8> raw;
+  FCEUSS_SaveRAW(&raw);
+
+  // Encode.
+  int blen = (basis == NULL) ? 0 : (min(basis->size(), raw.size()));
+  for (int i = 0; i < blen; i++) {
+    raw[i] -= (*basis)[i];
+  }
+
+  // Compress.
+  int error = Z_OK;
+  uint8* cbuf = &raw[0];
+  int len = raw.size();
+  // worst case compression: zlib says "0.1% larger than sourceLen plus 12 bytes"
+  uLongf comprlen = (len >> 9) + 12 + len;
+
+  // Make sure there is contiguous space. Need room for header too.
+  state->resize(4 + comprlen);
+
+  if (Z_OK != compress2(&(*state)[4], &comprlen, &raw[0], len, Z_DEFAULT_COMPRESSION)) {
+    fprintf(stderr, "Couldn't compress.\n");
[…]

### 2012-11-17 tasbot/emulator.h

@@ -44,8 +44,9 @@ struct Emulator {
   // Save and load with a basis vector. The vector can contain anything, and
   // doesn't even have to be the same length as an uncompressed save state,
   // but a state needs to be loaded with the same basis as it was saved.
-  static void SaveEx(vector<uint8> *out, const vector<uint8> &basis);
-  static void LoadEx(vector<uint8> *in, const vector<uint8> &basis);
+  // basis can be NULL, and then these behave the same as Save/Load.
+  static void SaveEx(vector<uint8> *out, const vector<uint8> *basis);
+  static void LoadEx(vector<uint8> *in, const vector<uint8> *basis);
 };
 
 

### 2012-11-17 tasbot/fceu/state.cpp

@@ -343,7 +343,69 @@ static bool ReadStateChunks(EMUFILE* is, int32 totalsize)
 int CurrentState=0;
 extern int geniestage;
 
+// Simplified save that does not compress.
+bool FCEUSS_SaveRAW(std::vector<uint8> *out) {
+  EMUFILE_MEMORY os(out);
 
+  uint32 totalsize = 0;
+
+  FCEUPPU_SaveState();
+  FCEUSND_SaveState();
+  totalsize = WriteStateChunk(&os,1,SFCPU);
+  totalsize += WriteStateChunk(&os,2,SFCPUC);
+  totalsize += WriteStateChunk(&os,3,FCEUPPU_STATEINFO);
+  // PERF: Do we need to save both old and new ppu infos?
+  totalsize += WriteStateChunk(&os,31,FCEU_NEWPPU_STATEINFO);
+  totalsize += WriteStateChunk(&os,4,FCEUCTRL_STATEINFO);
+  totalsize += WriteStateChunk(&os,5,FCEUSND_STATEINFO);
+
+  if(SPreSave) SPreSave();
+  // This allows other parts of the system to hook into things to be
+  // saved. It is indeed used for "WRAM", "LATC", "BUSC". -tom7
+  totalsize+=WriteStateChunk(&os,0x10,SFMDATA);
+  if(SPreSave) SPostSave();
+
+  // save the length of the file
+  int len = os.size();
+
+  // PERF trim?
+  
+  // sanity check: len and totalsize should be the same
+  if (len != totalsize) {
+    FCEUD_PrintError("sanity violation: len != totalsize");
+    return false;
+  }
+
+  return true;
+}
+
+bool FCEUSS_LoadRAW(std::vector<uint8> *in) {
+  EMUFILE_MEMORY is(in);
+
+  int totalsize = is.size();
+  // Assume current version; memory only.
+  int stateversion = FCEU_VERSION_NUMERIC;
+
+  // is->fread(memory_savestate.buf(), totalsize);
+
+  FCEUMOV_PreLoad();
+
+  bool success = (ReadStateChunks(&is, totalsize) != 0);
+
+  if(GameStateRestore) {
+    GameStateRestore(stateversion);
+  }
+
+  if (success) {
+    FCEUPPU_LoadState(stateversion);
+    FCEUSND_LoadState(stateversion);
+    return FCEUMOV_PostLoad();
+  } else {
+    return false;
+  }
+}
+
[…]

### 2012-11-17 trunk/tasbot/fceu/state.h
Now able to write basis savestate to make input to compression likely to have less entropy

@@ -54,6 +54,10 @@ struct SFORMAT
         char *desc;
 };
 
+// Tom 7's simplified versions. These should only be used for in-memory saves!
+bool FCEUSS_SaveRAW(std::vector<uint8> *out);
+bool FCEUSS_LoadRAW(std::vector<uint8> *in);
+
 void ResetExState(void (*PreSave)(void),void (*PostSave)(void));
 void AddExState(void *v, uint32 s, int type, char *desc);
 

### 2012-11-17 trunk/cc-lib/util.h
Utilities for TASbot

@@ -34,6 +34,9 @@ struct Util {
   // carriage returns, including ones not followed by newline.
   static vector<string> ReadFileToLines(const string &f);
 
+  static vector<unsigned char> ReadFileBytes(const string &f);
+  static bool WriteFileBytes(const string &f, const vector<unsigned char> &b);
+
   // XXX terrible names
   static int shout(int, string, unsigned int &);
   static string shint(int b, int i);

### 2012-11-17 cc-lib/util.cc
Utilities for TASbot

@@ -190,6 +190,18 @@ vector<string> Util::ReadFileToLines(const string &f) {
   return v;
 }
 
+// PERF memcpy
+vector<unsigned char> Util::ReadFileBytes(const string &f) {
+  string s = ReadFile(f);
+  vector<unsigned char> bytes;
+  bytes.reserve(s.size());
+  for (int i = 0; i < s.size(); i++) {
+    bytes.push_back((unsigned char)s[i]);
+  }
+  return bytes;
+}
+
+
 static bool hasmagicf(FILE * f, const string & mag) {
   char * hdr = (char*)malloc(mag.length());
   if (!hdr) return false;
@@ -264,6 +276,18 @@ bool Util::WriteFile(const string &fn, const string &s) {
 
   /* XXX check failure */
   fwrite(s.c_str(), 1, s.length(), f);
+
+  fclose(f);
+  
+  return true;
+}
+
+bool Util::WriteFileBytes(const string &fn, const vector<unsigned char> &bytes) {
+  FILE * f = fopen(fn.c_str(), "wb");
+  if (!f) return false;
+
+  /* XXX check failure */
+  fwrite(&bytes[0], 1, bytes.size(), f);
 
   fclose(f);
   

### 2012-11-18 trunk/cc-lib/heap.h
Implemented heaps by porting code from sml-lib

@@ -1,0 +1,237 @@
+/* Reimplementation of mutable heaps from scratch, based on
+   my sml implementation.
+*/
+
+#ifndef __CCLIB_HEAP_H
+#define __CCLIB_HEAP_H
+
+#include <vector>
+
+struct Heapable {
+  /* The Heap uses this to store the index of this element in the heap,
+     which allows you to update the Heapable value and then tell the
+     heap to fix the heap invariants.
+
+     You usually shouldn't read or write it.
+
+     If this becomes -1, then the item has been deleted. */
+  int location;
+};
+
+template<class Priority /* has comparison operators, value semantics */,
+         class Value /* :> Heapable */>
+class Heap {
+ public:
+  struct Cell {
+    Priority priority;
+    Value *value;
+  };
+
+  bool Valid(Value *v) { return v->location != -1; }
+
+  // With empty cells.
+  Heap() {}
+
+  void Insert(Priority p, Value *v) {
+    Cell c;
+    c.priority = p;
+    c.value = v;
+    // a la SetElem.
+    cells.push_back(c);
+    c.value->location = cells.size() - 1;
+
+    // No children, so invariant violations are always upward.
+    PercolateUp(cells.size() - 1);
+  }
+
+  void Delete(Value *v) {
+    int i = v->location;
+    if (i == -1) {
+      fprintf(stderr, "Tried to delete value more than once.\n");
+      abort();
+    }
+
+    // Invalidate v because it's being deleted.
+    v->location = -1;
+
+    // Need old priority to know which direction to percolate.
+    Priority pold = cells[i].priority;
+
+    /* if a handle is valid, then heap size > 0 */
+    Cell replacement = RemoveLast();
+    if (replacement.value == v) {
+      // Deleting the last element, so RemoveLast has taken
+      // care of everything for us.
+      return;
+    }
+
+    // Write the replacement over the deleted element.
+    SetElem(i, replacement);
+
+    /* might have violated heap invariant */
+    if (replacement.priority < pold) {
+      PercolateUp(i);
+    } else if (pold < replacement.priority) {
+      PercolateDown(i);
+    }
+  }
+
+  bool Empty() const {
+    return cells.empty();
+  }
+
+  // May not be empty.
+  Cell PopMinimum() {
+    if (cells.empty()) {
+      fprintf(stderr, "Can't PopMinimum on an empty heap.\n");
+      abort();
+    }
+
[…]

### 2012-11-18 cc-lib/heap_test.cc
Implemented heaps by porting code from sml-lib

@@ -1,0 +1,55 @@
+
+#include <stdio.h>
+#include <stdlib.h>
+#include <string>
+
+#include "heap.h"
+
+using namespace std;
+
+typedef unsigned long long uint64;
+
+struct TestValue : public Heapable {
+  TestValue(uint64 i) : i(i) {}
+  uint64 i;
+};
+
+static uint64 CrapHash(int a) {
+  uint64 ret = ~a;
+  ret *= 31337;
+  ret ^= 0xDEADBEEF;
+  ret = (ret >> 17) | (ret << (64 - 17));
+  ret -= 911911911911;
+  ret *= 65537;
+  ret ^= 0xCAFEBABE;
+  return ret;
+}
+
+int main () {
+  static const int kNumValues = 1000;
+  
+  Heap<uint64, TestValue> heap;
+  
+  vector<TestValue> values;
+  for (int i = 0; i < kNumValues; i++) {
+    values.push_back(TestValue(CrapHash(i)));
+  }
+
+  for (int i = 0; i < values.size(); i++) {
+    heap.Insert(values[i].i, &values[i]);
+  }
+
+  TestValue *last = heap.PopMinimumValue();
+  while (!heap.Empty()) {
+    TestValue *now = heap.PopMinimumValue();
+    fprintf(stderr, "%llu %llu\n", last->i, now->i);
+    if (now->i < last->i) {
+      printf("FAIL: %llu %llu\n", last->i, now->i);
+      return -1;
+    }
+    last = now;
+  }
+
+  printf("OK\n");
+  return 0;
+}

### 2012-11-18 trunk/tasbot/emu_test.cc
Cleanup, add heap

@@ -6,20 +6,10 @@
 
 #include <unistd.h>
 #include <sys/types.h>
-#include <signal.h>
-#include <sys/time.h>
-#include <sys/stat.h>
 #include <string.h>
-#include <strings.h>
 #include <errno.h>
 #include <stdio.h>
 #include <stdlib.h>
-#include <iostream>
-#include <fstream>
-#include <limits.h>
-#include <math.h>
-
-#include <zlib.h>
 
 #include "fceu/utils/md5.h"
 
@@ -30,52 +20,14 @@
 
 #include "fceu/state.h"
 
-#include "fceu/drivers/common/cheat.h"
 #include "fceu/fceu.h"
-#include "fceu/movie.h"
-#include "fceu/version.h"
-
-#include "fceu/oldmovie.h"
 #include "fceu/types.h"
 
 #include "../cc-lib/util.h"
 
 #include "simplefm2.h"
 #include "emulator.h"
-
-static void DoFun(int frameskip) {
-  uint8 *gfx;
-  int32 *sound;
-  int32 ssize;
-  static int opause = 0;
-
-  // fprintf(stderr, "In DoFun..\n");
-
-  // Limited ability to skip video and sound.
-  #define SKIP_VIDEO_AND_SOUND 2
-
-  // Emulate a single frame.
-  FCEUI_Emulate(NULL, &sound, &ssize, SKIP_VIDEO_AND_SOUND);
-
-  // This was the only useful thing from Update. It's called multiple
-  // times; I don't know why.
-  // --- update input! --
-
-  uint8 v = RAM[0x0009];
-  uint8 s = RAM[0x000B];  // Should be 77.
-  uint32 loc = (RAM[0x0080] << 24) |
-    (RAM[0x0081] << 16) |
-    (RAM[0x0082] << 8) |
-    (RAM[0x0083]);
-  // fprintf(stderr, "%02x %02x\n", v, s);
-}
-
-/**
- * Update the video, audio, and input subsystems with the provided
- * video (XBuf) and audio (Buffer) information.
- */
-void FCEUD_Update(uint8 *XBuf, int32 *Buffer, int Count) {
-}
+#include "basis-util.h"
 
 static int64 DumpMem() {
   for (int i = 0; i < 0x800; i++) {
@@ -146,34 +98,6 @@ static bool CompareByHash(int a, int b) {
   return CrapHash(a) < CrapHash(b);
 }
 
-static vector<uint8> LoadOrComputeBasis(const vector<uint8> &inputs,
[…]

### 2012-11-18 trunk/tasbot/headless-driver.cc

@@ -191,6 +191,13 @@ void FCEUD_TurboToggle() { }
 void FCEUD_DebugBreakpoint(int bp_num) {}
 void FCEUD_TraceInstruction(unsigned char*, int) {}
 
+
+/**
+ * Update the video, audio, and input subsystems with the provided
+ * video (XBuf) and audio (Buffer) information.
+ */
+void FCEUD_Update(uint8 *XBuf, int32 *Buffer, int Count) {}
+
 /**
  * Get the time in ticks.
  */

### 2012-11-18 trunk/tasbot/tasbot.cc

@@ -1,0 +1,89 @@
+/* Searches for solutions to Karate Kid. */
+
+#include <unistd.h>
+#include <sys/types.h>
+#include <string.h>
+#include <errno.h>
+#include <stdio.h>
+#include <stdlib.h>
+
+#include "fceu/utils/md5.h"
+
+#include "config.h"
+
+#include "fceu/driver.h"
+#include "fceu/drivers/common/args.h"
+
+#include "fceu/state.h"
+
+#include "fceu/fceu.h"
+#include "fceu/types.h"
+
+#include "../cc-lib/util.h"
+#include "../cc-lib/heap.h"
+
+#include "simplefm2.h"
+#include "emulator.h"
+#include "basis-util.h"
+
+
+
+
+static void DoFun(int frameskip) {
+  uint8 *gfx;
+  int32 *sound;
+  int32 ssize;
+  static int opause = 0;
+
+  // fprintf(stderr, "In DoFun..\n");
+
+  // Limited ability to skip video and sound.
+  #define SKIP_VIDEO_AND_SOUND 2
+
+  // Emulate a single frame.
+  FCEUI_Emulate(NULL, &sound, &ssize, SKIP_VIDEO_AND_SOUND);
+
+  // This was the only useful thing from Update. It's called multiple
+  // times; I don't know why.
+  // --- update input! --
+
+  uint8 v = RAM[0x0009];
+  uint8 s = RAM[0x000B];  // Should be 77.
+  uint32 loc = (RAM[0x0080] << 24) |
+    (RAM[0x0081] << 16) |
+    (RAM[0x0082] << 8) |
+    (RAM[0x0083]);
+  // fprintf(stderr, "%02x %02x\n", v, s);
+}
+
+/**
+ * The main loop for the SDL.
+ */
+int main(int argc, char *argv[]) {
+  Emulator::Initialize("karate.nes");
+
+  // The inputs are only needed for computing the basis.
+  vector<uint8> inputs = SimpleFM2::ReadInputs("karate.fm2");
+  vector<uint8> basis = BasisUtil::LoadOrComputeBasis(inputs, 4935, "karate.basis");
+  inputs.clear();
+
+  fprintf(stderr, "Running %d steps...\n", inputs.size());
+  for (int i = 0; i < inputs.size(); i++) {
+    // XXX don't think this should ever happen.
+    if (!GameInfo) {
[…]

### 2013-03-08 trunk/tasbot/paper/paper.tex
Sync paper

\documentclass[twocolumn]{article}
\usepackage[top=1.1in, left=0.85in, right=0.85in]{geometry}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{code}
\usepackage{graphicx}

\pagestyle{empty}

% XXX superscript
\newcommand\th{th}

% \usepackage{ulem}
% go back to italics for emphasis, though
% \normalem

\begin{document}

\title{The First Level of Super Mario Bros.~is Easy with Lexicographic Orderings and Time Travel
       {\normalsize \ldots after that it gets a little tricky.}}
\author{Dr.~Tom~Murphy~VII~Ph.D.\thanks{
Copyright \copyright\ 2013 the Regents of the Wikiplia
Foundation. Appears in SIGBOVIK 2013 with the reluctant sigh of the
Association for Computational Heresy; {\em IEEEEEE!} press,
Verlag-Verlag volume no.~0x40-2A.
CHF 0.00}
}

\renewcommand\>{$>$}
\newcommand\<{$<$}

\date{1 April 2013}

\maketitle

\begin{abstract}
This is where the Abstract goes.
\end{abstract}

\vspace{1em}
{\noindent \small {\bf Keywords}:
  computational super mario brothers, memory inspection, lexicographic induction, networked entertainment systems, pit-jumping, ...

}

\section*{Introduction}
The Nintendo Entertainment System is probably the best video game console, citation {\it not} needed. Like many, I have spent thousands of hours of my life playing NES games, including several complete playthroughs of classics like Super Mario Bros., Bionic Commando, Bubble Bobble, and other favorites. By the year 2013, home computers have become many orders of magnitude faster and more capacious than the NES hardware. This suggested to me that it may be time to automate the playing of NES games, in order to save time.\footnote{Rather, to replace it with time spent programming.} In this paper I present a generic technique for automating the playing of NES games. The approach is practical on a single computer, and succeeds on some games, such as Super Mario Bros.. The approach is amusingly elegant and surprisingly effective, requires no detailed knowledge of the game being played, and is capable of novel and impressive gameplay (for example, bug exploitation). {\bf Disclaimer for SIGBOVIK audience: This work is 100\% real.}

A short video version of this paper is available for those that hate reading, at \verb+http://tom7.org/mario+. This page also contains audiovisual material that makes this work more entertaining (for example, its output) and source code.

The basic idea is to deduce an objective function from a short recording of a player's inputs to the game. The objective function is then used to guide search over possible inputs, using an emulator. This allows the player's notion of progress to be generalized in order to produce novel gameplay. A design goal is that the objective function be amusingly elegant (not at all smart, fancy, or customized to the game) in order to demonstrate that the game is reducible to such a simple objective. The search needs to be game-agnostic and practical, but since the space is exponential ($256^{n}$), we need to be smart here.

The objective function, the algorithm to deduce it, the search strategy, and its implementation are all interesting and will be discussed in that order. To set the stage, I begin with a description of the NES hardware and emulation of it.

\subsection{The NES hardware and emulation}
The NES is based around an 8-bit processor running at 1.79~MHz, the Ricoh 2A03. 8 bits is really small. You can see them all right here: 00001111. It's no coincidence that each controller also has 8 buttons: Up, Down, Left, Right, Select, Start, B and A. It has only 2048 bytes of general purpose RAM. (There is also some special purpose RAM for graphics, which we ignore in this work.) 2048 bytes is really small. You can see them all in Figure~\ref{fig:bytes2048}. As a result, NES programs are written to use memory efficiently and straightforwardly; usually there are fixed memory locations used for all the critical game facts like the player's health, number of lives, coordinates on the screen, and so on. For example, in Super Mario Bros., the single byte at location \verb+0x757+ contains the number of lives the player has. The location \verb+0x75F+ contains the current world, and \verb+0x760+ the current level.

% (32kB of ROM is addressable at a time, and most games used bank switching to access hundreds of kilobytes of game included in the cartridge.)

There are a number of emulators for NES. These work by simulating the NES hardware, for example with a 2048-byte array for its memory, and simulating the steps of its 2A03 processor on some ROM, and hooking a keyboard or joystick into the 8 bits of input. (There are of course many details to work out! But in essence emulation is jut that.) This process is completely deterministic, so it is possible to record the sequence of inputs (the inputs can only be read once per video frame, so this sequence is 60 \comment{ XXX CHECK } bytes per second) and play them back and get the same result. This also means that an input sequence can be computed in non-real time, either much slower or much faster than a NES would normally run. In this work we use the FCEUX emulator, which is popular for its accuracy and advanced tools.

\begin{figure}
\begin{center}
% XXX make image
% \includegraphics[width=0.75 \linewidth]{bytes2048}
\end{center}\vspace{-0.1in}
\caption{2048 bytes, a 64x32 image.}
\label{fig:bytes2048}
\end{figure}

\section{Objective function}

Bytes in memory (and sometimes 16- and 32-bit words) can contain interesting game facts like the player's position in the level or score. The central idea of this paper is to use the value of memory locations to deduce when the player is ``winning''. The things that a human player perceives, like the video screen and sound effects, are completely ignored. As an additional simplification, we assume that winning always consists of a value {\it going up}---either the position in the level getting larger, the score getting larger, the number of lives, the world or level number getting bigger, etc.

This is actually a little bit too naive; for example, Mario's overall progress through the game is represented by a pair. You start in World 1-1 and the underground level that comes next is World 1-2 (we'll call this $w=1$ and $\ell=2$). But after you discover the princess is in another castle in World 1-4, the next level is 2-1.\footnote{In case you never realized this, the legendary ``Minus World'' of \verb+ -1+ is not actually a negative world, but World 36-1 being incorrectly rendered because there is no glyph for the 36\th\ digit. The trick used to get to the Minus World just happens to leave the value 36 in that slot rather than initializing it to a useful value. The ROM does not contain data for world 36 so it just interprets garbage data as a description of the world.} This can't be represented as a single byte going up (sometimes the second part $\ell$ goes down when we get to a new first part $w$), but it can be represented as a lexicographic order on the pair $\langle w, \ell \rangle$; that is, $\langle w_1, \ell_1 \rangle < \langle \w_2, \ell_2$ if $\w_1 = \w_2$ and $\ell_1 < \ell_2$, or if $\w_1 < \w_2$ no matter the values of $\ell_1$ and $\ell_2$. This matches our intuitive idea and is also mathematically nice. It also generalizes multi-byte encodings of things like your score (which can be larger than 8 bits and so is often stored in 16 or 32), including both big-endian and little-endian representations.\footnote{A possible additional simplification would be to just take lexicographic orderings over bits, which then generalizes to 8-bit bytes. This is probably too crazy, but just right now I am sort of feeling like maybe I should try it, though it may be the beer.}

More importantly, it allows the combination of semantically unrelated bytes, like: $\langle$ world, level, screen inside the world, $x$ position on the screen $\rangle$ or $\langle$ world, lives, low byte of score $\rangle$. Many orderings may describe gameplay. These orderings may be temporarily violated in normal play: Although the score always goes up, Mario's $x$ position may temporarily decrease if he needs to navigate around some obstacle.\footnote{Note to self: Maybe we should give a much higher score to globally preserved objectives than to locally preserved ones. But that may presuppose that the input represents a whole playthrough?} So, to ``faithfully'' represent gameplay, we will generate a set of lexicographic orderings on memory locations, with the idea that they ``generally go up'' but not necessarily at every step. These orderings will also have weights. The next section describes how we take a sequence of player inputs and deduce the orderings.

\subsection{Deriving the objective function}

\end{document}

### 2013-03-10 trunk/tasbot/paper/paper.tex
Slow progress on paper...

@@
 Bytes in memory (and sometimes 16- and 32-bit words) can contain interesting game facts like the player's position in the level or score.
-The central idea of this paper is to use the value of memory locations to deduce when the player is ``winning''.
+The central idea of this paper is to use (only) the value of memory locations to deduce when the player is ``winning''.
 The things that a human player perceives, like the video screen and sound effects, are completely ignored.
@@
 You start in World 1-1 and the underground level that comes next is World 1-2 (we'll call this $w=1$ and $\ell=2$).
-But after you discover the princess is in another castle in World 1-4, the next level is 2-1.\footnote{In case you never realized this, the legendary ``Minus World'' of \verb+ -1+ is not actually a negative world, but World 36-1 being incorrectly rendered because there is no glyph for the 36\th\ digit.
-The trick used to get to the Minus World just happens to leave the value 36 in that slot rather than initializing it to a useful value.
+But after you discover the princess is in another castle in World 1-4, the next level is 2-1.\footnote{In case you never realized this, it is time to learn that the legendary ``Minus World'' of \verb+ -1+ is not actually a negative world, but World 36-1 being incorrectly rendered because there is no glyph for the 36\th\ digit.
+The trick used to get to the Minus World just happens to leave the value 36 in that memory location rather than initializing it to a useful value.
 The ROM does not contain data for world 36 so it just interprets garbage data as a description of the world.} This can't be represented as a single byte going up (sometimes the second part $\ell$ goes down when we get to a new first part $w$), but it can be represented as a lexicographic order on the pair $\langle w, \ell \rangle$; that is, $\langle w_1, \ell_1 \rangle < \langle \w_2, \ell_2$ if $\w_1 = \w_2$ and $\ell_1 < \ell_2$, or if $\w_1 < \w_2$ no matter the values of $\ell_1$ and $\ell_2$.
@@
 
+In order to derive an objective function, we'll start with an abstract
+subroutine that finds a single lexicographic ordering
+nondeterministically. This function takes in an ordered list of $n$
+memories $M_1\ldots M_n$ which all have size $m$ bytes. For example,
+$m = 2048$ and $n = 100$, for the memories at each of the first $100$
+frames of someone playing Super Mario Bros.. It produces an ordered
+list of unique memory locations $L_1 \ldots L_k$ (where $0 \leq L_i <
+m$, that is, each is some spot in the 2048 bytes of RAM) that is a
+{\em maximal} {\em valid} lexicographic ordering of $M$. Let's start
+by defining those terms just to be careful.
+
+Given some list of memory locations $L_1 \ldots L_k$ and a pair of
+memories $M_a$ and $M_b$, we say that $M_a =_L M_b$ iff $M_a[L_1] =
+M_b[L_1]$ and $M_a[L_2] = M_b[L_2]$ and so on for every $L_i$; that
+is, the bytes must be equal at each of the locations. Easy. We say
+that $M_a <_L M_b$ iff there exists some $p \leq k$ where $M_a[L_1] =
+M_b[L_1] \ldots M_a[L_{p-1}] = M_b[L_{p-1}]$ and $M_a[L_p] <
+M_b[L_p]$. Put simply, if the two memories are not equal according to
+$L$ (have the same byte at every memory location) then there is a
+unique first location ($L_p$) where they have a different byte, and
+that byte determines the order. $M_a >_L M_b$ is just defined as $M_b <_L
+M_a$; $M_a \leq_L M_b$ is just $M_a <_L M_b$ or $M_a =_L M_b$, and similarly
+for $\geq_L$, and they mean what you think so don't worry.
+
+Every $L$ defines a lexicographic ordering ($<$ and $=$ operators).
+$L$ is a {\em valid} lexicographic ordering of $M$ if $M_i \leq_L M_{i + 1}$
+for $1 \leq i \leq n$; each memory is less than or equal to the next
+memory in the sequence. It follows that $M_i \leq_L M_j$ whenever $i < j$.
+
+Every prefix of a valid $L$ (including the empty prefix) is a valid
+lexicographic ordering as well. On a scale from useless to useful, the
+empty prefix is a 1 (it equates all memories), which suggests that
+some orderings are better than other. To give a primitive notion of
+``good'' lexicographic orderings, we define a {\em maximal} valid
+lexicographic ordering to be $L$ such that there are no extensions of
+$L$ that are valid. An extension of $L_1 \ldots L_k$ is just $L_1
+\ldots L_k, L_{k+1}$ (where $L_{k+1} \neq L_i$ for $1 \leq i \leq k$):
+Some new memory location that we put at the end of the order (in the
+least important position). We do not consider extending in the middle
+of the order or beginning, although that would make sense.
+
+On a scale from simple to fancy, this algorithm is a 3. Given those
+definitions, the idea is to start with the empty prefix, which is
+always a valid lexicographic ordering but usually not maximal. We then
+pick an extension that is valid; if none exists then we are done and
+have a maximal valid ordering. Now we have a new valid ordering, and
+we repeat.
+
+This works but has the bummer downside that it produces
+
+The function works like this:
+% XXX learnfun ...
+
+\subsection{Playfun}
+
+On a scale from ...
+
 \end{document}

### 2013-03-11 trunk/tasbot/paper/paper.tex
Slow progress on paper; sync ZRH...

@@
 m$, that is, each is some spot in the 2048 bytes of RAM) that is a
-{\em maximal} {\em valid} lexicographic ordering of $M$. Let's start
-by defining those terms just to be careful.
+{\em maximal} {\em tight} {\em valid} lexicographic ordering of $M$.
+Let's start by defining those terms just to be careful.
 
@@
 
+Maximal valid orderings are good and it is straightforward to produce
+them (a less constrained version of the algorithm below), but they
+have the bummer downside that memory locations that never change value
+for any $M$ can be included at any position of the ordering, and all
+such choices are equivalent. And in fact all locations {\em must} be
+included to make the ordering {\em maximal}. This is bad because when
+$M$ contains many locations with fixed values, we have boatloads of
+equivalent orderings, and they're also longer than necessary. An {\em
+  tight} valid ordering is one where each $L_i$ .. XXX HERE.
+
 On a scale from simple to fancy, this algorithm is a 3. Given those
@@
 we repeat.
-
-This works but has the bummer downside that it produces
 

### 2013-03-12 trunk/tasbot/paper/paper.tex
Progress on paper from plane

@@
 
+\newcommand[1]{comment}{}
+\newcommand[1]{parameteralert}{
+  % XXX this can get cooler.
+  {\bf Parameter Alert!}\,\, #1
+}
+
 % XXX superscript
@@
 
-\title{The First Level of Super Mario Bros.~is Easy with Lexicographic Orderings and Time Travel
+\title{The First Level of Super Mario Bros.~is Easy with
+       Lexicographic Orderings and Time Travel
        {\normalsize \ldots after that it gets a little tricky.}}
@@
 
-A short video version of this paper is available for those that hate reading, at \verb+http://tom7.org/mario+.
+On a scale from ``the title starts with Toward'' to ``Donald Knuth has finally finished the 8\th\ volume on the subject'', this work is a 3.
+The purpose of this paper is mainly as a careful record of the current status for repeatability and further development on this important research subject.
+A short video version of this paper is available for those that hate reading, at \verb+http://tom7.org/mario+, and is the more fun way to consume the results.
 This page also contains audiovisual material that makes this work more entertaining (for example, its output) and source code.
@@
 
-\subsection{Deriving the objective function}
+\subsection{Deriving the objective function} \label{sec:deriving}
 
@@
 equivalent orderings, and they're also longer than necessary. An {\em
-  tight} valid ordering is one where each $L_i$ .. XXX HERE.
+  tight} valid ordering is one where for each $L_i$ there exists at
+least one $M_a$ and $M_{a+1}$ where $M_a[L_i] < M_{a+1}[L_i]$ and
+$M_a[L_j] = M_{a+1}[L_j]$ for all $i < j$; that is, every location has
+to participate in the ordering at least once. The notion of {\em
+  maximal} has to be relative to this property as well---a {\em tight
+  extension} is one that produces a tight valid ordering, and a {\em
+  maximal tight valid ordering} permits no tight extensions.
 
@@
 definitions, the idea is to start with the empty prefix, which is
-always a valid lexicographic ordering but usually not maximal. We then
-pick an extension that is valid; if none exists then we are done and
-have a maximal valid ordering. Now we have a new valid ordering, and
-we repeat.
+always a tight valid lexicographic ordering but usually not maximal.
+We then pick a tight extension that is valid; if none exists then we are
+done and have a maximal tight valid ordering. Now we have a new tight valid
+ordering, and we repeat.
 
-The function works like this:
-% XXX learnfun ...
+The pseudocode gives a pseudoimplementation of the algorithm that is
+more pseudodetailed. The C++ implementation is in \verb+objective.*+.
+C++ is not a good language for this kind of program but we use it
+because the NES emulator is written in C++.
 
-\subsection{Playfun}
+\begin{figure}[htp]
+  \begin{code}
+    (* Prefix is the prefix so far (int list)
+       and remain is the list of memory locations
+       that we can still consider. Invt is that
+       prefix is a tight valid ordering on M. Returns
+       the memory locations from remain that are
+       a tight extension of the prefix. *)
+    fun candidates (prefix, remain) =
+      let lequal = (* list of indices $i$ where
+                      $M_i =_{\tt prefix} M_{i+1}$ *)
+      let notgreater = (* members $x$ of remain where
+                          $M_i[x] > M_{i + 1}[x]$ is
+                          not true for any $i$ in
+                          {\tt lequal} *)
+      let tight = (* members $y$ of notgreater where
+                     $M_i[x] < M_{i + 1}[x]$ is true
+                     for some $i$ in {\tt lequal} *)
+      in tight
 
-On a scale from ...
+    (* Returns a maximal tight valid ordering, given
+       a tight valid prefix and list of memory locations
+       we are permitted to consider. *)
+    fun ordering (prefix, remain) =
+      case candidates (prefix, remain) of
+        (* No extensions means it's maximal. *)
+        nil => prefix
+      | cand =>
+        let c = nondeterministically-choose-one cand
+        let remain' = remove-element (remain, c)
+        let prefix' = prefix @ [c]
+        in ordering (prefix', remain')
+  \end{code}
+
+  \caption{Pseudocodes for nondeterministically generating a maximal
+    tight valid lexicographic ordering on some memories $M$. The
+    recursive function {\tt ordering} just orchestrates the selection
+    of an extension until there are no possibilities remaining, and
+    returns it. The function {\tt candidates} finds all the possible
+    extensions for a prefix. First we compute {\tt lequal}, all of the
+    adjacent memory pairs (represented by the index of the first in
+    the pair) where the memories are equal on the prefix. Only pairs
+    that are equal on the prefix are interesting because these are the
+    only ones that will even depend on the extension to the prefix
+    when comparing the memories. We only need to consider adjacent
+    pairs because on an a scale of exercise for the reader to proof is
+    contained in the extended technical report, this statement is a
+    you can figure that one out yourself. Valid extension locations
+    are ones where the memory pairs are never increasing at that
+    location (note that in places where pairs are not equal but
+    strictly less on the prefix, it's perfectly fine for the extension
+    to be greater; this is the ``point'' of lexicographic orderings).
+    Finally, tight extensions are those valid ones that have at least
+    one pair of memories where the location has a value that is
+    strictly less.}
+  \label{fig:ordering}
+\end{figure}
+
+\subsection{The objective function, in practice}
+
+We can't just use a single objective function. Choosing objective
+functions nondeterministically, we may get a crap one like ``High byte
+of the score'' which only goes up once during all of our memory
+observations. We also can't just use all of the memory observations,
+because there may be brief moments that violate strict orderings, like
+Mario's $x$ coordinate temporarily decreasing to navigate around an
+obstacle. More starkly, the first few hundred frames of the game are
+almost always initialization where the memory temporarily takes on
+values that are not representative of later gameplay at all. In
+practice, we use the nondeterministic function from
+Section~\ref{sec:deriving} on multiple different slices of the memory
+observations. We also call it many times to nondeterministically
+generate many different objectives. Finally, we weight the objectives
+by how representative we think they are.
+
+\parameteralert{This one of the first places where we have some
+  arbitrary constants, which are the enemy of elegance. On a scale of
+  ballpark to obsessively overfit, these constants are a 2; I
+  basically looked at some graphs while developing the objective
+  function learning part of the code to decide whether they were
+  ``good enough'' and didn't tune them after starting to observe
+  actual performance. Some of those graphs appear in figures here.
+  For all I know, these are really bad choices, but it was important
+  for aesthetic reasons for the objective function to be brutish. The
+  only reason to permit these parameters at all is that it simply
+  does not work to have a single ordering or to use only orderings
+  that apply to the whole memory.}
+
+{\bf Skipping.}\,\, To avoid being confused by RAM initialization and
+menu, I ignore all memories up until the first input by the player.
+Including these would be especially suspicious because the RAM's
+contents are not even well-defined until the program writes something
+to them.\footnote{Emulators tend to fix them to a specific pattern so
+  that emulation is completely deterministic, but in real hardware
+  they are truly uninitialized, or retain the values from the last
+  reset or game inserted. Some semi-famous tricks involve removing
+  and inserting game cartridges while the system is running in order
+  to take advantage of this.}
+
+% XXXX good place for some graphs.
+{\bf Slicing.}\,\, I generate 50 orderings for $M_1 \ldots M_n$; the
+whole recording starting from the first keypress. During gameplay some
+values really are completely nondecreasing, like the player's score
+and world/level pair. I also generate 3 orderings for each tenth of
+the memory sequence, e.g. $M_1 \ldots M_{n/10}$ and $M_{n/10+1} \ldots
+M_{2n/10}$, etc. The intention is to capture orderings that are rarely
+violated, or sections of the game with unusual objectives (e.g. a
+minigame or swimming level). Orderings generated this way look pretty
+random, and on a scale from solid to suspicious, I can't vouch for
+them. Then I generate objectives from non-consecutive memories that
+are evenly spread out through the observations: Ten objectives
+chosen from every 100\th\ memory, starting from the 0\th\ frame,
+1\st\ frame, 2\nd\ frame, and so on up to the 9\th{}. Similarly
+for every 250\th\ frame, and a single objective for memory sliced
+to every 1000\th\ frame, with start positions of 0--9.
+
+{\bf Weighting.},\, To reduce the importance of randomness in the
+learning process, and the arbitrariness of the slicing, each objective
+function is also assigned a weight. An ideal objective function takes
+on its minimal value on the first frame and maximal on the last
+(according to the ordering), and increases steadily throughout the
+observed memories. This is ideal because it allows us to just follow
+the gradient to reach good states. A bad objective function freqently
+regresses (recall that although we produce valid orderings, an
+ordering that is valid on some slice may not be valid for the whole
+sequence of memories). To weight an objective $L$, we first collect
+all of the values (the vector of values of the memory locations $L_1
+\ldots L_k$) it takes on throughout the observations. These may not
+obey the ordering. We then sort the values lexicographically and
+remove duplicates.\footnote{Note to self: I'm not sure how to justify
+  removing duplicates here. It makes [0, 1, 1, 1, 1, 1, 10, 11] look
+  the same as [0, 1, 10, 11], which is probably not desirable?} Call
+this $V$. Now we can compute the {\em value fraction} for $L$ on some
+$M$: Extract the vector of locations $M[L_1], M[L_2], \ldots, M[L_k]$
+and find the lowest index $i$ in $V$ where the vector is less than or
+equal to $V_i$. The value fraction {\tt VF} is $i/|V|$, which is the
+normalized value of ``how big'' $M$ is, according to $L$, compared to
+all the values we ever saw for it. The value fraction is defined and
+in $[0, 1]$ for all memories in the observed set.\footnote{It is not
+  defined when the memory is greater than all observed memories, which
+  we will encounter later. The code returns $(i+1)/|V|$ in that case,
+  which is as good as anything.} This gives us the ability to compare
+objectives on an absolute scale.\footnote{This is certainly not the
+  only way to do it, and it has some questionable properties like
+  ignoring the magnitude of change. But it is very simple.} Weighting
+an objective is now simple:
+%
+$$ \Sigma_{i=1}^{n-1}  {\tt VF}($M_{i+1}$) - {\tt VF}($M_{i}$) $$
+%
+We sum the differences in value functions for each consecutive pair of
+memories. In the case of the ideal function this is \Sigma_{i=1}^{n-1}
+$1/n$, which approaches $1$.\footnote{The code implements a different
+  weighting scheme, which sums from $i=0$ and keeps track of the last
+  value fraction at each step, thus treating the value fraction of the
+  nonexistent memory $M_0$ as 0. This is wrong, because the first
+  value fraction may be very high, which credits the objective with a
+  positive value (e.g. 1) for that first part of the sum. Objectives
+  that start high on the first frame are not ideal; in fact, the worst
+  objectives start high on the first frame and steadily {\em
+    decrease}. I plan to correct this bug and study if it has any
+  noticeable consequences.} Objectives are constrained to have
+non-negative weights (I set the value to 0 if negative, which
+effectively disables it).
+
+% XXX good place to give some weights for some objectives.
+
+We save the objectives and their weights to a file and then we are
+done with the easy part.
+
+\subsection{Motifs} \label{sec:motifs}
+
+The very first thing I tried with the objective function is to just do
+some greedy search for input sequences that increased the objective.
+This works terribly, because the search space for inputs is large
+($2^8$ possibilities at each frame). Most are useless (it's almost
+impossible to press the left and right directions at the same time,
+and real players almost never do it); real input sequences usually do
+not change values 60 times per second (rather, the player holds the
+jump button for 10--20 frames); some button-presses and combinations
+are much more common than others (e.g. right+B is common for running
+right, but start pauses the game). Search quickly necessitates a model
+for inputs. Rather than do anything custom, I just use a really dumb
+approach: Take the observed inputs (the same ones that we learned the
+objective functions from) and split them into chunks of 10 inputs.
+Motifs are weighted by the number of times they occur in the input.
+There may be a single motif at the end that's fewer than 10 inputs.
+
+\parameteralert{Here I choose the magic number 10 for the size of input
+  motifs. On a scale from gravitational constant to pulled it out of
+  my ass, this is an 8. We could perhaps justify 10 as being close to
+  the speed of input change actually possible for a human (6 button
+  presses per second; 166ms). I believe it is possible to do much
+  better here and the code contains a few such false starts, but using
+  motifs was one of the biggest immediate improvements in the history
+  of this code, so I went with it. A natural thing to try is a Markov
+  model, although this has a free parameter too (the number of states
+  of history). It is likely possible to do some kid of abstract
+  interpretation where multiple different input sequences with
+  equivalent outcomes are explored simultaneously, which might obviate
+  the need for computing an input model from the observed play. The
+  playfun algorithm below takes motifs as primitive because of the way
+  it was developed; I'll use footnotes to describe my thinking about
+  how to remove this.}
+
+Motifs are written to a file too and then we're done with that. This
+concludes the learning we do from the example input; everything else
+is a matter of using the objective functions and motifs to play the
+game.
+
+\section{Now you're playing with power}
+
+In this section I'll describe how the objective functions are used to
+play the game. On a scale from canonical to Star Wars Christmas
+Special, this algorithm is an 7. So, rather than focus on the
+particulars of some of the heuristics, I'll try to give a story of the
+different things I tried, what motivated the ideas in the current
+version, and my intuitions about what is working well (or not) and
+why. This algorithm is called {\tt playfun} and it can be found
+implemented in C++ in {\tt playfun.cc}.
+
+\subsection{Basic software architecture}
+
+In order to use the emulator to search for good sequences of inputs, I
+needed deeper integration than just observing memory. The FCEUX
+emulator is about a jillion lines of C++-ish code, was intended as an
+interactive GUI application, contains support for multiple different
+platforms, and the code is, on a scale from a pile of horse shit to
+not horse shit, approximately a 2.\footnote{It is, however, an
+  excellent emulator to use, has fancy tools for recording and editing
+  movies, and is popular in the speedrun community; just don't read
+  the code.} With a medium amount of suffering I got the emulator
+compiling under mingw in 64-bit mode, and working behind a streamlined
+interface ({\tt emulator.h}). Once a game is initialized, it is always
+at an input frame---you can give it an 8-bit input (for the
+1\st\ player) to step a single frame, or read the 2048 bytes of
+memory. You can also save the complete state of the emulator into a
+vector of bytes, which allows you to restore the state to exactly that
+same point.\footnote{This contains the RAM, but also stuff we didn't
+  consider, like the Picture Processing Unit's state, and internal
+  emulator stuff.} These save-states are portable across sessions as
+long as the code was compiled and the game initialized the right
+way.\footnote{The original FCEUX supports portable save-states, but I
+  removed that guarantee in order to get the smallest possible byte
+  vectors. More on that below.} FCEUX must be single-threaded because
+it uses global variables galore. I made several enhancements to the
+emulator interface, which are discussed later.
+
+It's important to know that almost all the CPU time in all the
+algorithms discussed in this paper is spent emulating NES frames;
+it takes about XXX ms to process a single step. Lots of engineering
+effort goes into reducing the number of frames the algorithms emulate.
+
+\subsection{Naive attempts}
+
+The very first thing I tried, as mentioned in
+Section~\ref{sec:motifs}, was to just look at all $2^8$ different
+possible inputs at each step, and pick the best one. The inner
+loop looks pseudolike this:
+
+\begin{code}
+  for (;;) {
+    vector<uint8> before = GetMemory();
+    vector<uint8> state = GetState();
+    // Try every bitmask of the 8 inputs.
+    for (int i = 0; i < 256; i++) {
+      RestoreState(state);
+      Step((uint8)i);
+      vector<uint8> after = GetMemory();
+      double score = Score(before, after);
+      // Save the best-scoring i...
+    }
+    RestoreState(state);
+    Step(bestinput);
+  }
+\end{code}
+
+{\tt Score} computes a score of two memories using the objective
+functions, which was the point of all that. There are a few
+canonical-seeming ways to implement this; the simplest is to
+count the (weighted) number of objective functions $o$ where
+${\tt before} <_o {\tt after}$. We'll get to more later.
+
+I wish this worked, because that would be truly laughable, but
+on a scale from doesn't to does it's a 1. At best, Mario twitches
+in place. The inputs that it plays are insane. There are lots of
+reasons, but a clear one is that a single input rarely affects
+your progress in the game on the very next frame. I didn't really
+expect this approach to work and I already knew that the state
+space is too big to search exhaustively, which is why I implemented
+motifs. This drastically reduces the search space and makes each
+step more significant; the inner loop can now be:
+
+\begin{code}
+for (const Motif &m : motifs) {
+  RestoreState(state);
+  for (uint8 i : m.inputs()) Step(i);
+  vector<uint8> after = GetMemory();
+  double score = Score(before, after);
+  // Save the best-scoring motif...
+}
+\end{code}
+
+This works much better (anything would), though not much better than
+you'd expect from just weighted random playback of the motifs
+themselves (they mostly contain motifs like ``hold right'' and ``hold
+right and A''). Mario is bad at avoiding danger except by luck, and
+bad at jumping hard enough to get over pipes (the button needs to be
+held consecutively for maybe 40--50 frames to jump that high).
+
+These two things---danger avoidance and microplanning to string
+together multiple motifs in a useful way---are two sides of the same
+coin. At least, on a scale from one side of the coin to the other, it
+is a two. My attempts to address these two problems converged on a
+single idea that is the crux of the good part of {\tt playfun}. First
+let's start with avoiding bad futures, since that is somewhat simpler.
+
+\subsection{Avoiding bad futures}
+
+Scoring a move based on how much better it is than the previous state
+causes Mario to make sensible greedy moves to increase the objective
+function---until he is then faced with no good options. This happens
+very frequently in Super Mario Bros.~(and many other games) because
+death is not usually a single-frame affair. For example, once he's
+near a Goomba with some velocity, it's too late to slow down or jump
+out of the way; he'll die in a few frames. Similarly, he can be in the
+midst of a too-short jump over a pit, where he's destined to die no
+matter how he squirms. Moreover, in Mario and many other games, even
+death as recognizable to the player isn't an obvious loss to these
+objective functions; the game's memory doesn't change much until the
+interstitial screen and Mario being reset back to the nearest
+checkpoint. So in order to avoid danger, Mario needs to avoid states
+that make death a foregone conclusion, not just avoid death.
+
+This is nothing special; move search in Chess and pretty much any game
+involves evaluating an ending game state and not just the quality of
+the move itself (``I captured a piece! Must be a great move!'').
+Evaluating a Chess state is a delicate matter, but Goombas and gravity
+are very stupid opponents. For avoiding danger, the following works
+well: Take the state and run a few seconds (300--500 frames) worth of
+weighted random motifs, several times. This gives us a set of states
+that could follow our candidate state were we to keep playing. We
+judge the candidate state not on its immediate value compared to our
+start state, but based on the random futures that may ensue. In my
+first version I used the minimum value of all these random futures, so
+that if Mario was getting into a state where he could die, those
+states would have low scores. Later we'll find that this isn't the
+right way to think about it, but it gives us a big improvement in the
+quality of play---Mario cleanly jumps over Goombas. He also gets very
+nervous and all like analysis-paralysis when faced with pits of medium
+size, which is related to the next section. % XXX video of this?
+
+\subsection{Seeking good futures}
+
+The flipside of avoiding danger is seeking adventure. Mario can avoid
+danger for quite a long time by just waiting at the beginning of the
+game, until time runs out. He can dilly dally before a pit,
+contemplating the void. But princesses need rescuin'. The same
+approach as before works for evaluating a state in terms of its
+potential: Try some random futures and look at where we end up. We
+could take the max over those scores if we're being optimistic, or the
+average or sum if we're trying to judge the general value of the
+state. In my first version, which did work pretty well, I took the
+max; so basically I had the min of some random states plus the max of
+some other states. But why generate a bunch of futures and take the
+min of some and the max of some others? On a scale of Thank You Mario
+Your Quest Is Over We Present You A New Quest Push Button B to I'm
+Sorry, but the Princess is in Another Similarly-Shaped but Not Exactly
+that Samesuch Castle, this is an 8.
+
+\subsection{Stable futures}
+
+Taking the minimum of random futures is always silly (at least if we
+believe our objective function), because nothing other than our own
+bad memory can force us to take a series of steps if we know a
+different better series of steps. Taking the max is possibly foolish
+if we don't know how to reconstruct a rare good state that caused the
+maximum score to be high. Both lead us to the idea: Keep around
+a set of candidate futures that we use for evaluation and search,
+rather than just randomly generating futures when we want to evaluate
+a state. This turns out to work really well and be more efficient.
+
+HERE
+
+\subsection{Caching of emulation}
 

### 2013-03-14 tasbot/game.h
Fix bug in learnfun

@@ -2,15 +2,22 @@
 #ifndef __GAME_H
 #define __GAME_H
 
+/*
 #define GAME "mario"
-#define MOVIE "mario-cleantom.fm2"
+// #define MOVIE "mario-cleantom.fm2"
+#define MOVIE "mario-newtom2.fm2"
 // #define FASTFORWARD 256  // XXX cheats! -- should be 0
 #define FASTFORWARD 0
 #define BASE64 "base64:jjYwGG411HcjG/j9UOVM3Q=="
+*/
+
+#define GAME "adventureisland"
+#define MOVIE "adventureisland-tom.fm2"
+#define BASE64 "base64:JIuhrL/tKL5OaDmoWKXGCg=="
+#define FASTFORWARD 0
 
 // #define GAME "karate"
 // #define MOVIE "karate-tom.fm2"
-//#define FASTFORWARD 300
 // #define BASE64 "base64:6xX0UBv8pLORyg1PCzbWcA=="
 
 // #define GAME "pacman"

### 2013-03-14 tasbot/weighted-objectives.cc

@@ -66,9 +66,12 @@ void WeightedObjectives::SaveToFile(const string &filename) const {
   string out;
   for (Weighted::const_iterator it = weighted.begin(); 
        it != weighted.end(); ++it) {
-    const vector<int> &obj = it->first;
-    out += StringPrintf("%f %s\n", it->second,
-			ObjectiveToString(obj).c_str());
+    if (it->second->weight > 0) {
+      const vector<int> &obj = it->first;
+      const Info &info = *it->second;
+      out += StringPrintf("%f %s\n", info.weight,
+			  ObjectiveToString(obj).c_str());
+    }
   }
   Util::WriteFile(filename, out);
   printf("Saved weighted objectives to %s\n", filename.c_str());
@@ -255,26 +258,40 @@ void WeightedObjectives::WeightByExamples(const vector< vector<uint8> >
 					  &memories) {
   for (Weighted::iterator it = weighted.begin();
        it != weighted.end(); ++it) {
-    double score = 0.0;
-
     const vector<int> &obj = it->first;
+    Info *info = it->second;
     // All the distinct values this objective takes on, in order.
     vector< vector<uint8> > values = GetUniqueValues(memories, obj);
 
+
+    // Sum of deltas is just very last - very first.
+    CHECK(memories.size() > 0);
+    double score_end =
+      GetValueFrac(values, GetValues(memories[memories.size() - 1], obj));
+    double score_begin = 
+      GetValueFrac(values, GetValues(memories[0], obj));
+    CHECK(score_end >= 0 && score_end <= 1);
+    CHECK(score_begin >= 0 && score_begin <= 1);
+    double score = score_end - score_begin;
+
+    /*
     int lastvaluefrac = 0.0;
     for (int i = 0; i < memories.size(); i++) {
       vector<uint8> now = GetValues(memories[i], obj);
       double valuefrac = GetValueFrac(values, now);
+      CHECK(valuefrac >= 0);
+      CHECK(valuefrac <= 1);
       score += (valuefrac - lastvaluefrac);
       lastvaluefrac = valuefrac;
     }
+    */
 
     if (score <= 0.0) {
       printf("Bad objective lost more than gained: %f / %s\n",
 	     score, ObjectiveToString(obj).c_str());
-      it->second->weight = 0.0;
+      info->weight = 0.0;
     } else {
-      it->second->weight = score;
+      info->weight = score;
     }
   }
 }
[…]

### 2013-03-14 tasbot/weighted-objectives.h

@@ -36,12 +36,6 @@ struct WeightedObjectives {
   double Evaluate(const vector<uint8> &mem1,
                   const vector<uint8> &mem2) const;
 
-  // Old buggy version. Same as GetNumLess I think?
-#if 0
-  double BuggyEvaluate(const vector<uint8> &mem1,
-                       const vector<uint8> &mem2) const;
-#endif
-
   // Observe a game state. This informs us about the values that
   // the objective functions can take on, which lets us score the
   // magnitude of their changes. Not necessary for GetNumLess() or

### 2013-03-14 trunk/tasbot/learnfun.cc
Fix bug in learnfun

@@ -123,6 +123,7 @@ static void MakeObjectives(const vector< vector<uint8> > &memories) {
   // Weight them. Currently this is just removing duplicates.
   printf("There are %d objectives\n", objectives->size());
   WeightedObjectives weighted(*objectives);
+  printf("And %d example memories\n", memories.size());
   weighted.WeightByExamples(memories);
   printf("And %d unique objectives\n", weighted.Size());
 

### 2013-03-14 trunk/tasbot/playfun.cc

@@ -221,18 +221,18 @@ struct PlayFun {
 
   // Number of real futures to push forward.
   // XXX the more the merrier! Made this small to test backtracking.
-  static const int NFUTURES = 50;
+  static const int NFUTURES = 40;
 
   // Number of futures that should be generated from weighted
   // motifs as opposed to totally random.
-  static const int NWEIGHTEDFUTURES = 45;
+  static const int NWEIGHTEDFUTURES = 35;
 
   // Drop this many of the worst futures and replace them with
   // totally new futures.
   static const int DROPFUTURES = 5;
   // Drop this many of the worst futures and replace them with
   // variants on the best future.
-  static const int MUTATEFUTURES = 10;
+  static const int MUTATEFUTURES = 7;
 
   // Number of inputs in each future.
   static const int MINFUTURELENGTH = 50;

### 2013-03-14 trunk/tasbot/paper/paper.tex
Make this compile

@@
 \usepackage{graphicx}
+\usepackage{fancyvrb}
 
@@
 
-\newcommand[1]{comment}{}
-\newcommand[1]{parameteralert}{
+\newcommand\comment[1]{}
+\newcommand\parameteralert[1]{
   % XXX this can get cooler.
@@
 
-% XXX superscript
-\newcommand\th{th}
+\newcommand\st{$^{\mathrm{st}}$}
+\newcommand\nd{$^{\mathrm{nd}}$}
+\newcommand\rd{$^{\mathrm{rd}}$}
+\renewcommand\th{$^{\mathrm{th}}$}
+\newcommand\tm{$^{\mbox{\tiny \textsc{tm}}}$}
+
+% nice fractions
+\newcommand\sfrac[2]{{}\,$^{#1}$\!/{}\!$_{#2}$}
 
@@
 
-\begin{figure}
+\begin{figure}[ht]
 \begin{center}
-% XXX make image
-% \includegraphics[width=0.75 \linewidth]{bytes2048}
+% TODO: better if actual ram!
+\includegraphics[width=0.75 \linewidth]{bytes2048}
 \end{center}\vspace{-0.1in}
@@
 You start in World 1-1 and the underground level that comes next is World 1-2 (we'll call this $w=1$ and $\ell=2$).
-But after you discover the princess is in another castle in World 1-4, the next level is 2-1.\footnote{In case you never realized this, it is time to learn that the legendary ``Minus World'' of \verb+ -1+ is not actually a negative world, but World 36-1 being incorrectly rendered because there is no glyph for the 36\th\ digit.
+But after you discover the princess is in another castle in World 1-4, the next level is 2-1.\footnote{In case you never realized this, it is time to learn that the legendary ``Minus World'' of {\tt\, -1} is not actually a negative world, but World 36-1 being incorrectly rendered because there is no glyph for the 36\th\ digit.
 The trick used to get to the Minus World just happens to leave the value 36 in that memory location rather than initializing it to a useful value.
-The ROM does not contain data for world 36 so it just interprets garbage data as a description of the world.} This can't be represented as a single byte going up (sometimes the second part $\ell$ goes down when we get to a new first part $w$), but it can be represented as a lexicographic order on the pair $\langle w, \ell \rangle$; that is, $\langle w_1, \ell_1 \rangle < \langle \w_2, \ell_2$ if $\w_1 = \w_2$ and $\ell_1 < \ell_2$, or if $\w_1 < \w_2$ no matter the values of $\ell_1$ and $\ell_2$.
+The ROM does not contain data for world 36 so it just interprets garbage data as a description of the world.} This can't be represented as a single byte going up (sometimes the second part $\ell$ goes down when we get to a new first part $w$), but it can be represented as a lexicographic order on the pair $\langle w, \ell \rangle$; that is, $\langle w_1, \ell_1 \rangle < \langle w_2, \ell_2$ if $w_1 = w_2$ and $\ell_1 < \ell_2$, or if $w_1 < w_2$ no matter the values of $\ell_1$ and $\ell_2$.
 This matches our intuitive idea and is also mathematically nice.
@@
 
-\begin{figure}[htp]
+\begin{figure*}[htp]
   \begin{code}
@@
   \label{fig:ordering}
-\end{figure}
+\end{figure*}
 
@@
 %
-$$ \Sigma_{i=1}^{n-1}  {\tt VF}($M_{i+1}$) - {\tt VF}($M_{i}$) $$
+$$ \Sigma_{i=1}^{n-1}  {\tt VF}(M_{i+1}) - {\tt VF}(M_{i}) $$
 %
 We sum the differences in value functions for each consecutive pair of
-memories. In the case of the ideal function this is \Sigma_{i=1}^{n-1}
-$1/n$, which approaches $1$.\footnote{The code implements a different
+memories. In the case of the ideal function this is $\Sigma_{i=1}^{n-1}
+1/n$, which approaches $1$.\footnote{The code implements a different
   weighting scheme, which sums from $i=0$ and keeps track of the last
@@
 \begin{code}
-  for (;;) {
+  for (;;) \{
     vector<uint8> before = GetMemory();
@@
     // Try every bitmask of the 8 inputs.
-    for (int i = 0; i < 256; i++) {
+    for (int i = 0; i < 256; i++) \{
       RestoreState(state);
@@
       // Save the best-scoring i...
-    }
+    \}
     RestoreState(state);
     Step(bestinput);
-  }
+  \}
 \end{code}
@@
 \begin{code}
-for (const Motif &m : motifs) {
+for (const Motif &m : motifs) \{
   RestoreState(state);
@@
   // Save the best-scoring motif...
-}
+\}
 \end{code}

### 2013-03-16 trunk/tasbot/paper/paper.tex
Some results

@@
 
+% good parts!
+% pacman @6800
+
+\section{Results}
+
+These games were all played with the same settings developed and tuned
+for Super Mario Bros.; that is, this was the result of just running
+the program the first time on a recording of me playing the game briefly.
+
+\subsection{Hudson's Adventure Island}
+
+This is a very difficult game featuring a skateboarding cherubic
+island guy called Master Higgins, whose girlfriend has been kidnapped
+by King Quiller. He basically runs to the right and can throw or ride
+stuff that he finds in eggs. The controls are pretty soft and danger
+is everywhere. The objective functions that work are probably pretty
+similar to Super Mario Bros., but there are fewer obstacles to
+navigate---the difficulty for humans mostly comes from the speed and
+reaction time. Master Higgins doesn't care about bumping into rocks
+and dropping his health almost to nothing (but then is careful about
+health), and once he gets a weapon his aim anticipates off-screen
+enemies and he looks pretty savvy (Figure~\ref{fig:adventureisland}).
+His weakness regards holes in the ground, which he sometimes jumps
+into, probably for the same reason that Mario sometimes does. His
+``pot bonus'' was 16720. Master Higgins beats several levels and makes
+it into the cave before losing his last life by jumping straight into
+a vibrating skull with swirling fireballs around it, which on a scale
+of Darwin Award to dignified death is approximately a 6, in my
+opinion.
+
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[width=0.75 \linewidth]{adventureisland}
+\end{center}\vspace{-0.1in}
+\caption{Master Higgins putting safety first.}
+\label{fig:adventureisland}
+\end{figure}
+
+\subsection{Pac-Man}
+
+One of the smallest NES games at a mere 24kb, Pac-Man is a classic
+that I won't belabor. It has a fairly straightforward objective---your
+score---and there is no timer or anything to worry about but the
+ghosts. A good planner could probably solve Pac-Man with just score as
+an objective (keep in mind that with RAM inspection and because of
+ghost movement, this is not just a simple matter of using A$^{*}$ or
+something to plan a path through the Euclidean grid---states aren't
+repeated hardly ever). Automating this game works pretty well; Pac-Man
+doesn't have much trouble avoiding ghosts except when they trap him.
+Play is pretty strange: He doesn't necessarily grab nearby dots (hard
+to explain) and often switches direction in the middle of a corridor,
+unlike human play which is usually efficient sweeps of the dots when
+ghosts aren't near. However, automating has a huge advantage over human
+players with respect to ghosts, and Pac-Man is alarmingly fearless. He
+chases ghosts down corridors, turns straight into them, knowing that
+they will change direction in time to not run into him (this makes
+the time travel advantage quite stark). Figure~\ref{fig:pacman} shows
+gratuitous daredeviling as he ducks into a tiny sliver of space between
+two ghosts, then again, and survives.
+
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[width=0.75 \linewidth]{pacman1}
+\includegraphics[width=0.75 \linewidth]{pacman2}
+\includegraphics[width=0.75 \linewidth]{pacman3}
+\end{center}\vspace{-0.1in}
+\caption{Pac-Man showing utter disregard for the ghosts's personal
+  space. This occurs around frame 6700 of the output movie. Pac-Man
+  slips into the space between Blinky and Inky to touch the wall,
+  comes out unharmed, then momentarily teases Clyde before escaping
+  to vibrate some more in empty corridors.}
+\label{fig:pacman}
+\end{figure}
+
 \end{document}

### 2013-03-16 trunk/tasbot/game.h
Paper ignores

@@ -11,17 +11,20 @@
 #define BASE64 "base64:jjYwGG411HcjG/j9UOVM3Q=="
 */
 
+/*
 #define GAME "adventureisland"
 #define MOVIE "adventureisland-tom.fm2"
 #define BASE64 "base64:JIuhrL/tKL5OaDmoWKXGCg=="
 #define FASTFORWARD 0
+*/
 
 // #define GAME "karate"
 // #define MOVIE "karate-tom.fm2"
 // #define BASE64 "base64:6xX0UBv8pLORyg1PCzbWcA=="
 
-// #define GAME "pacman"
-// #define MOVIE "pacman-tom.fm2"
-// #define FASTFORWARD 600
+#define GAME "pacman"
+#define MOVIE "pacman-tom.fm2"
+#define FASTFORWARD 600
+#define BASE64 "base64:ymBr2Kh1o5bVJzXDu4T6Zw=="
 
 #endif

### 2013-03-16 tasbot/paper/paper.tex
Paper ignores

--- 2013-03-16 trunk/tasbot/paper/paper.tex
@@
 The location \verb+0x75F+ contains the current world, and \verb+0x760+ the current level.
+The NES outputs 60.0988 frames per second, which we will just call 60 in this paper.
 
@@
 (There are of course many details to work out!
-But in essence emulation is jut that.) This process is completely deterministic, so it is possible to record the sequence of inputs (the inputs can only be read once per video frame, so this sequence is 60 \comment{ XXX CHECK } bytes per second) and play them back and get the same result.
+But in essence emulation is jut that.) This process is completely deterministic, so it is possible to record the sequence of inputs (the inputs can only be read once per video frame, so this sequence is 60 bytes per second) and play them back and get the same result.
 This also means that an input sequence can be computed in non-real time, either much slower or much faster than a NES would normally run.
@@
 \begin{center}
-\includegraphics[width=0.75 \linewidth]{adventureisland}
+\includegraphics[width=0.95 \linewidth]{adventureisland}
 \end{center}\vspace{-0.1in}
@@
 unlike human play which is usually efficient sweeps of the dots when
-ghosts aren't near. However, automating has a huge advantage over human
-players with respect to ghosts, and Pac-Man is alarmingly fearless. He
-chases ghosts down corridors, turns straight into them, knowing that
-they will change direction in time to not run into him (this makes
-the time travel advantage quite stark). Figure~\ref{fig:pacman} shows
-gratuitous daredeviling as he ducks into a tiny sliver of space between
-two ghosts, then again, and survives.
+ghosts aren't near. However, automating has a huge advantage over
+human players with respect to ghosts, and Pac-Man is alarmingly
+fearless. He chases ghosts down corridors, turns straight into them,
+knowing that they will change direction in time to not run into him
+(this makes the time travel advantage quite stark).
+Figure~\ref{fig:pacman} shows gratuitous daredeviling as he ducks in
+and out of a tiny sliver of space between two ghosts, then does it
+again, and survives.
 
@@
 \begin{center}
-\includegraphics[width=0.75 \linewidth]{pacman1}
-\includegraphics[width=0.75 \linewidth]{pacman2}
+\includegraphics[width=0.75 \linewidth]{pacman1} \\[0.3em]
+\includegraphics[width=0.75 \linewidth]{pacman2} \\[0.3em]
 \includegraphics[width=0.75 \linewidth]{pacman3}

### 2013-03-16 tasbot/paper/paper.tex
Working on paper

@@
 \usepackage{fancyvrb}
+\usepackage{url}
+% \usepackage{inlinebib}
 
@@
 \newcommand\sfrac[2]{{}\,$^{#1}$\!/{}\!$_{#2}$}
+
+\newcommand\citef[1]{\addtocounter{footnote}{1}\footnotetext{\cite{#1}}\ensuremath{^{\mbox{\footnotesize [\thefootnote]}}}}
 
@@
 This suggested to me that it may be time to automate the playing of NES games, in order to save time.\footnote{Rather, to replace it with time spent programming.} In this paper I present a generic technique for automating the playing of NES games.
-The approach is practical on a single computer, and succeeds on some games, such as Super Mario Bros..
+The approach is practical on a single computer, and succeeds on several games, such as Super Mario Bros..
 The approach is amusingly elegant and surprisingly effective, requires no detailed knowledge of the game being played, and is capable of novel and impressive gameplay (for example, bug exploitation).
@@
 
-On a scale from ``the title starts with Toward'' to ``Donald Knuth has finally finished the 8\th\ volume on the subject'', this work is a 3.
+On a scale from ``the title starts with Toward'' to ``Donald Knuth has finally finished the 8\th\ volume on the subject,'' this work is a 3.
 The purpose of this paper is mainly as a careful record of the current status for repeatability and further development on this important research subject.
@@
 
-The objective function, the algorithm to deduce it, the search strategy, and its implementation are all interesting and will be discussed in that order.
-To set the stage, I begin with a description of the NES hardware and emulation of it.
+The objective function, the algorithm to deduce it, the search
+strategy, and its implementation are all interesting and will be
+discussed in that order. I then discuss the results of using the
+approach to automate several NES games. To set the stage, I begin with
+a description of the NES hardware and emulation of it.
 
@@
 
+% ends with 13 dots remaining
+
 \begin{figure}[ht]
@@
 
+\subsection{Bubble Bobble}
+
+Let's take a journey to the cave of monsters! This lovely game
+features two stout dinosaurs called Bub and Bob, who jump around in a
+series of single-screen caves. You can shoot bubbles to encase
+monsters and then pop the bubble to turn them into fruit or other
+treasure; clearing all the monsters takes you to the next stage.
+Bubbles are also necessary for navigating around, since you can bounce
+on your own bubbles when holding the jump button.
+
+I was surprised that automating this game works at all, since there is
+no obvious spatial gradient like in Super Mario Bros., and few local
+greedy rewards like in Pac-Man. Bub initially wastes his lives (a
+common theme in games where the respawn interval is low---it just
+looks like a fast way of warping to the corner). When he's near
+monsters he shows good tactics in shooting and popping them
+simultaneously, even doing so while facing away from them, which I
+didn't know was possible. At the beginning of the movie he prefers to
+hide in the bottom-left corner, but soon starts jumping all around the
+level, bouncing on his own bubbles and so on, almost like a real
+player would. He beats XXX levels, more than I would have guessed.
+Since it takes two start button presses to enter the game, the second
+START is part of the input motifs. Amusingly, Bub uses this to his
+advantage to change the synchronization of his futures, and when the
+situation gets really tight, he'll pause the game until a good
+future comes around (Figure~\ref{fig:bubblebobble}). Paused states
+probably look like modest progress since memory locations like the
+music cursor are still advancing.
+
+XXX what finally happens?
+
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[width=0.95 \linewidth]{bubblebobble}
+\end{center}\vspace{-0.1in}
+\caption{Bub navigates this game surprisingly well. Once he's on his
+  last life, he becomes careful and pauses the game when things are
+  looking grim---here pausing for about a thousand frames, burning
+  through the futures until one randomly comes along that looks good.
+  He then unpauses and executes that good future, killing three of
+  these monsters in short order.}
+\label{fig:bubblebobble}
+\end{figure}
+
+\section{Future Work}
+
+It is famous last words, but I actually intend to keep working on this
+project. Because of SIGBOVIK crunch pressure (and discovering some
+bugs as I was writing the paper) the approach got much better very
+recently and I'm mostly limited by CPU hours until conference. It's
+today in a state where whenever I run it on a new game I see something
+delightful. Even just running it on more of the NES classics is
+worthwhile. However, I have lots of ideas about how to extend the
+technique, either to make it more beautiful or to make it work better:
+
+\paragraph{Parameter reduction.} I've already pointed out the places
+where there are mysterious constants in the code. Although the fact
+that the same constants work on many different games without tuning
+is some solace, it would really be nicer to remove these. One of the
+hardest to remove is the length of the futures explored. XXX thoughts
+on how to improve this?
+
+\paragraph{Multiple players, multiple games.} Other than the particulars
+of the way it's built (\verb+vector<uint8>+ everywhere), there's no
+reason why the technique is limited to a single player's input. In a
+game like Bubble Bobble or Contra the two players can collaborate in
+interesting ways, and planning both simultaneously would probably lead
+to occasional awesome results. For example, in Contra, it might be
+that one player is shooting enemies for the other player, the bullets
+arriving from across the screen just in time to save him as he
+blithely jumps into danger. Another clever feat from the Tool Assisted
+Speedrun community is a sequence of inputs that beats multiple
+different {\em games} simultaneously. For example, human geniuses used
+tools to beat Mega Man (Mega Men?) 3, 4, 5, and 6 all at the same time
+using the exact same input sequence sent to all four
+games.\cite{BaxterTAS} I think the algorithms in this paper apply
+directly---it's just a matter of multiplexing the inputs to each of
+the games, and then taking the appropriate operation (min, max, sum) of
+their objective functions to produce a final objective function. The
+main obstacle is the architecture of the underlying emulator, which
+can only load one game into its global variables at once.
+
+\bibliographystyle{plain}
+% \bibliographycomment{} % nothing
+\bibliography{paper}
+
 \end{document}

### 2013-03-16 trunk/tasbot/paper/paper.tex
More about weighting objectives. Some figures

@@
 \usepackage{fancyvrb}
+\usepackage{url}
+\usepackage{textcomp}
+% \usepackage{inlinebib}
 
@@
 
-\section*{Introduction}
+\section{Introduction}
 The Nintendo Entertainment System is probably the best video game console, citation {\it not} needed.
@@
 
-There are a number of emulators for NES.
-These work by simulating the NES hardware, for example with a 2048-byte array for its memory, and simulating the steps of its 2A03 processor on some ROM, and hooking a keyboard or joystick into the 8 bits of input.
-(There are of course many details to work out!
-But in essence emulation is jut that.) This process is completely deterministic, so it is possible to record the sequence of inputs (the inputs can only be read once per video frame, so this sequence is 60 \comment{ XXX CHECK } bytes per second) and play them back and get the same result.
-This also means that an input sequence can be computed in non-real time, either much slower or much faster than a NES would normally run.
-In this work we use the FCEUX emulator, which is popular for its accuracy and advanced tools.
+There are a number of emulators for NES. These work by simulating the
+NES hardware, for example with a 2048-byte array for its memory, and
+simulating the steps of its 2A03 processor on some ROM, and hooking a
+keyboard or joystick into the 8 bits of input. (There are of course
+many details to work out! But in essence emulation is just that.) This
+process is completely deterministic, so it is possible to record the
+sequence of inputs (the inputs can only be read once per video frame,
+so this sequence is 60 bytes per second) and play them back and get
+the same result. This also means that an input sequence can be
+computed in non-real time, either much slower or much faster than a
+NES would normally run. In this work we use the FCEUX\cite{FCEUX}
+emulator, which is popular for its accuracy and advanced tools.
 
@@
 
-Bytes in memory (and sometimes 16- and 32-bit words) can contain interesting game facts like the player's position in the level or score.
-The central idea of this paper is to use (only) the value of memory locations to deduce when the player is ``winning''.
-The things that a human player perceives, like the video screen and sound effects, are completely ignored.
-As an additional simplification, we assume that winning always consists of a value {\it going up}---either the position in the level getting larger, the score getting larger, the number of lives, the world or level number getting bigger, etc.
+Bytes in memory (and sometimes 16- and 32-bit words) can contain
+interesting game facts like the player's position in the level or
+score. The central idea of this paper is to use (only) the value of
+memory locations to deduce when the player is ``winning''. The things
+that a human player perceives, like the video screen and sound
+effects, are completely ignored. As an additional simplification, we
+assume that winning always consists of a value {\it going up}---either
+the position in the level getting larger, the score getting larger,
+the number of lives, the world or level number getting bigger, and so
+on.
 
-This is actually a little bit too naive; for example, Mario's overall progress through the game is represented by a pair.
-You start in World 1-1 and the underground level that comes next is World 1-2 (we'll call this $w=1$ and $\ell=2$).
-But after you discover the princess is in another castle in World 1-4, the next level is 2-1.\footnote{In case you never realized this, it is time to learn that the legendary ``Minus World'' of {\tt\, -1} is not actually a negative world, but World 36-1 being incorrectly rendered because there is no glyph for the 36\th\ digit.
-The trick used to get to the Minus World just happens to leave the value 36 in that memory location rather than initializing it to a useful value.
-The ROM does not contain data for world 36 so it just interprets garbage data as a description of the world.} This can't be represented as a single byte going up (sometimes the second part $\ell$ goes down when we get to a new first part $w$), but it can be represented as a lexicographic order on the pair $\langle w, \ell \rangle$; that is, $\langle w_1, \ell_1 \rangle < \langle w_2, \ell_2$ if $w_1 = w_2$ and $\ell_1 < \ell_2$, or if $w_1 < w_2$ no matter the values of $\ell_1$ and $\ell_2$.
-This matches our intuitive idea and is also mathematically nice.
-It also generalizes multi-byte encodings of things like your score (which can be larger than 8 bits and so is often stored in 16 or 32), including both big-endian and little-endian representations.\footnote{A possible additional simplification would be to just take lexicographic orderings over bits, which then generalizes to 8-bit bytes.
-This is probably too crazy, but just right now I am sort of feeling like maybe I should try it, though it may be the beer.}
+This is actually a little bit too naive; for example, Mario's overall
+progress through the game is represented by a pair. You start in World
+{\tt 1-1} and the underground level that comes next is World {\tt 1-2}
+(we'll call this $w=1$ and $\ell=2$). But after you discover the
+princess is in another castle in World {\tt 1-4}, the next level is
+{\tt 2-1}.\footnote{In case you never realized this, it is time to learn
+  that the legendary ``Minus World'' of {\tt\, -1} is not actually a
+  negative world, but World {\tt 36-1} being incorrectly rendered because
+  there is no glyph for the 36\th\ digit. The trick used to get to the
+  Minus World just happens to leave the value 36 in that memory
+  location rather than initializing it to a useful value. The ROM does
+  not contain data for world 36 so it just interprets garbage data as
+  a description of the world.} This can't be represented as a single
+byte going up (sometimes the second part $\ell$ goes down when we get
+to a new first part $w$), but it can be represented as a lexicographic
+order on the pair $\langle w, \ell \rangle$; that is, $\langle w_1,
+\ell_1 \rangle < \langle w_2, \ell_2 \rangle$ if $w_1 = w_2$ and $\ell_1 <
+\ell_2$, or if $w_1 < w_2$ no matter the values of $\ell_1$ and
+$\ell_2$. This matches our intuitive idea and is also mathematically
+nice. It also generalizes multi-byte encodings of things like your
+score (which can be larger than 8 bits and so is often stored in 16 or
+32), including both big-endian and little-endian
+representations.\footnote{A possible additional simplification would
+  be to just take lexicographic orderings over bits, which then
+  generalizes to 8-bit bytes. This is probably too crazy, but just
+  right now I am sort of feeling like maybe I should try it, though it
+  may be the beer.}
 
-More importantly, it allows the combination of semantically unrelated bytes, like: $\langle$ world, level, screen inside the world, $x$ position on the screen $\rangle$ or $\langle$ world, lives, low byte of score $\rangle$.
-Many orderings may describe gameplay.
-These orderings may be temporarily violated in normal play: Although the score always goes up, Mario's $x$ position may temporarily decrease if he needs to navigate around some obstacle.\footnote{Note to self: Maybe we should give a much higher score to globally preserved objectives than to locally preserved ones.
-But that may presuppose that the input represents a whole playthrough?} So, to ``faithfully'' represent gameplay, we will generate a set of lexicographic orderings on memory locations, with the idea that they ``generally go up'' but not necessarily at every step.
-These orderings will also have weights.
-The next section describes how we take a sequence of player inputs and deduce the orderings.
+More importantly, it allows the combination of semantically unrelated
+bytes, like: $\langle$\!\!\!\! world, level, screen inside the world,
+$x$ position on the screen \!\!$\rangle$ or $\langle$\!\!\!\! world,
+lives, low byte of score \!\!$\rangle$. Many orderings may describe
+gameplay. These orderings may be temporarily violated in normal play:
+Although the score always goes up, Mario's $x$ position may
+temporarily decrease if he needs to navigate around some
+obstacle.\footnote{Note to self: Maybe we should give a much higher
+  score to globally preserved objectives than to locally preserved
+  ones. But that may presuppose that the input represents a whole
+  playthrough?} So, to ``faithfully'' represent gameplay, we will
+generate a set of lexicographic orderings on memory locations, with
+the idea that they ``generally go up'' but not necessarily at every
+step. These orderings will also have weights. The next section
+describes how we take a sequence of player inputs and deduce the
+orderings.
 
 \subsection{Deriving the objective function} \label{sec:deriving}
+
+\begin{figure}[htb]
+\begin{center}
+\includegraphics[width=0.95 \linewidth]{mario-global}
+\end{center}\vspace{-0.1in}
+\caption[lof not used]{
+  A single maximal tight valid lexicographic ordering for my
+  4,000-frame input training data to Super Mario Bros.. This function
+  is globally nondecreasing, and is the decimal memory locations
+  $\langle$\!~232, 57, 73, 74, 75, 115, 130, 155, 32, 184, 280,
+  491, 506, 1280, 1281, 1282, 1283, 1288, 1290, 1337, 1338, 1339,
+  1384, 1488, 1490, 1496, 1497, 1498, 1499, 1514, 1873, 1882, 1888,
+  1904, 1872, 1906, 112, 113, 114, 2009, 2010, 2011, 1539~\!$\rangle$.
+
+  This is not a great objective function; there are long spans where
+  all the memories are equal according to it, and the nice smooth slopes
+  are happening during level transitions where the game is ignoring
+  inputs (they are probably timers counting up each frame, which is
+  why they are so smooth). Other slicing produces better objectives.
+
+  For reasons unknown---I just discovered this while generating the
+  figure---{\em all} of the objective functions learned with this
+  method, regardless of the nondeterministic choices, appear to have
+  this same curve, despite using different memory locations. It may be
+  that they are different permutations of bytes that all change
+  simultaneously, only on the frames where there are jumps in this
+  picture, and there are no other orderings that are tight, valid, and
+  maximal. This is still surprising and warrants investigation. }
+\label{fig:global}
+\end{figure}
+
+\begin{figure}[htb]
+\begin{center}
+\includegraphics[width=0.95 \linewidth]{mario-tenths}
+\end{center}\vspace{-0.1in}
+\caption{Ten objective functions trained on different tenths of the
+  4,000 inputs for Super Mario Bros.. These functions are normalized
+  against the range of all values they take on during the movie; you
+  can see that most are increasing locally most of the time, but many
+  drop back to zero around the 3100\th\ frame, when Mario reaches
+  world {\tt 1-2}. Within its 400-frame window, each objective is
+  guaranteed to be nondecreasing.}
+\label{fig:tenths}
+\end{figure}
+
+\begin{figure}[htb]
+\begin{center}
+\includegraphics[width=0.95 \linewidth]{mario-strides}
+\end{center}\vspace{-0.1in}
+\caption{Ten objective functions trained on every 100\th\ memory,
+  starting at frame 0, frame 1, and so on up to frame 10. Normalized
+  as usual. These objectives exhibit excellent global progress, but
+  are locally very noisy. Among the methods I used to generate
+  objectives, this one produces the best results (based on my
+  intuition about what a good objective function looks like).
+}
+\label{fig:strides}
+\end{figure}
 
@@
 Every $L$ defines a lexicographic ordering ($<$ and $=$ operators).
-$L$ is a {\em valid} lexicographic ordering of $M$ if $M_i \leq_L M_{i + 1}$
-for $1 \leq i \leq n$; each memory is less than or equal to the next
-memory in the sequence. It follows that $M_i \leq_L M_j$ whenever $i < j$.
+$L$ is a {\em valid} lexicographic ordering of $M$ if $M_i \leq_L M_{i
+  + 1}$ for $1 \leq i \leq n$; each memory is less than or equal to
+the next memory in the sequence. It follows that $M_i \leq_L M_j$
+whenever $i < j$.
 
@@
 
-On a scale from simple to fancy, this algorithm is a 3. Given those
-definitions, the idea is to start with the empty prefix, which is
-always a tight valid lexicographic ordering but usually not maximal.
-We then pick a tight extension that is valid; if none exists then we are
-done and have a maximal tight valid ordering. Now we have a new tight valid
-ordering, and we repeat.
+On a scale from simple to fancy, the algorithm to generate $L$ from
+$M$ is a 3. Given those definitions, the idea is to start with the
+empty prefix, which is always a tight valid lexicographic ordering but
+usually not maximal. We then pick a tight extension that is valid; if
+none exists then we are done and have a maximal tight valid ordering.
+Otherwise we have a new tight valid ordering, and we repeat.
 
@@
   \begin{code}
-    (* Prefix is the prefix so far (int list)
-       and remain is the list of memory locations
-       that we can still consider. Invt is that
-       prefix is a tight valid ordering on M. Returns
-       the memory locations from remain that are
-       a tight extension of the prefix. *)
+    (* Prefix is the prefix so far (int list) and remain is the list of memory
+       locations that we can still consider. Invariant is that prefix is a
+       tight valid ordering on $M$. Returns the memory locations from remain 
+       that are a tight extension of the prefix. *)
     fun candidates (prefix, remain) =
       let lequal = (* list of indices $i$ where
-                      $M_i =_{\tt prefix} M_{i+1}$ *)
+                      $M_i =_{\tt\! prefix} M_{i+1}$ *)
       let notgreater = (* members $x$ of remain where
-                          $M_i[x] > M_{i + 1}[x]$ is
+                          $M_i[x] > M_{i+1}[x]$ is
                           not true for any $i$ in
@@
       let tight = (* members $y$ of notgreater where
-                     $M_i[x] < M_{i + 1}[x]$ is true
-                     for some $i$ in {\tt lequal} *)
+                     $M_i[x] < M_{i+1}[x]$ is true
+                     for some $i$ in{\tt lequal} *)
       in tight
 
-    (* Returns a maximal tight valid ordering, given
-       a tight valid prefix and list of memory locations
-       we are permitted to consider. *)
+    \
+
+    (* Returns a maximal tight valid ordering, given a tight valid prefix and 
+       list of memory locations we are permitted to consider. *)
     fun ordering (prefix, remain) =
@@
 
-{\bf Skipping.}\,\, To avoid being confused by RAM initialization and
+\paragraph{Skipping.} To avoid being confused by RAM initialization and
 menu, I ignore all memories up until the first input by the player.
@@
   they are truly uninitialized, or retain the values from the last
-  reset or game inserted. Some semi-famous tricks involve removing
-  and inserting game cartridges while the system is running in order
-  to take advantage of this.}
+  reset or game inserted. Some semi-famous tricks involve removing and
+  inserting game cartridges while the system is running in order to
+  take advantage of this.}
 
-% XXXX good place for some graphs.
-{\bf Slicing.}\,\, I generate 50 orderings for $M_1 \ldots M_n$; the
-whole recording starting from the first keypress. During gameplay some
-values really are completely nondecreasing, like the player's score
-and world/level pair. I also generate 3 orderings for each tenth of
-the memory sequence, e.g. $M_1 \ldots M_{n/10}$ and $M_{n/10+1} \ldots
-M_{2n/10}$, etc. The intention is to capture orderings that are rarely
-violated, or sections of the game with unusual objectives (e.g. a
-minigame or swimming level). Orderings generated this way look pretty
-random, and on a scale from solid to suspicious, I can't vouch for
-them. Then I generate objectives from non-consecutive memories that
-are evenly spread out through the observations: Ten objectives
-chosen from every 100\th\ memory, starting from the 0\th\ frame,
-1\st\ frame, 2\nd\ frame, and so on up to the 9\th{}. Similarly
-for every 250\th\ frame, and a single objective for memory sliced
-to every 1000\th\ frame, with start positions of 0--9.
+\paragraph{Slicing.} I generate 50 orderings for $M_1 \ldots M_n$; the
+whole recording starting immediately after the first keypress. During
+gameplay some values really are completely nondecreasing, like the
+player's score and world/level pair. Figure~\ref{fig:global} shows
+what a global ordering looks like. I also generate 3 orderings for
+each tenth of the memory sequence, e.g. $M_1 \ldots M_{n/10}$ and
+$M_{n/10+1} \ldots M_{2n/10}$, etc. The intention is to capture
+orderings that are rarely violated, or sections of the game with
+unusual objectives (e.g. a minigame or swimming level). Orderings
+generated this way look pretty random, and on a scale from solid to
+suspicious, I can't vouch for them. Then I generate objectives from
+non-consecutive memories that are evenly spread out through the
+observations: Ten objectives chosen from every 100\th\ memory,
+starting from the 0\th\ frame, 1\st\ frame, 2\nd\ frame, and so on up
+to the 9\th{}. Similarly for every 250\th\ frame, and a single
+objective for memory sliced to every 1000\th\ frame, with start
+positions of 0--9. The intention is to capture objectives that grow
+slowly over the course of a playthrough, without getting distracted by
+local noise.
 
-{\bf Weighting.},\, To reduce the importance of randomness in the
+\paragraph{Weighting.} To reduce the importance of randomness in the
 learning process, and the arbitrariness of the slicing, each objective
@@
 all the values we ever saw for it. The value fraction is defined and
-in $[0, 1]$ for all memories in the observed set.\footnote{It is not
+in $[0, 1)$ for all memories in the observed set.\footnote{It is not
   defined when the memory is greater than all observed memories, which
-  we will encounter later. The code returns $(i+1)/|V|$ in that case,
+  we will encounter later. The code returns $|V|/|V| = 1$ in that case,
   which is as good as anything.} This gives us the ability to compare
@@
 memories. In the case of the ideal function this is $\Sigma_{i=1}^{n-1}
-1/n$, which approaches $1$.\footnote{The code implements a different
-  weighting scheme, which sums from $i=0$ and keeps track of the last
-  value fraction at each step, thus treating the value fraction of the
-  nonexistent memory $M_0$ as 0. This is wrong, because the first
-  value fraction may be very high, which credits the objective with a
-  positive value (e.g. 1) for that first part of the sum. Objectives
-  that start high on the first frame are not ideal; in fact, the worst
-  objectives start high on the first frame and steadily {\em
-    decrease}. I plan to correct this bug and study if it has any
-  noticeable consequences.} Objectives are constrained to have
-non-negative weights (I set the value to 0 if negative, which
-effectively disables it).
+1/n$, which approaches $1$. Of course, when you think about it,
+this is actually the same as
+%
+\[
+\begin{array}{c}
+({\tt VF}(M_1) - {\tt VF}(M_0)) + ({\tt VF}(M_2) - {\tt VF}(M_1)) \,+ \\
+\vdots \\
+({\tt VF}(M_{m-1}) - {\tt VF}(M_{m-2})) + ({\tt VF}(M_{m}) - {\tt VF}(M_{m}))
+\end{array}
+\]
+%
+and all but $-{\tt VF}(M_0)$ and ${\tt VF}(M_m)$ cancel out. This
+means that the weight we use is just the final value minus the initial
+value, regardless of what happens in-between.\footnote{I think this
+  can be improved, for example by taking the deviation from the ideal
+  linear objective.} The mean value theorem or something is probably
+relevant here. {\bf Lesson about being careful:} I only realized that
+this code was kind of fishy when I started writing it up for SIGBOVIK.
+Not only did it loop over all the deltas as in the $\Sigma$ expression
+above, but it also summed from $i=0$ and kept track of the last value
+fraction at each step, thus treating the value fraction of the
+nonexistent memory $M_0$ as 0. This is wrong, because the first value
+fraction may be very high, which credits the objective with a positive
+value (e.g. 1) for that first part of the sum. Objectives that start
+high on the first frame are not ideal; in fact, the worst objectives
+start high on the first frame and steadily {\em decrease}. After
+writing this up I corrected it to the simple expression ${\tt VF}(M_m)
+- {\tt VF}(M_0)$ and the result was a huge breakthrough in the
+performance of the algorithm! I had spent much time tuning the {\tt
+  playfun} search procedure (Section~\ref{sec:playfun}) and not
+investigated whether the objectives were being weighted properly. More
+on this later, but the lesson is: Bugs matter, and thinking about your
+code and explaining it is a good way to find bugs.
 
-% XXX good place to give some weights for some objectives.
-
-We save the objectives and their weights to a file and then we are
-done with the easy part.
+Objectives are constrained to have non-negative weights (I set the
+value to 0 if negative, which effectively disables it). We save the
+objectives and their weights to a file and then we are done with the
+easy part.
 
@@
 
-\section{Now you're playing with power}
+\section{Now you're playing with power} \label{sec:playfun}
 
@@
 It's important to know that almost all the CPU time in all the
-algorithms discussed in this paper is spent emulating NES frames;
-it takes about XXX ms to process a single step. Lots of engineering
-effort goes into reducing the number of frames the algorithms emulate.
+algorithms discussed in this paper is spent emulating NES frames; it
+takes about 500\textmu s to process a single step. Lots of
+engineering effort goes into reducing the number of frames the
+algorithms emulate.
 
@@
 unlike human play which is usually efficient sweeps of the dots when
-ghosts aren't near. However, automating has a huge advantage over human
-players with respect to ghosts, and Pac-Man is alarmingly fearless. He
-chases ghosts down corridors, turns straight into them, knowing that
-they will change direction in time to not run into him (this makes
-the time travel advantage quite stark). Figure~\ref{fig:pacman} shows
-gratuitous daredeviling as he ducks into a tiny sliver of space between
-two ghosts, then again, and survives.
+ghosts aren't near. However, automating has a huge advantage over
+human players with respect to ghosts, and Pac-Man is alarmingly
+fearless. He chases ghosts down corridors, turns straight into them,
+knowing that they will change direction in time to not run into him
+(this makes the time travel advantage quite stark).
+Figure~\ref{fig:pacman} shows gratuitous daredeviling as he ducks in
+and out of a tiny sliver of space between two ghosts, then does it
+again, and survives.
+
+% ends with 13 dots remaining
 

### 2013-03-16 trunk/tasbot/paper/paper.tex
Inner loop

@@
 - {\tt VF}(M_0)$ and the result was a huge breakthrough in the
-performance of the algorithm! I had spent much time tuning the {\tt
+quality of the output! I had spent much time tuning the {\tt
   playfun} search procedure (Section~\ref{sec:playfun}) and not
@@
   model, although this has a free parameter too (the number of states
-  of history). It is likely possible to do some kid of abstract
+  of history). It is likely possible to do some kind of abstract
   interpretation where multiple different input sequences with
@@
 why. This algorithm is called {\tt playfun} and it can be found
-implemented in C++ in {\tt playfun.cc}.
+implemented in C++ in {\tt playfun.cc}; some historic versions are
+in {\tt playfun-*.cc}.
 
@@
   excellent emulator to use, has fancy tools for recording and editing
-  movies, and is popular in the speedrun community; just don't read
-  the code.} With a medium amount of suffering I got the emulator
-compiling under mingw in 64-bit mode, and working behind a streamlined
-interface ({\tt emulator.h}). Once a game is initialized, it is always
-at an input frame---you can give it an 8-bit input (for the
-1\st\ player) to step a single frame, or read the 2048 bytes of
-memory. You can also save the complete state of the emulator into a
-vector of bytes, which allows you to restore the state to exactly that
-same point.\footnote{This contains the RAM, but also stuff we didn't
-  consider, like the Picture Processing Unit's state, and internal
-  emulator stuff.} These save-states are portable across sessions as
-long as the code was compiled and the game initialized the right
-way.\footnote{The original FCEUX supports portable save-states, but I
-  removed that guarantee in order to get the smallest possible byte
-  vectors. More on that below.} FCEUX must be single-threaded because
-it uses global variables galore. I made several enhancements to the
-emulator interface, which are discussed later.
+  movies, and is popular in the speedrun community. I highly recommend
+  it; just don't read the code.} With a medium amount of suffering I
+got the emulator compiling under mingw in 64-bit mode, and working
+behind a streamlined interface ({\tt emulator.h}). Once a game is
+initialized, it is always at an input frame---you can give it an 8-bit
+input (for the 1\st\ player) to step a single frame, or read the 2048
+bytes of memory. You can also save the complete state of the emulator
+into a vector of bytes, which allows you to restore the state to
+exactly that same point.\footnote{This contains the RAM, but also
+  stuff we didn't consider, like registers, the Picture Processing
+  Unit's state, and internal emulator stuff.} These save-states are
+portable across sessions as long as the code was compiled and the game
+initialized the right way.\footnote{The original FCEUX supports
+  portable save-states, but I removed that guarantee in order to get
+  the smallest possible byte vectors. More on that below.} FCEUX must
+be single-threaded because it uses global variables galore. I made
+several enhancements to the emulator interface, which are discussed
+later.
 
@@
 algorithms discussed in this paper is spent emulating NES frames; it
-takes about 500\textmu s to process a single step. Lots of
-engineering effort goes into reducing the number of frames the
-algorithms emulate.
+takes about 500\textmu s to process a single step. Lots of engineering
+effort goes into reducing the number of frames the algorithms emulate.
+The {\tt playfun} program takes a ROM file, the learned objectives and
+motifs, and runs on the console for arbitrarily long, outputting a
+movie file consisting of the input sequence it think is best. The
+current {\tt playfun} algorithm is much too slow to run real-time, but
+it would be possible to have video output as it played. I disabled
+most of the video code, however, in an effort to make the emulation
+loop run as fast as possible.
 
-\subsection{Naive attempts}
+\subsection{Naive attempts} \label{sec:naive}
 
@@
 Section~\ref{sec:motifs}, was to just look at all $2^8$ different
-possible inputs at each step, and pick the best one. The inner
-loop looks pseudolike this:
+possible inputs at each step, and pick the best one. The inner loop
+looks pseudolike this:
 
@@
 
-I wish this worked, because that would be truly laughable, but
-on a scale from doesn't to does it's a 1. At best, Mario twitches
-in place. The inputs that it plays are insane. There are lots of
-reasons, but a clear one is that a single input rarely affects
-your progress in the game on the very next frame. I didn't really
-expect this approach to work and I already knew that the state
-space is too big to search exhaustively, which is why I implemented
-motifs. This drastically reduces the search space and makes each
-step more significant; the inner loop can now be:
+I wish this worked, because that would be truly laughable (and is fast
+enough to run real-time), but on a scale from doesn't to does it's a
+1. At best, Mario twitches in place. The inputs that it plays are
+insane. There are lots of reasons, but a clear one is that a single
+input rarely affects your progress in the game on the very next frame.
+I didn't really expect this approach to work and I already knew that
+the state space is too big to search exhaustively, which is why I
+implemented motifs. This drastically reduces the search space and
+makes each step more significant; the inner loop can now be:
 
@@
 Your Quest Is Over We Present You A New Quest Push Button B to I'm
-Sorry, but the Princess is in Another Similarly-Shaped but Not Exactly
+Sorry, but Our Princess is in Another Similarly-Shaped but Not Exactly
 that Samesuch Castle, this is an 8.
@@
 if we don't know how to reconstruct a rare good state that caused the
-maximum score to be high. Both lead us to the idea: Keep around
-a set of candidate futures that we use for evaluation and search,
-rather than just randomly generating futures when we want to evaluate
-a state. This turns out to work really well and be more efficient.
+maximum score to be high. Both lead us to the idea: Keep around a set
+of candidate futures that we use for evaluation and search, rather
+than just randomly generating futures when we want to evaluate a state
+and then discarding them. This turns out to work really well and be
+more efficient.
 
-HERE
+The basic version of the {\tt playfun} algorithm looks like this.
+Maintain {\tt NFUTURES} futures (this is 40 for the results presented
+in Section~\ref{sec:results}), originally just seeded with weighted
+random motifs. We aren't likely to execute any of these futures
+verbatim, but they are intended to give us high watermarks of what
+we're capable of, plus allow us to judge future danger. As we execute
+search, futures will be partly consumed and extended, and some discarded
+and replaced.
+
+Each future stores a desired length from 50--800 frames, and whenever
+it is shorter than that, we extend it (at its end) with random weighted
+motifs. The inner pseudoloop then looks like this:
+
+\begin{code}
+for (;;) \{
+  vector<uint8> before = GetMemory();
+  vector<uint8> state = GetState();
+\
+  set<vector<uint8>> nexts;
+  for (Future f : futures) \{
+   nexts.insert(f.First10Frames());
+   f.ChopOffFirst10Frames();
+  \}
+\ 
+  while (nexts.size() < NFUTURES)
+   nexts.push\_back(/* random motif */);
+\ 
+  for (vector<uint8> &next : nexts) \{
+    RestoreState(state);
+    for (uint8 i : next) Step(i);
+    double score = ScoreByFutures(before, futures);
+    // Save the best-scoring next...
+    // 
+  \}
+\
+  ReplaceBadFutures(futures);
+  ExtendFuturesToDesiredLengths(futures);
+\}
+\end{code}
+
+At each iteration of the loop we will find the best next 10-frame
+sequence to commit to. Rather than search over all motifs, we search
+over the first 10 frames of each future. This has several important
+consequences. First, since futures are generated by weighted motifs,
+it makes it more likely that we spend CPU time evaluating motifs that
+are common; the code from Section~\ref{sec:naive} always explores
+every motif, even rare ones. Second, it guarantees that if we pick
+some 10 frames on the basis of a single good future, we don't have to
+worry about recreating that future; it will still be among our futures
+for the next round. This is key: It allows us to use the determinism
+of the game to replay a really good future if we find one, not just
+use average random futures to assess how good a state will be.
+
+The function {\tt ScoreByFutures} saves the state, then runs each
+of the {\tt NFUTURES} futures to get a final state and memory. We
+score each final memory relative to the start memory. Scoring is
+a matter of
 
@@
 
-% good parts!
-% pacman @6800
+\subsection{MARIONET}
+
+\subsection{Backtracking}
 

### 2013-03-16 tasbot/paper/paper.tex
More work on the paper. Getting closer...

@@
 \usepackage{url}
+\usepackage{textcomp}
 % \usepackage{inlinebib}
@@
   % XXX this can get cooler.
+  \begin{quotation}
   {\bf Parameter Alert!}\,\, #1
+  \end{quotation}
 }
@@
 \begin{abstract}
-This is where the Abstract goes.
+% XXX better abstract
+This paper presents a simple, generic method for automating the play
+of Nintendo Entertainment System games.
 \end{abstract}
@@
 
-\section*{Introduction}
-The Nintendo Entertainment System is probably the best video game console, citation {\it not} needed.
-Like many, I have spent thousands of hours of my life playing NES games, including several complete playthroughs of classics like Super Mario Bros., Bionic Commando, Bubble Bobble, and other favorites.
-By the year 2013, home computers have become many orders of magnitude faster and more capacious than the NES hardware.
-This suggested to me that it may be time to automate the playing of NES games, in order to save time.\footnote{Rather, to replace it with time spent programming.} In this paper I present a generic technique for automating the playing of NES games.
-The approach is practical on a single computer, and succeeds on several games, such as Super Mario Bros..
-The approach is amusingly elegant and surprisingly effective, requires no detailed knowledge of the game being played, and is capable of novel and impressive gameplay (for example, bug exploitation).
-{\bf Disclaimer for SIGBOVIK audience: This work is 100\% real.}
+\section{Introduction}
+The Nintendo Entertainment System is probably the best video game
+console, citation {\it not} needed. Like many, I have spent thousands
+of hours of my life playing NES games, including several complete
+playthroughs of classics like Super Mario Bros., Bionic Commando,
+Bubble Bobble, and other favorites. By the year 2013, home computers
+have become many orders of magnitude faster and more capacious than
+the NES hardware. This suggested to me that it may be time to automate
+the playing of NES games, in order to save time.\footnote{Rather, to
+  replace it with time spent programming.} In this paper I present a
+generic technique for automating the playing of NES games. The
+approach is practical on a single computer, and succeeds on several
+games, such as Super Mario Bros.. The approach is amusingly elegant
+and surprisingly effective, requires no detailed knowledge of the game
+being played, and is capable of novel and impressive gameplay (for
+example, bug exploitation). {\bf Disclaimer for SIGBOVIK audience:
+  This work is 100\% real.}
 
-On a scale from ``the title starts with Toward'' to ``Donald Knuth has finally finished the 8\th\ volume on the subject,'' this work is a 3.
-The purpose of this paper is mainly as a careful record of the current status for repeatability and further development on this important research subject.
-A short video version of this paper is available for those that hate reading, at \verb+http://tom7.org/mario+, and is the more fun way to consume the results.
-This page also contains audiovisual material that makes this work more entertaining (for example, its output) and source code.
+On a scale from ``the title starts with Toward'' to ``Donald Knuth has
+finally finished the 8\th\ volume on the subject,'' this work is a 3.
+The purpose of this paper is mainly as a careful record of the current
+status for repeatability and further development on this important
+research subject. A short video version of this paper is available for
+those that hate reading, at \verb+http://tom7.org/mario+, and is the
+more fun way to consume the results. This page also contains
+audiovisual material that makes this work more entertaining (for
+example, its output) and source code.
 
-The basic idea is to deduce an objective function from a short recording of a player's inputs to the game.
-The objective function is then used to guide search over possible inputs, using an emulator.
-This allows the player's notion of progress to be generalized in order to produce novel gameplay.
-A design goal is that the objective function be amusingly elegant (not at all smart, fancy, or customized to the game) in order to demonstrate that the game is reducible to such a simple objective.
-The search needs to be game-agnostic and practical, but since the space is exponential ($256^{n}$), we need to be smart here.
+The basic idea is to deduce an objective function from a short
+recording of a player's inputs to the game. The objective function is
+then used to guide search over possible inputs, using an emulator.
+This allows the player's notion of progress to be generalized in order
+to produce novel gameplay. A design goal is that the objective
+function be amusingly elegant (not at all smart, fancy, or customized
+to the game) in order to demonstrate that the game is reducible to
+such a simple objective. The search needs to be game-agnostic and
+practical, but since the space is exponential ($256^{n}$), we need to
+be smart here.
 
@@
 \subsection{The NES hardware and emulation}
-The NES is based around an 8-bit processor running at 1.79~MHz, the Ricoh 2A03.
-8 bits is really small.
-You can see them all right here: 00001111.
-It's no coincidence that each controller also has 8 buttons: Up, Down, Left, Right, Select, Start, B and A.
-It has only 2048 bytes of general purpose RAM.
-(There is also some special purpose RAM for graphics, which we ignore in this work.) 2048 bytes is really small.
-You can see them all in Figure~\ref{fig:bytes2048}.
-As a result, NES programs are written to use memory efficiently and straightforwardly; usually there are fixed memory locations used for all the critical game facts like the player's health, number of lives, coordinates on the screen, and so on.
-For example, in Super Mario Bros., the single byte at location \verb+0x757+ contains the number of lives the player has.
-The location \verb+0x75F+ contains the current world, and \verb+0x760+ the current level.
-The NES outputs 60.0988 frames per second, which we will just call 60 in this paper.
+The NES is based around an 8-bit processor running at 1.79~MHz, the
+Ricoh 2A03. 8 bits is really small. You can see them all right here:
+00001111. It's no coincidence that each controller also has 8 buttons:
+Up, Down, Left, Right, Select, Start, B and A. It has only 2048 bytes
+of general purpose RAM. (There is also some special purpose RAM for
+graphics, which we ignore in this work.) 2048 bytes is really small.
+You can see them all in Figure~\ref{fig:bytes2048}. As a result, NES
+programs are written to use memory efficiently and straightforwardly;
+usually there are fixed memory locations used for all the critical
+game facts like the player's health, number of lives, coordinates on
+the screen, and so on. For example, in Super Mario Bros., the single
+byte at location \verb+0x757+ contains the number of lives the player
+has. The location \verb+0x75F+ contains the current world, and
+\verb+0x760+ the current level. The NES outputs 60.0988 frames per
+second, which we will just call 60 in this paper.
 
-% (32kB of ROM is addressable at a time, and most games used bank switching to access hundreds of kilobytes of game included in the cartridge.)
+% (32kB of ROM is addressable at a time, and most games used bank
+% switching to access hundreds of kilobytes of game included in the
+% cartridge.)
 
-There are a number of emulators for NES.
-These work by simulating the NES hardware, for example with a 2048-byte array for its memory, and simulating the steps of its 2A03 processor on some ROM, and hooking a keyboard or joystick into the 8 bits of input.
-(There are of course many details to work out!
-But in essence emulation is jut that.) This process is completely deterministic, so it is possible to record the sequence of inputs (the inputs can only be read once per video frame, so this sequence is 60 bytes per second) and play them back and get the same result.
-This also means that an input sequence can be computed in non-real time, either much slower or much faster than a NES would normally run.
-In this work we use the FCEUX emulator, which is popular for its accuracy and advanced tools.
+There are a number of emulators for NES. These work by simulating the
+NES hardware, for example with a 2048-byte array for its memory, and
+simulating the steps of its 2A03 processor on some ROM, and hooking a
+keyboard or joystick into the 8 bits of input. (There are of course
+many details to work out! But in essence emulation is just that.) This
+process is completely deterministic, so it is possible to record the
+sequence of inputs (the inputs can only be read once per video frame,
+so this sequence is 60 bytes per second) and play them back and get
+the same result. This also means that an input sequence can be
+computed in non-real time, either much slower or much faster than a
+NES would normally run. In this work we use the FCEUX\cite{FCEUX}
+emulator, which is popular for its accuracy and advanced tools.
 
@@
 
-Bytes in memory (and sometimes 16- and 32-bit words) can contain interesting game facts like the player's position in the level or score.
-The central idea of this paper is to use (only) the value of memory locations to deduce when the player is ``winning''.
-The things that a human player perceives, like the video screen and sound effects, are completely ignored.
-As an additional simplification, we assume that winning always consists of a value {\it going up}---either the position in the level getting larger, the score getting larger, the number of lives, the world or level number getting bigger, etc.
+Bytes in memory (and sometimes 16- and 32-bit words) can contain
+interesting game facts like the player's position in the level or
+score. The central idea of this paper is to use (only) the value of
+memory locations to deduce when the player is ``winning''. The things
+that a human player perceives, like the video screen and sound
+effects, are completely ignored. As an additional simplification, we
+assume that winning always consists of a value {\it going up}---either
+the position in the level getting larger, the score getting larger,
+the number of lives, the world or level number getting bigger, and so
+on.
 
-This is actually a little bit too naive; for example, Mario's overall progress through the game is represented by a pair.
-You start in World 1-1 and the underground level that comes next is World 1-2 (we'll call this $w=1$ and $\ell=2$).
-But after you discover the princess is in another castle in World 1-4, the next level is 2-1.\footnote{In case you never realized this, it is time to learn that the legendary ``Minus World'' of {\tt\, -1} is not actually a negative world, but World 36-1 being incorrectly rendered because there is no glyph for the 36\th\ digit.
-The trick used to get to the Minus World just happens to leave the value 36 in that memory location rather than initializing it to a useful value.
-The ROM does not contain data for world 36 so it just interprets garbage data as a description of the world.} This can't be represented as a single byte going up (sometimes the second part $\ell$ goes down when we get to a new first part $w$), but it can be represented as a lexicographic order on the pair $\langle w, \ell \rangle$; that is, $\langle w_1, \ell_1 \rangle < \langle w_2, \ell_2$ if $w_1 = w_2$ and $\ell_1 < \ell_2$, or if $w_1 < w_2$ no matter the values of $\ell_1$ and $\ell_2$.
-This matches our intuitive idea and is also mathematically nice.
-It also generalizes multi-byte encodings of things like your score (which can be larger than 8 bits and so is often stored in 16 or 32), including both big-endian and little-endian representations.\footnote{A possible additional simplification would be to just take lexicographic orderings over bits, which then generalizes to 8-bit bytes.
-This is probably too crazy, but just right now I am sort of feeling like maybe I should try it, though it may be the beer.}
+This is actually a little bit too naive; for example, Mario's overall
+progress through the game is represented by a pair. You start in World
+{\tt 1-1} and the underground level that comes next is World {\tt 1-2}
+(we'll call this $w=1$ and $\ell=2$). But after you discover the
+princess is in another castle in World {\tt 1-4}, the next level is
+{\tt 2-1}.\footnote{In case you never realized this, it is time to learn
+  that the legendary ``Minus World'' of {\tt\, -1} is not actually a
+  negative world, but World {\tt 36-1} being incorrectly rendered because
+  there is no glyph for the 36\th\ digit. The trick used to get to the
+  Minus World just happens to leave the value 36 in that memory
+  location rather than initializing it to a useful value. The ROM does
+  not contain data for world 36 so it just interprets garbage data as
+  a description of the world.} This can't be represented as a single
+byte going up (sometimes the second part $\ell$ goes down when we get
+to a new first part $w$), but it can be represented as a lexicographic
+order on the pair $\langle w, \ell \rangle$; that is, $\langle w_1,
+\ell_1 \rangle < \langle w_2, \ell_2 \rangle$ if $w_1 = w_2$ and $\ell_1 <
+\ell_2$, or if $w_1 < w_2$ no matter the values of $\ell_1$ and
+$\ell_2$. This matches our intuitive idea and is also mathematically
+nice. It also generalizes multi-byte encodings of things like your
+score (which can be larger than 8 bits and so is often stored in 16 or
+32), including both big-endian and little-endian
+representations.\footnote{A possible additional simplification would
+  be to just take lexicographic orderings over bits, which then
+  generalizes to 8-bit bytes. This is probably too crazy, but just
+  right now I am sort of feeling like maybe I should try it, though it
+  may be the beer.}
 
-More importantly, it allows the combination of semantically unrelated bytes, like: $\langle$ world, level, screen inside the world, $x$ position on the screen $\rangle$ or $\langle$ world, lives, low byte of score $\rangle$.
-Many orderings may describe gameplay.
-These orderings may be temporarily violated in normal play: Although the score always goes up, Mario's $x$ position may temporarily decrease if he needs to navigate around some obstacle.\footnote{Note to self: Maybe we should give a much higher score to globally preserved objectives than to locally preserved ones.
-But that may presuppose that the input represents a whole playthrough?} So, to ``faithfully'' represent gameplay, we will generate a set of lexicographic orderings on memory locations, with the idea that they ``generally go up'' but not necessarily at every step.
-These orderings will also have weights.
-The next section describes how we take a sequence of player inputs and deduce the orderings.
+More importantly, it allows the combination of semantically unrelated
+bytes, like: $\langle$\!\!\!\! world, level, screen inside the world,
+$x$ position on the screen \!\!$\rangle$ or $\langle$\!\!\!\! world,
+lives, low byte of score \!\!$\rangle$. Many orderings may describe
+gameplay. These orderings may be temporarily violated in normal play:
+Although the score always goes up, Mario's $x$ position may
+temporarily decrease if he needs to navigate around some
+obstacle.\footnote{Note to self: Maybe we should give a much higher
+  score to globally preserved objectives than to locally preserved
+  ones. But that may presuppose that the input represents a whole
+  playthrough?} So, to ``faithfully'' represent gameplay, we will
+generate a set of lexicographic orderings on memory locations, with
+the idea that they ``generally go up'' but not necessarily at every
+step. These orderings will also have weights. The next section
+describes how we take a sequence of player inputs and deduce the
+orderings.
 
 \subsection{Deriving the objective function} \label{sec:deriving}
+
+\begin{figure}[h!tb]
+\begin{center}
+\includegraphics[width=0.95 \linewidth]{mario-global}
+\end{center}\vspace{-0.1in}
+\caption[lof not used]{
+  A single maximal tight valid lexicographic ordering for my
+  4,000-frame input training data to Super Mario Bros.. This function
+  is globally nondecreasing, and is the decimal memory locations
+  $\langle$\!~232, 57, 73, 74, 75, 115, 130, 155, 32, 184, 280,
+  491, 506, 1280, 1281, 1282, 1283, 1288, 1290, 1337, 1338, 1339,
+  1384, 1488, 1490, 1496, 1497, 1498, 1499, 1514, 1873, 1882, 1888,
+  1904, 1872, 1906, 112, 113, 114, 2009, 2010, 2011, 1539~\!$\rangle$.
+
+  This is not a great objective function; there are long spans where
+  all the memories are equal according to it, and the nice smooth slopes
+  are happening during level transitions where the game is ignoring
+  inputs (they are probably timers counting up each frame, which is
+  why they are so smooth). Other slicing produces better objectives.
+
+  For reasons unknown---I just discovered this while generating the
+  figure---{\em all} of the objective functions learned with this
+  method, regardless of the nondeterministic choices, appear to have
+  this same curve, despite using different memory locations. It may be
+  that they are different permutations of bytes that all change
+  simultaneously, only on the frames where there are jumps in this
+  picture, and there are no other orderings that are tight, valid, and
+  maximal. This is still surprising and warrants investigation. }
+\label{fig:global}
+\end{figure}
+
+\begin{figure}[htb]
+\begin{center}
+\includegraphics[width=0.95 \linewidth]{mario-tenths}
+\end{center}\vspace{-0.1in}
+\caption{Ten objective functions trained on different tenths of the
+  4,000 inputs for Super Mario Bros.. These functions are normalized
+  against the range of all values they take on during the movie; you
+  can see that most are increasing locally most of the time, but many
+  drop back to zero around the 3100\th\ frame, when Mario reaches
+  world {\tt 1-2}. Within its 400-frame window, each objective is
+  guaranteed to be nondecreasing.}
+\label{fig:tenths}
+\end{figure}
+
+\begin{figure}[htb]
+\begin{center}
+\includegraphics[width=0.95 \linewidth]{mario-strides}
+\end{center}\vspace{-0.1in}
+\caption{Ten objective functions trained on every 100\th\ memory,
+  starting at frame 0, frame 1, and so on up to frame 10. Normalized
+  as usual. These objectives exhibit excellent global progress, but
+  are locally very noisy. Among the methods I used to generate
+  objectives, this one produces the best results (based on my
+  intuition about what a good objective function looks like).
+}
+\label{fig:strides}
+\end{figure}
 
@@
   they are truly uninitialized, or retain the values from the last
-  reset or game inserted. Some semi-famous tricks involve removing
-  and inserting game cartridges while the system is running in order
-  to take advantage of this.}
+  reset or game inserted. Some semi-famous tricks involve removing and
+  inserting game cartridges while the system is running in order to
+  take advantage of this.}
 
-% XXXX good place for some graphs.
-{\bf Slicing.}\,\, I generate 50 orderings for $M_1 \ldots M_n$; the
-whole recording starting from the first keypress. During gameplay some
-values really are completely nondecreasing, like the player's score
-and world/level pair. I also generate 3 orderings for each tenth of
-the memory sequence, e.g. $M_1 \ldots M_{n/10}$ and $M_{n/10+1} \ldots
-M_{2n/10}$, etc. The intention is to capture orderings that are rarely
-violated, or sections of the game with unusual objectives (e.g. a
-minigame or swimming level). Orderings generated this way look pretty
-random, and on a scale from solid to suspicious, I can't vouch for
-them. Then I generate objectives from non-consecutive memories that
-are evenly spread out through the observations: Ten objectives
-chosen from every 100\th\ memory, starting from the 0\th\ frame,
-1\st\ frame, 2\nd\ frame, and so on up to the 9\th{}. Similarly
-for every 250\th\ frame, and a single objective for memory sliced
-to every 1000\th\ frame, with start positions of 0--9.
+\paragraph{Slicing.} I generate 50 orderings for $M_1 \ldots M_n$; the
+whole recording starting immediately after the first keypress. During
+gameplay some values really are completely nondecreasing, like the
+player's score and world/level pair. Figure~\ref{fig:global} shows
+what a global ordering looks like. I also generate 3 orderings for
+each tenth of the memory sequence, e.g. $M_1 \ldots M_{n/10}$ and
+$M_{n/10+1} \ldots M_{2n/10}$, etc. The intention is to capture
+orderings that are rarely violated, or sections of the game with
+unusual objectives (e.g. a minigame or swimming level). Orderings
+generated this way look pretty random, and on a scale from solid to
+suspicious, I can't vouch for them. Then I generate objectives from
+non-consecutive memories that are evenly spread out through the
+observations: Ten objectives chosen from every 100\th\ memory,
+starting from the 0\th\ frame, 1\st\ frame, 2\nd\ frame, and so on up
+to the 9\th{}. Similarly for every 250\th\ frame, and a single
+objective for memory sliced to every 1000\th\ frame, with start
+positions of 0--9. The intention is to capture objectives that grow
+slowly over the course of a playthrough, without getting distracted by
+local noise.
 
-{\bf Weighting.},\, To reduce the importance of randomness in the
+\paragraph{Weighting.} \label{sec:objectiveweighting} 
+To reduce the importance of randomness in the
 learning process, and the arbitrariness of the slicing, each objective
@@
 all the values we ever saw for it. The value fraction is defined and
-in $[0, 1]$ for all memories in the observed set.\footnote{It is not
-  defined when the memory is greater than all observed memories, which
-  we will encounter later. The code returns $(i+1)/|V|$ in that case,
-  which is as good as anything.} This gives us the ability to compare
-objectives on an absolute scale.\footnote{This is certainly not the
-  only way to do it, and it has some questionable properties like
-  ignoring the magnitude of change. But it is very simple.} Weighting
-an objective is now simple:
+in $[0, 1)$ for all memories in the observed set.\footnote{It is not
+    defined when the memory is greater than all observed memories,
+    which we will encounter later. The code returns $|V|/|V| = 1$ in
+    that case, which is as good as anything.} This gives us the
+  ability to compare objectives on an absolute scale.\footnote{This is
+    certainly not the only way to do it, and it has some questionable
+    properties like ignoring the magnitude of change. But it is very
+    simple.} Weighting an objective is now simple:
 %
@@
 memories. In the case of the ideal function this is $\Sigma_{i=1}^{n-1}
-1/n$, which approaches $1$.\footnote{The code implements a different
-  weighting scheme, which sums from $i=0$ and keeps track of the last
-  value fraction at each step, thus treating the value fraction of the
-  nonexistent memory $M_0$ as 0. This is wrong, because the first
-  value fraction may be very high, which credits the objective with a
-  positive value (e.g. 1) for that first part of the sum. Objectives
-  that start high on the first frame are not ideal; in fact, the worst
-  objectives start high on the first frame and steadily {\em
-    decrease}. I plan to correct this bug and study if it has any
-  noticeable consequences.} Objectives are constrained to have
-non-negative weights (I set the value to 0 if negative, which
-effectively disables it).
+1/n$, which approaches $1$. Of course, when you think about it,
+this is actually the same as
+%
+\[
+\begin{array}{c}
+({\tt VF}(M_1) - {\tt VF}(M_0)) + ({\tt VF}(M_2) - {\tt VF}(M_1)) \,+ \\
+\vdots \\
+({\tt VF}(M_{m-1}) - {\tt VF}(M_{m-2})) + ({\tt VF}(M_{m}) - {\tt VF}(M_{m}))
+\end{array}
+\]
+%
+and all but $-{\tt VF}(M_0)$ and ${\tt VF}(M_m)$ cancel out. This
+means that the weight we use is just the final value minus the initial
+value, regardless of what happens in-between.\footnote{I think this
+  can be improved, for example by taking the deviation from the ideal
+  linear objective.} The mean value theorem or something is probably
+relevant here. {\bf Lesson about being careful:} I only realized that
+this code was kind of fishy when I started writing it up for SIGBOVIK.
+Not only did it loop over all the deltas as in the $\Sigma$ expression
+above, but it also summed from $i=0$ and kept track of the last value
+fraction at each step, thus treating the value fraction of the
+nonexistent memory $M_0$ as 0. This is wrong, because the first value
+fraction may be very high, which credits the objective with a positive
+value (e.g. 1) for that first part of the sum. Objectives that start
+high on the first frame are not ideal; in fact, the worst objectives
+start high on the first frame and steadily {\em decrease}. After
+writing this up I corrected it to the simple expression ${\tt VF}(M_m)
+- {\tt VF}(M_0)$ and the result was a huge breakthrough in the
+quality of the output! I had spent much time tuning the {\tt
+  playfun} search procedure (Section~\ref{sec:playfun}) and not
+investigated whether the objectives were being weighted properly. More
+on this later, but the lesson is: Bugs matter, and thinking about your
+code and explaining it is a good way to find bugs.
 
-% XXX good place to give some weights for some objectives.
-
-We save the objectives and their weights to a file and then we are
-done with the easy part.
+Objectives are constrained to have non-negative weights (I set the
+value to 0 if negative, which effectively disables it). We save the
+objectives and their weights to a file and then we are done with the
+easy part.
 
@@
   excellent emulator to use, has fancy tools for recording and editing
-  movies, and is popular in the speedrun community; just don't read
-  the code.} With a medium amount of suffering I got the emulator
-compiling under mingw in 64-bit mode, and working behind a streamlined
-interface ({\tt emulator.h}). Once a game is initialized, it is always
-at an input frame---you can give it an 8-bit input (for the
-1\st\ player) to step a single frame, or read the 2048 bytes of
-memory. You can also save the complete state of the emulator into a
-vector of bytes, which allows you to restore the state to exactly that
-same point.\footnote{This contains the RAM, but also stuff we didn't
-  consider, like the Picture Processing Unit's state, and internal
-  emulator stuff.} These save-states are portable across sessions as
-long as the code was compiled and the game initialized the right
-way.\footnote{The original FCEUX supports portable save-states, but I
-  removed that guarantee in order to get the smallest possible byte
-  vectors. More on that below.} FCEUX must be single-threaded because
-it uses global variables galore. I made several enhancements to the
-emulator interface, which are discussed later.
+  movies, and is popular in the speedrun community. I highly recommend
+  it; just don't read the code.} With a medium amount of suffering I
+got the emulator compiling under mingw in 64-bit mode, and working
+behind a streamlined interface ({\tt emulator.h}). Once a game is
+initialized, it is always at an input frame---you can give it an 8-bit
+input (for the 1\st\ player) to step a single frame, or read the 2048
+bytes of memory. You can also save the complete state of the emulator
+into a vector of bytes, which allows you to restore the state to
+exactly that same point.\footnote{This contains the RAM, but also
+  stuff we didn't consider, like registers, the Picture Processing
+  Unit's state, and internal emulator stuff.} These save-states are
+portable across sessions as long as the code was compiled and the game
+initialized the right way.\footnote{The original FCEUX supports
+  portable save-states, but I removed that guarantee in order to get
+  the smallest possible byte vectors. More on that below.} FCEUX must
+be single-threaded because it uses global variables galore. I made
+several enhancements to the emulator interface, which are discussed
+later.
 
 It's important to know that almost all the CPU time in all the
-algorithms discussed in this paper is spent emulating NES frames;
-it takes about XXX ms to process a single step. Lots of engineering
+algorithms discussed in this paper is spent emulating NES frames; it
+takes about 500\textmu s to process a single step. Lots of engineering
 effort goes into reducing the number of frames the algorithms emulate.
+The {\tt playfun} program takes a ROM file, the learned objectives and
+motifs, and runs on the console for arbitrarily long, outputting a
+movie file consisting of the input sequence it think is best. The
+current {\tt playfun} algorithm is much too slow to run real-time, but
+it would be possible to have video output as it played. I disabled
+most of the video code, however, in an effort to make the emulation
+loop run as fast as possible.
 
-\subsection{Naive attempts}
+\subsection{Naive attempts} \label{sec:naive}
 
@@
 if we don't know how to reconstruct a rare good state that caused the
-maximum score to be high. Both lead us to the idea: Keep around
-a set of candidate futures that we use for evaluation and search,
-rather than just randomly generating futures when we want to evaluate
-a state. This turns out to work really well and be more efficient.
+maximum score to be high. Both lead us to the idea: Keep around a set
+of candidate futures that we use for evaluation and search, rather
+than just randomly generating futures when we want to evaluate a state
+and then discarding them. This turns out to work really well and be
+more efficient.
 
-HERE
+The basic version of the {\tt playfun} algorithm looks like this.
+Maintain {\tt NFUTURES} futures (this is 40 for the results presented
+in Section~\ref{sec:results}), originally just seeded with weighted
+random motifs. We aren't likely to execute any of these futures
+verbatim, but they are intended to give us high watermarks of what
+we're capable of, plus allow us to judge future danger. As we execute
+search, futures will be partly consumed and extended, and some discarded
+and replaced.
 
-\subsection{Caching of emulation}
+Each future stores a desired length from 50--800 frames, and whenever
+it is shorter than that, we extend it (at its end) with random weighted
+motifs. The inner pseudoloop then looks like this:
 
-% good parts!
-% pacman @6800
+\begin{code}
+for (;;) \{
+  vector<uint8> before = GetMemory();
+  vector<uint8> state = GetState();
 
-\section{Results}
+ \
+
+  set<vector<uint8>> nexts;
+  for (Future f : futures) \{
+   nexts.insert(f.First10Frames());
+   f.ChopOffFirst10Frames();
+  \}
+
+ \
+
+  while (nexts.size() < NFUTURES)
+    nexts.push\_back(/* random motif */);
+
+ \
+
+  for (vector<uint8> &next : nexts) \{
+    RestoreState(state);
+    for (uint8 i : next) Step(i);
+    double score =
+      ScoreByFutures(before, futures);
+    // Save the best-scoring next...
+  \}
+
+ \
+
+  ReweightMotifs(best\_next, motifs);
+  ReplaceBadFutures(futures);
+  ExtendFuturesToDesiredLengths(futures);
+\}
+\end{code}
+
+At each iteration of the loop we will find the best next 10-frame
+sequence to commit to. Rather than search over all motifs, we search
+over the first 10 frames of each future. This has several important
+consequences. First, since futures are generated by weighted motifs,
+it makes it more likely that we spend CPU time evaluating motifs that
+are common; the code from Section~\ref{sec:naive} always explores
+every motif, even rare ones. Second, it guarantees that if we pick
+some 10 frames on the basis of a single good future, we don't have to
+worry about recreating that future; it will still be among our futures
+for the next round. This is key: It allows us to use the determinism
+of the game to replay a really good future if we find one, not just
+use average random futures to assess how good a state will be.
+
+The function {\tt ScoreByFutures} saves the state, then runs each
+of the {\tt NFUTURES} futures to get a final state and memory. We
+score each final memory relative to the start memory. The particulars
+of scoring are not as interesting as the general idea, which is:
+\begin{itemize}
+\item The potential 10-frame {\tt next} sequence that we're
+evaluating gets an {\em optimistic} score. This is based on the futures
+for which the objectives {\em go up}. This is always non-negative.
+\item Each future is also given a score, which is the sum of scores
+from all the different {\tt next} sequences, regardless of their sign.
+\end{itemize}
+
+The purpose of the first is to identify a good {\tt next} sequence. We
+take the sequence with the highest optimistic score. The idea is that
+this is the sequence that gives us the best outcome if we continue
+to control what we do in the future.
+
+The purpose of the second is to identify which futures are good. A
+good future tends to bring good outcomes regardless of the immediate
+next step. Random futures that make is walk left when the good stuff
+is to the right, or jump when there are spikes nearby above our head,
+will receive negative scores for many or all of the {\tt next}
+sequences.
+
+{\tt ReweightMotifs} changes the weight of motifs that match the best
+{\tt next} sequence that we just committed to, if we think that we
+made progress on this step. The idea is to learn which sequences tend
+to work well for us; this might be different from what the human
+player did. For example, in run-to-the-right-and-jump kinds of games,
+human players tend to hesitate more before obstacles and enemies than
+a computer does.\footnote{That said, this part was an early idea and
+  is probably not necessary. It's suspicious because these 10-frame
+  sequences are not necessarily motifs (10 is the same as the normal
+  motif length, and futures are generated by motifs, but they can
+  become desynchronized because of the final incomplete motif, future
+  mutation, and other things). So sometimes the chosen sequence
+  doesn't affect weights. I think this would be better if we kept a
+  Markov model } Knowing whether we made progress is kind of
+difficult, since we can almost always find some move that increases
+the objectives locally. For this I use the same approach of {\em value
+  fraction} from Section~\ref{sec:objectiveweighting} based on a
+sample of memories that we've seen so far. If we appear to be
+progressing then the motif is weighted up; if we're regressing then it
+is weighted down.
+
+{\tt ReplaceBadFutures} kicks out the the futures with the worst total
+scores, so that over time the random futures become good futures. Of
+course, we always have to keep randomly adding to the end of each
+future, and who knows what crap we'll find? A late addition to the
+algorithm replaces some proportion of these with mutated versions of
+the best scoring future. This helps us from messing up a good future
+by appending crap to it, since we have multiple copies of it. It also
+makes search more efficient, since futures that share common prefixes
+can often benefit from caching (Section~\ref{sec:caching}). Currently
+mutating a future involves chopping off its second half (it will get
+extended to the desired length in the next step), and a \sfrac{1}{7}
+chance of {\em dualizing} the entire sequence. Dualization swaps
+the left and right buttons, the up and down buttons, B and A, and 
+select and start. The idea of this is to occasionally introduce
+very different futures to help us get out of stuck situations where
+we should really be turning around or something like that.
+
+During the development and tuning of the algorithm, I sometimes
+observed {\tt ReweightMotifs} and {\tt ReplaceBadFutures} conspiring
+to create a total monoculture of futures, where the weight of a single
+motif (specifically ``press no buttons'') went out of control and all
+futures just consisted of waiting. Waiting is usually a local
+improvement to some objectives because of internal frame counters and
+the cursor used to control music playback. To guard against this,
+motifs have a maximum (and minimum) possible weight in {\tt
+  ReweightMotifs}, and {\tt ReplaceBadFutures} always ensure that
+some fraction of the futures are completely random (ignoring motif
+weights).
+
+\parameteralert{
+  This is really the worst part in terms of parameters. They are: 
+  {\tt NFUTURES}, the number of futures to maintain (40);
+  {\tt NWEIGHTEDFUTURES}, the number of futures that are weighted,
+  as opposed to totally random (35);
+  {\tt DROPFUTURES}, the number of the worst-scoring futures we
+  completely drop (5);
+  {\tt MUTATEFUTURES}, the number of worst-scoring futures that
+  we replace with mutants of the best future (7);
+  {\tt MINFUTURELENGTH} and {\tt MAXFUTURELENGTH}, which put bounds
+  on the sizes of futures (50 and 800);
+  {\tt OBSERVE\_EVERY}, the frequency with which we sample memory
+  for computing the value fraction for motif weighting (10);
+  {\tt ALPHA}, the multiplicative factor for up- or down-weighting
+  motifs (0.8);
+  {\tt MOTIF\_MIN\_FRAC} and {\tt MOTIF\_MAX\_FRAC}, the bounds
+  on motif weights (0.1 and 0.00001).
+
+  Each is a scarlet letter of personal shame and future work is to
+  eliminate them. In my opinion the least excusable are the bounds on
+  future length, since these are related to what portion of time is
+  ``interesting'' from a game perspective---for example, the max
+  future length must exceed the number of frames that transpire after
+  Mario dies but before he respawns, or the algorithm cannot properly
+  avoid death. In my opinion this requires too much knowledge of the
+  game. I don't have a good idea about how to compute this, at least
+  without a human providing an input movie of dying (which would
+  help for lots of other reasons)---but that would complicate the
+  output of the learning process, which violates a primary design goal.
+
+  {\tt NFUTURES} is another bothersome one, but there is more hope here.
+  Basically it trades off computation time for quality of play, and
+  more seems to always be better. We do have runtime metrics that could
+  be used to dynamically change NFUTURES. For example, we could use
+  the notion of improvability from Section~\ref{sec:backtracking}, or
+  the value fraction, the gauge the marginal value of searching an
+  additional future. Or something like that. This might actually help
+  performance because we do the same amount of work (modulo caching)
+  even during periods that the inputs are being ignored, like between
+  worlds in Super Mario Bros., as we do when the going gets rough,
+  and searching for the exact right delicate sequence of inputs would
+  benefit from more options. The appearance of pausing in the output
+  for Bubble Bobble (Section~\ref{sec:bubblebobble}) suggests that it
+  knows all the futures are bad and needs to search different ones,
+  and corroborates this idea.}
+
+\subsection{Backtracking} \label{sec:backtracking}
+
+The algorithm above always makes progress at the same rate of 10
+frames per iteration. Sometimes it can get stuck in local maxima.
+Super Mario Bros., my tuning game for this work, seems almost to have
+been designed to thwart {\tt playfun}. The first level is pretty easy
+once you have a decent algorithm, and early versions beat it without
+any trouble. It's mainly about avoiding Goombas and planning far
+enough ahead to make high jumps over pipes and pits (see the
+accompanying videos for amusing struggles with these). World {\tt 1-2}
+provides a challenge you may recognize from your youth, immediately
+followed by something that computers find much harder
+(Figure~\ref{fig:mario12}).
+
+I literally spent about six weekends on this problem. First, the
+overhang is set up so that enemies emerge from it just as you arrive.
+This means that loose play (imperfect enemy avoidance) tends to get
+you killed right here. Mario has found a number of different solutions
+to this problem, from waiting, to kicking turtles from the left to
+clear out the enemies, to timing his jump perfectly (it is possible)
+to stomp the turtle after bouncing his head off the ceiling.
+
+\begin{figure}[h!tb]
+\begin{center}
+\includegraphics[width=0.95 \linewidth]{mario12}
+\end{center}\vspace{-0.1in}
+\caption{This is where it starts to get hard, even with lexicographic
+ordering and time travel. This overhang is very tight in terms of the
+inputs that can get you through it; random play tends to kill you because
+of the advancing enemies that can't easily be jumped over. Greedy
+algorithms can't resist the ledge with the coins, but then need to
+turn around.}
+\label{fig:mario12}
+\end{figure}
+
+\subsection{Caching of emulation} \label{sec:caching}
+
+\subsection{MARIONET}
+
+\section{Results} \label{sec:results}
 
@@
 for Super Mario Bros.; that is, this was the result of just running
-the program the first time on a recording of me playing the game briefly.
+the program the first time on a recording of me playing the game
+briefly.
+
+\subsection{Super Mario Bros.}
 
@@
 
-% ends with 13 dots remaining
+Eventually, Pac-Man gets far enough away from the remaining dots that
+none of his futures bring him near, and without any other objective
+to seek out, runs into ghosts. He dies on the first level with 13 dots
+left.
 
@@
 
-\subsection{Bubble Bobble}
+\subsection{Bubble Bobble} \label{sec:bubblebobble}
 
@@
 level, bouncing on his own bubbles and so on, almost like a real
-player would. He beats XXX levels, more than I would have guessed.
-Since it takes two start button presses to enter the game, the second
-START is part of the input motifs. Amusingly, Bub uses this to his
-advantage to change the synchronization of his futures, and when the
-situation gets really tight, he'll pause the game until a good
-future comes around (Figure~\ref{fig:bubblebobble}). Paused states
-probably look like modest progress since memory locations like the
-music cursor are still advancing.
+player would. He beats more levels than my input movie did! Since it
+takes two start button presses to enter the game, the second START is
+part of the input motifs. Amusingly, Bub uses this to his advantage to
+change the synchronization of his futures, and when the situation gets
+really tight, he'll pause the game until a good future comes around
+(Figure~\ref{fig:bubblebobble}). Paused states probably look like
+modest progress since memory locations like the music cursor are still
+advancing.
 
-XXX what finally happens?
+After a harrowing escape from the ghost thing on level 4, Bub makes it
+to level 5 (my input movie quits at the beginning of level 4!), at
+which point I stopped the search because it was getting pretty pausey
+and I wanted to see some other games before the SIGBOVIK deadline.
 
@@
 \label{fig:bubblebobble}
+\end{figure}
+
+\subsection{Tetris}
+
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[width=0.95 \linewidth]{tetris}
+\end{center}\vspace{-0.1in}
+\caption{The joyful noise of objective functions learned for Tetris.
+  The first fifth of the movie is navigating menus and looks very
+  different from the remainder. There appear to be some frame counters
+  identified, which make perfect smooth progress throughout the entire
+  movie. I believe discontinuities correspond to placing of pieces;
+  some objectives monotonically increase at these points (and are noisy
+  in-between), suggesting that they incorporate the score.
+}
+\label{fig:tetris-objectives}
 \end{figure}

### 2013-03-17 trunk/tasbot/game.h
Section completed in paper

@@ -4,12 +4,17 @@
 
 /*
 #define GAME "mario"
-// #define MOVIE "mario-cleantom.fm2"
-#define MOVIE "mario-newtom2.fm2"
+#define MOVIE "mario-cleantom.fm2"
+// #define MOVIE "mario-newtom2.fm2"
 // #define FASTFORWARD 256  // XXX cheats! -- should be 0
 #define FASTFORWARD 0
 #define BASE64 "base64:jjYwGG411HcjG/j9UOVM3Q=="
 */
+
+#define GAME "tetris"
+#define MOVIE "tetristom2.fm2"
+#define FASTFORWARD 0
+#define BASE64 "base64:Ww5XFVjIx5aTe5avRpVhxg=="
 
 /*
 #define GAME "adventureisland"
@@ -22,9 +27,21 @@
 // #define MOVIE "karate-tom.fm2"
 // #define BASE64 "base64:6xX0UBv8pLORyg1PCzbWcA=="
 
-#define GAME "pacman"
-#define MOVIE "pacman-tom.fm2"
-#define FASTFORWARD 600
-#define BASE64 "base64:ymBr2Kh1o5bVJzXDu4T6Zw=="
+// #define GAME "pacman"
+// #define MOVIE "pacman-tom.fm2"
+// #define FASTFORWARD 600
+// #define BASE64 "base64:ymBr2Kh1o5bVJzXDu4T6Zw=="
+
+/*
+#define GAME "tetris"
+#define MOVIE "tetris-tom.fm2"
+*/
+
+/*
+#define GAME "bubblebobble"
+#define MOVIE "bubblebobble-tom.fm2"
+#define BASE64 "base64:Yp/O7FR2zdP6o3GJGQXqAg=="
+#define FASTFORWARD 0
+*/
 
 #endif

### 2013-03-17 trunk/tasbot/learnfun.cc

@@ -96,7 +96,10 @@ static void MakeObjectives(const vector< vector<uint8> > &memories) {
   // Some things will never violate the objective, like
   // [world number, stage number] or [score]. So generate
   // a handful of whole-game objectives.
-  
+
+  // TODO: In Mario, all 50 appear to be effectively the same
+  // when graphed. Are they all equivalent, and should we be
+  // accounting for that e.g. in weighting or deduplication?
   for (int i = 0; i < 50; i++) // was 10
     obj.EnumerateFullAll(PrintAndSave, 1, i);
 
@@ -162,6 +165,7 @@ int main(int argc, char *argv[]) {
 
   SaveMemory(&memories);
 
+  uint64 time_start = time(NULL);
   for (int i = start; i < movie.size(); i++) {
     if (i % 1000 == 0) {
       printf("  [% 3.1f%%] %d/%ld\n", 
@@ -172,8 +176,11 @@ int main(int argc, char *argv[]) {
     inputs.push_back(movie[i]);
     SaveMemory(&memories);
   }
+  uint64 time_end = time(NULL);
 
-  printf("Recorded %ld memories.\n", memories.size());
+  printf("Recorded %ld memories in %ld sec.\n", 
+         memories.size(),
+         time_end - time_start);
 
   MakeObjectives(memories);
   Motifs motifs;

### 2013-03-17 trunk/tasbot/playfun.cc

@@ -913,6 +913,13 @@ struct PlayFun {
 	  positive_scores + negative_scores;
       }
 
+      // TODO: I think maybe a better idea is to use the max over
+      // all futures? It's appropriate to be optimistic, but we
+      // shouldn't ignore components of the objective that go down,
+      // because we have to take them together. But integral_score
+      // probably dominates, here, so maybe positive_scores is a
+      // good tie-breaker?
+
       // Based on the idea that BuggyEvaluate only used the
       // positive_scores component. This 'next' doesn't have
       // to take the future if it has low score, so don't penalize
@@ -926,6 +933,7 @@ struct PlayFun {
 
       *futures_score += future_score;
 
+      // Unused except for diagnostics.
       if (future_score > *best_future_score)
 	*best_future_score = future_score;
       if (future_score < *worst_future_score)
@@ -1164,6 +1172,12 @@ struct PlayFun {
 	(*futures)[i].inputs.swap(newf);
       }
     }
+
+    // XXX: Don't drop the future if it was the one we got the
+    // max() score for. Right? It might have had very poor scores
+    // otherwise, but we might be relying on it existing.
+    // TODO: Consider duplicating the future that we got the max()
+    // score from.
 
     // Discard the futures with the worst total.
     // They'll be replaced the next time around the loop.

### 2013-03-17 trunk/tasbot/weighted-objectives.cc

@@ -318,10 +318,12 @@ void WeightedObjectives::SaveSVG(const vector< vector<uint8> > &memories,
 
     const string color = RandomColor(&rc);
     const string startpolyline =
-      StringPrintf("  <polyline fill=\"none\" opacity=\"0.5\" stroke=\"%s\""
+      StringPrintf("  <polyline fill=\"none\" "
+		   // opacity=\"0.5\" "
+		   "stroke=\"%s\""
 		   " stroke-width=\"%d\" points=\"", 
-		   (info.weight <= 0) ? "#f00" : "#0f0",
-		   // color.c_str(),
+		   // (info.weight <= 0) ? "#f00" : "#0f0",
+		   color.c_str(),
 		   1
 		   // (info.weight <= 0) ? 3 : 1
 		   );

### 2013-03-17 tasbot/paper/paper.tex
Section completed in paper

@@
 
-I literally spent about six weekends on this problem. First, the
-overhang is set up so that enemies emerge from it just as you arrive.
-This means that loose play (imperfect enemy avoidance) tends to get
-you killed right here. Mario has found a number of different solutions
-to this problem, from waiting, to kicking turtles from the left to
-clear out the enemies, to timing his jump perfectly (it is possible)
-to stomp the turtle after bouncing his head off the ceiling.
+I literally spent about six weekends and thousands of hours of CPU on
+this problem. First, the overhang is set up so that enemies emerge
+from it just as you arrive. This means that loose play (imperfect
+enemy avoidance) tends to get you killed right here. Mario has found a
+number of different solutions to this problem, from waiting, to
+kicking turtles from the left to clear out the enemies, to timing his
+jump perfectly (it is possible!) to stomp the turtle after bouncing his
+head off the ceiling. It's fun to see the different approaches and is
+a good benchmark for whether the search algorithm is actually doing a
+good job at tactics.
 
@@
 
+Immediately after this is an important test of longer-term planning.
+Most of my early solutions would jump up on this ledge with 4 coins,
+since the score is a component of many lexicographic
+orderings\footnote{Maybe even all of them, since it should be globally
+  obeyed; it's therefore a valid extension to any ordering that
+  doesn't already include it. It's not necessarily a tight extension,
+  however, so it may be excluded for that reason, or because during
+  initialization it is not actually zero and so not globally obeyed. I
+  never checked this stuff because I wanted to avoid any temptation to
+  overfit the learning algorithm to the particulars of any game.} and
+coins give you points. Down below is undesirable in the short term
+because of the enemies. Then Mario would feel stuck and just hump up
+against the wall jumping in this tiny dead end until time ran out. The
+game contains some bugs (you have probably seen them) that allow the
+screen to be scrolled without Mario actually progressing; these little
+mini scrolls, plus the givens like the frame counter and music cursor,
+prevent Mario from getting out of the local maximum. This is extremely
+frustrating. I decided to add some longer-term planning into the
+search procedure in order to try to help in these kinds of situations,
+as well as the occasional deaths I would see.
+
+Every {\tt CHECKPOINT\_EVERY} frames we save the state, and then every
+{\tt TRY\_BACKTRACK\_EVERY} rounds, we do a backtracking phase if we
+can. We just need to find a checkpoint at least {\tt
+  MIN\_BACKTRACK\_DISTANCE} frames in the past. Call that point
+{\tt start} and the current point {\tt now}. The backtracking
+routine looks like this:
+
+\begin{itemize}
+  \item Let {\tt improveme} be the inputs between {\tt start} and {\tt now}.
+  \item Get {\tt replacements}, a set of input sequences we might use
+    instead. These are usually based on {\tt improveme}.
+  \item Add {\tt improveme} to the set of replacements.
+  \item Truncate the movie to {\tt start}.
+  \item Now, do the normal playfun loop as though the (truncated) futures
+    are whatever our current futures are, and the set of {\tt next}
+    sequences are the {\tt replacements} array.
+  \item Whatever the best one among those is, keep it. Don't update motifs
+    or futures, however.
+\end{itemize}
+
+The idea is that we take a large sequence from our recent past, and the
+same futures we're already using, and see if that sequence can be improved,
+according to our objective functions, and using our same futures. Since
+the existing sequence is always one of those, if it is the best one,
+then the backtracking phase is a no-op. If we find something better, we
+slot it in and continue. So the only risk here is if our objective functions
+aren't good (we take as given that they are) and the only cost is time.
+
+Generating the candidate set of replacements uses a bunch of different
+techniques. They are:
+
+\paragraph{Random.} We generate a random sequence of the same length as
+the {\tt improveme} sequence.
+
+\paragraph{Opposites.} We dualize (swap left with right, up with down, start 
+with select, and B with A) and/or reverse random spans of the {\tt
+  improveme} sequence. The idea is to try to introduce some variety in
+cases where we're really getting stuck. This was specifically in hopes
+that Mario would walk off the coin ledge in {\tt 1-2} and then find that the
+course was clear below. I felt okay about this since this seems to be a
+generic notion (the buttons do have semantics that are common to most NES
+games where left and right are opposites), but that may just have been
+rationalization since I was getting desperate. It didn't work; see below.
+
+\paragraph{Ablation.} Randomly blanks out buttons from the input. For
+example, if we don't need to jump and jumping is slowing us down or
+something, this can remove that and make a small improvement to the
+sequence.
+
+\paragraph{Chop.} Removes random spans from the input. This iteratively
+removes spans as long as the previous span was an improvement (see
+below). This can cause the movie to get shorter by removing parts that
+weren't contributing anything.\footnote{However, in games where an
+  objective function includes something like a frame counter or the
+  music cursor, shorter sequences always score lower than
+  otherwise equivalent longer sequences.}
+
+We use the following formula to score a potential improvement:
+
+\begin{center}
+\includegraphics[width=0.95 \linewidth]{isimprovement}
+\end{center}
+
+The {\tt start} state is as discussed, {\tt old\_end} is the state at
+{\tt now}, and {\tt new\_end} is the state after the potential
+improvement sequence. $\int_s^o$ is the integral of score changes
+along the {\tt improveme} path and $\int_s^n$ along the candidate
+improvement. The integral is the weighted sum of the objectives
+increased minus the weighted sum of the objectives that decreased, at
+each step, all summed up. This works better than just computing the
+weighted score from e.g. {\tt start} to {\tt old\_end} when they are
+separated by many frames (typical backtrack distances are several
+hundred frames). The expression $n-o$\footnote{This is not numerical
+  minus but minus according to the set of objective functions. Just
+  roll with it.} is just the weighted sum of objectives that increased
+minus the weighted sum of objectives that decreased between the old
+end state and new end state; we have no option of computing the
+integral here because we don't have an input sequence that goes from
+the old end to the new end (and there almost certainly isn't one). We
+require that three conditions hold:
+\begin{enumerate}
+\item $\int_s^n \geq \int_s^o$,
+\item $\int_s^n > 0$
+\item $n - o > 0$
+\end{enumerate}
+
+If they all do, then the score for the improvement is 
+$\int_s^n - \int_s^o + (n - o)$, which is guaranteed to be positive.
+
+Even if we have many possible replacements, we try only the ones with
+the best scores; due to the particulars of the implementation there
+is not a simple constant describing this maximum, but it's around 220
+with the parameter settings used in this paper.
+
+\parameteralert{
+  Again! The humiliating appearance of constants! There are many here,
+  having to do with the number of potential replacements we use of
+  each type, the maximum number we keep, how often we bactrack, and
+  how far. I am not totally satisfied with how backtracking works (see
+  below), so I don't want to spend too much energy speculating about
+  how to reduce the parameter space here; I'd rather replace the
+  method wholesale.}
+
+\paragraph{Takes of woe, tales of joy.} 
+Backtracking was initially added in order to fix the mario coin ledge
+problem in {\tt 1-2}. It did not help with this. Older versions of the
+code would have Mario get past this spot, probably by luck, but in
+these he would get stuck in an embarrassing way on a small pit later,
+or jump straight into an enemy. Most of the approaches that actually
+looked solidly good elsewhere were failing badly here. As I started
+writing up this paper, on an airplane, I discovered the bug in the
+weighting of objective functions described in
+Section~\label{sec:objectiveweighting}. On my trip I had let {\tt
+  playfun} run with particularly high constants ({\tt NFUTURES} = 50,
+lots of backtracking candidates, etc.) and it had spent thousands of
+CPU hours playing to this point, grabbing the coins, and then dying of
+timeout, then losing all its lives, starting over, and getting to this
+point again! After fixing the bug, I tried again, and it was a huge
+breakthrough: Mario jumped up to get all the coins, and then almost
+immediately jumped back down to continue the level! On his first try
+he beat {\tt 1-2} and then immediately jumped in the pit at the
+beginning of {\tt 1-3} just as I was starting to feel suuuuuper smart
+(Figure~\ref{fig:mario13}). On a scale of OMFG to WTFLOL I was like
+{\em whaaaaaaat}?
+
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[width=0.95 \linewidth]{adventureisland}
+\end{center}\vspace{-0.1in}
+\caption{After making me feel so successful by finally cleanly
+  finishing world {\tt 1-2}, Mario immediately jumps into the first
+  trivial pit in {\tt 1-3}. I believe what's going on here is that
+  this is actually the best future, because resetting back a single
+  screen isn't a large loss, and probably the rest of the futures were
+  struggling to make progress. This level is very jumpy and probably
+  needs more futures to navigate its obstacles. In any case, Mario,
+  Touch\'e.}
+\label{fig:mario13}
+\end{figure}
+
+Seeing this huge improvement changed my idea about what part of the
+code needed work (I now believe that simpler search strategies might
+work and better lexicographic order generation and weighting is called
+for). But, this was already in the midst of writing up the paper, so
+instead I spent the CPU time running the current version on a bunch of
+games. Those results are described in Section~\ref{sec:results} and
+videos are available on \url{http://tom7.org/mario/}. In the next
+section I describe some of the performance improvements I made, and
+other modifications to the emulator interface.
+
+\section{Performance} \label{sec:performance}
+
 \subsection{Caching of emulation} \label{sec:caching}
@@
 \subsection{MARIONET}
+
+% better to run on n-1 cpus?
 

### 2013-03-17 tasbot/paper/paper.tex
Caching, compression, ...

@@
   \begin{quotation}
-  {\bf Parameter Alert!}\,\, #1
+  \noindent {\bf Parameter Alert!}\, #1
   \end{quotation}
@@
 of the advancing enemies that can't easily be jumped over. Greedy
-algorithms can't resist the ledge with the coins, but then need to
+algorithms can't resist the ledge with the coins, but then would need to
 turn around.}
@@
 
-If they all do, then the score for the improvement is 
-$\int_s^n - \int_s^o + (n - o)$, which is guaranteed to be positive.
+The integral has to be at least as good as before, the new integral
+has to be positive, and the new end state needs to look better than
+the old end state (the triangle inequality does not necessarily hold
+here). If they all do, then the score for the improvement is $\int_s^n
+- \int_s^o + (n - o)$, which is guaranteed to be positive.
 
@@
 weighting of objective functions described in
-Section~\label{sec:objectiveweighting}. On my trip I had let {\tt
+Section~\ref{sec:objectiveweighting}. On my trip I had let {\tt
   playfun} run with particularly high constants ({\tt NFUTURES} = 50,
@@
 \begin{center}
-\includegraphics[width=0.95 \linewidth]{adventureisland}
+\includegraphics[width=0.95 \linewidth]{mario13}
 \end{center}\vspace{-0.1in}
@@
   this is actually the best future, because resetting back a single
-  screen isn't a large loss, and probably the rest of the futures were
-  struggling to make progress. This level is very jumpy and probably
-  needs more futures to navigate its obstacles. In any case, Mario,
-  Touch\'e.}
+  screen isn't a large loss, lives aren't included in the objective
+  function, and probably the rest of the futures were struggling to
+  make progress. This level is very jumpy and probably needs more
+  futures to navigate its obstacles. In any case, Mario, Touch\'e.}
 \label{fig:mario13}
@@
 games. Those results are described in Section~\ref{sec:results} and
-videos are available on \url{http://tom7.org/mario/}. In the next
+videos are available at \url{http://tom7.org/mario/}. In the next
 section I describe some of the performance improvements I made, and
@@
 
+Performance is really important, both because the quality of the
+output is CPU-bound and because I am impatient. In its current state,
+running {\tt playfun} is an overnight affair; it takes about an hour
+of real time to produce 1000 output frames (16 seconds of gameplay) on
+a high-end desktop machine. Other than a trick that I didn't end up
+using, these performance ideas are not at all new, but they are
+documented here for completeness. I certainly spent lots of time on 'em!
+
 \subsection{Caching of emulation} \label{sec:caching}
 
+The most expensive part, by far, is running a step of emulation. It
+takes about 500\textmu s to emulate a single frame, though this can
+vary depending on what instructions are actually executed (recall that
+each frame corresponds to many many 2A03 instructions; there are
+approximately 16,000 clock cycles per frame!). Hardware like the
+Picture Processing Unit and sound chip are emulated as well, which may
+actually be less efficient than the actual hardware (for example, the
+sound hardware can implement filtering with passive physical
+components like capacitors, but FCEUX contains a finite impulse
+response filter running on IEEE floating point numbers). We want to
+avoid executing a step more than once, if at all possible.
+
+Caching is so canonical that some have called it (pejoratively) the
+only idea in Systems research. And it is what we do here. The emulator
+adds a call
+%
+\begin{code}
+void CachingStep(uint8 input);
+\end{code}
+%
+with exactly the same semantics as {\tt Step}. However, it first
+checks to see if this exact input has been executed on this exact
+start state before. If so, it just restores the cached result
+state instead of executing the step.
+
+I use the excellent CityHash function\cite{CityHash} as the hash code,
+and use a least-recently-used approach to clean out the hash table
+when it has more than a fixed {\em slop} amount more than the target
+size. States are large because they contain both the savestate of the
+before and after state. I have plenty of RAM, so I allow each process
+to store up to 100,000 states with 10,000 states of slop, which takes
+about 3 gigabytes per process.
+
+Caching adds a little bit of overhead (a single step is about 13\%
+slower, amortized), from saving the state, computing the hash code,
+copying the state into the table, and cleaning the hash table
+periodically. A cache hit is a zillion times faster than actually
+executing the step, however, so as long as the hit rate is more than
+13\%, it's worth doing. I use caching step in most places inside {\tt
+  playfun}, except when I know that the state can't have been computed
+yet. The {\tt playfun algorithm} is particularly suitable to caching:
+The set of potential next input sequences {\tt nexts} usually share
+several input prefixes, and we keep around futures for some time, so
+we can often end up evaluating the same one under the same conditions
+many times. Moreover, mutant futures usually share their first half,
+making them half price. A really important property of caching is that
+it's based on the state of memory but doesn't care about the actual
+sequence used to get there. This means that in cases where the input
+is ignored (Mario ignores the jump button most of the time while in
+the air, for example, and ignores all inputs between levels) we reach
+equivalent states and can use the cache if the next input is exactly
+the same. The ideal approach here would be to follow how the bits of
+the input are actually read by the code, and insert a more generic key
+into the hash table. For example, if we see that the code never even
+read the bit of the input corresponding to the up direction, then we
+can reuse the step for an input that matches all bits but the up bit!
+This of course requires a lot of mucking around in the internals,
+which on a scale of within the scope to beyond the scope of this
+article is a 9.9.
+
+{\bf Software engineering lesson:} Initial results from caching were
+disappointing and I thought the overhead was to blame. I made several
+low-level tweaks to avoid copying, etc., to reduce the overhead from
+36\% to 13\%. Then I discovered that there were 0 cache hits ever,
+because of a bug (I was inadvertantly using pointer equality on keys
+instead of value equality, so keys were never found). Check basic
+things about your code's correctness before going on an optimization
+soiree!
+
+\subsection{Space optimizations}
+
+Before I had the idea that is the actual subject of this paper, I was
+playing around with other emulator search ideas that involved storing
+very very large numbers of states. This necessitated minimal
+representations of savestates. FCEUX uses zlib internally to compress
+states, which works well; they shrink from something like 13,776
+bytes\footnote{I don't completely understand the NES architecture and
+  emulation strategy, but I believe some games include expansion chips
+  that require more space for save states. All this work is based on
+  the 2048 bytes of main NES RAM, as you already know. Perhaps clever
+  video game authors from the past can protect against this paper's
+  approach by storing important game facts inside expansion chips
+  in the cartridge.} to an average of 2266. I made some modifications
+to the save/load routines to make this as small as I could manage.
+I removed the backing buffer, which is only used for drawing graphics,
+movie info, header bytes that are constant or already known (savestate
+size), which shrunk the compressed size to an average of 2047.42 bytes.
+That meant I could fit about 32 million savestates in RAM at once, which
+is kind of amazing.
+
+Compression algoritms work well when there is lots of redundancy in
+their input. Comparing memories between two frames in some game,
+you'll often see only a few differences. Some of the memory contains
+stuff that's essentially read-only, even. To improve the
+compressibility of savestates, I introduce the idea of a {\em basis}
+state. This is any non-empty sequence of bytes which is an additional
+argument to the {\tt LoadState} and {\tt SaveState} functions. It's
+subtracted from the savestate prior to compression (if it's longer or
+shorter, it's treated as an infinite repetition of those bytes). If
+the basis is the same as the savestate, then the result is all zeroes,
+which compresses nicely (of course, now you need to have the basis in
+order to load the state, which is the same as the savestate, so that
+didn't save you much!). But memories during gameplay often don't
+differ much, so I save a state from any arbitrary frame in the game
+and then use that as the basis. This factors out any common parts
+and makes the states much more compressible: The average size drops
+to 1870.41 bytes.
+
+Using a better compression algorithm like the Burrows--Wheeler
+transform\cite{Burrows}\footnote{This is really one of the all-time
+  great algorithms. If you don't know it and you like this kind of
+  thing, you should check it out. It's so surprising and elegant.}
+would probably help too; zlip implements LZW which is only mediocre.
+However, on a scale of throwing my computer out the window to wanting
+to reimplement all software in my own private SML library, I just
+can't stand getting 3\rd-party libraries to compile and link into my
+applications, so I didn't even try. In any case, I abandoned this
+direction, which was not working well even with lean savestates.
+
+For {\tt playfun}, RAM is not a bottleneck at all. I disable
+compression completely on savestates, including the builtin zlib stuff
+(which is actually rather slow) and just use raw savestates. The
+removal of unnecessary headers and stuff is still there, and saves a
+few bytes and CPU cycles, but there was no reason to actually do it
+for this particular work. {\em But sometimes I do unnecessary things.}
+
 \subsection{MARIONET}
+
+The {\tt playfun} algorithm involves running 40 or so different
+10-sequence {\tt next} steps and then scoring them against {\tt
+  NFUTURES} different futures. Each {\tt next} and each future is
+completely independent and dominated by the cost of evaluating
+emulator steps. The ideal thing would be to use threads and shared
+memory to run these steps in parallel. As I mentioned earlier,
+FCEUX is hopelessly single-threaded.
+
+Eventually I realized I needed to search more states than I could in a
+single thread, and so I created MARIONET. It's a play on words, a
+double entendre of ``Mario network'' and ``Marionette'', which is
+obvious. This is a mode whereby {\tt playfun} can be started in
+``helper'' mode (listens on a port for work to do) or ``master'' mode
+(connects to a list of ports to run work on), in order to run
+many emulators in different processes.
 
@@
 \subsection{Super Mario Bros.}
+
+bug exploitation (Figure~\ref{fig:mario13bug1}).
+
+\begin{figure}[h!tb]
+\begin{center}
+\includegraphics[width=0.75 \linewidth]{mario13bug1a} \\[0.3em]
+\includegraphics[width=0.75 \linewidth]{mario13bug1b}
+\end{center}\vspace{-0.1in}
+\caption{Mario bounces off one Goomba up into the feet of another. Not
+  only doesn't he have enough velocity to reach the platform, but he's
+  about to hit that Goomba from below. Believe it or not, this ends
+  well: {\tt playfun} is happy to exploit bugs in the game; in this case,
+  that whenever Mario is moving downward (his jump crests just before the
+  Goomba hits him) this counts as ``stomping'' on an enemy, even if that
+  enemy is above him. The additional bounce from this Goomba also allows
+  him to make it up to the platform, which he wouldn't have otherwise!
+}
+\label{fig:mario13bug1}
+\end{figure}
 

### 2013-03-17 tasbot/paper/paper.tex
MARIONET section

@@
   \begin{quotation}
-  \noindent {\bf Parameter Alert!}\, #1
+  \noindent {\bf Parameter Alert!} #1
   \end{quotation}
@@
 Even if we have many possible replacements, we try only the ones with
-the best scores; due to the particulars of the implementation there
-is not a simple constant describing this maximum, but it's around 220
+the best scores; due to the particulars of the implementation there is
+not a simple constant describing this maximum, but it's around 220
 with the parameter settings used in this paper.
@@
   method wholesale.}
+
+The {\em improvability} of a state is the fraction of these potential
+improvements that are legal improvements, as above. Based on my
+analysis, states that are highly improvable ($> 30\%$) tend to be
+``worse'' (more stuck, closer to death, or whatever) than those that
+are not very improvable ($< 5\%$). This isn't being used for anything
+currently, but getting an absolute picture of how good a state is (as
+opposed to simply comparing it to an adjacent state) is one of the
+difficulties with this approach, so this may be a useful notion for
+future directions.
 
@@
   playfun} run with particularly high constants ({\tt NFUTURES} = 50,
-lots of backtracking candidates, etc.) and it had spent thousands of
-CPU hours playing to this point, grabbing the coins, and then dying of
-timeout, then losing all its lives, starting over, and getting to this
-point again! After fixing the bug, I tried again, and it was a huge
-breakthrough: Mario jumped up to get all the coins, and then almost
-immediately jumped back down to continue the level! On his first try
-he beat {\tt 1-2} and then immediately jumped in the pit at the
-beginning of {\tt 1-3} just as I was starting to feel suuuuuper smart
-(Figure~\ref{fig:mario13}). On a scale of OMFG to WTFLOL I was like
-{\em whaaaaaaat}?
-
-\begin{figure}[ht]
+lots of backtracking candidates, etc.) and it had spent about a
+thousand CPU hours playing to this point, grabbing the coins, and then
+dying of timeout, then losing all its lives, starting over, and
+getting to this point again! After fixing the bug, I tried again, and
+it was a huge breakthrough: Mario jumped up to get all the coins, and
+then almost immediately jumped back down to continue the level! On his
+first try he beat {\tt 1-2} and then immediately jumped in the pit at
+the beginning of {\tt 1-3} just as I was starting to feel suuuuuper
+smart (Figure~\ref{fig:mario13}). On a scale of OMFG to WTFLOL I was
+like {\em whaaaaaaat}?
+
+\begin{figure}[hb]
 \begin{center}
@@
 things about your code's correctness before going on an optimization
-soiree!
+sortie!
 
@@
 That meant I could fit about 32 million savestates in RAM at once, which
-is kind of amazing.
+would be kind of amazing if I could tell my 8 year-old self that.
 
@@
 argument to the {\tt LoadState} and {\tt SaveState} functions. It's
-subtracted from the savestate prior to compression (if it's longer or
-shorter, it's treated as an infinite repetition of those bytes). If
-the basis is the same as the savestate, then the result is all zeroes,
-which compresses nicely (of course, now you need to have the basis in
-order to load the state, which is the same as the savestate, so that
-didn't save you much!). But memories during gameplay often don't
-differ much, so I save a state from any arbitrary frame in the game
-and then use that as the basis. This factors out any common parts
-and makes the states much more compressible: The average size drops
-to 1870.41 bytes.
+subtracted from the savestate prior to compression (it doesn't have to
+be the same size; it's treated as an infinite repetition of those
+bytes). If the basis is the same as the savestate, then the result is
+all zeroes, which compresses nicely (of course, now you need to have
+the basis in order to load the state, which is the same as the
+savestate, so that didn't save you much!). But memories during
+gameplay often don't differ much, so I save a state from any arbitrary
+frame in the game and then use that as the basis for everything else.
+This factors out any common parts and makes the states much more
+compressible: The average size drops to 1870.41 bytes.
 
@@
   thing, you should check it out. It's so surprising and elegant.}
-would probably help too; zlip implements LZW which is only mediocre.
+would probably help too; zlib implements LZW which is only mediocre.
 However, on a scale of throwing my computer out the window to wanting
@@
 
-% better to run on n-1 cpus?
+The processes communicate over TCP/IP. I only run it on my local
+computer, but it would be reasonable to run it on a cluster or
+possibly even the Internet. I used SDL's SDL\_Net as a portable
+network interface in the hopes of keeping it platform-agnostic (FCEUX
+is basically portable so it would be a shame to make this
+Windows-specific, though I am not eager to try to get this thing to
+compile on other platforms, I gotta be honest). SDL is a nightmare to
+get working with mingw in a 64-bit compile, as usual, and SDL\_Net
+contained some bugs\footnote{In particular, the {\tt
+    SDLNet\_TCP\_Recv} call is supposed to block until it has received
+  the entire requested length, but it occasionally returns early.}
+that I had to spend some time working around. Anyway, I'm just
+complaining. For serializing requests and responses to bytes, I used
+Google's Protocol Buffer library\cite{ProtoBuf}, which is designed for
+this kind of thing.
+
+Helpers all start up with the same ROM and motif and objectives
+loaded, so that they can simulate whatever the master would
+compute.\footnote{Only the master reweights motifs, since it needs a
+  single sample of memories that we've observed. That means that in
+  cases where a helper generates random motifs, it does so with
+  respect to the original human weighting. There are some other small
+  things I simply avoided doing because they would require too much
+  coordination between processes or make the RPCs too large; MARIONET
+  was part of the system for much of its development.} They just wait
+on a port for the master to send a request, which can be either ``try
+this set of {\tt nexts} with these futures and send me the results'' or
+``try to find $n$ improvements of these inputs {\tt improveme} using
+so-and-so approach.'' In either case we send the base state as a
+serialized savestate.
+
+The master does the outer loop of the algorithm and all the writing to
+disk and so on. When it needs to do an expensive step like the inner
+loop, it prepares a set of jobs and sends them to workers using a little
+fork-join library ({\tt netutil.*}). The library keeps track of the
+outstanding jobs and which helpers are working, and feeds them work
+when they are idle, until all the jobs are complete. It's even got a
+color ANSI progress bar and stuff like that.
+
+Utilization is very good (Figure~\ref{fig:utilization}), and we get
+almost a linear speedup with the number of processes, up to the number
+of hardware threads the system supports (12). Actually, in addition
+to the computer remaining usable for other stuff, $n-1$ helpers seems
+to work a bit better than $n$, because each of the helpers is then able
+to avoid preemption. Since the driver loop is synchronous, having one
+laggard process slows the whole darn thing down.
+
+\begin{figure}[htb]
+\begin{center}
+\includegraphics[width=0.75 \linewidth]{utilization}
+\end{center}\vspace{-0.1in}
+\caption{Utilization with 12 helpers and one master on a 6-core (12
+  hyperthreads) processor. Looks good. Keeps the bedroom warm.}
+\label{fig:utilization}
+\end{figure}
+
+MARIONET is a huge improvement and was what enabled using 40+ futures
+in a reasonable amount of time, which is key to high-quality output. It's
+obviously the way you want to run the software, if you can get it to
+compile. There is one downside, though, which is that it reduces our
+ability to get lucky cache hits across {\tt nexts} that are similar (or
+have the same effect), and when replaying the winner as we commit to it.
+It's worth it, but smartly deciding which helper gets which tasks (because
+they share prefixes, for example) would save time. Running it all in one
+process with a shared memory cache would be the best, as long as the
+lock contention could be kept under control.
 
@@
 
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[width=0.95 \linewidth]{tetris-tower}
+\end{center}\vspace{-0.1in}
+\caption{Would you hire this construction company? Death is imminent,
+so {\tt playfun} pauses the game shortly after this and then doesn't
+unpause it.}
+\label{fig:tetris-tower}
+\end{figure}
+
+Tetris is a block dropping game, also known to the ancients. The
+Nintendo implementation is known for being inferior to the unlicensed
+Tengen version, which Nintendo tried to sue out of existence. Here I
+try to automate the Nintendo version as a tribute to litigation.
+Unsurprisingly, {\tt playfun} is awful at this game. Tetris has
+several screens of menus, where the player can select between
+different modes and theme musics and stuff like that. As an amusing
+prelude to this disaster in tetromino planning, {\tt playfun} keeps
+entering the menu and exiting back to the title screen rapidly, taking
+several seconds to even start the game. (Like in Bubble Bobble, this
+means that the start button is among the motifs.) Although the piece
+dropping looks more or less natural (but it's hard to not be, as the
+game drops the pieces for you), the placement is idiotic. This may be
+because placing a piece gives you a small amount of points, which
+probably looks like progress (Figure~\ref{fig:tetris-objectives}). As
+the screen fills, there's a tiny amount of tetris-like gameplay,
+probably as the end of the game becomes visible in some of the
+futures. The end result is awful and {\tt playfun} makes zero lines
+and certainly no Tetrises (Figure~\ref{fig:tetris-tower}). The only
+cleverness is pausing the game right before the next piece causes the
+game to be over, and leaving it paused. Truly, the only winning move
+is not to play.
+
 \section{Future Work}

### 2013-03-17 trunk/tasbot/paper/paper.tex
About Super Mario

@@
 \newcommand\parameteralert[1]{
-  % XXX this can get cooler.
-  {\bf Parameter Alert!}\,\, #1
+  \begin{quotation}
+  \noindent {\bf Parameter Alert!} #1
+  \end{quotation}
 }
@@
 \section{Introduction}
-The Nintendo Entertainment System is probably the best video game console, citation {\it not} needed.
-Like many, I have spent thousands of hours of my life playing NES games, including several complete playthroughs of classics like Super Mario Bros., Bionic Commando, Bubble Bobble, and other favorites.
-By the year 2013, home computers have become many orders of magnitude faster and more capacious than the NES hardware.
-This suggested to me that it may be time to automate the playing of NES games, in order to save time.\footnote{Rather, to replace it with time spent programming.} In this paper I present a generic technique for automating the playing of NES games.
-The approach is practical on a single computer, and succeeds on several games, such as Super Mario Bros..
-The approach is amusingly elegant and surprisingly effective, requires no detailed knowledge of the game being played, and is capable of novel and impressive gameplay (for example, bug exploitation).
-{\bf Disclaimer for SIGBOVIK audience: This work is 100\% real.}
-
-On a scale from ``the title starts with Toward'' to ``Donald Knuth has finally finished the 8\th\ volume on the subject,'' this work is a 3.
-The purpose of this paper is mainly as a careful record of the current status for repeatability and further development on this important research subject.
-A short video version of this paper is available for those that hate reading, at \verb+http://tom7.org/mario+, and is the more fun way to consume the results.
-This page also contains audiovisual material that makes this work more entertaining (for example, its output) and source code.
-
-The basic idea is to deduce an objective function from a short recording of a player's inputs to the game.
-The objective function is then used to guide search over possible inputs, using an emulator.
-This allows the player's notion of progress to be generalized in order to produce novel gameplay.
-A design goal is that the objective function be amusingly elegant (not at all smart, fancy, or customized to the game) in order to demonstrate that the game is reducible to such a simple objective.
-The search needs to be game-agnostic and practical, but since the space is exponential ($256^{n}$), we need to be smart here.
+The Nintendo Entertainment System is probably the best video game
+console, citation {\it not} needed. Like many, I have spent thousands
+of hours of my life playing NES games, including several complete
+playthroughs of classics like Super Mario Bros., Bionic Commando,
+Bubble Bobble, and other favorites. By the year 2013, home computers
+have become many orders of magnitude faster and more capacious than
+the NES hardware. This suggested to me that it may be time to automate
+the playing of NES games, in order to save time.\footnote{Rather, to
+  replace it with time spent programming.} In this paper I present a
+generic technique for automating the playing of NES games. The
+approach is practical on a single computer, and succeeds on several
+games, such as Super Mario Bros.. The approach is amusingly elegant
+and surprisingly effective, requires no detailed knowledge of the game
+being played, and is capable of novel and impressive gameplay (for
+example, bug exploitation). {\bf Disclaimer for SIGBOVIK audience:
+  This work is 100\% real.}
+
+On a scale from ``the title starts with Toward'' to ``Donald Knuth has
+finally finished the 8\th\ volume on the subject,'' this work is a 3.
+The purpose of this paper is mainly as a careful record of the current
+status for repeatability and further development on this important
+research subject. A short video version of this paper is available for
+those that hate reading, at \verb+http://tom7.org/mario+, and is the
+more fun way to consume the results. This page also contains
+audiovisual material that makes this work more entertaining (for
+example, its output) and source code.
+
+The basic idea is to deduce an objective function from a short
+recording of a player's inputs to the game. The objective function is
+then used to guide search over possible inputs, using an emulator.
+This allows the player's notion of progress to be generalized in order
+to produce novel gameplay. A design goal is that the objective
+function be amusingly elegant (not at all smart, fancy, or customized
+to the game) in order to demonstrate that the game is reducible to
+such a simple objective. The search needs to be game-agnostic and
+practical, but since the space is exponential
+($256^{n}$)\cite{GeneralizedSuper}, we need to be smart here.
 
@@
 \subsection{The NES hardware and emulation}
-The NES is based around an 8-bit processor running at 1.79~MHz, the Ricoh 2A03.
-8 bits is really small.
-You can see them all right here: 00001111.
-It's no coincidence that each controller also has 8 buttons: Up, Down, Left, Right, Select, Start, B and A.
-It has only 2048 bytes of general purpose RAM.
-(There is also some special purpose RAM for graphics, which we ignore in this work.) 2048 bytes is really small.
-You can see them all in Figure~\ref{fig:bytes2048}.
-As a result, NES programs are written to use memory efficiently and straightforwardly; usually there are fixed memory locations used for all the critical game facts like the player's health, number of lives, coordinates on the screen, and so on.
-For example, in Super Mario Bros., the single byte at location \verb+0x757+ contains the number of lives the player has.
-The location \verb+0x75F+ contains the current world, and \verb+0x760+ the current level.
-The NES outputs 60.0988 frames per second, which we will just call 60 in this paper.
-
-% (32kB of ROM is addressable at a time, and most games used bank switching to access hundreds of kilobytes of game included in the cartridge.)
+The NES is based around an 8-bit processor running at 1.79~MHz, the
+Ricoh 2A03. 8 bits is really small. You can see them all right here:
+00001111. It's no coincidence that each controller also has 8 buttons:
+Up, Down, Left, Right, Select, Start, B and A. It has only 2048 bytes
+of general purpose RAM. (There is also some special purpose RAM for
+graphics, which we ignore in this work.) 2048 bytes is really small.
+You can see them all in Figure~\ref{fig:bytes2048}. As a result, NES
+programs are written to use memory efficiently and straightforwardly;
+usually there are fixed memory locations used for all the critical
+game facts like the player's health, number of lives, coordinates on
+the screen, and so on. For example, in Super Mario Bros., the single
+byte at location \verb+0x757+ contains the number of lives the player
+has. The location \verb+0x75F+ contains the current world, and
+\verb+0x760+ the current level. The NES outputs 60.0988 frames per
+second, which we will just call 60 in this paper.
+
+% (32kB of ROM is addressable at a time, and most games used bank
+% switching to access hundreds of kilobytes of game included in the
+% cartridge.)
 
@@
 
-\begin{figure}[htb]
+\begin{figure}[h!tb]
 \begin{center}
@@
 
-\paragraph{Weighting.} To reduce the importance of randomness in the
+\paragraph{Weighting.} \label{sec:objectiveweighting} 
+To reduce the importance of randomness in the
 learning process, and the arbitrariness of the slicing, each objective
@@
 in $[0, 1)$ for all memories in the observed set.\footnote{It is not
-  defined when the memory is greater than all observed memories, which
-  we will encounter later. The code returns $|V|/|V| = 1$ in that case,
-  which is as good as anything.} This gives us the ability to compare
-objectives on an absolute scale.\footnote{This is certainly not the
-  only way to do it, and it has some questionable properties like
-  ignoring the magnitude of change. But it is very simple.} Weighting
-an objective is now simple:
+    defined when the memory is greater than all observed memories,
+    which we will encounter later. The code returns $|V|/|V| = 1$ in
+    that case, which is as good as anything.} This gives us the
+  ability to compare objectives on an absolute scale.\footnote{This is
+    certainly not the only way to do it, and it has some questionable
+    properties like ignoring the magnitude of change. But it is very
+    simple.} Weighting an objective is now simple:
 %
@@
   vector<uint8> state = GetState();
+
  \
+
   set<vector<uint8>> nexts;
@@
   \}
+
  \
+
   while (nexts.size() < NFUTURES)
     nexts.push\_back(/* random motif */);
+
  \
+
   for (vector<uint8> &next : nexts) \{
@@
     for (uint8 i : next) Step(i);
-    double score = ScoreByFutures(before, futures);
+    double score =
+      ScoreByFutures(before, futures);
     // Save the best-scoring next...
-    // 
   \}
+
  \
+
+  ReweightMotifs(best\_next, motifs);
   ReplaceBadFutures(futures);
@@
 of the {\tt NFUTURES} futures to get a final state and memory. We
-score each final memory relative to the start memory. Scoring is
-a matter of
-
-\subsection{Caching of emulation}
+score each final memory relative to the start memory. The particulars
+of scoring are not as interesting as the general idea, which is:
+\begin{itemize}
+\item The potential 10-frame {\tt next} sequence that we're
+evaluating gets an {\em optimistic} score. This is based on the futures
+for which the objectives {\em go up}. This is always non-negative.
+\item Each future is also given a score, which is the sum of scores
+from all the different {\tt next} sequences, regardless of their sign.
+\end{itemize}
+
+The purpose of the first is to identify a good {\tt next} sequence. We
+take the sequence with the highest optimistic score. The idea is that
+this is the sequence that gives us the best outcome if we continue
+to control what we do in the future.
+
+The purpose of the second is to identify which futures are good. A
+good future tends to bring good outcomes regardless of the immediate
+next step. Random futures that make is walk left when the good stuff
+is to the right, or jump when there are spikes nearby above our head,
+will receive negative scores for many or all of the {\tt next}
+sequences.
+
+{\tt ReweightMotifs} changes the weight of motifs that match the best
+{\tt next} sequence that we just committed to, if we think that we
+made progress on this step. The idea is to learn which sequences tend
+to work well for us; this might be different from what the human
+player did. For example, in run-to-the-right-and-jump kinds of games,
+human players tend to hesitate more before obstacles and enemies than
+a computer does.\footnote{That said, this part was an early idea and
+  is probably not necessary. It's suspicious because these 10-frame
+  sequences are not necessarily motifs (10 is the same as the normal
+  motif length, and futures are generated by motifs, but they can
+  become desynchronized because of the final incomplete motif, future
+  mutation, and other things). So sometimes the chosen sequence
+  doesn't affect weights. I think this would be better if we kept a
+  Markov model } Knowing whether we made progress is kind of
+difficult, since we can almost always find some move that increases
+the objectives locally. For this I use the same approach of {\em value
+  fraction} from Section~\ref{sec:objectiveweighting} based on a
+sample of memories that we've seen so far. If we appear to be
+progressing then the motif is weighted up; if we're regressing then it
+is weighted down.
+
+{\tt ReplaceBadFutures} kicks out the the futures with the worst total
+scores, so that over time the random futures become good futures. Of
+course, we always have to keep randomly adding to the end of each
+future, and who knows what crap we'll find? A late addition to the
+algorithm replaces some proportion of these with mutated versions of
+the best scoring future. This helps us from messing up a good future
+by appending crap to it, since we have multiple copies of it. It also
+makes search more efficient, since futures that share common prefixes
+can often benefit from caching (Section~\ref{sec:caching}). Currently
+mutating a future involves chopping off its second half (it will get
+extended to the desired length in the next step), and a \sfrac{1}{7}
+chance of {\em dualizing} the entire sequence. Dualization swaps
+the left and right buttons, the up and down buttons, B and A, and 
+select and start. The idea of this is to occasionally introduce
+very different futures to help us get out of stuck situations where
+we should really be turning around or something like that.
+
+During the development and tuning of the algorithm, I sometimes
+observed {\tt ReweightMotifs} and {\tt ReplaceBadFutures} conspiring
+to create a total monoculture of futures, where the weight of a single
+motif (specifically ``press no buttons'') went out of control and all
+futures just consisted of waiting. Waiting is usually a local
+improvement to some objectives because of internal frame counters and
+the cursor used to control music playback. To guard against this,
+motifs have a maximum (and minimum) possible weight in {\tt
+  ReweightMotifs}, and {\tt ReplaceBadFutures} always ensure that
+some fraction of the futures are completely random (ignoring motif
+weights).
+
+\parameteralert{
+  This is really the worst part in terms of parameters. They are: 
+  {\tt NFUTURES}, the number of futures to maintain (40);
+  {\tt NWEIGHTEDFUTURES}, the number of futures that are weighted,
+  as opposed to totally random (35);
+  {\tt DROPFUTURES}, the number of the worst-scoring futures we
+  completely drop (5);
+  {\tt MUTATEFUTURES}, the number of worst-scoring futures that
+  we replace with mutants of the best future (7);
+  {\tt MINFUTURELENGTH} and {\tt MAXFUTURELENGTH}, which put bounds
+  on the sizes of futures (50 and 800);
+  {\tt OBSERVE\_EVERY}, the frequency with which we sample memory
+  for computing the value fraction for motif weighting (10);
+  {\tt ALPHA}, the multiplicative factor for up- or down-weighting
+  motifs (0.8);
+  {\tt MOTIF\_MIN\_FRAC} and {\tt MOTIF\_MAX\_FRAC}, the bounds
+  on motif weights (0.1 and 0.00001).
+
+  Each is a scarlet letter of personal shame and future work is to
+  eliminate them. In my opinion the least excusable are the bounds on
+  future length, since these are related to what portion of time is
+  ``interesting'' from a game perspective---for example, the max
+  future length must exceed the number of frames that transpire after
+  Mario dies but before he respawns, or the algorithm cannot properly
+  avoid death. In my opinion this requires too much knowledge of the
+  game. I don't have a good idea about how to compute this, at least
+  without a human providing an input movie of dying (which would
+  help for lots of other reasons)---but that would complicate the
+  output of the learning process, which violates a primary design goal.
+
+  {\tt NFUTURES} is another bothersome one, but there is more hope here.
+  Basically it trades off computation time for quality of play, and
+  more seems to always be better. We do have runtime metrics that could
+  be used to dynamically change NFUTURES. For example, we could use
+  the notion of improvability from Section~\ref{sec:backtracking}, or
+  the value fraction, the gauge the marginal value of searching an
+  additional future. Or something like that. This might actually help
+  performance because we do the same amount of work (modulo caching)
+  even during periods that the inputs are being ignored, like between
+  worlds in Super Mario Bros., as we do when the going gets rough,
+  and searching for the exact right delicate sequence of inputs would
+  benefit from more options. The appearance of pausing in the output
+  for Bubble Bobble (Section~\ref{sec:bubblebobble}) suggests that it
+  knows all the futures are bad and needs to search different ones,
+  and corroborates this idea.}
+
+\subsection{Backtracking} \label{sec:backtracking}
+
+The algorithm above always makes progress at the same rate of 10
+frames per iteration. Sometimes it can get stuck in local maxima.
+Super Mario Bros., my tuning game for this work, seems almost to have
+been designed to thwart {\tt playfun}. The first level is pretty easy
+once you have a decent algorithm, and early versions beat it without
+any trouble. It's mainly about avoiding Goombas and planning far
+enough ahead to make high jumps over pipes and pits (see the
+accompanying videos for amusing struggles with these). World {\tt 1-2}
+provides a challenge you may recognize from your youth, immediately
+followed by something that computers find much harder
+(Figure~\ref{fig:mario12}).
+
+I literally spent about six weekends and thousands of hours of CPU on
+this problem. First, the overhang is set up so that enemies emerge
+from it just as you arrive. This means that loose play (imperfect
+enemy avoidance) tends to get you killed right here. Mario has found a
+number of different solutions to this problem, from waiting, to
+kicking turtles from the left to clear out the enemies, to timing his
+jump perfectly (it is possible!) to stomp the turtle after bouncing his
+head off the ceiling. It's fun to see the different approaches and is
+a good benchmark for whether the search algorithm is actually doing a
+good job at tactics.
+
+\begin{figure}[h!tb]
+\begin{center}
+\includegraphics[width=0.95 \linewidth]{mario12}
+\end{center}\vspace{-0.1in}
+\caption{This is where it starts to get hard, even with lexicographic
+ordering and time travel. This overhang is very tight in terms of the
+inputs that can get you through it; random play tends to kill you because
+of the advancing enemies that can't easily be jumped over. Greedy
+algorithms can't resist the ledge with the coins, but then would need to
+turn around.}
+\label{fig:mario12}
+\end{figure}
+
+Immediately after this is an important test of longer-term planning.
+Most of my early solutions would jump up on this ledge with 4 coins,
+since the score is a component of many lexicographic
+orderings\footnote{Maybe even all of them, since it should be globally
+  obeyed; it's therefore a valid extension to any ordering that
+  doesn't already include it. It's not necessarily a tight extension,
+  however, so it may be excluded for that reason, or because during
+  initialization it is not actually zero and so not globally obeyed. I
+  never checked this stuff because I wanted to avoid any temptation to
+  overfit the learning algorithm to the particulars of any game.} and
+coins give you points. Down below is undesirable in the short term
+because of the enemies. Then Mario would feel stuck and just hump up
+against the wall jumping in this tiny dead end until time ran out. The
+game contains some bugs (you have probably seen them) that allow the
+screen to be scrolled without Mario actually progressing; these little
+mini scrolls, plus the givens like the frame counter and music cursor,
+prevent Mario from getting out of the local maximum. This is extremely
+frustrating. I decided to add some longer-term planning into the
+search procedure in order to try to help in these kinds of situations,
+as well as the occasional deaths I would see.
+
+Every {\tt CHECKPOINT\_EVERY} frames we save the state, and then every
+{\tt TRY\_BACKTRACK\_EVERY} rounds, we do a backtracking phase if we
+can. We just need to find a checkpoint at least {\tt
+  MIN\_BACKTRACK\_DISTANCE} frames in the past. Call that point
+{\tt start} and the current point {\tt now}. The backtracking
+routine looks like this:
+
+\begin{itemize}
+  \item Let {\tt improveme} be the inputs between {\tt start} and {\tt now}.
+  \item Get {\tt replacements}, a set of input sequences we might use
+    instead. These are usually based on {\tt improveme}.
+  \item Add {\tt improveme} to the set of replacements.
+  \item Truncate the movie to {\tt start}.
+  \item Now, do the normal playfun loop as though the (truncated) futures
+    are whatever our current futures are, and the set of {\tt next}
+    sequences are the {\tt replacements} array.
+  \item Whatever the best one among those is, keep it. Don't update motifs
+    or futures, however.
+\end{itemize}
+
+The idea is that we take a large sequence from our recent past, and the
+same futures we're already using, and see if that sequence can be improved,
+according to our objective functions, and using our same futures. Since
+the existing sequence is always one of those, if it is the best one,
+then the backtracking phase is a no-op. If we find something better, we
+slot it in and continue. So the only risk here is if our objective functions
+aren't good (we take as given that they are) and the only cost is time.
+
+Generating the candidate set of replacements uses a bunch of different
+techniques. They are:
+
+\paragraph{Random.} We generate a random sequence of the same length as
+the {\tt improveme} sequence.
+
+\paragraph{Opposites.} We dualize (swap left with right, up with down, start 
+with select, and B with A) and/or reverse random spans of the {\tt
+  improveme} sequence. The idea is to try to introduce some variety in
+cases where we're really getting stuck. This was specifically in hopes
+that Mario would walk off the coin ledge in {\tt 1-2} and then find that the
+course was clear below. I felt okay about this since this seems to be a
+generic notion (the buttons do have semantics that are common to most NES
+games where left and right are opposites), but that may just have been
+rationalization since I was getting desperate. It didn't work; see below.
+
+\paragraph{Ablation.} Randomly blanks out buttons from the input. For
+example, if we don't need to jump and jumping is slowing us down or
+something, this can remove that and make a small improvement to the
+sequence.
+
+\paragraph{Chop.} Removes random spans from the input. This iteratively
+removes spans as long as the previous span was an improvement (see
+below). This can cause the movie to get shorter by removing parts that
+weren't contributing anything.\footnote{However, in games where an
+  objective function includes something like a frame counter or the
+  music cursor, shorter sequences always score lower than
+  otherwise equivalent longer sequences.}
+
+We use the following formula to score a potential improvement:
+
+\begin{center}
+\includegraphics[width=0.95 \linewidth]{isimprovement}
+\end{center}
+
+The {\tt start} state is as discussed, {\tt old\_end} is the state at
+{\tt now}, and {\tt new\_end} is the state after the potential
+improvement sequence. $\int_s^o$ is the integral of score changes
+along the {\tt improveme} path and $\int_s^n$ along the candidate
+improvement. The integral is the weighted sum of the objectives
+increased minus the weighted sum of the objectives that decreased, at
+each step, all summed up. This works better than just computing the
+weighted score from e.g. {\tt start} to {\tt old\_end} when they are
+separated by many frames (typical backtrack distances are several
+hundred frames). The expression $n-o$\footnote{This is not numerical
+  minus but minus according to the set of objective functions. Just
+  roll with it.} is just the weighted sum of objectives that increased
+minus the weighted sum of objectives that decreased between the old
+end state and new end state; we have no option of computing the
+integral here because we don't have an input sequence that goes from
+the old end to the new end (and there almost certainly isn't one). We
+require that three conditions hold:
+\begin{enumerate}
+\item $\int_s^n \geq \int_s^o$,
+\item $\int_s^n > 0$
+\item $n - o > 0$
+\end{enumerate}
+
+The integral has to be at least as good as before, the new integral
+has to be positive, and the new end state needs to look better than
+the old end state (the triangle inequality does not necessarily hold
+here). If they all do, then the score for the improvement is $\int_s^n
+- \int_s^o + (n - o)$, which is guaranteed to be positive.
+
+Even if we have many possible replacements, we try only the ones with
+the best scores; due to the particulars of the implementation there is
+not a simple constant describing this maximum, but it's around 220
+with the parameter settings used in this paper.
+
+\parameteralert{
+  Again! The humiliating appearance of constants! There are many here,
+  having to do with the number of potential replacements we use of
+  each type, the maximum number we keep, how often we bactrack, and
+  how far. I am not totally satisfied with how backtracking works (see
+  below), so I don't want to spend too much energy speculating about
+  how to reduce the parameter space here; I'd rather replace the
+  method wholesale.}
+
+The {\em improvability} of a state is the fraction of these potential
+improvements that are legal improvements, as above. Based on my
+analysis, states that are highly improvable ($> 30\%$) tend to be
+``worse'' (more stuck, closer to death, or whatever) than those that
+are not very improvable ($< 5\%$). This isn't being used for anything
+currently, but getting an absolute picture of how good a state is (as
+opposed to simply comparing it to an adjacent state) is one of the
+difficulties with this approach, so this may be a useful notion for
+future directions.
+
+\paragraph{Takes of woe, tales of joy.} 
+Backtracking was initially added in order to fix the mario coin ledge
+problem in {\tt 1-2}. It did not help with this. Older versions of the
+code would have Mario get past this spot, probably by luck, but in
+these he would get stuck in an embarrassing way on a small pit later,
+or jump straight into an enemy. Most of the approaches that actually
+looked solidly good elsewhere were failing badly here. As I started
+writing up this paper, on an airplane, I discovered the bug in the
+weighting of objective functions described in
+Section~\ref{sec:objectiveweighting}. On my trip I had let {\tt
+  playfun} run with particularly high constants ({\tt NFUTURES} = 50,
+lots of backtracking candidates, etc.) and it had spent about a
+thousand CPU hours playing to this point, grabbing the coins, and then
+dying of timeout, then losing all its lives, starting over, and
+getting to this point again! After fixing the bug, I tried again, and
+it was a huge breakthrough: Mario jumped up to get all the coins, and
+then almost immediately jumped back down to continue the level! On his
+first try he beat {\tt 1-2} and then immediately jumped in the pit at
+the beginning of {\tt 1-3} just as I was starting to feel suuuuuper
+smart (Figure~\ref{fig:mario13}). On a scale of OMFG to WTFLOL I was
+like {\em whaaaaaaat}?
+
+\begin{figure}[hb]
+\begin{center}
+\includegraphics[width=0.95 \linewidth]{mario13}
+\end{center}\vspace{-0.1in}
+\caption{After making me feel so successful by finally cleanly
+  finishing world {\tt 1-2}, Mario immediately jumps into the first
+  trivial pit in {\tt 1-3}. I believe what's going on here is that
+  this is actually the best future, because resetting back a single
+  screen isn't a large loss, lives aren't included in the objective
+  function, and probably the rest of the futures were struggling to
+  make progress. This level is very jumpy and probably needs more
+  futures to navigate its obstacles. In any case, Mario, Touch\'e.}
+\label{fig:mario13}
+\end{figure}
+
+Seeing this huge improvement changed my idea about what part of the
+code needed work (I now believe that simpler search strategies might
+work and better lexicographic order generation and weighting is called
+for). But, this was already in the midst of writing up the paper, so
+instead I spent the CPU time running the current version on a bunch of
+games. Those results are described in Section~\ref{sec:results} and
+videos are available at \url{http://tom7.org/mario/}. In the next
+section I describe some of the performance improvements I made, and
+other modifications to the emulator interface.
+
+\section{Performance} \label{sec:performance}
+
+Performance is really important, both because the quality of the
+output is CPU-bound and because I am impatient. In its current state,
+running {\tt playfun} is an overnight affair; it takes about an hour
+of real time to produce 1000 output frames (16 seconds of gameplay) on
+a high-end desktop machine. Other than a trick that I didn't end up
+using, these performance ideas are not at all new, but they are
+documented here for completeness. I certainly spent lots of time on 'em!
+
+\subsection{Caching of emulation} \label{sec:caching}
+
+The most expensive part, by far, is running a step of emulation. It
+takes about 500\textmu s to emulate a single frame, though this can
+vary depending on what instructions are actually executed (recall that
+each frame corresponds to many many 2A03 instructions; there are
+approximately 16,000 clock cycles per frame!). Hardware like the
+Picture Processing Unit and sound chip are emulated as well, which may
+actually be less efficient than the actual hardware (for example, the
+sound hardware can implement filtering with passive physical
+components like capacitors, but FCEUX contains a finite impulse
+response filter running on IEEE floating point numbers). We want to
+avoid executing a step more than once, if at all possible.
+
+Caching is so canonical that some have called it (pejoratively) the
+only idea in Systems research. And it is what we do here. The emulator
+adds a call
+%
+\begin{code}
+void CachingStep(uint8 input);
+\end{code}
+%
+with exactly the same semantics as {\tt Step}. However, it first
+checks to see if this exact input has been executed on this exact
+start state before. If so, it just restores the cached result
+state instead of executing the step.
+
+I use the excellent CityHash function\cite{CityHash} as the hash code,
+and use a least-recently-used approach to clean out the hash table
+when it has more than a fixed {\em slop} amount more than the target
+size. States are large because they contain both the savestate of the
+before and after state. I have plenty of RAM, so I allow each process
+to store up to 100,000 states with 10,000 states of slop, which takes
+about 3 gigabytes per process.
+
+Caching adds a little bit of overhead (a single step is about 13\%
+slower, amortized), from saving the state, computing the hash code,
+copying the state into the table, and cleaning the hash table
+periodically. A cache hit is a zillion times faster than actually
+executing the step, however, so as long as the hit rate is more than
+13\%, it's worth doing. I use caching step in most places inside {\tt
+  playfun}, except when I know that the state can't have been computed
+yet. The {\tt playfun algorithm} is particularly suitable to caching:
+The set of potential next input sequences {\tt nexts} usually share
+several input prefixes, and we keep around futures for some time, so
+we can often end up evaluating the same one under the same conditions
+many times. Moreover, mutant futures usually share their first half,
+making them half price. A really important property of caching is that
+it's based on the state of memory but doesn't care about the actual
+sequence used to get there. This means that in cases where the input
+is ignored (Mario ignores the jump button most of the time while in
+the air, for example, and ignores all inputs between levels) we reach
+equivalent states and can use the cache if the next input is exactly
+the same. The ideal approach here would be to follow how the bits of
+the input are actually read by the code, and insert a more generic key
+into the hash table. For example, if we see that the code never even
+read the bit of the input corresponding to the up direction, then we
+can reuse the step for an input that matches all bits but the up bit!
+This of course requires a lot of mucking around in the internals,
+which on a scale of within the scope to beyond the scope of this
+article is a 9.9.
+
+{\bf Software engineering lesson:} Initial results from caching were
+disappointing and I thought the overhead was to blame. I made several
+low-level tweaks to avoid copying, etc., to reduce the overhead from
+36\% to 13\%. Then I discovered that there were 0 cache hits ever,
+because of a bug (I was inadvertantly using pointer equality on keys
+instead of value equality, so keys were never found). Check basic
+things about your code's correctness before going on an optimization
+sortie!
+
+\subsection{Space optimizations}
+
+Before I had the idea that is the actual subject of this paper, I was
+playing around with other emulator search ideas that involved storing
+very very large numbers of states. This necessitated minimal
+representations of savestates. FCEUX uses zlib internally to compress
+states, which works well; they shrink from something like 13,776
+bytes\footnote{I don't completely understand the NES architecture and
+  emulation strategy, but I believe some games include expansion chips
+  that require more space for save states. All this work is based on
+  the 2048 bytes of main NES RAM, as you already know. Perhaps clever
+  video game authors from the past can protect against this paper's
+  approach by storing important game facts inside expansion chips
+  in the cartridge.} to an average of 2266. I made some modifications
+to the save/load routines to make this as small as I could manage.
+I removed the backing buffer, which is only used for drawing graphics,
+movie info, header bytes that are constant or already known (savestate
+size), which shrunk the compressed size to an average of 2047.42 bytes.
+That meant I could fit about 32 million savestates in RAM at once, which
+would be kind of amazing if I could tell my 8 year-old self that.
+
+Compression algoritms work well when there is lots of redundancy in
+their input. Comparing memories between two frames in some game,
+you'll often see only a few differences. Some of the memory contains
+stuff that's essentially read-only, even. To improve the
+compressibility of savestates, I introduce the idea of a {\em basis}
+state. This is any non-empty sequence of bytes which is an additional
+argument to the {\tt LoadState} and {\tt SaveState} functions. It's
+subtracted from the savestate prior to compression (it doesn't have to
+be the same size; it's treated as an infinite repetition of those
+bytes). If the basis is the same as the savestate, then the result is
+all zeroes, which compresses nicely (of course, now you need to have
+the basis in order to load the state, which is the same as the
+savestate, so that didn't save you much!). But memories during
+gameplay often don't differ much, so I save a state from any arbitrary
+frame in the game and then use that as the basis for everything else.
+This factors out any common parts and makes the states much more
+compressible: The average size drops to 1870.41 bytes.
+
+Using a better compression algorithm like the Burrows--Wheeler
+transform\cite{Burrows}\footnote{This is really one of the all-time
+  great algorithms. If you don't know it and you like this kind of
+  thing, you should check it out. It's so surprising and elegant.}
+would probably help too; zlib implements LZW which is only mediocre.
+However, on a scale of throwing my computer out the window to wanting
+to reimplement all software in my own private SML library, I just
+can't stand getting 3\rd-party libraries to compile and link into my
+applications, so I didn't even try. In any case, I abandoned this
+direction, which was not working well even with lean savestates.
+
+For {\tt playfun}, RAM is not a bottleneck at all. I disable
+compression completely on savestates, including the builtin zlib stuff
+(which is actually rather slow) and just use raw savestates. The
+removal of unnecessary headers and stuff is still there, and saves a
+few bytes and CPU cycles, but there was no reason to actually do it
+for this particular work. {\em But sometimes I do unnecessary things.}
 
@@
 
-\subsection{Backtracking}
-
-\section{Results}
-
-These games were all played with the same settings developed and tuned
-for Super Mario Bros.; that is, this was the result of just running
-the program the first time on a recording of me playing the game briefly.
+The {\tt playfun} algorithm involves running 40 or so different
+10-sequence {\tt next} steps and then scoring them against {\tt
+  NFUTURES} different futures. Each {\tt next} and each future is
+completely independent and dominated by the cost of evaluating
+emulator steps. The ideal thing would be to use threads and shared
+memory to run these steps in parallel. As I mentioned earlier,
+FCEUX is hopelessly single-threaded.
+
+Eventually I realized I needed to search more states than I could in a
+single thread, and so I created MARIONET. It's a play on words, a
+double entendre of ``Mario network'' and ``Marionette'', which is
+obvious. This is a mode whereby {\tt playfun} can be started in
+``helper'' mode (listens on a port for work to do) or ``master'' mode
+(connects to a list of ports to run work on), in order to run
+many emulators in different processes.
+
+The processes communicate over TCP/IP. I only run it on my local
+computer, but it would be reasonable to run it on a cluster or
+possibly even the Internet. I used SDL's SDL\_Net\cite{SDLNet} as a
+portable network interface in the hopes of keeping it
+platform-agnostic (FCEUX is basically portable so it would be a shame
+to make this Windows-specific, though I am not eager to try to get
+this thing to compile on other platforms, I gotta be honest). SDL is a
+nightmare to get working with mingw in a 64-bit compile, as usual, and
+SDL\_Net contained some bugs\footnote{In particular, the {\tt
+    SDLNet\_TCP\_Recv} call is supposed to block until it has received
+  the entire requested length, but it occasionally returns early.}
+that I had to spend some time working around. Anyway, I'm just
+complaining. For serializing requests and responses to bytes, I used
+Google's Protocol Buffer library\cite{ProtoBuf}, which is designed for
+this kind of thing.
+
+Helpers all start up with the same ROM and motif and objectives
+loaded, so that they can simulate whatever the master would
+compute.\footnote{Only the master reweights motifs, since it needs a
+  single sample of memories that we've observed. That means that in
+  cases where a helper generates random motifs, it does so with
+  respect to the original human weighting. There are some other small
+  things I simply avoided doing because they would require too much
+  coordination between processes or make the RPCs too large; MARIONET
+  was part of the system for much of its development.} They just wait
+on a port for the master to send a request, which can be either ``try
+this set of {\tt nexts} with these futures and send me the results'' or
+``try to find $n$ improvements of these inputs {\tt improveme} using
+so-and-so approach.'' In either case we send the base state as a
+serialized savestate.
+
+The master does the outer loop of the algorithm and all the writing to
+disk and so on. When it needs to do an expensive step like the inner
+loop, it prepares a set of jobs and sends them to workers using a
+little fork-join library ({\tt netutil.*}). The library keeps track of
+the outstanding jobs and which helpers are working, and feeds them
+work when they are idle, until all the jobs are complete. It's even
+got a color ANSI progress bar and stuff like that.
+
+Utilization is very good (Figure~\ref{fig:utilization}), and we get
+almost a linear speedup with the number of processes, up to the number
+of hardware threads the system supports (twelve). Actually, in
+addition to the computer remaining usable for other stuff, $n-1$
+helpers may work a bit better than $n$, because each of the
+helpers is then able to avoid preemption. Since the driver loop is
+synchronous, having one laggard process slows the whole darn thing
+down while everything waits for it to finish.
+
+\begin{figure}[htb]
+\begin{center}
+\includegraphics[width=0.75 \linewidth]{utilization}
+\end{center}\vspace{-0.1in}
+\caption{Utilization with 12 helpers and one master on a 6-core (12
+  hyperthreads) processor. Looks good. Keeps the bedroom warm.}
+\label{fig:utilization}
+\end{figure}
+
+MARIONET is a huge improvement and was what enabled using 40+ futures
+in a reasonable amount of time, which is key to high-quality output. It's
+obviously the way you want to run the software, if you can get it to
+compile. There is one downside, though, which is that it reduces our
+ability to get lucky cache hits across {\tt nexts} that are similar (or
+have the same effect), and when replaying the winner as we commit to it.
+It's worth it, but smartly deciding which helper gets which tasks (because
+they share prefixes, for example) would save time. Running it all in one
+process with a shared memory cache would be the best, as long as the
+lock contention could be kept under control.
+
+\section{Results} \label{sec:results}
+
+In this section I describe the results of running {\tt learnfun} and
+{\tt playfun} on several NES games. These games were all played with
+the same settings developed and tuned for Super Mario Bros.; that is,
+this was the result of just running the program the first time on a
+recording of me playing the game briefly. Every game I tried is
+documented here; lack of CPU time before conference is the main reason
+your favorite is not included. The website \url{http://tom7.org/mario/}
+contains video versions of some of these, which may be more fun. Let's
+begin with the {\em ur} game, Super Mario Bros..
+
+\subsection{Super Mario Bros.}
+
+This game is an obvious benchmark and the one I used during the
+development of {\tt learnfun} and {\tt playfun}. Some parts of the
+algorithm (see e.g. Section~\ref{sec:backtracking}) were developed
+specifically to get around problems in the ability to play this game,
+though I believe simpler methods would also work on this game, now
+that some important bugs are fixed.
+
+Automated Mario is fun to watch, and definitely my favorite. The
+familiarity of the game, the combination of human-like maneuvers and
+completely bizarre stuff, daredevil play, and bug exploitation
+(Figure~\ref{fig:mario13bug1}) are all a delight. It's even funny when
+it fails, watching Mario struggling with obstacles like tiny pits, or
+making a heroic move followed by a trivial mistake. I recommend
+watching the videos.
+
+This algorithm works well on Super Mario Bros., and I think that with
+some improvements it could play quite far into the game. It probably
+can't beat the game; Worlds {\tt 7-4} and {\tt 8-4} require some weird
+puzzling that we all needed {\em Nintendo Power Magazine}'s help to
+solve as kids, and are not just a matter of running to the right and
+avoiding obstacles. In the current incarnation, Mario breezes through
+World {\tt 1-1} every time, consistently beats {\tt 1-2} (I've only
+tried about three times with the newest version of the code, but each
+time he's succeeded) and has limited success on {\tt 1-3} but
+eventually suicides too many times. I'll keep trying.
+
+\begin{figure}[h!tb]
+\begin{center}
+\includegraphics[width=0.75 \linewidth]{mario13bug1a} \\[0.3em]
+\includegraphics[width=0.75 \linewidth]{mario13bug1b}
+\end{center}\vspace{-0.1in}
+\caption{Mario bounces off one Goomba up into the feet of another. Not
+  only doesn't he have enough velocity to reach the platform, but he's
+  about to hit that Goomba from below. Believe it or not, this ends
+  well: {\tt playfun} is happy to exploit bugs in the game; in this case,
+  that whenever Mario is moving downward (his jump crests just before the
+  Goomba hits him) this counts as ``stomping'' on an enemy, even if that
+  enemy is above him. The additional bounce from this Goomba also allows
+  him to make it up to the platform, which he wouldn't have otherwise!
+}
+\label{fig:mario13bug1}
+\end{figure}
+
+The Mortal Kombat-style {\em Finish Him!} to any machine learning
+algorithm is overfitting, however. Does the technique work on other
+games, without endless tweaking as I did with Super Mario Bros.? On a
+scale of no to yes, this is both a 1 and a 10; some games work even
+better than Mario and some are a disaster.
 
@@
 
-This is a very difficult game featuring a skateboarding cherubic
-island guy called Master Higgins, whose girlfriend has been kidnapped
-by King Quiller. He basically runs to the right and can throw or ride
-stuff that he finds in eggs. The controls are pretty soft and danger
-is everywhere. The objective functions that work are probably pretty
-similar to Super Mario Bros., but there are fewer obstacles to
-navigate---the difficulty for humans mostly comes from the speed and
-reaction time. Master Higgins doesn't care about bumping into rocks
-and dropping his health almost to nothing (but then is careful about
-health), and once he gets a weapon his aim anticipates off-screen
-enemies and he looks pretty savvy (Figure~\ref{fig:adventureisland}).
-His weakness regards holes in the ground, which he sometimes jumps
-into, probably for the same reason that Mario sometimes does. His
-``pot bonus'' was 16720. Master Higgins beats several levels and makes
-it into the cave before losing his last life by jumping straight into
-a vibrating skull with swirling fireballs around it, which on a scale
-of Darwin Award to dignified death is approximately a 6, in my
-opinion.
+This is a bad, difficult, but likable game featuring a skateboarding
+cherubic island guy called Master Higgins, whose girlfriend has been
+kidnapped by King Quiller. He basically runs to the right and can
+throw or ride stuff that he finds in eggs. The controls are pretty
+soft and danger is everywhere. The objective functions that work are
+probably pretty similar to Super Mario Bros., but there are fewer
+obstacles to navigate---the difficulty for humans mostly comes from
+the speed and reaction time. Master Higgins doesn't care about bumping
+into rocks and dropping his health almost to nothing (but then is
+careful about health), and once he gets a weapon his aim anticipates
+off-screen enemies and he looks pretty savvy
+(Figure~\ref{fig:adventureisland}). His weakness regards holes in the
+ground, which he sometimes jumps into, probably for the same reason
+that Mario sometimes does. His ``pot bonus'' was 16720. Master Higgins
+beats several levels and makes it into the cave before losing his last
+life by jumping straight into a vibrating skull with swirling
+fireballs around it, which on a scale of Darwin Award to dignified
+demise is approximately a 6, in my opinion.
 
@@
 
-\subsection{Bubble Bobble}
+\subsection{Karate Kid, The} \label{sec:karate}
+
+\begin{figure}[htb]
+\begin{center}
+\includegraphics[width=0.95 \linewidth]{karate}
+\end{center}\vspace{-0.1in}
+\caption{
+  Daniel-San blowing his last Crane Kick on a karate noob like there's
+  no tomorrow, which there won't be if he uses up all his power moves
+  like that. Note that the {\bf Chop} approach to backtracking was
+  used to generate this movie frame, appropriately.}
+\label{fig:karate}
+\end{figure}
+
+Karate Kid, The, is the typical trainwreck that follows when a beloved
+movie is turned into a video game. The game has awful controls and
+integer-only physics with massive throw-back upon collision, strange
+mini-games with no explanation, annoying debris flying through the
+sky, and luck-based fighting. It was probably only ever finished by
+kids with extreme discipline and self-loathing, or kids with only one
+video game available to them. It begins with a karate tournament which
+is even less fun than the main game, which is sort of a platformer
+with very few platforms.
+
+In this game I played 1,644 frames, just the first two of four
+opponents in the karate tournament. Automated by {\tt playfun}
+Daniel-San is able to punch and kick the low-level karate noobs,
+preferring to spend all his super-strong Crane Kick power moves right
+away. His style doesn't make much sense, sometimes facing away from
+the opponent, or throwing punches and kicks when the opponent isn't
+near. He doesn't worry much about taking damage. He gets to the final
+round and makes a valiant effort, at one point taking himself and the
+karate boss to 0 health simultaneously. But in this game, tie goes to
+the computer-controlled player without feelings, so it's back to the
+title screen for Daniel-San. The result is not impressive at all; the
+main goal here is to reduce the opponent's health, but our objective
+function can only track bytes that go {\em up}. Still, the automated
+version gets farther than my input did.
+
+\subsection{Bubble Bobble} \label{sec:bubblebobble}
 
@@
 level, bouncing on his own bubbles and so on, almost like a real
-player would. He beats XXX levels, more than I would have guessed.
-Since it takes two start button presses to enter the game, the second
-START is part of the input motifs. Amusingly, Bub uses this to his
-advantage to change the synchronization of his futures, and when the
-situation gets really tight, he'll pause the game until a good
-future comes around (Figure~\ref{fig:bubblebobble}). Paused states
-probably look like modest progress since memory locations like the
-music cursor are still advancing.
-
-XXX what finally happens?
+player would. He beats more levels than my input movie did! Since it
+takes two start button presses to enter the game, the second START is
+part of the input motifs. Amusingly, Bub uses this to his advantage to
+change the synchronization of his futures, and when the situation gets
+really tight, he'll pause the game until a good future comes around
+(Figure~\ref{fig:bubblebobble}). Paused states probably look like
+modest progress since memory locations like the music cursor are still
+advancing.
+
+After a harrowing escape from the ghost thing on level 4, Bub makes it
+to level 5 (my input movie quits at the beginning of level 4!), at
+which point I stopped the search because it was getting pretty pausey
+and I wanted to see some other games before the SIGBOVIK deadline.
 
@@
 
+\subsection{Tetris}
+
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[width=0.95 \linewidth]{tetris}
+\end{center}\vspace{-0.1in}
+\caption{The joyful noise of objective functions learned for Tetris.
+  The first fifth of the movie is navigating menus and looks very
+  different from the remainder. There appear to be some frame counters
+  identified, which make perfect smooth progress throughout the entire
+  movie. I believe discontinuities correspond to placing of pieces;
+  some objectives monotonically increase at these points (and are noisy
+  in-between), suggesting that they incorporate the score.
+}
+\label{fig:tetris-objectives}
+\end{figure}
+
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[width=0.95 \linewidth]{tetris-tower}
+\end{center}\vspace{-0.1in}
+\caption{Would you hire this construction company? Death is imminent,
+so {\tt playfun} pauses the game shortly after this and then doesn't
+unpause it.}
+\label{fig:tetris-tower}
+\end{figure}
+
+Tetris is a block dropping game, also known to the ancients. The
+Nintendo implementation is known for being inferior to the unlicensed
+Tengen version, which Nintendo tried to sue out of existence. Here I
+try to automate the Nintendo version as a tribute to litigation.
+Unsurprisingly, {\tt playfun} is awful at this game. Tetris has
+several screens of menus, where the player can select between
+different modes and theme musics and stuff like that. As an amusing
+prelude to this disaster in tetromino stacking, {\tt playfun} keeps
+entering the menu and exiting back to the title screen rapidly, taking
+several seconds to even start the game. (Like in Bubble Bobble, this
+means that the start button is among the motifs.) Although the piece
+dropping looks more or less natural (but it's hard to not be, as the
+game drops the pieces for you), the placement is idiotic---worse than
+random. This may be because placing a piece gives you a small amount
+of points, which probably looks like progress
+(Figure~\ref{fig:tetris-objectives}), so there is incentive to stack
+the pieces as soon as possible rather than pack them in. As the screen
+fills, there's a tiny amount of tetris-like gameplay, probably as the
+end of the game becomes visible in some of the futures. The end result
+is awful and {\tt playfun} makes zero lines and certainly no Tetrises
+(Figure~\ref{fig:tetris-tower}). The only cleverness is pausing the
+game right before the next piece causes the game to be over, and
+leaving it paused. Truly, the only winning move is not to play.
+
 \section{Future Work}
@@
 is some solace, it would really be nicer to remove these. One of the
-hardest to remove is the length of the futures explored. XXX thoughts
-on how to improve this?
+hardest to remove is the length of the futures explored. And I like
+a challenge!
+
+\paragraph{Generalized orderings.} Right now, lexicographic orderings
+are limited to vectors of unsigned bytes that get larger. Many games
+treat bytes as signed (I believe this is true of Mario's velocity, for
+example). For other bytes, like one representing your enemy's health
+(c.f. Karate Kid), winning means making the byte {\em go down}, not
+up. It is possible to generalize the lexicographic orderings we
+generate to a vector of ($L_i$, $<_i$) pairs, where $<_i$ tells us
+how to compare the bytes at that location. Things to try would be
+two's-complement signed comparison, or unsigned greater-than. I think
+this is a great avenue; the dangers are overfitting (now most short
+sequences can be explained in one way or another) and being too fancy.
+
+\paragraph{Input models.}
+
+\paragraph{Better backtracking.}
+XXX Keep backtracking until improvability is below a certain limit.
+XXX backtrack farther in some conditions.
 
@@
 
+\section{Conclusion}
+
+XXX
+
 \bibliographystyle{plain}

### 2013-03-17 trunk/tasbot/paper/paper.tex
Almost done. Maybe one more game, abstract, layout polish

@@
   doesn't affect weights. I think this would be better if we kept a
-  Markov model } Knowing whether we made progress is kind of
-difficult, since we can almost always find some move that increases
-the objectives locally. For this I use the same approach of {\em value
-  fraction} from Section~\ref{sec:objectiveweighting} based on a
-sample of memories that we've seen so far. If we appear to be
-progressing then the motif is weighted up; if we're regressing then it
-is weighted down.
+  Markov model and updated it no matter what sequences we generated.}
+Knowing whether we made progress is kind of difficult, since we can
+almost always find some move that increases the objectives locally.
+For this I use the same approach of {\em value fraction} from
+Section~\ref{sec:objectiveweighting} based on a sample of memories
+that we've seen so far. If we appear to be progressing then the motif
+is weighted up; if we're regressing then it is weighted down.
 
@@
 
-\paragraph{Parameter reduction.} I've already pointed out the places
-where there are mysterious constants in the code. Although the fact
-that the same constants work on many different games without tuning
-is some solace, it would really be nicer to remove these. One of the
-hardest to remove is the length of the futures explored. And I like
-a challenge!
-
-\paragraph{Generalized orderings.} Right now, lexicographic orderings
-are limited to vectors of unsigned bytes that get larger. Many games
-treat bytes as signed (I believe this is true of Mario's velocity, for
-example). For other bytes, like one representing your enemy's health
-(c.f. Karate Kid), winning means making the byte {\em go down}, not
-up. It is possible to generalize the lexicographic orderings we
-generate to a vector of ($L_i$, $<_i$) pairs, where $<_i$ tells us
-how to compare the bytes at that location. Things to try would be
-two's-complement signed comparison, or unsigned greater-than. I think
-this is a great avenue; the dangers are overfitting (now most short
-sequences can be explained in one way or another) and being too fancy.
+\paragraph{Parameter reduction.}
+I've already pointed out the places where there are mysterious
+constants in the code. Although the fact that the same constants work
+on many different games without tuning is some solace, it would really
+be nicer to remove these. One of the hardest to remove is the length
+of the futures explored. And I like a challenge!
+
+\paragraph{Generalized orderings.}
+Right now, lexicographic orderings are limited to vectors of unsigned
+bytes that get larger. Many games treat bytes as signed (I believe
+this is true of Mario's velocity, for example). For other bytes, like
+one representing your enemy's health (c.f. Karate Kid), winning means
+making the byte {\em go down}, not up. It is possible to generalize
+the lexicographic orderings we generate to a vector of ($L_i$, $<_i$)
+pairs, where $<_i$ tells us how to compare the bytes at that location.
+Things to try would be two's-complement signed comparison, or unsigned
+greater-than. I think this is a great avenue; the dangers are
+overfitting (now most short sequences can be explained in one way or
+another) and being too fancy.
 
 \paragraph{Input models.}
+I'm unsatisfied with the motif approach. As mentioned earlier, the
+obvious thing to try instead is a Markov model, which would probably
+be simpler and would allow re-weighting from inputs regardless of what
+we concocted while playing (the current version can only reweight the
+human motifs, not learn new sequences discovered while running). I
+would also like some solution to the start button problem---if it is
+among the motifs, it often shows up in gameplay in annoying ways. I
+don't feel good about simply blacklisting it, however. In Bubble
+Bobble, start appears to be used to burn away futures when all of them
+are bad. Maybe a simple improvement would be to allow the inner loop
+of {\tt playfun} to skip the turn (empty {\tt next} sequence) in
+order to simulate this.
 
 \paragraph{Better backtracking.}
-XXX Keep backtracking until improvability is below a certain limit.
-XXX backtrack farther in some conditions.
-
-\paragraph{Multiple players, multiple games.} Other than the particulars
-of the way it's built (\verb+vector<uint8>+ everywhere), there's no
-reason why the technique is limited to a single player's input. In a
-game like Bubble Bobble or Contra the two players can collaborate in
-interesting ways, and planning both simultaneously would probably lead
-to occasional awesome results. For example, in Contra, it might be
-that one player is shooting enemies for the other player, the bullets
-arriving from across the screen just in time to save him as he
-blithely jumps into danger. Another clever feat from the Tool Assisted
-Speedrun community is a sequence of inputs that beats multiple
-different {\em games} simultaneously. For example, human geniuses used
-tools to beat Mega Man (Mega Men?) 3, 4, 5, and 6 all at the same time
-using the exact same input sequence sent to all four
-games.\cite{BaxterTAS} I think the algorithms in this paper apply
+Backtracking is a powerful idea, and there's lots more to do here. The
+fixed-size backtracking window is disappointing since it's a
+parameter, and doesn't even seem right for most games. It would make
+sense to do something like lengthen the window until the {\em
+  improvability} starts dropping; basically, find the amount of recent
+past that's comparable to random play, and try improving that.
+Moreover, it should be possible to re-backtrack, since some parts of
+the game are harder than others. Ideally, when Mario has so few
+options that he contemplates suicide, he should be backtracking over
+and over and farther and farther until he finds a future that's good.
+This is the power of time travel. Use it, man.
+
+\paragraph{Efficiency in search}
+Quality is directly related to the number of states searched. Some
+sequences are easier to search than others, because they share a prefix
+with something we've already done and can be cached. It would be worth
+looking into algorithms that explicitly take into account the cost of
+branching, so that we explore some tree (of futures, for example) rather
+than disjoint linear futures. The effective number of futures explored
+could be much higher for the same CPU.
+
+\paragraph{Multiple players, multiple games.}
+Other than the particulars of the way it's built (\verb+vector<uint8>+
+everywhere), there's no reason why the technique is limited to a
+single player's input. In a game like Bubble Bobble or Contra the two
+players can collaborate in interesting ways, and planning both
+simultaneously would probably lead to occasional awesome results. For
+example, in Contra, it might be that one player is shooting enemies
+for the other player, the bullets arriving from across the screen just
+in time to save him as he blithely jumps into danger. Another clever
+feat from the Tool Assisted Speedrun community is a sequence of inputs
+that beats multiple different {\em games} simultaneously. For example,
+human geniuses used tools to beat Mega Man (Mega Men?) 3, 4, 5, and 6
+all at the same time using the exact same input sequence sent to all
+four games.\cite{BaxterTAS} I think the algorithms in this paper apply
 directly---it's just a matter of multiplexing the inputs to each of
-the games, and then taking the appropriate operation (min, max, sum) of
-their objective functions to produce a final objective function. The
-main obstacle is the architecture of the underlying emulator, which
-can only load one game into its global variables at once.
+the games, and then taking the appropriate operation (min, max, sum)
+of their objective functions to produce a final objective function.
+The main obstacle is the architecture of the underlying emulator,
+which can only load one game into its global variables at once.
 
@@
 
-XXX
+In this paper I showed how lexicographic orderings and time travel can
+be used to automate classic video games. The approach uses an
+amusingly simple and mathematically elegant model to learn what
+constitutes ``winning'' according to a human player's inputs. It then
+uses hundreds of CPU hours to search different input sequences that
+seem to ``win'', inspecting only the RAM of the simulated game,
+ignoring any of the human outputs like video and sound. The technique
+works great on some games, getting farther into the game than the
+human's inputs did, and produces novel gameplay (for example, bug
+exploitation and super-human daredevil timing). On other games, it is
+garbage, but on a scale of recycling symbol 1 to recycling symbol 7,
+it is at least hilarious garbage.
 

### 2013-03-18 trunk/tasbot/game.h
This is the first submitted version

@@ -11,10 +11,12 @@
 #define BASE64 "base64:jjYwGG411HcjG/j9UOVM3Q=="
 */
 
+/*
 #define GAME "tetris"
 #define MOVIE "tetristom2.fm2"
 #define FASTFORWARD 0
 #define BASE64 "base64:Ww5XFVjIx5aTe5avRpVhxg=="
+*/
 
 /*
 #define GAME "adventureisland"
@@ -23,9 +25,24 @@
 #define FASTFORWARD 0
 */
 
-// #define GAME "karate"
-// #define MOVIE "karate-tom.fm2"
-// #define BASE64 "base64:6xX0UBv8pLORyg1PCzbWcA=="
+/*
+#define GAME "karate"
+#define MOVIE "karate-tom.fm2"
+#define BASE64 "base64:6xX0UBv8pLORyg1PCzbWcA=="
+#define FASTFORWARD 0
+*/
+
+/*
+#define GAME "dinosaur"
+#define MOVIE "dinosaur-tom.fm2"
+#define BASE64 "base64:md0dI6RuaLcWPbxbtpuZKQ=="
+#define FASTFORWARD 0
+*/
+
+#define GAME "contra"
+#define MOVIE "contra-tom.fm2"
+#define BASE64 "base64:WlwvTxyvsfVajcDVrUVQ5Q=="
+#define FASTFORWARD 0
 
 // #define GAME "pacman"
 // #define MOVIE "pacman-tom.fm2"

### 2013-03-18 trunk/tasbot/learnfun.cc

@@ -165,6 +165,12 @@ int main(int argc, char *argv[]) {
 
   SaveMemory(&memories);
 
+  {
+    vector<uint8> save;
+    Emulator::SaveUncompressed(&save);
+    printf("Save states are %ld bytes.\n", save.size());
+  }
+
   uint64 time_start = time(NULL);
   for (int i = start; i < movie.size(); i++) {
     if (i % 1000 == 0) {

### 2013-09-22 wishlist/paper.tex
Typo in paper

@@
 % \usepackage{everypage}
+
+% \pagestyle{empty}
 
@@
 
-\begin{figure}
+\begin{figure}[t]
 \includegraphics[width=\linewidth]{wishlist-cdf}
@@
 
-\begin{figure}
+\begin{figure}[t]
 \includegraphics[width=\linewidth]{coin1}
@@
 
-\begin{figure}
+\begin{figure}[t]
 \includegraphics[width=\linewidth]{italiancoin1}
@@
 example, {\it congenital} is a much nicer word than the plain ugly
-{\it congenial}, and is reflected by the fact that {\it congenial} is
+{\it congenial}, and is reflected by the fact that {\it congenital} is
 used five times more frequently than {\it
@@
 updates since March~2006. There were 1,386 status updates containing
-such jems as ``Tom Murphy VII thinks mathfrak is straight ballin''
+such gems as ``Tom Murphy VII thinks mathfrak is straight ballin''
 and ``Tom Murphy VII global L\_50 reused for unused\_36166!!''. The
@@
 like ``0 is even'', and then you prove that all numbers are either
-even or odd by assuming ``$n$ is even or odd'' and proving ``$n + 1$
-is even or odd''.
-
-most people remember induction, perhaps 
-coinduction: never gonna give you up. once you pop you can't stop.
+even or odd by assuming ``$n - 1$ is even or odd'' and proving ``$n$
+is even or odd''. From this we conclude that every number is either
+even or odd. The idea is the proof shows how to, for any given number
+$m$, count down to the base case ``0 is even'', and then repeatedly
+apply the $n-1$ step (inductive step) to get back up to $m$. This is a
+great way to prove facts about finite things like numbers. Think of
+induction as a way you prove a statement like ``Good to the last
+drop,'' or ``There's always room for Jello.''
+
+Coinduction is a good proof technique for infinite things, like a
+sorted infinite list of possible strings. The idea behind coinduction
+is kind of like, you prove something like ``0 is a number'' (the base
+case), then prove something like ``if $n$ is a number, then $n+1$ is a
+larger number'', and then conclude that there exists an infinite
+series of numbers, each larger than the previous one. Think of
+coinduction as a way you prove a statement like ``Once you pop, you
+can't stop,'' or ``Never gonna give you up.''
+
+To sort the infinite list we don't actually use coinduction (we're
+not going to prove anything, just implement it), but its computational
+counterpart, corecursion. I just can't resist the ``coin'' pun.
+
+What we do is define a function ``most probable paths'', which returns
+a (possibly infinite) stream of strings for a given starting state.
+Each string is finite and ends with the terminal symbol \>, and they
+appear sorted by decreasing probability. (The most probable words
+overall will be just the first elements from the stream returned by
+this function when using a starting state like \<\<\< for $n=3$.)
+Since we don't want to explore all possible strings in order to
+produce this list (there are infinitely many), the trick is to put a
+lower bound on the probability of the words that will be included.
+There are always finitely many words with probability greater than a
+given positive value, unless the Markov chain contains a cycle where
+each edge has probability 1. (This is impossible for Markov chains
+created only by observing finite strings, such as all the ones in this
+paper.) It is efficient to use a very small lower bound with this
+algorithm, like 0.00000000000001.
+
+So the specification for ``most probable paths'' is to return all of
+the strings (that end with \>) that exceed the given lower bound in
+probability, sorted in descending probability order. It is easy to
+check the path directly to \>; we compare its probability to the lower
+bound by just looking it up in the table, and consider it if it
+exceeds the lower bound. For any other symbol {\it sym}, we will
+proceed (co)recursively: Call the probability of seeing {\it sym} next
+{\it p}, and then compute {\it tails}, all of the most probable paths
+starting in the state we would be in upon seeing {\it sym}. We turn
+{\it tails} into the sorted stream for the current state by just
+adding {\it sym} to the beginning of each string in it, and
+multiplying the probability by {\it p}. It remains sorted because
+multiplying by the same {\it p} is monotonic. The most important
+thing, which makes the algorithm practical (indeed terminate at all),
+is that we pass in a new lower bound: The current lower bound divided
+by {\it p}. After all, the outputs will be multipled by {\it p}, so
+they have to exceed this in order to meet the lower bound. This tends
+to increase the lower bound (sometimes over 1) since probabilities are
+between 0 and 1. This way, we only need to search a few symbols deep
+before it's clear that no string can exceed the lower bound.
+
+Now we have a list of sorted streams, at most one for each symbol in
+our alphabet. It is fairly straightforward to merge these into a
+single sorted stream, by only looking at the first element from each
+one. Pseudocode for {\tt most\_probable\_paths} appears in
+Figure~\ref{fig:mostprobablepaths} and for {\tt merge\_sorted} in
+Figure~\ref{fig:mergesorted}. Performance of this code is great;
+building the Markov chains (or even just reading the dictionary files)
+dominates the latency of the analyses in this paper.
+
+\begin{figure*}
+\begin{code}
+fun most\_probable\_paths \{ lower_bound : real, state : state \}
+       : \{ string : symbol list, p : real \} stream =
+  let
+    fun nexts i =
+      case symbol\_from\_int i of 
+      NONE => nil
+    | SOME sym =>
+      let 
+        val p = (* probability of seeing sym in this state *)
+      in
+        if p < lower\_bound
+        then nexts (i + 1)
+        else if sym = end\_symbol
+             then S.singleton \{ string = nil, p = p \} :: nexts (i + 1)
+             else 
+               let
+                   val lb' = lower\_bound / p
+                   val tails =
+                       most\_probable\_paths \{ lower\_bound = lb',
+                                             state = advance\_state (state, sym) \}
+               in
+                   (* Now multiply through the probabilities and add the symbol
+                      to the head of the strings. *)
+                   Stream.map (fn \{ string = t, p = p' \} =>
+                               \{ string = sym :: t, p = p * p' \}) tails ::
+                   nexts (i + 1)
+               end
+      end
+
+    (* Try all next symbols. *)
+    val streams = nexts 0
+  in
+    S.merge\_sorted bysecond\_real\_descending streams
+  end
+\end{code}
+\caption{Pseudocode for {\tt most\_probable\_paths}. {\tt advance\_state}
+gives a new state from a previous state and symbol observed, so that
+for example {\tt advance\_state(abc, z)} gives {\tt bcz}. The pseudocode
+for {\tt merge\_sorted} is given in Figure~\ref{fig:mergesorted}.}
+\label{fig:mostprobablepaths}
+\end{figure*}
+
+\begin{figure*}[t]
+\begin{code}
+fun merge\_sorted cmp l =
+  let
+      fun ms nil () = Nil
+        | ms (s :: t) () =
+          case force s of
+              Nil => ms t ()
+            | Cons (v, ss) =>
+              ms\_insert v [ss] t
+
+      and ms\_insert bv sg nil =
+          Cons (bv, delay (ms sg))
+        | ms\_insert bv sg (s :: t) =
+          case force s of
+              Nil => ms\_insert bv sg t
+            | Cons (v, ss) =>
+                case cmp (bv, v) of
+                    GREATER =>
+                      ms\_insert v (singleton bv :: ss :: sg) t
+                  | \_ => ms\_insert bv (s :: sg) t
+  in
+      delay (ms l)
+  end
+\end{code}
+\caption{Pseudocode for {\tt merge\_sorted}. {\tt ms} merges a
+sorted list, and {\tt ms\_insert} is a helper where we have a
+candidate best value {\tt bv} which will either be the one we
+return at the head of the stream, or we'll replace it and then
+stick {\tt bv} somewhere to be returned later. (This algorithm
+can be improved by making a data structure like a (co)heap;
+this is just a simple first pass.)}
+\label{fig:mergesorted}
+\end{figure*}
 
@@
 These words did not come with definitions, except for {\it gruntle}, which Jessica
-gives as ``the opposite of distgruntle''. This is actually already a word, but
+gives as ``the opposite of disgruntle''. This is actually already a word, but
 it was the inspiration for Section~\ref{sec:backformation}. {\it etsy} is the name
@@
 
-% TODO: backformation, stripping dis- and anti- and un- from words
-% then checking to see if they're still words?
-
 % TODO: portmanteautally
@@
 \section{Conclusion}
-
-\subsection{Recommendations}
-
-sweeeeeeeeeeeeeeeeeeet with 19 {\it e}s, which means ``Really sweet.''
-
-``rane'' sounds too much like ``rain'', but ``sare'' has a unique pronunciation and many people seem to think it's already a word.
-``cho'' is similarly easy to pronounce and spell.
-I propose that it be defined as ``A kind of cheese,'' so that we can really nail the new triple entendre on the classic joke.
 
@@
 words, and some don't work that well, but nonetheless we have several
-rich sources of words, each time scientifically justified.
-
-Nobody really likes writing conclusions, so I will just now summarize
-the paper by the giving the most likely sentences that I have not
-already written.
-
-% XXXX
-
-% \begin{Verbatim}
-facebook wall posts, 4-grams
-100 top paths:
-0.002525504 pittsburgh
-0.002093510 steelers
-0.002093510 sfo
-0.001953943 it's
-0.001256106 bdl
-0.001116539 sigbovik
-0.001099196 facebook
-0.000976971 mic
-0.000976971 s
-0.000976971 can't
-0.000837404 i'm
-0.000837404 icfp
-0.000837404 app
-0.000697837 x
-0.000697837 drunj
-0.000697837 g
-0.000616021 ther
-0.000558269 doesn't
-0.000558269 's
-0.000558269 sr
-0.000558269 pm
-0.000558269 fyi
-0.000558269 wtf
-0.000558269 k
-0.000558269 don't
-0.000558269 today's
-0.000558269 m
-0.000558269 what's
-0.000456766 iphone
-0.000429438 inted
-0.000418702 oh
-0.000418702 phl
-0.000418702 gonna
-0.000418702 pre
-0.000418702 zrh
-0.000418702 th
-0.000418702 h
-0.000418702 b
-0.000418702 n
-0.000418702 ok
-0.000418702 d
-0.000418702 that's
-0.000418702 rex
-0.000418702 fanzibar
-0.000418702 st
-0.000418702 kinda
-0.000418702 didn't
-0.000355262 cally
-0.000334962 brillobox
-0.000314027 superburrito
-0.000305304 weath
-0.000279135 techno
-0.000279135 vii
-
-trigrams, including comments:
-100 top paths:
-0.003013897 it's
-0.001856422 ther
-0.001704432 i'm
-0.001503910 sfo
-0.001102867 -
-0.001102867 oh
-0.001050350 don't
-0.001002607 haha
-0.001002607 s
-0.000902346 bdl
-0.000808769 mic
-0.000808769 that's
-0.000776981 pittsburgh
-0.000705812 thes
-0.000701825 ok
-0.000645063 can't
-0.000601564 icfp
-0.000601564 k
-0.000583335 didn't
-0.000507201 runj
-0.000501303 g
-0.000501303 omg
-0.000501303 x
-0.000485262 ins
-0.000453126 sout
-0.000450580 ands
-0.000449950 coff
-0.000425932 reat
-0.000425133 ning
-0.000416189 pittle
-0.000412468 getty
-0.000407151 ster
-0.000401043 sr
-0.000401043 pm
-0.000401043 m
-0.000401043 wtf
-0.000401043 aw
-0.000401043 d
-0.000401043 fyi
-0.000401043 ple
-0.000401043 's
-0.000400791 phon
-0.000386315 weat
-0.000384116 befor
-0.000375978 els
-0.000374307 als
-0.000357782 nothe
-0.000354769 app
-0.000354769 appy
-0.000354422 they're
-0.000353750 wate
-0.000345727 there's
-0.000340508 res
-0.000339344 mes
-0.000334202 ove
-0.000326473 wher
-0.000324373 lity
-0.000319011 hotos
-0.000316613 dea
-0.000315777 sigbovik
-0.000315105 ver
-0.000314965 thand
-0.000310607 outh
-0.000300782 n
-
-facebook, n = 2
-100 top paths:
-0.010628671 ing
-0.007144197 th
-0.005025442 st
-0.004291375 re
-0.002932937 ne
-0.002662094 se
-0.002546194 whe
-0.002108786 ther
-0.001923496 al
-0.001669458 con
-0.001652820 wor
-0.001498222 ch
-0.001495485 wis
-0.001379778 i'm
-0.001336166 jus
-0.001302469 sh
-0.001280888 yout
-0.001244352 fo
-0.001225250 le
-0.001224422 pre
-0.001221357 ar
-0.001201881 com
-0.001200574 ge
-0.001165417 thes
-0.001116295 som
-0.001108305 fing
-0.001104851 ang
-0.001102867 -
-0.001102706 it's
-0.001039704 ist
-0.001036886 hing
-0.001007930 ou
-0.001002607 s
-0.000967282 de
-0.000932965 en
-0.000926568 tor
-0.000827260 wat
-0.000827151 oh
-0.000814235 cor
-0.000812961 ned
-0.000783696 lis
-0.000762020 ple
-0.000760162 whis
-0.000759898 sed
-0.000755396 res
-0.000728412 mor
-0.000728049 ong
-0.000727818 don
-0.000720098 cand
-0.000710277 ning
-0.000708380 ithe
-0.000693547 ame
-0.000659248 ty
-0.000656211 dow
-0.000635771 ha
-0.000633225 ext
-0.000601564 k
-% \end{Verbatim}
+rich sources of words, each time scientifically justified. I conclude
+with a section of recommendations for words that ought to exist, along
+with definitions.
+
+\subsection{Recommendations}
+
+{\it Sweeeeeeeeeeeeeeeeeeet} with 19 {\it e}s is the clear favorite
+based on analysis of usage, so this one should be introduced. It means
+``Really sweet.''
+
+{\it Rane} sounds too much like {\it rain}, but {\it sare} has a
+unique pronunciation and many people seem to think it's already a
+word. I propose that {\it sare} be introduced as a noun meaning, ``a
+word that sounds real but isn't.''
+
+{\it Cho} is similarly easy to pronounce and spell. I propose that it
+be defined as ``A kind of cheese,'' so that we can really nail the new
+triple entendre on the classic joke. {\it Chomaker} is someone who
+makes that kind of cheese.
+
+{\it Unders} was one of the most frequently occuring words towards the
+top of many analyses. This word should be a colloquialism for
+underwear, which would probably already be understood from context.
+
+{\it Dise} is suggested by both the Markov model (as {\it dises}, {\it
+  dising}) and backformation (as {\it dises}). I like thinking of it
+as being the root of {\it paradise}, where {\it para--} means
+something like ``along side of'' or ``resembling''. So {\it dise} is
+the place you're really looking for when you get to paradise and
+realize it's just a mediocre country club.
+
+{\it Helafelo} is one hell of a fellow.
+
+\medskip
+{\bf Addendum.} During the preparation of this paper, the Scrallbe
+game has converged on a culture where the words played are real-seeming,
+with creative definitions. Examples: {\it frodeo} (``Gandalf is the
+clown.'') {\it pridefax} (``An unproven treatment for telephone
+anxiety.'') {\it eeeee} (``eeeee'') {\it orzigato} (``Move mr. roboto.
+for great justice.'') {\it stovebed} (``Now you don't have to get out
+from under the covers to make breakfast.'') {\it ovawiki} (``The free
+egg cell that anyone can edit.'') {\it gaptave} (``Two discontinuous
+musical intervals.'') {\it achoolane} (``Nostril (colloq.)'') {\it
+  gplerious} (``Completely overcome by software licenses.) {\it
+  bestcano} (``Some eruptions are better than others.'') Thanks to the
+players for their contributions, especially Ben~Blum, Chrisamaphone,
+Rob~Simmons, and Flammin~Fingers.
+
+\section*{Appendix}
+
+Here are the most likely words induced by the English Wikipedia, with $n=3$.
+I have left off the probabilities; they can be reproduced by downloading
+Wikipedia and running the software yourself, which only takes like 9 hours.
+
+s
+ther
+w
+en
+t
+c
+d
+f
+m
+b
+ser
+e
+diff
+u
+sup
+p
+n
+de
+tal
+othe
+th
+wikipedia
+rd
+con
+id
+r
+ame
+edia
+cal
+uk
+wiki
+und
+als
+st
+l
+km
+wp
+g
+x
+ext
+j
+des
+td
+ter
+h
+co
+k
+sout
+don
+o
+cound
+v
+ver
+wer
+med
+res
+ind
+la
+em
+re
+ins
+afd
+ands
+div
+htm
+infor
+rom
+wher
+ple
+mi
+acces
+ii
+ent
+unive
+ff
+refere
+ger
+bution
+lish
+ince
+wikiped
+auth
+dist
+oth
+smally
+rea
+al
+y
+est
+cle
+fr
+lation
+tral
+gov
+nd
+mily
+isbn
+cond
+bel
+befor
+hould
+ber
+pers
+inder
+les
+gree
+mation
+lar
+chan
+dia
+ove
+stary
+compan
+mon
+val
+san
+cate
+nothe
+pdf
+tes
+ster
+nort
+pril
+hist
+sq
+clast
+ful
+ment
+q
+aft
+tr
+nown
+lign
+del
+z
+ave
+dify
+stor
+sity
+artic
+pring
+rese
+mas
+li
+ord
+ca
+thes
+oldid
+lin
+es
+ress
+flage
+blace
+pland
+mar
+mes
+conce
+exter
+wate
+mp
+jame
+lity
+ft
+cology
+spany
+mic
+ve
+han
+thist
+spam
+cand
+sher
+smal
+subdiv
+tem
+jan
+ture
+inded
+bothe
+ared
+cour
+rect
+thers
+ope
+mm
+dom
+der
+mand
+tt
+dr
+phy
+cd
+websity
+stude
+lor
+bbc
+ru
+mer
+flor
+succes
+ence
+tely
+aust
+stan
+clas
+et
+outh
+pl
+ints
+vfd
+fc
+wikipedit
+sus
+forman
+contry
+au
+edu
+inding
+el
+lic
+bord
+mr
+ston
+whic
+ide
+hous
+mich
+auser
+google
+pt
+le
+doi
+engle
+ed
+gen
+userved
+cols
+num
+nate
+cer
+withis
+clude
+shous
+inted
+wikipedian
+thered
+andex
+chool
+ning
+attle
+chael
+sta
+plation
+ampion
+crea
+fontry
+lege
+prese
+grough
+iii
+mader
+augh
+ll
+ching
+fb
+furt
+yearch
+examp
+octor
+nove
+furth
+bein
+tra
+nover
+texter
+mont
+sted
+gener
+quary
+inclub
+ander
+sco
+ast
+gr
+wome
+ality
+lon
+hough
+ano
+exas
+polid
+sease
+dec
+sor
+ga
+nov
+unty
+shor
+imdb
+offic
+partic
+oved
+forn
+lan
+fm
+undex
+rus
+arted
+alia
+cong
+vill
+eason
+gue
+shough
+cc
+vil
+tember
+octory
+jr
+gove
+writy
+thand
+anding
+rev
+colog
+gover
+edian
+publist
+mics
+comple
+syster
+olding
+coor
+useum
+spect
+expan
+puble
+bate
+tribut
+oned
+coibot
+betwo
+conter
+cles
+leason
+fina
+freen
+eng
+fa
+que
+fied
+tre
+hamp
+guage
+wered
+nic
+awar
+yout
+usic
+cgi
+andy
+nrhp
+clar
+arting
+bc
+cated
+rober
+feb
+ames
+ven
+jun
+thould
+defere
+artics
+sen
+frand
+cred
+ada
+cana
+rection
+modiff
+sal
+los
+ret
+ched
+decial
+collow
+notalk
+nl
+yor
+elete
+sain
+ance
+cens
+chand
+apr
+delect
+dea
+stant
+bruary
+commer
+righ
+ch
+tor
+peral
+andia
+arth
+roup
+founty
+hor
+scies
+alth
+pete
+footnot
+html
+ath
+tribs
+ac
+chas
+eith
+pa
+tl
+refor
+bes
+bor
+cently
+fonter
+neer
+tration
+reat
+hi
+awa
+chris
+da
+wikipedits
+atter
+pp
+ents
+dvd
+arge
+wan
+sch
+sr
+fere
+inds
+norts
+tha
+nament
+anot
+oper
+aspx
+rence
+website
+minal
+brity
+dts
+apring
+goalso
+alber
+rown
+graphy
+anded
+min
+alian
+ength
+mility
+abbr
+talign
+bec
+com
+authe
+hite
+pic
+pation
+jone
+ords
+demy
+alt
+wated
+alife
+totalk
+lond
+tage
+ad
+ana
+evers
+amg
+gened
+colo
+engton
+mina
+lates
+porth
+fff
+lifor
+footbally
+stry
+shing
+hase
+lis
+eletion
+rol
+mally
+inton
+ded
+doesn
+ny
+mus
+pro
+du
+pre
+somes
+dired
+na
+becamp
+ire
+inces
+cologo
+mong
+swith
+acts
+aree
+ang
+els
+noth
+unic
+mous
+dest
+ali
+deat
+smale
+poing
+dc
+ission
+iv
+oct
+posity
+pred
+poss
+kore
+anda
+ja
+ms
+gard
+coi
+frome
+afc
+nhl
+publice
+mothe
+jers
+maded
+qual
+geory
+roy
+vid
+pur
+studing
+ip
+ign
+fath
+cally
+coundex
+col
+abouth
+twork
+threen
+img
+ess
+lishe
+lable
+reland
+cing
+chi
+thern
+dete
+ar
+ving
+di
+eur
+wers
+il
+dio
+hl
+ando
+seast
+sation
+orge
+ble
+sep
+db
+classion
+cm
+las
+conver
+evel
+ing
+popula
+uns
+polic
+vs
+frans
+ker
+ann
+phich
+ovember
+alish
+cher
+therence
+prock
+ife
+feate
+uservice
+indid
+unded
+powere
+hs
+phot
+assion
+von
+dece
+theight
+proview
+ama
+prote
+kg
+birthe
+specian
+userve
+se
+nois
+cfm
+ma
+len
+ause
+gle
+jp
+flagic
+differe
+counder
+rance
+swed
+lee
+tel
+ont
+sould
+whics
+artion
+rans
+ric
+parly
+overy
+ves
+sa
+grea
+ans
+une
+tary
+ss
+jul
+ne
+fami
+dever
+ange
+hms
+oright
+dation
+cription
+prothe
+univer
+lat
+tota
+un
+adming
+aaa
+ampan
+perfor
+athe
+seased
+pov
+becaust
+alson
+thed
+methe
+aff
+spand
+inc
+sis
+stral
+refered
+endex
+xfd
+texamp
+valign
+sha
+dists
+fron
+heade
+lood
+thest
+unived
+italk
+brary
+colled
+ucfirst
+ite
+gan
+ep
+rary
+oney
+wikipedition
+ra
+ex
+scord
+cition
+hout
+clos
+bon
+lp
+imple
+acade
+flagico
+bost
+isn
+spection
+samp
+ral
+tead
+sant
+roces
+yorks
+ories
+bords
+unition
+che
+dra
+epison
+eff
+stal
+abour
+ach
+earch
+busic
+reate
+conves
+ses
+hims
+mone
+fran
+carly
+nated
+lished
+fra
+pm
+oh
+att
+delet
+unior
+reporth
+daving
+emple
+tworld
+noved
+trand
+ross
+ie
+vall
+orld
+vi
+becauser
+caust
+dard
+exand
+feder
+ches
+yound
+gence
+bol
+stribut
+rement
+gf
+sugges
+tity
+forder
+aug
+systed
+mathe
+websited
+abover
+ltd
+recember
+colle
+wils
+mos
+sman
+lete
+dow
+eles
+throup
+ba
+unt
+expr
+formany
+df
+ments
+tect
+pape
+hool
+exper
+butions
+hited
+delection
+deces
+coung
+gon
+gers
+stors
+abc
+partics
+writion
+furthe
+rade
+mult
+bb
+lism
+afries
+frence
+bish
+arence
+tems
+musion
+colore
+ethe
+bute
+vie
+ia
+hing
+oute
+hom
+clain
+nom
+missue
+hus
+ns
+chich
+oxfor
+contribut
+vict
+autom
+boile
+novers
+aus
+airst
+jos
+proble
+werence
+georgan
+rel
+eas
+opedia
+rs
+ned
+aut
+iss
+pg
+ct
+othis
+indo
+shi
+dst
+alf
+ian
+grounty
+hile
+mor
+jor
+grap
+parth
+elet
+regory
+thems
+ell
+septed
+pc
+url
+gnf
+unds
+appy
+studed
+offere
+gb
+timent
+sed
+dary
+eved
+hally
+deate
+hr
+taly
+noming
+duri
+cented
+thead
+aliam
+mored
+pmid
+hered
+hong
+recore
+vious
+wates
+fathe
+goved
+squary
+soment
+mout
+rela
+eview
+unives
+shere
+trate
+betweek
+gia
+mory
+cas
+dal
+fi
+enge
+stribs
+colspan
+vol
+facted
+amer
+reser
+loc
+fination
+deven
+andid
+leade
+tand
+protes
+gar
+dely
+til
+alled
+sary
+tro
+alist
+ng
+cha
+charly
+ab
+hel
+begion
+mana
+sk
+powed
+ted
+calign
+wikipe
+dar
+nm
+churce
+mil
+ren
+dise
+rever
+sel
+howere
+nexter
+commen
+sonal
+oly
+subser
+coibothe
+shool
+hal
+froman
+eu
+waship
+pr
+rease
+ph
+rical
+op
+assed
+bility
+ording
+conces
+mal
+stitle
+iucn
+sri
+os
+mond
+lations
+mance
+artment
+rigin
+rt
+brite
+ffff
+jsp
+lb
+dd
+wikipeding
+lant
+sett
+arter
+te
+yource
+sh
+contribs
+footbal
+zh
+aution
+acter
+addit
+govers
+coln
+useries
+ract
+thich
+tain
+inten
+sec
+aa
+blo
+janual
+hon
+nowing
+letion
+tition
+pute
+goldid
+communite
+wors
+fontribut
+areat
+titute
+rected
+summent
+stational
+hu
+counded
+sp
+eld
+ples
+yeare
+nating
+fing
+foll
+reque
+fiel
+pol
+josed
+atland
+phile
+manuary
+civision
+ok
+feded
+ult
+dj
+sas
+alle
+cg
+buildid
+kee
+redia
+ap
+unk
+ko
+cr
+inver
+nomy
+sund
+toon
+reg
+aaaa
+cath
+dll
+lc
+ters
+communive
+bal
+prope
+refef
+ania
+mrs
+hased
+worder
+pria
+arry
+centing
+spon
+ade
+vand
+dor
+sourch
+tooks
+seph
+capan
+pbb
+colong
+sql
+petion
+shors
+fift
+prily
+prial
+brition
+fl
+seption
+nz
+fonto
+cbs
+fally
+whill
+losed
+cluded
+mations
+monal
+sland
+pict
+russion
+profilm
+coundo
+reture
+harly
+soute
+sc
+rr
+titution
+therent
+shouser
+pager
+outhe
+quent
+spons
+fn
+ited
+proces
+bable
+atte
+whicle
+sile
+si
+rece
+unitor
+yan
+grought
+cos
+pian
+comput
+trans
+jource
+arthe
+tury
+specia
+artice
+provid
+raction
+deleter
+alo
+fored
+ti
+eague
+wiking
+conth
+ree
+musly
+ars
+mf
+ups
+schood
+rey
+donal
+sected
+writed
+freet
+formate
+cre
+colution
+opers
+chen
+bd
+hp
+sember
+folled
+pagest
+smales
+wen
+socies
+cs
+engin
+aften
+chur
+delp
+bia
+ce
+nors
+cource
+ussion
+fontribs
+summan
+arties
+jon
+dism
+lang
+sking
+augus
+theren
+xs
+seve
+deball
+er
+uning
+truction
+nomic
+creat
+gian
+acco
+sv
+pera
+alls
+rict
+relete
+rects
+coord
+stre
+ences
+aread
+spane
+meth
+brea
+ener
+thernal
+thire
+whical
+rounty
+thu
+cl
+pola
+una
+thround
+uservation
+volved
+hund
+aga
+cames
+menter
+froma
+uset
+publis
+frey
+libe
+posite
+northe
+ska
+appedia
+recial
+ro
+fant
+hought
+mination
+geo
+presult
+andon
+occup
+ala
+nas
+thists
+ew
+maring
+secore
+nee
+dian
+inders
+inves
+infore
+hower
+sume
+scal
+kans
+indust
+eding
+ka
+aux
+brid
+var
+orded
+gived
+backgrough
+hugh
+ado
+prive
+eb
+rading
+joing
+hown
+euro
+lim
+kan
+chus
+alifor
+wille
+deced
+conven
+blogy
+nbc
+quall
+coller
+partion
+folle
+eleter
+reven
+reques
+lition
+humber
+moder
+backgroup
+tribution
+flages
+lears
+forch
+reight
+pault
+evely
+mayork
+sition
+feath
+recont
+guit
+metion
+hony
+deview
+nume
+confere
+fle
+unitory
+rican
+ima
+accore
+deside
+ris
+pedia
+lisher
+acture
+aland
+userview
+thate
+polor
+kar
+nel
+sents
+pany
+cluding
+vard
+bing
+woul
+havel
+greer
+thang
+genry
+effere
+locall
+desh
+thors
+empt
+sult
+quence
+mir
+blook
+porthe
+tran
+rew
+bording
+shought
+kely
+neight
+lam
+stion
+publishe
+hors
+wis
+apan
+nony
+sourn
+nage
+orce
+profest
+jonal
+fonth
+younty
+deta
+pd
+wa
+lo
+peries
+flign
+fource
+ros
+rfc
+mely
+mt
+smalled
+ge
+retion
+relect
+vern
+alid
+aligh
+userves
+speer
+willa
+infl
+undon
+dels
+aftern
+gres
+namedia
+advan
+distor
+oria
+subdiving
+lx
+presign
+blocal
+edway
+def
+scho
+fic
+counds
+belf
+sily
+metry
+cread
+pi
+novery
+occes
+cribut
+holor
+reces
+inte
+howed
+cout
+chis
+ency
+actory
+lange
+cf
+nc
+bornia
+adven
+revie
+sames
+newspan
+shan
+orial
+userry
+suff
+eth
+moto
+showere
+hunty
+nucle
+witz
+ider
+mune
+flagicont
+aire
+shtm
+fri
+guary
+rigion
+mome
+afted
+heate
+mo
+formedia
+oble
+cnn
+nea
+ha
+canage
+borning
+tol
+navbox
+ava
+octors
+md
+frent
+offical
+wal
+arent
+anon
+artish
+cration
+rell
+dir
+auding
+ira
+butional
+labout
+bott
+southe
+dence
+sporth
+formall
+ps
+intry
+thamp
+stube
+unus
+usering
+ron
+usert
+conse
+brote
+lib
+oces
+bg
+proad
+tas
+werent
+chese
+lyn
+atta
+pris
+oile
+drage
+unre
+whenry
+ement
+stada
+chara
+raphy
+tection
+rica
+reletion
+foothe
+titler
+lassion
+areen
+theim
+conship
+formand
+ution
+unio
+ided
+guid
+riging
+brited
+buch
+septem
+eleted
+conds
+preser
+ta
+ort
+feld
+ine
+actional
+alize
+tex
+fer
+aused
+engly
+eque
+sevel
+atted
+smalle
+thome
+alty
+reath
+gation
+prefere
+hd
+nr
+peration
+covember
+govember
+livision
+fren
+dq
+aren
+deve
+va
+ffcc
+jo
+prian
+afl
+gion
+medy
+stategory
+nomich
+tured
+mel
+ene
+appa
+licy
+linal
+rought
+areate
+bethe
+othey
+eral
+opport
+ara
+kw
+whild
+cance
+communited
+varing
+brothe
+pia
+mu
+sched
+socian
+poin
+vis
+obs
+squal
+paking
+sma
+convers
+noing
+maken
+cathe
+ni
+solor
+alry
+flagicons
+aaaaa
+weren
+artical
+mtv
+flagion
+chall
+alm
+upons
+dister
+searly
+sur
+towned
+oring
+ests
+avg
+whis
+publicy
+ec
+thange
+gp
+ced
+pes
+whick
+figh
+peted
+fed
+sween
+discore
+sil
+ds
+sournal
+tay
+dier
+kim
+adver
+reak
+mote
+wouldn
+paign
+inal
+medit
+pointo
+cata
+sper
+pts
+idw
+mlb
+guis
+cational
+ul
+wasn
+septer
+aird
+throm
+fe
+feated
+collo
+itary
+uni
+uple
+distory
+disa
+regist
+vere
+effor
+whican
+ober
+bettle
+anning
+cution
+speciate
+palign
+ai
+cit
+moview
+flagicon
+af
+covery
+govery
+lina
+reast
+creen
+texand
+exten
+vic
 

### 2013-09-22 tasbot/scopefun.cc
Remove DOS line endings everywhere

@@ -62,11 +62,12 @@ struct Graphic {
 struct ScopeFun {
   ScopeFun(const string &game,
 	   const string &moviename,
-	   int sf, int mf, bool so) : 
+	   int sf, int mf, bool so, bool fo) : 
     controller("controller.png"),
     controllerdown("controllerdown.png"),
     game(game),
-    soundonly(so) {
+    soundonly(so),
+    fatobjectivelines(fo) {
 
     Emulator::Initialize(game + ".nes");
     objectives = WeightedObjectives::LoadFromFile(game + ".objectives");
@@ -181,12 +182,14 @@ struct ScopeFun {
     }
   }
 
+  // XXX This function is used inconsistently; either remove
+  // the alpha param or use it...
   inline void WritePixel(int x, int y,
 			 uint8 r, uint8 g, uint8 b, uint8 a) {
-    CHECK(x >= 0);
-    CHECK(y >= 0);
-    CHECK(x < width);
-    CHECK(y < height);
+    DCHECK(x >= 0);
+    DCHECK(y >= 0);
+    DCHECK(x < width);
+    DCHECK(y < height);
 
     rgba[y * width * 4 + x * 4 + 0] = r;
     rgba[y * width * 4 + x * 4 + 1] = g;
@@ -194,14 +197,27 @@ struct ScopeFun {
     rgba[y * width * 4 + x * 4 + 3] = 0xFF;
   }
 
-  inline void WritePixelTo(int x, int y,
-			   uint8 r, uint8 g, uint8 b, uint8 a,
-			   uint8 *vec,
-			   int ww, int hh) {
-    CHECK(x >= 0);
-    CHECK(y >= 0);
-    CHECK(x < ww);
-    CHECK(y < hh);
+  inline uint32 GetPixel(int x, int y) {
+    DCHECK(x >= 0);
+    DCHECK(y >= 0);
+    DCHECK(x < width);
+    DCHECK(y < height);
+
+    uint32 r = rgba[y * width * 4 + x * 4 + 0];
+    uint32 g = rgba[y * width * 4 + x * 4 + 1];
+    uint32 b = rgba[y * width * 4 + x * 4 + 2];
+    uint32 a = rgba[y * width * 4 + x * 4 + 3];
+    return (r << 24) | (g << 16) | (b << 8) | a;
+  }
+
+  static inline void WritePixelTo(int x, int y,
+				  uint8 r, uint8 g, uint8 b,
+				  uint8 *vec,
[…]

### 2013-09-22 tasbot/tasbot.h

@@ -17,6 +17,8 @@
 
 #define TASBOT_SAMPLE_RATE 44100
 
+#define DEBUGGING 0
+
 #ifdef __GNUC__
 namespace std {
 using namespace __gnu_cxx;
@@ -37,6 +39,14 @@ struct hash< unsigned long long > {
 #define CHECK(condition) \
   while (!(condition)) {                                    \
     fprintf(stderr, "%s:%d. Check failed: %s\n",            \
+            __FILE__, __LINE__, #condition                  \
+            );                                              \
+    abort();                                                \
+  }
+
+#define DCHECK(condition) \
+  while (DEBUGGING && !(condition)) {                       \
+    fprintf(stderr, "%s:%d. DCheck failed: %s\n",           \
             __FILE__, __LINE__, #condition                  \
             );                                              \
     abort();                                                \

### 2013-09-22 tasbot/weighted-objectives.cc

@@ -57,18 +57,20 @@ WeightedObjectives::LoadFromFile(const string &filename) {
   WeightedObjectives *wo = new WeightedObjectives;
   vector<string> lines = Util::ReadFileToLines(filename);
   for (int i = 0; i < lines.size(); i++) {
-    stringstream ss(lines[i], stringstream::in);
-    double d;
-    ss >> d;
-    vector<int> locs;
-    while (!ss.eof()) {
-      int i;
-      ss >> i;
-      locs.push_back(i);
+    string line = lines[i];
+    Util::losewhitel(line);
+    if (!line.empty() && !Util::startswith(line, "#")) {
+      stringstream ss(line, stringstream::in);
+      double d;
+      ss >> d;
+      vector<int> locs;
+      while (!ss.eof()) {
+	int i;
+	ss >> i;
+	locs.push_back(i);
+      }
+      wo->weighted.insert(make_pair(locs, new Info(d)));
     }
-
-    // printf("GOT: %f | %s\n", d, ObjectiveToString(locs).c_str());
-    wo->weighted.insert(make_pair(locs, new Info(d)));
   }
 
   return wo;

### 2013-09-22 trunk/tasbot/learnfun.cc
Remove DOS line endings everywhere

@@ -139,6 +139,12 @@ static void MakeObjectives(const string &game,
 
 int main(int argc, char *argv[]) {
   map<string, string> config = Util::ReadFileToMap("config.txt");
+  if (config.empty()) {
+    fprintf(stderr, "You need a file called config.txt; please "
+	    "take a look at the README.\n");
+    abort();
+  }
+
   const string game = config["game"];
   const string moviename = config["movie"];
 

### 2013-09-22 trunk/tasbot/playfun.cc

@@ -165,6 +165,12 @@ static void SaveFuturesHTML(const vector<Future> &futures,
 struct PlayFun {
   PlayFun() : watermark(0), log(NULL), rc("playfun") {
     map<string, string> config = Util::ReadFileToMap("config.txt");
+    if (config.empty()) {
+      fprintf(stderr, "You need a file called config.txt; please "
+	      "take a look at the README.\n");
+      abort();
+    }
+
     game = config["game"];
     const string moviename = config["movie"];
     CHECK(!game.empty());

### 2013-09-22 cc-lib/util.cc
Style and credits

@@ -45,6 +45,7 @@ string dtos(double d) {
   return (string)s;
 }
 
+namespace {
 struct linereal : public line {
   int x0, y0, x1, y1;
   int dx, dy;
@@ -52,26 +53,26 @@ struct linereal : public line {
   int frac;
 
   virtual ~linereal() {}
-  
-  linereal(int x0_, int y0_, int x1_, int y1_) : 
+
+  linereal(int x0_, int y0_, int x1_, int y1_) :
     x0(x0_), y0(y0_), x1(x1_), y1(y1_) {
-    
+
 
     dy = y1 - y0;
     dx = x1 - x0;
 
-    if (dy < 0) { 
+    if (dy < 0) {
       dy = -dy;
-      stepy = -1; 
+      stepy = -1;
     } else {
-      stepy = 1; 
+      stepy = 1;
     }
 
     if (dx < 0) {
       dx = -dx;
       stepx = -1;
-    } else { 
-      stepx = 1; 
+    } else {
+      stepx = 1;
     }
 
     dy <<= 1;
@@ -113,7 +114,7 @@ struct linereal : public line {
       }
     }
   }
-  
+
   virtual void destroy() {
     delete this;
   }
@@ -123,6 +124,7 @@ struct linereal : public line {
 line * line::create(int a, int b, int c, int d) {
   return new linereal(a, b, c, d);
 }
+}  // namespace
 
 bool Util::isdir(string f) {
   struct stat st;
@@ -159,7 +161,6 @@ string Util::ReadFile(const string &s) {
   if (s == "") return "";
 
   FILE * f = fopen(s.c_str(), "rb");
-
   if (!f) return "";
   fseek(f, 0, SEEK_END);
   int size = ftell(f);
@@ -234,7 +235,7 @@ static bool hasmagicf(FILE * f, const string & mag) {
       return false;
     }
   }
-  
+
   free(hdr);
   return true;
 }
@@ -244,7 +245,7 @@ bool Util::hasmagic(string s, const string &mag) {
   if (!f) return false;
 
   bool hm = hasmagicf(f, mag);
-  
+
   fclose(f);
   return hm;
 }
@@ -294,11 +295,12 @@ bool Util::WriteFile(const string &fn, const string &s) {
   fwrite(s.c_str(), 1, s.length(), f);
 
   fclose(f);
-  
+
   return true;
 }
 
-bool Util::WriteFileBytes(const string &fn, const vector<unsigned char> &bytes) {
+bool Util::WriteFileBytes(const string &fn,
[…]

### 2013-12-23 powerhour/paper/paper.tex
Wrote an introduction and about the solo part

\documentclass[twocolumn]{article}
\usepackage[top=1.1in, left=0.85in, right=0.85in]{geometry}

\usepackage{cite}
% \usepackage{eclbkbox}
\usepackage{amsmath}
\usepackage{amssymb}
% \usepackage{code}
% \usepackage{amscd}
% \usepackage{xy}
\usepackage{graphicx}
% \usepackage{fancyhdr}
% \usepackage{color}
% \usepackage[dark,all,bottom,landscape,timestamp]{draftcopy}
% \usepackage{everypage}

% \pagestyle{empty}

\usepackage{ulem}
% go back to italics for emphasis, though
\normalem

% \usepackage{inlinebib}

\newcommand\sfrac[2]{{}\,^{#1}\!/{}\!_{#2}}

\begin{document}

\title{New results in $\sfrac{k}{n}$ Power-Hours}
\author{Dr.~Tom~Murphy~VII~Ph.D.\thanks{
Copyright \copyright\ 2014 the Regents of the Wikiplia
Foundation. Appears in SIGBOVIK 2014 with the chagrin of the
Association for Computational Heresy; {\em IEEEEEE!} press,
Verlag-Verlag volume no.~0x40-2A.
\yen 0.00}
}

\renewcommand\>{$>$}
\newcommand\<{$<$}
\newcommand\kn{\ensuremath{\sfrac{k}{n}\,}}

\newcommand\any{\ensuremath{\textrm{?}}}
\newcommand\nocup{\text{\sout{\ensuremath{\cup}}}}
\newcommand\fullcup{\ensuremath{\uplus}}
\newcommand\emptycup{\ensuremath{\cup}}
\newcommand\overcup{\ensuremath{\cap}}

\newcommand\nodrink{\ensuremath{\Rightarrow}}
\newcommand\drink{\ensuremath{\stackrel{{}^{\textrm{+}}}{\Rightarrow}}}

\date{1 April 2014}

\maketitle

\begin{abstract}
Something about the k-n hours

\end{abstract}

\vspace{1em}
{\noindent \small {\bf Keywords}:
  generalized binge drinking, maths,
  finite-state automata,
  abstract interpretation
}

\section*{Introduction}
A 2012 paper by Blum, Martens, Murphy, and Lovas\cite{algorithms}
introduced the \kn Power-Hour, a fractional variant on the well-known
drinking game. In a traditional Power-Hour, participants drink one
shot of beer per minute for 60 minutes. Since 5 beers in an hour
sometimes have adverse effects, some players opt for an attenuated
version of the game wherein fewer than 60 shots are consumed. However,
since the game is frantic and played simultaneous with others, it is
critical to have a mechanical procedure for performing the attenuated
Hour. The framework by Blum {\em et al.}, hereafter BMML, gives a
handful of simple operations that can be used to define a state
machine among $p$ players:
\begin{itemize}
  \item At the beginning of each minute, each player has at most one
    shot glass in front of him or her
  \item The shot glass must be in one of three states: Filled \fullcup,
    empty \emptycup, or overturned \overcup
  \item Atomically, each player performs an action based only on the
    state of his or her cup. If not in possession of a cup (written
    \nocup), the only action is to do nothing. With a cup:
  \begin{itemize}
    \item The player may drink \drink, or not drink \nodrink
    \item The player may pass the cup in any state to any player (a
      fixed player per action)
    \item However: If the cup is filled and the player did not drink,
      it must be passed in the filled state
    \item A player may not receive more than one cup in the same round
  \end{itemize}
\end{itemize}

Every assignment of rules and starting condition to $p$ players yields
a deterministic outcome, though some of these are illegal (because
they result in two or more cups being passed to the same player in
some round). For legal games, the outcome is that the $p$ players have
consumed $k_i$ shots of beer where $1 \leq i \leq p$ and $0 \leq k_i
\leq 60$. For the traditional power hour, the player starts with an
empty cup, at each step drinks,\!\footnote{In practice, this is done
  by filling the cup and then drinking it.} leaves the cup empty, and
passes to herself.

While the authors made a mostly clear definition of BMML and presented
some initial results, these results contained multiple serious errors
and the paper abruptly switches notation and assumptions several
times, and rambles incoherently. By their own admission, the authors
were drinking while they wrote it, taking only one hour to do so.
Don't drink and derive, kids!

This paper revisits the problem of BMML from a modern, sober
perspective, clarifies some of the original results, and presents
several new ones and a few conjectures.

\section{One-player \kn Power-Hours}

The goal of the \kn Power-Hour is to attenuate the number of drinks
consumed by the $p$ players, and its expressive power comes from the
ability to encode some state in the orientation of the cups, and
propagate that state via passing them from player to player. Even
without passing cups, the ability for a single player to attenuate his
drinking is nontrivial. Playing drinking games alone is sad indeed,
but the solo \kn Power-Hour still has practical applications. When
playing a Power Hour with others, if each player's desired $k$ is
attainable through solo methods then there is no need for passing
cups, which simplifies the ergonomics considerably. A common case is
where some of the players would like to do half--Power-Hours, which is
easily achieved in BMML by transitioning \emptycup\ to \fullcup\ without
drinking and \fullcup\ to \emptycup\ by drinking, and passing to
oneself.\!\footnote{There are many variations, but this was the
  strategy used many times in practice before being generalized to
  BMML.}

A full list of attainable \kn Power-Hours where $p=1$ appears in
Figure~\ref{fig:solo}. Possible values of $k$ are $\{ 0, 1, 2, 20, 29,
30, 31, 40, 58, 59, 60 \}$. The BMML paper claimed that the possible
values were $\{0, 1, 2, 20, 30, 40, 58, 59, 60 \}$, describing 31 for
example as ``super impossible.'' Achieving 31 is somewhat interesting.
One way to do it is to start with \emptycup, and use the rule that
\emptycup\ means drink and then fill the cup. We then use the rule
that \fullcup\ means drink and flip the cup, and \overcup\ means don't
drink and fill the cup. Essentially we use \emptycup\ to mean ``this
is the very first state'' and then take shots on alternating minutes
by using \overcup\ and \fullcup\ to encode the parity. Exploiting
non-steady-states like this (Figure~\ref{fig:solo31}) is how we
achieve $k$ that does not divide $n$.

\begin{figure}[ht]
\begin{center}
\includegraphics[width=0.90 \linewidth]{solo31.pdf}
\end{center}\vspace{-0.1in}
\caption{State machine that achieves $k=31$ in a solo BMML Power Hour.
  $+$ on an edge means the player drinks. The player always passes to
  herself.}
\label{fig:solo31}
\end{figure}

It is tractable to work out the possibilities for the solo case by
hand, though apparently not while drinking~\cite{algorithms}. These
results were generated by a computer program, which is probably
necessary for $p>1$. The next section describes how we explore this
space computationally.

\begin{figure}[ht]
\begin{center}
\begin{tabular}{c|ll}
$k$ & start & rules \\
\hline
0  &   \nocup    &  \emptycup \nodrink \any     , \overcup \nodrink \any      , \fullcup \nodrink \any \\
1  &   \emptycup &  \emptycup \drink \overcup   , \overcup \nodrink \overcup  , \fullcup \nodrink \any \\
2  &   \emptycup &  \emptycup \drink \fullcup   , \overcup \nodrink \overcup  , \fullcup \drink \overcup \\
20 &   \emptycup &  \emptycup \nodrink \overcup , \overcup \nodrink \fullcup  , \fullcup \drink \emptycup \\
29 &   \emptycup &  \emptycup \nodrink \overcup , \overcup \nodrink \fullcup  , \fullcup \drink \overcup \\
30 &   \emptycup &  \emptycup \drink \overcup   , \overcup \nodrink \emptycup , \fullcup \nodrink \any \\
31 &   \emptycup &  \emptycup \drink \fullcup   , \overcup \nodrink \fullcup  , \fullcup \drink \overcup \\
40 &   \emptycup &  \emptycup \drink \overcup   , \overcup \nodrink \fullcup  , \fullcup \drink \emptycup \\
58 &   \emptycup &  \emptycup \nodrink \overcup , \overcup \nodrink \fullcup  , \fullcup \drink \fullcup \\
59 &   \emptycup &  \emptycup \nodrink \overcup , \overcup \drink \overcup    , \fullcup \nodrink \any \\
60 &   \emptycup &  \emptycup \drink \emptycup  , \overcup \nodrink \any      , \fullcup \nodrink \any \\
\end{tabular}
\end{center}
\label{fig:solo}
\caption{All the possible $k$ for a solo Power-Hour in BMML. A
  superscript $+$ means that the player drinks. The symbol \any\ means
  that any cup state can be used in that position. Note that 29 and 58
  require wasting a shot of beer; all the others but 31 permit a
  variant where a shot is wasted as well.}
\end{figure}

\section{Two-player \kn Power-Hours}

For two players, it is easy to see how one of them could

\begin{figure}
\begin{center}
\includegraphics[width=0.90 \linewidth]{3and2.pdf}
\end{center}\vspace{-0.1in}
\caption{Outcomes possible for the first two players in all different
  3-player power hours (black), overlayed by all possible outcomes
  outcomes for 2-player power hours (red). Mainly included because
  it looks pretty sweet.
}
\label{fig:3and2}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=0.90 \linewidth]{powerday3.pdf}
\end{center}\vspace{-0.1in}
\caption{All possible outcomes for the first two players in 3-player
  power days. These are the same games as the 3-player power hours,
  but at this scale makes it clear the groupings and their sparsity
  in the limit. Lines plotted from (0,0) to (60,*) and (*,60) show
  significant structure, but don't explain some of the interior points.
  (TODO: what are these games? what did the other player drink?)
}
\label{fig:powerday3}
\end{figure}

\section{Some stuffff}

Hi

\bibliography{paper}{}
\bibliographystyle{plain}
\end{document}

### 2013-12-23 trunk/powerhour/paper/paper.tex
More work on paper. Implemented optimization but it didn't reduce the number of states even

@@
 
+\usepackage{code}
+\usepackage{cite}
 % \usepackage{eclbkbox}
@@
 
-% \usepackage{ulem}
+\usepackage{ulem}
 % go back to italics for emphasis, though
-% \normalem
+\normalem
 
+% \usepackage{inlinebib}
+
+\newcommand\comment[1]{}
 \newcommand\sfrac[2]{{}\,^{#1}\!/{}\!_{#2}}
@@
 \author{Dr.~Tom~Murphy~VII~Ph.D.\thanks{
-Copyright \copyright\ 2012 the Regents of the Wikiplia
-Foundation. Appears in SIGBOVIK 2013 with the chagrin of the
+Copyright \copyright\ 2014 the Regents of the Wikiplia
+Foundation. Appears in SIGBOVIK 2014 with the chagrin of the
 Association for Computational Heresy; {\em IEEEEEE!} press,
@@
 
+\renewcommand\th{\ensuremath{{}^{\textrm{th}}}}
+\newcommand\st{\ensuremath{{}^{\textrm{st}}}}
+\newcommand\rd{\ensuremath{{}^{\textrm{rd}}}}
+\newcommand\nd{\ensuremath{{}^{\textrm{nd}}}}
+\newcommand\at{\ensuremath{\scriptstyle @}}
+
 \renewcommand\>{$>$}
 \newcommand\<{$<$}
+\newcommand\kn{\ensuremath{\sfrac{k}{n}\,}}
 
-\date{1 April 2011}
+\newcommand\any{\ensuremath{\textrm{?}}}
+\newcommand\nocup{\text{\sout{\ensuremath{\cup}}}}
+\newcommand\fullcup{\ensuremath{\uplus}}
+\newcommand\emptycup{\ensuremath{\cup}}
+\newcommand\overcup{\ensuremath{\cap}}
+
+\newcommand\nodrink{\ensuremath{\Rightarrow}}
+\newcommand\drink{\ensuremath{\stackrel{{}^{\textrm{+}}}{\Rightarrow}}}
+\newcommand\qdrink{\ensuremath{\stackrel{{}^{\textrm{?}}}{\Rightarrow}}}
+
+\date{1 April 2014}
 
@@
 \section*{Introduction}
-Here's where I tell you about doo-dah and so-so.
+A 2012 paper by Blum, Martens, Murphy, and Lovas\cite{algorithms}
+introduced the \kn Power-Hour, a fractional variant on the well-known
+drinking game. In a traditional Power-Hour, participants drink one
+shot of beer per minute for 60 minutes. Since 5--6 beers in an hour
+sometimes have adverse effects, some players opt for an attenuated
+version of the game wherein fewer than 60 shots are consumed. However,
+since the game is frantic and played simultaneous with others, it is
+critical to have a mechanical procedure for performing the attenuated
+Hour. The framework by Blum {\em et al.}, hereafter BMML, gives a
+handful of simple operations that can be used to define a state
+machine among $p$ players:
+\begin{itemize}
+  \item At the beginning of each minute, each player has at most one
+    shot glass in front of him or her
+  \item The shot glass must be in one of three states: Filled \fullcup,
+    empty \emptycup, or overturned \overcup
+  \item Atomically, each player performs an action based only on the
+    state of his or her cup. If not in possession of a cup (written
+    \nocup), the only action is to do nothing. With a cup:
+  \begin{itemize}
+    \item The player may drink \drink, or not drink \nodrink
+    \item The player may pass the cup in any state to any player (a
+      fixed player per action)
+    \item However: If the cup is filled and the player did not drink,
+      it must be passed in the filled state
+    \item A player may not receive more than one cup in the same round
+  \end{itemize}
+\end{itemize}
+
+Every assignment of rules and starting condition to $p$ players yields
+a deterministic outcome, though some of these are illegal (because
+they result in two or more cups being passed to the same player in
+some round). For legal games, the outcome is that the $p$ players have
+consumed $k_i$ shots of beer where $1 \leq i \leq p$ and $0 \leq k_i
+\leq 60$. For the traditional power hour, the player starts with an
+empty cup, at each step drinks,\!\footnote{In practice, this is done
+  by filling the cup and then drinking it.} leaves the cup empty, and
+passes to herself.
+
+While the authors made a mostly clear definition of BMML and presented
+some initial results, these results contained multiple serious errors
+and the paper abruptly switches notation and assumptions several
+times, and rambles incoherently. By their own admission, the authors
+were drinking while they wrote it, taking only one hour to do so.
+Don't drink and derive, kids!
+
+This paper revisits the problem of BMML from a modern, sober
+perspective, clarifies some of the original results, and presents
+several new ones and a few conjectures.
+
+\section{One-player \kn Power-Hours}
+
+The goal of the \kn Power-Hour is to attenuate the number of drinks
+consumed by the $p$ players, and its expressive power comes from the
+ability to encode some state in the orientation of the cups, and
+propagate that state via passing them from player to player. Even
+without passing cups, the ability for a single player to attenuate his
+drinking is nontrivial. Playing drinking games alone is sad indeed,
+but the solo \kn Power-Hour still has practical applications. When
+playing a Power Hour with others, if each player's desired $k$ is
+attainable through solo methods then there is no need for passing
+cups, which simplifies the ergonomics considerably. A common case is
+where some of the players would like to do half--Power-Hours, which is
+easily achieved in BMML by transitioning \emptycup\ to \fullcup\ without
+drinking and \fullcup\ to \emptycup\ by drinking, and passing to
+oneself.\!\footnote{There are many variations, but this was the
+  strategy used many times in practice before being generalized to
+  BMML.}
+
+A full list of attainable \kn Power-Hours where $p=1$ appears in
+Figure~\ref{fig:solo}. Possible values of $k$ are $\{ 0, 1, 2, 20, 29,
+30, 31, 40, 58, 59, 60 \}$. The BMML paper claimed that the possible
+values were $\{0, 1, 2, 20, 30, 40, 58, 59, 60 \}$, describing 31 for
+example as ``super impossible.'' Achieving 31 is somewhat interesting.
+One way to do it is to start with \emptycup, and use the rule that
+\emptycup\ means drink and then fill the cup. We then use the rule
+that \fullcup\ means drink and flip the cup, and \overcup\ means don't
+drink and fill the cup. Essentially we use \emptycup\ to mean ``this
+is the very first state'' and then take shots on alternating minutes
+by using \overcup\ and \fullcup\ to encode the parity. Exploiting
+non-steady-states like this (Figure~\ref{fig:solo31}) is how we
+achieve $k$ that does not divide $n$.
+
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[width=0.90 \linewidth]{solo31.pdf}
+\end{center}\vspace{-0.1in}
+\caption{State machine that achieves $k=31$ in a solo BMML Power Hour.
+  $+$ on an edge means the player drinks. The disembodied incoming
+  edge is the start state. The player always passes to herself.}
+\label{fig:solo31}
+\end{figure}
+
+It is tractable to work out the possibilities for the solo case by
+hand, though apparently not while drinking~\cite{algorithms}. These
+results were generated by a computer program, which is probably
+necessary for $p>1$. The next section describes how we explore this
+space computationally. XXX no it doesn't
+
+\begin{figure}[ht]
+\begin{center}
+\begin{tabular}{c|ll}
+$k$ & start & rules \\
+\hline
+0  &   \nocup    &  \emptycup \nodrink \any     , \overcup \nodrink \any      , \fullcup \nodrink \any \\
+1  &   \emptycup &  \emptycup \drink \overcup   , \overcup \nodrink \overcup  , \fullcup \nodrink \any \\
+2  &   \emptycup &  \emptycup \drink \fullcup   , \overcup \nodrink \overcup  , \fullcup \drink \overcup \\
+20 &   \emptycup &  \emptycup \nodrink \overcup , \overcup \nodrink \fullcup  , \fullcup \drink \emptycup \\
+29 &   \emptycup &  \emptycup \nodrink \overcup , \overcup \nodrink \fullcup  , \fullcup \drink \overcup \\
+30 &   \emptycup &  \emptycup \drink \overcup   , \overcup \nodrink \emptycup , \fullcup \nodrink \any \\
+31 &   \emptycup &  \emptycup \drink \fullcup   , \overcup \nodrink \fullcup  , \fullcup \drink \overcup \\
+40 &   \emptycup &  \emptycup \drink \overcup   , \overcup \nodrink \fullcup  , \fullcup \drink \emptycup \\
+58 &   \emptycup &  \emptycup \nodrink \overcup , \overcup \nodrink \fullcup  , \fullcup \drink \fullcup \\
+59 &   \emptycup &  \emptycup \nodrink \overcup , \overcup \drink \overcup    , \fullcup \nodrink \any \\
+60 &   \emptycup &  \emptycup \drink \emptycup  , \overcup \nodrink \any      , \fullcup \nodrink \any \\
+\end{tabular}
+\end{center}
+\label{fig:solo}
+\caption{All the possible $k$ for a solo Power-Hour in BMML. A
+  superscript $+$ means that the player drinks. The symbol \any\ means
+  that any cup state can be used in that position. Note that 29 and 58
+  require wasting a shot of beer; all the others but 31 permit a
+  variant where a shot is wasted as well.}
+\end{figure}
+
+\section{Two-player \kn Power-Hours}
+
+For more players, the number of possible configurations explodes.
+Let's make the following definitions to bound the size:
+\begin{itemize}
+\item $s = 4$, the number of starting states (\fullcup, \emptycup,
+  \overcup, \nocup)
+\item $a = 2 * p * 3$, the number of actions given a cup. The player
+  can drink or not drink, pass to any player, and in 3 configurations
+  (\fullcup, \emptycup, \overcup)
+\end{itemize}
+Then the number of configurations is bounded by $(s * a^3)^p$. For
+$p=1$ this was just 864. For $p=2$ it is 47,775,744; for $p=3$ it's
+12,694,994,583,552, already beyond the limits of straight enumeration.
+
+\comment{
+fun ct (p : IntInf.int) =
+  let val s = 4
+      fun pow n 0 = 1
+        | pow n m = n * pow n (m - 1)
+      val a = 2 * p * 3
+   in
+      pow (s * pow a 3) p
+  end
+}
+
+However, this is just an upper bound. For one thing, the base of the
+exponent is actually bounded by
+$$
+  s * a^2 * a_{\textrm{filled}}
+$$
+where
+$a_{\textrm{filled}} = (p * 3) + p$ (the actions that can be taken on
+\fullcup, where if the player does not drink, then he must pass the
+cup \fullcup).
+
+\comment{
+fun ctb (p : IntInf.int) =
+  let val s = 4
+      fun pow n 0 = 1
+        | pow n m = n * pow n (m - 1)
+      val af = p * 3 + p
+      val a = 2 * p * 3
+   in
+      pow (s * pow a 2 * af) p
+  end
+}
+
+The values for $p \in \{1,2,3\}$ are still 576; 21,233,664;
+3,761,479,876,608. There are a few other simplifications possible.
+Many of these games are illegal because they result in multiple cups
+being passed to the same player in some turn. These are difficult to
+exclude analytically, but there are some sufficient conditions; for
+example, if two players pass to the same player no matter their input
+state, then their cups always collide. There are also many games
+that are isomorphic. For one thing, \emptycup\ and \overcup\ are not
+distinguished in the rules at all, so any two configurations where
+these are simply swapped has the exact same outcome. Likewise for
+permuting the players.
+
+21 million configurations is no big deal for a modern computer. A
+simple SML program computes all of the configurations and runs them;
+ones that are found to be illegal are rejected. (It implements the
+first simplification having to do with \fullcup\ when generating the
+configurations, since it can be done statically.) All of the possible
+outcomes are shown as black squares in Figure~\ref{fig:powerhour2player}.
+
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[width=0.90 \linewidth]{powerhour2player.pdf}
+\end{center}\vspace{-0.1in}
+\caption{All of the possible outcomes ($k$) for the two players in a
+  BMML Power-Hour. The matrix is symmetric, of course, since the players
+  are interchangeable.}
+\label{fig:powerhour2player}
+\end{figure}
+
+Each cell represents a pair of $\langle k_1, k_2 \rangle$ for the
+number of shots imbibed by players 1 and 2. 454 of the $61^2 = 3721$
+combinations are achievable. Note that column 0 represents the case
+where player 1 drinks nothing. It dominates the matrix in the sense
+that if $\langle k_1, k_2 \rangle$ is achievable, then $\langle 0, k_2
+\rangle$ is as well. Most of the time it is easy to see how this is
+done: Take the configuration that produces $\langle k_1, k_2 \rangle$
+and do the same, but player 1 simply performs her actions without
+drinking. This works except for the case where player 1 receives a
+\fullcup\ and passes it in a state other than \fullcup. The player
+can't simply not drink, as this is illegal (the beer must be emptied,
+and BMML does not permit such messy reductions). It is curious that
+this does not affect the result; I discuss this further in
+Section~\ref{sec:conjectures}. Another interesting column is the last
+one, which represents outcomes of the form $\langle 60, k_2 \rangle$,
+where the first player achieves a full Power-Hour. This of course
+includes all of the $k_2$ achievable solo (the players can just do
+their thing without interacting). But some new $k$ are now achievable:
+$\{ 3, 4, 15, 28, 32, 45, 56, 57 \}$. Interacting with a player doing
+a full Power-Hour still affords us a few additional bits of
+information that can be used to attenuate the other player's
+consumption. The solution for $45$ is instructive, and appears in
+Figure~\ref{fig:dual45}.
+
+This is a useful result, but it may be the case that someone wants to
+drink exactly 27 shots of beer, which is not possible with just two
+players in BMML. There are two avenues to explore: Adding more 
+players, and generalizing BMML. We begin with the three-player case.
+
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[width=0.90 \linewidth]{dual45.pdf}
+\end{center}\vspace{-0.1in}
+\caption{State machine that achieves $\langle k_1=45, k_2=60 \rangle$
+  in a two-player BMML Power Hour. The bottom row of states are for
+  player 1, who drinks 45, and the top for player 2, who drinks 60.
+  Clearly, player 2 must drink at every step. The players always pass
+  to each other, with the two cups exchanging hands each turn. The
+  cycles for the two cups are disconnected; one alternates between
+  \fullcup\ in player 2's hand and \emptycup\ in player 1's (cycle
+  of length 2), drinking on each turn. This cycle yields 30 drinks
+  for each player. The other cycle is of length 4; player 2 drinks
+  on every step (as we know), and player 1 every 4\th\ step, yielding
+  15 more drinks for a total of 45.
+% [60,45] wasting 2: 12 game(s) like 
+% [(start F, U=>D*@1 D=>F*@1 F=>U*@1),
+%  (start F, U=>F*@0 D=>D@0 F=>U*@0)]
+}
+\label{fig:dual45}
+\end{figure}
+
+\section{Three-player \kn Power-Hours}
+
+With 3.7 trillion possible configurations, enumeration is not
+feasible. But as we observed before, many of these combinations are
+illegal (they result in a player recieving two cups), and many are
+isomorphic to one another. By being clever about how we explore the
+configurations, testing ``all'' the three-player configurations
+becomes feasible.
+
+Here is a one-player BMML configuration that illustrates a particular
+kind of redundancy:
+\vspace{1em}
+\begin{tabular}{cc}
+start \emptycup & \emptycup \drink \emptycup 
+ \quad \fullcup \drink \overcup
+ \quad \overcup \nodrink \fullcup
+\end{tabular}
+
+The cup starts empty, and at each step the player fills it and drinks
+(traditional Power-Hour). The player also has rules for the case that
+she observes a full or overturned cup. {\em It does not matter what
+  these are} because they can never be used. This example is trivial,
+but there are many ways that the execution of a configuration can be
+indifferent to some of its content. Another is a two player
+configuration like
+\vspace{1em}
+\begin{tabular}{rcccc}
+player 1 & start \emptycup & \emptycup \drink \overcup \at 1
+ & \fullcup \drink \overcup \at 2
+ & \overcup \nodrink \fullcup \at 1 \\
+player 2 & start \emptycup & \emptycup \nodrink \emptycup \at 1
+ & \fullcup \drink \overcup \at 1
+ & \overcup \drink \fullcup \at 2 \\
+\end{tabular}
+where the $\at n$ notation means to pass the cup in that state to
+player $n$. In this case, the first thing the players do is to
+pass both of their cups to player 1, which is illegal and ends the
+game. Again, none of the other rules are ever used.
+
+In order to explore what is possible in three-player games, we exploit
+this redundancy with a technique like abstract
+interpretation~\cite{abstract}. The start state is always used, so we
+begin by enumerating all assignments of start states to players. There
+are only $4^p$. Every other rule starts out undetermined, maybe written
+like this:
+
+\begin{tabular}{cc}
+start \emptycup & \emptycup \qdrink \any
+ \quad \fullcup \qdrink \any
+ \quad \overcup \qdrink \any
+\end{tabular}
+
+Now we execute programs as before, and hope that we never encounter a
+situation where we depend on a rule. If we finish without ever using
+one of the \any\ rules, we evaluated a potentially large group of
+configurations all at once. During the execution of a configuration,
+if we need to use a rule that is currently marked \any, we explore all
+of the possibilities. This is accomplished by a loop that looks like
+the following (in Pseudo SML):
+
+\begin{code}
+val queue = (* all abstract configurations *)
+val results = (* map from (k_1, k_2) 
+                 to example *)
+\codeskip
+fun loop nil = (* done *)
+  | loop (h :: t) =
+      let
+        val res = evaluate h
+      in
+        insert (results, res);
+        loop t
+      end handle Expand l => loop (l @ t)
+\codeskip
+fun evaluate config =
+   (* ... *) 
+   case rulefor cup of 
+      QuestionMark =>
+         raise Expand expandedconfigs
+    | (* ... *)
+\codeskip
+val () = loop queue
+\end{code}
+
+The key trick here for keeping the code under control is to
+iteratively evaluate the configurations as usual, but if we find a
+\any, then we abort the current simulation with an exception that
+carries along the set of configurations that expand the current one in
+just that position. This wastes some work (and we often need to
+restart multiple times per abstract configuration), but not much: If a
+rule is used at all, it is usually used in one of the first few
+rounds.
+
+With this technique, we can simulate all possible two-player games
+with just 15,744,259 game-minutes simulated (with naive enumeration it
+would be 1.2 billion) in less than 2 seconds on a crappy old computer.
+
+% Total steps 15744259. Took 1982 ms
+% Total steps 15744259. Took 2080 ms
 
@@
 
+\section{Known bounds}
+
+Define BMnL, where $n$ is the number of non-filled states. BMML=BM2L.
+Find $n$ such that a solo or dual power hour achieves any $k$.
+
+\section{Conjectures} \label{sec:conjectures}
+
+With two friends, you can drink any amount
+
+Teetotaller: When $\langle k_1, \ldots, k_i, \ldots, k_n \rangle$ is possible,
+so is $\langle k_1, \ldots, 0, \ldots, k_n \rangle$ (under what conditions)?
+
 \section{Some stuffff}
@@
 
+\bibliography{paper}{}
+\bibliographystyle{plain}
 \end{document}

### 2013-12-24 powerhour/paper/paper.tex
Add n-cupstate generalization for 1 player

@@
 
+\usepackage{code}
 \usepackage{cite}
@@
 
+\newcommand\comment[1]{}
 \newcommand\sfrac[2]{{}\,^{#1}\!/{}\!_{#2}}
@@
 
+\renewcommand\th{\ensuremath{{}^{\textrm{th}}}}
+\newcommand\st{\ensuremath{{}^{\textrm{st}}}}
+\newcommand\rd{\ensuremath{{}^{\textrm{rd}}}}
+\newcommand\nd{\ensuremath{{}^{\textrm{nd}}}}
+\newcommand\at{\ensuremath{\scriptstyle @}}
+
 \renewcommand\>{$>$}
@@
 \newcommand\drink{\ensuremath{\stackrel{{}^{\textrm{+}}}{\Rightarrow}}}
+\newcommand\qdrink{\ensuremath{\stackrel{{}^{\textrm{?}}}{\Rightarrow}}}
 
@@
 drinking game. In a traditional Power-Hour, participants drink one
-shot of beer per minute for 60 minutes. Since 5 beers in an hour
+shot of beer per minute for 60 minutes. Since 5--6 beers in an hour
 sometimes have adverse effects, some players opt for an attenuated
@@
 \caption{State machine that achieves $k=31$ in a solo BMML Power Hour.
-  $+$ on an edge means the player drinks. The player always passes to
-  herself.}
+  $+$ on an edge means the player drinks. The disembodied incoming
+  edge is the start state. The player always passes to herself.}
 \label{fig:solo31}
@@
 necessary for $p>1$. The next section describes how we explore this
-space computationally.
+space computationally. XXX no it doesn't
 
@@
 
-For two players, it is easy to see how one of them could
+For more players, the number of possible configurations explodes.
+Let's make the following definitions to bound the size:
+\begin{itemize}
+\item $s = 4$, the number of starting states (\fullcup, \emptycup,
+  \overcup, \nocup)
+\item $a = 2 * p * 3$, the number of actions given a cup. The player
+  can drink or not drink, pass to any player, and in 3 configurations
+  (\fullcup, \emptycup, \overcup)
+\end{itemize}
+Then the number of configurations is bounded by $(s * a^3)^p$. For
+$p=1$ this was just 864. For $p=2$ it is 47,775,744; for $p=3$ it's
+12,694,994,583,552, already beyond the limits of straight enumeration.
+
+\comment{
+fun ct (p : IntInf.int) =
+  let val s = 4
+      fun pow n 0 = 1
+        | pow n m = n * pow n (m - 1)
+      val a = 2 * p * 3
+   in
+      pow (s * pow a 3) p
+  end
+}
+
+However, this is just an upper bound. For one thing, the base of the
+exponent is actually bounded by
+$$
+  s * a^2 * a_{\textrm{filled}}
+$$
+where
+$a_{\textrm{filled}} = (p * 3) + p$ (the actions that can be taken on
+\fullcup, where if the player does not drink, then he must pass the
+cup \fullcup).
+
+\comment{
+fun ctb (p : IntInf.int) =
+  let val s = 4
+      fun pow n 0 = 1
+        | pow n m = n * pow n (m - 1)
+      val af = p * 3 + p
+      val a = 2 * p * 3
+   in
+      pow (s * pow a 2 * af) p
+  end
+}
+
+The values for $p \in \{1,2,3\}$ are still 576; 21,233,664;
+3,761,479,876,608. There are a few other simplifications possible.
+Many of these games are illegal because they result in multiple cups
+being passed to the same player in some turn. These are difficult to
+exclude analytically, but there are some sufficient conditions; for
+example, if two players pass to the same player no matter their input
+state, then their cups always collide. There are also many games
+that are isomorphic. For one thing, \emptycup\ and \overcup\ are not
+distinguished in the rules at all, so any two configurations where
+these are simply swapped has the exact same outcome. Likewise for
+permuting the players.
+
+21 million configurations is no big deal for a modern computer. A
+simple SML program computes all of the configurations and runs them;
+ones that are found to be illegal are rejected. (It implements the
+first simplification having to do with \fullcup\ when generating the
+configurations, since it can be done statically.) All of the possible
+outcomes are shown as black squares in Figure~\ref{fig:powerhour2player}.
+
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[width=0.90 \linewidth]{powerhour2player.pdf}
+\end{center}\vspace{-0.1in}
+\caption{All of the possible outcomes ($k$) for the two players in a
+  BMML Power-Hour. The matrix is symmetric, of course, since the players
+  are interchangeable.}
+\label{fig:powerhour2player}
+\end{figure}
+
+Each cell represents a pair of $\langle k_1, k_2 \rangle$ for the
+number of shots imbibed by players 1 and 2. 454 of the $61^2 = 3721$
+combinations are achievable. Note that column 0 represents the case
+where player 1 drinks nothing. It dominates the matrix in the sense
+that if $\langle k_1, k_2 \rangle$ is achievable, then $\langle 0, k_2
+\rangle$ is as well. Most of the time it is easy to see how this is
+done: Take the configuration that produces $\langle k_1, k_2 \rangle$
+and do the same, but player 1 simply performs her actions without
+drinking. This works except for the case where player 1 receives a
+\fullcup\ and passes it in a state other than \fullcup. The player
+can't simply not drink, as this is illegal (the beer must be emptied,
+and BMML does not permit such messy reductions). It is curious that
+this does not affect the result; I discuss this further in
+Section~\ref{sec:conjectures}. Another interesting column is the last
+one, which represents outcomes of the form $\langle 60, k_2 \rangle$,
+where the first player achieves a full Power-Hour. This of course
+includes all of the $k_2$ achievable solo (the players can just do
+their thing without interacting). But some new $k$ are now achievable:
+$\{ 3, 4, 15, 28, 32, 45, 56, 57 \}$. Interacting with a player doing
+a full Power-Hour still affords us a few additional bits of
+information that can be used to attenuate the other player's
+consumption. The solution for $45$ is instructive, and appears in
+Figure~\ref{fig:dual45}.
+
+This is a useful result, but it may be the case that someone wants to
+drink exactly 27 shots of beer, which is not possible with just two
+players in BMML. There are two avenues to explore: Adding more 
+players, and generalizing BMML. We begin with the three-player case.
+
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[width=0.90 \linewidth]{dual45.pdf}
+\end{center}\vspace{-0.1in}
+\caption{State machine that achieves $\langle k_1=45, k_2=60 \rangle$
+  in a two-player BMML Power Hour. The bottom row of states are for
+  player 1, who drinks 45, and the top for player 2, who drinks 60.
+  Clearly, player 2 must drink at every step. The players always pass
+  to each other, with the two cups exchanging hands each turn. The
+  cycles for the two cups are disconnected; one alternates between
+  \fullcup\ in player 2's hand and \emptycup\ in player 1's (cycle
+  of length 2), drinking on each turn. This cycle yields 30 drinks
+  for each player. The other cycle is of length 4; player 2 drinks
+  on every step (as we know), and player 1 every 4\th\ step, yielding
+  15 more drinks for a total of 45.
+% [60,45] wasting 2: 12 game(s) like 
+% [(start F, U=>D*@1 D=>F*@1 F=>U*@1),
+%  (start F, U=>F*@0 D=>D@0 F=>U*@0)]
+}
+\label{fig:dual45}
+\end{figure}
+
+\section{Three-player \kn Power-Hours}
+
+With 3.7 trillion possible configurations, enumeration is not
+feasible. But as we observed before, many of these combinations are
+illegal (they result in a player recieving two cups), and many are
+isomorphic to one another. By being clever about how we explore the
+configurations, testing ``all'' the three-player configurations
+becomes feasible.
+
+Here is a one-player BMML configuration that illustrates a particular
+kind of redundancy:
+\vspace{1em}
+\begin{tabular}{cc}
+start \emptycup & \emptycup \drink \emptycup 
+ \quad \fullcup \drink \overcup
+ \quad \overcup \nodrink \fullcup
+\end{tabular}
+
+The cup starts empty, and at each step the player fills it and drinks
+(traditional Power-Hour). The player also has rules for the case that
+she observes a full or overturned cup. {\em It does not matter what
+  these are} because they can never be used. This example is trivial,
+but there are many ways that the execution of a configuration can be
+indifferent to some of its content. Another is a two player
+configuration like
+\vspace{1em}
+\begin{tabular}{rcccc}
+player 1 & start \emptycup & \emptycup \drink \overcup \at 1
+ & \fullcup \drink \overcup \at 2
+ & \overcup \nodrink \fullcup \at 1 \\
+player 2 & start \emptycup & \emptycup \nodrink \emptycup \at 1
+ & \fullcup \drink \overcup \at 1
+ & \overcup \drink \fullcup \at 2 \\
+\end{tabular}
+where the $\at n$ notation means to pass the cup in that state to
+player $n$. In this case, the first thing the players do is to
+pass both of their cups to player 1, which is illegal and ends the
+game. Again, none of the other rules are ever used.
+
+In order to explore what is possible in three-player games, we exploit
+this redundancy with a technique like abstract
+interpretation~\cite{abstract}. The start state is always used, so we
+begin by enumerating all assignments of start states to players. There
+are only $4^p$. Every other rule starts out undetermined, maybe written
+like this:
+
+\begin{tabular}{cc}
+start \emptycup & \emptycup \qdrink \any
+ \quad \fullcup \qdrink \any
+ \quad \overcup \qdrink \any
+\end{tabular}
+
+Now we execute programs as before, and hope that we never encounter a
+situation where we depend on a rule. If we finish without ever using
+one of the \any\ rules, we evaluated a potentially large group of
+configurations all at once. During the execution of a configuration,
+if we need to use a rule that is currently marked \any, we explore all
+of the possibilities. This is accomplished by a loop that looks like
+the following (in Pseudo SML):
+
+\begin{code}
+val queue = (* all abstract configurations *)
+val results = (* map from (k_1, k_2) 
+                 to example *)
+\codeskip
+fun loop nil = (* done *)
+  | loop (h :: t) =
+      let
+        val res = evaluate h
+      in
+        insert (results, res);
+        loop t
+      end handle Expand l => loop (l @ t)
+\codeskip
+fun evaluate config =
+   (* ... *) 
+   case rulefor cup of 
+      QuestionMark =>
+         raise Expand expandedconfigs
+    | (* ... *)
+\codeskip
+val () = loop queue
+\end{code}
+
+The key trick here for keeping the code under control is to
+iteratively evaluate the configurations as usual, but if we find a
+\any, then we abort the current simulation with an exception that
+carries along the set of configurations that expand the current one in
+just that position. This wastes some work (and we often need to
+restart multiple times per abstract configuration), but not much: If a
+rule is used at all, it is usually used in one of the first few
+rounds.
+
+With this technique, we can simulate all possible two-player games
+with just 15,744,259 game-minutes simulated (with naive enumeration it
+would be 1.2 billion) in less than 2 seconds on a crappy old computer.
+
+% Total steps 15744259. Took 1982 ms
+% Total steps 15744259. Took 2080 ms
+
+% #100000 queue 31 min 16 6059166 states/sec 100000 config/sec
+% #200000 queue 20 min 16 8762059 states/sec 200000 config/sec
+% #300000 queue 35 min 16 12125162 states/sec 300000 config/sec
+% #400000 queue 25 min 16 7613270 states/sec 200000 config/sec
+% Total steps 15744259. Took 2101 ms
+
+% There are 16 games before splitting.
+% #100000 queue 31 min 16 6059166 states/sec 100000 config/sec
+% #200000 queue 20 min 16 8762059 states/sec 200000 config/sec
+% #300000 queue 35 min 16 6062581 states/sec 150000 config/sec
+% #400000 queue 25 min 16 7613270 states/sec 200000 config/sec
+% Wrote 454 possibilities to possible-2.txt
+% Total steps 15744259. Took 2162 ms
+
+% Wrote 454 possibilities to possible-2.txt
+% Total steps 895743. Took 292 ms
+
+% Wrote 2787 possibilities to possible-5cup-60min-2player.txt
+% Total steps 6890640495. Took 8161324 ms
+
+% on metroid
+% Wrote 37560 possibilities to possible-3.txt
+% Total steps 1393736865. Took 1,545,217 ms
 
@@
 
+\section{Generalized BMML}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.90 \linewidth]{soloncups.pdf}
+\end{center}\vspace{-0.1in}
+\caption{Possible outcomes for BM$n$L with a single player. With no
+  cup states, it is only possible to drink nothing. By convention, the
+  0\th\ cup is the special ``filled'' state, so it is only possible to
+  drink on every turn (60) or never (0). As we add more cup states,
+  the number of achievable states strictly increases; with 8 cup states
+  we can drink any amount. 7 and 53 drinks are the most difficult to
+  achieve, and can only be done with 8 cup states.
+}
+\label{fig:soloncups}
+\end{figure}
+
+\section{Known bounds}
+
+Define BMnL, where $n$ is the number of non-filled states. BMML=BM2L.
+Find $n$ such that a solo or dual power hour achieves any $k$.
+
+\section{Conjectures} \label{sec:conjectures}
+
+With two friends, you can drink any amount
+
+Teetotaller: When $\langle k_1, \ldots, k_i, \ldots, k_n \rangle$ is possible,
+so is $\langle k_1, \ldots, 0, \ldots, k_n \rangle$ (under what conditions)?
+
 \section{Some stuffff}

### 2013-12-27 trunk/powerhour/paper/paper.tex
Remove code duplication in powerhour.sml, oops. Implement another performance improvement that slows it down
Cube rendering, looks terrible
New sampling-based enumeration

@@
 
+% #100000 queue 31 min 16 6059166 states/sec 100000 config/sec
+% #200000 queue 20 min 16 8762059 states/sec 200000 config/sec
+% #300000 queue 35 min 16 12125162 states/sec 300000 config/sec
+% #400000 queue 25 min 16 7613270 states/sec 200000 config/sec
+% Total steps 15744259. Took 2101 ms
+
+% There are 16 games before splitting.
+% #100000 queue 31 min 16 6059166 states/sec 100000 config/sec
+% #200000 queue 20 min 16 8762059 states/sec 200000 config/sec
+% #300000 queue 35 min 16 6062581 states/sec 150000 config/sec
+% #400000 queue 25 min 16 7613270 states/sec 200000 config/sec
+% Wrote 454 possibilities to possible-2.txt
+% Total steps 15744259. Took 2162 ms
+
+% Wrote 454 possibilities to possible-2.txt
+% Total steps 895743. Took 292 ms
+
+% Wrote 2787 possibilities to possible-5cup-60min-2player.txt
+% Total steps 6890640495. Took 8161324 ms
+
+% on metroid
+% Wrote 37560 possibilities to possible-3.txt
+% Total steps 1393736865. Took 1,545,217 ms
+
 \begin{figure}
@@
 
+\section{Generalized BMML}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.90 \linewidth]{soloncups.pdf}
+\end{center}\vspace{-0.1in}
+\caption{Possible outcomes for BM$n$L with a single player. With no
+  cup states, it is only possible to drink nothing. By convention, the
+  0\th\ cup is the special ``filled'' state, so it is only possible to
+  drink on every turn (60) or never (0). As we add more cup states,
+  the number of achievable states strictly increases; with 8 cup states
+  we can drink any amount. 7 and 53 drinks are the most difficult to
+  achieve, and can only be done with 8 cup states.
+}
+\label{fig:soloncups}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.90 \linewidth]{composite2playermock.png}
+\end{center}\vspace{-0.1in}
+\caption{What's achievable for the first two players in a three-player
+  generalized BM$n$L game, with $n$ ranging from 1 cup state (darkest)
+  to 5 (lightest). TODO WRITEME
+% 6 cupstates ran overnight and got here:
+% #26186200000 queue 109 min 49 466729 states/sec 796272 config/sec
+}
+\label{fig:composite2player}
+\end{figure}
+
 \section{Known bounds}
@@
 Define BMnL, where $n$ is the number of non-filled states. BMML=BM2L.
-Find $n$ such that a solo or dual power hour achieves any $k$.
+(XXX actually I think it's the total number of states including filled
+everywhere but that sentence, right?)  Find $n$ such that a solo or
+dual power hour achieves any $k$.
+
+\begin{enumerate}
+  \item With one player, we have exact bounds on what is possible in
+    the generalized case. With 8 cup states, a single player can
+    drink 0--60 shots. Since each player can just play independently,
+    this result extends to any number of players in BM8L.
+  \item With two players, we know that BM5L (XXX check) does not allow
+    all combinations of $\langle k_1, k_2 \rangle$ shots, though BM5L
+    can achieve any $\langle k_1, k_2 \rangle$ where $\k_2 \in \{ 0, 1
+    \}$. No other $k_2$ can be used universally, though of course many
+    other combinations are possible
+    (Figure~\ref{fig:compositeplayer2mock}).
+  \item Open: Can BM6L achieve all $\langle k_1, k_2 \rangle$?
+  \item Open: Can BM7L achieve all $\langle k_1, k_2 \rangle$? Seems
+    likely, given how close BM7L is for a single player.
+  \item With three players one player can drink any number of shots in
+    BMML with two other players helping.
+\end{enumerate}
 

### 2013-12-28 powerhour/paper/paper.tex
Add composite placeholder
Made thing that can plot sampled grids. Made powerhour properly honor symmetry when filling in the grid, at least

@@
 
+% XXX TODO: Display in 3D cube.
+It is also feasible overnight to enumerate all three-player games. The
+results are three-dimensional, of course, but we can display the
+outcomes for two of the players in the familiar presentation
+(Figure~\ref{cig:3and2}). Note that $\langle k_1, k_2, k_3 \rangle$ is
+achievable for any $k_1$, $k_2 \in \{ 0, 1, 2 \}$, and some unknown
+$k_3$ (projected out of this display). This is a significant
+improvement over what was achievable in BMML with two players. It
+suggests that with enough friends willing to follow a program, some
+set of people are likely to be able to achieve any amount of drinks
+between them (if they have the ability to construct the right rule
+set!); see Section~\ref{sec:conjectures}.
+
+The sheer number of configurations for 4 or more players makes these
+exact enumeration techniques infeasible. However, we have other
+avenues for generalization (and exploration), which are explored in
+the next chapter.
+
 % Total steps 15744259. Took 1982 ms
@@
   outcomes for 2-player power hours (red). Mainly included because
-  it looks pretty sweet.
-}
+  it looks pretty sweet. The outcomes that are possible with three
+  players are a superset of those with two, which is intuitive: We
+  can add a third player to any game who just does nothing.}
 \label{fig:3and2}
@@
   (TODO: what are these games? what did the other player drink?)
+  (TODO: Incorporate this into the text, perhaps after discussing
+  the linear combination trick.)
 }
@@
 \section{Generalized BMML}
+
+Like any drinking game, there are several arbitrary things about BMML.
 
@@
 
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.90 \linewidth]{composite2playermock.png}
+\end{center}\vspace{-0.1in}
+\caption{What's achievable for the first two players in a three-player
+  generalized BM$n$L game, with $n$ ranging from 1 cup state (darkest)
+  to 5 (lightest). TODO WRITEME
+% 6 cupstates ran overnight and got here:
+% #26186200000 queue 109 min 49 466729 states/sec 796272 config/sec
+}
+\label{fig:composite2player}
+\end{figure}
+
+\section{Samping games}
+
 \section{Known bounds}
@@
 Define BMnL, where $n$ is the number of non-filled states. BMML=BM2L.
-Find $n$ such that a solo or dual power hour achieves any $k$.
+(XXX actually I think it's the total number of states including filled
+everywhere but that sentence, right?)  Find $n$ such that a solo or
+dual power hour achieves any $k$.
+
+\begin{enumerate}
+  \item With one player, we have exact bounds on what is possible in
+    the generalized case. With 8 cup states, a single player can
+    drink 0--60 shots. Since each player can just play independently,
+    this result extends to any number of players in BM8L.
+  \item With two players, we know that BM5L (XXX check) does not allow
+    all combinations of $\langle k_1, k_2 \rangle$ shots, though BM5L
+    can achieve any $\langle k_1, k_2 \rangle$ where $\k_2 \in \{ 0, 1
+    \}$. No other $k_2$ can be used universally, though of course many
+    other combinations are possible
+    (Figure~\ref{fig:compositeplayer2mock}).
+  \item Open: Can BM6L achieve all $\langle k_1, k_2 \rangle$? Seems
+    unlikely, based on how many samples we need to get close to 100\%
+    for BM7L.
+  \item Open: Can BM7L achieve all $\langle k_1, k_2 \rangle$? Seems
+    likely, given how close BM7L is for a single player. We know it
+    can get 96.67\%, at least!
+  \item With three players one player can drink any number of shots in
+    BMML with two other players helping.
+\end{enumerate}
 

### 2013-12-28 powerhour/paper/paper.tex
Sync metroid. 1D grid rendering (old), some work on paper and sampler (new)

@@
 if we need to use a rule that is currently marked \any, we explore all
-of the possibilities. This is accomplished by a loop that looks like
-the following (in Pseudo SML):
+of the possibilities for that spot. This is accomplished by a loop
+that looks like the following (in Pseudo SML):
 
@@
 outcomes for two of the players in the familiar presentation
-(Figure~\ref{cig:3and2}). Note that $\langle k_1, k_2, k_3 \rangle$ is
-achievable for any $k_1$, $k_2 \in \{ 0, 1, 2 \}$, and some unknown
-$k_3$ (projected out of this display). This is a significant
-improvement over what was achievable in BMML with two players. It
-suggests that with enough friends willing to follow a program, some
-set of people are likely to be able to achieve any amount of drinks
-between them (if they have the ability to construct the right rule
-set!); see Section~\ref{sec:conjectures}.
+(Figure~\ref{fig:3and2}). Note that $\langle k_1, k_2, k_3 \rangle$ is
+achievable for any $k_1 \in \{ 0 .. 60 \}$, $k_2 \in \{ 0, 1, 2 \}$,
+and some unknown $k_3$ (projected out of this display). This is a
+significant improvement over what was achievable in BMML with two
+players. It suggests that with enough friends willing to follow a
+program, some set of people are likely to be able to achieve any
+amount of drinks between them (if they have the ability to construct
+the right rule set!); see Section~\ref{sec:conjectures}.
 
@@
 exact enumeration techniques infeasible. However, we have other
-avenues for generalization (and exploration), which are explored in
-the next chapter.
+avenues for generalization (and exploration), which are investigated
+in the next chapter.
 
@@
 \begin{center}
-\includegraphics[width=0.90 \linewidth]{soloncups.pdf}
+\includegraphics[width=\linewidth]{soloncups.pdf}
 \end{center}\vspace{-0.1in}
@@
     all combinations of $\langle k_1, k_2 \rangle$ shots, though BM5L
-    can achieve any $\langle k_1, k_2 \rangle$ where $\k_2 \in \{ 0, 1
+    can achieve any $\langle k_1, k_2 \rangle$ where $k_2 \in \{ 0, 1
     \}$. No other $k_2$ can be used universally, though of course many
     other combinations are possible
-    (Figure~\ref{fig:compositeplayer2mock}).
+    (Figure~\ref{fig:composite2player}).
   \item Open: Can BM6L achieve all $\langle k_1, k_2 \rangle$? Seems
@@
   \item Open: Can BM7L achieve all $\langle k_1, k_2 \rangle$? Seems
-    likely, given how close BM7L is for a single player. We know it
-    can get 96.67\%, at least!
+    possible, given how close BM7L is for a single player. We know it
+    can get 99.79\%, at least!
   \item With three players one player can drink any number of shots in

### 2013-12-28 powerhour/paper/paper.tex
Writing paper

@@
 \newcommand\comment[1]{}
-\newcommand\sfrac[2]{{}\,^{#1}\!/{}\!_{#2}}
+\newcommand\sfrac[2]{\!{}\,^{#1}\!/{}\!_{#2}}
 
@@
 \newcommand\kn{\ensuremath{\sfrac{k}{n}\,}}
+\newcommand\bmsl{BM$s$L}
 
@@
 \newcommand\overcup{\ensuremath{\cap}}
+\newcommand\scup[1]{\ensuremath{\cup\!\!\!{}^{{}_{#1}}}}
 
@@
 (Figure~\ref{fig:3and2}). Note that $\langle k_1, k_2, k_3 \rangle$ is
-achievable for any $k_1 \in \{ 0 .. 60 \}$, $k_2 \in \{ 0, 1, 2 \}$,
+achievable for any $k_1 \in \{ 0 \ldots 60 \}$, $k_2 \in \{ 0, 1, 2 \}$,
 and some unknown $k_3$ (projected out of this display). This is a
@@
 
+\section{Generalized BMML}
+
+Like any drinking game, there are several arbitrary things about BMML.
+While we will not tamper with its essence (for example, allowing beer
+to be spilled from a cup without drinking it), there are some other
+variables to adjust. The most naturally flexible is the number of cup
+states. We will always have \fullcup, a cup with beer in it. In BMML
+we also have \emptycup\ and \overcup. But why not $\supset$ (cup
+turned on its side, facing west) or $\stackrel{\wedge}{\cup}$ (an
+upright cup with a cocktail umbrella on it)?
+
+We define \bmsl, where $s$ is the number of distinct cup states. By
+convention, the 0\th\ cup state will be the filled cup \fullcup\ since
+it has special rules. The remainder will be \scup{i} for $i \in \{ 1,
+\ldots, s - 1 \}$. BMML is BM3L where we've just renamed \emptycup\ to
+\scup{1}, and \overcup\ to \scup{2}.
+
+Clearly, more cups give us more expressive power, and should allow us
+to reach more outcomes. To illustrate, recall the construction of
+$k=31$ in the solo BM3L game (Figure~\ref{fig:solo31}). It has a
+length-2 cycle alternating between two cup states, where the player
+drinks on every other turn. The third state is just used once as a
+drinking lead-in to make the total 31 rather than 30. With an
+additional cup state, we can straightforwardly transform this into a
+game with an outcome of $k=32$ by extending the prelude with another
+state where the player drinks. Of course, the expressive power is not
+just limited to such extensions; we now can create cycles of new
+lengths, admit the possibility of more disconnected cycles, and so on.
+
+It is easy to enumerate the 1-player \bmsl\ games. These appear in
+Figure~\ref{fig:soloncups}. The interesting range for $s$ is $\{ 1
+\ldots 8 \}$.\!\footnote{It's not clear that BM0L should be considered
+  legal as the rules speak of a 0\th\ cup, but it is degenerate
+  anyway.} At 8 cup states, the player can achieve any amount in a
+solo game. Importantly, this extends to any number of players in BM8L,
+because the players can pass to themselves and not even interact.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=\linewidth]{soloncups.pdf}
+\end{center}\vspace{-0.1in}
+\caption{Possible outcomes for \bmsl\ with a single player. The
+  vertical axis shows an increasing number of cup states, and the
+  horizontal axis shows the achievable values of $k$ drinks. With no
+  cup states, it is only possible to drink nothing. By convention, the
+  0\th\ cup is the special ``filled'' state, so it is only possible to
+  drink on every turn (60) or never (0). As we add more cup states,
+  the number of achievable states strictly increases; with 8 cup
+  states we can drink any amount. 7 and 53 drinks are the most
+  elusive, and can only be done with 8 cup states.}
+\label{fig:soloncups}
+\end{figure}
+
+The two-player case is much more interesting. We've already enumerated
+all the possible outcomes for BM3L
+(Figure~\ref{fig:powerhour2player}). It is computationally tractable
+to enumerate them for $s < 6$. The set of achievable outcomes in
+BM$s$L is always contained within BM$(s+1)$L, because we can embed a
+game from the former into the latter by just never producing the
+additional cup state and having any arbitrary rule for it. Therefore,
+we show these results in a composite grid where more and more states
+are reachable as we increase $s$ (Figure~\ref{fig:composite2player}).
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.90 \linewidth]{composite2playermock.png}
+\end{center}\vspace{-0.1in}
+\caption{What's achievable for two players in a generalized \bmsl\ game,
+  with $s$ ranging from 1 cup state (darkest) to 5 (lightest).
+% 6 cupstates ran overnight and got here:
+% #26186200000 queue 109 min 49 466729 states/sec 796272 config/sec
+}
+\label{fig:composite2player}
+\end{figure}
+
+In order to efficiently enumerate BM5L, I improved the algorithm
+again. Observe that the ``drink'' action associated with a rule
+usually does not affect anything but the final outcome. The only
+exception is that in the rule for \fullcup, the player must drink if
+passing in a state other than \fullcup. Putting this aside for a
+moment, note that we can just count the number of times each rule was
+executed for each player, producing an $s$-dimensional vector $[ d_1,
+  \ldots d_s ]$ for each player. That player is able to achieve many
+different drink totals, specifically, $d_1 \times r_1 + \ldots + d_s
+\times r_s$ where $r_i$ is 1 if the player should drink on that rule
+and 0 if not. Simulating a game this way is even more like abstract
+interpretation (we leave a concrete value free and compute a formula
+rather than an integer), and allows us to evaluate many concrete games
+at once. At the end, we simply plug in every legal value for $r_i$ for
+each player and insert those games into the database. This last step
+is where we must tend to the exception around \fullcup. We may not set
+$r_0$ to 0 if the player ever passes \fullcup\ in a non-full state. A
+very close approximation would be to insist that $r_0 = 1$ if the rule
+in the \fullcup\ position does not output as \fullcup, but this is
+inexact, as that rule may never be executed\footnote{We may be able to
+  argue that in that case, there always exists another game that does
+  not violate this condition. But I think it is simpler to just
+  implement the rules.} Instead, during simulation we keep track of
+whether each player ever actually passed a non-full cup from the
+\fullcup\ state. If so, then we force $r_0 = 1$, which attends to this
+special case.
+
+Although this makes earlier enumerations extremely fast and BM5L
+quite quick, 2-player BM6L ran 26 billion concrete states overnight
+and made only modest progress. In the absence of fancier techniques
+for reducing the state space, we must resort to different, inexact
+approaches.
+
+\section{Samping games}
+
+\section{The fractal geometry of \kn\ Power-Hours}
+
+Note that all of the two-dimensional figures resemble one another even
+though they are fundamentally different (adding players, adding
+states, adding random trials). This suggests that the combinatorial
+problem (``what outcomes are reachable from finite state machines that
+look kind of like this?'') has some geometric structure.
+
 \begin{figure}
@@
 
-\section{Generalized BMML}
-
-Like any drinking game, there are several arbitrary things about BMML.
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=\linewidth]{soloncups.pdf}
-\end{center}\vspace{-0.1in}
-\caption{Possible outcomes for BM$n$L with a single player. With no
-  cup states, it is only possible to drink nothing. By convention, the
-  0\th\ cup is the special ``filled'' state, so it is only possible to
-  drink on every turn (60) or never (0). As we add more cup states,
-  the number of achievable states strictly increases; with 8 cup states
-  we can drink any amount. 7 and 53 drinks are the most difficult to
-  achieve, and can only be done with 8 cup states.
-}
-\label{fig:soloncups}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=0.90 \linewidth]{composite2playermock.png}
-\end{center}\vspace{-0.1in}
-\caption{What's achievable for the first two players in a three-player
-  generalized BM$n$L game, with $n$ ranging from 1 cup state (darkest)
-  to 5 (lightest). TODO WRITEME
-% 6 cupstates ran overnight and got here:
-% #26186200000 queue 109 min 49 466729 states/sec 796272 config/sec
-}
-\label{fig:composite2player}
-\end{figure}
-
-\section{Samping games}
-
 \section{Known bounds}
 
-Define BMnL, where $n$ is the number of non-filled states. BMML=BM2L.
-(XXX actually I think it's the total number of states including filled
-everywhere but that sentence, right?)  Find $n$ such that a solo or
-dual power hour achieves any $k$.
+Define \bmsl, where $s$ is the number of cup states. BMML=BM3L. Find
+$s$ such that a solo or dual power hour achieves any $k$.
 
@@
   \item Open: Can BM6L achieve all $\langle k_1, k_2 \rangle$? Seems
-    unlikely, based on how many samples we need to get close to 100\%
-    for BM7L.
+    unlikely, given that random exploration plateaus at about 95.54\%.
   \item Open: Can BM7L achieve all $\langle k_1, k_2 \rangle$? Seems
     possible, given how close BM7L is for a single player. We know it
-    can get 99.79\%, at least!
+    can get 99.89\%, at least!
   \item With three players one player can drink any number of shots in
@@
 
-Teetotaller: When $\langle k_1, \ldots, k_i, \ldots, k_n \rangle$ is possible,
-so is $\langle k_1, \ldots, 0, \ldots, k_n \rangle$ (under what conditions)?
+Teetotaller: When $\langle k_1, \ldots, k_i, \ldots, k_p \rangle$ is possible,
+so is $\langle k_1, \ldots, 0, \ldots, k_p \rangle$ (under what conditions)?
 

### 2013-12-28 powerhour/paper/paper.tex
Getting close on paper

@@
 
+\usepackage{url}
 \usepackage{code}
-\usepackage{cite}
+% \usepackage{cite}
 % \usepackage{eclbkbox}
@@
 
+\usepackage{natbib}
 % \usepackage{inlinebib}
@@
 \fullcup\ state. If so, then we force $r_0 = 1$, which attends to this
-special case.
+special case. \label{sec:linear}
 
@@
 
-\section{Samping games}
+\section{Sampling games}
+
+To establish a result like ``BM5L cannot achieve $\langle 47, 27
+\rangle$'' we really need to enumerate all the BM5L games. (Or make
+some ad hoc proof of the fact, which seems quite difficult.) However,
+to prove an existence result like ``BM7L can achieve $\langle 33, 49
+\rangle$'' we only need to have a single example configuration that
+produces that result. Therefore, we may be able to improve our bounds
+on what is possible (or generate conjectures) by sampling random
+configurations.
+
+Sampling is actually much easier than enumeration. There is no need to
+leave rules abstract. It is also easy to stop and restart because
+there is no state other than the matrix of what we've found. I use the
+SML {\tt textformat} library~\cite{textformat} to serialize and
+deserialize the matrix (which then makes it easy to generate these
+graphics in a separate program). There are a handful of interesting
+aspects:
+
+\paragraph{Generating a random configuration.} To generate a random
+game, we can just fill in all of the slots (destination and cup state
+for each rule, starting cup state) uniformly at random. Many of these
+will be illegal, but they fail very quickly at runtime; a lazy and
+pragmatic way to ``filter'' to legal game. It is not simply a matter
+of generating all the permutations on $p \times s$ nodes, by the way.
+Multiple cups can pass through the same player on cycles of different
+periods, as long as they do not collide within the 60 steps, and
+acyclic preludes (Figure~\ref{fig:solo31}) are important and useful.
+For a uniformly random 2-player BM7L configuration, 29.23\% (measured
+empirically) are legal. However, we will see later that we do not want
+to spend so much time exploring configurations where one or both
+players start without a cup; these are very limited. Therefore, the
+configuration generator is biased towards producing a cup in the
+starting states most of the time.
+
+\paragraph{Symmetry.} We can get more bang for the buck by considering
+some obvious symmetries. When a simulation finishes and we have an
+outcome $\langle k_1, \ldots, k_p \rangle$, it is clear that any
+permutation of $k_1 \ldots k_p$ is also achievable. We insert every
+permutation of the drink counts into the database, along with the
+permuted example configuration. Better still would be to only store
+the outcomes in some normalized form (e.g. require that $k_1 \leq
+\ldots \leq k_p$).
+
+We already have exact results for two players in BM5L, so the next
+uncharted territory is BM6L. The result after apparent convergence
+appears in Figure~\ref{fig:samplebm6l}. The sampling procedure runs
+for many hours before plateauing overnight with 95.54\% of the matrix
+filled. This suggests that BM6L is not universal for two players, or
+else the configurations for the missing cells are extremely rare.
+% XXX never give up! It just jumped to 95.65.
+
+This approach scales much better than enumeration and is efficient for
+all sorts of generalizations (it works best when the dimensionality is
+low---i.e., two players---and the expressiveness is high---i.e., many
+cup states). Since we already know BM8L is universal, the remaining
+open problem is BM7L, whose results are in
+Figure~\ref{fig:samplebm7l}. This is tantalizingly close to a complete
+matrix, with only 4 cells missing! It may be possible to solve these
+two cases manually, though a proof of non-existence would be
+difficult.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.90 \linewidth]{sample60min2players6states.png}
+\end{center}\vspace{-0.1in}
+\caption{ 25.6 billion samples of legal two-player BM6L
+  configurations. Darker cells represent outcomes that occur more
+  often; cells that are pure white never occurred and are likely to be
+  unattainable. Note that the intensity represents the rank of
+  occurrence, not the magnitude; in actuality, outcomes like $\langle
+  0, 0 \rangle$ occur {\em much} more often than others.
+  %
+  95.54\% of the cells are filled.
+  % XXX new version with more filled...!
+}
+\label{fig:samplebm6l}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.90 \linewidth]{sample60min2players7states.png}
+\end{center}\vspace{-0.1in}
+\caption{ 22.1 billion samples of legal two-player BM7L
+  configurations. Fewer configurations were sampled than in
+  Figure~\ref{fig:samplebm6l} because they take somewhat longer than
+  6-state games to simulate, and a smaller proportion of random games
+  are legal.
+  %
+  All but four of the $(60 + 1)^2 = 3721$ cells are attainable: 
+  Permutations of $\langle 11, 53 \rangle$ and $\langle 49, 53 \rangle$
+  remain elusive. Note that 53 drinks was also unattainable in a solo
+  BM7L power-hour; this may in some sense be the ``hardest'' number of
+  shots to drink in \bmsl.
+}
+\label{fig:samplebm7l}
+\end{figure}
 
@@
 
+Some of the patterns are easy to explain. The top-left half of the
+matrix is more populous than the bottom right, for example. This is
+because we can bound the total number of drinks by $60 / c$, where $c$
+is the number of cups active in the game (same as the number of cups
+in the starting states). The top-left half is the region where this
+sum is less than or equal to 30; in two player games, both players
+must start with a cup in order to get an outcome in the bottom-right
+half. We also see distinct clumps around 0, 15, 30, 45, 60; divisors
+of 60. This is intuitive because the expressive power of BMML comes
+from the ability to form cycles of cup states and drink on some
+fraction of them. Clumps are formed around these values because of the
+possibility of preludes leading into the cycles
+(Figure~\ref{fig:solo31}) that either drink (adding to the total) or
+don't (subtracting from it). Minor clumps form as echoes between the
+major ones, because a player may participate in two cycles of
+different length (Figure~\ref{fig:dual45}.
+
+Of course, discretization effects compound and so the exact values of
+cells are not neatly predictable. Moreover, clumps interfere by
+overlapping; there are many different strategies for achieving
+$\langle 33, 33 \rangle$. One way to make the basic structure more
+visible is to extend the number of minutes that the game is played
+for. Figure~\ref{fig:powerday3} shows the utterly unhealthy
+three-player Power Day (BM3L). In it, the clumps become tiny dots,
+but some relationship among them along lines is clear. Interior points
+can probably be found as linear combinations of two of these lines;
+we exploit that exact structure in Section~\ref{sec:linear}, in fact.
+
 \begin{figure}
@@
   power days. These are the same games as the 3-player power hours,
-  but at this scale makes it clear the groupings and their sparsity
-  in the limit. Lines plotted from (0,0) to (60,*) and (*,60) show
-  significant structure, but don't explain some of the interior points.
-  (TODO: what are these games? what did the other player drink?)
-  (TODO: Incorporate this into the text, perhaps after discussing
-  the linear combination trick.)
+  but at this scale makes it clear the groupings and their sparsity in
+  the limit. Lines plotted from $\langle 0,0 \rangle$ to $langle 60,
+  k_2 \rangle$ and $\langle k_1, 60 \rangle$ show significant
+  structure, but don't explain some of the interior points. These are
+  probably games where a player participates in two cycles of
+  different periods.
 }

### 2013-12-29 trunk/powerhour/paper/paper.tex
Commit before overwrite

@@
 
+\usepackage{url}
 \usepackage{code}
-\usepackage{cite}
-% \usepackage{eclbkbox}
+% \usepackage{cite}
 \usepackage{amsmath}
 \usepackage{amssymb}
-% \usepackage{code}
-% \usepackage{amscd}
-% \usepackage{xy}
 \usepackage{graphicx}
-% \usepackage{fancyhdr}
-% \usepackage{color}
-% \usepackage[dark,all,bottom,landscape,timestamp]{draftcopy}
-% \usepackage{everypage}
 
@@
 
-% \usepackage{inlinebib}
+\usepackage{natbib}
 
 \newcommand\comment[1]{}
-\newcommand\sfrac[2]{{}\,^{#1}\!/{}\!_{#2}}
+\newcommand\sfrac[2]{\!{}\,^{#1}\!/{}\!_{#2}}
 
@@
 \begin{abstract}
-Something about the k-n hours
-
+We correct for inebriated missteps, using computational methods to
+establish new bounds in generalized \kn Power-Hour theory.
 \end{abstract}
@@
 perspective, clarifies some of the original results, and presents
-several new ones and a few conjectures.
+several new ones and a few conjectures. It is based on several pieces
+of software, whose source is available online.\!\footnote{ \url{http://sourceforge.net/p/tom7misc/svn/HEAD/tree/trunk/powerhour/} }
 
@@
 non-steady-states like this (Figure~\ref{fig:solo31}) is how we
-achieve $k$ that does not divide $n$.
+achieve $k$ that does not share many factors with $n$.
 
@@
 results were generated by a computer program, which is probably
-necessary for $p>1$. The next section describes how we explore this
-space computationally. XXX no it doesn't
+necessary for $p>1$. In the remainder of the paper, I'll describe
+several different approaches for exploring this space, and
+generalizations of it, computationally.
 
@@
   that any cup state can be used in that position. Note that 29 and 58
-  require wasting a shot of beer; all the others but 31 permit a
-  variant where a shot is wasted as well.}
+  require wasting a shot of beer (the game ends with the shotglass
+  full); all the others but 31 permit a variant where a shot is wasted
+  as well. We do not concern ourselves much in this report with these
+  leftover shots.}
 \end{figure}
@@
 \begin{itemize}
-\item $s = 4$, the number of starting states (\fullcup, \emptycup,
+\item $t = 4$, the number of starting states (\fullcup, \emptycup,
   \overcup, \nocup)
-\item $a = 2 * p * 3$, the number of actions given a cup. The player
-  can drink or not drink, pass to any player, and in 3 configurations
-  (\fullcup, \emptycup, \overcup)
+\item $a = 2 \times p \times 3$, the number of actions given a cup.
+  The player can drink or not drink, pass to any player, and in 3
+  configurations (\fullcup, \emptycup, \overcup)
 \end{itemize}
-Then the number of configurations is bounded by $(s * a^3)^p$. For
+Then the number of configurations is bounded by $(t \times a^3)^p$. For
 $p=1$ this was just 864. For $p=2$ it is 47,775,744; for $p=3$ it's
@@
 $$
-  s * a^2 * a_{\textrm{filled}}
+  t \times a^2 \times a_{\textrm{filled}}
 $$
 where
-$a_{\textrm{filled}} = (p * 3) + p$ (the actions that can be taken on
+$a_{\textrm{filled}} = (p \times 3) + p$ (the actions that can be taken on
 \fullcup, where if the player does not drink, then he must pass the
@@
 example, if two players pass to the same player no matter their input
-state, then their cups always collide. There are also many games
-that are isomorphic. For one thing, \emptycup\ and \overcup\ are not
-distinguished in the rules at all, so any two configurations where
-these are simply swapped has the exact same outcome. Likewise for
-permuting the players.
+state, and every player starts with a cup, then their cups always
+collide. There are also many games that are isomorphic. For one thing,
+\emptycup\ and \overcup\ are not distinguished in the rules at all, so
+any two configurations where these are simply swapped has the exact
+same outcome. Likewise for permuting the players.
 
@@
 
+% XXX TODO: Display in 3D cube.
+It is also feasible overnight to enumerate all three-player games. The
+results are three-dimensional, of course, but we can display the
+outcomes for two of the players in the familiar presentation
+(Figure~\ref{fig:3and2}). Note that $\langle k_1, k_2, k_3 \rangle$ is
+achievable for any $k_1 \in \{ 0 \ldots 60 \}$, $k_2 \in \{ 0, 1, 2 \}$,
+and some unknown $k_3$ (projected out of this display). This is a
+significant improvement over what was achievable in BMML with two
+players. It suggests that with enough friends willing to follow a
+program, some set of people are likely to be able to achieve any
+amount of drinks between them (if they have the ability to construct
+the right rule set!); see Section~\ref{sec:conjectures}.
+
+The sheer number of configurations for 4 or more players makes these
+exact enumeration techniques infeasible. However, we have other
+avenues for generalization (and exploration), which are investigated
+in the next chapter.
+
 % Total steps 15744259. Took 1982 ms
@@
   outcomes for 2-player power hours (red). Mainly included because
-  it looks pretty sweet.
-}
+  it looks pretty sweet. The outcomes that are possible with three
+  players are a superset of those with two, which is intuitive: We
+  can add a third player to any game who just does nothing.}
 \label{fig:3and2}
 \end{figure}
+
+\section{Generalized BMML}
+
+Like any drinking game, there are several arbitrary things about BMML.
+While we will not tamper with its essence (for example, allowing beer
+to be spilled from a cup without drinking it), there are some other
+variables to adjust. The most naturally flexible is the number of cup
+states. We will always have \fullcup, a cup with beer in it. In BMML
+we also have \emptycup\ and \overcup. But why not $\supset$ (cup
+turned on its side, facing west) or $\stackrel{\wedge}{\cup}$ (an
+upright cup with a cocktail umbrella on it)?
+
+We define \bmsl, where $s$ is the number of distinct cup states. By
+convention, the 0\th\ cup state will be the filled cup \fullcup\ since
+it has special rules. The remainder will be \scup{i} for $i \in \{ 1,
+\ldots, s - 1 \}$. BMML is BM3L where we've just renamed \emptycup\ to
+\scup{1}, and \overcup\ to \scup{2}.
+
+Clearly, more cups give us more expressive power, and should allow us
+to reach more outcomes. To illustrate, recall the construction of
+$k=31$ in the solo BM3L game (Figure~\ref{fig:solo31}). It has a
+length-2 cycle alternating between two cup states, where the player
+drinks on every other turn. The third state is just used once as a
+drinking lead-in to make the total 31 rather than 30. With an
+additional cup state, we can straightforwardly transform this into a
+game with an outcome of $k=32$ by extending the prelude with another
+state where the player drinks. Of course, the expressive power is not
+just limited to such extensions; we now can create cycles of new
+lengths, admit the possibility of more disconnected cycles, and so on.
+
+It is easy to enumerate the 1-player \bmsl\ games. These appear in
+Figure~\ref{fig:soloncups}. The interesting range for $s$ is $\{ 1
+\ldots 8 \}$.\!\footnote{It's not clear that BM0L should be considered
+  legal as the rules speak of a 0\th\ cup, but it is degenerate
+  anyway.} At 8 cup states, the player can achieve any amount in a
+solo game. Importantly, this extends to any number of players in BM8L,
+because the players can pass to themselves and not even interact.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=\linewidth]{soloncups.pdf}
+\end{center}\vspace{-0.1in}
+\caption{Possible outcomes for \bmsl\ with a single player. The
+  vertical axis shows an increasing number of cup states, and the
+  horizontal axis shows the achievable values of $k$ drinks. With no
+  cup states, it is only possible to drink nothing. By convention, the
+  0\th\ cup is the special ``filled'' state, so it is only possible to
+  drink on every turn (60) or never (0). As we add more cup states,
+  the number of achievable states strictly increases; with 8 cup
+  states we can drink any amount. 7 and 53 drinks are the most
+  elusive, and can only be done with 8 cup states.}
+\label{fig:soloncups}
+\end{figure}
+
+The two-player case is much more interesting. We've already enumerated
+all the possible outcomes for BM3L
+(Figure~\ref{fig:powerhour2player}). It is computationally tractable
+to enumerate them for $s < 6$. The set of achievable outcomes in
+BM$s$L is always contained within BM$(s+1)$L, because we can embed a
+game from the former into the latter by just never producing the
+additional cup state and having any arbitrary rule for it. Therefore,
+we show these results in a composite grid where more and more states
+are reachable as we increase $s$ (Figure~\ref{fig:composite2player}).
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.90 \linewidth]{composite2playermock.png}
+\end{center}\vspace{-0.1in}
+\caption{What's achievable for two players in a generalized \bmsl\ game,
+  with $s$ ranging from 1 cup state (darkest) to 5 (lightest).
+% 6 cupstates ran overnight and got here:
+% #26186200000 queue 109 min 49 466729 states/sec 796272 config/sec
+}
+\label{fig:composite2player}
+\end{figure}
+
+In order to efficiently enumerate BM5L, I improved the algorithm
+again. Observe that the ``drink'' action associated with a rule
+usually does not affect anything but the final outcome. The only
+exception is that in the rule for \fullcup, the player must drink if
+passing in a state other than \fullcup. Putting this aside for a
+moment, note that we can just count the number of times each rule was
+executed for each player, producing an $s$-dimensional vector $[ d_1,
+  \ldots, d_s ]$ for each player. That player is able to achieve many
+different drink totals, specifically, $d_1 \times r_1 + \ldots + d_s
+\times r_s$ where $r_i$ is 1 if the player should drink on that rule
+and 0 if not. Simulating a game this way is even more like abstract
+interpretation (we leave a concrete value free and compute a formula
+rather than an integer), and allows us to evaluate many concrete games
+at once. At the end, we simply plug in every legal value for $r_i$ for
+each player and insert those games into the database. This last step
+is where we must tend to the exception around \fullcup. We may not set
+$r_0$ to 0 if the player ever passes \fullcup\ in a non-full state. A
+very close approximation would be to insist that $r_0 = 1$ if the rule
+in the \fullcup\ position does not output as \fullcup, but this is
+inexact, as that rule may never be executed.\!\footnote{We may be able to
+  argue that in that case, there always exists another game that does
+  not violate this condition. But I think it is simpler to just
+  implement the rules.} Instead, during simulation we keep track of
+whether each player ever actually passed a non-full cup from the
+\fullcup\ state. If so, then we force $r_0 = 1$, which attends to this
+special case. \label{sec:linear}
+
+Although this makes earlier enumerations extremely fast and BM5L
+quite quick, 2-player BM6L ran 26 billion concrete states overnight
+and made only modest progress. In the absence of fancier techniques
+for reducing the state space, we must resort to different, inexact
+approaches.
+
+\section{Sampling games}
+
+To establish a result like ``BM5L cannot achieve $\langle 47, 27
+\rangle$'' we really need to enumerate all the BM5L games. (Or make
+some ad hoc proof of the fact, which seems quite difficult.) However,
+to prove an existence result like ``BM7L can achieve $\langle 33, 49
+\rangle$'' we only need to have a single example configuration that
+produces that result. Therefore, we may be able to improve our bounds
+on what is possible (or generate conjectures) by sampling random
+configurations.
+
+Sampling is actually much easier than enumeration. There is no need to
+leave rules abstract. It is also easy to stop and restart because
+there is no state other than the matrix of what we've found. I use the
+SML {\tt textformat} library~\cite{textformat} to serialize and
+deserialize the matrix (which then makes it easy to generate these
+graphics in a separate program). There are a handful of interesting
+aspects:
+
+\paragraph{Generating a random configuration.} To generate a random
+game, we can just fill in all of the slots (destination and cup state
+for each rule, starting cup state) uniformly at random. Many of these
+will be illegal, but they fail very quickly at runtime; a lazy and
+pragmatic way to ``filter'' to legal game. It is not simply a matter
+of generating all the permutations on $p \times s$ nodes, by the way.
+Multiple cups can pass through the same player on cycles of different
+periods, as long as they do not collide within the 60 steps, and
+acyclic preludes (Figure~\ref{fig:solo31}) are important and useful.
+For a uniformly random 2-player BM7L configuration, 29.23\% (measured
+empirically) are legal. However, we will see later that we do not want
+to spend so much time exploring configurations where one or both
+players start without a cup; these are very limited. Therefore, the
+configuration generator is biased towards producing a cup in the
+starting states most of the time.
+
+\paragraph{Symmetry.} We can get more bang for the buck by considering
+some obvious symmetries. When a simulation finishes and we have an
+outcome $\langle k_1, \ldots, k_p \rangle$, it is clear that any
+permutation of $k_1 \ldots k_p$ is also achievable. We insert every
+permutation of the drink counts into the database, along with the
+permuted example configuration. Better still would be to only store
+the outcomes in some normalized form (e.g. require that $k_1 \leq
+\ldots \leq k_p$).
+
+We already have exact results for two players in BM5L, so the next
+uncharted territory is BM6L. The result after apparent convergence
+appears in Figure~\ref{fig:samplebm6l}. The sampling procedure runs
+for many hours before plateauing overnight with 95.65\% of the matrix
+filled. This suggests that BM6L is not universal for two players, or
+else the configurations for the missing cells are extremely
+rare.\!\footnote{This is definitely a possibility, as new cells were
+still appearing after exploring tens of billions of samples. However,
+the gap here seems quite large.}
+
+This approach scales much better than enumeration and is efficient for
+all sorts of generalizations (it works best when the dimensionality is
+low---i.e., two players---and the expressiveness is high---i.e., many
+cup states). Since we already know BM8L is universal, the remaining
+open problem is BM7L, whose results are in
+Figure~\ref{fig:samplebm7l}. This is tantalizingly close to a complete
+matrix, with only 4 cells missing! It may be possible to solve these
+two cases manually, though a proof of non-existence would be
+difficult.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.90 \linewidth]{sample60min2players6states.png}
+\end{center}\vspace{-0.1in}
+\caption{ 37.1 billion samples of legal two-player BM6L
+  configurations. Darker cells represent outcomes that occur more
+  often; cells that are pure white never occurred and are likely to be
+  unattainable. Note that the intensity represents the rank of
+  occurrence, not the magnitude; in actuality, outcomes like $\langle
+  0, 0 \rangle$ occur {\em much} more often than others.
+  %
+  95.65\% of the cells are filled.
+}
+\label{fig:samplebm6l}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.90 \linewidth]{sample60min2players7states.png}
+\end{center}\vspace{-0.1in}
+\caption{ 22.1 billion samples of legal two-player BM7L
+  configurations. Fewer configurations were sampled than in
+  Figure~\ref{fig:samplebm6l} because they take somewhat longer than
+  6-state games to simulate, and a smaller proportion of random games
+  are legal.
+  %
+  All but four of the $(60 + 1)^2 = 3721$ cells are attainable: 
+  Permutations of $\langle 11, 53 \rangle$ and $\langle 49, 53 \rangle$
+  remain elusive. Note that 53 drinks was also unattainable in a solo
+  BM7L power-hour; this may in some sense be the ``hardest'' number of
+  shots to drink in \bmsl.
+}
+\label{fig:samplebm7l}
+\end{figure}
+
+\section{The fractal geometry of \kn\ Power-Hours}
+
+Note that all of the two-dimensional figures resemble one another even
+though they are fundamentally different (adding players, adding
+states, adding random trials). This suggests that the combinatorial
+problem (``what outcomes are reachable from finite state machines that
+look kind of like this?'') has some geometric structure.
+
+Some of the patterns are easy to explain. The top-left half of the
+matrix is more populous than the bottom right, for example. This is
+because we can bound the total number of drinks by $60 / c$, where $c$
+is the number of cups active in the game (same as the number of cups
+in the starting states). The top-left half is the region where this
+sum is less than or equal to 30; in two player games, both players
+must start with a cup in order to get an outcome in the bottom-right
+half. We also see distinct clumps around 0, 15, 30, 45, 60; these
+correspond to simple fractions (``drink every other time''; ``drink
+three of four times'') of 60. This is intuitive because the expressive
+power of BMML comes from the ability to form cycles of cup states and
+drink on some fraction of them. Clumps are formed around these values
+because of the possibility of preludes leading into the cycles
+(Figure~\ref{fig:solo31}) that either drink (adding to the total) or
+don't (subtracting from it). Minor clumps form as echoes between the
+major ones, because a player may participate in two cycles of
+different length (Figure~\ref{fig:dual45}).
+
+Of course, discretization effects compound and so the exact values of
+cells are not neatly predictable. Moreover, clumps interfere by
+overlapping; there are many different strategies for achieving
+$\langle 33, 33 \rangle$. One way to make the basic structure more
+visible is to extend the number of minutes that the game is played
+for. Figure~\ref{fig:powerday3} shows the utterly unhealthy
+three-player Power Day (BM3L). In it, the clumps become tiny dots,
+but some relationship among them along lines is clear. Interior points
+can probably be found as linear combinations of two of these lines;
+we exploit that exact structure in Section~\ref{sec:linear}, in fact.
 
@@
   power days. These are the same games as the 3-player power hours,
-  but at this scale makes it clear the groupings and their sparsity
-  in the limit. Lines plotted from (0,0) to (60,*) and (*,60) show
-  significant structure, but don't explain some of the interior points.
-  (TODO: what are these games? what did the other player drink?)
+  but at this scale makes it clear the groupings and their sparsity in
+  the limit. Lines plotted from $\langle 0,0 \rangle$ to $\langle 60,
+  k_2 \rangle$ and $\langle k_1, 60 \rangle$ show significant
+  structure, but don't explain some of the interior points. These are
+  probably games where a player participates in two cycles of
+  different periods.
 }
@@
 
-\section{Generalized BMML}
+The less extreme \kn\ Power-Three-Hours appears in
+Figure~\ref{fig:powerthreehours}.
 
@@
 \begin{center}
-\includegraphics[width=0.90 \linewidth]{soloncups.pdf}
+\includegraphics[width=0.90 \linewidth]{sample180min2players7states.png}
 \end{center}\vspace{-0.1in}
-\caption{Possible outcomes for BM$n$L with a single player. With no
-  cup states, it is only possible to drink nothing. By convention, the
-  0\th\ cup is the special ``filled'' state, so it is only possible to
-  drink on every turn (60) or never (0). As we add more cup states,
-  the number of achievable states strictly increases; with 8 cup states
-  we can drink any amount. 7 and 53 drinks are the most difficult to
-  achieve, and can only be done with 8 cup states.
+\caption{6,456,764,116 samples of two-player BM7L Power Three-Hours.
 }
-\label{fig:soloncups}
+\label{fig:powerthreehours}
 \end{figure}
 
-\begin{figure}
-\begin{center}
-\includegraphics[width=0.90 \linewidth]{composite2playermock.png}
-\end{center}\vspace{-0.1in}
-\caption{What's achievable for the first two players in a three-player
-  generalized BM$n$L game, with $n$ ranging from 1 cup state (darkest)
-  to 5 (lightest). TODO WRITEME
-% 6 cupstates ran overnight and got here:
-% #26186200000 queue 109 min 49 466729 states/sec 796272 config/sec
-}
-\label{fig:composite2player}
-\end{figure}
+\section{Conclusion}
 
-\section{Known bounds}
+This section summarizes the known bounds for \bmsl, and states some
+conjectures, before concluding.
 
-Define BMnL, where $n$ is the number of non-filled states. BMML=BM2L.
-(XXX actually I think it's the total number of states including filled
-everywhere but that sentence, right?)  Find $n$ such that a solo or
-dual power hour achieves any $k$.
-
+\subsection{Known results}
 \begin{enumerate}
@@
     drink 0--60 shots. Since each player can just play independently,
-    this result extends to any number of players in BM8L.
-  \item With two players, we know that BM5L (XXX check) does not allow
-    all combinations of $\langle k_1, k_2 \rangle$ shots, though BM5L
-    can achieve any $\langle k_1, k_2 \rangle$ where $\k_2 \in \{ 0, 1
-    \}$. No other $k_2$ can be used universally, though of course many
+    this result extends to any number of players in BM8L. With fewer
+    than 8 cup states, not every $k$ can be achieved alone.
+  \item With two players in BM3L or BM4L, it is not possible for one
+    of the players to drink every $k$ even if the other player helps her
+    out.
+  \item With two players, we know that BM5L does allow one player to
+    drink any $k_1$ if the other player assists. In fact we can
+    achieve any $\langle k_1, k_2 \rangle$ where $k_2 \in \{ 0, 1 \}$.
+    No other $k_2$ can be used universally, though of course many
     other combinations are possible
-    (Figure~\ref{fig:compositeplayer2mock}).
-  \item Open: Can BM6L achieve all $\langle k_1, k_2 \rangle$?
+    (Figure~\ref{fig:composite2player}).
+  \item Open: Can BM6L achieve all $\langle k_1, k_2 \rangle$? Seems
+    unlikely, given that random exploration plateaus with about 95.65\%
+    of the grid filled.
   \item Open: Can BM7L achieve all $\langle k_1, k_2 \rangle$? Seems
-    likely, given how close BM7L is for a single player.
-  \item With three players one player can drink any number of shots in
-    BMML with two other players helping.
+    possible, but there are four (really two with symmetry) unsolved
+    combinations.
+  \item With three players in BM3L, one player can drink any number of
+    shots if the other two players help.
 \end{enumerate}
 
-\section{Conjectures} \label{sec:conjectures}
+\subsection{Conjectures} \label{sec:conjectures}
 
-With two friends, you can drink any amount
+\paragraph{Freedom: With two friends, you can drink any amount.} We know
+that in a three-player game of BM3L, one of the three players can
+drink the $k$ of her choice. This straightforwardly extends to $3
+\times p$-player BM3L games. The {\bf Freedom} conjecture is that
+with $p + 2$ players in BM3L, $p$ of them may have their choice of
+$k_1 \ldots k_n$ drinks. If this conjecture fails, it probably fails
+for 4 players, which might have a feasible enumeration strategy.
 
-Teetotaller: When $\langle k_1, \ldots, k_i, \ldots, k_n \rangle$ is possible,
-so is $\langle k_1, \ldots, 0, \ldots, k_n \rangle$ (under what conditions)?
+\paragraph{Teetotaller: Someone can drink nothing.} When $\langle k_1, 
+\ldots, k_i, \ldots, k_p \rangle$ is achievable in \bmsl, so is
+$\langle k_1, \ldots, 0, \ldots, k_p \rangle$. This conjecture would
+be trivial if not for the rule that requires us to drink the contents
+of a full cup if we want to pass it in a different state. This
+conjecture is true for all the graphics presented in this
+paper;\footnote{Actually, it is not established for (only) 11 minutes
+  in Figure~\ref{fig:powerthreehours}, but this is not a proper
+  \bmsl\ game as it takes place over 180 minutes. There may very well
+  be solutions for 11 anyway; this is just a sample.} we can see that
+cells in the 0 column are always filled when some other cell in that
+row is filled. If this conjecture fails, it probably fails for BM1L or
+BM2L which have the least freedom per player.
 
-\section{Some stuffff}
+\medskip
 
-Hi
+In this paper I present some new results in \kn\ Power-Hour theory, as
+well as correct the historical record of some inebriated missteps. We
+see that stochastic simulation, abstract interpration, and sampling
+are powerful tools for solving combinatorial drinking problems. We
+established some firm results for the classic game and some bounds for
+generalizations, as well as informally looked at some visualizations
+of its geometric structure. However, there are still several open
+problems in this field that demand further study.
 

### 2013-12-29 powerhour/paper/paper.tex
First submission

@@
 % \usepackage{cite}
-% \usepackage{eclbkbox}
 \usepackage{amsmath}
 \usepackage{amssymb}
-% \usepackage{code}
-% \usepackage{amscd}
-% \usepackage{xy}
 \usepackage{graphicx}
-% \usepackage{fancyhdr}
-% \usepackage{color}
-% \usepackage[dark,all,bottom,landscape,timestamp]{draftcopy}
-% \usepackage{everypage}
 
-% \pagestyle{empty}
+\pagestyle{empty}
 
@@
 \usepackage{natbib}
-% \usepackage{inlinebib}
 
@@
 
-\maketitle
+\maketitle \thispagestyle{empty}
 
 \begin{abstract}
-Something about the k-n hours
-
+We correct for inebriated missteps, using computational methods to
+establish new bounds in generalized \kn Power-Hour theory.
 \end{abstract}
@@
 executed for each player, producing an $s$-dimensional vector $[ d_1,
-  \ldots d_s ]$ for each player. That player is able to achieve many
+  \ldots, d_s ]$ for each player. That player is able to achieve many
 different drink totals, specifically, $d_1 \times r_1 + \ldots + d_s
@@
 in the \fullcup\ position does not output as \fullcup, but this is
-inexact, as that rule may never be executed\footnote{We may be able to
+inexact, as that rule may never be executed.\!\footnote{We may be able to
   argue that in that case, there always exists another game that does
@@
 appears in Figure~\ref{fig:samplebm6l}. The sampling procedure runs
-for many hours before plateauing overnight with 95.54\% of the matrix
+for many hours before plateauing overnight with 95.65\% of the matrix
 filled. This suggests that BM6L is not universal for two players, or
-else the configurations for the missing cells are extremely rare.
-% XXX never give up! It just jumped to 95.65.
+else the configurations for the missing cells are extremely
+rare.\!\footnote{This is definitely a possibility, as new cells were
+still appearing after exploring tens of billions of samples. However,
+the gap here seems quite large.}
 
@@
 \end{center}\vspace{-0.1in}
-\caption{ 25.6 billion samples of legal two-player BM6L
+\caption{ 37.1 billion samples of legal two-player BM6L
   configurations. Darker cells represent outcomes that occur more
@@
   %
-  95.54\% of the cells are filled.
-  % XXX new version with more filled...!
+  95.65\% of the cells are filled.
 }
@@
 
-\begin{figure}
+\begin{figure}[ht!]
 \begin{center}
@@
 though they are fundamentally different (adding players, adding
-states, adding random trials). This suggests that the combinatorial
+states, adding random trials). Even samples from BMML with three
+players (3D projected to 2D), which is shown in Figure~\ref{fig:3d},
+produces a similar pattern. This suggests that the combinatorial
 problem (``what outcomes are reachable from finite state machines that
@@
 must start with a cup in order to get an outcome in the bottom-right
-half. We also see distinct clumps around 0, 15, 30, 45, 60; divisors
-of 60. This is intuitive because the expressive power of BMML comes
-from the ability to form cycles of cup states and drink on some
-fraction of them. Clumps are formed around these values because of the
-possibility of preludes leading into the cycles
+half. We also see distinct clumps around 0, 15, 30, 45, 60; these
+correspond to simple fractions (``drink every other time''; ``drink
+three of four times'') of 60. This is intuitive because the expressive
+power of BMML comes from the ability to form cycles of cup states and
+drink on some fraction of them. Clumps are formed around these values
+because of the possibility of preludes leading into the cycles
 (Figure~\ref{fig:solo31}) that either drink (adding to the total) or
@@
 major ones, because a player may participate in two cycles of
-different length (Figure~\ref{fig:dual45}.
+different length (Figure~\ref{fig:dual45}).
 
@@
 \begin{center}
+\includegraphics[width=0.90 \linewidth]{sample60min3players3states.png}
+\end{center}\vspace{-0.1in}
+\caption{490 million samples of three-player BMML Power Hours. The
+  cube is rotated $15\,^{\circ}$ along each axis, the top-left corner is
+  $\langle 0, 0, 0 \rangle$, and the bottom right is $\langle 60, 60,
+  60 \rangle$. }
+\label{fig:3d}
+\end{figure}
+
+\begin{figure}
+\begin{center}
 \includegraphics[width=0.90 \linewidth]{powerday3.pdf}
@@
   but at this scale makes it clear the groupings and their sparsity in
-  the limit. Lines plotted from $\langle 0,0 \rangle$ to $langle 60,
+  the limit. Lines plotted from $\langle 0,0 \rangle$ to $\langle 60,
   k_2 \rangle$ and $\langle k_1, 60 \rangle$ show significant
@@
 
-\section{Known bounds}
+The less extreme \kn\ Power-Three-Hours appears in
+Figure~\ref{fig:powerthreehours}.
 
-Define \bmsl, where $s$ is the number of cup states. BMML=BM3L. Find
-$s$ such that a solo or dual power hour achieves any $k$.
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.90 \linewidth]{sample180min2players7states.png}
+\end{center}\vspace{-0.1in}
+\caption{6,456,764,116 samples of two-player BM7L Power Three-Hours.
+}
+\label{fig:powerthreehours}
+\end{figure}
 
+\section{Conclusion}
+
+This section summarizes the known bounds for \bmsl, and states some
+conjectures, before concluding.
+
+\subsection{Known results}
 \begin{enumerate}
@@
     drink 0--60 shots. Since each player can just play independently,
-    this result extends to any number of players in BM8L.
-  \item With two players, we know that BM5L (XXX check) does not allow
-    all combinations of $\langle k_1, k_2 \rangle$ shots, though BM5L
-    can achieve any $\langle k_1, k_2 \rangle$ where $k_2 \in \{ 0, 1
-    \}$. No other $k_2$ can be used universally, though of course many
+    this result extends to any number of players in BM8L. With fewer
+    than 8 cup states, not every $k$ can be achieved alone.
+  \item With two players in BM3L or BM4L, it is not possible for one
+    of the players to drink every $k$ even if the other player helps her
+    out.
+  \item With two players, we know that BM5L does allow one player to
+    drink any $k_1$ if the other player assists. In fact we can
+    achieve any $\langle k_1, k_2 \rangle$ where $k_2 \in \{ 0, 1 \}$.
+    No other $k_2$ can be used universally, though of course many
     other combinations are possible
@@
   \item Open: Can BM6L achieve all $\langle k_1, k_2 \rangle$? Seems
-    unlikely, given that random exploration plateaus at about 95.54\%.
+    unlikely, given that random exploration plateaus with about 95.65\%
+    of the grid filled.
   \item Open: Can BM7L achieve all $\langle k_1, k_2 \rangle$? Seems
-    possible, given how close BM7L is for a single player. We know it
-    can get 99.89\%, at least!
-  \item With three players one player can drink any number of shots in
-    BMML with two other players helping.
+    possible, but there are four (really two with symmetry) unsolved
+    combinations.
+  \item With three players in BM3L, one player can drink any number of
+    shots if the other two players help.
 \end{enumerate}
 
-\section{Conjectures} \label{sec:conjectures}
+\subsection{Conjectures} \label{sec:conjectures}
 
-With two friends, you can drink any amount
+\paragraph{Freedom: With two friends, you can drink any amount.} We know
+that in a three-player game of BM3L, one of the three players can
+drink the $k$ of her choice. This straightforwardly extends to $3
+\times p$-player BM3L games. The {\bf Freedom} conjecture is that
+with $p + 2$ players in BM3L, $p$ of them may have their choice of
+$k_1 \ldots k_n$ drinks. If this conjecture fails, it probably fails
+for 4 players, which might have a feasible enumeration strategy.
 
-Teetotaller: When $\langle k_1, \ldots, k_i, \ldots, k_p \rangle$ is possible,
-so is $\langle k_1, \ldots, 0, \ldots, k_p \rangle$ (under what conditions)?
+\paragraph{Teetotaller: Someone can drink nothing.} When $\langle k_1, 
+\ldots, k_i, \ldots, k_p \rangle$ is achievable in \bmsl, so is
+$\langle k_1, \ldots, 0, \ldots, k_p \rangle$. This conjecture would
+be trivial if not for the rule that requires us to drink the contents
+of a full cup if we want to pass it in a different state. This
+conjecture is true for all the graphics presented in this
+paper;\footnote{Actually, it is not established for (only) 11 minutes
+  in Figure~\ref{fig:powerthreehours}, but this is not a proper
+  \bmsl\ game as it takes place over 180 minutes. There may very well
+  be solutions for 11 anyway; this is just a sample.} we can see that
+cells in the 0 column are always filled when some other cell in that
+row is filled. If this conjecture fails, it probably fails for BM1L or
+BM2L which have the least freedom per player.
 
-\section{Some stuffff}
+\medskip
 
-Hi
+In this paper I present some new results in \kn\ Power-Hour theory, as
+well as correct the historical record of some inebriated missteps. We
+see that stochastic simulation, abstract interpration, and sampling
+are powerful tools for solving combinatorial drinking problems. We
+established some firm results for the classic game and some bounds for
+generalizations, as well as informally looked at some visualizations
+of its geometric structure. However, there are still several open
+problems in this field that demand further study.
 

### 2013-12-31 trunk/powerhour/paper/paper.tex
Revise paper, establishing BM7L is complete with two players, yay

@@
 
-% \pagestyle{empty}
+\pagestyle{empty}
 
@@
 
-\maketitle
+\maketitle \thispagestyle{empty}
 
@@
 open problem is BM7L, whose results are in
-Figure~\ref{fig:samplebm7l}. This is tantalizingly close to a complete
-matrix, with only 4 cells missing! It may be possible to solve these
-two cases manually, though a proof of non-existence would be
-difficult.
+Figure~\ref{fig:samplebm7l}. Indeed, after more than 30 billion
+samples the matrix is completely filled in; we have found an example
+configuration that achieves every outcome. Some of these were
+extremely rare, such as the solution for $\langle 11, 53 \rangle$
+(Figure~\ref{fig:dual11-53}). In BM7L, two players can drink any
+amount.
 
@@
 \end{center}\vspace{-0.1in}
-\caption{ 22.1 billion samples of legal two-player BM7L
+\caption{ 30.7 billion samples of legal two-player BM7L
   configurations. Fewer configurations were sampled than in
@@
   6-state games to simulate, and a smaller proportion of random games
-  are legal.
-  %
-  All but four of the $(60 + 1)^2 = 3721$ cells are attainable: 
-  Permutations of $\langle 11, 53 \rangle$ and $\langle 49, 53 \rangle$
-  remain elusive. Note that 53 drinks was also unattainable in a solo
-  BM7L power-hour; this may in some sense be the ``hardest'' number of
-  shots to drink in \bmsl.
+  are legal. Moreover, we stop after finding a solution for every
+  cell, proving that BM7L is complete! The last cells found---an
+  earlier version of this paper held these as open problems!---were
+  permutations of $\langle 11, 53 \rangle$
+  (Figure~\ref{fig:dual11-53}) and $\langle 49, 53 \rangle$. Note that
+  53 drinks was also unattainable in a solo BM7L power-hour; this may
+  in some sense be the ``hardest'' number of shots to drink in \bmsl.
 }
 \label{fig:samplebm7l}
+\end{figure}
+
+\begin{figure}[ht!]
+\begin{center}
+\includegraphics[width=0.90 \linewidth]{dual11-53.pdf}
+\end{center}\vspace{-0.1in}
+\caption{ A solution for the elusive $\langle 11, 53 \rangle$ two-player
+  BM7L power hour! This one was found after 30.7 billion outcomes sampled,
+  making it the most rare. By the end of the game, the two players are
+  just passing back and forth two cups in state 5. A long lead-in beginning
+  on the left-player's state 2 spans 12 different configurations (the 6
+  cup states for the two players) before entering the length-two cycle.
+  Both cups travel along this lead-in, with one three steps ahead of the
+  other. The right player drinks on every step until reaching the cycle
+  (and then never again) for 11; the left player drinks during the cycle
+  plus a little extra during the lead-in for 53. This configuration is
+  quite flexible because the two players can make fine adjustments to
+  their drink total by drinking or not drinking on the lead-in rules,
+  which are executed just once or twice.}
+\label{fig:dual11-53}
 \end{figure}
@@
 matrix is more populous than the bottom right, for example. This is
-because we can bound the total number of drinks by $60 / c$, where $c$
-is the number of cups active in the game (same as the number of cups
-in the starting states). The top-left half is the region where this
-sum is less than or equal to 30; in two player games, both players
-must start with a cup in order to get an outcome in the bottom-right
-half. We also see distinct clumps around 0, 15, 30, 45, 60; these
-correspond to simple fractions (``drink every other time''; ``drink
-three of four times'') of 60. This is intuitive because the expressive
-power of BMML comes from the ability to form cycles of cup states and
-drink on some fraction of them. Clumps are formed around these values
-because of the possibility of preludes leading into the cycles
-(Figure~\ref{fig:solo31}) that either drink (adding to the total) or
-don't (subtracting from it). Minor clumps form as echoes between the
-major ones, because a player may participate in two cycles of
-different length (Figure~\ref{fig:dual45}).
+because we can bound the total number of drinks by $60 \times c$,
+where $c$ is the number of cups active in the game (same as the number
+of cups in the starting states). The top-left half is the region where
+this sum is less than or equal to 60; in two player games, both
+players must start with a cup in order to get an outcome in the
+bottom-right half. We also see distinct clumps around 0, 15, 30, 45,
+60; these correspond to simple fractions (``drink every other time'';
+``drink three of four times'') of 60. This is intuitive because the
+expressive power of BMML comes from the ability to form cycles of cup
+states and drink on some fraction of them. Clumps are formed around
+these values because of the possibility of preludes leading into the
+cycles (Figures~\ref{fig:solo31}, \ref{fig:dual11-53}) that either
+drink (adding to the total) or don't (subtracting from it). Minor
+clumps form as echoes between the major ones, because a player may
+participate in two cycles of different length
+(Figure~\ref{fig:dual45}).
 
@@
 we exploit that exact structure in Section~\ref{sec:linear}, in fact.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.90 \linewidth]{sample60min3players3states.png}
+\end{center}\vspace{-0.1in}
+\caption{490 million samples of three-player BMML Power Hours. The
+  cube is rotated $15\,^{\circ}$ along each axis, the top-left corner is
+  $\langle 0, 0, 0 \rangle$, and the bottom right is $\langle 60, 60,
+  60 \rangle$. }
+\label{fig:3d}
+\end{figure}
 
@@
     other combinations are possible
-    (Figure~\ref{fig:composite2player}).
+    (Figure~\ref{fig:composite2player}). Many pairs $\langle k_1, k_2
+    \rangle$ are known to be unattainable; this was established by
+    exhaustively testing all possible configurations.
   \item Open: Can BM6L achieve all $\langle k_1, k_2 \rangle$? Seems
@@
     of the grid filled.
-  \item Open: Can BM7L achieve all $\langle k_1, k_2 \rangle$? Seems
-    possible, but there are four (really two with symmetry) unsolved
-    combinations.
+  \item BM7L can achieve all $\langle k_1, k_2 \rangle$. This was
+    established by sampling random games until we found an example
+    for every $\langle k_1, k_2 \rangle$.
   \item With three players in BM3L, one player can drink any number of
@@
   in Figure~\ref{fig:powerthreehours}, but this is not a proper
-  \bmsl\ game as it takes place over 180 minutes. There may very well
-  be solutions for 11 anyway; this is just a sample.} we can see that
+  \bmsl\ game as it takes place over 180 minutes. There should be
+  solutions for 11, like in Figure~\ref{fig:dual11-53}; this is just
+  a sample.} we can see that
 cells in the 0 column are always filled when some other cell in that
@@
 
-In this paper I present some new results in \kn\ Power-Hour theory, as
+In this paper I presented some new results in \kn\ Power-Hour theory, as
 well as correct the historical record of some inebriated missteps. We
-see that stochastic simulation, abstract interpration, and sampling
+saw that stochastic simulation, abstract interpration, and sampling
 are powerful tools for solving combinatorial drinking problems. We

### 2014-01-20 trunk/midinverse/paper.tex
From plane: Tag 'range' version. Begin work on paper.

\documentclass[twocolumn]{article}
\usepackage[top=1.1in, left=0.85in, right=0.85in]{geometry}

\usepackage{url}
\usepackage{code}
% \usepackage{cite}
\usepackage{amsmath}
\usepackage{amssymb}
% \usepackage{graphicx}

\pagestyle{empty}

\usepackage{ulem}
% go back to italics for emphasis, though
\normalem

\usepackage{natbib}

\newcommand\comment[1]{}
\newcommand\sfrac[2]{\!{}\,^{#1}\!/{}\!_{#2}}

\begin{document}

\title{What is the opposite of a song?}
\author{Dr.~Tom~Murphy~VII~Ph.D.\thanks{
Copyright \copyright\ 2014 the Regents of the Wikiplia
Foundation. Appears in SIGBOVIK 2014 with whatever the
opposite of permission of the 
Association for Computational Heresy is; {\em IEEEEEE!} press,
Verlag-Verlag volume no.~0x40-2A.
\$0.00}
}

\renewcommand\th{\ensuremath{{}^{\textrm{th}}}}
\newcommand\st{\ensuremath{{}^{\textrm{st}}}}
\newcommand\rd{\ensuremath{{}^{\textrm{rd}}}}
\newcommand\nd{\ensuremath{{}^{\textrm{nd}}}}
\newcommand\at{\ensuremath{\scriptstyle @}}

\date{1 April 2014}

\maketitle \thispagestyle{empty}

\begin{abstract}
what do I put here
\end{abstract}

\vspace{1em}
{\noindent \small {\bf Keywords}:
  musical instrument digital inverse,
  ...
}

\section*{Introduction}

Opposite Day, which is known in mathematics as {\it duality}, is a
powerful technique for turning good things into bad things. For
example, it can turn the perfectly straightforward reasoning technique
of {\it induction} into the real head-scratcher {\it coinduction}, or
the righteous and valid {\it true} into the muschevious and degenerate
{\it false}. Dualization can also turn bad things into good, of
course. In this paper, I investigate what kinds of formal notions of
duality can be employed to answer the question: What is the opposite
of a song?

\paragraph{Note.} Not since the 80s when in extreme novelty circumstances a magazine or
Happy Meal would come with a crap phonograph engraved into flimsy acetate has
it been possible to include audio material in printed conference
proceedings, and as you might imagine, the work presented here does
involve sound. Therefore I recommend reading this paper with
simultaneous access to the audio recordings at {XXX TODO}, or watching
the video summary linked to from that site. For those who've become
marooned on a desert isle with the complete SIGBOVIK proceedings, the
hearing-impaired, post-human societies attempting to recreate 2014-era
science and technology from the academic record of prestigious
conferences, or the moderately lazy, I have included over-the-top
textual descriptions of the songs in this work.

% ideas: Graphical transpose of piano roll -- long notes (horizontal
% bar) become many short notes played at once (vertical bar).

% convert MIDI to bitmap and perform invertible photoshop-like filters.
% the above becomes a 90\deg rotation of 127-tick-long chunks, for
% example.

% AES encryption?!

\section{What is a song?}

It should be easy to find dozens of wankerholes who have tried to
define ``song'' and some other art jokers who have pushed the
boundaries of these definitions. The author approves of this
mischief---the prions of music---but this paper is not concerned with
that tired analysis. For the purposes of this paper, we'll use only a
simple and practical high-level encoding of music, MIDI.

\paragraph{MIDI.} It stands for Musical Instrument Digital Interface
and you can tell from the need to clarify that it is ``digital'' that
it is as old as old dead dry dust and just barely adequate for modern
applications, and so has never been successfully superseded. It
permits 16 {\it channels}, each more-or-less a separate instrument,
with 128 {\it notes} per channel. Each of the 128 MIDI notes has a
fixed meaning, from 0 = C1 \comment{XXX?}, which is 5 octaves below
middle C (about 8 Hz\footnote{A note like ``middle C'' has a fixed
  semantics in some wishy-washy music sense, but the actual frequency
  of the tone is determined by mapping that meaning into one of
  several conventional scales. For example, in ``concert pitch'',
  middle A is fixed to 220 Hz (so middle C is XXX Hz), but there is
  also ``computer science pitch'', where middle C is 256 Hz. The
  techniques here are mostly insensitive to the absolute pitch values,
  but we use computer science pitch because what the hey. }) to 127 =
G11 (about 12,274 Hz). A track consists of a time-ordered series of
events, with the main events being ``turn note $n$ on with at volume
$v$'' and ``turn note $n$ off''\footnote{I still haven't decided how I
  feel about this, but the note-off event also includes a velocity,
  which I guess means something like ``how hard you yank your finger
  off the piano key''. It is usually ignored and in fact the
  convention actually seems to encode note-off as ``note-on with
  volume 0''. In this paper I just use note-off to mean whatever
  concrete event results in the note being inaudible.}; volumes also
range 0--127, allowing for every possible nuance of expression. Notes
can be turned on and off arbitrarily, but we restrict ourselves to the
sensible subset where a note-on event always precedes exactly one
note-off event for the same note on the same track, with no nesting or
overlaps or other shenanigans. For such MIDIs, there exists a simple
isomorphism between the event-based representation and a ``bitmap''
representation where for each MIDI tick, each note in each track has a
volume level 0--127, where 0 is predominant and means ``off''. Of course
there are a jillion other oddities to MIDI, but this will suffice as
a definition of a ``song'' for now.

%  127 g11
%  126 f#
%  125 f
%  124 e 
%  123 d# 
%  122 d 
%  121 c#
%  120 c11
%  108 c10
%  96  c9 
%  84  c8
%  72  c7
%  60 = middle c
%  48  c5
%  36  c4
%  24  c3
%  12  c2
%  0   c1

\section{What is an opposite?}

There are many formal and informal ideas of ``opposite''. Here I will
use a very loose interpretation of opposite so that we can explore
many different reasonable ideas. We define not the opposite of a
specific thing, but a method for getting the opposite of all of the
things of some type. So an opposite of some thing of type $A$ will be a
function $f$ of type $A \rightarrow A$ where
%
\begin{itemize}
\item $\exists x{:}A. f(x) \neq x$
\item $\forall x{:}A. f(f(x)) = x$
\end{itemize}

The first condition simply says that the opposite function must not be
the identity function; there must be some things that are not their
own opposites. Requiring that nothing be its own opposite is quite a
strong condition and excludes some things we would normally think of
as opposites. (For example, it seems reasonable to say the opposite of
$3$ is $-3$, but for negation, the opposite of $0$ is still $0$.) For
songs that are their own opposites, see
Section~\ref{sec:selfopposites}.

The second condition is that f is its own inverse, which expresses
the idea that items ``pair up'' into their opposites. So if the opposite
of day is night, then we expect the opposite of night to be day.

In the case of this paper, the type $A$ will be ``song'' (MIDI file)
and so we'll look at a bunch of different functions $f$ for turning
one MIDI file into another. If we run the function on a MIDI file and
get its opposite, then run it again to get the opposite of that thing,
we'll be back where we started, like if there were an Opposite Hour
within Opposite Day where everything is normal.

(XXX about what this idea is in mathematics--is it an involution?)

\section{Well-known opposites}

Others have asked this question before (XXX cite beatles, etc.). One
idea is so old we'll just mention it and move on: Playing the song
in reverse. (XXX cover it briefly?)

% any rotation mod 127 (except 0)

\section{Self-opposites} \label{sec:selfopposites}

\section{Conclusion}

\bibliography{paper}{}
\bibliographystyle{plain}
\end{document}

### 2014-01-25 trunk/midinverse/paper.tex
Add ability to permute notes within an octave

@@
 \usepackage{url}
-\usepackage{code}
+% \usepackage{code}
 % \usepackage{cite}
@@
 \usepackage{amssymb}
-% \usepackage{graphicx}
+\usepackage{graphicx}
 
@@
 applications, and so has never been successfully superseded. It
-permits 16 {\it channels}, each more-or-less a separate instrument,
-with 128 {\it notes} per channel. Each of the 128 MIDI notes has a
-fixed meaning, from 0 = C1 \comment{XXX?}, which is 5 octaves below
-middle C (about 8 Hz\footnote{A note like ``middle C'' has a fixed
-  semantics in some wishy-washy music sense, but the actual frequency
-  of the tone is determined by mapping that meaning into one of
-  several conventional scales. For example, in ``concert pitch'',
-  middle A is fixed to 220 Hz (so middle C is XXX Hz), but there is
-  also ``computer science pitch'', where middle C is 256 Hz. The
-  techniques here are mostly insensitive to the absolute pitch values,
-  but we use computer science pitch because what the hey. }) to 127 =
-G11 (about 12,274 Hz). A track consists of a time-ordered series of
-events, with the main events being ``turn note $n$ on with at volume
-$v$'' and ``turn note $n$ off''\footnote{I still haven't decided how I
-  feel about this, but the note-off event also includes a velocity,
-  which I guess means something like ``how hard you yank your finger
-  off the piano key''. It is usually ignored and in fact the
-  convention actually seems to encode note-off as ``note-on with
-  volume 0''. In this paper I just use note-off to mean whatever
-  concrete event results in the note being inaudible.}; volumes also
-range 0--127, allowing for every possible nuance of expression. Notes
-can be turned on and off arbitrarily, but we restrict ourselves to the
-sensible subset where a note-on event always precedes exactly one
-note-off event for the same note on the same track, with no nesting or
-overlaps or other shenanigans. For such MIDIs, there exists a simple
+permits 16 {\it channels}, each more-or-less a separate
+instrument,\!\footnote{It's really more like 16 members of an
+  orchestra, each of whom can only play one instrument at a time, but
+  we'll make the additional simplification in this paper that each
+  channel corresponds to a single instrument.} with 128 {\it notes}
+per channel. Each of the 128 MIDI notes has a fixed meaning, from 0 =
+C$_{-1}$, which is 5 octaves below middle C (about 8~Hz\footnote{A
+  note like ``middle C'' has a fixed semantics in some wishy-washy
+  music sense, but the actual frequency of the tone is determined by
+  mapping that meaning into one of several conventional scales. For
+  example, in ``concert pitch'', middle A is fixed to 220 Hz (so
+  middle C is 261.63~Hz), but there is also ``computer science
+  pitch'', where middle C is 256~Hz. The techniques here are mostly
+  insensitive to the absolute pitch values, but we use computer
+  science pitch because what the hey. }) to 127 = G9 (about
+12,274~Hz). A track consists of a time-ordered series of events, with
+the main events being ``turn note $n$ on with at volume $v$'' and
+``turn note $n$ off'';\footnote{I still haven't decided how I feel
+  about this, but the note-off event also includes a velocity, which I
+  guess means something like ``how hard you yank your finger off the
+  piano key''. It is usually ignored and in fact the convention
+  actually seems to encode note-off as ``note-on with volume 0''. In
+  this paper I just use note-off to mean whatever concrete event
+  results in the note being inaudible.} volumes also range 0--127,
+allowing for every possible nuance of expression. Notes can be turned
+on and off arbitrarily, but we restrict ourselves to the sensible
+subset where a note-on event always precedes exactly one note-off
+event for the same note on the same track, with no nesting or overlaps
+or other shenanigans. For such MIDIs, there exists a simple
 isomorphism between the event-based representation and a ``bitmap''
 representation where for each MIDI tick, each note in each track has a
-volume level 0--127, where 0 is predominant and means ``off''. Of course
-there are a jillion other oddities to MIDI, but this will suffice as
-a definition of a ``song'' for now.
+volume level 0--127, where 0 is predominant and means ``off''. Of
+course there are a jillion other oddities to MIDI, but this will
+suffice as a definition of a ``song'' for now.
 
@@
 things of some type. So an opposite of some thing of type $A$ will be a
-function $f$ of type $A \rightarrow A$ where
+total function $f$ of type $A \rightarrow A$ where
 %
@@
 the idea that items ``pair up'' into their opposites. So if the opposite
-of day is night, then we expect the opposite of night to be day.
+of day is night, then we expect the opposite of night to be day. This
+is called an ``involution'' in mathematics.
 
@@
 within Opposite Day where everything is normal.
-
-(XXX about what this idea is in mathematics--is it an involution?)
 

### 2014-02-22 trunk/epsilon/paper.tex
cxpoint to hacksession

\documentclass[twocolumn]{article}
\usepackage[utf8]{inputenc}
\usepackage[top=1.1in, left=0.85in, right=0.85in]{geometry}

% \usepackage{eclbkbox}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{code}
% \usepackage{amscd}
% \usepackage{xy}
\usepackage{graphicx}

% \pagestyle{empty}

% \usepackage{ulem}
% go back to italics for emphasis, though
% \normalem

\usepackage{natbib}

\begin{document}

\title{What, if anything, is epsilon?}
\author{Dr.~Tom~Murphy~VII~Ph.D.\thanks{
Copyright \copyright\ 2013 the Regents of the Wikiplia
Foundation. Appears in SIGBOVIK 2013 with the careless accounting of the
Association for Computational Heresy; {\em IEEEEEE!} press,
Verlag-Verlag volume no.~0x40-2A.
\yen 0.00}
}

\renewcommand\>{$>$}
\newcommand\<{$<$}

\date{1 April 2014}

\maketitle

\begin{abstract}
what do I put here
\end{abstract}

\vspace{1em}
{\noindent \small {\bf Keywords}:
 computational archaeology, epsilon, very-small and medium-small numbers
}

\section*{Introduction}

Epsilon, the all-spelled-out version of $\epsilon$, although properly
``epsilon'' because it's the {\em lowercase} Greek letter, though it's
not like I'm going to start my paper with a lowercase letter even if
it's technically correct, since I like to wait at least until the
second or third letter of the paper before the reader starts doubting
that I can write or spell or have shift-keys on my keyboard, anyway
epsilon is a mathematical symbol denoting a {\em very small number}.

% (intro about math definition in \epsilon--\delta formulation of limits.
% it's not that complicated)

In mathematics, $\epsilon$ usually refers to the
$\epsilon--\delta$ formulation of limits. This is pretty simple and a
reminder appears in Figure~\ref{fig:epsilondelta}. This paper is not
about that kind of math.

In computing, $\epsilon$ is used in a much more general sense to just
mean some small number or error bound. For example, two numbers are
often considered equal if their absolute difference is less than
$\epsilon$.

In IEEE-754 floating-point~\cite{ieee754}, the standard way that
computers represent ``real numbers,'' there is a specific formal value
called ``machine epsilon'', or ``unit roundoff''. It is the maximum
(relative) amount of error from a single rounding operation. This
number is useful if you want to do numerical programming and be
careful about what you're doing.

Most programmers find this subject too tedious and simply pick a
number that seems pretty small. Thus in practice, $\epsilon$ is an
application-specific choice. Of well-known ``constants,''
$\epsilon$ may be the least agreed-upon, with values seen in the wild
spanning {\em 50 orders of magnitude}. This paper explores the practical
values of $\epsilon$ in real software.

\begin{figure}[ht]
\begin{center}
\includegraphics[width=0.90 \linewidth]{epsilondelta.pdf}
\end{center}\vspace{-0.1in}
\caption{explain the graph}
\label{fig:epsilondelta}
\end{figure}

\section{Methodology}

Did you know that the SPEC benchmarks~\cite{spec2006} cost \$800? Like
they literally expect me to pay them money to download the source code
so that I could grep for {\tt const double epsilon} or test my
compiler out on that. Mmany are even based on open-source software like
Sphinx and POV-Ray. Ridiculous. I refuse. Values of epsilon for the
SPEC benchmarks do not appear in Figure~\ref{fig:spec}.

\begin{figure}[ht]
\begin{center}
\includegraphics[width=0.99 \linewidth]{spec.pdf}
\end{center}\vspace{-0.1in}
\caption{\$800? Fuck that!}
\label{fig:spec}
\end{figure}

\section{Results}

\bibliography{paper}{}
\bibliographystyle{plain}
\end{document}

### 2014-02-22 chesstego/paper.tex
Create Chesstego with search

coauthors:

Jim McCann, in italics ...... oh! .... nervous laughter .... no, that's not part of the name still. I stopped saying the name. Why are you still typing? This isn't the name any more. End quote.

### 2014-03-01 epsilon/paper.tex
Progress on paper, tables, etc.

--- 2014-02-22 trunk/epsilon/paper.tex
@@
 In mathematics, $\epsilon$ usually refers to the
-$\epsilon--\delta$ formulation of limits. This is pretty simple and a
+$\epsilon$--$\delta$ formulation of limits. This is pretty simple and a
 reminder appears in Figure~\ref{fig:epsilondelta}. This paper is not
@@
 $\epsilon$ may be the least agreed-upon, with values seen in the wild
-spanning {\em 50 orders of magnitude}. This paper explores the practical
+spanning {\em 300 orders of magnitude}. This paper explores the practical
 values of $\epsilon$ in real software.
@@
 \end{center}\vspace{-0.1in}
-\caption{explain the graph}
+\caption{ The curvy line is the function $\frac{\sin x}{x}$, which is
+  undefined at $0$. However, its limit as $x$ approaches $0$ is $1$.
+  The $\epsilon$--$\delta$ formulation of the limit is this: For any
+  positive choice of $\epsilon$, there exists some $\delta$ such that
+  $\frac{\sin (0 + \delta)}{0 + \delta}$ is less than $1 + \epsilon$.
+  In other words, for an arbirarily small error (your choice), I
+  can produce a delta from $0$ (my choice) that brings the result
+  within the error of the limit. This has nothing to do with the
+  subject of the paper.}
 \label{fig:epsilondelta}
@@
 
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[width=0.99 \linewidth]{chart-c-double}
+\end{center}\vspace{-0.1in}
+\caption{ Values for {\tt const double epsilon} in the C programming
+  language. In this chart, the blue (lower) bars are the distinct values
+  of epsilon seen. The vertically aligned red (upper) bar is its count.
+  Epsilon values are plotted on a logarithmic scale, where the minimum
+  observed value $\log(-708)$ is $303 \times 10^{-308}$, and the largest
+  $\log(1.609)$ is $5$. 
+  %
+  Notes: One programmer used the value \tt{-1e10}, which is
+  -10,000,000,000, problably meaning \tt{1e-10}. This value was excluded
+  because it has no real logarithm.
+}
+\label{fig:cdouble}
+\end{figure}
+
+\begin{figure}[ht]
+\begin{center}
+\begin{tabular}{l}
+\verb+0.5/ELEC_REST_ENERGY+ \\
+\verb+alpha/beta+ \\
+\verb+4 / MULT32+ \\
+\verb+exact_epsilon(true)+ \\
+\verb+fmass_Epsilon * EPS_EXTRA+ \\
+\verb+((Lj_Parameters*) parameters)->+ \\
+\verb+scalar_traits<+ \\
+\verb+EpsArray[prec]+ \\
+\verb+hfwfn_->+ \\
+\verb+fl.net_.opt_.epsilon+ \\
+\verb+Tolerance+ \\
+\end{tabular}
+\end{center}\vspace{-0.1in}
+\caption{ Other uninterpretable values of {\tt const double epsilon}
+  in C. Who knows what these are supposed to be?}
+\label{fig:cdoubleuninterpretable}
+\end{figure}
+
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[width=0.99 \linewidth]{chart-c-float}
+\end{center}\vspace{-0.1in}
+\caption{ Values for {\tt const float epsilon} in the C programming
+  language. An $x$--$y$ scatter plot where the $y$ coordinate is the
+  count of the number of times that specific value occurred, and the
+  $x$ coordinate is the log of the value. Values take on a less
+  extreme range than with type {\tt double}, naturally, ranging ``only''
+  31 orders of magnitude from $3.9 \times 10^{-31}$ to $0$.}
+\label{fig:cfloat}
+\end{figure}
+
+\begin{figure}[ht]
+\begin{center}
+% \includegraphics[width=0.99 \linewidth]{chart-c-double}
+\end{center}\vspace{-0.1in}
+\caption{
+  Values for {\tt const double epsilon} and {\tt constexpr
+    double epsilon} in the C++ programming language. The {\tt
+    constexpr} qualifier, a new feature of C++11, is used very rarely
+  (less than 1\% of the time).
+  Notes:
+  Five times, the computed value is actually equal to 0.0,
+
+  The value {\tt pow(10,-13)} is annotated in German
+  ``Genauigkeitsziel bei der Nullstellensuche,'' or ``Accuracy goal in
+  the search for zeros.''
+}
+\label{fig:cppdouble}
+\end{figure}
+
 \bibliography{paper}{}

### 2014-03-01 epsilon/printlog.cc

@@ -1,0 +1,306 @@
+
+#include <limits>
+#include <cfloat>
+#include <vector>
+#include <algorithm>
+#include <stdio.h>
+#include <cmath>
+#include <map>
+
+using namespace std;
+
+// #define DBL_EPSILON std::numeric_limits<double>::epsilon()
+#define EPSILON FLT_EPSILON
+
+#define INIT(s) #s, s
+struct NamedEntry {
+  const char *expr;
+  double value;
+  int count;
+};
+
+NamedEntry entries_cdouble[] = {
+  { INIT(0.00000000000001), 1 },
+  { INIT(1.0E-6), 1 },
+  { INIT(1.0e-1), 1 },
+  { INIT(1.0E-20), 1 },
+  { INIT(2.0e-16), 1 },
+  { INIT(1e-13), 1 },
+  { INIT(1e-5), 1 },
+  { INIT(1.0E-32), 1 },
+  { INIT(-1e10), 1 },
+  { INIT(.00000000000000000000001), 1 },
+  { INIT(1e-3), 1 },
+  { INIT(1.776356839400251e-15), 1 },
+  { INIT(1.00e-07), 1 },
+  { INIT(.02), 1 },
+  { INIT(1.0e-50), 1 },
+  { INIT(2e-4), 1 },
+  { INIT(0.33), 1 },
+  { INIT(0.9), 1 },
+  { INIT(.000001), 1 },
+  { INIT(DBL_MIN), 1 },
+  { INIT(1.0E-10), 1 },
+  { INIT(EPSILON * EPSILON), 1 },
+  { INIT(1.0e-8), 1 },
+  { INIT(1.192092896e-06), 1 },
+  { INIT(5.0), 1 },
+  { INIT(1.0E-100), 1 },
+  { INIT(1e-7), 1 },
+  { INIT(1e-4), 1 },
+  { INIT(1e-100), 2 },
+  { INIT(1e-10f), 2 },
+  { INIT(1e-9), 2 },
+  { INIT(1.0e-10), 2 },
+  { INIT(1.0E-200), 2 },
+  { INIT(0.00000001f), 2 },
+  { INIT(1.0E-300), 2 },
+  { INIT(1.0e-6), 2 },
+  { INIT(1e-20), 2 },
+  { INIT(1.0E-14), 2 },
+  { INIT(1.0E-16), 2 },
+  { INIT(5.0E-16), 2 },
[…]

### 2014-03-01 trunk/epsilon/paper.tex
Did C++ double

@@
 number that seems pretty small. Thus in practice, $\epsilon$ is an
-application-specific choice. Of well-known ``constants,''
-$\epsilon$ may be the least agreed-upon, with values seen in the wild
-spanning {\em 50 orders of magnitude}. This paper explores the practical
-values of $\epsilon$ in real software.
+application-specific choice. Of well-known ``constants,'' $\epsilon$
+may be the least agreed-upon, with values seen in the wild spanning
+more than {\em 300 orders of magnitude}. This paper explores the
+practical values of $\epsilon$ in real software.
 
@@
 
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[width=0.99 \linewidth]{chart-c-double}
+\end{center}\vspace{-0.1in}
+\caption{ Values for {\tt const double epsilon} in the C programming
+  language. In this chart, the blue (lower) bars are the distinct values
+  of epsilon seen. The vertically aligned red (upper) bar is its count.
+  Epsilon values are plotted on a logarithmic scale, where the minimum
+  observed value $\log(-708)$ is $303 \times 10^{-308}$, and the largest
+  $\log(1.609)$ is $5$. 
+  %
+  Notes: One programmer used the value \tt{-1e10}, which is
+  -10,000,000,000, problably meaning \tt{1e-10}. This value was excluded
+  because it has no real logarithm.
+}
+\label{fig:cdouble}
+\end{figure}
+
+\begin{figure}[ht]
+\begin{center}
+\begin{tabular}{l}
+\verb+0.5/ELEC_REST_ENERGY+ \\
+\verb+alpha/beta+ \\
+\verb+4 / MULT32+ \\
+\verb+exact_epsilon(true)+ \\
+\verb+fmass_Epsilon * EPS_EXTRA+ \\
+\verb+((Lj_Parameters*) parameters)->+ \\
+\verb+scalar_traits<+ \\
+\verb+EpsArray[prec]+ \\
+\verb+hfwfn_->+ \\
+\verb+fl.net_.opt_.epsilon+ \\
+\verb+Tolerance+ \\
+\end{tabular}
+\end{center}\vspace{-0.1in}
+\caption{ Other uninterpretable values of {\tt const double epsilon}
+  in C. Who knows what these are supposed to be?}
+\label{fig:cdoubleuninterpretable}
+\end{figure}
+
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[width=0.99 \linewidth]{chart-c-float}
+\end{center}\vspace{-0.1in}
+\caption{ Values for {\tt const float epsilon} in the C programming
+  language. An $x$--$y$ scatter plot where the $y$ coordinate is the
+  count of the number of times that specific value occurred, and the
+  $x$ coordinate is the log of the value. Values take on a less
+  extreme range than with type {\tt double}, naturally, ranging ``only''
+  31 orders of magnitude from $3.9 \times 10^{-31}$ to $0$.}
+\label{fig:cfloat}
+\end{figure}
+
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[width=0.99 \linewidth]{chart-cpp}
+\end{center}\vspace{-0.1in}
+\caption{
+  Values for {\tt const double epsilon} and {\tt constexpr
+    double epsilon} in the C++ programming language. The {\tt
+    constexpr} qualifier, a new feature of C++11, is used very rarely
+  (less than 1\% of the time).
+  %
+  Notes:
+  Five times, the programmer used 0.0 for epsilon, which is possibly
+  the only unjustifiable value. The value {\tt pow(10,-13)} is
+  annotated in German ``Genauigkeitsziel bei der Nullstellensuche,''
+  or ``Accuracy goal in the search for zeros.'' }
+\label{fig:cppdouble}
+\end{figure}
+
 \bibliography{paper}{}

### 2014-03-01 trunk/epsilon/printlog.cc

@@ -285,7 +285,7 @@ static bool CompareByFirst(const NamedEntry &a, const NamedEntry &b) {
   return a.value < b.value;
 }
 
-#define entries entries_cfloat
+#define entries entries_cppdouble
 int main () {
   map<double, int> all;
   for (int i = 0; i < sizeof (entries) / sizeof (NamedEntry); i++) {

### 2014-03-01 chesstego/paper.tex
Add COPYING to epsilon

@@
-coauthors:
+\documentclass[twocolumn]{article}
+\usepackage[top=1.1in, left=0.85in, right=0.85in]{geometry}
 
-Jim McCann, in italics ...... oh! .... nervous laughter .... no, that's not part of the name still. I stopped saying the name. Why are you still typing? This isn't the name any more. End quote.
+\usepackage{url}
+% \usepackage{code}
+% \usepackage{cite}
+\usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{graphicx}
+\usepackage{chessboard}
+
+\pagestyle{empty}
+
+\usepackage{ulem}
+% go back to italics for emphasis, though
+\normalem
+
+\usepackage{natbib}
+
+% \newcommand\comment[1]{}
+\newcommand\sfrac[2]{\!{}\,^{#1}\!/{}\!_{#2}}
+
+\begin{document}
+
+\title{It still seems that black has hope in these extremely unfair variants of chess}
+\author{Dr.~Tom~Murphy~VII~Ph.D. \and Ben Blum \and
+Jim McCann, in italics ......
+oh!
+....
+nervous laughter ....
+no, that's not part of the name still.
+I stopped saying the name.
+Why are you still typing?
+This isn't the name any more.
+End quote.\thanks{
+Copyright \copyright\ 2014 the Regents of the Wikiplia
+Foundation. Appears in SIGBOVIK 2014 with the undivided
+attention of the Association for Computational Heresy; 
+{\em IEEEEEE!} press, Verlag-Verlag volume no.~0x40-2A.
+\$0.00}}
+
+\renewcommand\th{\ensuremath{{}^{\textrm{th}}}}
+\newcommand\st{\ensuremath{{}^{\textrm{st}}}}
+\newcommand\rd{\ensuremath{{}^{\textrm{rd}}}}
+\newcommand\nd{\ensuremath{{}^{\textrm{nd}}}}
+\newcommand\at{\ensuremath{\scriptstyle @}}
+
+\date{1 April 2014}
+
+\maketitle \thispagestyle{empty}
+
+\begin{abstract}
+what do I put here
+\end{abstract}
+
+\setchessboard{tinyboard,showmover=false}
+% \chessboard[
+%   setfen=rnbqkbnr/pppppppp/8/8/8/8/NNNBBNNN/QQQQKQQQ]
+
+\input{king2}
+
+% results:
+
+% No mate in 12 for 0-7. 6 ran for like 2 full days!
+% regular board.
+
+% No mate in 14 for 1,2,3,4,5,7. didn't run 7 because 6 is
+% still running overnight! 13h51m! 0 also running for 42h00m.
+
+% No mate in 10 for king at pawn index 0-7, even when
+% white's pieces are all queens (but the king)
+% (6 also finished last. Is this the weakest spot?)
+
+% Now mate in 10 for white having ``mega'' pieces
+
+% Mate in 10 for pawn idx 0! 441450 tree 
+%                 and idx 1! 210316 tree
+
+% TODO: Generate ``book'' where white is making strong moves
+% (black has few options) but black keeps it going for 10/12 moves?
+
+% \bibliography{paper}{}
+% \bibliographystyle{plain}
+\end{document}
+
+% rnbqkbnr/pppppppp/8/8/8/8/NNNBBNNN/QQQQKQQQ w KQkq - 0 1

### 2014-03-01 epsilon/paper.tex
Submitted version of Epsilon paper

@@
 \usepackage{code}
+\usepackage{url}
 % \usepackage{amscd}
@@
 
+\newcommand\CC{{C\nolinebreak[4]\hspace{-.05em}\raisebox{.4ex}{\tiny\bf ++}}}
+
 \date{1 April 2014}
@@
 \begin{abstract}
-what do I put here
+We present a sample of the values of the programming constant {\tt
+  epsilon} as found on the internet, for several different programming
+languages and with a variety of visualizations.
 \end{abstract}
@@
 number that seems pretty small. Thus in practice, $\epsilon$ is an
-application-specific choice. Of well-known ``constants,''
-$\epsilon$ may be the least agreed-upon, with values seen in the wild
-spanning {\em 300 orders of magnitude}. This paper explores the practical
-values of $\epsilon$ in real software.
+application-specific choice. Of well-known ``constants,'' $\epsilon$
+may be the least agreed-upon, with values seen in the wild spanning
+more than {\em 300 orders of magnitude}. This paper explores the
+practical values of $\epsilon$ in real software.
 
@@
 
-Did you know that the SPEC benchmarks~\cite{spec2006} cost \$800? Like
-they literally expect me to pay them money to download the source code
-so that I could grep for {\tt const double epsilon} or test my
-compiler out on that. Mmany are even based on open-source software like
-Sphinx and POV-Ray. Ridiculous. I refuse. Values of epsilon for the
-SPEC benchmarks do not appear in Figure~\ref{fig:spec}.
-
-\begin{figure}[ht]
-\begin{center}
-\includegraphics[width=0.99 \linewidth]{spec.pdf}
-\end{center}\vspace{-0.1in}
-\caption{\$800? Fuck that!}
-\label{fig:spec}
-\end{figure}
-
-\section{Results}
-
-\begin{figure}[ht]
-\begin{center}
-\includegraphics[width=0.99 \linewidth]{chart-c-double}
-\end{center}\vspace{-0.1in}
-\caption{ Values for {\tt const double epsilon} in the C programming
-  language. In this chart, the blue (lower) bars are the distinct values
-  of epsilon seen. The vertically aligned red (upper) bar is its count.
-  Epsilon values are plotted on a logarithmic scale, where the minimum
-  observed value $\log(-708)$ is $303 \times 10^{-308}$, and the largest
-  $\log(1.609)$ is $5$. 
-  %
-  Notes: One programmer used the value \tt{-1e10}, which is
-  -10,000,000,000, problably meaning \tt{1e-10}. This value was excluded
-  because it has no real logarithm.
-}
-\label{fig:cdouble}
-\end{figure}
+\paragraph{Github.} Github is the hub that contains all gits, approximately
+10~million of them, as of the beginning of 2014. The site has search
+functionality, which ``allowed'' me to scrape one hundred pages of
+results for queries like {\tt "const double epsilon ="} for various
+languages, as long as I didn't do it too fast. I scraped the
+programming languages C, \CC, C$\sharp$, JavaScript, and Objective~C.
+Each language has its own idiosyncracies about how constants are
+defined, so I used one (or several) appropriate to each language. For
+example, in JavaScript, I looked for {\tt "var epsilon ="}. From these
+HTML files I extracted all of the right-hand-side expressions,
+manually excluded the ones that could not be evaluated (for example
+because they depended on other symbols; see
+Figure~\ref{fig:cdoubleuninterpretable} for some examples), and then
+computed the actual values for the rest. The source code to do the
+scraping, extract the expressions, and tally the results is available
+online.\!\footnote{In the Subversion repository at:
+  \url{https://sourceforge.net/p/tom7misc/svn/HEAD/tree/trunk/epsilon/}}
 
@@
 
+\paragraph{SPEC benchmarks.} Did you know that the SPEC 
+benchmarks~\cite{spec2006} cost \$800? Like they literally expect me
+to pay them money to download the source code so that I could grep for
+{\tt const double epsilon} or test my compiler out on that. Many are
+even based on open-source software like Sphinx and POV-Ray.
+Ridiculous. I refuse. Values of epsilon for the SPEC benchmarks do not
+appear in Figure~\ref{fig:spec}.
+
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[width=0.99 \linewidth]{spec.pdf}
+\end{center}\vspace{-0.1in}
+\caption{\$800? Fuck that!}
+\label{fig:spec}
+\end{figure}
+
+\section{Results}
+
+The results of the analysis appear in several figures which are
+interspersed haphazardly with this text. Each figure presents the data
+in a different way, since this diversity in presentation should
+maximize the chance that one of the charts makes sense to you. The C
+programming language has two numeric types that could reasonably be
+used to represent $\epsilon$: {\tt float} and {\tt double}. The
+results for {\tt double} appear in Figure~\ref{fig:cdouble} and for
+{\tt float} in Figure~\ref{fig:cfloat}. \CC\ has those same two types,
+but I decided arbitrarily to only look at {\tt double}, which is in
+Figure~\ref{fig:cppdouble}. Programmers in C$\sharp$ are very
+creative; their results are presented in Figure~\ref{fig:csharp}.
+Objective~C proved unpopular for use of $\epsilon$, its sparse data
+are in Figure~\ref{fig:objectivec}. Finally, the ineffable JavaScript
+has its results in Figure~\ref{fig:javascript}.
+
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[width=0.99 \linewidth]{chart-c-double}
+\end{center}\vspace{-0.1in}
+\caption{ Values for {\tt const double epsilon} in the C programming
+  language. In this chart, the blue (lower) bars are the distinct values
+  of epsilon seen. The vertically aligned red (upper) bar is its count.
+  Epsilon values are plotted on a logarithmic scale, where the minimum
+  observed value $\log(-708)$ is $303 \times 10^{-308}$, and the largest
+  $\log(1.609)$ is $5$. 
+  %
+  Notes: One programmer used the value {\tt -1e10}, which is
+  -10,000,000,000, probably meaning {\tt 1e-10}. This value was excluded
+  because it has no real logarithm.
+}
+\label{fig:cdouble}
+\end{figure}
+
 \begin{figure}[ht]
@@
 \begin{center}
-% \includegraphics[width=0.99 \linewidth]{chart-c-double}
+\includegraphics[width=0.99 \linewidth]{chart-cpp}
 \end{center}\vspace{-0.1in}
@@
   (less than 1\% of the time).
+  %
   Notes:
-  Five times, the computed value is actually equal to 0.0,
+  Five times, the programmer used 0.0 for epsilon, which is possibly
+  the only unjustifiable value. The value {\tt pow(10,-13)} is
+  annotated in German ``Genauigkeitsziel bei der Nullstellensuche,''
+  or ``Accuracy goal in the search for zeros.'' }
+\label{fig:cppdouble}
+\end{figure}
 
-  The value {\tt pow(10,-13)} is annotated in German
-  ``Genauigkeitsziel bei der Nullstellensuche,'' or ``Accuracy goal in
-  the search for zeros.''
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[width=0.99 \linewidth]{chart-csharp}
+\end{center}\vspace{-0.1in}
+\caption{ Values for {\tt const double epsilon} in the C$\sharp$
+  programming language. In this chart, the left edge of greyscale
+  blocks on the left-facing flag represent the cumulative value of the
+  logarithm of all values of epsilon seen so far. The right edge of
+  the block with the exact same greyscale value on the bottom-right
+  flag is the cumulative count of all definitions of epsilon so far.
+  It is easy to see that there are 1,000 total definitions, as
+  expected.
+  %
+  Notes: One programmer defined epsilon as {\tt 1.0-18}, that is,
+  -17, probably meaning {\tt 1.0e-18}. Four used literal {\tt 0} for
+  epsilon. Another defined epsilon as {\tt 1 / 100000}, which uses
+  integer division and results again in 0. Owing to the strength and
+  creativity of C$\sharp$ developers, here we saw our largest
+  value of epsilon so far, {\tt 700}, and the declaration with the most
+  significant digits: {\tt
+    0.0000000000000002220446049250313080847263336181640625}.
 }
-\label{fig:cppdouble}
+\label{fig:csharp}
+\end{figure}
+
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[width=0.99 \linewidth]{chart-objectivec}
+\end{center}\vspace{-0.1in}
+\caption{Results for the language Objective~C for {\tt const double epsilon}.
+Perhaps because Objective~C programmers do not use this idiom for defining
+constants, or because it is only used to make iPhone games where epsilon is
+not a concept of interest, there were only 8 distinct values of epsilon
+observed. This allows us to present the data completely in the chart. The
+$x$ axis is $\sin(\epsilon)$ and the $y$ axis is $\cos(f)$ where $f$ is
+the total number of times that the given value occurred in the code.}
+\label{fig:objectivec}
+\end{figure}
+
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[width=0.70 \linewidth]{chart-js}
+\end{center}\vspace{-0.1in}
+\caption{Finally, JavaScript, the official language of the Internet.
+  In this ``radar'' plot, the data are arranged around a circle, as
+  tuples of $\langle \log(\epsilon), \epsilon, f \rangle$ where $f$ is
+  the frequency of the value being observed. Numbers in boxes ascend
+  along the $0\deg$ axis, displaying each of the multiples of 5
+  between 0 and 100, inclusive. Boxes stack so that only part of the
+  number is visible, but you know what's under there if you've looked
+  at numbers before. A spider-web-like network of interlacing lines
+  ascribe some additional meaning to some of the points on the clock
+  face. The smallest nonzero value observed was $10^{-32}$.
+%
+  Notes: JavaScript programmers have the highest tolerance for error
+  of all languages tested, with over 1,000 using epsilon of 0.5 or
+  higher. One programmer used $\epsilon = 1024$, and another
+  6,000,000! }
+\label{fig:javascript}
 \end{figure}

### 2014-03-01 epsilon/printlog.cc

@@ -281,11 +281,171 @@ NamedEntry entries_cppdouble[] = {
   { INIT(0.00001), 6 },
 };
 
+NamedEntry entries_csharp[] = {
+  { INIT(1.1102230246251565E-16), 1 },
+  { INIT(10e-6), 1 },
+  { INIT(1.401298E-45d), 1 },
+  { INIT(1e-100), 1 },
+  { INIT(10e-7), 1 },
+  { INIT(1E-8), 1 },
+  { INIT(700), 1 },
+  { INIT(0.00000000000001), 1 },
+  { INIT(0.000005), 1 },
+  { INIT(1E-3), 1 },
+  { INIT(1e-11), 1 },
+  { INIT(1E-14), 1 },
+  { INIT(1.0E-5), 1 },
+  { INIT(0.000001f), 1 },
+  { INIT(30), 1 },
+  { INIT(216.0 / 24389.0), 1 },
+  { INIT(2.7105054312137610850186320021749e-20), 1 },
+  { INIT(0.01d), 1 },
+  { INIT(1e-13), 1 },
+  { INIT(.00001), 1 },
+  { INIT(.000001d), 1 },
+  { INIT(0.1d), 1 },
+  { INIT(2.22044604925031E-15), 1 },
+  { INIT(1e-4), 1 },
+  { INIT(0.05), 1 },
+  { INIT(4.9406564584124650e-324d), 1 },
+  { INIT(5.0), 1 },
+  { INIT(.0000000001), 1 },
+  { INIT(0.0000001f), 1 },
+  { INIT(1 / 100000), 2 },
+  { INIT(1.0/1024.0), 2 },
+  { INIT(216. / 24389.), 2 },
+  { INIT(0.000000000001), 2 },
+  { INIT(1.0e-5), 2 },
+  { INIT(1E-5), 2 },
+  { INIT(1E-13), 2 },
+  { INIT(1e-7), 2 },
+  { INIT(0.000000000000000000001), 2 },
+  { INIT(1.0-18), 2 },
+  { INIT(3.141592653589 * 0.01), 2 },
+  { INIT(0.008856), 2 },
+  { INIT(1E-11), 2 },
+  { INIT(1e-14), 2 },
+  { INIT(.001), 2 },
+  { INIT(0.0001f), 2 },
+  { INIT(0.005), 2 },
+  { INIT(0), 2 },
+  { INIT(0.25), 2 },
+  { INIT(9.9999997473787516E-05), 2 },
+  { INIT(0.001d), 2 },
[…]

### 2014-03-08 chesstego/paper.tex
Some notes from long-running processes and TODO
Fix bug where king/queen were swapped (hehe), write a little about Chesstego, some more idea. n.b. this version can reproduce the winning solution for chesstego when black has no pieces but white is normal

@@
 
+\setchessboard{showmover=false}
+
+\newcommand\checkmate{\hspace{-.05em}\raisebox{.4ex}{\tiny\bf ++}}
+
 \renewcommand\th{\ensuremath{{}^{\textrm{th}}}}
@@
 \begin{abstract}
-what do I put here
+CHECKMATE.
 \end{abstract}
+
+\section*{Introduction}
+
+Chess is an old-timey game that you already know. One problem with
+Chess is that it is hard; both players may struggle mightily in a
+game, expending their brain-sugars, and it is not clear who the winner
+will be. Another problem with Chess is that it isn't other games, and
+we're pretty much over it. In this paper we attempt to address both
+problems, with limited success. We show how to combine Chess with
+several other board games, in order to make it more predictable.
+
+As usual for a Tom~7 SIGBOVIK joint, the results herein are real.
+The source code that was used to solve the games or prove that no
+winning strategy exists up to some depth can be found on the inter-net.\!\footnote{In the Subversion repository at:
+  \url{https://sourceforge.net/p/tom7misc/svn/HEAD/tree/trunk/chesstego/}}
+
+\section{Chesstego}
+
+{\em Chesstego} is a combination of the games Chess and Stratego. You
+already should know that every game in this paper is a combination of
+Chess and something, but I wanted to emphasize the portmanteau. All of
+the games will be named with portmanteau, and some of the names will
+be achingly bad.
+
+In Stratego, each player begins the game by arranging his or her
+Stratego-pieces in a secret fashion on the board. In the world of
+Stratego, civization is ruled by a leader known as Flag. The player's
+goal is to assassinate the opponent Flag, using a member of his or her
+army. However, which piece represents Flag is unknown!
+
+There are many ways we might apply the Stratego system of governance
+to Chess.
+
+\paragraph{Chesstego variation I.} In this variation, each player decides
+in secret, before the game begins, which of his or her pieces is the
+actual King, that is, the Flag. If this piece is checkmated, the game
+is lost. There are two subvariations: {\em I(a)}, where a player must
+announce {\em Check!} when the Flag is under attack, and all of the
+normal rules about moving into (or castling through) check must be
+obeyed. In subvariation {\em I(b)}, which I prefer, the piece that is
+Flag may slip silently into and out of check, and the game only ends
+when that piece is captured.\footnote{In Chess proper, these
+  formulations are nearly equivalent---except for rules like castling
+  through check and some corner cases---but it is deemed important for
+  movie drama that players be able to announce {\em Check!} and {\em
+    Checkmate!} at times.}
+
+{\em Chesstego variation I} is a reasonable if slightly silly game,
+and it is difficult. It may even be more difficult than Chess due to the
+psychological mind-games that are possible. It can be hard to tell which
+player is winning, let alone which player will win. In this paper we
+are interested in variants of Chess that both {\em are other games} and
+{\em are predictable}. We'd like to give one of the players a clear
+strategy for winning.
+
+\paragraph{Chesstego variation II.} In this variation, each player
+decides in secret, before the game, which of her {\em opponent's} pieces
+is Flag. Same as before, if this piece is captured, the game ends
+instantly. But now, players don't even know which of their own pieces
+is their glorious ruler, Flag. This can be very exciting or tittilating.
+
+Unfortunately, there is no known winning strategy for White, and there
+are no winning strategies with fewer than 6 moves. This was proved
+by computer program. In fact, it was proved for a stronger case:
+
+\paragraph{Chesstego variation III.} In this variation, the first player,
+known as White, chooses in secret both the identity of Flag for his own
+populace, as well as the identity of Flag for his opponent. Even in
+this very unfair setup, Black can always survive for at least 6 moves.
+
+% TODO: Example of black escaping.
+
+This will not do! Black just has so many options with all those pieces.
+Perhaps if he were handicapped somewhat?
+
+\paragraph{Chesstego variation IV.} In this variation, White chooses
+the identity of both Flags again, in secret without telling her
+opponent, and also Black does not get any good pieces, just pawns. Like this:
+
+\chessboard[
+  setfen=4k3/pppppppp/8/8/8/8/PPPPPPPP/RNBQKBNR]
+
+Lo! This version is finally satisfactory. White shall pick one of her
+safe pieces (e.g., one of the rooks) as Flag, and choose Black's b7~pawn
+as Black~Flag. White's winning move is c2c4 and then Qb3, with Black~Flag
+unable to escape the B file. Choosing the f7~pawn works as well.
+
+% make chesstego.exe && ./chesstego.exe -blackpieces justking 1 > king1.tex
+% \input{king1}
+
+Example hopeless match: 1. c4 Kf8~~2. Qb3 b5~~3. c4xb4\checkmate \\
+\chessboard[setfen=5k2/p1pppppp/8/1P6/8/1Q6/PP1PPPPP/RNB1KBNR]
+
+\paragraph{Chesstego variation V.} This variation is just like {\em IV}
+but White gets super good pieces everywhere instead of having some
+dumb ones, and Black still gets bupkiss:
+
+\chessboard[
+  setfen=4k3/pppppppp/8/8/8/8/NNNBBNNN/QQQQKQQQ]
+
+Now there are many more winning strategies, but the one from {\em IV}
+works as well.
+
+\paragraph{Other variations.} Incidentally, my on-line version of Chess 
+called SICO\footnote{ On the internet at:
+  \url{http://snoot.org/toys/sico/}}---where you just play a single
+move in a random game---has for about six years had variations called
+``center wall'' and ``barricades''\cite{sico} which were inspired by
+the layout of the Stratego board. However, these cannot be considered
+worthy of portmanteau, as the rules are basically just Chess rules.
+
+% Chessy Crush Saga
+
+\section{Cluess}
+
+This game, known as {\em Chuessdo} in the United Kingdom, is a cross
+between Chess and {\em Clue}. In this version, the player called White
+is known as Mrs.~White, and the player called Black is known as
+Professor~Plum.
 
@@
 
-\input{king2}
+% \input{king2}
 
@@
 
-% No mate in 14 for 1,2,3,4,5,7. didn't run 7 because 6 is
-% still running overnight! 13h51m! 0 also running for 42h00m.
+% Regular board still:
+% No mate in 14 for 1,2,3,4,5,7. didn't run 6 because 6-ply took
+% 2 full days!
+% 0 finally finished with no mates found.
 
@@
 
-% Now mate in 10 for white having ``mega'' pieces
+% No mate in 10 for white having ``mega'' pieces
 
@@
 
+% TODO: We always have to explore all of black's moves, but we
+% can push the search much deeper (if we just want to prove the
+% existence of mate) by prioritizing white's moves. e.g., on the
+% last move, only move to capture the king; and on the second-to-last,
+% only move into check.
+
 % TODO: Generate ``book'' where white is making strong moves
@@
 
-% \bibliography{paper}{}
-% \bibliographystyle{plain}
+\bibliography{paper}{}
+\bibliographystyle{plain}
+
 \end{document}

### 2014-03-10 chesstego/paper.tex
Sync for mtv trip

@@
 
-Lo! This version is finally satisfactory. White shall pick one of her
-safe pieces (e.g., one of the rooks) as Flag, and choose Black's b7~pawn
-as Black~Flag. White's winning move is c2c4 and then Qb3, with Black~Flag
+Lo! This version is finally satisfactory. In this version we need not
+equivocate over who shall win. White picks one of her very safe pieces
+(e.g., one of the rooks) as Flag, and chooses Black's b7~pawn as
+Black~Flag. White's winning move is c2c4 and then Qb3, with Black~Flag
 unable to escape the B file. Choosing the f7~pawn works as well.
@@
 
-% Chessy Crush Saga
-
 \section{Cluess}
@@
 Professor~Plum.
+
+\section{Chess Who?}
+
+The board game {\em Guess Who?} disappointed millions of children with
+its delightful television commercial that far outstripped the
+capabilities of the actual game. This required them to add a
+disclaimer to the commercial, ``Game cards do not actually talk.'' In
+{\em Guess Who?}, players take turns trying to guess the identity of
+the opponent, among a fixed set of personalities with traits like
+curly hair, straight hair, brown hair, short hair, bangs, long hair,
+blonde hair, red hair, or glasses. One player asks, ``Do you have
+curly hair?'' and the other says ``Yes'' or ``No'' and then the first
+player can eliminate all the remaining personalities that don't have
+the trait. It's basically binary search, but for kids.
+
+In {\em Chess Who?}, which is the chess version, players have two
+different options on each turn. They can either move a piece like
+normal, or ask a question of their opponent. The question must be
+about the physical characteristics of the pieces, for example, ``Does
+your piece have curly hair?'' No pieces have curly hair, so nothing
+happens. The opponent might ask a different question, on the next turn,
+like ``Does your piece have straight hair?'' No pieces have straight
+hair, so nothing happens. On the next turn, suppose the player asks,
+``Does your piece have brown hair?'' No pieces have brown hair,
+nor any hair at all, and no pieces have any brown parts at all either.
+Stop asking. Next, Black asks, ``Does your piece have crennelations?''
+Both of White's rooks have crennelations, so they are eliminated from
+the board (Figure~\ref{fig:crennelations}).
+
+\begin{figure}[ht]
+\begin{center}
+\chessboard[
+   setfen=rnbqkbnr/pppppppp/8/8/8/8/PPPPPPPP/1NBQKBN1]
+\end{center}\vspace{-0.1in}
+\caption{
+  The board after
+  1. Does your piece have curly hair?
+     \ldots Does your piece have straight hair?
+  2. Does your piece have brown hair?
+     \ldots Does your piece have crennelations?
+}
+\label{fig:crennelations}
+\end{figure}
+
+Next, White might ask, ``Does your piece have a cross on his hat?''
+This matches both Bishops and the King, but Black responds ``Check!''
+because it is illegal to ask a question that would eliminate the King.
+
+This variation produces draws easily. An example Nash Equilibrium:
+White begins by asking, ``Does your piece have three lobes like a
+truncated snowman, {\em or} crennelations, {\em or} horse ears, {\em
+  or} two weird long shoulder-ribbons that I don't know what they are,
+{\em or} five pointy tines on her crown?'' This matches all of Black's
+pieces except his king, so he removes them all. But then Black asks:
+``Does your piece {\em not} have a lumpy crown?'' which matches all of
+White's pieces except his king. So all we have left is the two kings,
+which is known to be a draw~\cite{pawnless}.
+
+% Chessy Crush Saga
+% Battleship + Chess = Battlechess
+% Chess Tac Toe
+% Chesslers of Catan
+% Chesstris
+% Hungry Hungry Hipposchess
+
+% Monopoly?
 

### 2014-03-16 chesstego/paper.tex
Chessy crush v2 (v1 looks nearly fair). Mainline support

@@
 
+\section{Chessy Crush Saga}
+
+% \begin{itemize} % white
+% \end{itemize} % white
+
+\begin{figure}[ht]
+\begin{center}
+\chessboard[setfen=4k3/1ppppppp/8/8/8/1p1PP3/P1P2PPP/4K3]
+\end{center}\vspace{-0.1in}
+\caption{
+  Black interfering with White's Omega Weapon main line:
+  b3, d3, e3, c3 {\em (making candy)}, Rc2, Re2++. Black
+  can progress pawns fast enough to capture White's. She
+  can also move her King, making the candied rook less
+  devastating. Ultimately, White's one-tempo advantage
+  is not obviously winning.
+}
+\label{fig:chessyv1}
+\end{figure}
+
+\chessboard[setfen=4k3/ppp1pppp/8/8/3p4/2PP4/PP2PPPP/4K3]
+
 % Chessy Crush Saga

### 2014-03-16 chesstego/paper.tex
Progress on paper

@@
 
-% \begin{itemize} % white
-% \end{itemize} % white
+{\em Chessy Crush Saga} is a hybrid of Chess and the popular and
+lucrative phone-game {\em Candy Crush Saga}. In this game, the
+regular moves of chess are allowed, but it is also possible to
+{\em crush} chess pieces into candies. {\em Chessy Crush v1}
+follows the rules of {\em Candy Crush} closely: When a piece moves
+such that three pieces of the same color are in a row (either
+horizontally or vertically), those pieces are all removed. Usually
+this is a tactical misstep in Chess, as three of one's own pieces
+are destroyed for nothing. However, as in {\em Candy Crush}, rewards
+are given when more than three like pieces are destroyed at once.
+The simplest is to award the player with a Candy Rook at at the
+place where the moved piece ended up. A Candy Rook moves like a
+rook, but if it is crushed, it destroys all of the opponent's
+pieces on the file (column) it currently occupies.\!\footnote{In
+formal {\em Candy Crush}, this is the vertical striped candy,
+which is awarded when the piece was moved vertically. To keep
+this variant simple, we always award a ``vertically striped''
+Candy Rook, since it seems with only pawns, only vertical moves
+are likely. (This later turns out to be a bad assumption.)}
+
+As another simplification, we remove all of the proper pieces from the
+black files, except for the King, in honor of the software company
+that makes {\em Candy Crush}, ``King.'' A densely packed board with
+pieces that can move backward is dangerous, for it is easy to destroy
+one's own king by moving out of and then back into the back file,
+crushing the whole thing (Figure~\ref{fig:selfcrush}).
+
+\begin{figure}[ht]
+\begin{center}
+\chessboard[
+  setfen=rnbqkbnr/1ppppppp/p7/8/8/8/PPPPPPPP/RNBQKBNR,
+  color=black,
+  pgfstyle=knightmove,
+  markmoves={c3-b1},
+  pgfstyle=cross,
+  color=red,
+  backrank=1]
+\end{center}\vspace{-0.1in}
+\caption{
+  ``Fool's Mate'' in the rejected {\em Candy Crush v0},
+  played with normal pieces in the back ranks. White self-mates
+  in two moves, crushing all his own pieces {\em e.g.} with:
+  1. Nc3 a6 2. Nb1 (??) 0--1
+}
+\label{fig:selfcrush}
+\end{figure}
 
@@
 
-\chessboard[setfen=4k3/ppp1pppp/8/8/3p4/2PP4/PP2PPPP/4K3]
+In {\em v2}, the Candy Rook is extremely powerful, since
+not only does it destroy an entire file (the e file being
+the likely target, since it contains the immovable King),
+but when it is used (by crushing) it is also replaced,
+since a crush always yields candy. Therefore, White's
+most straightforward strategy is to quickly create a
+Candy Rook, then crush it in the e file, which wins. This
+should be a line where White's tempo advantage makes it
+mostly immune from interference, since even a check is
+answered by destroying the opponent's king.
+
+% \chessboard[setfen=4k3/ppp1pppp/8/8/3p4/2PP4/PP2PPPP/4K3]
+% \chessboard[setfen=4k3/ppp1pppp/8/8/8/4p3/PP3PPP/4K3]
+% \chessboard[setfen=4k3/1ppppppp/8/8/8/8/pR2PPPP/4K3]
+\begin{figure}[ht]
+\begin{center}
+\chessboard[setfen=4k3/3ppppp/1r6/8/8/2R5/PP3PPP/4K3]
+(a) \\
+\chessboard[
+  setfen=4k3/3ppppp/4r3/8/8/8/2R2PPP/4K3,
+  pgfstyle=border,
+  color=blue!50,
+  markfile=e,
+  color=black,
+  pgfstyle=straightmove,
+  markmoves={b6-e6},
+  pgfstyle=cross,
+  color=red,
+  backfields={e8,e7,e6}]
+(b)
+\end{center}\vspace{-0.1in}
+\caption{ 
+%
+  Black's surprising draw strategy in {\em Chessy Crush v2}. White
+  begins her Omega Weapon: 1. e3 c6 2. d3 a6 3. c3 {\em making candy}
+  b6 {\em making candy} we have the board in (a). It seems that Black
+  is stuck after 4. Rc2, since 5. Re2 will crush White's candy rook
+  and destroy the entire e file, including Black's king. Black
+  defending with 4. \ldots Re6 initially seems pointless (White will
+  destroy the rook as well so the check is irrelevant---this indeed
+  refutes 4. \ldots Rb1) but Re6 actually creates a {\em vertical}
+  three-piece crush, sacrificing Black's king to destroy the e file
+  first, which also destroys White's king! Such mutual destruction
+  is not possible in traditional chess, but it seems appropriate to
+  consider it a draw in Chessy Crush Saga v2. It appears that White
+  can head off this line with 4. Rc8++, a totally vanilla back-rank mate,
+  but this is not actually mate with the same 4. \ldots Rb1 escaping
+  to a draw.
+%
+}
+\label{fig:chessyv2}
+\end{figure}
 

### 2014-03-17 chesstego/paper.tex
Submitted version, delete unused chesswho

@@
 {\em IEEEEEE!} press, Verlag-Verlag volume no.~0x40-2A.
-\$0.00}}
+\$0.00}. Badness: 12,844 }
 
@@
 Stratego-pieces in a secret fashion on the board. In the world of
-Stratego, civization is ruled by a leader known as Flag. The player's
-goal is to assassinate the opponent Flag, using a member of his or her
-army. However, which piece represents Flag is unknown!
+Stratego, civilization is ruled by a leader known as Flag. The
+player's goal is to assassinate the opponent Flag, using a member of
+his or her army. However, which piece represents Flag is unknown!
 
@@
 in secret, before the game begins, which of his or her pieces is the
-actual King, that is, the Flag. If this piece is checkmated, the game
+actual King, that is, Flag. If this piece is checkmated, the game
 is lost. There are two subvariations: {\em I(a)}, where a player must
-announce {\em Check!} when the Flag is under attack, and all of the
+announce {\em Check!} when Flag is under attack, and all of the
 normal rules about moving into (or castling through) check must be
@@
 Flag may slip silently into and out of check, and the game only ends
-when that piece is captured.\footnote{In Chess proper, these
+when that piece is captured.\!\footnote{In Chess proper, these
   formulations are nearly equivalent---except for rules like castling
@@
 instantly. But now, players don't even know which of their own pieces
-is their glorious ruler, Flag. This can be very exciting or tittilating.
+is their glorious ruler, Flag. This can be very exciting or titillating.
 
@@
 
+\begin{center}
 \chessboard[
+  smallboard,
   setfen=4k3/pppppppp/8/8/8/8/PPPPPPPP/RNBQKBNR]
+\end{center}
 
@@
 Black~Flag. White's winning move is c2c4 and then Qb3, with Black~Flag
-unable to escape the B file. Choosing the f7~pawn works as well.
+unable to escape the B file~(Figure~\ref{fig:hopeless}). Choosing the f7~pawn works as well.
 
@@
 
-Example hopeless match: 1. c4 Kf8~~2. Qb3 b5~~3. c4xb4\checkmate \\
-\chessboard[setfen=5k2/p1pppppp/8/1P6/8/1Q6/PP1PPPPP/RNB1KBNR]
+\begin{figure}[ht]
+\begin{center}
+\chessboard[setfen=5k2/p1pppppp/8/1P6/8/1Q6/PP1PPPPP/RNB1KBNR,
+  pgfstyle=straightmove,
+  markmoves={c4-b5}
+]
+\end{center}\vspace{-0.1in}
+\caption{
+Example hopeless match: 1. c4 Kf8~~2. Qb3 b5~~3. xb5 1--0
+}
+\label{fig:hopeless}
+\end{figure}
 
@@
 
+\begin{center}
 \chessboard[
+  smallboard,
   setfen=4k3/pppppppp/8/8/8/8/NNNBBNNN/QQQQKQQQ]
+\end{center}
 
@@
 nor any hair at all, and no pieces have any brown parts at all either.
-Stop asking. Next, Black asks, ``Does your piece have crennelations?''
-Both of White's rooks have crennelations, so they are eliminated from
-the board (Figure~\ref{fig:crennelations}).
+Stop asking. Next, Black asks, ``Does your piece have crenellations?''
+Both of White's rooks have crenellations, so they are eliminated from
+the board (Figure~\ref{fig:crenellations}).
 
@@
   2. Does your piece have brown hair?
-     \ldots Does your piece have crennelations?
+     \ldots Does your piece have crenellations?
 }
-\label{fig:crennelations}
+\label{fig:crenellations}
 \end{figure}
@@
 White begins by asking, ``Does your piece have three lobes like a
-truncated snowman, {\em or} crennelations, {\em or} horse ears, {\em
+truncated snowman, {\em or} crenellations, {\em or} horse ears, {\em
   or} two weird long shoulder-ribbons that I don't know what they are,
@@
 are given when more than three like pieces are destroyed at once.
-The simplest is to award the player with a Candy Rook at at the
+The simplest is to award the player with a Candy Rook at the
 place where the moved piece ended up. A Candy Rook moves like a
@@
 
+With just pawns, it seems that the best way to gain advantage for
+White is to crush four pawns to create a Candy Rook before Black does,
+then menace Black with it. Unfortunately, moving four pawns into
+formation before black can interfere is not possible; Black can react
+to White's first move and advance one of his own pawns towards the
+construction (Figure~\ref{fig:chessyv1}). Moreover, if he does, and
+trades for one of White's pawns (or better, the resulting candy), then
+Black is probably in a superior position, White having crapified his
+pawn structure in pursuit of candy. Worse still, crushing a successful
+Candy Rook is not that menacing, since Black can simply move his king
+out of the way.
+
 \begin{figure}[ht]
 \begin{center}
-\chessboard[setfen=4k3/1ppppppp/8/8/8/1p1PP3/P1P2PPP/4K3]
+\chessboard[setfen=4k3/1ppppppp/8/8/8/1p1PP3/P1P2PPP/4K3,
+  pgfstyle=straightmove,
+  markmoves={a4-b3},
+]
 \end{center}\vspace{-0.1in}
@@
 
-In {\em v2}, the Candy Rook is extremely powerful, since
-not only does it destroy an entire file (the e file being
-the likely target, since it contains the immovable King),
-but when it is used (by crushing) it is also replaced,
-since a crush always yields candy. Therefore, White's
-most straightforward strategy is to quickly create a
-Candy Rook, then crush it in the e file, which wins. This
-should be a line where White's tempo advantage makes it
-mostly immune from interference, since even a check is
-answered by destroying the opponent's king.
+To correct these problems with balance ({\em i.e.}, that the game may
+be balanced), we simplify the candying rules: Any crush produces a
+Candy Rook, including ones of length 3. This allows White to create a
+Candy Rook before Blacks pawns can reach her. Furthermore, we
+stipulate that the kings cannot move (they don't have legs anyway,
+this is plain to see, so let's be realistic). In {\em v2}, the Candy
+Rook is extremely powerful, since not only does it destroy an entire
+file (the e file being the likely target, since it contains the
+immovable King), but when it is used (by crushing) it is also
+replaced, since a crush always yields candy. Therefore, White's most
+straightforward strategy is to quickly create a Candy Rook, then crush
+it in the e file, which wins. This should be a line where White's
+tempo advantage makes it mostly immune from interference, since even a
+check is answered by destroying the opponent's king.
 
-% \chessboard[setfen=4k3/ppp1pppp/8/8/3p4/2PP4/PP2PPPP/4K3]
+The idea behind White's Omega Weapon is to create a triplet of pawns
+on rank 3, either c3--e3 or e3--g3. White begins by moving her king's
+pawn to e3. Black can interfere with one side of the triplet
+(Figure~\ref{fig:twopaths}), but only one. After 1. e3, if \ldots f5
+or h5, then White will continue creating the triplet in c3--e3.
+Otherwise, she is safe to create it in e3--g3. Once the Omega Weapon
+is constructed, she need only move it to e2 (which is free due to the
+first move, and will be adjacent to at least two pawns). This destroys
+the Black king, which cannot have moved, due to the rules.
+Unfortunately, this strategy can be foiled by Black with a
+clever\footnote{This strategy was discovered by computer, like most
+  clever things these days.} mutual suicide, as shown in
+Figure~\ref{fig:chessyv2}. This strategy is not easily defended
+against, nor is it straightforward to fix the rules of the
+game,\!\footnote{ It seems perhaps we can eliminate Black's draw
+  strategy by restoring the idea of horizontally striped Candy Rooks
+  (see footnote above). I implemented this. Unfortunately, although
+  the move that ends the game in a draw is a horizontal move, the move
+  that creates the Candy Rook is a vertical one. Therefore, Black's
+  Candy Rook is vertically striped, and can successfully destroy the
+  White king (though it is then replaced with a horizontally striped
+  Candy Rook to observe the empty thrones).} so it seems {\em Chessy
+  Crush Saga v2} also retains hope for Black, even though the odds
+initially seemed stacked against her.
+
+\begin{figure}[ht]
+\begin{center}
+\chessboard[setfen=4k3/ppp1pppp/8/8/3p4/2PP4/PP2PPPP/4K3]
+\end{center}\vspace{-0.1in}
+\caption{ 
+%
+  White may not use a fixed strategy to construct her Omega Weapon.
+  The position shown is after 1. c3 d5 2. d3 d4. If 3. e3 {\em (making
+    candy)} then \ldots xe3, obviously. White can make candy with 3.
+  b3, but now e2 is not clear to crush the Candy Rook here and destroy
+  the Black king.
+%
+}
+\label{fig:twopaths}
+\end{figure}
+
 % \chessboard[setfen=4k3/ppp1pppp/8/8/8/4p3/PP3PPP/4K3]
@@
   Black's surprising draw strategy in {\em Chessy Crush v2}. White
-  begins her Omega Weapon: 1. e3 c6 2. d3 a6 3. c3 {\em making candy}
-  b6 {\em making candy} we have the board in (a). It seems that Black
-  is stuck after 4. Rc2, since 5. Re2 will crush White's candy rook
-  and destroy the entire e file, including Black's king. Black
-  defending with 4. \ldots Re6 initially seems pointless (White will
-  destroy the rook as well so the check is irrelevant---this indeed
-  refutes 4. \ldots Rb1) but Re6 actually creates a {\em vertical}
-  three-piece crush, sacrificing Black's king to destroy the e file
-  first, which also destroys White's king! Such mutual destruction
-  is not possible in traditional chess, but it seems appropriate to
-  consider it a draw in Chessy Crush Saga v2. It appears that White
-  can head off this line with 4. Rc8++, a totally vanilla back-rank mate,
-  but this is not actually mate with the same 4. \ldots Rb1 escaping
-  to a draw.
+  begins her Omega Weapon: 1. e3 c6 2. d3 a6 3. c3 {\em (making
+    candy)} b6 {\em (making candy)} we have the board in (a). It seems
+  that Black is stuck after 4. Rc2 {\em (making candy)}, since 5. Re2
+  {\em (making candy)} will crush White's candy rook and destroy the
+  entire e file, including Black's king. Black defending with 4.
+  \ldots Re6 initially seems pointless (White will destroy the rook as
+  well so the check is irrelevant---this indeed refutes 4. \ldots Rb1)
+  but Re6 actually creates a {\em vertical} three-piece crush,
+  sacrificing Black's king to destroy the e file first, which also
+  destroys White's king! Such mutual destruction is not possible in
+  traditional chess, but it seems appropriate to consider it a draw in
+  Chessy Crush Saga v2. It appears that White can head off this line
+  with 4. Rc8++, a totally vanilla back-rank mate, but this is not
+  actually mate with the same 4. \ldots Re6 {\em (making candy)}
+  escaping to a draw.
 %
@@
 
-% Chessy Crush Saga
-% Battleship + Chess = Battlechess
-% Chess Tac Toe
-% Chesslers of Catan
-% Chesstris
-% Hungry Hungry Hipposchess
+\section{Future work and conclusions}
 
-% Monopoly?
+Many other games can be combined with Chess to ruin Chess. For example,
+consider {\em Chess Tac Toe}, {\em Chesslers of Catan}, {\em Chesstris},
+and {\em Hungry Hungry Hipposchess}. Combining Chess and Sokoban is
+impossible, obviously. The combination of Chess and Battleship is
+{\em Battlechess}, published in 1988 by Interplay.
 
-\setchessboard{tinyboard,showmover=false}
+\medskip
+\medskip
+In conclusion, San Dimas High School Footchess rules!
+
+% \setchessboard{tinyboard,showmover=false}
 % \chessboard[
@@
 
+% TODO: Use mainline to see if any more chesstego is solvable.
+%
+
 % TODO: We always have to explore all of black's moves, but we

### 2015-03-14 redi/redi.cc
Safe SDL exit, text network and layer printing

@@ -50,6 +50,12 @@ static_assert((NPP >= 3), "Must have at least R, G, B pixels.");
 
 // static_assert(RANDOM == 0, "random not yet supported.");
 
+std::mutex print_mutex;
+#define Printf(fmt, ...) do {		\
+  MutexLock Printf_ml(&print_mutex);		\
+  printf(fmt, ##__VA_ARGS__);			\
+  } while (0);
+
 static SDL_Surface *screen = nullptr;
 #define SCREENW 1920
 #define SCREENH 1080
@@ -98,9 +104,15 @@ struct ImageRGBA {
 
 // Using standard region of -1 to 1.
 static uint8 FloatByte(float f) {
-  if (f < -1.0f) return 0;
-  else if (f > 1.0f) return 255;
-  else return (uint8)((f + 1.0f) * 127.5);
+  if (f <= -1.0f) return 0;
+  else if (f >= 1.0f) return 255;
+  else {
+    // Now we know f is in [-1, 1].
+    float zero_to_two = f + 1.0f;
+    float zero_to_one = zero_to_two * 0.5f;
+    return (uint8)(zero_to_one * 255.0);
+    // return (uint8)((f + 1.0f) * 127.5);
+  }
 }
 
 static float ByteFloat(uint8 b) {
@@ -118,7 +130,20 @@ static void WriteLayerAsImage(const string &filename, const vector<float> &value
   CHECK_EQ(rgba.size(), SIZE * SIZE * 4);
   ImageRGBA img(rgba, SIZE, SIZE);
   img.Save(filename);
-  printf("Wrote %s...\n", filename.c_str());
+  Printf("Wrote %s...\n", filename.c_str());
+ 
+  // XXX PERF and as text file.
+  {
+    string tf = (string)"text-" + filename + (string)".txt";
+    FILE *ftxt = fopen(tf.c_str(), "wb");
+    for (int i = 0; i < values.size(); i++) {
+      if (i % 8 == 0) fprintf(ftxt, "\n");
+      fprintf(ftxt, "% f ", values[i]);
+    }
+    fclose(ftxt);
+    Printf("And %s.\n", tf.c_str());
+  }
+
 }
 
 // Single-channel bitmap.
@@ -264,6 +289,26 @@ struct Network {
   // a permutation of 0..(NUM_NODES * INDICES_PER_NODE - 1).
   vector<vector<uint32>> inverted_indices;
 };
+
[…]

### 2015-03-14 trunk/redi/constants.h
Safe SDL exit, text network and layer printing

@@ -18,6 +18,10 @@
 // This size could later be layer-specific, btw.
 //
 // Does NOT include bias term.
-#define INDICES_PER_NODE 1024
+
+// #define INDICES_PER_NODE 1024
+// #define NEIGHBORHOOD 5
+#define INDICES_PER_NODE 4
+#define NEIGHBORHOOD 0
 
 #define NUM_NODES (SIZE * SIZE * NPP)

### 2015-03-14 trunk/redi/threadutil.h

@@ -146,5 +146,19 @@ struct MutexLock {
   std::mutex *m;
 };
 
+// Read with the mutex that protects it. T must be copyable,
+// obviously!
+template<class T>
+T ReadWithLock(std::mutex *m, const T *t) {
+  MutexLock ml(m);
+  return *t;
+}
+
+// Write with the mutex that protects it. T must be copyable.
+template<class T>
+void WriteWithLock(std::mutex *m, T *t, const T &val) {
+  MutexLock ml(m);
+  *t = val;
+}
 
 #endif

### 2015-03-14 redi/constants.h
Make backwards pass a kernel too

@@ -1,6 +1,6 @@
 // Nodes per pixel. Beyond the R, G, B values, the inputs will be zero and
 // the outputs ignored.
-#define NPP 3
+#define NPP 4
 
 // Size of of image in pixels (width/height). Note that there are
 // three nodes per pixel (RGB).
@@ -21,7 +21,7 @@
 
 // #define INDICES_PER_NODE 1024
 // #define NEIGHBORHOOD 5
-#define INDICES_PER_NODE 4
-#define NEIGHBORHOOD 0
+#define INDICES_PER_NODE 64
+#define NEIGHBORHOOD 1
 
 #define NUM_NODES (SIZE * SIZE * NPP)

### 2015-03-14 redi/redi.cc

@@ -1,6 +1,8 @@
 #include "../cc-lib/sdl/sdlutil.h"
 #include "SDL.h"
 #include "SDL_main.h"
+#include "../cc-lib/sdl/chars.h"
+#include "../cc-lib/sdl/font.h"
 
 #include <CL/cl.h>
 #include <string.h>
@@ -41,8 +43,29 @@ using uchar = uint8_t;
 using uint32 = uint32_t;
 using uint64 = uint64_t;
 
+#if 1
+#define ECHECK(a)
+#define ECHECK_EQ(a, b)
+#define ECHECK_LT(a, b)
+#define ECHECK_GT(a, b)
+#define ECHECK_LE(a, b)
+#define ECHECK_GE(a, b)
+#else
+#error implement these
+#endif
+
+// Graphics.
+#define FONTWIDTH 9
+#define FONTHEIGHT 16
+static Font *font = nullptr;
+#define SCREENW 1920
+#define SCREENH 1080
+static SDL_Surface *screen = nullptr;
+
 // You can read the previous versions (redi-random.cc) to see some of
 // the history / thoughts.
+
+static constexpr int NUM_LAYERS = 2;
 
 static_assert((NPP >= 3), "Must have at least R, G, B pixels.");
 // static_assert((NEIGHBORHOOD & 1) == 1, "neighborhood must be odd.");
@@ -55,10 +78,6 @@ std::mutex print_mutex;
   MutexLock Printf_ml(&print_mutex);		\
   printf(fmt, ##__VA_ARGS__);			\
   } while (0);
-
-static SDL_Surface *screen = nullptr;
-#define SCREENW 1920
-#define SCREENH 1080
 
 template<class C>
 static void DeleteElements(C *cont) {
@@ -102,30 +121,25 @@ struct ImageRGBA {
   vector<uint8> rgba;
 };
 
-// Using standard region of -1 to 1.
 static uint8 FloatByte(float f) {
-  if (f <= -1.0f) return 0;
-  else if (f >= 1.0f) return 255;
-  else {
-    // Now we know f is in [-1, 1].
-    float zero_to_two = f + 1.0f;
-    float zero_to_one = zero_to_two * 0.5f;
-    return (uint8)(zero_to_one * 255.0);
-    // return (uint8)((f + 1.0f) * 127.5);
-  }
+  if (f <= 0.0f) return 0;
+  if (f >= 1.0f) return 255;
+  else return f * 255.0;
 }
 
[…]

### 2015-03-14 redi/redi.cc
make updateweights a kernel but I did it wrong

@@ -847,6 +847,79 @@ struct BackwardLayerCL {
   std::mutex m;
 };
 
+struct UpdateWeightsCL {
+  explicit UpdateWeightsCL(CL *cl) : cl(cl) {
+    const string kernel_src = 
+      Util::ReadFile("constants.h") + "\n" +
+      Util::ReadFile("updateweights.cl");
+
+    std::tie(program, kernel) = cl->BuildOneKernel(kernel_src, "UpdateWeights");
+  }
+
+  void Update(float learning_rate,
+	      const Stimulation &stim, const Errors &err, Network *net, int layer) {
+    MutexLock ml(&m);
+
+    cl_mem layer_error = MoveMemoryToGPUConst(cl->context, cl->queue,
+					      err.error[layer]);
+    cl_mem layer_indices = MoveMemoryToGPUConst(cl->context, cl->queue,
+						net->indices[layer]);
+    cl_mem layer_values = MoveMemoryToGPUConst(cl->context, cl->queue,
+					       stim.values[layer]);
+    cl_mem layer_weights = MoveMemoryToGPU(cl->context, cl->queue,
+					   false, &net->weights[layer]);
+    cl_mem layer_biases = MoveMemoryToGPU(cl->context, cl->queue,
+					  false, &net->biases[layer]);
+
+    CHECK_SUCCESS(clSetKernelArg(kernel, 0, sizeof(cl_float), (void *)&learning_rate));
+    CHECK_SUCCESS(clSetKernelArg(kernel, 1, sizeof(cl_mem), (void *)&layer_error));
+    CHECK_SUCCESS(clSetKernelArg(kernel, 2, sizeof(cl_mem), (void *)&layer_indices));
+    CHECK_SUCCESS(clSetKernelArg(kernel, 3, sizeof(cl_mem), (void *)&layer_values));
+    CHECK_SUCCESS(clSetKernelArg(kernel, 4, sizeof(cl_mem), (void *)&layer_weights));
+    CHECK_SUCCESS(clSetKernelArg(kernel, 5, sizeof(cl_mem), (void *)&layer_biases));
+
+    size_t global_work_offset[] = { 0 };
+    size_t global_work_size[] = { (size_t)NUM_NODES };
+    CHECK(CL_SUCCESS == clEnqueueNDRangeKernel(cl->queue, kernel,
+					       // work dimensions
+					       1, 
+					       // global work offset
+					       global_work_offset,
+					       // global work size
+					       global_work_size, 
+					       // local work size
+					       nullptr, 
[…]

### 2015-03-14 trunk/redi/redi.cc
Sweet, round takes 2.7s now after doing updateweights a smarter way

@@ -856,56 +856,73 @@ struct UpdateWeightsCL {
     std::tie(program, kernel) = cl->BuildOneKernel(kernel_src, "UpdateWeights");
   }
 
-  void Update(float learning_rate,
-	      const Stimulation &stim, const Errors &err, Network *net, int layer) {
-    MutexLock ml(&m);
+  struct UpdateContext {
+    UpdateContext(UpdateWeightsCL *parent, Network *net, int layer) :
+      parent(parent), net(net), layer(layer) {
+      CL *cl = parent->cl;
 
-    cl_mem layer_error = MoveMemoryToGPUConst(cl->context, cl->queue,
-					      err.error[layer]);
-    cl_mem layer_indices = MoveMemoryToGPUConst(cl->context, cl->queue,
-						net->indices[layer]);
-    cl_mem layer_values = MoveMemoryToGPUConst(cl->context, cl->queue,
-					       stim.values[layer]);
-    cl_mem layer_weights = MoveMemoryToGPU(cl->context, cl->queue,
-					   false, &net->weights[layer]);
-    cl_mem layer_biases = MoveMemoryToGPU(cl->context, cl->queue,
-					  false, &net->biases[layer]);
+      layer_indices = MoveMemoryToGPUConst(cl->context, cl->queue, net->indices[layer]);
+      layer_weights = MoveMemoryToGPU(cl->context, cl->queue, false, &net->weights[layer]);
+      layer_biases = MoveMemoryToGPU(cl->context, cl->queue, false, &net->biases[layer]);
+    }
 
-    CHECK_SUCCESS(clSetKernelArg(kernel, 0, sizeof(cl_float), (void *)&learning_rate));
-    CHECK_SUCCESS(clSetKernelArg(kernel, 1, sizeof(cl_mem), (void *)&layer_error));
-    CHECK_SUCCESS(clSetKernelArg(kernel, 2, sizeof(cl_mem), (void *)&layer_indices));
-    CHECK_SUCCESS(clSetKernelArg(kernel, 3, sizeof(cl_mem), (void *)&layer_values));
-    CHECK_SUCCESS(clSetKernelArg(kernel, 4, sizeof(cl_mem), (void *)&layer_weights));
-    CHECK_SUCCESS(clSetKernelArg(kernel, 5, sizeof(cl_mem), (void *)&layer_biases));
+    void Update(float learning_rate, const Stimulation &stim, const Errors &err, int layer) {
+      CL *cl = parent->cl;
 
-    size_t global_work_offset[] = { 0 };
[…]

### 2015-03-15 trunk/redi/make-corpus.sh
Saving and loading of network data

@@ -76,6 +76,16 @@ make fetcher.exe
 # ./fetcher.exe pyramids
 # ./fetcher.exe iceberg
 
-./fetcher.exe capitol &
-./fetcher.exe "jack o lantern" &
-./fetcher.exe "tennis racket"
+# ./fetcher.exe capitol &
+# ./fetcher.exe "jack o lantern" &
+# ./fetcher.exe "tennis racket"
+
+# ./fetcher.exe beer &
+# ./fetcher.exe bumblebee &
+# ./fetcher.exe grapefruit &
+# ./fetcher.exe park &
+# ./fetcher.exe architecture
+
+./fetcher.exe waterslide &
+./fetcher.exe tarantula &
+./fetcher.exe "ferris wheel"

### 2015-03-15 trunk/redi/resample.cc

@@ -163,7 +163,8 @@ struct ResamplerCL {
 };
 
 int main(int argc, char* argv[])  {
-  Timer setup_timer;
+  ArcFour rc(__func__);
+  rc.Discard(2000);
 
   // Reports 1, currently, ugh.
   printf("[CPU] Has %d hardware concurrency\n", std::thread::hardware_concurrency());
@@ -251,10 +252,15 @@ int main(int argc, char* argv[])  {
 
   {
     std::mutex m, output;
+    vector<uint8> seeds;
+    seeds.reserve(images.size());
+    for (int i = 0; i < images.size(); i++) seeds.push_back(rc.Byte());
+
     int num = (int)images.size();
     int done = 0;
-    std::function<void(const vector<uint8> &v)> ResampleSave =
-      [&resampler1, &resampler2, &m, &output, num, &done](const vector<uint8> &v) {
+    auto ResampleSave =
+      [&resampler1, &resampler2, &m, &output, num, &done, &seeds](int i, const vector<uint8> &v) {
+      const uint8 seed = seeds[i];
       vector<uchar> newimg = resampler1.Resample(v);
       vector<uchar> newimg2 = resampler2.Resample(newimg);
       
@@ -265,7 +271,12 @@ int main(int argc, char* argv[])  {
 	done++;
       }
 
-      SaveImage(StringPrintf("corpus256/img%d.png", local_done), newimg2, 256);
+      // About 4% of corpus goes to held-out set.
+      if (seed < 11) {
+	SaveImage(StringPrintf("eval256/img%d.png", local_done), newimg2, 256);
+      } else {
+	SaveImage(StringPrintf("corpus256/img%d.png", local_done), newimg2, 256);
+      }
 
       if (local_done % 10 == 0) {
 	MutexLock ml(&output);
@@ -274,7 +285,7 @@ int main(int argc, char* argv[])  {
       }
     };
 
-    ParallelApp(images, ResampleSave, 8);
+    ParallelAppi(images, ResampleSave, 12);
   }
 
   CHECK_SUCCESS(clReleaseCommandQueue(command_queue));

### 2015-03-15 trunk/redi/threadutil.h

@@ -21,9 +21,9 @@
 // data structures. Return value of function is ignored.
 //
 // TODO: Implement this in tems of ParallelComp
-template<class T>
+template<class T, class F>
 void ParallelAppi(const vector<T> &vec, 
-		  std::function<void(int i, const T&)> &f,
+		  const F &f,
 		  int max_concurrency) {
   // TODO: XXX This cast may really be unsafe, since these vectors
   // could exceed 32 bit ints in practice.

### 2015-03-15 redi/redi.cc
Saving and loading of network data

@@ -59,13 +59,13 @@ using uint64 = uint64_t;
 #define FONTHEIGHT 16
 static Font *font = nullptr;
 #define SCREENW 1920
-#define SCREENH 1080
+#define SCREENH 1280
 static SDL_Surface *screen = nullptr;
 
 // You can read the previous versions (redi-random.cc) to see some of
 // the history / thoughts.
 
-static constexpr int NUM_LAYERS = 2;
+static constexpr int NUM_LAYERS = 3;
 
 static_assert((NPP >= 3), "Must have at least R, G, B pixels.");
 // static_assert((NEIGHBORHOOD & 1) == 1, "neighborhood must be odd.");
@@ -171,7 +171,7 @@ struct ImageA {
 };
 
 // Pretty much only useful for this application since it loads the whole font each time.
-ImageA *FontChar(const string &filename, char c, int size) {
+static ImageA *FontChar(const string &filename, char c, int size) {
   fflush(stdout);
 
   FILE *file = fopen(filename.c_str(), "rb");
@@ -202,9 +202,9 @@ ImageA *FontChar(const string &filename, char c, int size) {
   return new ImageA(ret, width, height);
 }
 
-void BlitChannel(uint8 r, uint8 g, uint8 b, const ImageA &channel, 
-		 int xpos, int ypos,
-		 ImageRGBA *dest) {
+static void BlitChannel(uint8 r, uint8 g, uint8 b, const ImageA &channel, 
+			int xpos, int ypos,
+			ImageRGBA *dest) {
   for (int y = 0; y < channel.height; y++) {
     int dsty = ypos + y;
     if (dsty >= 0 && dsty < dest->height) {
@@ -226,12 +226,32 @@ void BlitChannel(uint8 r, uint8 g, uint8 b, const ImageA &channel,
 	  Blend(didx + 1, g, ch);
 	  Blend(didx + 2, b, ch);
 	  dest->rgba[didx + 3] = 0xFF;
-	  /*
-	  dest->rgba[didx + 0] = r;
-	  dest->rgba[didx + 1] = g;
-	  dest->rgba[didx + 2] = b;
-	  dest->rgba[didx + 3] = ch;
-	  */
+	}
+      }
+    }
+  }
+}
+
+// This doesn't blend; it assumes we have 0 in every slot to start.
+// (This is only used to draw into the unused image channel in the
+// training data to try to coax it to recognize the position of the 'i').
[…]

### 2015-03-15 portmantout/timer.h
performance timer for windows

@@ -1,0 +1,30 @@
+#include <windows.h>
+
+// Computes time elapsed in ms. Windows only.
+struct Timer {
+  Timer() {
+    QueryPerformanceFrequency(&freq);
+    Start();
+  }
+
+  void Start() {
+    QueryPerformanceCounter(&starttime);
+  }
+
+  // This looks buggy (MS also "stops"); don't call it!
+  // void Stop() {
+  // QueryPerformanceCounter(&stoptime);
+  // }
+
+  double MS() {
+    QueryPerformanceCounter(&stoptime);
+    // Stop();
+    double elapsed_ms = (stoptime.QuadPart - starttime.QuadPart) *
+      1000.0 / freq.QuadPart;
+    return elapsed_ms;
+  }
+
+ private:
+  LARGE_INTEGER freq;
+  LARGE_INTEGER starttime, stoptime;
+};

### 2015-03-15 trunk/portmantout/threadutil.h
Threads for portmantout

@@ -1,0 +1,164 @@
+
+#ifndef __THREADUTIL_H
+#define __THREADUTIL_H
+
+#include <thread>
+#include <mutex>
+
+#include "mingw.thread.h"
+#include "mingw.mutex.h"
+// ugh, conflict...
+#undef ARRAYSIZE
+
+// Do progress meter.
+// It should be thread safe and have a way for a thread to register a sub-meter.
+
+// TODO: Make these functions take a "callable" template param F, rather
+// than requiring std::function.
+
+// Run the function f on each element of vec in parallel, with its
+// index. The caller must of course synchronize any accesses to shared
+// data structures. Return value of function is ignored.
+//
+// TODO: Implement this in tems of ParallelComp
+template<class T, class F>
+void ParallelAppi(const vector<T> &vec, 
+		  const F &f,
+		  int max_concurrency) {
+  // TODO: XXX This cast may really be unsafe, since these vectors
+  // could exceed 32 bit ints in practice.
+  max_concurrency = min((int)vec.size(), max_concurrency);
+  std::mutex index_m;
+  int next_index = 0;
+  
+  // Thread applies f repeatedly until there are no more indices.
+  // PERF: Can just start each thread knowing its start index, and avoid
+  // the synchronization overhead at startup.
+  auto th = [&index_m, &next_index, &vec, &f]() {
+    for (;;) {
+      index_m.lock();
+      if (next_index == vec.size()) {
+	// All done. Don't increment counter so that other threads can
+	// notice this too.
+	index_m.unlock();
+	return;
+      }
+      int my_index = next_index++;
+      index_m.unlock();
+
+      // Do work, not holding mutex.
+      (void)f(my_index, vec[my_index]);
+    }
+  };
+
+  vector<std::thread> threads;
+  threads.reserve(max_concurrency);
+  for (int i = 0; i < max_concurrency; i++) {
+    threads.emplace_back(th);
+  }
+  // Now just wait for them all to finish.
+  for (std::thread &t : threads) t.join();
+}
+
+// Same, but the typical case that the index is not needed.
+template<class T>
+void ParallelApp(const vector<T> &vec, 
+		 std::function<void(const T&)> &f,
+		 int max_concurrency) {
+  std::function<void(int, const T &)> ff =
+    [&f](int i_unused, const T &arg) { return f(arg); };
+  ParallelAppi(vec, ff, max_concurrency);
+}
+
[…]

### 2015-03-15 portmantout/mingw.condition_variable.h
Threads for portmantout

@@ -1,0 +1,192 @@
+/**
+ * @file mingw.condition_variable.h
+ * @brief std::condition_vaiable implementation for MinGW
+ *
+ * (c) 2014 by Mega Limited, Wellsford, New Zealand
+ *
+ * This file is part of the standard threads implementation for MinGW.
+ *
+ * This code is distributed in the hope that it will be useful,
+ * but WITHOUT ANY WARRANTY; without even the implied warranty of
+ * MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.
+ *
+ * @copyright GNU LGPL version 2.1 License.
+ *
+ * You should have received a copy of the license along with this
+ * program.
+ */
+#ifndef MINGW_CONDITIONAL_VARIABLE_H
+#define MINGW_CONDITIONAL_VARIABLE_H
+#include <atomic>
+#include <assert.h>
+#include <condition_variable>
+namespace std
+{
+
+enum class cv_status { no_timeout, timeout };
+class condition_variable_any
+{
+protected:
+    recursive_mutex mMutex;
+    atomic<int> mNumWaiters;
+    HANDLE mSemaphore;
+    HANDLE mWakeEvent;
+public:
+    typedef HANDLE native_handle_type;
+    native_handle_type native_handle() {return mSemaphore;}
+    condition_variable_any(const condition_variable_any&) = delete;
+    condition_variable_any& operator=(const condition_variable_any&) = delete;
+    condition_variable_any()
+        :mNumWaiters(0), mSemaphore(CreateSemaphore(NULL, 0, 0xFFFF, NULL)),
+         mWakeEvent(CreateEvent(NULL, FALSE, FALSE, NULL))
+    {}
+    ~condition_variable_any() {  CloseHandle(mWakeEvent); CloseHandle(mSemaphore);  }
+protected:
+    template <class M>
+    bool wait_impl(M& lock, DWORD timeout)
+    {
+        {
+            lock_guard<recursive_mutex> guard(mMutex);
+            mNumWaiters++;
+        }
+        lock.unlock();
+            DWORD ret = WaitForSingleObject(mSemaphore, timeout);
+
+        mNumWaiters--;
+        SetEvent(mWakeEvent);
+        lock.lock();
+        if (ret == WAIT_OBJECT_0)
+            return true;
+        else if (ret == WAIT_TIMEOUT)
+            return false;
+//2 possible cases:
+//1)The point in notify_all() where we determine the count to
[…]

### 2015-03-15 portmantout/mingw.mutex.h

@@ -1,0 +1,251 @@
+/**
+ * @file mingw.mutex.h
+ * @brief std::mutex et al implementation for MinGW
+ *
+ * (c) 2014 by Mega Limited, Wellsford, New Zealand
+ *
+ * This file is part of the standard threads implementation for MinGW.
+ *
+ * This code is distributed in the hope that it will be useful,
+ * but WITHOUT ANY WARRANTY; without even the implied warranty of
+ * MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.
+ *
+ * @copyright GNU LGPL version 2.1 License.
+ *
+ * You should have received a copy of the license along with this
+ * program.
+ */
+#ifndef WIN32STDMUTEX_H
+#define WIN32STDMUTEX_H
+
+#if !defined(STDTHREAD_STRICT_NONRECURSIVE_LOCKS) && !defined(NDEBUG)
+    #define STDTHREAD_STRICT_NONRECURSIVE_LOCKS
+#endif
+
+namespace std
+{
+class recursive_mutex
+{
+protected:
+    CRITICAL_SECTION mHandle;
+public:
+    typedef LPCRITICAL_SECTION native_handle_type;
+    native_handle_type native_handle() {return &mHandle;}
+    recursive_mutex() noexcept
+    {
+        InitializeCriticalSection(&mHandle);
+    }
+    recursive_mutex (const recursive_mutex&) = delete;
+    ~recursive_mutex() noexcept
+    {
+        DeleteCriticalSection(&mHandle);
+    }
+    void lock()
+    {
+        EnterCriticalSection(&mHandle);
+    }
+    void unlock()
+    {
+        LeaveCriticalSection(&mHandle);
+    }
+    bool try_lock()
+    {
+        return (TryEnterCriticalSection(&mHandle)!=0);
+    }
+};
+template <class B>
+class _NonRecursiveMutex: protected B
+{
+protected:
+    typedef B base;
+#ifdef STDTHREAD_STRICT_NONRECURSIVE_LOCKS
+    DWORD mOwnerThread;
+#endif
+public:
+    using base::native_handle_type;
+    using base::native_handle;
+    _NonRecursiveMutex() noexcept :base()
+#ifdef STDTHREAD_STRICT_NONRECURSIVE_LOCKS
+    , mOwnerThread(0)
+#endif
+    {}
[…]

### 2015-03-15 trunk/portmantout/portmantout.cc
Score some words!

@@ -1,0 +1,132 @@
+
+#include <vector>
+#include <string>
+#include <unordered_map>
+#include <unordered_set>
+#include <algorithm>
+
+#include "util.h"
+#include "timer.h"
+#include "threadutil.h"
+
+using namespace std;
+
+struct Word;
+using smap = unordered_map<string, vector<Word *>>;
+using sset = unordered_set<string>;
+
+struct Word {
+  explicit Word(const string &s) : w(s) {}
+  const string w;
+  int used = 0;
+  int accessibility = 0;
+  int exitability = 0;
+  int score = 0;
+  sset prefixes;
+  sset suffixes;
+};
+
+int main () {
+  vector<string> dict = Util::ReadFileToLines("wordlist.asc");
+  printf("%d words.\n", (int)dict.size());
+  
+  vector<Word> words;
+  words.reserve(dict.size());
+  for (const string &s : dict) {
+    // Should strip.
+    if (s.empty()) continue;
+    for (char c : s) {
+      if (c < 'a' || c > 'z') {
+	printf("Bad: [%s]\n", s.c_str());
+      }
+    }
+    words.push_back(Word{s});
+  }
+
+  smap fwd;
+  smap bwd;
+  vector<Word *> empty;
+  auto Forward = [&empty, &fwd](const string &s) -> const vector<Word *> & {
+    auto it = fwd.find(s);
+    if (it == fwd.end()) return empty;
+    else return it->second;
+  };
+
+  auto Backward = [&empty, &bwd](const string &s) -> const vector<Word *> & {
+    auto it = bwd.find(s);
+    if (it == bwd.end()) return empty;
+    else return it->second;
+  };
+
+  for (Word &word : words) {
+    // XXX Prefix and suffix should of course be nonempty, but should
+    // we include the entire word? I mean, in the case of 'a' we kind
+    // of have to?
+    for (int i = 1; i <= word.w.length(); i++) {
+      string prefix = word.w.substr(0, i);
+      string suffix = word.w.substr(i - 1, string::npos);
+      word.prefixes.insert(prefix);
+      word.suffixes.insert(suffix);
+      fwd[prefix].push_back(&word);
+      bwd[suffix].push_back(&word);
+    }
+  }
+  printf("Built prefix/suffix map.\n");
+
+  Timer onepass;
+  for (Word &word : words) {
+    word.accessibility = 0;
+    word.exitability = 0;
+  }
[…]

### 2015-03-15 trunk/portmantout/portmantout.cc
Greedy approach gets stuck, obviously

@@ -4,16 +4,25 @@
 #include <unordered_map>
 #include <unordered_set>
 #include <algorithm>
+#include <deque>
 
 #include "util.h"
 #include "timer.h"
 #include "threadutil.h"
+#include "arcfour.h"
+#include "randutil.h"
 
 using namespace std;
 
 struct Word;
 using smap = unordered_map<string, vector<Word *>>;
 using sset = unordered_set<string>;
+
+std::mutex print_mutex;
+#define Printf(fmt, ...) do {		\
+  MutexLock Printf_ml(&print_mutex);		\
+  printf(fmt, ##__VA_ARGS__);			\
+  } while (0);
 
 struct Word {
   explicit Word(const string &s) : w(s) {}
@@ -26,11 +35,10 @@ struct Word {
   sset suffixes;
 };
 
-int main () {
-  vector<string> dict = Util::ReadFileToLines("wordlist.asc");
-  printf("%d words.\n", (int)dict.size());
-  
+int Attempt (ArcFour *rc, const vector<string> &dict) {
   vector<Word> words;
+  int total_letters = 0;
+  int max_length = 0;
   words.reserve(dict.size());
   for (const string &s : dict) {
     // Should strip.
@@ -41,7 +49,11 @@ int main () {
       }
     }
     words.push_back(Word{s});
+    total_letters += s.size();
+    max_length = std::max((int)s.size(), max_length);
   }
+
+  printf("Total size %d, max length %d\n", total_letters, max_length);
 
   smap fwd;
   smap bwd;
@@ -59,9 +71,6 @@ int main () {
   };
 
   for (Word &word : words) {
-    // XXX Prefix and suffix should of course be nonempty, but should
-    // we include the entire word? I mean, in the case of 'a' we kind
-    // of have to?
     for (int i = 1; i <= word.w.length(); i++) {
       string prefix = word.w.substr(0, i);
       string suffix = word.w.substr(i - 1, string::npos);
@@ -88,13 +97,95 @@ int main () {
       }
     }
   }
+  for (Word &word : words) {
+    word.score = word.score + word.exitability;
+  }
+
   double op_ms = onepass.MS();
-  printf("Took %fms\n", op_ms);
+  printf("Scoring took %fms\n", op_ms);
+  
+  vector<Word *> sorted;
+  sorted.reserve(words.size());
+  for (Word &word : words) sorted.push_back(&word);
+  std::sort(sorted.begin(), sorted.end(),
[…]

### 2015-03-15 portmantout/randutil.h
Greedy approach gets stuck, obviously

@@ -3,6 +3,11 @@
 
 #include "arcfour.h"
 #include <cmath>
+#include <cstdint>
+
+using uint8 = uint8_t;
+using uint64 = uint64_t;
+using uint32 = uint32_t;
 
 // Caller owns new-ly allocated pointer.
 ArcFour *Substream(ArcFour *rc, int n) {

### 2015-03-15 portmantout/paper.tex
Makes particles successfully

\documentclass{article}
\usepackage[top=0.4in, left=0.4in, right=0.4in]{geometry}

\usepackage{setspace}
\usepackage{seqsplit}
\usepackage{amsmath}
\usepackage{amssymb}
% \usepackage{code}
\usepackage{graphicx}
\usepackage{fancyvrb}
\usepackage{url}
\usepackage{textcomp}

\pagestyle{empty}

\newcommand\comment[1]{}

\newcommand\st{$^{\mathrm{st}}$}
\newcommand\nd{$^{\mathrm{nd}}$}
\newcommand\rd{$^{\mathrm{rd}}$}
\renewcommand\th{$^{\mathrm{th}}$}
\newcommand\tm{$^{\mbox{\tiny \textsc{tm}}}$}

\newcommand\sfrac[2]{{}\,$^{#1}$\!/{}\!$_{#2}$}

\newcommand\citef[1]{\addtocounter{footnote}{1}\footnotetext{\cite{#1}}\ensuremath{^{\mbox{\footnotesize [\thefootnote]}}}}
\begin{document}

\title{Portmantout}
\author{Dr.~Tom~Murphy~VII~Ph.D.\thanks{
WIKIPEDIA!!!!
}
}

\renewcommand\>{$>$}
\newcommand\<{$<$}

\date{15 Mar 2015}

\maketitle

\begin{abstract}
Long word ok cool
\end{abstract}

\vspace{1em}
{\noindent \small {\bf Keywords}:
 super
}

\newcommand\foo[1]{%
  \begin{minipage}{7.6in}
  \seqsplit{#1}
  \end{minipage}
  }

\section{Introduction}

A {\em portmanteau}, henceforewith non-italicized, is a
stringin'-together of two words to make a new word, like
``brogrammer'' ({\sf brother} + {\sf programmer}; a programmer who is
your brother), or (other example).

It is natural to think of generalizations of the portmanteau, such as
the {\em portmantrois},\!\footnote{Graham Smith, personal
  communication.} (itself a portmanteau of {\sf portmanteau} + {\sf
  trois}, French-language for three) the human-centipedification of
{\em three} words, such as ``anachillaxis'' ({\sf anaphylaxis} + {\sf
  chill} + {\sf relax}; a severe allergic reaction to idleness) or
``brogrammermaid'' ({\sf brother} + {\sf programmer} + {\sf mermaid};
a programmer who is your brother and a mermaid).

In this paper I present the world's first (?)\footnote{I did a couple
  Google searches; seems good enough.} {\em portmantout}, a
portmanteau of all English-language words.

overmagnifyingearlesshrimpierabbinicalcicadaeratorshrimpiestandardizableaseholdershrimpingementshrimpsychedelically ({\sf overmagnifying} + {\sf gearless} + {\sf shrimpier} + {\sf rabbinical} + {\sf calc} + {\sf cicada} + {\sf aerator} + {\sf shrimpiest} + {\sf standardizable} + {\sf leaseholders} + {\sf shrimping} + {\sf impingements} + {\sf shrimps} + {\sf psychedelically}; presumably having something to do with shrimp).

\section{The big ol' word}
\tiny
\setstretch{0.1}
biiiiig word.

.... put big word here ....

\end{document}

### 2015-03-15 trunk/portmantout/join.cc
Makes particles successfully

@@ -70,7 +70,6 @@ int Attempt (ArcFour *rc, const vector<string> &dict) {
     if (it == bwd.end()) return empty;
     else return it->second;
   };
-  (void)Backward;
 
   for (Word &word : words) {
     for (int i = 1; i <= word.w.length(); i++) {

### 2015-03-15 portmantout/join.cc
Works. Let's write it up and then seek improvements

@@ -19,42 +19,105 @@ struct Word;
 using smap = unordered_map<string, vector<Word *>>;
 using sset = unordered_set<string>;
 
+struct Word {
+  explicit Word(const string &s) : w(s) {}
+  const string w;
+  sset prefixes;
+  sset suffixes;
+};
+
 std::mutex print_mutex;
 #define Printf(fmt, ...) do {		\
   MutexLock Printf_ml(&print_mutex);		\
   printf(fmt, ##__VA_ARGS__);			\
   } while (0);
 
-struct Word {
-  explicit Word(const string &s) : w(s) {}
-  const string w;
-  int used = 0;
-  int accessibility = 0;
-  int exitability = 0;
-  int score = 0;
-  sset prefixes;
-  sset suffixes;
-};
+int main () {
+  vector<string> dict = Util::ReadFileToLines("wordlist.asc");
+  printf("%d words.\n", (int)dict.size());
+  vector<string> particles = Util::ReadFileToLines("particles.txt");
+  printf("%d particles.\n", (int)particles.size());
+  ArcFour rc("portmantout");
 
-int Attempt (ArcFour *rc, const vector<string> &dict) {
-  vector<Word> words;
   int total_letters = 0;
-  int max_length = 0;
-  words.reserve(dict.size());
-  for (const string &s : dict) {
-    // Should strip.
-    if (s.empty()) continue;
-    for (char c : s) {
-      if (c < 'a' || c > 'z') {
-	printf("Bad: [%s]\n", s.c_str());
+  for (const string &w : dict) {
+    total_letters += w.size();
+  }
+  
+  // Build a complete graph telling us how to efficiently get from the
+  // source letter to destination, where both source and destination
+  // are in [a-z]. We'll fill in the diagonal too, but we won't need
+  // to use it.
+
+  struct Path {
+    Path() {}
+    bool has = false;
+    // Must start with src character and end with dst character.
+    string word;
+  };
+
+  // Note that source letter will be ending a particle we're trying
+  // to join, and dst will be beginning one.
+  // matrix[src][dst]
+  vector<vector<Path>> matrix(26, vector<Path>(26, Path()));
+
+  auto SaveTable = [&matrix](const string &filename) {
+    FILE *f = fopen(filename.c_str(), "wb");
+    fprintf(f, "<!doctype html>\n"
[…]

### 2015-03-15 trunk/portmantout/portmantout.cc
Works. Let's write it up and then seek improvements

@@ -238,23 +238,6 @@ int Attempt (ArcFour *rc, const vector<string> &dict) {
   }
 
   return 1;
-
-#if 0
-  for (int idx = 0; idx < sorted.size(); idx++) {
-
-
-    if (idx % 100 == 0) {
-      printf("Portmantout is %d chars, working on [%s].c_str()");
-
-    }
-    const string &thisword = sorted[i]->w;
-    // If the word is already in the portmantout, then we are done.
-    if (portmantout.find(thisword) == string::npos) {
-      already_substr++;
-    }
-
-  }
-#endif
 }
 
 int main () {
@@ -264,40 +247,5 @@ int main () {
   
   Attempt(&rc, dict);
 
-#if 0
-  vector<const Word *> sorted;
-  for (const Word &word : words) sorted.push_back(&word);
-  std::sort(sorted.begin(), sorted.end(),
-	    [](const Word *l, const Word *r) {
-	      return l->score < r->score;
-	    });
-  for (const Word *word : sorted) {
-    printf("%d -> %s -> %d\n", word->accessibility, word->w.c_str(), word->exitability);
-  }
-#endif
-
   return 0;
 }
-
-#if 0
-  int starts[26] = {};
-  int ends[26] = {};
-
-  for (const string &s : dict) {
-    // Should strip.
-    if (s.empty()) continue;
-    int st = s[0] - 'a';
-    int en = s[s.size() - 1] - 'a';
-    if (st < 0 || st >= 26 ||
-	en < 0 || en >= 26) {
-      printf("Bad: [%s]\n", s.c_str());
-    } else {
-      starts[st]++;
-      ends[en]++;
-    }
-  }
-
-  for (int i = 0; i < 26; i++) {
-    printf("%6d   %c %6d\n", starts[i], 'a' + i, ends[i]);
-  }
-#endif

### 2015-03-15 portmantout/paper.tex
First submitted version, sunday night

@@
 \documentclass{article}
-\usepackage[top=0.4in, left=0.4in, right=0.4in]{geometry}
+\usepackage[top=0.25in, left=0.25in, right=0.25in, bottom=0.25in]{geometry}
 
@@
 
-\title{Portmantout}
+\title{The Portmantout}
 \author{Dr.~Tom~Murphy~VII~Ph.D.\thanks{
-WIKIPEDIA!!!!
+Copyright \copyright\ 2015 the Regents of the Wikiplia
+Foundation. Appears in SIGBOVIK 2015 with the {\bf etaoin shrdlu} of the
+Association for Computational Heresy; {\em IEEEEEE!} press,
+Verlag-Verlag volume no.~0x40-2A.
+BTC~0.00
 }
@@
 
-\date{15 Mar 2015}
+\date{1 April 2015}
 
 \maketitle
-
-\begin{abstract}
-Long word ok cool
-\end{abstract}
-
-\vspace{1em}
-{\noindent \small {\bf Keywords}:
- super
-}
 
@@
 ``brogrammer'' ({\sf brother} + {\sf programmer}; a programmer who is
-your brother), or (other example).
+your brother), ``hupset'' ({\sf hungry} + {\sf upset}; a bit more
+passive than hangry), or ``webinar'' ({\sf web} + {\sf nerd}).
+Portmanteaus were invented by Lewis Carroll, the Jabberwock of
+wordplay.
 
@@
 In this paper I present the world's first (?)\footnote{I did a couple
-  Google searches; seems good enough.} {\em portmantout}, a
-portmanteau of all English-language words.
+  Google searches; seems good enough.}\footnote{Enjoy source code: {\tt http://sourceforge.net/p/tom7misc/svn/HEAD/tree/trunk/portmantout}} {\em portmantout}, a
+portmanteau of all English-language words ({\em tout} means ``all'' in
+French-language). I also considered calling this a {\em portmantotal},
+{\em portmantotale}, etc., as well as {\em portmantoutal} (a
+portmantroix of the first three) or even {\em portmantoutale}. You
+kind of see how this can get out of hand. The word is
+%% XXX check final version
+630,408
+%% XXX
+letters long and contains all 108,709 words in a particular wordlist
+called {\tt wordlist.asc}.\footnote{Tom Murphy VII, ``What words ought
+  to exist?'', SIGBOVIK~2011} Even though nobody can really agree
+what all the words in English are, the technique used to generate this
+portmantout should work for most very long word lists, although we will
+see in Section~\ref{sec:join} that a handful of words are very important.
 
-overmagnifyingearlesshrimpierabbinicalcicadaeratorshrimpiestandardizableaseholdershrimpingementshrimpsychedelically ({\sf overmagnifying} + {\sf gearless} + {\sf shrimpier} + {\sf rabbinical} + {\sf calc} + {\sf cicada} + {\sf aerator} + {\sf shrimpiest} + {\sf standardizable} + {\sf leaseholders} + {\sf shrimping} + {\sf impingements} + {\sf shrimps} + {\sf psychedelically}; presumably having something to do with shrimp).
+Since the word itself is
+%% XXX check final version
+11 pages
+%% XXX
+long in 4pt type with $.75$ linespacing, and this SIGBOVIK proceedings is positively overfull
+hbox with content, we should probably get on with it.
 
-\section{The big ol' word}
-\tiny
-\setstretch{0.1}
-biiiiig word.
+\section{Computationalizing ``portmanteau''}
 
-.... put big word here ....
+A real portmanteau is usually phonetic, like ``portmantotally'' is
+about the sound of ``---teau'' being the same as ``to---''. It's also
+not unusual for part of the word to be completely dropped, as in
+``chillax'', which drops the ``re---'' from {\sf relax}. They are also
+usually clever or meaningful. For the sake of computing a portmantout,
+we need to make some rules about what it is, and it can't require
+cleverness or semantic/phonemic interpretation of words if I'm going to
+start and complete this project on the day of the SIGBOVIK deadline.
+
+{\bf Generalized portmanteau.}\quad For a set of strings $L$, a string
+$s$ is a generalized portmanteau if the entire string can be covered
+by strings in $L$. A cover is a set of word occurrences $W = \langle m, n \rangle$,
+where $s_m$--$s_n$ (the substring starting at offset $m$ and ending at $n$, inclusive)
+is in $L$, and, taken sorted by the $m$ component, $W_i.n >= W_{i+1}.m$ for
+each $i$ in range. For example,
+
+\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\includegraphics[width=2in]{rewrotempered}
+
+This string can be covered by {\sf rewrote}, {\sf temper}, and {\sf
+  red}, so it is a generalized portmanteau if these three strings are
+in $L$. (Spoiler alert: $L$ is English-language, so they are.)
+Importantly, the covering strings overlap: {\sf rewro}, {\sf temper}
+and {\sf ed} on their own would not cover this string (and {\sf rewro}
+is not a word). Therefore, we cannot simply concatenate the entire
+dictionary.
+
+{\bf Portmantout.}\quad A generalized portmanteau is a portmantout if
+it contains every string in $L$ as a substring. The words need not be
+in its cover, as there may be muliple covers (In fact I conspicuously
+did not choose the simpler one {\sf rewrote} + {\sf tempered}.) Other
+words, like {\sf wrote}, {\sf rote} and {\sf rot} are in there
+too ``for free'', even though they may not be able to participate in
+a legal cover.
+
+A word may appear multiple times; we just have to get them all. This
+is fairly unavoidable---the word {\sf a} appears
+%% XXX check final
+60,374 times
+% XXX
+in the portmantout. Perhaps more surprising is that the word {\sf
+  iraq} appears
+%%% XXX check final
+315 times.
+%% XXX
+
+Note that a ``generalized'' portmanteau does not actually include most
+colloquial portmanteaus; {\sf brogrammer} cannot be covered since we
+dropped the ``p---'' in {\sf programmer}. {\sf brogrammar} is also not
+a generalized portmanteau since {\sf bro} and {\sf grammar} do not
+overlap, yo, but I think it would be accepted colloquially by most
+dudes. Disrupt!
+
+\section{Generating a portmantout}
+
+It's fairly straightforward handwaving to see that generating the
+shortest portmantout is NP-complete. Seeing that it is in NP is easy;
+we just need to check the cover and look up all the substrings, which
+is clearly polynomial. It is probably NP-hard because the traveling
+salesman problem can be reduced to it; for each node in the graph,
+generate a two-symbol word $xy$ where $x$ and $y$ are fresh symbols;
+for each edge between cities $x_iy_i$ and $x_jy_j$ generate a string
+$y_iu^{w}x_j$, where $u$ is also a fresh symbol repeated $w$ times,
+the cost of the edge. This allows us only to join two city words by
+using a corresponding edge word.\footnote{There are some rubs: TSP
+  requires that nodes only be visited once but a portmantout can use
+  words multiple times. I believe that the multi-visit generalization
+  of TSP is also NP-hard. The portmantout solution also requires visiting
+  every edge, but we can relax this by concatenating all edge words $e$ into a
+  new mega-long word like $e_0 z e_1 z \ldots e_k$ where $z$ is also
+  a fresh symbol; since this word must appear and all edges are substrings
+  of it, we now have no requirement that the rest of the solution (containing
+  our TSP embedding) contains all edge words. This kind of trick also lets us
+  set the start node for TSP.}
+
+OK but good news! Since it's NP-complete, we know that we can come up
+with a solution that's both non-optimal and slow, and we can still
+feel pretty good about it. We proceed in two steps: Generating
+particles eagerly, and then joining them together.
+
+\subsection{Generating particles} \label{sec:gen}
+
+For the first step, we load up all the words, and insert each word
+into a multimap, keyed by each of its non-empty prefixes. We then
+start by initializing a particle from any word; we choose {\sf
+  portmanteau} to start, obviously. Then, repeatedly:
+\begin{itemize}
+\item Check each suffix of the particle in descending length-order,
+\begin{itemize}
+  \item If we have a word that has not already been used, strip the suffix from its start and append the remainder to the particle
+\end{itemize}
+\item Otherwise, emit the particle and start a new one with any unused word.
+\end{itemize}
+
+As an additional optimization, we discard words that are
+substrings of any particle. This search makes the program much slower,
+but it produces a much more efficient portmantout.
+
+This always makes progress, using up one word at each step: We either
+append it to our current particle, or we start a new particle with a
+word. The particles are all generalized portmanteaus by construction,
+because each added word has non-empty overlap with the previous one.
+Here's an example particle: {\sf overmagnify\-ingearlesshrimp\-ierabbinicalci\-cadaeratorshrimpiestandard\-izablease\-holdershrimp\-inge\-mentshrimpsychedelically} ({\sf overmagnifying} + {\sf gearless} + {\sf shrimpier} + {\sf rabbinical} + {\sf calc} + {\sf cicada} + {\sf aerators} + {\sf shrimpiest} + {\sf standardizable} + {\sf leaseholders} + {\sf shrimping} + {\sf impingements} + {\sf shrimps} + {\sf psychedelically}; presumably having something to do with shrimp).
+
+At this point, we have about 38,000 particles, many of which are a
+single word. English contains many imbalances, like vastly more words
+ending with ``---y'' (10,071) than beginning ``y---'' (only 338), so
+it is not hard to see how we may get stuck with no new words to add to
+a particle. We've also used each word only once, and locally maximized
+the amount of overlap. If we can join these particles together in a
+valid way, we'll have a portmantout.
+
+\subsection{Joining particles} \label{sec:join}
+
+Since we've already used every word, and, by construction, these
+particles cannot be adjoined directly, we know we will need to reuse
+some words to join them together. A simple way to do this is to
+construct a table of size $26^2$ that for every pair of letters $a$
+and $b$, contains a short word that starts with $a$ and ends with $b$.
+86\% of the table entries can be filled in, but some letters are very
+tricky: For example, almost no words end with ``q'' (we have only {\sf
+  colloq} and {\sf iraq} in our dictionary), and there are no words
+that start with ``v'' and end with ``f''. Fortunately, if we consider
+all two-word (generalized) portmanteaus, using basically the same
+algorithm as in Section~\ref{sec:gen}, we can fill the table
+completely (Figure~\ref{fig:table}).
+
+It is lucky for the existence of words like {\sf iraq}; they are used
+for many of the entries in the ``q'' column. In fact, without a
+handful of such words, it might be the case that English would not
+permit a portmantotal! There are probably some less irregular
+languages that cannot achieve this lexical feat. \verb+:'-(+
+
+This table alone would allow us a very simple algorithm for generating
+a valid portmantotal: Just take words from the dictionary and
+concatenate them, but when putting e.g. {\sf tv} and {\sf farm}
+together, we use the v--f linker {\sf vetof} ({\sf veto} + {\sf of})
+from the table. We can't ever fail! However, this would produce a
+portmantout that's bigger than the dictionary itself, which isn't very
+economical. Rather than use the whole dictionary this way, we instead
+join all of the 38,000 particles from the previous section. These are
+much more compact. And now we are done!
+
+\section{The portmantout}
+
+This portmantout is
+%% XXX check final version
+630,408
+%% XXX
+letters long; there are 931,823 total letters in the dictionary so this
+is a compresion ratio of 
+%% XXX
+1.47:1.
+%% XXX
+In comparison, ``brogrammermaid'' (although an illegal generalized
+portmanteau) has a compression ratio of \sfrac{24}{14} = 1.71:1. So it
+is fair to say that we are in the ballpark of a ``solid portmanteau.''
+Of course, the gold standard is a compression ratio of $\infty : 1$---for
+the case that we have the word {\sf portmanteau}, a totally overlapping portmanteau of
+{\sf portmanteau} + {\sf portmanteau},\!\footnote{Cara Gillotti, personal communication, 2015.} iterated infinitely.
+
+\begin{figure}[htp]
+\input{table-two}
+\caption{Minimal joining strings for every letter (rows) to every other letter (columns).} \label{fig:table}
+\end{figure}
+
+%  \begin{sloppypar}
+%  \tiny
+%  \setstretch{0.1}
+%  \hyphenpenalty=1
+%  \pretolerance=2
+%  \tolerance=2
+%  
+%  \noindent \input{portmantout}
+%  \end{sloppypar}
 

### 2015-03-15 portmantout/join.cc

@@ -12,6 +12,7 @@
 #include "arcfour.h"
 #include "randutil.h"
 #include "base/logging.h"
+#include "base/stringprintf.h"
 
 using namespace std;
 
@@ -43,6 +44,7 @@ int main () {
   for (const string &w : dict) {
     total_letters += w.size();
   }
+  printf("%d total letters.\n", total_letters);
   
   // Build a complete graph telling us how to efficiently get from the
   // source letter to destination, where both source and destination
@@ -62,31 +64,70 @@ int main () {
   vector<vector<Path>> matrix(26, vector<Path>(26, Path()));
 
   auto SaveTable = [&matrix](const string &filename) {
-    FILE *f = fopen(filename.c_str(), "wb");
-    fprintf(f, "<!doctype html>\n"
-	    "<style> td { padding: 0 4px } body { font: 13px Verdana } </style>\n"
-	    "<table>");
-    
-    fprintf(f, "<tr><td>&nbsp;</td> ");
-    for (int i = 0; i < 26; i++)
-      fprintf(f, "<td>%c</td>", 'a' + i);
-    fprintf(f, " </tr>\n");
-    
-    for (int src = 0; src < 26; src++) {
-      fprintf(f, "<tr><td>%c</td> ", 'a' + src);
-      for (int dst = 0; dst < 26; dst++) {
-	const Path &p = matrix[src][dst];
-	if (p.has) {
-	  fprintf(f, "<td>%s</td>", p.word.c_str());
-	} else {
-	  fprintf(f, "<td>&mdash;</td>");
-	}
-      }
+    {
+      FILE *f = fopen(StringPrintf("%s.html", filename.c_str()).c_str(), "wb");
+      fprintf(f, "<!doctype html>\n"
+	      "<style> td { padding: 0 4px } body { font: 13px Verdana } </style>\n"
+	      "<table>");
+
+      fprintf(f, "<tr><td>&nbsp;</td> ");
+      for (int i = 0; i < 26; i++)
+	fprintf(f, "<td>%c</td>", 'a' + i);
       fprintf(f, " </tr>\n");
+
+      for (int src = 0; src < 26; src++) {
+	fprintf(f, "<tr><td>%c</td> ", 'a' + src);
+	for (int dst = 0; dst < 26; dst++) {
+	  const Path &p = matrix[src][dst];
+	  if (p.has) {
[…]

### 2015-03-18 trunk/redi/make-corpus.sh
Paper submission, though some graphics are non-free so not added to sourceforge. Forked jpeg dequantization one too

@@ -86,6 +86,18 @@ make fetcher.exe
 # ./fetcher.exe park &
 # ./fetcher.exe architecture
 
-./fetcher.exe waterslide &
-./fetcher.exe tarantula &
-./fetcher.exe "ferris wheel"
+# ./fetcher.exe waterslide &
+# ./fetcher.exe tarantula &
+# ./fetcher.exe "ferris wheel"
+
+./fetcher.exe highway
+./fetcher.exe snakes
+./fetcher.exe airplane
+./fetcher.exe cocktail
+./fetcher.exe "cocktail party"
+./fetcher.exe "dinner party"
+./fetcher.exe superyacht
+./fetcher.exe helicopter
+./fetcher.exe coffee
+./fetcher.exe "dog on skateboard"
+./fetcher.exe "mri machine"

### 2015-03-18 redi/paper/paper.tex
Paper submission, though some graphics are non-free so not added to sourceforge. Forked jpeg dequantization one too

\documentclass[twocolumn]{article}
\usepackage[top=1.1in, left=0.85in, right=0.85in]{geometry}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{fancyvrb}
\usepackage{url}
\usepackage{textcomp}
% \usepackage{inlinebib}

\pagestyle{empty}

\newcommand\comment[1]{}

\newcommand\st{$^{\mathrm{st}}$}
\newcommand\nd{$^{\mathrm{nd}}$}
\newcommand\rd{$^{\mathrm{rd}}$}
\renewcommand\th{$^{\mathrm{th}}$}
\newcommand\tm{$^{\mbox{\tiny \textsc{tm}}}$}

% nice fractions
\newcommand\sfrac[2]{{}\,$^{#1}$\!/{}\!$_{#2}$}

\newcommand\citef[1]{\addtocounter{footnote}{1}\footnotetext{\cite{#1}}\ensuremath{^{\mbox{\footnotesize [\thefootnote]}}}}

% XXX TODO: before submit, search ``red eye'' and ``reduction'' vs removal
% we -> I

\begin{document}

\title{Red i removal with artificial retinal networks}
\author{Dr.~Tom~Murphy~VII~Ph.D.\thanks{
Copyright \copyright\ 2015 the Regents of the Wikiplia
Foundation. Appears in SIGBOVIK 2015 with the increasingly askance visage of the
Association for Computational Heresy; {\em IEEEEEE!} press,
Verlag-Verlag volume no.~0x40-2A.
0.00 Australian Neo-Dollars}
}

\renewcommand\>{$>$}
\newcommand\<{$<$}

\date{1 April 2015}

\maketitle

\begin{abstract}
We present a GPU-accelerated means for red i removal in photographs.
\end{abstract}

\vspace{1em}
{\noindent \small {\bf Keywords}:
  computational photography, image processing, generalized photoshop, artificial retinal networks, types
}

\section{Introduction}

When light---such as the bright flashbulb of a camera---strikes the
human eye, it illuminates the retina. Some of that light bounces back
out of the eye, but most of it stimulates neurons in the retina to
produce electrical signals. These signals stimulate other neurons to
which they are connected, and so on, until the brain (which is
technically part of the eye) perceives an image, as a two-dimensional
array of neurons with different activation levels. Humans often use
these images to sense the world, for example, in reading research papers.

This research paper concerns a particular feature of this process,
which is that humans are able to view an image and ignore certain
details of it. For example, Figure~\ref{fig:cbs} contains a printout
of an image file of a photograph of a television displaying a recorded
video of an actor. The video contains a superimposed eye in the
corner, the logo of the network CBS. Most viewers are not tormented by
this everpresent eye staring at them! In fact, most viewers are able
to completely ignore the eye, and view the scene as though it didn't
contain the stimulus, even if details such as the actor's sweatshirt's
collar pass beneath the stimulus and are occluded by it.

\begin{figure}[t]
\begin{center}
\includegraphics[width=0.85 \linewidth]{cbs}
\end{center}
\caption{Q.~Who watches the TV~watchers?
A.~CBS's all-seeing eye.} \label{fig:cbs}
\end{figure}

Some stimulus is more everpresent than others. The Clay Mathematics
Institute lists among its unsolved {\em Millennium Prize} problems the
``red i removal problem.'' This concerns the removal of stimulus (a
red letter ``i'') from images (Figure~\ref{fig:redi}). The problem
is particularly difficult because the information occluded by the
i is completely gone, and because the authors of papers about the
problem are persistently agitated because it seems like the letter
should be capitalized.

In this paper I show how red i removal can be solved in certain
specialized cases, using an artificial retinal network patterned after
the brain contained within the human eye. Training this artificial
retinal network is feasible on a single powerful desktop machine. Both
training and execution of the model (a mere 400 megabytes) are GPU
accelerated. The model presented in this paper was trained in about 3
days, and executing it in parallel on a suite of images takes about
100 milliseconds per image.\!\footnote{Source code is available on
the World Wide Web at:
{\tt http://sourceforge.net/p/tom7misc/svn/HEAD/tree/trunk/redi}}

\begin{figure}[tb]
\begin{center}
\includegraphics[width=0.65 \linewidth]{rotor}
\end{center}
\caption{An image of an Enigma machine rotor with a red i
 superimposed. Solving this instance of the red i removal problem
 would mean producing an image without the red i. One way to do this
 would be to steal a floppy disk containing an original, unimposed
 image of the rotor, from someone in possession of it, for example
 the paper's author.} \label{fig:redi}
\end{figure}

\section{Artificial retinal networks}

As I expertly foreshadowed in the previous section, an artificial
retinal network works just like the brain inside a human eye. The
retina is itself a rectangular 2D array of neurons, which turn photons
into IEEE-754 floating point values between 0.0f and 1.0f. Behind this
is another 2D array of exactly the same size, except the pixels are in
a weird jumbled order, and then another layer. This series of layers
is known as the optic nerve. Finally, the brain perceives the image as
an array of pixels, again of the same size (Figure~\ref{fig:retinal}).

It is easy to reconceptualize this process as an array of pixels
undergoing several transformations. Obviously the story is more
complicated: Humans see in color, so each pixel is actually three
different nodes, one for red, green and blue. In fact, since some
scientists hypothesize of certain ``superseers''---that is, people who
can perceive more than just the three wavelengths of light---we
actually allow an arbitrary number of nodes per pixel. In this work,
we used $N=4$.

In a real human eye, each node is fed inputs from every node in the
previous layer. For computational efficiency, in this work we allow
only 64 inputs to each node from the previous layer. Because we
suspect that layers are spatially related, a node is always connected
to its {\em neighborhood} in the previous layer (each node from the 9
pixels within Manhattan distance 1). The rest of the inputs are
selected randomly from a Gaussian distribution, as long as the samples
fall within the image (using rejection sampling---the sides and
corners do not ``wrap around''). By the way, the images are always
256x256, because numbers that are a power of two are
faster.\!\footnote{This is true on computers, because computers count
in binary. In the human eye, powers of ten are faster, because humans
have ten fingers.} The connection from one node to another is
modulated by a {\em weight}, again an IEEE-754 floating point number.
A node outputs the sum of its input values, passed through a
smoothulator, specifically the one found in Gray's~Anatomy (the book,
not the TV show),

$$ \frac{1}{1 + e^{-v}} $$

\begin{figure}[tb]
\includegraphics[width=0.95 \linewidth]{retinal}
\caption{ How eyes work, and thus, artificial retinal networks. } \label{fig:retinal}
\end{figure}

This function is {\em biologically plausible}.

We learn by backpropagated stochastic gradient descent, like babies
do. Specifically: The network is presented with an image on its
retina, and then we run the floating point values through the layers,
summing them up and applying the smoothulator function, to produce a
final image within the brain. Like a baby, it compares this image to
what it expected to see, node by node. Where each node does not agree
with the image, the error is computed. The baby propagates the partial
derivative of the change in error with respect to the change in
stimulus to the previous layer proportional to its weighted impact;
fortunately the smoothulator has a simple derivative that is easily
computed at a point from its output values. Error is not propagated
into the real world (i.e., by sending light off of the retina back to
the physical object that created the stimulus; that would be
ridiculous).

\subsection{Training data}

One of the insights of this paper is that although artificial retinal
networks require a lot of data to train, for certain problems the
training data can be easily {\em generated}. For the red i removal
problem, we begin with a corpus of about 4,000 images that were
scraped from Google Image Search. Scraping images is easy; one just
needs to list a bunch of queries for things that babies would want to
view in order to learn what the world looks like. In this experiment I
used terms such as [snakes], [dog on skateboard], [guitar],
[stonehenge] and [superyacht]. Following this, I manually cleaned the
training data. I deleted images that were not photographic (drawings,
etc.; for example, most images of guitars are actually 3D rendered
cartoon guitars, fantasy images of guitars on fire, the Guitar~Hero logo,
and so on) or that were {\em too}, uh, pornographic (most queries for
things that can have sex, like tapirs, contain prominent images of the
things having sex). These are not appropriate for babies.

All images are cropped to a square and resized to 256x256. Then we
generate training instances: An input image and the expected result.
An image that does not contain a red i should just be transformed into
the image itself (indeed, when we peer directly into the brain of a
baby looking at a TV show, we find a region of the brain where the TV
show is clearly visible). It is also easy to generate instances of the
red i removal problem along with their solutions---we simply put a red
i randomly on the source image and keep the destination image
unchanged. In this way, we can easily generate a large amount of
training instances (in actual practice, this procedure had a small bug;
see Section~\ref{sec:evaluation}). One unexpected phenomenon is that
I had to be careful to remove images that already contained a red i,
like many images of casinos, which are often called ``CASINO''.

% notes on corpus:
% - my own picture of library in there

In order to coax networks into recognizing the i, we also place an i
into the 4th color channel in the same position in the {\em expected
output}. This is an invisible color channel which we discard, and
which is always zero in the input. In essence we giving a hint to the
eye's brain that it should not just remove the red i, but it should
also {\em perceive} it. I have not performed enough experiments to
know if this is helpful.

For repeatability's sake, important constants used in this experiment:
There were 2 hidden layers. Gaussan samples were produced with a
standard deviation of 16 pixels. The red i was rendered in Comic Sans,
at a height of 80 pixels. I used a variety of learning rates,
including an expontentially decreasing one (the standard advice of
0.05 is too large for constant learning rates on this kind of task, and
limits the sharpness of resultant images).

\subsection{CUDA, SHUDA, WUDA}

Because we are working with graphical data, we should use the Graphics
Processing Unit of the computer, not its Central Processing Unit (we
are not processing centers). I implemented high-performance OpenCL
kernels for each phase of training: The {\em forward} pass (signals
flowing from the retina to the eye's brain), the {\em backpropagation}
step (when the eye computes the error and partial derivatives) and the
{\em weight update} step (when the eye rewires its neurons so that it
sees the right thing next time). The phases have different parallelism
constraints. Because the connectivity is sparse, we represent both
forward and inverted index maps, which are decoded on the GPU. We take
care to only load a single layer of the retinal network into the GPU's
RAM at once, to enable very large models, but we run many training
instances in parallel for a single round. Some other operations, like
the preparation of training data, are performed on the CPU. These are
also frequently done in parallel, using C++11's new {\tt std::thread}
with some crazy-ass wrappers to allow them to function in mingw32's
64-bit gcc port. On a good day, training uses all 6 CPU cores and all
2800 GPU cores and about 14 GB of RAM and warms the home office like a
1kW space heater.

\section{Results}

After 4 rounds of training, the network produces an excellent-looking
image that could be a {\em Cure} album cover, regardless of the stimulus
cast upon its retina (Figure~\ref{fig:cure}).

\begin{figure}[bth]
\begin{center}
\includegraphics[width=0.65 \linewidth]{cure}
\end{center}
\caption{Result after 4 rounds of training. Looks great, and there is no
red i to be seen, but it loses some points for not resembling the
input image at all.} \label{fig:cure}
\end{figure}

\begin{figure}[htb]
\begin{center}
\includegraphics[width=0.95 \linewidth]{frosted}
\end{center}
\caption{Result after 80 rounds of training, with the input image at the top
and the signal proceeding downward through two hidden layers. The hidden layers
aspire to crazy noise-terror glitch art versions of the stimulus as well.} \label{fig:frosted}
\end{figure}

\begin{figure}[htb]
\begin{center}
\includegraphics[width=0.95 \linewidth]{instagram}
\end{center}
\caption{After about 9,000 rounds. Top row is the input image, the
second row is the output (no longer showing hidden layers because
they all just look like firefly raves); the bottom row shows 4x
magnified detail of the region formerly containing the red i.
Images are somewhat desaturated and blurry, but the red i is removed.
Note how in the right image, the retinal network successfully continued
both the horizontal and vertical bookshelves into the occluded region.
This is not a trick.} \label{fig:instagram}
\end{figure}

\begin{figure}[htb]
\begin{center}
\includegraphics[width=0.95 \linewidth]{eval}
\end{center}
\caption{Evaluation on new images after 30,000 rounds of training. Top row
is the input image, the second row is the output, and third is 4x magnified
defail. The first image (no i) shows the high amount of detail preserved.
In the latter two, the i is successfully removed; the quality of the replacement
is not perfect, but certainly reasonable.} \label{fig:eval}
\end{figure}

\begin{figure}[htb]
\begin{center}
\includegraphics[width=0.95 \linewidth]{jpeg}
\end{center}
\caption{Evaluation of an early model for J peg dequantization. The model
still contains a lot of noise pixels, which sometimes take a long time to
converge, but it is already easy to see how quantization artifacts have been reduced (left).
Actually, there is no reason why such dequantization must only be applied to J pegs;
the right column shows it working on a nice rainbow picture.
} \label{fig:jpeg}
\end{figure}

It's not long before the network learns that it should not produce the
same result for every input, and the output starts to mimic the
input. These images look sort of like the world viewed through frosted
glass (Figure~\ref{fig:frosted}), simulating how a baby first learns
to see the world through the 1cm bulletproof Lexan of its translucent
BabyLearn incubation cylinder.

Soon thereafter, the network begins to converge on something like the
identity function, as this drastically reduces error (even if some
error is incurred by the persistence of the red i). Left overnight
(about 9,000 rounds), we start to see the network both produce images
much like the original (perhaps through a hip ``vintage'' Instagram
filter), as well as removing the red i stimulus (Figure~\ref{fig:instagram}).

\subsection{Evaluation} \label{sec:evaluation}

With a further 30,000 rounds of training, the output images sharpen
and lose their Instagram quality (maybe only a small amount of
``grain''), and the i is still successfully removed. However, since
we've now made many passes over each image in the training set (and
the model has about 100,000 degrees of freedom), it is certainly
possible that we've simply overfit to this set of images (that is,
that the baby's eye's brain is simply memorizing the i-free images and
then recalling the one that looks closest to the stimulus). To evaluate
fairly, we need to apply the model to totally new images that it was
not trained on. These are called ``eval'' images.

Firing this up, I observed that the model successfully reproduced eval
images that did not contain a red i; this is good because it means
that it is not simply memorizing the training set images. I then
started placing red i stimulus on the images with the mouse, and my
heart sank: It wasn't removing the red i at all! Dejected, I tried
loading up the training images and putting a red i on them---it also
did not remove the i, which did not make sense! Even for babies!
Eventually, I discovered that the red i would be removed, as expected,
but only when the i was in a handful of very specific locations. This
was found to be a bug in the random i placing code; can you find it too?

\smallskip
\noindent \verb+        uint8 x_dice = seed & 0x255;+ \\
          \verb+        seed >>= 8;+ \\
          \verb+        uint8 y_dice = seed & 0x255;+
\smallskip

As a result, there are only 16 different possible $x$ coordinates, and same
for $y$. Nonetheless, this is still 256 different i locations that work,
which implies considerable generality is possible. Due to Draconian SIGBOVIK
deadlines, I have not yet been able to test a debugged training procedure.

Once the evaluation code only places an i at expected locations, the artificial
retinal network works well (Figure~\ref{fig:eval})!

\section{Conclusions}

We find that the supposedly impossible red i removal problem is in
fact solvable, at least in some forms, using artificial retinal
networks. There are some limitations of the current model:
\begin{itemize}
\item It has only been tested to remove the letter i when it is rendered
  in bright red, in 30 point Comic Sans.
\item It probably also removes letters like j, but maybe also in some
  other fonts, which is a sword that cuts both ways.
\item Due to a bug, the red i must be at a position whose coordinates are exactly
  $\langle 12 + x_0 + 4x_1 + 16x_2 + 64x_3, 12 + y_0 + 4y_1 + 16y_2 + 63y_3 \rangle$,
for $x_j$ and $y_j$ in $\{0, 1\}$.
\item It automatically and non-optionally applies Instagram-style filters.
\end{itemize}

This technique can probably be applied to other image processing
problems, for example, J peg dequantization. Here, we take an image
and badly quantize it (for example, to 4 bits per color channel), and
the training instance consists of the quantized image as input and the
original image as the expected output; the retinal network learns how
to fill in detail. Figure~\ref{fig:jpeg} shows the early stages (about
4000 rounds) of training such a model.

A related, still unsolved problem is ``red i reduction''; here we do
simply remove the i but replace it with a smaller i. For example, we
could replace a capital I with a lowercase one, or replace a lowercase
30pt Comic Sans i with a lowercase 29pt Comic Sans i. This is an
offshoot of the text ure compression field, which seeks to make the
text ``ure'' smaller wherever it appears.

\medskip
Biologically-inspired computer algorithms hold many wonders for those
that seek to tap into the limitless potential of the 85\% of the human
eye's brain that is currently unused. Perhaps humans even contain
graphics processing units!

\bigskip
For higher-fidelity images and source code, please consult {\tt
http://tom7.org/redi}.

\end{document}

### 2015-03-18 redi/jpeg.cc

@@ -205,6 +205,17 @@ static ImageA *FontChar(const string &filename, char c, int size) {
   return new ImageA(ret, width, height);
 }
 
+// Reduce to 4-bit color.
+static void Quantize(ImageRGBA *img) {
+  // three bit color
+  static constexpr uint8 MASK = 0xE0;
+  for (int i = 0; i < img->rgba.size(); i++) {
+    // Don't change alpha channel.
+    if (((i + 1) % 4) == 0) continue;
+    img->rgba[i] = MASK & img->rgba[i];
+  }
+}
+
 static void BlitChannel(uint8 r, uint8 g, uint8 b, const ImageA &channel, 
 			int xpos, int ypos,
 			ImageRGBA *dest) {
@@ -1230,10 +1241,10 @@ static void App(const vector<A> &vec, const F &f) {
   for (const auto &elt : vec) f(elt);
 }
 
-static constexpr int NUM_VIDEO_STIMULATIONS = 7;
+static constexpr int NUM_VIDEO_STIMULATIONS = 4;
 std::mutex video_export_m;
 int current_round = 0;
-vector<Stimulation> current_stimulations(7, Stimulation(NUM_LAYERS));
+vector<Stimulation> current_stimulations(NUM_VIDEO_STIMULATIONS, Stimulation(NUM_LAYERS));
 Network current_network(NUM_LAYERS);
 static bool allow_updates = true;
 
@@ -1375,7 +1386,7 @@ void UIThread() {
       // In this mode, draw the different nodes per pixel as separate rectangles.
       // We just show one example at a time.
       const int ex = offset % current_stimulations.size();
-      int show_channels = std::min(NPP, 7);
+      int show_channels = std::min(NPP, NUM_VIDEO_STIMULATIONS);
       for (int ch = 0; ch < show_channels; ch++) {
 	int sx = MARGIN + ch * (SIZE + GAP);
 	for (int layer = 0; layer < NUM_LAYERS + 1; layer++) {
@@ -1586,7 +1597,7 @@ void UIThread() {
       }
     } else {
       // PERF
-      SDL_Delay(1000);
+      SDL_Delay(10);
     }
   }
 }
@@ -1621,7 +1632,7 @@ void TrainThread() {
   // MakeIndices(&rc, &net);
   (void)RandomizeNetwork;
   (void)MakeIndices;
-  ReadNetworkBinary("net.val", &net);
+  ReadNetworkBinary("jnet.val", &net);
 
   printf("Invert index:\n");
[…]

### 2016-02-27 trunk/snacks/sim.js
Port of cc-lib's RandomGamma

@@ -51,6 +51,59 @@ var RandomGaussian = function() {
   };
 }();
 
+function RandomExponential() {
+  // n.b. JS Math.random() cannot return 1.0 exactly.
+  return -Math.log(1.0 - Math.random());
+}
+
+// Random variables from Gamma(s) distribution, for any
+// shape parameter (scale can be applied by just multiplying
+// the resulting variate by the scale).
+// Adapted from numpy, based on Marsaglia & Tsang's method.
+// Please see NUMPY.LICENSE.
+var RandomGamma = function(shape) {
+  if (shape == 1.0) {
+    return RandomExponential();
+  } else if (shape < 1.0) {
+    var one_over_shape = 1.0 / shape;
+    for (;;) {
+      var u = Math.random();
+      var v = RandomExponential();
+      if (u < 1.0 - shape) {
+	var x = Math.pow(u, one_over_shape);
+	if (x <= v) {
+	  return x;
+	}
+      } else {
+	var y = -Math.log((1.0 - u) / shape);
+	var x = Math.pow(1.0 - shape + shape * y, one_over_shape);
+	if (x <= v + y) {
+	  return x;
+	}
+      }
+    }
+  } else {
+    var b = shape - (1.0 / 3.0);
+    var c = 1.0 / Math.sqrt(9.0 * b);
+    for (;;) {
+      var x, v;
+      do {
+	x = RandomGaussian();
+	v = 1.0 + c * x;
+      } while (v <= 0.0);
+
+      var v_cubed = v * v * v;
+      var x_squared = x * x;
+      var u = Math.random();
+      if (u < 1.0 - 0.0331 * x_squared * x_squared ||
+	  Math.log(u) < 0.5 * x_squared + 
+	  b * (1.0 - v_cubed - Math.log(v_cubed))) {
+	return b * v_cubed;
+      }
+    }
+  }
+}
+
 // Possible policies:
 //  - Always sort all shelves (encountered) in reverse preference
 //    order (does this have to be paired with a selection method?)
@@ -80,6 +133,8 @@ var snacks = [];
 //   depth : int  - number of snacks skipped
 //   eaten : int  - number of snacks consumed
 //   utils : double - total number of snack delection points
+//   next : double - time until next craving. Can be thought of
+//     as seconds but should be scale invariant?
 //   prefs : array(array(double)) - preference function for varieties. 
 //     Outer array is length |snacks|; inner is length
[…]

### 2016-02-27 snacks/sim.js
Simple display, thoughts on snack selection mechanism, implementation of recurrence.

@@ -56,9 +56,10 @@ function RandomExponential() {
   return -Math.log(1.0 - Math.random());
 }
 
-// Random variables from Gamma(s) distribution, for any
-// shape parameter (scale can be applied by just multiplying
+// Random variables from Gamma(s) distribution, for any shape
+// parameter k (scale, aka theta, can be applied by just multiplying
 // the resulting variate by the scale).
+//
 // Adapted from numpy, based on Marsaglia & Tsang's method.
 // Please see NUMPY.LICENSE.
 var RandomGamma = function(shape) {
@@ -133,11 +134,18 @@ var snacks = [];
 //   depth : int  - number of snacks skipped
 //   eaten : int  - number of snacks consumed
 //   utils : double - total number of snack delection points
+//   mean_wait : double - mean time until craving for this person.
+//     Some people crave more often than others.
 //   next : double - time until next craving. Can be thought of
 //     as seconds but should be scale invariant?
 //   prefs : array(array(double)) - preference function for varieties. 
 //     Outer array is length |snacks|; inner is length
 //     |varieties of that snack|; elements are utility.
+//   exrec : array(array(double)) - Outer length is length |snacks|;
+//     inner is length MAX_ITEMS + 1. exrec[s][n] gives the value
+//     of the ExRec(N) function, which is the expected value of
+//     the snack we get for applying the commit-or-discard snack
+//     choosing procedure on N items for snack s.
 var people = [];
 
 // Generate a random preference function for n items. This is
@@ -163,6 +171,52 @@ function RandomPreferenceFn(n) {
   return pref;
 }
 
+// OK, so now we have the situation where the player has some
+// item in hand, with the given value. What we want to know is,
+// do I have a better expected value if I discard it and search
+// the rest? This should only depend on the number of remaining
+// items, N, and the player's preference function. Call this
+// expected value ExRec(N).
+
+// ExRec(0) is 0, because we get no item.
+//
+// ExRec(1) is the expected value of a single item, which (because
+// we assume for simplicity that each item is stocked in equal
+// numbers) is just the mean output value of our preference
+// function. (In fact, not really: we can always discard the snack,
+// so it's really the expectation over max(pref[k], 0) for each k.
[…]

### 2016-02-27 snacks/sim.js
Actually do snack selection, note potential bug

@@ -245,6 +245,7 @@ function Init() {
       stock.push(RandTo(varieties));
     }
 
+    // Bug: Soemtimes we get non-contiguous snacks this way?
     var rpref = RandomPreferenceFn(num_stocked);
     SortByPreference(stock, rpref);
     shelves[i] = stock;
@@ -278,6 +279,14 @@ function Init() {
   }
 
   Redraw();
+
+  // Loop();
+}
+
+function Loop() {
+  Step();
+  Redraw();
+  window.setTimeout(Loop, 10);
 }
 
 var COLORS = [
@@ -374,19 +383,23 @@ function Step() {
   // up to N times? (XXX todo)
 
   var shelf_idx = RandTo(shelves.length);
-  console.log('player ' + best_i + ' starts on shelf ' + shelf_idx);
   var shelf = shelves[shelf_idx];
   var prefs = player.prefs[shelf_idx];
   var exrec = player.exrec[shelf_idx];
+  console.log('player ' + best_i + ' starts on shelf ' + shelf_idx +
+	      ', which has ' + shelf.length + ' snacks of ' +
+	      prefs.length + ' varieties');
   if (shelf.length != 0) {
     var cur = 0;
     var cur_value = prefs[shelf[cur]];
     for (;;) {
       var num_left = shelf.length - cur;
       if (num_left > 0 &&
-	  cur_value > 0 &&
-	  cur_value > exrec[num_left]) {
+	  cur_value < exrec[num_left]) {
 	// swap for next one
+	console.log('Player ' + best_i + ' skips depth ' + cur +
+		    ' because E=' + exrec[num_left].toFixed(2) + 
+		    ' > cur=' + cur_value.toFixed(2));
 	cur++;
 	cur_value = prefs[shelf[cur]];
       } else {
@@ -396,15 +409,22 @@ function Step() {
 	  // next shelf here based on an estimate of its
 	  // value?
 	  shelf.splice(cur, 1);
+	  console.log('Player ' + best_i + ' goes ' + cur +
+		      ' snacks deep, and eats for ' + cur_value.toFixed(2) +
+		      ' utils.');
+	  player.eaten++;
+	  player.utils += cur_value;
+	  break;
 	} else {
-	  // TODO: Some chance of going to next shelf,
+	  console.log('Player ' + best_i + ' gets no snack.');
+	  // TODO: Some chance of going to next shelf--
 	  // if it's just a fixed probability it's really
 	  // no different than just waiting for the next
[…]

### 2016-02-28 snacks/sim.js
Fix bug, bug check code

@@ -227,6 +227,21 @@ function SortByPreference(stock, pref) {
   });
 }
 
+function CheckStock(s) {
+  if (s.length === 0) return;
+  var seen = {};
+  var prev = s[0];
+  for (var i = 0; i < s.length; i++) {
+    if (seen[s[i]]) {
+      throw('non-contiguous ' + s[i] + ' in: ' +
+	    s.join(','));
+    }
+    if (s[i] != prev) {
+      seen[prev] = true;
+    }
+  }
+}
+
 // Entry point.
 function Init() {
   for (var i = 0; i < NUM_SHELVES; i++) {
@@ -245,10 +260,21 @@ function Init() {
       stock.push(RandTo(varieties));
     }
 
-    // Bug: Soemtimes we get non-contiguous snacks this way?
-    var rpref = RandomPreferenceFn(num_stocked);
+    // Bug: Sometimes we get non-contiguous snacks this way?
+    var rpref = RandomPreferenceFn(varieties);
     SortByPreference(stock, rpref);
     shelves[i] = stock;
+
+    // Sanity check
+    try { 
+      CheckStock(stock);
+    } catch (e) {
+      console.log('ns: ' + num_stocked + ' len: ' + stock.length + 
+		  ' pref: ' + rpref.length);
+      console.log(rpref);
+      console.log(e);
+      throw e;
+    }
   }
   
   for (var i = 0; i < NUM_PEOPLE; i++) {

### 2016-02-28 trunk/snacks/sim.js
Pull out into separate function so that player can also advance shelves

@@ -227,7 +227,7 @@ function SortByPreference(stock, pref) {
   });
 }
 
-function CheckStock(s) {
+function CheckStockContiguous(s) {
   if (s.length === 0) return;
   var seen = {};
   var prev = s[0];
@@ -248,33 +248,36 @@ function Init() {
     var varieties = RandTo(MAX_VARIETY) + 1;
     snacks.push({ varieties: varieties });
     var num_stocked = RandTo(MAX_ITEMS - MIN_ITEMS) + MIN_ITEMS;
-    // We stock each item uniformly from the varieties, but in
-    // contiguous order. One easy way to do that is to generate a
-    // random preference function and sort with it.
-    //
-    // XXX, actually, I think maybe these should just be random,
-    // since players behave as though they are, so we have some kind
-    // of mismatched realism here
     var stock = [];
     for (var s = 0; s < num_stocked; s++) {
       stock.push(RandTo(varieties));
     }
 
-    // Bug: Sometimes we get non-contiguous snacks this way?
-    var rpref = RandomPreferenceFn(varieties);
-    SortByPreference(stock, rpref);
-    shelves[i] = stock;
+    // We stock each item uniformly from the varieties, but in
+    // contiguous order. One easy way to do that is to generate a
+    // random preference function and sort with it.
 
-    // Sanity check
-    try { 
-      CheckStock(stock);
-    } catch (e) {
-      console.log('ns: ' + num_stocked + ' len: ' + stock.length + 
-		  ' pref: ' + rpref.length);
-      console.log(rpref);
-      console.log(e);
-      throw e;
-    }
+    /*
+      // Actually, I think maybe these should just be random,
+      // since players behave as though they are, so we have some kind
+      // of mismatched realism here
+
+      // Bug: Sometimes we get non-contiguous snacks this way?
+      var rpref = RandomPreferenceFn(varieties);
+      SortByPreference(stock, rpref);
+
+      // Sanity check
+      try { 
+	CheckStockContiguous(stock);
+      } catch (e) {
+	console.log('ns: ' + num_stocked + ' len: ' + stock.length + 
+		    ' pref: ' + rpref.length);
+	console.log(rpref);
+	console.log(e);
+	throw e;
+      }
+    */
+    shelves[i] = stock;
   }
   
[…]

### 2016-03-06 trunk/snacks/paper.tex
More snack sorting policies, notes for paper

\documentclass[twocolumn]{article}
\usepackage[top=1.1in, left=0.85in, right=0.85in]{geometry}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{code}
\usepackage{graphicx}
\usepackage{fancyvrb}
\usepackage{url}
\usepackage{textcomp}

\pagestyle{empty}

\newcommand\comment[1]{}

\newcommand\st{$^{\mathrm{st}}$}
\newcommand\nd{$^{\mathrm{nd}}$}
\newcommand\rd{$^{\mathrm{rd}}$}
\renewcommand\th{$^{\mathrm{th}}$}
\newcommand\tm{$^{\mbox{\tiny \textsc{tm}}}$}

% nice fractions
\newcommand\sfrac[2]{{}\,$^{#1}$\!/{}\!$_{#2}$}

\newcommand\citef[1]{\addtocounter{footnote}{1}\footnotetext{\cite{#1}}\ensuremath{^{\mbox{\footnotesize [\thefootnote]}}}}

\usepackage{ulem}
% go back to italics for emphasis, though
\normalem

\begin{document}

\title{Snacking policies}
\author{Dr.~Tom~Murphy~VII~Ph.D.\thanks{
    Copyright \copyright\ 2016 the Regents of the Wikiplia Foundation.
    Appears in SIGBOVIK 2016 with the haunting memory of the
    Association for Computational Heresy; {\em IEEEEEE!} press,
    Verlag--Verlag volume no.~0x2016. ANG 0.00} }

\renewcommand\>{$>$}
\newcommand\<{$<$}

\date{1 April 2016}

\maketitle

\begin{abstract}
% XXX better abstract
what do I put here
\end{abstract}

\vspace{1em}
{\noindent \small {\bf Keywords}:
  snacking, consequentialism, comestability theory
}

\section{Introduction}

Some workplaces of the future

308000 rounds
Without reordering, utils: [589.02, 780.55] mean 684.87

Player 0 always sorts from the shelf he takes:

% 315000 rounds
% utils: [589.57, 792.09] mean 684.96
% p0: [9.11, 36.71] mean 22.59
% ineq: [20.34, 41.60] mean 30.60

total: 684.73 mean, 585.76--781.39
p0: 22.58, 8.82--36.77
ineq: 30.59, 19.93--41.84

No resorting:
total: 685.02 mean, 586.49--781.99
p0: 22.86, 9.50--37.97
ineq: 30.60, 19.92--41.79

Everyone resorts:
total: 684.46 mean, 586.26 -- 780.25
p0: 22.83 mean, 8.96-37.40
ieq: 30.55 19.94--41.27

--------------------

with 0.05 cost to look

total utils: 594.49 mean, 509.82-681.81
p0: 19.83 mean, 7.22-33.24
ieq: 28.07, 18.30-38.49

only p0 sorts:
total utils: 585.52 mean, 499-669
p0: 18.65 7.07-31.26
ieq: 27.92 18.33-38.38

Everyone puts favorite last:
total: 546.61 [466.36, 626.97]
p0: 18.24 [6.39, 31.27]
ieq: 26.92 [17.73, 37.05]

--------------------------------------------------
Now using 100k iterations. everyone puts snack last:

snacks at end × 100000 = Mean: 201.47 [116.62 , 443.72] Median: 208.81
total utils × 100000 = Mean: 546.75 [466.20 , 625.91] Median: 539.81
p0 utils × 100000 = Mean: 18.24 [6.06 , 30.91] Median: 16.52
inequality × 100000 = Mean: 26.91 [17.90 , 37.38] Median: 25.14

nobody sorts:
snacks at end × 100000 = Mean: 201.65 [116.65 , 421.74] Median: 217.99
total utils × 100000 = Mean: 594.17 [508.66 , 680.64] Median: 590.98
p0 utils × 100000 = Mean: 19.82 [7.52 , 33.50] Median: 19.63
inequality × 100000 = Mean: 28.05 [18.23 , 38.42] Median: 26.23

we eat almost exactly the same number of snacks, but are less happy.

nobody sorts; measure snack utils specifically:

snacks at end × 100000 = Mean: 201.26 [116.59 , 417.74] Median: 203.90
total utils × 100000 = Mean: 594.45 [506.68 , 679.40] Median: 588.88
p0 utils × 100000 = Mean: 19.83 [7.53 , 33.60] Median: 17.61
p0 snack utils × 100000 = Mean: 22.30 [8.49 , 36.57] Median: 20.62
inequality × 100000 = Mean: 28.05 [18.31 , 38.37] Median: 26.00

only p0 sorts:

snacks at end × 100000 = Mean: 201.52 [117.71 , 421.74] Median: 198.99
total utils × 100000 = Mean: 590.13 [507.03 , 678.05] Median: 590.58
p0 utils × 100000 = Mean: 18.98 [7.34 , 31.92] Median: 18.44
p0 snack utils × 100000 = Mean: 22.19 [9.17 , 37.02] Median: 20.87
inequality × 100000 = Mean: 27.99 [18.52 , 38.53] Median: 25.96

only p0 places outliers:

snacks at end × 100000 = Mean: 201.54 [117.79, 435.73] Median: 198.50
total utils × 100000 = Mean: 592.27 [509.32, 681.56] Median: 593.03
p0 utils × 100000 = Mean: 19.48 [7.14, 32.58] Median: 17.97
p0 snack utils × 100000 = Mean: 22.28 [8.53, 36.59] Median: 19.70
inequality × 100000 = Mean: 28.01 [18.51, 38.60] Median: 27.01

Credible interval\cite{krushke2014doing}

\end{document}

### 2016-03-06 trunk/snacks/sim.js

@@ -128,13 +128,17 @@ var SIMULATIONS_PER_FRAME = 100;
 //  - Only put your favorite snack at the back.
 
 // Each shelf holds a different item.
-var NUM_SHELVES = 10;
-var NUM_PEOPLE = 50;
+var NUM_SHELVES = 15;
+var NUM_PEOPLE = 30;
 // The number of varieties of a snack is uniform in [1, MAX_VARIETY]
 var MAX_VARIETY = 8;
 // TODO: Argument for minimum, based on plurality, etc.
 var MIN_ITEMS = 3;
 var MAX_ITEMS = 50;
+
+var OUTLIER_RATIO = 2.0;
+
+var COST_TO_LOOK = 0.05;
 
 // The current contents of the shelves: An ordered list of
 // variety ints.
@@ -158,6 +162,8 @@ var snacks = [];
 //   prefs : array(array(double)) - preference function for varieties. 
 //     Outer array is length |snacks|; inner is length
 //     |varieties of that snack|; elements are utility.
+//   favorite : variety of the favorite snack for each shelf
+//   outlier : true if the favorite snack is an outlier
 //   exrec : array(array(double)) - Outer length is length |snacks|;
 //     inner is length MAX_ITEMS + 1. exrec[s][n] gives the value
 //     of the ExRec(N) function, which is the expected value of
@@ -182,7 +188,7 @@ function RandomPreferenceFn(n) {
   var MEAN = 0.75;
   // Should certainly include negative values to indicate an
   // aversion to a snack.
-  var STDDEV = 1.0;
+  var STDDEV = 2.0;
   var pref = [];
   for (var i = 0; i < n; i++) {
     pref.push(RandomGaussian() * STDDEV + MEAN);
@@ -223,7 +229,8 @@ function MakeExRec(pref) {
   // ExRec[0] is 0, because we get no snack.
   var ret = [0];
   for (var i = 1; i <= MAX_ITEMS; i++) {
-    var ex_rec_next = ret[i - 1];
+    // Discounted because we have to do work to switch.
+    var ex_rec_next = ret[i - 1] - COST_TO_LOOK;
     var sum = 0.0;
     for (var s = 0; s < pref.length; s++) {
       sum += Math.max(ex_rec_next, pref[s]);
@@ -246,6 +253,26 @@ function SortByPreference(stock, pref) {
   });
 }
 
+function PutFavoriteLast(stock, fav) {
+  var dst = 0;
+  var n = 0;
[…]

### 2016-03-07 trunk/snacks/random.js
Major code reorganization, set up for experiments and hyperparameter search. Bibtex for paper

@@ -1,0 +1,89 @@
+// TODO: Clean up and start js-util with this.
+
+// Generate a uniformly distributed random integer in [0, n).
+var RandTo = function(n) {
+  // XXX to get uniform random numbers efficiently, generate
+  // random bit patterns, mod by the next power of two, then
+  // reject.
+  return Math.trunc(Math.random() * n);
+}
+
+// Generates two Gaussians at once, keeping state. Pretty much
+// a direct port of Marsaglia's pseudocode.
+var RandomGaussian = function() {
+  var have = false;
+  var next = 0.0;
+  return function() {
+    if (have) {
+      have = false;
+      return next;
+    } else {
+      var v1 = 0.0, v2 = 0.0, sqnorm = 0.0;
+      // Generate a non-degenerate random point in the unit circle by
+      // rejection sampling.
+      do {
+	v1 = 2.0 * Math.random() - 1.0;
+	v2 = 2.0 * Math.random() - 1.0;
+	sqnorm = v1 * v1 + v2 * v2;
+      } while (sqnorm >= 1.0 || sqnorm == 0.0);
+      var multiplier = Math.sqrt(-2.0 * Math.log(sqnorm) / sqnorm);
+      next = v2 * multiplier;
+      have = true;
+      return v1 * multiplier;
+    }
+  };
+}();
+
+function RandomExponential() {
+  // n.b. JS Math.random() cannot return 1.0 exactly.
+  return -Math.log(1.0 - Math.random());
+}
+
+// Random variables from Gamma(s) distribution, for any shape
+// parameter k (scale, aka theta, can be applied by just multiplying
+// the resulting variate by the scale).
+//
+// Adapted from numpy, based on Marsaglia & Tsang's method.
+// Please see NUMPY.LICENSE.
+var RandomGamma = function(shape) {
+  if (shape == 1.0) {
+    return RandomExponential();
+  } else if (shape < 1.0) {
+    var one_over_shape = 1.0 / shape;
+    for (;;) {
+      var u = Math.random();
+      var v = RandomExponential();
+      if (u < 1.0 - shape) {
+	var x = Math.pow(u, one_over_shape);
+	if (x <= v) {
+	  return x;
+	}
+      } else {
+	var y = -Math.log((1.0 - u) / shape);
[…]

### 2016-03-07 trunk/snacks/sim.js

@@ -9,7 +9,6 @@ document.onkeydown = function(e) {
 
   // ESC
   if (e.keyCode == 27)  {
-    // document.body.innerHTML = '(SILENCED. RELOAD TO PLAY)';
     EXIT = true;
   }
 }
@@ -20,7 +19,6 @@ function makeElement(what, cssclass, elt) {
   if (elt) elt.appendChild(e);
   return e;
 }
-function IMG(cssclass, elt) { return makeElement('IMG', cssclass, elt); }
 function DIV(cssclass, elt) { return makeElement('DIV', cssclass, elt); }
 function CANVAS(cssclass, elt) { return makeElement('CANVAS', cssclass, elt); }
 function SPAN(cssclass, elt) { return makeElement('SPAN', cssclass, elt); }
@@ -31,101 +29,11 @@ function TEXT(contents, elt) {
   return e;
 }
 
-// Generate a uniformly distributed random integer in [0, n).
-var RandTo = function(n) {
-  // XXX to get uniform random numbers efficiently, generate
-  // random bit patterns, mod by the next power of two, then
-  // reject.
-  return Math.trunc(Math.random() * n);
-}
-
-// Generates two Gaussians at once, keeping state. Pretty much
-// a direct port of Marsaglia's pseudocode.
-var RandomGaussian = function() {
-  var have = false;
-  var next = 0.0;
-  return function() {
-    if (have) {
-      have = false;
-      return next;
-    } else {
-      var v1 = 0.0, v2 = 0.0, sqnorm = 0.0;
-      // Generate a non-degenerate random point in the unit circle by
-      // rejection sampling.
-      do {
-	v1 = 2.0 * Math.random() - 1.0;
-	v2 = 2.0 * Math.random() - 1.0;
-	sqnorm = v1 * v1 + v2 * v2;
-      } while (sqnorm >= 1.0 || sqnorm == 0.0);
-      var multiplier = Math.sqrt(-2.0 * Math.log(sqnorm) / sqnorm);
-      next = v2 * multiplier;
-      have = true;
-      return v1 * multiplier;
-    }
-  };
-}();
-
-function RandomExponential() {
-  // n.b. JS Math.random() cannot return 1.0 exactly.
-  return -Math.log(1.0 - Math.random());
-}
-
-// Random variables from Gamma(s) distribution, for any shape
-// parameter k (scale, aka theta, can be applied by just multiplying
-// the resulting variate by the scale).
-//
[…]

### 2016-03-08 snacks/sim.js
Fixed bugs. hyperparam search

@@ -130,6 +130,8 @@ function SortByPreference(stock, pref) {
 }
 
 function PutFavoriteLast(stock, fav) {
+  var startlen = stock.length;
+  var sorted = stock.slice();
   var dst = 0;
   var n = 0;
   for (var src = 0; src < stock.length; src++) {
@@ -147,11 +149,19 @@ function PutFavoriteLast(stock, fav) {
     stock[dst] = fav;
     dst++;
   }
+  if (stock.length != startlen) throw 'NO';
+  var res = stock.slice();
+  sorted.sort(function(a, b) { return a - b; });
+  res.sort(function(a, b) { return a - b; });
+  for (var i = 0; i < res.length; i++) {
+    if (sorted[i] != res[i]) throw ('No: ' + sorted.join(',') + ' vs ' +
+				    res.join(','));
+  }
 }
 
 // Policies.
-var ALWAYS_SORT = 1;
 var NO_SORT = 0;
+var ALWAYS_SORT = 1;
 var FAVORITE_LAST = 2;
 var OUTLIER_LAST = 3;
 
@@ -177,7 +187,12 @@ function CloneSimulation(s) {
   for (var i = 0; i < s.people.length; i++) {
     p.push(ClonePerson(s.people[i]));
   }
-  return { shelves: s.shelves.slice(),
+  var h = [];
+  for (var i = 0; i < s.shelves.length; i++) {
+    h.push(s.shelves[i].slice());
+  }
+  return { shelves: h,
+	   // XXX deep clone objects.
 	   snacks: s.snacks.slice(),
 	   people: p,
 	   time_left: s.time_left };
@@ -264,7 +279,7 @@ function NewSimulation() {
       
       exrecs.push(MakeExRec(pref));
     }
-
+    
     people.push({
       policy: NO_SORT,
       eaten: 0,
@@ -296,7 +311,18 @@ function NewStats() {
 }
 
 function NewTask() {
-  // XXX hyper-tune parameters
+  // hyper-tune parameters, globally
+  NUM_SHELVES = RandTo(20) + 2;
+  NUM_PEOPLE = RandTo(40) + 2;
+  MAX_VARIETY = RandTo(12) + 2;
+  MIN_ITEMS = 3;
+  MAX_ITEMS = 5 + RandTo(75);
+  OUTLIER_RATIO = 1.0 + Math.random() * 2.5;
[…]

### 2016-03-08 trunk/snacks/sim.js
Keep track of best. Now search for two positive ratios

@@ -300,7 +300,7 @@ function NewSimulation() {
 
 function NewStats() {
   return { snacks_left: [],
-	   // Total happiness of all players.
+	   // Total (snack) happiness of all players.
 	   total_utils: [],
 	   // Total happiness of player 0, who may have a selfish policy.
 	   p0_utils: [],
@@ -323,6 +323,21 @@ function NewTask() {
   PREF_MEAN = RandomGamma(2.0) * 0.75;
   PREF_STDDEV = RandomGamma(2.0) * 2.0;
 
+  // Half the time, average with the current best.
+  // This spends more time searching near local maxima.
+  if (Math.random() < 0.5 && best.num_shelves != undefined) {
+    NUM_SHELVES = Math.round((NUM_SHELVES + best.num_shelves) / 2);
+    NUM_PEOPLE = Math.round((NUM_PEOPLE + best.num_people) / 2);
+    MAX_VARIETY = Math.round((MAX_VARIETY + best.max_variety) / 2);
+    MIN_ITEMS = Math.round((MIN_ITEMS + best.min_items) / 2);
+    MAX_ITEMS = Math.round((MAX_ITEMS + best.max_items) / 2);
+    OUTLIER_RATIO = (OUTLIER_RATIO + best.outlier_ratio) / 2;
+    COST_TO_LOOK = (COST_TO_LOOK + best.cost_to_look) / 2;
+    MEAN_WAIT = (MEAN_WAIT + best.mean_wait) / 2;
+    PREF_MEAN = (PREF_MEAN + best.pref_mean) / 2;
+    PREF_STDDEV = (PREF_STDDEV + best.pref_stddev) / 2;
+  }
+  
   return {
     experiment_sims_left: SIMULATIONS_PER_EXPERIMENT,
     cstats: NewStats(),
@@ -350,16 +365,16 @@ function GetStats(sim) {
   for (var s = 0; s < sim.shelves.length; s++)
     snacks_left += sim.shelves[s].length;
 
-  var utils = 0;
-  var min = sim.people[0].utils, max = sim.people[0].utils;
+  var snack_utils = 0;
+  var min = sim.people[0].snack_utils, max = sim.people[0].snack_utils;
   for (var p = 0; p < sim.people.length; p++) {
-    utils += sim.people[p].utils;
-    min = Math.min(min, sim.people[p].utils);
[…]

### 2016-03-09 trunk/snacks/paper.tex
Some real progress on paper, add license

@@
 
+\usepackage{relsize}
 \usepackage{amsmath}
 \usepackage{amssymb}
-\usepackage{code}
+% \usepackage{code}
 \usepackage{graphicx}
@@
 
-Some workplaces of the future
+Some workplaces of the future now offer a feature called Snacks.
+With Snacks, a kitchenette is provided near workers, which supplies running water and an array of small food packs.
+These foods are free of (dollar) charge, but various spoken and unspoken rules govern worker interactions with the foods.
+
+This presents a challenge, since some foods are more desirable than others.
+Specifically, say that one yogurt food comes in four varieties: Classic, Diet, Cherry-Vanilla, and Caffeine Free.
+Furthermore, say that one user \verb+<3+'s Cherry-Vanilla variety yogurt food and \verb+>=3+'s Caffeine Free variety yogurt food.
+This worker is happiest when he begins his office work while eating Cherry-Vanilla.
+If there is no Cherry-Vanilla, the worker may resort to Classic yogurt food, reducing his task-ready disposition, and thus performance.
+If all flavors have been exhausted but Caffeine Free, then the worker may take no yogurt at all, knowing that Caffeine Free brings more displeasure than even hunger.
+This creates a negative affect, which may cause the worker to actively sabotage the work of his peers.
+This provides poor Return on Investment (ROI).
+
+One strategy is for this worker, who we will call Sal, to take all of the favored Cherry-Vanilla foods from the kitchenette to his desk at the beginning of the day.
+This strategy is called hoarding.
+This ensures that Sal may eat all his Cherry-Vanilla flavors.
+However, this behavior is considered unfair, for one reason that all other workers are completely deprived of Cherry-Vanilla flavor.
+It is perceived that Sal and all other workers should have equivalent access to the shared resource, except for the moment that he is selecting a food (for he is "first in line", and lines are fair).
+Moreover, Sal should take only one food at a time (for it is "Please help yourself.
+We ask that you take only one piece so that others may enjoy it as well," which is fair).
+We also perceive that Sal should take food only when he is actually hungry (for it is "waste not, want not," which is fair).
+
+Given these rules, there are still things that Sal can do to influence the chance that he gets the Cherry-Vanilla flavors.
+In particular, this paper investigates the strategy of Reordering, where Sal selects his favored snack, and also changes the order of the foods in the kitchenette.
+The thought is that while everyone retains equivalent access to the foods, other workers are less likely to select Cherry-Vanilla due to the decreased visibility and/or increased effort in finding them.
+
+Note that the author does not reorder habitually reorder snacks; this question is of abstract philosophical interest. We consulted the wisdom of Judge John~Hodgman, who wrote~\cite{hodgman2016snacks}:
+
+\begin{quotation}
+  \noindent Why don't you just go out to lunch and buy the food you want with
+  your own money?
+
+  \noindent \ldots\ Stop with the personal e-mails and get back to work.
+\end{quotation}
+
+We did not find this to be a satisfactory argument.
+
+\smallskip
+In this paper, I first provide a short argument why Sal's behavior may be considered fair, using an unjustifiable but common assumption.
+I then give a formal model for Snacks, which can be used to conduct controlled experiments.
+I then show that under suitable conditions, Sal's behavior benefits both him and the workplace, in a Utilitarian sense.
+
+To investigate whether Reordering helps Sal and the rest of the workplace, we could run an experiment.
+Unfortunately this would be very expensive; we would need to find many work forces that are comparable, with similar food preferences, randomly assign some to the experimental group (where some fraction implement the Reordering policy) and then somehow judge their happiness.
+This would take a long time, and if the policy or experimental controls turn out to be harmful, might impact real GDP.
+For the effect sizes we see later, a live experiment is unlikely to show significance.
+
+Instead, we develop a simple model that captures important aspects of the Snacks program, implement this on a computer, and then run millions of simulations.
+
+The simplified model is as follows.
+
+A simulation consists of an array $S$ of shelves, each of which is stocked with different varieties of food.
+The varieties are just given as integers from 0--$n_i$, where each food type (shelf) may have a different number of varieties.
+In an early version of the simulation each food and variety is given a name, so we might have a shelf consisting of many sodas, like
+
+\begin{itemize}
+\item Diet Kake 
+\item Diet Thuck Lite
+\item Mango Slooch 
+\item Mr. Sleepe Black
+\item Caffeine-Free Droob Ultra
+\item Caffeine-Free Spask Black
+\item Dr. Drarb Classic
+\item Drorp Lite
+\item Diet Grobe
+\item Strawberry Sad 
+\item Vanilla Grerb
+\item Cherry Prote Lite
+\item Duq Ultra
+\item Dr. Brosh Lite
+\item Caffeine-Free Choq Black
+\item Grape Ding Lite
+\item Mrs. Broop 
+\item Diet Pap 
+\item Grape Drax Classic
+\end{itemize}
+
+We also have an array $W$ of workers.
+Each worker has a preference function $P_{ij}$, one floating point value for each snack variety.
+This value may be negative, indicating an aversion to that snack.
+Nominally, these values are in dollars, for scale.
+
+For simplicity, workers all get hungry at the same rate (although their hunger strikes randomly). When a worker hungers, she
+
+\begin{enumerate}
+\item Selects a shelf at random.
+\item Sets her gaze upon the foremost variety on that shelf. She has some value $v$ for this item, given by $P$. \label{step:start}
+\item She can see the number of items on the shelf, but not what varieties they are.
+From this number, she estimates what value $v'$ she would get from skipping the current variety (for this round) and setting her gaze upon the next one.
+\label{step:skip}
+\item If $v' > v$, she does so, and repeats from step \ref{step:skip}.
+\item If not, this is the food provisionally selected for this shelf, with value $v$.
+\item If there are shelves remaining, she estimates the value of abandoning this shelf and trying the next one, $v''$.
+\item If $v'' > v$, she moves to the next shelf and returns to step \ref{step:start}.
+\item If she finishes with a selected food, she may reorder the items on the shelf of that food arbitrarily. She removes the selected food, if any, and eats it.
+\end{enumerate}
+
+No player retains any knowledge of the organization of a shelf between rounds.\footnote{This is not an accurate assumption in reality, but seems to only disadvantage those who reorder snacks from benefiting from their own treachery.
+A very advanced strategy might rearrange items on the shelf in order to encode information about what has been reordered, for example, by coding a specific unlikely pattern at the front of a shelf (or prior shelf) to foreshadow the hidden booty.
+Ultra-advanced strategies might place misleading codes to confuse other workers and cause them to make suboptimal choices.
+Hyper-advanced strategies might use steganographic techniques or cryptographic signatures to hide codes or make them tamper-proof.
+Of course, this does not matter in a real workplace because workers can remember extremely simple facts themselves.}
+
+We have not yet said how the worker estimates the value of a shelf. But observe the following properties:
+
+\begin{itemize}
+\item If estimates are accurate, workers select a rational choice of food to maximize their own happiness.
+\item If a user has a dramatically favored snack, she is willing to search deep within a shelf for it.
+\item If a user has some snacks she favors and some she does not, she will be less willing to give up a good snack to find her favorite snack, because she might get stuck with a worse snack.
+\end{itemize}
+
+However, it also has an undesirable property:
+
+\begin{itemize}
+\item If a worker has a flat distribution of preferences, she will search the whole shelf.
+This is because there is no risk of getting stuck with a bad snack; she likes them all.
+This extends in a soft way to nearly flat distributions.
+\end{itemize}
+
+This does not match our intuitions of how real workers behave.
+Most of the time, an indifferent worker will just take a food that is "good enough"; this is known as "satisficing."\cite{simon1956rational} To prevent this, a worker's estimate of the value of continuing to search the shelf will include a small cost to search each item.
+This can be thought of as the cost of the physical labor or the displaced opportunity cost, or an estimate of the risk that an interruption causes her to have to stop searching before she selects a snack.
+
+\medskip
+The above requires an estimate of a shelf's value, both for the case where the worker may continue searching a shelf and the case where she continues to the next shelf.
+This can be computed with a recurrence relation.
+Since the worker cannot see beyond the snack her gaze is upon, this only depends on which shelf this is and the number of items on it.
+The expected value $E_s(n)$ for looking through $n$ items on shelf number $s$ is
+
+$$
+\begin{array}{c}
+E_s(0) = 0 \\
+E_s(n) = \mathlarger{\mathlarger{\sum}}_{i=1}^{V_s}\ \mathrm{P}(\mathrm{item}\ i) \times \mathrm{max}(P_{si}, E_s(n - 1) - c)
+\end{array}
+$$
+
+where $V_s$ is the number varieties for shelf $s$, $\mathrm{P}(\mathrm{item}\ i)$ is the probability of selecting variety $i$ in the next slot, $P_{si}$ is the worker's preference for variety $i$ from shelf $s$, and $c$ is the small cost of looking at all.
+The content of the recurrence is simple: At each step, for each possible item, the worker can either can take that item with the value given by the preference function $P$, or keep going (but now there will be one fewer item).
+
+Since we stipulate that the worker remembers nothing between rounds, the only probability distribution that makes sense for $\mathrm{P}(\mathrm{item}\ i)$ is the uniform one, so the general case becomes
+
+$$
+E_s(n) = \mathlarger{\mathlarger{\sum}}_{i=1}^{V_s}\ \frac{\mathrm{max}(P_{si}, E_s(n - 1) - c)}{V_s}
+$$
+
+Since this only depends on the preference function, we can compute this when the worker is born and print it on their birth certificate and employee badge.
+
+--------------------------------------------------------------------------------
 
@@
 
-Credible interval\cite{krushke2014doing}
+Credible interval\cite{kruschke2014doing}
+
+\bibliographystyle{IEEEtran}
+% \bibliographycomment{} % nothing
+\bibliography{paper}
 

### 2016-03-09 trunk/snacks/sim.js

@@ -325,7 +325,7 @@ function NewTask() {
 
   // Half the time, average with the current best.
   // This spends more time searching near local maxima.
-  if (Math.random() < 0.5 && best.num_shelves != undefined) {
+  while (Math.random() < 0.5 && best.num_shelves != undefined) {
     NUM_SHELVES = Math.round((NUM_SHELVES + best.num_shelves) / 2);
     NUM_PEOPLE = Math.round((NUM_PEOPLE + best.num_people) / 2);
     MAX_VARIETY = Math.round((MAX_VARIETY + best.max_variety) / 2);
@@ -418,20 +418,20 @@ function PolicyString(sim) {
 var task = null;
 
 var best = {
-  cost_to_look: 0.0654450536472723,
-  max_items: 17,
-  max_variety: 8,
-  mean_wait: 23638.810553936702,
+  cost_to_look: 0.0013740099966526031,
+  max_items: 37,
+  max_variety: 4,
+  mean_wait: 36305.715396286374,
   min_items: 3,
-  num_people: 20,
-  num_shelves: 3,
-  outlier_ratio: 3.134223625762388,
-  params: "NUM_SHELVES: 3, NUM_PEOPLE: 20, MAX_VARIETY: 8, MIN_ITEMS: 3, MAX_ITEMS: 17, OUTLIER_RATIO: 3.134223625762388, COST_TO_LOOK: 0.0654450536472723, MEAN_WAIT: 23638.810553936702, PREF_MEAN: 0.6997781433386749, PREF_STDDEV: 4.5844644380407145",
-  pref_mean: 0.6997781433386749,
-  pref_stddev: 4.5844644380407145,
-  ratio: 1.0247744252866777,
-  tratio: 1.0,
-  mratio: 1.0
+  mratio: 1.0106439712837496,
+  num_people: 2,
+  num_shelves: 15,
+  outlier_ratio: 2.1996847920818254,
+  params: "NUM_SHELVES: 15, NUM_PEOPLE: 2, MAX_VARIETY: 4, MIN_ITEMS: 3, MAX_ITEMS: 37, OUTLIER_RATIO: 2.1996847920818254, COST_TO_LOOK: 0.0013740099966526031, MEAN_WAIT: 36305.715396286374, PREF_MEAN: 1.3201387708679042, PREF_STDDEV: 5.625407250754956",
+  pref_mean: 1.3201387708679042,
+  pref_stddev: 5.625407250754956,
+  ratio: 1.02090727236298,
[…]

### 2016-03-09 trunk/snacks/paper.tex
almost done with this poo

@@
 
+\usepackage{amsthm}
 \usepackage{relsize}
@@
 However, this behavior is considered unfair, for one reason that all other workers are completely deprived of Cherry-Vanilla flavor.
-It is perceived that Sal and all other workers should have equivalent access to the shared resource, except for the moment that he is selecting a food (for he is "first in line", and lines are fair).
-Moreover, Sal should take only one food at a time (for it is "Please help yourself.
-We ask that you take only one piece so that others may enjoy it as well," which is fair).
-We also perceive that Sal should take food only when he is actually hungry (for it is "waste not, want not," which is fair).
+It is perceived that Sal and all other workers should have equivalent access to the shared resource, except for the moment that he is selecting a food (for he is ``first in line,'' and lines are fair).
+Moreover, Sal should take only one food at a time (for it is ``Please help yourself.
+We ask that you take only one piece so that others may enjoy it as well,'' which is fair).
+We also perceive that Sal should take food only when he is actually hungry (for it is ``waste not, want not,'' and you can't spell ``aphorism'' without ``a fair is,'' mmm?).
 
@@
 
+\subsection{Game of Scones}
+
 To investigate whether Reordering helps Sal and the rest of the workplace, we could run an experiment.
-Unfortunately this would be very expensive; we would need to find many work forces that are comparable, with similar food preferences, randomly assign some to the experimental group (where some fraction implement the Reordering policy) and then somehow judge their happiness.
+Unfortunately this would be very expensive; we would need to find many workforces that are comparable, with similar food preferences, randomly assign some to the experimental group (where some fraction implement the Reordering policy) and then somehow judge their happiness.
 This would take a long time, and if the policy or experimental controls turn out to be harmful, might impact real GDP.
@@
 The varieties are just given as integers from 0--$n_i$, where each food type (shelf) may have a different number of varieties.
-In an early version of the simulation each food and variety is given a name, so we might have a shelf consisting of many sodas, like
+In an early version of the simulation each food and variety is given a name, so we might have a shelf consisting of many sodas, like\footnote{All of these brands now Copyright Trademark Patent Pending Dr.~Tom Murphy VII Ph.D., 2016!}
 
@@
 \item Mr. Sleepe Black
-\item Caffeine-Free Droob Ultra
-\item Caffeine-Free Spask Black
+\item Caffeine--Free Droob Ultra
+\item Caffeine--Free Spask Black
 \item Dr. Drarb Classic
 \item Drorp Lite
-\item Diet Grobe
 \item Strawberry Sad 
@@
 \item Cherry Prote Lite
+\item Diet Grobe
+\item Dr. Brosh Lite
 \item Duq Ultra
-\item Dr. Brosh Lite
-\item Caffeine-Free Choq Black
 \item Grape Ding Lite
@@
 This does not match our intuitions of how real workers behave.
-Most of the time, an indifferent worker will just take a food that is "good enough"; this is known as "satisficing."\cite{simon1956rational} To prevent this, a worker's estimate of the value of continuing to search the shelf will include a small cost to search each item.
+Most of the time, an indifferent worker will just take a food that is ``good enough;'' this is known as ``satisficing.''\cite{simon1956rational} To prevent this, a worker's estimate of the value of continuing to search the shelf will include a small cost to search each item.
 This can be thought of as the cost of the physical labor or the displaced opportunity cost, or an estimate of the risk that an interruption causes her to have to stop searching before she selects a snack.
@@
 The expected value $E_s(n)$ for looking through $n$ items on shelf number $s$ is
-
+%
 $$
@@
 $$
-
+%
 where $V_s$ is the number varieties for shelf $s$, $\mathrm{P}(\mathrm{item}\ i)$ is the probability of selecting variety $i$ in the next slot, $P_{si}$ is the worker's preference for variety $i$ from shelf $s$, and $c$ is the small cost of looking at all.
@@
 
+\subsection{Experimenting}
+
+I implemented the rules above in about 1000 lines of JavaScript,
+including a UI, random number generation, and the experimentation harness.\footnote{It can be found at \url{sourceforge.net/p/tom7misc/svn/HEAD/tree/trunk/snacks/}.}
+
+Some other parameters need to be set: \verb+NUM_SHELVES+, \verb+NUM_PEOPLE+ and \verb+MAX_VARIETIES+, all self-explanatory; \verb+MIN_ITEMS+ and \verb+MAX_ITEMS+, the bounds on the number of snacks initially stocked per shelf (it seems that \verb+MIN+ should be at least 2, for the program is Snacks, plural); \verb+COST_TO_LOOK+, the penalty from the recurrence for estimating a shelf's value.
+We also have \verb+PREF_MEAN+ and \verb+PREF_STDDEV+ which are the parameters for the generation of preference functions (each is Gaussian---preferences are allowed to be negative).
+Finally, \verb+MEAN_WAIT+ gives the average amount of time between hunger events, wait times cannot be negative and so are distributed as $\Gamma(2.0, \verb+MEAN_WAIT+)$.
+We run each simulation for three eight-hour work days.
+
+I then generated random scenarios and evaluated different snack
+reordering policies. One extreme policy sorts the shelf in reverse
+preference order, so that the worker's favorite snacks are at the
+back. This seems unnecessary (takes $\mathrm{O}(n \mathrm{lg} n)$
+time, unless perhaps using ``radix sort'') and unrealistic. It also
+leads to substantial interference between workers, as one worker's
+resorting completely undoes another's. Another relaxation of this,
+more realistic, is where the worker only moves his favorite variety to
+the depths of the shelf. This is analogous to a ``Move-to-Back
+List''~\cite{rivest1976self}, except not really, since the worker
+somehow finds all of the instances of their favorite variety in the
+whole shelf before moving them to the back. The most realistic policy
+generalizes this last one, and only moves the favorite snack when it
+passes some ``outlier'' threshold for how much it is preferred over
+the other snacks. This is the policy I ran many experiments for.
+
+\medskip
+The first thing to notice from the experiments is that the policy
+doesn't even work! OMFG! For the reasonable parameter settings I
+started with, not only is the overall happiness (value of snacks
+eaten) neutral to slightly negative when the policy is in effect,
+but the worker employing the reordering strategy \emph{also eats
+  neutral to worse snacks}! This doesn't concur with anyone's
+perspective on the problem (except maybe John~Hodgman's)---at least
+the ``selfish'' worker who reorders snacks should be benefiting
+from it, right?
+
+Normally, this is where a heroic scientist would (a) implicate the
+model, seeing as how it fails to have intuitive properties\footnote{I
+  think that the model does not capture two important aspects. First
+  is that Sal \emph{does know} that he's reordered the snacks and can
+  likely find his favorites by just looking immediately to the back.
+  Second is that traces of the model suggest that most workers look
+  through nearly all of the snacks due to the (rational) expectation
+  that they will improve their selection. There does not seem to be a
+  good setting of the {\tt COST\_TO\_LOOK} penalty that suitably
+  discourages workers from looking through the whole shelf while still
+  allowing a resorter to find his deeply-placed favorites.} or (b)
+give up. But owing to Draconian SIGBOVIK deadlines, this heroic
+scientist resorts to another unjustifiable but common technique:
+``Tweaking'' the model parameters until the experiments turn out as
+expected (i.e., positive). Better yet, this process can be automated!
+
+\medskip
+Overnight in 4 separate browser tabs I ran a simple ``hyper-parameter
+search'' to try to find the best settings of the parameters described
+previously. I selected parameter values uniformly at random, and then
+interpolated against the current local maximum (by averaging the
+random point with the best point for each successive ``heads''
+coin-flip during intialization). The goal was to find parameters that
+simultaneously improved two scores in the experiment: The value of the
+snacks eaten by the resorting worker, and the total value of snacks
+eaten across all workers. (Specifically, I maximized the minimum of
+these two). The best settings of the model had these values:
+
+\begin{tabular}{@{\hspace{0.5in}}lr}
+\verb+COST_TO_LOOK+ & 0.001 \\
+\verb+MAX_ITEMS+ & 37 \\
+\verb+MAX_VARIETY+ & 4 \\
+\verb+MEAN_WAIT+ & 36305.715 \\
+\verb+MIN_ITEMS+ & 3 \\
+\verb+MRATIO+ & 1.010 \\
+\verb+NUM_PEOPLE+ & 2 \\
+\verb+NUM_SHELVES+ & 15 \\
+\verb+OUTLIER_RATIO+ & 2.199 \\
+\verb+PREF_MEAN+ & 1.320 \\
+\verb+PREF_STDDEV+ & 5.625 \\
+\end{tabular}
+
+This an interesting instance: There are only two workers, but lots of
+shelves and lots of snacks.
+Most interestingly, the mean hunger time is very unusually large (in the highest 97th percentile of the probability distribution): 36,305~seconds is over 10~hours!
+This means that in the course of a three-day simulation, we only expect each player to eat about 2.5~times.
+There is a very good chance that the workers never interact (never visit the same shelf) and more than a \sfrac{1}{4} chance that a worker never eats anything.
+
+Repeating many more simulations with these parameter values
+suggests---but does not prove---that this may just be a nearly
+degenerate case in the simulation where the policies of the workers do
+not matter, and what we are seeing is pure noise. However, when I
+stopped the simulation arbitrarily after 1 million rounds, it produced
+nice smooth-looking distributions, consistent with a good sample size
+(Figure~\ref{figure:smooth}). The player implementing the resorting
+policy ate 0.015\% better snacks on average, and the overall workplace
+ate 0.031\% better snacks on average! This truly is a victory for
+snack reordering and the scientific method! \qed
+
+\begin{figure*}[p]
+\begin{center}
+\includegraphics[width=3.5in]{utils}
+\end{center}\vspace{-0.1in}
+\caption{Histogram of one million experiments.} \label{figure:smooth}
+\smallskip
+The top group is the
+  control (no resorting) and the bottom is the experiment (worker 0
+  moves his favorite to the back of the shelf if its value exceeds the
+  outlier ratio). Note how smooth the simulated outcomes are. This
+  kind of graph is just the kind of statistics you write home about.
+  Within each group, the big lump is the total snacks eaten by the two
+  workers, and then the snacks eaten by each worker individually. Note
+  a dramatic peak at the value 0 for the two workers, corresponding to
+  {\it no snacks}; this happens because of the extreme hunger inteval
+  of 10~hours. A overall value of zero is very unlikely because the
+  simulation can only advance time (and thus end) when a worker has a
+  chance to eat. Each histogram shows its mean and median, as well as
+  darkening the 95\% highest density
+  interval~\cite{kruschke2014doing}. An advanced technique would
+  difference the values of the experiment and control and show a
+  distribution of that statistic, but again, Draconian SIGBOVIK page
+  size limitations preclude this. Normally we would use the 95\%
+  intervals to make the comparison between control and experiment, but
+  this appears to be unreadable due to the very smooth, nice looking
+  distribution occluding it. Therefore we compare the means, seeing
+  0.031\%, 0.015\%, and 0.046\% improvement in snack consumption
+  respectively for the three pairs. Note that resorting is actually
+  {\it altruistic} in this experiment, helping most the worker who does not
+  sort!
+
+  \smallskip
+
+  Without getting into too much math, regarding the frequentist
+  standard of statistical significance typically used, we can say with
+  confidence that $p > 0.05$.
+% } 
+\end{figure*}
+
+  \comment{
 --------------------------------------------------------------------------------
@@
 Credible interval\cite{kruschke2014doing}
+}
 

### 2016-03-09 trunk/snacks/onesim.js

@@ -31,23 +31,39 @@ function TEXT(contents, elt) {
 
 var simulation_num = 0;
 var SIMULATIONS_PER_FRAME = 100;
-var SIMULATIONS_PER_EXPERIMENT = 10000;
+var SIMULATIONS_PER_EXPERIMENT = 10000000;
 
 
+var best = {
+  cost_to_look: 0.0013740099966526031,
+  max_items: 37,
+  max_variety: 4,
+  mean_wait: 36305.715396286374,
+  min_items: 3,
+  mratio: 1.0106439712837496,
+  num_people: 2,
+  num_shelves: 15,
+  outlier_ratio: 2.1996847920818254,
+  params: "NUM_SHELVES: 15, NUM_PEOPLE: 2, MAX_VARIETY: 4, MIN_ITEMS: 3, MAX_ITEMS: 37, OUTLIER_RATIO: 2.1996847920818254, COST_TO_LOOK: 0.0013740099966526031, MEAN_WAIT: 36305.715396286374, PREF_MEAN: 1.3201387708679042, PREF_STDDEV: 5.625407250754956",
+  pref_mean: 1.3201387708679042,
+  pref_stddev: 5.625407250754956,
+  ratio: 1.02090727236298,
+  tratio: 1.0106439712837496
+};
 
 // Each shelf holds a different item.
-var NUM_SHELVES = 15;
-var NUM_PEOPLE = 30;
+var NUM_SHELVES = best.num_shelves;
+var NUM_PEOPLE = best.num_people;
 // The number of varieties of a snack is uniform in [1, MAX_VARIETY]
-var MAX_VARIETY = 8;
+var MAX_VARIETY = best.max_variety;
 // TODO: Argument for minimum, based on plurality, etc.
-var MIN_ITEMS = 3;
-var MAX_ITEMS = 50;
-var OUTLIER_RATIO = 2.0;
-var COST_TO_LOOK = 0.05;
+var MIN_ITEMS = best.min_items;
+var MAX_ITEMS = best.max_items;
+var OUTLIER_RATIO = best.outlier_ratio;
+var COST_TO_LOOK = best.cost_to_look;
 // This will be the average number of seconds between
 // snack events.
-var MEAN_WAIT = 2 * 60 * 60;
+var MEAN_WAIT = best.mean_wait;
 // Maybe should depend on snack type, like assuming that
 // people have similar values for "premium" snacks, based on
[…]

### 2016-03-10 trunk/snacks/paper.tex
First submitted version

@@
 
+\usepackage{supertabular}
 \usepackage{amsthm}
@@
 
-\title{Snacking policies}
+\title{Reordering snacks is effective and just}
 \author{Dr.~Tom~Murphy~VII~Ph.D.\thanks{
@@
 
-\begin{abstract}
-% XXX better abstract
-what do I put here
-\end{abstract}
+\section*{Concrete}
+It's time to get serious about {\em snacks} and {\em stats}.
 
@@
 {\noindent \small {\bf Keywords}:
-  snacking, consequentialism, comestability theory
+  snacking, permutations, comestability theory
 }
@@
 One strategy is for this worker, who we will call Sal, to take all of the favored Cherry-Vanilla foods from the kitchenette to his desk at the beginning of the day.
-This strategy is called hoarding.
+This strategy is called Hoarding.
 This ensures that Sal may eat all his Cherry-Vanilla flavors.
@@
 
-Note that the author does not reorder habitually reorder snacks; this question is of abstract philosophical interest. We consulted the wisdom of Judge John~Hodgman, who wrote~\cite{hodgman2016snacks}:
+Note that the author does not reorder habitually reorder snacks; this question is of abstract philosophical interest. We consulted the wisdom of Judge John~Hodgman, who ruled~\cite{hodgman2016snacks}:
 
@@
 
+\subsection{Why Reorder?}
+
+As suggested before, Reordering benefits Sal because his favorite snacks are less likely to be eaten by others.
+The policy can be thought of as fair because it still provides everyone equivalent access to the snacks.
+But did Sal make the snacks worse for everyone by putting the good snacks at the back?
+The problem seems intractible until we consider an unjustifiable but standard simplifying assumption: Each worker's preferences for snacks are independent and identically distributed.
+This means that Sal's preference for Cherry-Vanilla gives us no information about the rest of the workforce's preferences.
+Since all preferences are equally likely, reordering the snacks improves access to some snacks and hinders access to others; these effects cancel out.
+Thus under suitable assumptions, {\it any} reordering of snacks is fair and harmless to other workers.
+The reason why we believe that this particular policy is beneficial is that other workers may be indifferent among various flavors, and simply take the one that is most convenient.
+When indifferent, taking a variety that someone else substantially favors is globally sub-optimal.
+This is similar to the principle of ``decoy beer'', whereby cheap OK beer (e.g.
+Red Stripe) is placed at the front of the fridge in front of the premium good beer (which may be unsuitable for casual drinkers anyway).
+
+We can go further and demonstrate this scientifically.
+
 \subsection{Game of Scones}
@@
 
-A simulation consists of an array $S$ of shelves, each of which is stocked with different varieties of food.
-The varieties are just given as integers from 0--$n_i$, where each food type (shelf) may have a different number of varieties.
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[width=\linewidth]{shelves}
+\end{center}\vspace{-0.1in}
+\caption{Two shelves with snacks.
+Shelf~1 has four snacks; potentially different types of Y\"o!gurt brand yogurt.
+Shelf~2 has six snacks, potentially different types CHIPz brand natural particle chips.
+I can tell you that the first three on Shelf~1 are Cherry flavor, and the last one is the Diet Cherry variety.
+All of the CHIPz are different varieties: Dust flavor, Dirt flavor, Sticks flavor, Cobweb flavor, Sand flavor, and Powder flavor.
+But because of the way snacks are arranged, you can only see the variety of the snack in front.} \label{figure:shelves}
+\end{figure}
+
+A simulation consists of an array $S$ of shelves, each of which is stocked with different varieties of food (Figure~\ref{figure:shelves}).
+The varieties are just given as integers from 0--$V_i$, where each food type (shelf) may have a different number of varieties.
 In an early version of the simulation each food and variety is given a name, so we might have a shelf consisting of many sodas, like\footnote{All of these brands now Copyright Trademark Patent Pending Dr.~Tom Murphy VII Ph.D., 2016!}
@@
 \item If not, this is the food provisionally selected for this shelf, with value $v$.
-\item If there are shelves remaining, she estimates the value of abandoning this shelf and trying the next one, $v''$.
+\item If there are unvisited shelves remaining, she estimates the value of abandoning this shelf and trying the next one, $v''$.
 \item If $v'' > v$, she moves to the next shelf and returns to step \ref{step:start}.
@@
 This does not match our intuitions of how real workers behave.
-Most of the time, an indifferent worker will just take a food that is ``good enough;'' this is known as ``satisficing.''\cite{simon1956rational} To prevent this, a worker's estimate of the value of continuing to search the shelf will include a small cost to search each item.
+Most of the time, an indifferent worker will just take a food that is ``good enough;'' this is known as ``satisficing.''\cite{simon1956rational} The argument for the global value of Sal reordering hinges on such indifference, in fact.
+To prevent this, a worker's estimate of the value of continuing to search the shelf will include a small cost to search each item.
 This can be thought of as the cost of the physical labor or the displaced opportunity cost, or an estimate of the risk that an interruption causes her to have to stop searching before she selects a snack.
@@
 preference order, so that the worker's favorite snacks are at the
-back. This seems unnecessary (takes $\mathrm{O}(n \mathrm{lg} n)$
+back. This seems unnecessary (takes $\mathrm{O}(n\ \mathrm{lg}\ n)$
 time, unless perhaps using ``radix sort'') and unrealistic. It also
@@
   allowing a resorter to find his deeply-placed favorites.} or (b)
-give up. But owing to Draconian SIGBOVIK deadlines, this heroic
-scientist resorts to another unjustifiable but common technique:
-``Tweaking'' the model parameters until the experiments turn out as
-expected (i.e., positive). Better yet, this process can be automated!
+appeal that ``it seems more work is needed in this fertile area of
+study'' or (c) give up. But owing to Draconian SIGBOVIK deadlines,
+this heroic scientist resorts to another unjustifiable but common
+technique: ``Tweaking'' the model parameters until the experiments
+turn out as expected (i.e., positive). Better yet, this process can be
+automated!
 
 \medskip
-Overnight in 4 separate browser tabs I ran a simple ``hyper-parameter
+{\bf Search}.\ \ \ Overnight in 4 separate browser tabs I ran a simple ``hyper-parameter
 search'' to try to find the best settings of the parameters described
@@
 
-\begin{tabular}{@{\hspace{0.5in}}lr}
+\begin{supertabular}{@{\hspace{0.5in}}lr}
 \verb+COST_TO_LOOK+ & 0.001 \\
@@
 \verb+PREF_STDDEV+ & 5.625 \\
-\end{tabular}
+\end{supertabular}
 

### 2016-03-12 trunk/smeight/smeight.cc
Tom make many bug today   (-;

@@ -2,6 +2,7 @@
 #include <string>
 #include <set>
 #include <memory>
+#include <unordered_map>
 
 #include <cstdio>
 #include <cstdlib>
@@ -91,8 +92,10 @@ struct Tilemap {
       string key = Util::chop(line);
       if (key.empty() || key[0] == '#') continue;
       
-      if (key.size() != 2 || !Util::IsHexDigit(key[0]) || !Util::IsHexDigit(key[1])) {
-	printf("Bad key in tilemap. Should be exactly 2 hex digits: [%s]\n", key.c_str());
+      if (key.size() != 2 || !Util::IsHexDigit(key[0]) ||
+	  !Util::IsHexDigit(key[1])) {
+	printf("Bad key in tilemap. "
+	       "Should be exactly 2 hex digits: [%s]\n", key.c_str());
 	continue;
       }
 
@@ -165,7 +168,8 @@ struct SM {
 
     if (!moviefile.empty()) {
       inputs = SimpleFM2::ReadInputs(moviefile);
-      printf("There are %d inputs in %s.\n", (int)inputs.size(), moviefile.c_str());
+      printf("There are %d inputs in %s.\n",
+	     (int)inputs.size(), moviefile.c_str());
       const int warmup = atoi(config["warmup"].c_str());
       CHECK(inputs.size() >= warmup);
       // If we have warmup, advance to that point and discard them from inputs.
@@ -213,6 +217,13 @@ struct SM {
     
     vector<uint8> start_state = emu->SaveUncompressed();
 
+    // Just allocate the maximum size that this can ever be.
+    // It's only 16MB.
+    sprite_textures.resize(64);
+    for (int i = 0; i < 64; i++) {
+      sprite_textures[i].resize(256 * 256 * 4);
+    }
+    
     int current_angle = 0;
 
     bool paused = false;
@@ -344,20 +355,187 @@ struct SM {
     const uint8 ppu_ctrl2 = ppu->PPU_values[1];
     // Are sprites 16 pixels tall?
     const bool tall_sprites = !!(ppu_ctrl1 & (1 << 5));
-    printf("SPRITES ARE TALL.\n");
+    const int sprite_height = tall_sprites ? 16 : 8;
+    if (tall_sprites) {
+      printf("SPRITES ARE TALL %x.\n", ppu_ctrl1);
+    } else {
+      printf("8x8.\n");
+    }
 
[…]

### 2016-03-12 smeight/smeight.cc
Extracting and fusing sprite data. Try rendering next...

@@ -44,6 +44,11 @@ static_assert(TILEST >= TILESW, "TILEST must hold TILESW");
 static_assert(TILEST >= TILESW, "TILEST must hold TILESH");
 static_assert(0 == (TILEST & (TILEST - 1)), "TILEST must be power of 2");
 
+#define SPRTEXW 256
+#define SPRTEXH 256
+static_assert(0 == (SPRTEXW & (SPRTEXW - 1)), "SPRTEXW must be a power of 2");
+static_assert(0 == (SPRTEXH & (SPRTEXH - 1)), "SPRTEXH must be a power of 2");
+
 // I don't understand why Palette::FCEUD_GetPalette isn't working,
 // but the NES palette is basically constant (emphasis aside), so
 // let's just inline it to save time. RGB triplets.
@@ -68,6 +73,7 @@ static constexpr uint8 ntsc_palette[] = {
 
 // XXX
 static GLuint bg_texture = 0;
+static GLuint sprite_texture[64] = {};
 
 typedef void (APIENTRY *glWindowPos2i_t)(int, int);
 glWindowPos2i_t glWindowPos2i = nullptr;
@@ -219,9 +225,9 @@ struct SM {
 
     // Just allocate the maximum size that this can ever be.
     // It's only 16MB.
-    sprite_textures.resize(64);
+    sprite_rgba.resize(64);
     for (int i = 0; i < 64; i++) {
-      sprite_textures[i].resize(256 * 256 * 4);
+      sprite_rgba[i].resize(SPRTEXW * SPRTEXH * 4);
     }
     
     int current_angle = 0;
@@ -350,31 +356,31 @@ struct SM {
     vector<Sprite> ret;
 
     const PPU *ppu = emu->GetFC()->ppu;
-    const uint8 *palette_table = ppu->PALRAM;
+
     const uint8 ppu_ctrl1 = ppu->PPU_values[0];
     const uint8 ppu_ctrl2 = ppu->PPU_values[1];
     // Are sprites 16 pixels tall?
     const bool tall_sprites = !!(ppu_ctrl1 & (1 << 5));
     const int sprite_height = tall_sprites ? 16 : 8;
+    #if 0
     if (tall_sprites) {
       printf("SPRITES ARE TALL %x.\n", ppu_ctrl1);
     } else {
       printf("8x8.\n");
     }
-
+    #endif
+    
[…]

### 2016-03-12 smeight/smeight.cc
In-plane sprites sort of work. Need to get alpha blending, and some sprite data is wrong (flipping?)

@@ -49,6 +49,8 @@ static_assert(0 == (TILEST & (TILEST - 1)), "TILEST must be power of 2");
 static_assert(0 == (SPRTEXW & (SPRTEXW - 1)), "SPRTEXW must be a power of 2");
 static_assert(0 == (SPRTEXH & (SPRTEXH - 1)), "SPRTEXH must be a power of 2");
 
+#define BOX_DIM 2
+
 // I don't understand why Palette::FCEUD_GetPalette isn't working,
 // but the NES palette is basically constant (emphasis aside), so
 // let's just inline it to save time. RGB triplets.
@@ -70,6 +72,9 @@ static constexpr uint8 ntsc_palette[] = {
   0xFF,0xF7,0x9C, 0xD7,0xE8,0x95, 0xA6,0xED,0xAF, 0xA2,0xF2,0xDA,
   0x99,0xFF,0xFC, 0xDD,0xDD,0xDD, 0x11,0x11,0x11, 0x11,0x11,0x11,
 };
+
+static bool draw_sprites = true;
+static bool draw_boxes = true;
 
 // XXX
 static GLuint bg_texture = 0;
@@ -155,6 +160,8 @@ struct SM {
   Tilemap tilemap;
   
   SM() : rc("sm") {
+    InitTextures();
+
     map<string, string> config = Util::ReadFileToMap("config.txt");
     if (config.empty()) {
       fprintf(stderr, "Missing config.txt.\n");
@@ -205,13 +212,15 @@ struct SM {
   };
   
   struct Sprite {
-    // center of sprite -- different types will treat this
-    // location differently
+    // location of sprite -- different types will treat this
+    // differently
     Vec3 loc{0.0f, 0.0f, 0.0f};
     // Pixels.
-    int width, height;
-    int texture_x, texture_y;
-    SpriteType type;
+    int width = 0, height = 0;
+    int texture_x = 0, texture_y = 0;
+    // The texture id to use, in [0, 63].
+    int texture_num = 0;
+    SpriteType type = IN_PLANE;
   };
   
   void Loop() {
@@ -238,7 +247,16 @@ struct SM {
       printf("Start loop!\n");
       emu->LoadUncompressed(start_state);
 
-      for (int input_idx = 0; input_idx < inputs.size(); input_idx += !paused) {
+      #define START_IDX 550
+      if (START_IDX) {
+	printf("SKIP to %d\n", START_IDX);
[…]

### 2016-03-12 trunk/smeight/smeight.cc
Ok, that bug was quite shallow. Clean up code a little. Now needs transparency

@@ -384,13 +384,6 @@ struct SM {
     // Are sprites 16 pixels tall?
     const bool tall_sprites = !!(ppu_ctrl1 & (1 << 5));
     const int sprite_height = tall_sprites ? 16 : 8;
-    #if 0
-    if (tall_sprites) {
-      printf("SPRITES ARE TALL %x.\n", ppu_ctrl1);
-    } else {
-      printf("8x8.\n");
-    }
-    #endif
     
     const bool sprites_enabled = !!(ppu_ctrl2 && (1 << 4));  
     const bool sprites_clipped = !!(ppu_ctrl2 && (1 << 2));
@@ -434,15 +427,7 @@ struct SM {
     };
     vector<PreSprite> presprites{64};
 
-    auto GetRoot = [&presprites](int a) -> int {
-      if (0) {
-	printf("----------------\nGetroot(%d)\n", a);
-	for (int i = 0; i < presprites.size(); i++) {
-	  printf("%d -> %d   ", i, presprites[i].parent);
-	  if (i % 4 == 3) printf("\n");
-	}
-      }
-      
+    auto GetRoot = [&presprites](int a) -> int {      
       int root = a;
       while (presprites[root].parent != -1) {
 	CHECK(root != presprites[root].parent);
@@ -461,15 +446,6 @@ struct SM {
 	a = na;
       }
 
-      if (0 && changed) {
-	printf("Now on getroot(%d)\n", a);
-	for (int i = 0; i < presprites.size(); i++) {
-	  printf("%d -> %d   ", i, presprites[i].parent);
-	  if (i % 4 == 3) printf("\n");
-	}
-      }
-      
-      //      printf("Return: %d\n", root);
       return root;
     };
     
@@ -482,9 +458,9 @@ struct SM {
       // back to itself. Can't create cycles!
       if (ra == rb) return;
       CHECK(presprites[rb].parent == -1)
-      << rb << " -> " << presprites[rb].parent;
+          << rb << " -> " << presprites[rb].parent;
       CHECK(presprites[ra].parent == -1)
-      << ra << " -> " << presprites[ra].parent;
+          << ra << " -> " << presprites[ra].parent;
       presprites[rb].parent = ra;
     };
     
[…]

### 2016-03-12 trunk/smeight/smeight.cc
It works! GL_MODULATE was the key.

@@ -437,11 +437,9 @@ struct SM {
       CHECK(presprites[root].parent == -1);
       
       // Path compression.
-      bool changed = false;
       while (a != root) {
 	const int na = presprites[a].parent;
 	// printf("Path compression %d -> %d, new root %d\n", a, na, root);
-	changed = true;
 	presprites[a].parent = root;
 	a = na;
       }
@@ -619,7 +617,10 @@ struct SM {
 	    // alpha 0. The palette doesn't matter; 0 means transparent
 	    // in every palette.
 	    if (value == 0) {
-	      rgba[pixel + 3] = 0x00;
+	      rgba[pixel + 0] = rc.Byte();
+	      rgba[pixel + 1] = rc.Byte();
+	      rgba[pixel + 2] = rc.Byte();
+	      rgba[pixel + 3] = 0x20;
 	    } else {
 	      // Offset with palette table. Sprite palette entries come
 	      // after the bg ones, so add 0x10.
@@ -687,7 +688,6 @@ struct SM {
 			0, 0, SPRTEXW, SPRTEXH,
 			GL_RGBA, GL_UNSIGNED_BYTE,
 			sprite_rgba[n].data());
-
 	
 	// TODO: Decide on BILLBOARD vs IN_PLANE etc. Probably
 	// should not merge sprites of different types.
@@ -916,9 +916,13 @@ struct SM {
 #endif
 
     glEnable(GL_TEXTURE_2D);
-    glTexEnvf(GL_TEXTURE_ENV, GL_TEXTURE_ENV_MODE, GL_DECAL);
+    // glTexEnvi(GL_TEXTURE_ENV, GL_TEXTURE_ENV_MODE, GL_DECAL);
+    // GL_MODULATE is needed for alpha blending, not sure why.
+    glTexEnvi(GL_TEXTURE_ENV, GL_TEXTURE_ENV_MODE, GL_MODULATE);
    
     if (draw_boxes) {
+      glDisable(GL_BLEND);
+
       // One texture for the whole scene.
       glBindTexture(GL_TEXTURE_2D, bg_texture);
       glBegin(GL_TRIANGLES);
@@ -1007,6 +1011,11 @@ struct SM {
     
     // Now sprites.
     if (draw_sprites) {
+
+      glEnable(GL_BLEND);
+      // Should only be 1.0 or 0.0.
[…]

### 2016-03-13 smeight/smeight.cc
tom fix many bug    ;-)

@@ -36,6 +36,9 @@
 #define HEIGHT 1080
 static constexpr double ASPECT_RATIO = WIDTH / (double)HEIGHT;
 
+static constexpr float DEGREES_TO_RADS =
+  (1.0f / 360.0f) * 2.0f * 3.1415926535897932384626433832795f;
+
 // Size of NES nametable
 #define TILESW 32
 #define TILESH 30
@@ -241,13 +244,13 @@ struct SM {
     
     int current_angle = 0;
 
-    bool paused = false;
-    
+    bool paused = false, single_step = false;
+
     for (;;) {
       printf("Start loop!\n");
       emu->LoadUncompressed(start_state);
 
-      #define START_IDX 550
+      #define START_IDX 0
       if (START_IDX) {
 	printf("SKIP to %d\n", START_IDX);
 	paused = true;
@@ -256,7 +259,7 @@ struct SM {
 	}
       }
       
-      for (int input_idx = START_IDX; input_idx < inputs.size(); input_idx += !paused) {
+      for (int input_idx = START_IDX; input_idx < inputs.size(); ) {
 	const uint8 input = inputs[input_idx];
 	frame++;
 
@@ -267,6 +270,9 @@ struct SM {
 	    return;
 	  case SDL_KEYUP:
 	    switch (event.key.keysym.sym) {
+	    case SDLK_PERIOD:
+	      single_step = true;
+	      break;
 	    case SDLK_b:
 	      draw_boxes = !draw_boxes;
 	      break;
@@ -291,12 +297,20 @@ struct SM {
 	  }
 	}
 	  
-	SDL_Delay(1000.0 / 60.0);
+	// SDL_Delay(1000.0 / 60.0);
 
-	if (!paused) {
+	if (!paused || single_step) {
+	  if (single_step) {
+	    printf("Frame %d\n", input_idx);
+	    single_step = false;
+	  }
 	  emu->StepFull(input, 0);
+	  input_idx++;
 	}
 
+	// printf("---\n");
+	
+	// PERF
 	vector<uint8> image = emu->GetImageARGB();
 
 	glClear(GL_COLOR_BUFFER_BIT | GL_DEPTH_BUFFER_BIT);
@@ -306,6 +320,15 @@ struct SM {
 	vector<Box> boxes = GetBoxes();
 
 	vector<Sprite> sprites = GetSprites();
+
+	// XXX Make optional with key
+	#if 0
+	int z = 240;
[…]

### 2016-03-13 smeight/paper/paper.tex
add blank paper; depth sorting which maybe doesn't work

--- 2016-03-09 trunk/snacks/paper.tex
@@
 
+\usepackage{supertabular}
+\usepackage{amsthm}
 \usepackage{relsize}
@@
 
-\title{Snacking policies}
+\title{The glEnd() of Zelda}
 \author{Dr.~Tom~Murphy~VII~Ph.D.\thanks{
     Copyright \copyright\ 2016 the Regents of the Wikiplia Foundation.
-    Appears in SIGBOVIK 2016 with the haunting memory of the
+    Appears in SIGBOVIK 2016 with the danger of going alone of the
     Association for Computational Heresy; {\em IEEEEEE!} press,
-    Verlag--Verlag volume no.~0x2016. ANG 0.00} }
+    Verlag--Verlag volume no.~0x2016. 255 Rupees} }
 
@@
 
-\begin{abstract}
-% XXX better abstract
-what do I put here
-\end{abstract}
+\section*{Abstract}
+3D ZELDA
 
@@
 {\noindent \small {\bf Keywords}:
-  snacking, consequentialism, comestability theory
+  small keys, boss keys, dungeon keys
 }
@@
 
-Some workplaces of the future now offer a feature called Snacks.
-With Snacks, a kitchenette is provided near workers, which supplies running water and an array of small food packs.
-These foods are free of (dollar) charge, but various spoken and unspoken rules govern worker interactions with the foods.
-
-This presents a challenge, since some foods are more desirable than others.
-Specifically, say that one yogurt food comes in four varieties: Classic, Diet, Cherry-Vanilla, and Caffeine Free.
-Furthermore, say that one user \verb+<3+'s Cherry-Vanilla variety yogurt food and \verb+>=3+'s Caffeine Free variety yogurt food.
-This worker is happiest when he begins his office work while eating Cherry-Vanilla.
-If there is no Cherry-Vanilla, the worker may resort to Classic yogurt food, reducing his task-ready disposition, and thus performance.
-If all flavors have been exhausted but Caffeine Free, then the worker may take no yogurt at all, knowing that Caffeine Free brings more displeasure than even hunger.
-This creates a negative affect, which may cause the worker to actively sabotage the work of his peers.
-This provides poor Return on Investment (ROI).
-
-One strategy is for this worker, who we will call Sal, to take all of the favored Cherry-Vanilla foods from the kitchenette to his desk at the beginning of the day.
-This strategy is called hoarding.
-This ensures that Sal may eat all his Cherry-Vanilla flavors.
-However, this behavior is considered unfair, for one reason that all other workers are completely deprived of Cherry-Vanilla flavor.
-It is perceived that Sal and all other workers should have equivalent access to the shared resource, except for the moment that he is selecting a food (for he is "first in line", and lines are fair).
-Moreover, Sal should take only one food at a time (for it is "Please help yourself.
-We ask that you take only one piece so that others may enjoy it as well," which is fair).
-We also perceive that Sal should take food only when he is actually hungry (for it is "waste not, want not," which is fair).
-
-Given these rules, there are still things that Sal can do to influence the chance that he gets the Cherry-Vanilla flavors.
-In particular, this paper investigates the strategy of Reordering, where Sal selects his favored snack, and also changes the order of the foods in the kitchenette.
-The thought is that while everyone retains equivalent access to the foods, other workers are less likely to select Cherry-Vanilla due to the decreased visibility and/or increased effort in finding them.
-
-Note that the author does not reorder habitually reorder snacks; this question is of abstract philosophical interest. We consulted the wisdom of Judge John~Hodgman, who wrote~\cite{hodgman2016snacks}:
-
-\begin{quotation}
-  \noindent Why don't you just go out to lunch and buy the food you want with
-  your own money?
-
-  \noindent \ldots\ Stop with the personal e-mails and get back to work.
-\end{quotation}
-
-We did not find this to be a satisfactory argument.
-
+\begin{figure*}[p]
+\begin{center}
+\includegraphics[width=3.5in]{link3d}
+\end{center}\vspace{-0.1in}
+\caption{Histogram of one million experiments.} \label{figure:link3d}
 \smallskip
-In this paper, I first provide a short argument why Sal's behavior may be considered fair, using an unjustifiable but common assumption.
-I then give a formal model for Snacks, which can be used to conduct controlled experiments.
-I then show that under suitable conditions, Sal's behavior benefits both him and the workplace, in a Utilitarian sense.
-
-To investigate whether Reordering helps Sal and the rest of the workplace, we could run an experiment.
-Unfortunately this would be very expensive; we would need to find many work forces that are comparable, with similar food preferences, randomly assign some to the experimental group (where some fraction implement the Reordering policy) and then somehow judge their happiness.
-This would take a long time, and if the policy or experimental controls turn out to be harmful, might impact real GDP.
-For the effect sizes we see later, a live experiment is unlikely to show significance.
-
-Instead, we develop a simple model that captures important aspects of the Snacks program, implement this on a computer, and then run millions of simulations.
-
-The simplified model is as follows.
-
-A simulation consists of an array $S$ of shelves, each of which is stocked with different varieties of food.
-The varieties are just given as integers from 0--$n_i$, where each food type (shelf) may have a different number of varieties.
-In an early version of the simulation each food and variety is given a name, so we might have a shelf consisting of many sodas, like
-
-\begin{itemize}
-\item Diet Kake 
-\item Diet Thuck Lite
-\item Mango Slooch 
-\item Mr. Sleepe Black
-\item Caffeine-Free Droob Ultra
-\item Caffeine-Free Spask Black
-\item Dr. Drarb Classic
-\item Drorp Lite
-\item Diet Grobe
-\item Strawberry Sad 
-\item Vanilla Grerb
-\item Cherry Prote Lite
-\item Duq Ultra
-\item Dr. Brosh Lite
-\item Caffeine-Free Choq Black
-\item Grape Ding Lite
-\item Mrs. Broop 
-\item Diet Pap 
-\item Grape Drax Classic
-\end{itemize}
-
-We also have an array $W$ of workers.
-Each worker has a preference function $P_{ij}$, one floating point value for each snack variety.
-This value may be negative, indicating an aversion to that snack.
-Nominally, these values are in dollars, for scale.
-
-For simplicity, workers all get hungry at the same rate (although their hunger strikes randomly). When a worker hungers, she
-
-\begin{enumerate}
-\item Selects a shelf at random.
-\item Sets her gaze upon the foremost variety on that shelf. She has some value $v$ for this item, given by $P$. \label{step:start}
-\item She can see the number of items on the shelf, but not what varieties they are.
-From this number, she estimates what value $v'$ she would get from skipping the current variety (for this round) and setting her gaze upon the next one.
-\label{step:skip}
-\item If $v' > v$, she does so, and repeats from step \ref{step:skip}.
-\item If not, this is the food provisionally selected for this shelf, with value $v$.
-\item If there are shelves remaining, she estimates the value of abandoning this shelf and trying the next one, $v''$.
-\item If $v'' > v$, she moves to the next shelf and returns to step \ref{step:start}.
-\item If she finishes with a selected food, she may reorder the items on the shelf of that food arbitrarily. She removes the selected food, if any, and eats it.
-\end{enumerate}
-
-No player retains any knowledge of the organization of a shelf between rounds.\footnote{This is not an accurate assumption in reality, but seems to only disadvantage those who reorder snacks from benefiting from their own treachery.
-A very advanced strategy might rearrange items on the shelf in order to encode information about what has been reordered, for example, by coding a specific unlikely pattern at the front of a shelf (or prior shelf) to foreshadow the hidden booty.
-Ultra-advanced strategies might place misleading codes to confuse other workers and cause them to make suboptimal choices.
-Hyper-advanced strategies might use steganographic techniques or cryptographic signatures to hide codes or make them tamper-proof.
-Of course, this does not matter in a real workplace because workers can remember extremely simple facts themselves.}
-
-We have not yet said how the worker estimates the value of a shelf. But observe the following properties:
-
-\begin{itemize}
-\item If estimates are accurate, workers select a rational choice of food to maximize their own happiness.
-\item If a user has a dramatically favored snack, she is willing to search deep within a shelf for it.
-\item If a user has some snacks she favors and some she does not, she will be less willing to give up a good snack to find her favorite snack, because she might get stuck with a worse snack.
-\end{itemize}
-
-However, it also has an undesirable property:
-
-\begin{itemize}
-\item If a worker has a flat distribution of preferences, she will search the whole shelf.
-This is because there is no risk of getting stuck with a bad snack; she likes them all.
-This extends in a soft way to nearly flat distributions.
-\end{itemize}
-
-This does not match our intuitions of how real workers behave.
-Most of the time, an indifferent worker will just take a food that is "good enough"; this is known as "satisficing."\cite{simon1956rational} To prevent this, a worker's estimate of the value of continuing to search the shelf will include a small cost to search each item.
-This can be thought of as the cost of the physical labor or the displaced opportunity cost, or an estimate of the risk that an interruption causes her to have to stop searching before she selects a snack.
-
-\medskip
-The above requires an estimate of a shelf's value, both for the case where the worker may continue searching a shelf and the case where she continues to the next shelf.
-This can be computed with a recurrence relation.
-Since the worker cannot see beyond the snack her gaze is upon, this only depends on which shelf this is and the number of items on it.
-The expected value $E_s(n)$ for looking through $n$ items on shelf number $s$ is
-
-$$
-\begin{array}{c}
-E_s(0) = 0 \\
-E_s(n) = \mathlarger{\mathlarger{\sum}}_{i=1}^{V_s}\ \mathrm{P}(\mathrm{item}\ i) \times \mathrm{max}(P_{si}, E_s(n - 1) - c)
-\end{array}
-$$
-
-where $V_s$ is the number varieties for shelf $s$, $\mathrm{P}(\mathrm{item}\ i)$ is the probability of selecting variety $i$ in the next slot, $P_{si}$ is the worker's preference for variety $i$ from shelf $s$, and $c$ is the small cost of looking at all.
-The content of the recurrence is simple: At each step, for each possible item, the worker can either can take that item with the value given by the preference function $P$, or keep going (but now there will be one fewer item).
-
-Since we stipulate that the worker remembers nothing between rounds, the only probability distribution that makes sense for $\mathrm{P}(\mathrm{item}\ i)$ is the uniform one, so the general case becomes
-
-$$
-E_s(n) = \mathlarger{\mathlarger{\sum}}_{i=1}^{V_s}\ \frac{\mathrm{max}(P_{si}, E_s(n - 1) - c)}{V_s}
-$$
-
-Since this only depends on the preference function, we can compute this when the worker is born and print it on their birth certificate and employee badge.
-
---------------------------------------------------------------------------------
-
-308000 rounds
-Without reordering, utils: [589.02, 780.55] mean 684.87
-
-Player 0 always sorts from the shelf he takes:
-
-% 315000 rounds
-% utils: [589.57, 792.09] mean 684.96
-% p0: [9.11, 36.71] mean 22.59
-% ineq: [20.34, 41.60] mean 30.60
-
-total: 684.73 mean, 585.76--781.39
-p0: 22.58, 8.82--36.77
-ineq: 30.59, 19.93--41.84
-
-No resorting:
-total: 685.02 mean, 586.49--781.99
-p0: 22.86, 9.50--37.97
-ineq: 30.60, 19.92--41.79
-
-Everyone resorts:
-total: 684.46 mean, 586.26 -- 780.25
-p0: 22.83 mean, 8.96-37.40
-ieq: 30.55 19.94--41.27
-
---------------------
-
-with 0.05 cost to look
-
-total utils: 594.49 mean, 509.82-681.81
-p0: 19.83 mean, 7.22-33.24
-ieq: 28.07, 18.30-38.49
-
-only p0 sorts:
-total utils: 585.52 mean, 499-669
-p0: 18.65 7.07-31.26
-ieq: 27.92 18.33-38.38
-
-Everyone puts favorite last:
-total: 546.61 [466.36, 626.97]
-p0: 18.24 [6.39, 31.27]
-ieq: 26.92 [17.73, 37.05]
-
---------------------------------------------------
-Now using 100k iterations. everyone puts snack last:
-
-snacks at end × 100000 = Mean: 201.47 [116.62 , 443.72] Median: 208.81
-total utils × 100000 = Mean: 546.75 [466.20 , 625.91] Median: 539.81
-p0 utils × 100000 = Mean: 18.24 [6.06 , 30.91] Median: 16.52
-inequality × 100000 = Mean: 26.91 [17.90 , 37.38] Median: 25.14
-
-nobody sorts:
-snacks at end × 100000 = Mean: 201.65 [116.65 , 421.74] Median: 217.99
-total utils × 100000 = Mean: 594.17 [508.66 , 680.64] Median: 590.98
-p0 utils × 100000 = Mean: 19.82 [7.52 , 33.50] Median: 19.63
-inequality × 100000 = Mean: 28.05 [18.23 , 38.42] Median: 26.23
-
-we eat almost exactly the same number of snacks, but are less happy.
-
-nobody sorts; measure snack utils specifically:
-
-snacks at end × 100000 = Mean: 201.26 [116.59 , 417.74] Median: 203.90
-total utils × 100000 = Mean: 594.45 [506.68 , 679.40] Median: 588.88
-p0 utils × 100000 = Mean: 19.83 [7.53 , 33.60] Median: 17.61
-p0 snack utils × 100000 = Mean: 22.30 [8.49 , 36.57] Median: 20.62
-inequality × 100000 = Mean: 28.05 [18.31 , 38.37] Median: 26.00
-
-only p0 sorts:
-
-snacks at end × 100000 = Mean: 201.52 [117.71 , 421.74] Median: 198.99
-total utils × 100000 = Mean: 590.13 [507.03 , 678.05] Median: 590.58
-p0 utils × 100000 = Mean: 18.98 [7.34 , 31.92] Median: 18.44
-p0 snack utils × 100000 = Mean: 22.19 [9.17 , 37.02] Median: 20.87
-inequality × 100000 = Mean: 27.99 [18.52 , 38.53] Median: 25.96
-
-only p0 places outliers:
-
-snacks at end × 100000 = Mean: 201.54 [117.79, 435.73] Median: 198.50
-total utils × 100000 = Mean: 592.27 [509.32, 681.56] Median: 593.03
-p0 utils × 100000 = Mean: 19.48 [7.14, 32.58] Median: 17.97
-p0 snack utils × 100000 = Mean: 22.28 [8.53, 36.59] Median: 19.70
-inequality × 100000 = Mean: 28.01 [18.51, 38.60] Median: 27.01
-
-Credible interval\cite{kruschke2014doing}
+The top group is the
+% } 
+\end{figure*}
 

### 2016-03-13 smeight/matrices.h

@@ -82,6 +82,11 @@ inline Vec2 Vec2Plus(const Vec2 &a, const Vec2 &b) {
   return Vec2(a.x + b.x, a.y + b.y);
 }
 
+inline float Vec3Distance(const Vec3 &a, const Vec3 &b) {
+  float dx = b.x - a.x, dy = b.y - a.y, dz = b.z - a.z;
+  return sqrt(dx * dx + dy * dy + dz * dz);
+}
+
 inline Vec3 ScaleVec3(const Vec3 &v, float s) {
   return Vec3(v.x * s, v.y * s, v.x * s);
 }

### 2016-03-13 trunk/smeight/smeight.cc
add blank paper; depth sorting which maybe doesn't work

@@ -1,3 +1,4 @@
+#include <algorithm>
 #include <vector>
 #include <string>
 #include <set>
@@ -78,6 +79,7 @@ static constexpr uint8 ntsc_palette[] = {
 
 static bool draw_sprites = true;
 static bool draw_boxes = true;
+static bool draw_nes = true;
 
 // XXX
 static GLuint bg_texture = 0;
@@ -224,6 +226,8 @@ struct SM {
     // The texture id to use, in [0, 63].
     int texture_num = 0;
     SpriteType type = IN_PLANE;
+    // Used temporarily during rendering.
+    float distance = 0.0f;
   };
   
   void Loop() {
@@ -243,14 +247,17 @@ struct SM {
     }
     
     int current_angle = 0;
-
+    int angle_offset = 0;
+    
     bool paused = false, single_step = false;
 
     for (;;) {
       printf("Start loop!\n");
       emu->LoadUncompressed(start_state);
 
-      #define START_IDX 0
+      // still alpha bug?
+      // #define START_IDX 520
+      #define START_IDX 2000
       if (START_IDX) {
 	printf("SKIP to %d\n", START_IDX);
 	paused = true;
@@ -270,11 +277,20 @@ struct SM {
 	    return;
 	  case SDL_KEYUP:
 	    switch (event.key.keysym.sym) {
+	    case SDLK_LEFTBRACKET:
+	      angle_offset += 15;
+	      break;
+	    case SDLK_RIGHTBRACKET:
+	      angle_offset -= 15;
+	      break;
 	    case SDLK_PERIOD:
 	      single_step = true;
 	      break;
 	    case SDLK_b:
 	      draw_boxes = !draw_boxes;
+	      break;
+	    case SDLK_n:
+	      draw_nes = !draw_nes;
 	      break;
 	    case SDLK_SPACE:
 	      paused = !paused;
@@ -310,12 +326,13 @@ struct SM {
 
 	// printf("---\n");
 	
-	// PERF
-	vector<uint8> image = emu->GetImageARGB();
-
 	glClear(GL_COLOR_BUFFER_BIT | GL_DEPTH_BUFFER_BIT);
-	  
-	BlitNESGL(image, 0, HEIGHT);
+
+	if (draw_nes) {
+	  // PERF
+	  vector<uint8> image = emu->GetImageARGB();
+	  BlitNESGL(image, 0, HEIGHT);
+	}
 	
 	vector<Box> boxes = GetBoxes();
 
@@ -333,6 +350,11 @@ struct SM {
[…]

### 2016-03-13 trunk/smeight/paper/paper.tex
start on paper

--- 2016-03-13 smeight/paper/paper.tex
@@
 
+{\bf 1986.
+Hyrule.}\quad The Legend of {\it frickin'}~Zelda for the Nintendo {\it freakin'}~Entertainment System.
+Need I say more?
+A {\it god damn} \uline{gold cartridge}.
+Fortunes made just from saving the gold cartridge and melting it down to make gold teeth grilles, after carefully extracting the even more precious ROM inside.
+A die-cut hole in the box so that you could get a peek of the cartridge and presage that you were getting some solid gold.
+A die cut little window that you could palpate through the wrapping paper on Christmas~Eve, presaging some epic thumb blisters in store for the coming weeks.
+Koji {\it freckin'} Kondo.
+Koji Kondo whipping up a nice 8-bit arrangement of Bol\'ero as the theme music until realizing at the last minute that this music was copyrighted\footnote{Perhaps ironic, since Ravel's {Bol\'ero} itself was composed as a result of Ravel getting cop-blocked.\cite{wikipedia2016bolero} And surely some Zelda knock-off since then has included a Muzak ersatz of the Zelda theme!
+What is the longest documented sequence of compositions due to Copyright restrictions?} and so instead composing its epic theme in {\it one day}??
+
+A gold cartridge that contained ROMs and a little swallowable battery to keep the onboard SRAM powered up so that it could retain your epic save game.
+A battery designed to last 70~years.
+Nothing could cause you to lose your save game, even once you were half way through the Second Quest.
+Unless your little brother starts a completely new game and saves over your slot, earning him one of the most righteously deserved clobberings this side of Inigo Montoya.
+Saves right on top of your slot with a player called just \verb+A   +.
+Right over your slot, erasing it, and hasn't even picked up the {\it sword} yet.
+Saves right over yours even though there are TWO OTHER UNUSED SLOTS and you even wisely put your game in the third slot {\it exactly to avoid this kind of calamity}.
+
+Need I say more? Shall I say it?
+
+technical limitations...
+Like looking at tree people with polyhedra for hands is some kind of ...
+
 \begin{figure*}[p]
@@
 \end{center}\vspace{-0.1in}
-\caption{Histogram of one million experiments.} \label{figure:link3d}
+\caption{Technical diagram of mathematical equations.} \label{figure:link3d}
 \smallskip
-The top group is the
+% caption can go here...
 % } 
 \end{figure*}
+
+\nocite{murphy2013first}
 

### 2016-03-13 trunk/smeight/paper/paper.tex
progress on paper, ppu tutorial

@@
 
-{\bf 1986.
-Hyrule.}\quad The Legend of {\it frickin'}~Zelda for the Nintendo {\it freakin'}~Entertainment System.
-Need I say more?
-A {\it god damn} \uline{gold cartridge}.
-Fortunes made just from saving the gold cartridge and melting it down to make gold teeth grilles, after carefully extracting the even more precious ROM inside.
-A die-cut hole in the box so that you could get a peek of the cartridge and presage that you were getting some solid gold.
-A die cut little window that you could palpate through the wrapping paper on Christmas~Eve, presaging some epic thumb blisters in store for the coming weeks.
-Koji {\it freckin'} Kondo.
-Koji Kondo whipping up a nice 8-bit arrangement of Bol\'ero as the theme music until realizing at the last minute that this music was copyrighted\footnote{Perhaps ironic, since Ravel's {Bol\'ero} itself was composed as a result of Ravel getting cop-blocked.\cite{wikipedia2016bolero} And surely some Zelda knock-off since then has included a Muzak ersatz of the Zelda theme!
-What is the longest documented sequence of compositions due to Copyright restrictions?} and so instead composing its epic theme in {\it one day}??
+{\bf 1986. Hyrule.}\quad The Legend of {\it frickin'}~Zelda for the
+Nintendo {\it freakin'}~Entertainment System. Need I say more? A {\it
+  god damn} \uline{gold cartridge}. Fortunes made just from melting
+the gold cartridge down to make gold teeth grilles, after carefully
+extracting the even more precious ROM inside. A die-cut hole in the
+box so that you could get a peek of the cartridge and presage that you
+were getting some solid gold. A die cut little window that you could
+palpate through the wrapping paper on Christmas~Eve, presaging some
+epic thumb blisters in store for the coming weeks. Koji {\it freckin'}
+Kondo. Koji Kondo whipping up a nice 8-bit arrangement of Bol\'ero as
+the theme music until realizing at the last minute that this music was
+copyrighted\footnote{Perhaps ironic, since Ravel's {Bol\'ero} itself
+  was composed as a result of Ravel getting
+  cop-blocked.\cite{wikipedia2016bolero} And surely some Zelda
+  knock-off since then has included a Muzak ersatz of the Zelda theme!
+  What is the longest documented sequence of compositions due to
+  Copyright restrictions?} and so instead composing its epic theme in
+{\it one day}??
 
-A gold cartridge that contained ROMs and a little swallowable battery to keep the onboard SRAM powered up so that it could retain your epic save game.
-A battery designed to last 70~years.
-Nothing could cause you to lose your save game, even once you were half way through the Second Quest.
-Unless your little brother starts a completely new game and saves over your slot, earning him one of the most righteously deserved clobberings this side of Inigo Montoya.
-Saves right on top of your slot with a player called just \verb+A   +.
-Right over your slot, erasing it, and hasn't even picked up the {\it sword} yet.
-Saves right over yours even though there are TWO OTHER UNUSED SLOTS and you even wisely put your game in the third slot {\it exactly to avoid this kind of calamity}.
-
-Need I say more? Shall I say it?
-
-technical limitations...
-Like looking at tree people with polyhedra for hands is some kind of ...
-
-\begin{figure*}[p]
+\begin{figure}[ht]
 \begin{center}
-\includegraphics[width=3.5in]{link3d}
+\includegraphics[width=\linewidth]{link3d}
 \end{center}\vspace{-0.1in}
-\caption{Technical diagram of mathematical equations.} \label{figure:link3d}
-\smallskip
+\caption{Technical diagram of mathematical equations.} \label{fig:link3d}
+% \smallskip
 % caption can go here...
 % } 
-\end{figure*}
+\end{figure}
+
+A gold cartridge that contained ROMs and a little swallowable battery
+to keep the onboard SRAM powered up so that it could retain your epic
+save game. A battery designed to last 70~years. Nothing could cause
+you to lose your save game, even once you were half way through the
+Second Quest. Unless your little brother starts a completely new game
+and saves over your slot, earning him one of the most righteously
+deserved clobberings this side of Inigo Montoya. Saves right on top of
+your slot with a player called just \verb+A +. Right over your slot,
+erasing it, and hasn't even picked up the {\it sword} yet. All you
+have to do is step on the black square in the first room and the guy
+gives you the sword and then dies. Jeez pooleaze. Saves right over
+yours with no sword even though there are TWO OTHER UNUSED SLOTS and
+you even wisely put your game in the third slot {\it exactly to avoid
+  this kind of calamity}.
+
+\medskip
+Need I say more? Shall I say it? Apparently so. KIDS THESE DAYS. Kids
+these days and their three dimensions. You can certainly make a
+respectable case that the Zenith of the series was {\it Legend of
+  Zelda III: A Link To The Past}, or even {\it Legend of Game Boy
+  Zelda: I forgot What It's Called}. But there seems to be an
+established consensus that {\it Legend Of Zelda LXIV: The Ocarina of
+  Time} is the true masterpiece; that while the NES Original indeed
+invented the genre, the technical limitations of the system
+handicapped it. You're talking to me about technical {\it frockin'}
+limitations? Like somehow appreciating the subtle wonder of the
+universe and its mechanics requires me to look at tree people with
+tetrahedra clubs for hands. And they say this of the 64\th\ game in
+the series? Please jouize.
+
+A fine game, don't get me wrong. But was its theme song composed in
+one day? Did it come in a gold cartridge?
+
+And the thing is: The original Zelda was built with enough forethought
+that it {\em can} be rendered in first-person 3D. So here I will make
+the game 3D so that it can be enjoyed by these kids and they can get
+off my lawn, tapdancing around playing their titular Ocarinas. I will
+make it possible to play the game in glorious HD, or even 4K
+resolution, and then some (Figure~\ref{fig:link3d}). And while
+we're at it, why not try to do it in a way that can play loads of NES
+games in three {\it fr\"ockin'} dimensions? What's the worst that
+could happen?
+
+\section{Technical limitations}
+
+As presaged, the NES does have certain technical limitations, which I
+prefer to think of as constraints. Some basic details are important
+for understanding how this works.
+
+The NES has a modest CPU that executes game logic. It's a little 8-bit
+baby puppy with access to 2 kilobytes of RAM. If you think about it,
+the NES's 256x240 pixel screen is already 61,440 pixels, which if it
+used the entirety of RAM to represent it, we'd have only
+$0.2\overline{6}$ bits of RAM per pixel, which makes no sense. At XXX
+MhZ, and a generous 1 instruction per pixel, we'd only be able to render
+XXX frames per second! This doesn't make sense, and this is because
+the CPU does not render the graphics on a NES.
+
+The NES also contains a custom Picture Processing Unit---PPU---which
+outputs 60 frames of NTSC or PAL video per second. In fact, since the
+PPU also drives a CPU interrupt for each scanline and vblank (XXX
+true?), and most games use this for timing,%
+%
+\footnote{Fun fact: NTSC, used in the Americas, has a ~60Hz frame
+  rate. But PAL, used in Europe and Asia, has a 50Hz refresh rate.
+  This means that many games run 16\% slower in Europe than the US, so
+  if you grew up playing there, you were playing on {\em Easy} mode.
+  {\tt\ ;-)}}
+%
+in many ways the CPU is subservient to the PPU, kind of like the PPU
+is the {\it real} heart of the NES. In the context of this paper, you
+could say that the CPU is the Ocarina of Time and the PPU is the
+Legend of Zelda. It's just an analogy. The PPU is very complicated and
+has tetrahedra for hands, like clubs. But the CPU communicates with
+the PPU not by drawing pixels, but by giving a high-level description
+of the screen (by writing it into PPU memory). The chief insight of
+this work is that the PPU representation can be interpreted and
+rendered in 3D instead of 2D. This allows us to use the exact original
+game logic and only change the way it's viewed. Sometimes this even
+allows us to see aspects of the game that are actually present but
+were not visible on NES hardware. Often it makes the game more
+difficult by making things invisible (enemies behind the player or
+behind walls) or inducing motion sickness.
+
+\subsubsection{Background}
+The PPU has two major drawing facilities: Background and Sprites. Both
+are made of ``tiles'', which are 8x8 graphics that are
+usually\footnote{Some games trick the NES hardware by remapping tile
+memory (e.g. in response to a memory read), even during a scanline. I
+do not allow for such chicanery in this work.} stored in ROM. There
+are two tile sets, each with 256 tiles. Figure~\ref{fig:ppu} shows the
+tile sets for the starting location in the Zelda overworld.
+
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[width=\linewidth]{ppu1} \\
+\vspace{1em}
+\includegraphics[width=\linewidth]{ppu2}
+\end{center}\vspace{-0.1in}
+\caption{Pattern tables for the Zelda overworld. In Zelda, the top tiles
+are used for sprites, and the bottom are used for the background graphics.
+All tiles are 8x8 and use 4 color values (color 0 is transparent). These
+color values index into 4 sprite palettes and 4 background palettes, with
+some limitations.
+} \label{fig:ppu}
+\end{figure}
+
+The NES background is described by the ``nametable'', which is an
+array of 32x30 bytes giving the tile numbers to fill the screen with.
+It's two rows shy of 256 probably for some reason having to do with
+the aspect ratio of TVs or number of actual NTSC scanlines, but also
+to make room for 64 bytes of attribute data. There are just two bits
+per four tiles. These two bits select which of the four palettes to
+use for each 2x2 group of tiles on the screen. It's interesting to
+look for the tricks that artists use to work around the constraint on
+the number of background colors. % XXX figure showing this off?
+
+Because of this array of tiles, NES games have to be built of
+high-level, block-like structures. In Zelda, this structure is
+particularly (but not unusually) regular. Since each 2x2 tile group
+(16x16 pixels) shares the same palette, the screens are built around
+16x16 pixel grid-aligned blocks.
+
+\subsubsection{Sprites}
+
+Sprites, named in honor of the forest pixies whose homes were razed in
+order to make space inside the PPU for them, are more complicated.
+
+There are 64 sprites, which are all turned on or off together. Each
+sprite is described by four bytes: Its $x$ and $y$ coordinates (these
+can be placed at arbitrary integer pixel locations), the index of
+the tile to draw, and an attribute byte:
+
+\begin{tabular}{rrl}
+    & bit \# & bit deets \\
+\hline \\
+MSB & 7      & Vertical flip \\
+    & 6      & Horizontal flip \\
+    & 5      & something weird \\
+    & 4      & I don't know \\
+    & 3      & unused? who can say \\
+    & 2      & bit 2; could be anything \\
+LSB & 1,0    & Palette index \\
+\end{tabular}
+
+Additionally, there are some global sprite settings. But
+
+Sprite Zero has some special properties. I don't want to tell you
+about the properties, I just wanted to make a soda joke.
 

### 2016-03-14 smeight/paper/paper.tex
More in paper

@@
 
-\begin{figure*}[p]
+{\bf 1986. Hyrule.}\quad The Legend of {\it frickin'}~Zelda for the
+Nintendo {\it freakin'}~Entertainment System. Need I say more? A {\it
+  god damn} \uline{gold cartridge}. Fortunes made just from melting
+the gold cartridge down to make gold teeth grilles, after carefully
+extracting and preserving the even more precious ROM inside. A die-cut hole in the
+box so that you could get a peek of the cartridge and presage that you
+were getting some solid gold. A die cut little window that you could
+palpate through the wrapping paper on Christmas~Eve, presaging some
+epic thumb blisters in store for the coming weeks. Koji {\it freckin'}
+Kondo. Koji Kondo whipping up a nice 8-bit arrangement of Bol\'ero as
+the theme music until realizing at the last minute that this music was
+copyrighted\footnote{Perhaps ironic, since Ravel's {Bol\'ero} itself
+  was composed as a result of Ravel getting
+  cop-blocked.\cite{wikipedia2016bolero} And surely some Zelda
+  knock-off since then has included a Muzak ersatz of the Zelda theme!
+  What is the longest documented sequence of compositions due to
+  Copyright restrictions?} and so instead composing its epic theme in
+{\it one day}??
+
+\begin{figure}[ht]
 \begin{center}
-\includegraphics[width=3.5in]{link3d}
+\includegraphics[width=\linewidth]{link3d}
 \end{center}\vspace{-0.1in}
-\caption{Histogram of one million experiments.} \label{figure:link3d}
-\smallskip
-The top group is the
+\caption{Technical diagram of mathematical equations.} \label{fig:link3d}
+% \smallskip
+% caption can go here...
 % } 
-\end{figure*}
+\end{figure}
+
+A gold cartridge that contained ROMs and a little swallowable battery
+to keep the onboard SRAM powered up so that it could retain your epic
+save game. A battery designed to last 70~years. Nothing could cause
+you to lose your save game, even once you were half way through the
+Second Quest. Unless your little brother starts a completely new game
+and saves over your slot, earning him one of the most righteously
+deserved clobberings this side of Inigo Montoya. Saves right on top of
+your slot with a player called just \verb+A +. Right over your slot,
+erasing it, and hasn't even picked up the {\it sword} yet. All you
+have to do is step on the black square in the first room and the guy
+gives you the sword and then dies. Jeez pooleaze. Saves right over
+yours with no sword even though there are TWO OTHER UNUSED SLOTS and
+you even wisely put your game in the third slot {\it exactly to avoid
+  this kind of calamity}.
+
+\medskip
+Need I say more? Shall I say it? Apparently so. KIDS THESE DAYS. Kids
+these days and their three dimensions. You can certainly make a
+respectable case that the Zeldest of the series was {\it Legend of
+  Zelda III: A Link To The Past}, or even {\it Legend of Game Boy
+  Zelda: I forgot What It's Called}. But there seems to be an
+established consensus that {\it Legend of Zelda LXIV: The Ocarina of
+  Time} is the true masterpiece; that while the NES Original indeed
+invented the genre, the technical limitations of the system
+handicapped it. You're talking to me about technical {\it frockin'}
+limitations? Like somehow appreciating the subtle wonder of the
+universe and its mechanics requires me to look at 3D tree people with
+tetrahedric clubs for hands. And they say this of the 64\th\ game in
+the series? Please jouize.
+
+A fine game, don't get me wrong. But was its theme song composed in
+one day? Did it come in a gold cartridge?
+
+And the thing is: The original Zelda was built with enough forethought
+that it {\em can} be rendered in first-person 3D. So here I will make
+the game 3D so that it can be enjoyed by these kids and they can get
+off my lawn, tapdancing around playing their titular Ocarinas. I will
+make it possible to play the game in glorious HD, or even 4K
+resolution, and then some (Figure~\ref{fig:link3d}). And while
+we're at it, why not try to do it in a way that can play loads of NES
+games in three {\it fr\"ockin'} dimensions? What's the worst that
+could happen?
+
+\section{Technical limitations}
+
+As presaged, the NES does have certain technical limitations, which I
+prefer to think of as constraints. Some basic details are important
+for understanding how this works.
+
+The NES has a modest CPU that executes game logic. It's a little 8-bit
+baby puppy with access to 2 kilobytes of RAM. If you think about it,
+the NES's 256x240 pixel screen is already 61,440 pixels, which if it
+used the entire 2k of RAM to represent it, we'd have only
+$0.2\overline{6}$ bits of RAM per pixel, which makes no sense. At XXX
+MhZ, and a generous 1 instruction per pixel, we'd only be able to render
+XXX frames per second! This doesn't make sense, and this is because
+the CPU does not render the graphics on a NES.
+
+The NES also contains a custom Picture Processing Unit, or PPU, the
+8-bit baby puppy's baby ppupy brother. This thing outputs 60 frames of
+NTSC or PAL video per second. In fact, since the PPU also drives a CPU
+interrupt for each scanline and vblank (XXX true?), and most games use
+this for timing,%
+%
+\footnote{Fun fact: NTSC, the video standard used in the Americas, has
+  % XXX how u make ``approx'' symbol
+  a \~60Hz frame rate. But PAL, used in Europe and Asia, has a 50Hz
+  refresh rate. This means that many games run 16\% slower in Europe
+  than the US, so if you grew up playing there, you were playing on
+  {\em Easy} mode. {\tt\ ;-)}}
+%
+in many ways the CPU is subservient to the PPU, kind of like the PPU
+is the {\it true} pure heart of the NES. In the context of this paper, you
+could say that the CPU is the Ocarina of Time and the PPU is the
+Legend of Zelda. It's just an analogy. The PPU is very complicated and
+has tetrahedra for hands, like clubs. The CPU communicates with
+the PPU not by drawing pixels, but by giving a high-level description
+of the screen (by writing it into PPU memory). The chief insight of
+this work is that the PPU representation can be interpreted and
+rendered in 3D instead of 2D. This allows us to use the exact original
+game logic and only change the way it's viewed. Sometimes this even
+allows us to see aspects of the game that are actually present but
+were not visible on NES hardware. Often it makes the game more
+difficult by making things invisible (enemies behind the player or
+behind walls) and inducing motion sickness.
+
+\subsubsection{Background}
+The PPU has two major drawing facilities: Background and Sprites. Both
+are made of ``tiles'', which are 8x8 graphics that are
+usually\footnote{Some games trick the NES hardware by remapping tile
+memory (e.g. in response to a memory read), even during a scanline. I
+do not allow for such chicanery in this work.} stored in ROM. There
+are two tile sets, each with 256 tiles. Figure~\ref{fig:ppu} shows the
+tile sets for the starting location in the Zelda overworld.
+
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[width=\linewidth]{ppu1} \\
+\vspace{1em}
+\includegraphics[width=\linewidth]{ppu2}
+\end{center}\vspace{-0.1in}
+\caption{Pattern tables for the Zelda overworld. In Zelda, the top tiles
+are used for sprites, and the bottom are used for the background graphics.
+All tiles are 8x8 and use 4 color values (color 0 is transparent). These
+color values index into 4 sprite palettes and 4 background palettes, with
+some limitations.
+} \label{fig:ppu}
+\end{figure}
+
+The NES background is described by the ``nametable'', which is an
+array of 32x30 bytes giving the tile numbers to fill the screen with.
+It's two rows shy of 256 probably for some reason having to do with
+the aspect ratio of TVs or number of actual NTSC scanlines, but also
+to make room for 64 bytes of attribute data. There are just two bits
+per four tiles. These two bits select which of the four palettes to
+use for each 2x2 group of tiles on the screen. It's an interesting hobby to
+look for the tricks that artists use to work around the constraint on
+the number of background colors. % XXX figure showing this off?
+
+Because of this array of tiles, NES games have to be built of
+high-level, block-like structures. In Zelda, this structure is
+particularly (but not unusually) regular. Since each 2x2 tile group
+(16x16 pixels) shares the same palette, the screens are built around
+16x16 pixel grid-aligned blocks.
+
+\subsubsection{Sprites} \label{sec:sprites}
+
+Sprites, named in honor of the forest pixies whose homes were razed in
+order to make space inside the PPU for them, are more complicated.
+
+There are 64 sprites, which are all turned on or off together. Each
+sprite is described by four bytes: Its $x$ and $y$ coordinates (these
+can be placed at arbitrary integer pixel locations), the index of
+the tile to draw, and an attribute byte:
+
+\begin{tabular}{rrl}
+    & bit \# & bit deets \\
+\hline \\
+MSB & 7      & Vertical flip \\
+    & 6      & Horizontal flip \\
+    & 5      & something weird \\
+    & 4      & I don't know \\
+    & 3      & unused? who can say \\
+    & 2      & bit 2; could be anything \\
+LSB & 1,0    & Palette index \\
+\end{tabular}
+
+Additionally, there are some global sprite settings. The only one that
+is important for this presentation is the ``tall sprites'' flag. If
+it's set, then sprites are 8x16 (drawn of two consecutive tiles)
+instead of 8x8. Sprites are drawn in a particular order. Barring some
+bizarre mischief, all of the sprites are drawn; you can't turn them
+off individually. Instead, you position the ones you don't want off
+the screen or make them be completely transparent. In order to make
+things more challenging, if there are exactly 8 sprites on a
+particular scan line, some really weird stuff happens. % XXX explain
+
+Sprite Zero has some special properties. I don't want to tell you
+about the properties, I just wanted to make a soda joke.
+
+\section{Three to D---$u$ and $v$}\footnote{Spoiler alert because I think
+this joke may be too abstruse. ``Free to Be You and Me'' was a
+semi-famous kids record (here meaning actual vinyl) by 
+% XXX
+Milo Thomas. D is ``dimensions.'' $u$ and $v$ are common names of
+the axes for texture coordinates in computer graphics.}
+
+The game runs in an emulator, a version of FCEUX\cite{fceux} that I
+have hacked up almost beyond recognition. In normal operation, it
+
+\begin{itemize}
+\item Emulates a complete frame of the NES. This includes the CPU and
+  PPU operation for all 240 scanlines. The execution of a frame depends
+  on the current controller input (which doesn't change as it runs);
+\item Extracts scene geometry and textures. This consists of two
+  pieces: Boxes and Sprites;
+\item Extracts the player location and orientation;
+\item Moves texture data to the GPU;
+\item Transforms the scene geometry into a textured model, sets the camera
+  based on the player location, and renders.
+\end{itemize}
+
+I'll describe the interesting parts of these.
+
+\paragraph{Emulation.} 
+Well, this is not that interesting. I give the emulator its controller
+input and it does its thing, modifying the RAM and PPU and CPU state.
+I have been working on improvements to emulator technology but these
+are independent of this paper and better described elsewhere.
+
+\paragraph{Extracting boxes.} 
+Yeah! Okay! Boxes are the three-dimensional version of squares, and
+are what give the game its 3Dness. I read the nametable in chunks of
+2x2 tiles, which is the size of one overworld tree (and the same size
+as Link). Recall that Zelda uses this grid size, probably because
+color attributes can only be set at this scale. These become $2x2x2$
+cubes. The $x$ and $y$ coordinates are obvious (pixel positions
+divided by 16, since each 2x2 tile block is 16x16 pixels), but what
+should $z$ be? Zelda and other games tend to use tiles quite
+consistently to represent walls, obstacles, floor, empty space, and so
+on. This makes sense because a tree graphic, for example, suggests
+something to the player. So we simply build a file,
+\verb+zelda.tiles+, which maps tiles (by index) to one of \verb+wall+,
+\verb+floor+, \verb+rut+, or \verb+unknown+. Floor blocks are placed
+at $z=0$, which will be under the player's feet (we typically only see
+the top face). Walls are placed at $z=2$, so that their bottoms are
+coplanar with the floor. Rut blocks are placed at $z=-0.5$; this is
+used for blocks that are part of the ``floor'' but that can't be
+walked on, like water. Because each box is actually made of 4 tiles,
+we just use the maximum $z$ value for the constituent tiles (they should
+normally be the same). It would make sense to develop heuristics for
+when games use tile-level resolution, even to generate primitives
+other than cubes.
+
+This approach is a bit unsatisfying because it requires us to have
+specific knowledge of the game as well as do some manual labor.
+However, there really aren't that many tiles, and it's kind of fun.
+(In fact, you really only need to pick out a handful of floor tiles
+before the game is totally playable.) There is another problem, which
+is that many games switch tile sets, which may change whether tile
+\verb+0x84+ (``desert dots'' in the overworld) is floor or wall. What
+happens if you go into a dungeon in Zelda, for example? I didn't even
+try that yet but I bet it's nuts! This can easily be fixed by
+identifying distinct tile sets (for example by using the source ROM
+address, or a hash of their content) and mapping them all.
+
+However, there's some hope that this can be done in a much more
+automatic way. I describe an idea in Section~\ref{sec:cast}.
+
+\paragraph{Box textures.}
+The game is pretty understandable with just TRON-cubes, but it is
+better to texture them. We put the same texture on each face of a box;
+that texture is just the 16x16 pixel graphic consisting of the 4 tiles
+that made it, in the appropriate palette. It does {\it not} work to
+generate a bunch of tile-sized boxes (with the individual tile texture
+on all sides); these will only look right on the top face. Indeed, it
+is possible to get weird texture mismatches between these 2x2x2 boxes
+for basically the same reason. It usually looks pretty good, though.
+
+Conceivably, every box on the screen could have a different texture,
+and conceivably these all change every frame. It is inefficient to
+move 960 distinct textures to video memory every frame, so instead we
+actually just move the rendered background (256x240) itself and record
+with each box its texture coordinates within. This is {\it much}
+faster, though it can create artifacts at the edges of boxes as an
+adjacent pixel peeks into the texture. I checked out Ocarina of Time
+and it is {\it loaded} with texture seam issues, so I figure this is
+how they like it.
+
+\paragraph{Extracting sprites.}
+Sprites are the most difficult. Aside from just being more fiddly to
+implement, it is tricky to recover an important high-level structure.
+Although we know the location of each individual sprite and the two tiles
+(Zelda uses ``tall' 8x16 sprites; see Section~\ref{sec:sprites}) that
+make it up, most game objects (like enemies) are in fact drawn with
+{\it two} sprites. These sprites could be in any of the 64 sprite
+slots (it's not at all obvious how Zelda allocates these, or that it
+would be similar for other games). Adding to the complexity, sprites
+like Link's sword or the pulsing balls of plasma that frogmen shoot
+at you from the water {\it just for being an elf} are just a single
+sprite, and bosses can be like XXX sprites (XXX add figure). If we
+promote sprites to 3D without knowing which ones are supposed to be
+part of the same object, then the enemies get cut up along their
+sprite seems and it looks super wrong.
+
+To fix this, we perform {\it sprite fusion}. We say that two sprites
+are part of the same object if they share a complete edge
+(transitively). Since this allows the construction of very tall ($64
+\times 16$) or wide ($64 \times 8$) sprites, we have to make sure that
+our textures are large enough to contain large sprites. Rather than
+try to pack them into a texture at runtime (computationally expensive
+since they can also be weird shapes), we just pre-allocate one texture
+of the maximum size for each sprite slot (16MB video ram; nothing).
+Sprite fusion, computation of the sprite size, drawing the sprites,
+extracting the geometry, and moving the textures to the video card are
+all intertwined. The simplified algorithm is
+
+\begin{itemize}
+\item Create 64 ``pre-sprites'', which are baby sprites. These know
+  their coordinates and have the index of a ``parent'' pre-sprite
+  (initialized to $-1$).
+\item While filling these in, consult a hash-map to see if there are
+  already pre-sprites exactly aligned in each of the cardinal
+  directions. If so, use a union-find--like algorithm (XXX cite)
+  to efficiently update its parent index; each fused sprite has one
+  canonical representative whose parent index is $-1$. Add the
+  pre-sprite to the hash-map.
+\item Another pass pushes the min/max bounds to the canonical representative,
+  so that we know how big the fused sprite is.
+\item Another pass draws all sprites into the texture data for their
+  canonical representative, at the appropriate location. This pass must
+  happen in the same order as on the NES hardware, so that overlay effects
+  are correct.
+\item Finally, create proper sprites to return for each of the fused sprites.
+  Copy the used region of the texture data for these to the video card.
+\end{itemize}
+
+% XXX diagram of this algorithm would be nice
+
+Two types of sprite rendering are supported: in-plane and billboard.
+In-plane just draws the sprites on the floor or some other $xy$ plane
+of your choice. (This does not actually need sprite fusion, but it is
+not harmful.) Billboard sprites are drawn so that they face the
+player, like in Wolfenstein~3D. These are positioned at their $xy$
+centroid and positioned in $z$ so that the bottom of the sprite is
+touching the floor.
+
+\paragraph{Extracting location.}
+
+Most NES games use very simple layouts for their game facts (i.e.,
+global variables at fixed memory locations), since the processor is
+simple and memory light.\cite{murphy2013first} In Zelda, Link's
+$x$ coordinate is at \verb+0x70+ and $y$ at \verb+0x84+. Since this
+is his top-left corner, we add 8 to each to get to the middle of the
+16x16 pixel boy.
+
+We also have to face in the same direction as link, to make the
+perspective first-person. Link's orientation is at memory location
+\verb+0x98+.
+
+Link moves smoothly (with pixel resolution) around the screen, but
+turns abruptly. This is sickening to watch. I apply some smoothing
+to Link's angle, ...
+
+(This also involves manual steps..)
+
+\paragraph{Rendering.}
+
+......
+
+Note that some of the tiles on screen are used to draw a menu at the
+top, which is not part of the gameplay area. This still gets interpreted
+as
+
+Problem with link always hovering behind you
+
+\section{A Link to the cast} \label{sec:cast}
+
+\nocite{murphy2013first}
 

### 2016-03-14 trunk/smeight/paper/paper.tex
so sleepy

@@
 the gold cartridge down to make gold teeth grilles, after carefully
-extracting the even more precious ROM inside. A die-cut hole in the
+extracting and preserving the even more precious ROM inside. A die-cut hole in the
 box so that you could get a peek of the cartridge and presage that you
@@
 
-\begin{figure}[ht]
+\begin{figure}[h!t]
 \begin{center}
@@
 these days and their three dimensions. You can certainly make a
-respectable case that the Zenith of the series was {\it Legend of
+respectable case that the Zeldest of the series was {\it Legend of
   Zelda III: A Link To The Past}, or even {\it Legend of Game Boy
   Zelda: I forgot What It's Called}. But there seems to be an
-established consensus that {\it Legend Of Zelda LXIV: The Ocarina of
+established consensus that {\it Legend of Zelda LXIV: The Ocarina of
   Time} is the true masterpiece; that while the NES Original indeed
@@
 limitations? Like somehow appreciating the subtle wonder of the
-universe and its mechanics requires me to look at tree people with
-tetrahedra clubs for hands. And they say this of the 64\th\ game in
+universe and its mechanics requires me to look at 3D tree people with
+tetrahedric clubs for hands. And they say this of the 64\th\ game in
 the series? Please jouize.
@@
 
+\begin{figure}[h!t]
+\begin{center}
+\includegraphics[width= \linewidth]{ppu1} \\
+\vspace{1em}
+\includegraphics[width= \linewidth]{ppu2}
+\end{center}\vspace{-0.1in}
+\caption{Pattern tables for the Zelda overworld. In Zelda, the top tiles
+are used for sprites, and the bottom are used for the background graphics.
+All tiles are 8x8 and use 4 color values (color 0 is transparent). These
+color values index into 4 sprite palettes and 4 background palettes, with
+some limitations.
+} \label{fig:ppu}
+\end{figure}
+
 As presaged, the NES does have certain technical limitations, which I
@@
 the NES's 256x240 pixel screen is already 61,440 pixels, which if it
-used the entirety of RAM to represent it, we'd have only
-$0.2\overline{6}$ bits of RAM per pixel, which makes no sense. At XXX
-MhZ, and a generous 1 instruction per pixel, we'd only be able to render
-XXX frames per second! This doesn't make sense, and this is because
-the CPU does not render the graphics on a NES.
+used the entire 2k of RAM to represent it, we'd have only
+$0.2\overline{6}$ bits of RAM per pixel, which makes no sense. At
+1.7~MhZ, and a generous 1 instruction per pixel, we'd only be able to
+render 28 frames per second! This doesn't make sense, and this is
+because the CPU does not render the graphics on a NES.
 
-The NES also contains a custom Picture Processing Unit---PPU---which
-outputs 60 frames of NTSC or PAL video per second. In fact, since the
-PPU also drives a CPU interrupt for each scanline and vblank (XXX
-true?), and most games use this for timing,%
+The NES also contains a custom Picture Processing Unit, or PPU, the
+8-bit baby puppy's baby ppupy brother. This thing outputs 60 frames of
+NTSC video per second. In fact, since the PPU also drives a CPU
+interrupt for each scanline and vblank, and most games use
+this for timing,%
 %
-\footnote{Fun fact: NTSC, used in the Americas, has a ~60Hz frame
-  rate. But PAL, used in Europe and Asia, has a 50Hz refresh rate.
-  This means that many games run 16\% slower in Europe than the US, so
-  if you grew up playing there, you were playing on {\em Easy} mode.
-  {\tt\ ;-)}}
+\footnote{Fun fact: NTSC (``Never Twice the Same Color''), the video
+  standard used in the Americas, has a $\sim 60$ Hz frame rate. But
+  PAL (``Pointless and Lousy''), used in Europe and Asia, has a $\sim
+  50$ Hz refresh rate. This means that many games run 16\% slower in
+  Europe than the US, so if you grew up playing there, you were
+  playing on {\em Easy} mode. {\tt\ ;-)}}
 %
 in many ways the CPU is subservient to the PPU, kind of like the PPU
-is the {\it real} heart of the NES. In the context of this paper, you
+is the {\it true} pure heart of the NES. In the context of this paper, you
 could say that the CPU is the Ocarina of Time and the PPU is the
 Legend of Zelda. It's just an analogy. The PPU is very complicated and
-has tetrahedra for hands, like clubs. But the CPU communicates with
+has tetrahedra for hands, like clubs. The CPU communicates with
 the PPU not by drawing pixels, but by giving a high-level description
@@
 difficult by making things invisible (enemies behind the player or
-behind walls) or inducing motion sickness.
+behind walls) and inducing motion sickness.
 
@@
 
-\begin{figure}[ht]
-\begin{center}
-\includegraphics[width=\linewidth]{ppu1} \\
-\vspace{1em}
-\includegraphics[width=\linewidth]{ppu2}
-\end{center}\vspace{-0.1in}
-\caption{Pattern tables for the Zelda overworld. In Zelda, the top tiles
-are used for sprites, and the bottom are used for the background graphics.
-All tiles are 8x8 and use 4 color values (color 0 is transparent). These
-color values index into 4 sprite palettes and 4 background palettes, with
-some limitations.
-} \label{fig:ppu}
-\end{figure}
-
 The NES background is described by the ``nametable'', which is an
@@
 per four tiles. These two bits select which of the four palettes to
-use for each 2x2 group of tiles on the screen. It's interesting to
+use for each 2x2 group of tiles on the screen. It's an interesting hobby to
 look for the tricks that artists use to work around the constraint on
@@
 
-\subsubsection{Sprites}
+\subsubsection{Sprites} \label{sec:sprites}
 
@@
 
-Additionally, there are some global sprite settings. But
+Additionally, there are some global sprite settings. The only one that
+is important for this presentation is the ``tall sprites'' flag. If
+it's set, then sprites are 8x16 (drawn of two consecutive tiles)
+instead of 8x8. Sprites are drawn in a particular order. Barring some
+bizarre mischief, all of the sprites are drawn; you can't turn them
+off individually. Instead, you position the ones you don't want off
+the screen or make them be completely transparent. In order to make
+things more challenging, if there are exactly 8 sprites on a
+particular scan line, some really weird stuff happens. % XXX explain
 
@@
 about the properties, I just wanted to make a soda joke.
+
+\section{Three to D---$u$ and $v$}\footnote{Spoiler alert because I think
+this joke may be too abstruse. ``Free to Be You and Me'' was a
+semi-famous kids record (here meaning actual vinyl) by 
+Marlo~Thomas. D is ``dimensions.'' $u$ and $v$ are common names of
+the axes for texture coordinates in computer graphics.}
+
+The game runs in an emulator, a version of FCEUX\cite{fceux} that I
+have hacked up almost beyond recognition. In normal operation, it
+
+\begin{itemize}
+\item Emulates a complete frame of the NES. This includes the CPU and
+  PPU operation for all 240 scanlines. The execution of a frame depends
+  on the current controller input (which doesn't change as it runs);
+\item Extracts scene geometry and textures. This consists of two
+  pieces: Boxes and Sprites;
+\item Extracts the player location and orientation;
+\item Moves texture data to the GPU;
+\item Transforms the scene geometry into a textured model, sets the camera
+  based on the player location, and renders.
+\end{itemize}
+
+I'll describe the interesting parts of these.
+
+\paragraph{Emulation.} 
+Well, this is not that interesting. I give the emulator its controller
+input and it does its thing, modifying the RAM and PPU and CPU state.
+I have been working on improvements to emulator technology but these
+are independent of this paper and better described elsewhere.
+
+\paragraph{Extracting boxes.} 
+Yeah! Okay! Boxes are the three-dimensional version of squares, and
+are what give the game its 3Dness. I read the nametable in chunks of
+2x2 tiles, which is the size of one overworld tree (and the same size
+as Link). Recall that Zelda uses this grid size, probably because
+color attributes can only be set at this scale. These become $2x2x2$
+cubes. The $x$ and $y$ coordinates are obvious (pixel positions
+divided by 16, since each 2x2 tile block is 16x16 pixels), but what
+should $z$ be? Zelda and other games tend to use tiles quite
+consistently to represent walls, obstacles, floor, empty space, and so
+on. This makes sense because a tree graphic, for example, suggests
+something to the player. So we simply build a file,
+\verb+zelda.tiles+, which maps tiles (by index) to one of \verb+wall+,
+\verb+floor+, \verb+rut+, or \verb+unknown+. Floor blocks are placed
+at $z=0$, which will be under the player's feet (we typically only see
+the top face). Walls are placed at $z=2$, so that their bottoms are
+coplanar with the floor. Rut blocks are placed at $z=-0.5$; this is
+used for blocks that are part of the ``floor'' but that can't be
+walked on, like water. Because each box is actually made of 4 tiles,
+we just use the maximum $z$ value for the constituent tiles (they should
+normally be the same). It would make sense to develop heuristics for
+when games use tile-level resolution, even to generate primitives
+other than cubes.
+
+This approach is a bit unsatisfying because it requires us to have
+specific knowledge of the game as well as do some manual labor.
+However, there really aren't that many tiles, and it's kind of fun.
+(In fact, you really only need to pick out a handful of floor tiles
+before the game is totally playable.) There is another problem, which
+is that many games switch tile sets, which may change whether tile
+\verb+0x84+ (``desert dots'' in the overworld) is floor or wall. What
+happens if you go into a dungeon in Zelda, for example? I didn't even
+try that yet but I bet it's nuts! This can easily be fixed by
+identifying distinct tile sets (for example by using the source ROM
+address, or a hash of their content) and mapping them all.
+
+However, there's some hope that this can be done in a much more
+automatic way. I describe an idea in Section~\ref{sec:cast}.
+
+\paragraph{Box textures.}
+The game is pretty understandable with just TRON-cubes, but it is
+better to texture them. We put the same texture on each face of a box;
+that texture is just the 16x16 pixel graphic consisting of the 4 tiles
+that made it, in the appropriate palette. It does {\it not} work to
+generate a bunch of tile-sized boxes (with the individual tile texture
+on all sides); these will only look right on the top face. Indeed, it
+is possible to get weird texture mismatches between these 2x2x2 boxes
+for basically the same reason. It usually looks pretty good, though.
+
+Conceivably, every box on the screen could have a different texture,
+and conceivably these all change every frame. It is inefficient to
+move 960 distinct textures to video memory every frame, so instead we
+actually just move the rendered background (256x240) itself and record
+with each box its texture coordinates within. This is {\it much}
+faster, though it can create artifacts at the edges of boxes as an
+adjacent pixel peeks into the texture. I checked out Ocarina of Time
+and it is {\it loaded} with texture seam issues, so I figure this is
+how they like it.
+
+\paragraph{Extracting sprites.}
+Sprites are the most difficult. Aside from just being more fiddly to
+implement, it is tricky to recover an important high-level structure.
+Although we know the location of each individual sprite and the two tiles
+(Zelda uses ``tall' 8x16 sprites; see Section~\ref{sec:sprites}) that
+make it up, most game objects (like enemies) are in fact drawn with
+{\it two} sprites. These sprites could be in any of the 64 sprite
+slots (it's not at all obvious how Zelda allocates these, or that it
+would be similar for other games). Adding to the complexity, sprites
+like Link's sword or the pulsing balls of plasma that frogmen shoot
+at you from the water {\it just for being an elf} are just a single
+sprite, and bosses can be like XXX sprites (XXX add figure). If we
+promote sprites to 3D without knowing which ones are supposed to be
+part of the same object, then the enemies get cut up along their
+sprite seems and it looks super wrong.
+
+To fix this, we perform {\it sprite fusion}. We say that two sprites
+are part of the same object if they share a complete edge
+(transitively). Since this allows the construction of very tall ($64
+\times 16$) or wide ($64 \times 8$) sprites, we have to make sure that
+our textures are large enough to contain large sprites. Rather than
+try to pack them into a texture at runtime (computationally expensive
+since they can also be weird shapes), we just pre-allocate one texture
+of the maximum size for each sprite slot (16MB video ram; nothing).
+Sprite fusion, computation of the sprite size, drawing the sprites,
+extracting the geometry, and moving the textures to the video card are
+all intertwined. The simplified algorithm is
+
+\begin{itemize}
+\item Create 64 ``pre-sprites'', which are baby sprites. These know
+  their coordinates and have the index of a ``parent'' pre-sprite
+  (initialized to $-1$).
+\item While filling these in, consult a hash-map to see if there are
+  already pre-sprites exactly aligned in each of the cardinal
+  directions. If so, use a union-find--like algorithm\cite{tarjan1975efficiency}
+  to efficiently update its parent index; each fused sprite has one
+  canonical representative whose parent index is $-1$. Add the
+  pre-sprite to the hash-map.
+\item Another pass pushes the min/max bounds to the canonical representative,
+  so that we know how big the fused sprite is.
+\item Another pass draws all sprites into the texture data for their
+  canonical representative, at the appropriate location. This pass must
+  happen in the same order as on the NES hardware, so that overlay effects
+  are correct.
+\item Finally, create proper sprites to return for each of the fused sprites.
+  Copy the used region of the texture data for these to the video card.
+\end{itemize}
+
+% XXX diagram of this algorithm would be nice
+
+Two types of sprite rendering are supported: in-plane and billboard.
+In-plane just draws the sprites on the floor or some other $xy$ plane
+of your choice. (This does not actually need sprite fusion, but it is
+not harmful.) Billboard sprites are drawn so that they face the
+player, like in Wolfenstein~3D. These are positioned at their $xy$
+centroid and positioned in $z$ so that the bottom of the sprite is
+touching the floor.
+
+\paragraph{Extracting location.}
+
+Most NES games use very simple layouts for their game facts (i.e.,
+global variables at fixed memory locations), since the processor is
+simple and memory light.\cite{murphy2013first} In Zelda, Link's
+$x$ coordinate is at \verb+0x70+ and $y$ at \verb+0x84+. Since this
+is his top-left corner, we add 8 to each to get to the middle of the
+16x16 pixel boy.
+
+We also have to face in the camera in the same direction as Link, to
+make the perspective first-person. Link's orientation is at memory
+location \verb+0x98+.
+
+Link moves smoothly (with pixel resolution) around the screen, but
+turns abruptly. This is sickening to watch. I apply some smoothing to
+Link's angle, by treating the location from memory as a ``target''
+angle and twisting towards it at a limited speed.
+
+Knowing what memory locations hold the player's position and
+orientation also involves some manual work. You could imagine
+heuristics for trying to determine this, by seeing which bytes tend to
+go up when pressing right on the controller, and so on. It's easy to
+find these values manually or to even look them up for popular games.
+
+\paragraph{Rendering.}
+
+Finally, we can render the scene. I used ``classic'' OpenGL interface,
+even though it is terrible, for the sake of making good on the
+excellent pun that is this paper's title. The final result looks like
+Figure~\ref{fig:alone}.
+
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[width=\linewidth]{alone}
+\end{center}\vspace{-0.1in}
+\caption{The first cave, where this man gives you a sword and then dies,
+rendered in 3D. Letters are rendered as walls so you can see 'em upright,
+but then of course the second line is in front of the first. The 2D NES
+screen is superimposed on the top left for comparison.
+} \label{fig:alone}
+\end{figure}
+
+......
+
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[width=\linewidth]{offscreen}
+\end{center}\vspace{-0.1in}
+\caption{
+XXX write description
+} \label{fig:offscreen}
+\end{figure}
+
+Note that some of the tiles on screen are used to draw a menu at the
+top, which is not part of the gameplay area. This still gets
+interpreted as part of the geometry of the world, of course
+(Figure~\ref{fig:offscreen}). It would be a simple matter to crop the
+region of interest, and natural to even overlay it on a heads-up
+display. But I like it over there. Why shouldn't menus be 3D as well?
+
+Problem with link always hovering behind you (Figure~\ref{fig:megalink}).
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[width=\linewidth]{megalink}
+\end{center}\vspace{-0.1in}
+\caption{look who's behind u
+} \label{fig:megalink}
+\end{figure}
+
+\section{Other games}
+
+\section{A Link to the cast} \label{sec:cast}
 

### 2016-03-15 smeight/paper/paper.tex
First submitted version

@@
 used the entire 2k of RAM to represent it, we'd have only
-$0.2\overline{6}$ bits of RAM per pixel, which makes no sense. At XXX
-MhZ, and a generous 1 instruction per pixel, we'd only be able to render
-XXX frames per second! This doesn't make sense, and this is because
-the CPU does not render the graphics on a NES.
+$0.2\overline{6}$ bits of RAM per pixel, which makes no sense. At
+1.7~MhZ, and a generous 1 instruction per pixel, we'd only be able to
+render 28 frames per second! This doesn't make sense, and this is
+because the CPU does not render the graphics on a NES.
 
@@
 8-bit baby puppy's baby ppupy brother. This thing outputs 60 frames of
-NTSC or PAL video per second. In fact, since the PPU also drives a CPU
-interrupt for each scanline and vblank (XXX true?), and most games use
+NTSC video per second. In fact, since the PPU also drives a CPU
+interrupt for each scanline and vblank, and most games use
 this for timing,%
 %
-\footnote{Fun fact: NTSC, the video standard used in the Americas, has
-  % XXX how u make ``approx'' symbol
-  a \~60Hz frame rate. But PAL, used in Europe and Asia, has a 50Hz
-  refresh rate. This means that many games run 16\% slower in Europe
-  than the US, so if you grew up playing there, you were playing on
-  {\em Easy} mode. {\tt\ ;-)}}
+\footnote{Fun fact: NTSC (``Never Twice the Same Color''), the video
+  standard used in the Americas, has a $\sim\!\!60$ Hz frame rate. But
+  PAL (``Pointless and Lousy''), used in Europe and Asia, has a $\sim\!\!50$
+  Hz refresh rate. This means that many games run 16\% slower in
+  Europe than the US, so if you grew up playing there, you were
+  playing on {\em Easy} mode. {\tt\ ;-)}}
 %
@@
 tile sets for the starting location in the Zelda overworld.
-
-\begin{figure}[ht]
-\begin{center}
-\includegraphics[width=\linewidth]{ppu1} \\
-\vspace{1em}
-\includegraphics[width=\linewidth]{ppu2}
-\end{center}\vspace{-0.1in}
-\caption{Pattern tables for the Zelda overworld. In Zelda, the top tiles
-are used for sprites, and the bottom are used for the background graphics.
-All tiles are 8x8 and use 4 color values (color 0 is transparent). These
-color values index into 4 sprite palettes and 4 background palettes, with
-some limitations.
-} \label{fig:ppu}
-\end{figure}
 
@@
     & bit \# & bit deets \\
-\hline \\
+\hline
 MSB & 7      & Vertical flip \\
@@
 
-\section{Three to D---$u$ and $v$}\footnote{Spoiler alert because I think
+\section[Three to D---$u$ and $v$]{Three to D---$u$ and $v$\footnote{Spoiler alert because I think
 this joke may be too abstruse. ``Free to Be You and Me'' was a
 semi-famous kids record (here meaning actual vinyl) by 
-% XXX
-Milo Thomas. D is ``dimensions.'' $u$ and $v$ are common names of
-the axes for texture coordinates in computer graphics.}
+Marlo~Thomas. D is ``dimensions.'' $u$ and $v$ are common names of
+the axes for texture coordinates in computer graphics.}}
 
@@
 at you from the water {\it just for being an elf} are just a single
-sprite, and bosses can be like XXX sprites (XXX add figure). If we
+sprite, and bosses can be like 6 or more sprites (Figure~\ref{fig:spider}). If we
 promote sprites to 3D without knowing which ones are supposed to be
@@
 sprite seems and it looks super wrong.
+
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[width=\linewidth]{spider}
+\end{center}\vspace{-0.1in}
+\caption{Link's spider nemesis, made of six 8x16 sprites. The sprites
+have adjacent edges, so they end up being fused into one merged sprite.
+} \label{fig:spider}
+\end{figure}
 
@@
   already pre-sprites exactly aligned in each of the cardinal
-  directions. If so, use a union-find--like algorithm (XXX cite)
+  directions. If so, use a union-find--like algorithm\cite{tarjan1975efficiency}
   to efficiently update its parent index; each fused sprite has one
@@
 
-We also have to face in the same direction as link, to make the
-perspective first-person. Link's orientation is at memory location
-\verb+0x98+.
+We also have to face in the camera in the same direction as Link, to
+make the perspective first-person. Link's orientation is at memory
+location \verb+0x98+.
 
 Link moves smoothly (with pixel resolution) around the screen, but
-turns abruptly. This is sickening to watch. I apply some smoothing
-to Link's angle, ...
+turns abruptly. This is sickening to watch. I apply some smoothing to
+Link's angle, by treating the location from memory as a ``target''
+angle and twisting towards it (along the shortest angular path) at a
+limited speed.
 
-(This also involves manual steps..)
+Knowing what memory locations hold the player's position and
+orientation also involves some manual work. You could imagine
+heuristics for trying to determine this, by seeing which bytes tend to
+go up when pressing right on the controller, and so on. However, it's
+easy to find these values manually or to even look them up for popular
+games.
 
@@
 
-......
+\begin{figure*}[t]
+\begin{center}
+\includegraphics[width=\linewidth]{bridgewide}
+\end{center}\vspace{-0.1in}
+\caption{A representative view of the Legend of Zelda in glorious
+  3D.} \label{fig:bridgewide}
+\end{figure*}
 
-Note that some of the tiles on screen are used to draw a menu at the
-top, which is not part of the gameplay area. This still gets interpreted
-as
+Finally, we can render the scene. I used ``classic'' OpenGL interface,
+even though it is terrible, for the sake of making good on the
+excellent pun that is this paper's title. The final result looks like
+Figures~\ref{fig:bridgewide} and~\ref{fig:alone}.
 
-Problem with link always hovering behind you
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[width=\linewidth]{alone}
+\end{center}\vspace{-0.1in}
+\caption{The first cave, where this man gives you a sword and then dies,
+rendered in 3D. Letters are treated as walls so you can see 'em upright,
+but then of course the second line is in front of the first. The 2D NES
+screen is superimposed on the top left for comparison.
+} \label{fig:alone}
+\end{figure}
 
-\section{A Link to the cast} \label{sec:cast}
+Rendering with OpenGL is quite straightforward given the geometry
+extracted from the previous sections. I try to preserve the pixel
+aesthetic by using \verb+GL_TEXTURE_MAG_FILTER+ = \verb+GL_NEAREST+,
+but at this point we can use whatever approach to 3D graphics most
+titillates the youth. Full-screen multisample anti-aliasing? Sure.
+Screen Space Ambient Occlusion? I hear that this creates physically
+realistic shadows on the cheap, sure. Stencil buffer shadows?
+Volumetric fog? God rays? Depth-of-field? Lens flares? You got it.
 
-\nocite{murphy2013first}
+\subsection{Quirks}
+There are some problems or quirks with the current approach.
+
+\paragraph{Extramural geometry.}
+In Zelda, some of the tiles on screen are used to draw a menu at the
+top, which is not part of the gameplay area. This still gets
+interpreted as part of the geometry of the world, of course
+(Figure~\ref{fig:offscreen}). It would be a simple matter to crop the
+region of interest (especially if done manually), and it would even be
+natural to overlay it on a heads-up display. But I like it over there.
+Why shouldn't menus be 3D as well?
+
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[width=\linewidth]{offscreen}
+\end{center}\vspace{-0.1in}
+\caption{ The game menu visible to the North because the ocean does
+  not get in its way. Did you know that if you were able to see beyond
+  the cluttering trees and mountains at the horizon of Earth, you'd
+  see your own status menu in real life? } \label{fig:offscreen}
+\end{figure}
+
+\paragraph{Positive self-image.}
+We place the camera at Link's position and orientation, so that the
+player ``becomes'' Link. However, Link is still represented by a
+sprite in the game, and we do not treat it specially. This means that
+the fused Link billboard is always in the scene geometry right at the
+location of the camera. The upshot is that whenever the camera is
+facing south, an enormous Link graphic fills the entire screen,
+reminding you that he has always been following a few pixels behind
+you, haunting you (Figure~\ref{fig:megalink}). This sprite could be
+rejected by seeing that it is coincident with the player's location
+(which we extract) or by its particular content. We could also set the
+``near plane'' so that nearby sprites are not visible. However, the
+everpresent looming spectre of a ghost mimicking our every move is a
+good reminder of our mortality, so I kept it.
+
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[width=\linewidth]{megalink}
+\end{center}\vspace{-0.1in}
+\caption{Look who's behind you! WE ARE FRIENDS FOREVER {\tt:-) :-)}
+} \label{fig:megalink}
+\end{figure}
+
+\paragraph{Surprising sprite fusion.}
+Sprite fusion works well for merging sprites that are part of the same
+logical object in Zelda. (It may not work as well in other games where
+e.g. a monster's legs are moved independently from its body.) However,
+it occasionally mistakenly fuses two sprites that just incidentally
+share an edge (Figure~\ref{fig:badfusion}). This usually persists for
+just a tiny moment since the sprites are moving. But who's to say that
+as those monsters high-five, they aren't forming a momentary totem?
+Only Shigeru {\it froakin'} Miyamoto can say that isn't what he intended.
+
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[width=\linewidth]{bad-sprite-fusion}
+\end{center}\vspace{-0.1in}
+\caption{Two Octoroks that were incorrectly fused because they have
+constituent sprites that have adjacent edges (the fused sprite shape is
+not rectangular). The formation is a Hexadecirok.
+} \label{fig:badfusion}
+\end{figure}
+
+\paragraph{Texture orientation.} The textures of both backgrounds and sprites
+often have strange and inconsistent perspective ideas in video games.
+For example, the Octorok points its nose in the direction it's moving,
+because it is always viewed ``from above''. However, when Link is
+facing south, you see his front (not the top of his head facing down),
+and when facing west you see his side. This doesn't make sense, and
+makes it impossible to apply any consistent rule for mapping sprites
+into world orientation. This mostly just produces humorous results,
+like when Link throws his sword and it moves away from him in first
+person view: It may be pointing towards the sky, towards the floor, or
+to either side, but never in the direction it's thrown. Even if we
+were to render the sword ``in plane'', its handle should always be
+oriented towards the camera (which would not be the case). No single
+policy will work for all cases, but perhaps heuristics can look for
+the use of the horizontal and vertical flip bits to figure out what
+sprites are rotatable, and then rotate them relative to the camera.
+
+\section{Future work}
+
+This work is literally in progress, and due to Draconian SIGBOVIK
+deadlines, this paper is figuratively already out of date by the time
+you read it. You can check out the project's home page
+
+\medskip
+\begin{center}
+\url{http://tom7.org/zelda}
+\end{center}
+\medskip
+
+and maybe download it and play it, and probably watch a video
+demonstration. There may be some new features. Depends what I
+get to, you know?
+
+\subsection{Control}
+Currently, the game is viewed first person but played with native
+controls; this means that pressing ``up'' always moves link to the
+North, turning the camera if necessary. This is super weird. However,
+it is pretty easy to remap first person controls into native controls,
+as long as there is a way of implementing that intent (for example, it
+is not possible to ``strafe'' in Zelda). We can change the camera
+angle arbitrary, implementing ``mouselook'' or virtual reality
+support. Since the z axis is otherwise unused, we could allow the player
+to jump, though this would do nothing in the game itself.
+
+\subsection{Other games}
+This program was built in such a way that it should take minimal
+effort make other NES games 3D, perhaps with humorous results. There
+are some issues that need to be dealt with. A common type of game is
+the side-scrolling platformer, like Super Mario Bros.. At least two
+important issues are not yet handled: Mario has a different coordinate
+system than Zelda (it is ``side view'' rather than ``top down''), so
+we need to support that rendering style and control/camera scheme. No
+big deal. Second, this game scrolls the screen rather than using
+single-screen layouts; this uses the PPU in a different and somewhat
+more complex way. Many games will use advanced PPU tricks that cannot
+be interpreted in a generic high-level way.
+
+\subsection{A Link to the cast} \label{sec:cast}
+
+One dissatisfying problem with the current approach is that it
+requires some manual effort to map different tile types to geometry.
+One could imagine various heuristics (even a image classifier) to
+provide defaults or to remove the manual input entirely. An exciting
+angle is to use the game's own understanding of these tiles. The main
+thing that differentiates a ``wall'' from a ``floor'' is whether the
+player can walk over it. Clearly somewhere in the game's code it
+``knows'' what is floor and what is wall. We could extract that
+directly, but this would require a different manual step.
+Alternatively, we could determine it experimentally. A simple
+heuristic would keep track of what tiles the player has walked on. But
+we are not limited to what the player actually does; since we are
+emulating, we can also emulate counterfactual scenarios. Imagine that
+as we draw a first person view, we cast lines out from the camera
+towards the geometry, as in a ray tracer. Except that instead of
+tracing light rays until an intersection with scene geometry, we trace
+a hypothetical player path by emulating the player moving in that
+direction, until Link can move no more. This only requires
+understanding the bytes that correspond to the player's location
+(which we already need) and how the controller works.\footnote{Most
+  rays are diagonal. We can implement a north-northwest diagonal by
+  repeatedly moving two pixels north, then one west. We could also set
+  the player's position directly to a putative destination and then
+  move in cardinal directions to test solidness.} Testing contact with
+sprites might give us useful heuristics about how they should be
+rendered as well.
+
+There are many ways we can imagine this going wrong, but probably
+spectacularly so. Interestingly, this would allow us to render parts
+of the game that are not even visible in the PPU yet, since emulating
+the player's path ahead far enough would cause the game to switch
+to future scenes.
+
+\section{Conclusion}
+
+In this paper I made Zelda three-dimensional so that it can be enjoyed
+by the kids. I am fully aware that these ``kids'' are actually also
+full-grown adults who themselves resent a slightly younger generation
+who inexplicably think of Legend of Zelda Twilight Princess HD as the
+canonical episode; these people should get to work at enhancing Ocarina
+of Time to add Amiibo support. While you're at it, get rid of the
+really long section where you have to be a dog. I didn't like that part.
 
